### 2020-11-21 paper/intro.tex

\section{Introduction}

This paper builds on
\cite{2019-sp} and
\cite{10.1145/3428262}.

Pomsets are a nice model of concurrent computation.
Predicate transformers are a nice model of sequential computation.
We show how these interact.

We do this by attaching predicate transformers to \emph{configurations}.

Thus the amount of transformation depends on which read events are seen in
the configuration.

We develop the theory of pomsets with predicate transformers, then apply the
theory to model relaxed memory.

Other things we could model include Map/Reduce and Fork/Join.

### 2020-12-19 paper/arm.tex
some agda questions

\section{Efficient Implementation on ARMv8}
\label{sec:arm}

Alternate characterization of ARM:
\cite[\textsection B2.3.6]{arm-reference-manual}
\cite{alglave-git-alternate}
\cite{armed-cats}

Internal visibility:
If $\bEv\xpoloc\aEv$ then either
\begin{itemize}
\item %$\bEv\xca\aEv$:
  $\bEv:\DWP{x}{1}\xco\aEv:\DWP{x}{2}$ 
\item %$\bEv\xrfx\aEv$:
  $\bEv:\DWP{x}{1}\xrfx\aEv:\DRP{x}{1}$
\item %$\bEv\mathrel{\xco\xrfx}\aEv$:
  $\bEv:\DWP{x}{1}\xco\DWP{x}{2}\xrfx\aEv:\DRP{x}{2}$
\item
  $\bEv:\DRP{x}{0}\xfr\aEv:\DWP{x}{1}$
\item %$\bEv\mathrel{\xfr\xrfx}\aEv$:
  $\bEv:\DRP{x}{0}\xfr\DWP{x}{1}\xrfx\aEv:\DRP{x}{1}$
\item %$\bEv\mathrel{\xrfxinv\xrfx}\aEv$:
  $\bEv:\DRP{x}{1}\xrfxinv\DWP{x}{1}\xrfx\aEv:\DRP{x}{1}$
  % \begin{tikzinline}
  %   \node(a){$\DWP{x}{1}$};
  %   \node(b)[right=1.5em of a]{$\DRP{x}{1}$};
  %   \node(c)[right=1.5em of b]{$\DRP{x}{1}$};
  %   \rfx{a}{b}
  %   \rfx[out=15,in=165]{a}{c}
  % \end{tikzinline}
\end{itemize}
If $\bEv\xpoloc\aEv$ then the following are impossible:
\begin{itemize}
\item $\bEv:\DWP{x}{2}\xcoinv\aEv:\DWP{x}{1}$ 
\item $\bEv:\DWP{x}{1}\xfrinv\aEv:\DRP{x}{0}$ 
\item $\bEv:\DRP{x}{1}\xrfxinv\aEv:\DWP{x}{1}$ 
\item $\bEv:\DRP{x}{2}\xrfxinv\DWP{x}{2}\xcoinv\aEv:\DWP{x}{1}$
\item $\bEv:\DRP{x}{1}\xrfxinv\DWP{x}{1}\xfrinv\aEv:\DRP{x}{0}$
\end{itemize}

\begin{figure*}
\begin{verbatim}
let IM0 = loc & ((IW * (M\IW)) | ((W\FW) * FW))

(* Local read successor *)
let lrs = [W]; po-loc \ intervening-write(po-loc); [R]

(* Local write successor *)
let lws = po-loc; [W]

(* Dependency-ordered-before *)
let dob =  addr | data               
        | (addr | data); lrs
        | addr; po; [W]         |  addr; po; [ISB]; po; [R]
        | ctrl;     [W]         |  ctrl;     [ISB]; po; [R]
our dob = addr; [W]
        | data; [W]
        | ctrl; [W]

(* Barrier-ordered-before *)
let bob = [A | Q]; po            | po; [dmb.full]; po          | po; ([A];amo;[L]); po
        | po; [L]                | [R]; po; [dmb.ld]; po
        | [L]; po; [A]           | [W]; po; [dmb.st]; po; [W]

(* Locally-ordered-before *)                let aob = rmw 
let rec lob = lws | dob | bob | lob; lob            | [W & range(rmw)]; lrs; [A | Q]

(* Internal visibility requirement *)
acyclic po-loc | fr | co | rf as internal

(* External global completion requirement *)
let gc-req = (W * _) | (R * _) & ((range(rfe) * _) | (rfi^-1; lob))
let preorder-gcb = IM0 | lob & gc-req

with gcb from linearisations(M, preorder-gcb)
~empty gcb

let gcbl = gcb & loc
let rf-gcb = (W * R) & gcbl \ intervening-write(gcbl)
let co-gcb = (W * W) & gcbl

call equal(rf, rf-gcb)
call equal(co, co-gcb)

James:
let po-conflict = po-loc \ (R * R)
Claim we can add po-conflict to preorder-gcb without changing things
let preorder-gcb = IM0 | lob & gc-req | po-conflict
\end{verbatim}
  \caption{ARM}
  \label{fig:arm}
\end{figure*}

\textcolor{red}{[This is out of date, from an old submission]}

In this section, we consider the fragment of our language without
restriction.  For simplicity, we allow release and acquire synchronization
but ban fences.  We assume that all memory locations are initialized to $0$
and parallel-composition occurs only at top level.  We take the set of memory
locations to be finite.  In other words, we assume that programs have the
form
\begin{displaymath}
  {\aLoc_1}\GETS{0}\SEMI
  \cdots\SEMI
  {\aLoc_m}\GETS{0}\SEMI
  (\aCmd^1 \PAR \cdots \PAR \aCmd^n)
\end{displaymath}
where $\aCmd^1$, \ldots, $\aCmd^n$ do not include composition, restriction or
fence operations.

Our language can be translated to ARM using \texttt{ldr} for relaxed read,
\texttt{ldar} for acquiring read, \texttt{str} for relaxed write, and
\texttt{stlr} for releasing write.  Relative to the ARM specification, we
have removed loops and read-modify-write (RMW) operations, in addition to
fences\footnote{We leave out fences for simplicity.  Following
  \citet{DBLP:journals/pacmpl/PodkopaevLV19}, our $\FENCE$ instruction can be
  translated to \texttt{dmb.sy}, since it has release-acquire semantics.
  Acquire fences map to \texttt{dmb.ld}, and release fences to
  \texttt{dmb.sy} --- \texttt{dmb.st} does not provide order to prior
  reads.}.

We show that any ARM-consistent execution graph for this sublanguage can be
considered an execution of our semantics.  
%
% Syntactically, we drop the superscript \textsf{rlx} on relaxed reads and
% writes; in addition, we use structured conditionals rather than the more
% general \textsf{goto}.  We refer to this sublanguage as $\muIMM$.
% (Because the source language lacks RMW operations, the ``is
% exclusive'' flag on every read will be \textsf{not-ex} and the RMW mode on
% every write will be \textsf{normal}.)
%
Due to space limitations, we do not include a full description ARM
consistency in the main text .
Here we provide a birds eye view of the details, drawing on the intuitions gleaned from~\citep{DBLP:journals/pacmpl/PulteFDFSS18}.  
Interested readers should see \textsection\ref{sec:arm:proof}
for further details.

An ARM execution graph $G$ defines many relations, including program order
($\rpox$), reads-from ($\rrfx$), coherence ($\rco$) and several dependency
orders.  From these are derived:
\begin{itemize}
\item ${\rpoloc}$, which is the subrelation of $\rpox$ that only relates
  actions on the same location,
\item ${\rob}$, which is required to be acyclic (\ref{external}), and
\item $\reco$, with the requirement that ${\rpoloc}\cup{\reco}$ be acyclic (\ref{sc-per-loc}).
\end{itemize}
% Let $G$ be an execution graph satisfying the ARM consistency 
%requirements.
Given an execution graph $G$, we say that $\aEv$ is an \emph{internal read} if $\aEv\in\fcodom({\rpox}\cap {\rrfx})$.

The ${\rob}$ order is an acyclic global order on events, agreed upon by all threads, reflecting the progress of time in an \armeight\ execution.  The cross thread component of the ordering is induced by the ordering on conflicting actions on the same location from different threads.    The intra thread component of the ordering is induced by barrier ordering and data ordering.  Notably, these dependencies  are determined syntactically.  In particular, $\rob$
may not necessarily include the intra thread component of $\rpoloc$ ordering.

This motivates the translation of an \armeight\ execution into our setting.  In our setting, the progress of time is given by $\lt$.   We accommodate intra-thread reordering by internal read actions, thus excusing us from the obligation of placing them on the global $\lt$-timeline.

Formally, from $G$ we construct a candidate pomset $\aPS$ as follows:
\begin{itemize}
\item $\Event= \textsf{E}$,
\item $\labelingAct(\aEv)=\tau \mathsf{lab}(e)$, if $\aEv$ is a relaxed
  internal read, 
\item $\labelingAct(\aEv)=\mathsf{lab}(e)$, if $\aEv$ is not a relaxed
  internal read,
\item $\labelingForm(\aEv)=\TRUE$,
% \item ${\le} = {\rob}^?$, where $?$ denotes reflexive closure, and
\item ${\gtN} = ({\rob} \cup {\reco})^*$, where $*$ denotes reflexive and transitive closure.
\end{itemize}

\begin{theorem}
  If $G$ is ARM consistent, the constructed candidate satisfies the
  requirements for a top-level memory model pomset.
\end{theorem}
Any $\lt$-ordering imposed in our model
is enforced by \armeight, since our notion of semantic dependency is more
permissive than \armeight 's syntactic dependency.  So, the heart of the proof is showing the acyclicity of $({\rob} \cup {\reco})^*$ for the events under consideration.  Since the cross thread portion of $\reco$ ($\rcoe,\rfre,\rrfe$) is included in $\rob$, this result is really about the influence of $\reco \cap \rpox$.  Our translation of \armeight 's \rrfi\  as silent internal actions removes them from order considerations.   Consequently, we only have to consider the suborder of ${\rob}$ derived without ever using  $\rrfi$ for the following key property demonstrated in \textsection\ref{sec:arm:proof}.
\begin{lemma}\label{extendob}
Let $\aEv, \bEv$ be distinct events and $\bEv'\ (\xob\cap \xpox) \ \bEv\ ((\xeco\cap \xpox) \setminus \xrfi) \  \aEv\ (\xob \cap \xpox)  \ \aEv'$.  Then $\bEv' \xob \aEv'$.
\end{lemma}

\begin{remark}[Proof for TSO]
  The proof for compilation into \tso\ is very similar.  In particular the facts listed above hold for \tso, where $\rob$ is replaced by (the
  transitive closure of) the propagation relation defined for \tso\ 
  \citep{alglave}.
\end{remark}

\section{Proof of compilation for ARMv8}
\label{sec:arm:proof}
\textcolor{red}{[This is out of date, from an old submission]}

In this section, we develop the proof of correctness of compilation to \armeight.  In order to ease readability, we reproduce the definitions from the main text.

Given a relation $R$, $R^?$ denotes reflexive closure, $R^+$ denotes
transitive closure and $R^*$ denotes reflexive and transitive closure.  Given relations $R$ and $S$, $R;S$ denotes composition.

The ARMv8 model is described using the following relations.
\begin{itemize}
\item $\IDR$, $\IDW$, $\IDAcq$, $\IDRel$: identity on reads, writes, acquires
  and releases.
% \item $\IDR$ identity on reads
% \item $\IDW$: identity on writes
% \item $\IDAcq$: identity on acquires
% \item $\IDRel$: identity on releases
\item $\IDLoc$: relates any two events that touch the same location.
\item $\rpox$: program order.
\item $\rdata$, $\rctrl$, $\raddr$: data, control and address dependencies.
\item $\rrfx$: reads-from. $\rrfx^{-1}$ relates each read to a matching write
  on the same location.
\item $\rco$: coherence, which is a total order on the writes to a single
  location.
\item ${\rfr}\eqdef{\rco};\rrfx^{-1}$: from-read, which relates reads to
  subsequent writes.
\end{itemize}
For any relation, the cross-thread subrelation is denoted by appending $e$;
the intra-thread subrelation is denoted by appending $i$.  For example,
${\rrfe}\eqdef{\rrfx}\setminus{\rpox}$ and ${\rrfi}\eqdef{\rrfx}\cap{\rpox}$.
The subrelation restriction attention to actions on the same location is
given by appending $\mathsf{loc}$.  For example, ${\rpoloc}\eqdef{\rpox}\cap{\IDLoc}$.

The ARMv8 model defines the following relations.
In our presentation, we have elided rules concerning fences and RMW operations.
\begin{align*}
  \tag{Extended coherence}
  {\reco} &\eqdef {\rrf} \cup {\rfr} \cup {\rco}
  \\
  \tag{Observed externally}
  {\robs} &\eqdef \smash{
    {\rrfe} \cup {\rfre} \cup {\rcoe}
  }
  \\
  \tag{Dependency order}
  {\rdob} &\eqdef\smash{
    ({\raddr}\cup{\rdata}); {\rrfi}^?
    \cup ({\rctrl}\cup{\rdata}); {\IDW}; {\rcoi}^?
    \cup {\raddr}; {\rpox}; {\IDW}
  }
  \\
  \tag{Barrier order}
  {\rbob} &\eqdef\smash{
    {\IDAcq}; {\rpox}
    \cup {\rpox};{\IDRel}; {\rcoi}^?
  }
  \\
  \tag{Acyclic order}
  {\rob} &\eqdef\smash{
    ({\robs} \cup {\rdob} \cup {\rbob})^+
  }
\end{align*}
\begin{definition}
  An RMW-free and fence-free execution is \emph{ARM-consistent} if
  \begin{align*}&
    \tag{\textsc{$\rrfx$-completeness}}\label{rf-comp}
    \fcodom(\rrfx)=\fdom(\rreads)
    \\[-1ex]&
    \tag{\textsc{$\rco$-totality}}\label{co-tot}
    \text{For every location $\aLoc$, $\rco$ totally orders the writes of $\aLoc$}  
    \\[-1ex]&
    \tag{\textsc{sc-per-loc}}\label{sc-per-loc}
    {\rpoloc} \cup {\rrfx} \cup {\rfr} \cup {\rco}\;\text{is acyclic}
    \\[-1ex]&
    \tag{\textsc{external}}\label{external}
    {\rob}\;\text{is acyclic}
  \end{align*}
\end{definition}

% Use these to refer to the rules in text:
%\ref{rf-comp} 
%\ref{co-tot}
%\ref{sc-per-loc}
%\ref{external}

Given an execution graph $G$, we say that $\aEv$ is an \emph{internal read} if
$\aEv\in\fcodom(\mathsf{po}\cap \mathsf{rf})$.    We are going to translate internal reads of execution graphs into internal reads of the semantics.

From $G$ we construct a candidate pomset $\aPS$ as follows:
\begin{itemize}
\item $\Event= \textsf{E}$,
\item $\labelingAct(\aEv)=\tau \mathsf{lab}(e)$, if $\aEv$ is a relaxed
  internal read, 
\item $\labelingAct(\aEv)=\mathsf{lab}(e)$, if $\aEv$ is not a relaxed
  internal read,
\item $\labelingForm(\aEv)=\TRUE$,
%\item ${\le} = {\rob}$, and
\item ${\gtN} = ({\rob} \cup {\reco})^*$
\end{itemize}
To reempphasize, in this candidate pomset, $\rob$ is calculated by considering the definition of $\rob$ without $\rrfi$, ie.:
\begin{align*}
  \tag{Dependency order}
  {\rdob} &\eqdef\smash{
    ({\raddr}\cup{\rdata});
    \cup ({\rctrl}\cup{\rdata}); {\IDW}; {\rcoi}^?
    \cup {\raddr}; {\rpox}; {\IDW}
  }
  \\
  \tag{Barrier order}
  {\rbob} &\eqdef\smash{
    {\IDAcq}; {\rpox}
    \cup {\rpox};{\IDRel}; {\rcoi}^?
  }
  \\
  \tag{Acyclic order}
  {\rob} &\eqdef\smash{
    ({\robs} \cup {\rdob} \cup {\rbob})^+
  }
\end{align*}

We show that $\aPS$ is a top-level pomset, reasoning as follows.
% We establish the criteria for a top-level memory-model pomset:
\begin{itemize}
\item ${\le}$ is a partial order.  This holds since $G.{\rar}$ is acyclic.
% \item If $\bEv \le \aEv$ then $\bEv \gtN \aEv$.  By construction.
% \item If $\bEv \le \aEv$ and $\aEv \gtN \bEv$ then $\bEv = \aEv$.  Proved below.
% \item If $\cEv \le \bEv \gtN \aEv$ or $\cEv \gtN \bEv \le \aEv$ then
%   $\cEv \gtN \aEv$. By construction.
% \end{itemize}

% Next, we establish the criteria for a 3-valued pomset with preconditions (Definition~\ref{def:3pre}).
% \begin{itemize}
\item $\labelingForm(\aEv)$ implies $\labelingForm(\bEv)$ whenever
  $\bEv\le\aEv$.   Trivial, since every formula is $\TRUE$.
% \item $\aPS$ is $\aLoc$-coherent; that is, when restricted to events that
%   read or write $\aLoc$, $\gtN$ forms a partial order.
% \end{itemize}

% Finally, we establish the criteria for a top-level pomset
% (Definition~\ref{def:x-closed}).
% \begin{itemize}
\item $\aEv$ is location independent. Trivial, since every formula is $\TRUE$.
\item If $\aEv$ reads $\aVal$ from $\aLoc$, then there is some $\bEv$ such that
  \begin{itemize}
  \item $\bEv \lt \aEv$,  
  \item $\bEv$ writes $\aVal$ to $\aLoc$, and
  \item if $\cEv$ writes to $\aLoc$
    then either $\cEv \gtN \bEv$ or $\aEv \gtN \cEv$.
  \end{itemize}    
\end{itemize}

\subsection{Proof that  $({\rob} \cup {\reco})^*$ is irreflexive. }

\paragraph*{Proof of lemma~\ref{extendob}. }

Let $\aEv, \bEv$ be distinct events and $\bEv'\ (\xob\cap \xpox) \ \bEv\ ((\xeco\cap \xpox) \setminus \xrfi) \  \aEv\ (\xob \cap \xpox)  \ \aEv'$.  Then $\bEv' \xob \aEv'$.

\begin{proof}
If $\bEv'$ is an acquire,  or $\aEv$ is an release, or $\aEv'$ is a release, result is immediate.

We next consider the case where $\aEv$ is a read.  In this case,  $\bEv$ is a write.  Since $\bEv\ ((\xeco\cap \xpox) \setminus \xrfi) \  \aEv$, there is a write $\bEv_1$ such that $ \bEv \xcoe\ \bEv_1 \ \xrfe\ \aEv' $.  So, $\bEv \xob \aEv$ and result follows in this case.

So, it suffices to prove the following assuming that $\bEv'$ is not an acquire and $\aEv'$ is not a release and $\aEv$ is not a release or a read and $\aEv, \bEv$ are distinct.
\begin{itemize}
\item If $\bEv'\ (\xob\cap \xpox)  \ \bEv(\xeco\cap\xpox)\aEv$ then $\bEv'\xob\aEv$.
\item If $\bEv\ (\xeco\cap\xpox) \ \aEv(\xob\cap\xpox)\aEv'$ then $\bEv\xob\aEv'$.
\end{itemize}

We first prove that if $\bEv'\ (\xob \cap \xpox) \ \bEv\ (\xeco \cap \xpox) \ \aEv$ then $\bEv'\xob\aEv$.   Proof proceeds by cases on the witness for $\bEv'\ (\xob\cap \xpox) \ \bEv$. 
\begin{itemize}
\item  If $\bEv' \xbob  \bEv$, then: 
\[ \bEv'\ (\smash{
    {\IDAcq}; {\rpox}
    \cup {\rpox};{\IDRel}; {\rcoi}^?) \ 
  }
\bEv
\]
Since $\bEv'$ is not an acquire, $\bEv' ({\rpox};{\IDRel}; {\rcoi}^?) \bEv$, so $\bEv$ is a write.  Since $\aEv$ is not a read,  $\bEv \xcoi\ \aEv$. Thus, result follows.

\item If $\bEv' \xdob  \bEv$, then: 
\[ \bEv'\ 
\smash{
    ( ({\rctrl}\cup{\rdata}); {\IDW}; {\rcoi}^?
    \cup {\raddr}; {\rpox}; {\IDW}
  } \
\bEv
\]
So, $\bEv$ is a write.  Since $\aEv$ is also a write, we deduce that 
\[ \bEv'\ 
\smash{
    ( ({\rctrl}\cup{\rdata}); {\IDW}; {\rcoi}^?
    \cup {\raddr}; {\rpox}; {\IDW}
  } \
\aEv
\]
\end{itemize}

We next prove  that if $\bEv\ (\xeco \cap \xpox) \ \aEv\ (\xob\cap \xpox) \ \aEv'$ then $\bEv\xob\aEv'$, under the assumptions that  $\aEv'$ is not a release and $\aEv$ is not a release or a read and $\aEv, \bEv$ are distinct.

 Proof proceeds by cases on the witness for $\aEv (\xob\cap \xpox) \aEv'$.

\begin{itemize}
\item  If $\aEv \xbob  \aEv'$, then: 
\[ \aEv\ (\smash{
    {\IDAcq}; {\rpox}
    \cup {\rpox};{\IDRel}; {\rcoi}^?) \ 
  }
\aEv'
\]
Since $\aEv$ is not a read, $\aEv ({\rpox};{\IDRel}; {\rcoi}^?) \aEv'$.  Result follows since  $\bEv \xpox\ \aEv$.

\item If $\aEv \xdob  \aEv'$, then $\aEv$ is a read.  \qedhere

\end{itemize}
\end{proof}

\begin{lemma}\label{obeco1}
If $\bEv\xob\aEv$ then $\lnot(\aEv\xeco\bEv)$.

\begin{proof}
Proof by contradiction.  Let 
\[ \aEv \xob \aEv' \xeco \bEv' \xob \bEv \xob \cEv \xob \cEv' \xob \aEv \]
where $\aEv' \xpox \bEv'$.

By lemma~\ref{extendob}, if $\aEv \not=\aEv'$, we deduce $\aEv \xob \bEv'$, and thus $\aEv \xob \bEv$.  If $\bEv \not=\bEv'$, we deduce $\aEv' \xob \bEv$ and thus $\aEv \xob \bEv$.

Thus, if $\aEv \not=\aEv'$ or $\bEv \not=\bEv'$, then there is a cycle $\aEv \xob \bEv \xob \cEv \xob \cEv' \xob \aEv$.

So we can assume that  $\aEv' = \aEv$, $\bEv' = \bEv$ and 
\[ \aEv  \xeco \bEv \xob \cEv \xob \cEv' \xeco \aEv \]
where all of $\aEv, \bEv, \cEv, \cEv'$ access the same location and at least one of $\aEv,\bEv$ is a write, at least one of $\aEv,\cEv'$ is a write, and at least one of $\bEv,\cEv$ is a write.

We reason by cases.
\begin{itemize}
\item If $\cEv'$ is a write or both $(\aEv, \bEv)$ are writes.

We deduce that $\bEv \xeco \cEv' \xeco \aEv$ and thus $\bEv \xeco \aEv$.

\item $\cEv'$ is a read.  $\aEv$ is a write.  $\bEv$ is a read.

In this case $\cEv$ is a write.  From $\cEv \xob \aEv$, we deduce $\cEv \xeco \aEv$. Combining with $\bEv \xeco \cEv$, we deduce that $\bEv \xeco \aEv$.

\end{itemize}
In either case, there is a contradiction $\aEv \xeco \bEv \xeco \aEv$.
\end{proof}
\end{lemma}

\begin{lemma}\label{obeco2}
$({\rob} \cup {\reco})^*$ is irrreflexive.
\end{lemma}
\begin{proof}
The simple case that $\rob; \reco$ is irreflexive is proved above.  The full proof by contradiction.

Let $n \geq 1$ be the minimum such that:
\begin{align*} 
&\aEv^0_0 \xob \aEv^0_1 \xeco \bEv^0_0 \xob \bEv^0_1  \\
(\xeco \cap \xob) &  \   \aEv^1_0 \xob \aEv^1_1 \xeco \bEv^1_0 \xob \bEv^1_1 \\
(\xeco \cap \xob) & \ \ldots \\
& \ldots \bEv^n_1 \\
 (\xeco \cap \xob) & \  \aEv^0_0
\end{align*}
where  for all $i$, we have:
\[ \aEv^i_0 \xpox \aEv^i_1 (\xeco \cap \xpox) \bEv^i_0 \xpox \bEv^i_1\] and 
\[ \neg (\bEv^i_1 \xpox (\aEv^{(i+1) \mod n}_0 \]

For any $i$, if $\aEv^i_0 \not= \aEv^i_1$ or $\bEv^i_0 \xpox \bEv^i_1$, via lemma~\ref{extendob}, we deduce that $\aEv^i_0  \xob \bEv^i_1$, contradicting minimality of $n$.

So, we can assume that $n \geq 1$ is such that:
\begin{align*} 
&\ \aEv^0 \xeco  \bEv^0 \\
(\xeco \cap \xob) &  \   \aEv^1  \xeco \bEv^1 \\
(\xeco \cap \xob) & \ \ldots \\
& \ldots \bEv^n \\
 (\xeco \cap \xob) & \ \aEv^0
\end{align*}
which is a contradiction since it is a cycle in $\xeco$ and since at least one of $\aEv^i ,\bEv^i$ is a write for all $i$. 
\end{proof}

### 2021-01-08 paper/appendix.tex
Initial definitions.

\appendix
% \appendices
% \section{}

This paper addresses several errors in \citep{10.1145/3428262}, which we
henceforth refer to as \jjr{}.

In \jjr{\textsection2.4}, parallel composition is defined allowing coalescing
of events.  Here we have forbidden coalescing.  This difference appears to be
arbitrary.  In \jjr{}, however, there is a mistake in the handling of
termination actions.  The predicates should be joined using $\land$, not
$\lor$.

In \jjr{\textsection2.6} the semantics of read is defined as follows:
\begin{align*}
  \sem{\aReg\GETS\aLoc^\amode\SEMI \aCmd} & \eqdef \textstyle\bigcup_\aVal\;
  (\DRmode\aLoc\aVal) \prefix \sem{\aCmd} [\aLoc/\aReg]
\end{align*}
The definition of prefixing$((\aForm \mid \aAct) \prefix \aPSS)$ has several clauses.
The most relevant are as follows, where $\bEv$ is the new event labeled with
$(\aForm \mid \aAct)$ and $\aEv$ is an event from $\aPSS$:
\begin{description}
\item[{\labeltextsc[P4c]{(P4c)}{4c}}]
  If $\bEv$ reads $\aVal$ from $\aLoc$ then either $\aEv=\bEv$ or
  $\labelingForm'(\aEv)$ implies $\labelingForm(\aEv)[\aVal/\aLoc]$.
\item[{\labeltextsc[P5a]{(P5a)}{5a}}]\labeltextsc[P5]{}{5}%
  If $\bEv$ reads and $\aEv$ writes then either $\labelingForm'(\aEv)$
  implies $\labelingForm(\aEv)$ or $\bEv\le'\aEv$.
% \item[{\labeltextsc[P5b]{(P5b)}{5b}}]
%   If $\bEv$ and $\aEv$ are in conflict then $\bEv\le'\aEv$.
\end{description}

We have discovered two issues with this definition.

The first issue concerns the substitution $[\aLoc/\aReg]$.  It should be
$[\aReg/\aLoc]$.  We noticed this error while developing the alternative
characterization presented here.  The error causes redundant read elimination
to fail in \jjr{}.  As a result, common subexpression elimination also fails.
The problem can be seen in \ref{TC2}.
\begin{gather*}
  \taglabel{TC2}
  r\GETS x\SEMI
  s\GETS x\SEMI
  \IF{r{=}s}\THEN y\GETS 1\FI
  \PAR
  x\GETS y
\end{gather*}
% In \jjr{\textsection3.1},
We claimed that \ref{TC2} allowed the following
execution:
\begin{gather*}
  \hbox{\begin{tikzinline}[node distance=1.5em]
  \event{a1}{\DR{x}{1}}{}
  \event{a2}{\DR{x}{1}}{right=of a1}
  \event{a3}{\DW{y}{1}}{right=of a2}
  % \po{a2}{a3}
  % \po[out=15,in=165]{a1}{a3}
  \event{b1}{\DR{y}{1}}{right=3em of a3}
  \event{b2}{\DW{x}{1}}{right=of b1}
  \rf{a3}{b1}
  \po{b1}{b2}
  \rf[out=169,in=11]{b2}{a2}
  \rf[out=169,in=11]{b2}{a1}
    \end{tikzinline}}
\end{gather*}
But this execution is not possible using the semantics of \jjr{}:
$\DWP{y}{1}$ has precondition $r{=}s$ in
\begin{math}
  \sem{\IF{r{=}s}\THEN y\GETS 1\FI}.
\end{math}
Given the lack of order in the execution, the precondition of $\DWP{y}{1}$
must entail $r{=}1\land r{=}x$ in 
\begin{math}
  \sem{s\GETS x\SEMI
  \IF{r{=}s}\THEN y\GETS 1\FI}.
\end{math}
\ref{4c} imposes $r{=}1$, and \ref{5a} imposes $r{=}x$.  Adding the second
read, the precondition of $\DWP{y}{1}$ must entail both $1{=}1\land 1{=}x$
and also $x{=}1\land x{=}x$.  This can be simplified to $x{=}1$.  This leaves
a requirement that must be satisfied by a preceding write.  Since the
preceding write is the initialization to $0$, the requirement cannot be
satisfied, and the execution is impossible.\footnote{In \jjr{} we ignore the
  middle terms, mistakenly simplifying this to $1{=}1\land x{=}x$.
  Correcting the error, the attempted execution is:
\begin{gather*}
  \hbox{\begin{tikzinline}[node distance=1.5em]
  \event{a1}{\DR{x}{1}}{}
  \event{a2}{\DR{x}{1}}{right=of a1}
  \event{a3}{\DW{y}{1}}{right=of a2}
  \po{a2}{a3}
  \po[out=-20,in=-160]{a1}{a3}
  \event{b1}{\DR{y}{1}}{right=3em of a3}
  \event{b2}{\DW{x}{1}}{right=of b1}
  \rf{a3}{b1}
  \po{b1}{b2}
  \rf[out=169,in=11]{b2}{a2}
  \rf[out=169,in=11]{b2}{a1}
    \end{tikzinline}}
\end{gather*}}

The substitution $[\aLoc/\aReg]$ leaves the obligation on $\aLoc$ to be
fulfilled by the preceding write.  Thus, the read does not update the
\emph{value} of $\aLoc$ in subsequent predicates.  The substitution
$[\aReg/\aLoc]$, instead, does update the value of $\aLoc$, thus removing any
obligation on $\aLoc$ for preceding code.

In order to write this, we must update the definition of prefixing reads to
include the register.  Then \ref{4c} becomes:
\begin{description}
\item[\textsc{(p4c)}] If $\bEv$ reads $\aVal$ from $\aLoc$ then either
  $\aEv=\bEv$ or $\labelingForm'(\aEv)$ implies
  $\labelingForm(\aEv)[\aVal/\aReg]$.
\end{description}

We can then reason with \ref{TC2} as follows: $\DWP{y}{1}$ has precondition
$r{=}s$ in
\begin{math}
  \sem{\IF{r{=}s}\THEN y\GETS 1\FI}.
\end{math}
To avoid introducing order in the execution, the precondition of $\DWP{y}{1}$
must entail $r{=}1\land r{=}s$ in 
\begin{math}
  \sem{s\GETS x\SEMI
  \IF{r{=}s}\THEN y\GETS 1\FI}.
\end{math}
\ref{4c} imposes $r{=}1$, and \ref{5a} imposes $r{=}x$.  Adding the second
read, the precondition of $\DWP{y}{1}$ must entail both $1{=}1\land 1{=}x$
and also $x{=}1\land x{=}x$.  This can be simplified to $x{=}1$.  This leaves
a requirement that must be satisfied by a preceding write.

With read elimination, the rule for relaxed reads is as follows:
\begin{align*}
  \sem{\PR{\aLoc}{\aReg} \SEMI \aCmd} &\eqdef
  \sem{\aCmd}[\aLoc/\aReg]
  \cup
  \textstyle\bigcup_\aVal\;
  \DRP{\aLoc}{\aVal} \prefix_{\aReg} %\Rdis{\aLoc}{\aVal}
  \sem{\aCmd}[\aReg/\aLoc]
\end{align*}
It is interesting to note that the substitution is $[\aLoc/\aReg]$ on
eliminated reads, and $[\aReg/\aLoc]$ on non-eliminated reads.  Intuitively,
the subsequent value of $\aLoc$ is fixed by an explicit read, but not for an
eliminated read.  In the latter case, the value is fixed by some preceding
action.  The preceding action may itself be a read. This gives rise to some
fear that we might introduce thin-air reads, since we do not enforce
read-read coherence.  But this is not the case.  Consider the following example:
\begin{gather*}
  r\GETS x\SEMI
  s\GETS x\SEMI
  y\GETS s
  \PAR
  x\GETS y
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
  \event{a1}{\DR{x}{1}}{}
  \event{a2}{\DR{x}{1}}{right=of a1}
  \event{a3}{\DW{y}{1}}{right=of a2}
  %\po{a2}{a3}
  \po[out=-20,in=-160]{a1}{a3}
  \event{b1}{\DR{y}{1}}{right=3em of a3}
  \event{b2}{\DW{x}{1}}{right=of b1}
  \rf{a3}{b1}
  \po{b1}{b2}
  \rf[out=169,in=11]{b2}{a2}
  \rf[out=169,in=11]{b2}{a1}
    \end{tikzinline}}
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
  \event{a1}{\DR{x}{1}}{}
  \internal{a2}{\DR{x}{1}}{right=of a1}
  \event{a3}{\DW{y}{1}}{right=of a2}
  %\po{a2}{a3}
  \po[out=-20,in=-160]{a1}{a3}
  \event{b1}{\DR{y}{1}}{right=3em of a3}
  \event{b2}{\DW{x}{1}}{right=of b1}
  \rf{a3}{b1}
  \po{b1}{b2}
  %\rf[out=169,in=11]{b2}{a2}
  \rf[out=169,in=11]{b2}{a1}
    \end{tikzinline}}
\end{gather*}
But this is not a problem, since fulfillment requires that $\DWP{x}{1}$
precede both reads of $x$.

The second issue concerns acquiring reads.  Shortly after publication,
\citet{anton} noticed a shortcoming of the implementation on \armeight{} in
\jjr{\textsection 7}.  The proof given there assumes that all internal reads
can be dropped.  However, this is not the case for acquiring reds.  For
example, \jjr{} disallows the following execution, which is allowed by
\armeight{} and \tso{}.
\begin{gather*}
  \PW{x}{2}\SEMI 
  \PR[\mRA]{x}{r}\SEMI
  \PR{y}{s} \PAR
  \PW{y}{2}\SEMI
  \PW[\mRA]{x}{1}
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a}{\DW{x}{2}}{}
      \event{b}{\DR[\mRA]{x}{2}}{right=of a}
      \event{c}{\DR{y}{0}}{right=of b}
      \event{d}{\DW{y}{2}}{right=2.5em of c}
      \event{e}{\DW[\mRA]{x}{1}}{right=of d}
      \rf{a}{b}
      \sync{b}{c}
      \wk{c}{d}
      \sync{d}{e}
      \wk[out=-165,in=-15]{e}{a}
      % \rfi{a}{b}
      % \bob{b}{c}
      % \fre{c}{d}
      % \bob{d}{e}
      % \coe[out=-165,in=-15]{e}{a}
    \end{tikzinline}}
\end{gather*}
The solution we have adopted is to allow an acquiring read to be downgraded
to a relaxed read when it is preceded (sequentially) by a relaxed write that
could fulfill it.  Back-porting this solution to \jjr{} requires that we add
access predicates to the logic and allow

In addition, the \armeight{} implementation result given here does not rely
on read elimination.  Instead we use a recent alternative characterization of
\armeight{} \citep{alglave-git-alternate,arm-reference-manual,armed-cats}.

### 2021-01-09 paper/appendix.tex
Small edits.
Draft of full semantics.

@@
 
+\subsection{Reads Update Local State}
+In the rule for read prefixing we have substituted $[r/x]$, rather than
+$[x/r]$.  This means that reads clobber local state.  We assume registers are
+only used once---otherwise, one needs to generate a fresh register for the
+substitution.
+
+With read-read dependencies, this difference can be seen.  For example, the
+following execution is allowed with $[x/r]$, but not $[r/x]$.
+\begin{gather*}
+  \PW{x}{0}\SEMI
+  \PR{x}{r}\SEMI
+  \IF{r}\THEN \PR{x}{s}\FI\SEMI
+  \PW{y}{s{+}1}
+  \PAR
+  \PW{x}{\PR{y}{}}
+  \\[-1ex]
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a1}{\DW{x}{0}}{}
+      \event{a2}{\DW{x}{1}}{right=of a1}
+      \event{a3}{\DR{x}{0}}{right=of a2}
+      \event{a4}{\DW{y}{1}}{right=of a3}
+      \event{b1}{\DR{y}{1}}{right=3em of a4}
+      \event{b2}{\DW{x}{1}}{right=of b1}
+      \wk{a1}{a2}
+      \po{a2}{a3}
+      \po{b1}{b2}
+      \rf{a4}{b1}
+      \rf[out=165,in=15]{b2}{a2}
+    \end{tikzinline}}
+\end{gather*}
+[Is there a difference w/o read-read dependencies?]
+
+\subsection{Parallel Composition}
+
 In \jjr{\textsection2.4}, parallel composition is defined allowing coalescing
@@
 $\lor$.
+
+\subsection{Redundant Read Elimination}
 
@@
 
+\subsection{Internal Acquiring Reads}
+
 The second issue concerns acquiring reads.  Shortly after publication,
@@
 
+The notion of data-race is incorrect in \jjr{}.
+\begin{gather*}
+  \PW{x}{1}\SEMI
+  \PW[\mRA]{y}{1}\SEMI
+  \PR[\mRA]{x}{r}
+  \PAR
+  \IF{\PR[\mRA]{y}{}}\THEN \PW[\mRA]{x}{2}\FI
+  \\
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a1}{\DW{x}{1}}{}
+      \event{a2}{\DW[\mRA]{y}{1}}{right=of a1}
+      \event{a3}{\DR[\mRA]{x}{1}}{right=of a2}
+      \event{b1}{\DR[\mRA]{y}{1}}{right=3em of a3}
+      \event{b2}{\DW[\mRA]{x}{2}}{right=of b1}
+      \sync{a1}{a2}
+      \rf[out=20,in=160]{a1}{a3}
+      \rf[out=20,in=160]{a2}{b1}
+      \sync{b1}{b2}
+    \end{tikzinline}}
+  \\
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a1}{\DW{x}{1}}{}
+      \event{a2}{\DW[\mRA]{y}{1}}{right=of a1}
+      \event{a3}{\DR[\mRA]{x}{2}}{right=of a2}
+      \event{b1}{\DR[\mRA]{y}{1}}{right=3em of a3}
+      \event{b2}{\DW[\mRA]{x}{2}}{right=of b1}
+      \sync{a1}{a2}
+      \rf[out=20,in=160]{a2}{b1}
+      \rf[out=160,in=20]{b2}{a3}
+      \sync{b1}{b2}
+    \end{tikzinline}}
+\end{gather*}
+Bug is in \citep[Lemma A.4]{DBLP:conf/ppopp/DongolJR19}.  It assumes that
+$\DRP[\mRA]{x}{1}$ and $\DWP[\mRA]{x}{2}$ are racing in the first execution
+because they are not ordered by happens-before.  But this is false since
+neither is plain.
+
 In addition, the \armeight{} implementation result given here does not rely

### 2021-01-13 paper/intro.tex
edits

@@
 \section{Introduction}
+
+Our approach is only vaguely related to the theories of
+\citet{DBLP:journals/cacm/Dijkstra75}, which provide an alternative
+characterization of Hoare logic by mapping postconditions to preconditions
+($\fwp$ and $\fwlp$) or preconditions to postconditions ($\fsp$).  Instead
+our transformers $\aTr$ map preconditions to preconditions. The relation to
+Hoare triples is that
+\begin{align*}
+  \hoare{\aForm}{\bCmd}{\bForm}
+  \;\;\Leftrightarrow\;\;
+  \hoare{\aTr(\aCmd,\aForm)}{\aCmd\SEMI\bCmd}{\bForm}
+  .
+\end{align*}
+For example
+\begin{gather*}
+  \hoare{r{=}1}{\PW{y}{r}}{y{=}1}
+  \\
+  \hoare{x{=}1}{\PR{x}{r}\SEMI\PW{y}{r}}{y{=}1}
+\end{gather*}
+
+\endinput
 

### 2021-01-22 paper/abstract.tex
More or less feature complete in the paper.

% Program logics and semantics tell us that in order to derive
% \begin{math}
%   ((S_1\SEMI S_2), \sigma_0) \Downarrow \sigma_2,
% \end{math}
% we derive 
% \begin{math}
%   (S_1, \sigma_0) \Downarrow \sigma_1
% \end{math}
% and then
% \begin{math}
%   (S_2, \sigma_1) \Downarrow \sigma_2.
% \end{math}
Program logics and semantics tell us that when executing ((S1; S2), state0),
we execute (S1, state0) to arrive at state1, then execute (S2, state1) to
arrive at the final state2.

This is a delightfully simple story that can be explained to children.  It is
also a lie.

Processors execute instructions out of order, due to pipelines and caches.
Compilers reorder programs even more dramatically.  All of this reordering is
meant to be unobservable in single-threaded code.  In multi-threaded code,
however, all bets are off.  A formal attempt to understand the resulting mess
is known as a ``relaxed memory model.''

Most of the relaxed memory models that have been proposed are designed to
help us understand whole program execution: they have no compositionality
properties whatsoever.  Recently, denotation models have appeared that treat
\emph{concurrent} execution compositionally.  One such model is ``Pomsets with
Preconditions''.  It remains an open question, however, whether it is
possible to treat \emph{sequential} execution compositionally in such a model,
without overly restricting processors and compilers.

We propose adding families of predicate transformers to Pomsets with
Preconditions.  The resulting model is denotational, supporting both parallel
and sequential composition.  When composing (S1;S2), the predicate
transformer used to validate the precondition of an event in S2 is chosen
based on the dependency order from S1 into this event.  As usual in work on
relaxed memory, we have not handled loops or recursion.

Happily, most of the results expected of a relaxed memory model can be
established by appeal to prior work.  So here we are able to concentrate on
the model itself.  The model is formalized in Agda, where we have established
associativity for sequential composition.

For the memory-model specialist, we retain the good properties of the prior
work on Pomsets with Preconditions, fixing some errors along the way.  These
properties include efficient implementation on ARMv8, support for compiler
optimizations, support for logics that prove the absence of thin-air
behaviors, and a local data race freedom theorem.

### 2021-01-22 paper/arm.tex
Initial definitions.
More or less feature complete in the paper.

@@
 \label{sec:arm}
+
+\subsection{Down-Grading Acquires ($\xDGA$)}
+\label{sec:internal}
+
+We use the uninterpreted symbols $\Dx{\aLoc}$.
+
+\begin{definition}
+  \label{def:DS}
+  Let $[\FALSE/\D]$ be the substitution that performs $[\FALSE/\Dx{\aLoc}]$
+  for every $\aLoc$.
+  Let $\DS{\aLoc}{\amode}$ and $\DL{\aLoc}{\amode}$ be defined:
+  \begin{align*}
+    \DS{\aLoc}{\mRLX}{\bForm}&=\bForm[\TRUE/\Dx{\aLoc}] 
+    &\DL{\aLoc}{\mRLX}&=\TRUE
+    \\
+    \DS{\aLoc}{\mRA}{\bForm}&=\bForm[\FALSE/\D]
+    &\DL{\aLoc}{\mRA}&=\Dx{\aLoc}
+    \\
+    \DS{\aLoc}{\mSC}{\bForm}&=\bForm[\FALSE/\D]
+    &\DL{\aLoc}{\mSC}&=\Dx{\aLoc}
+  \end{align*}
+\end{definition}
+
+\begin{definition}[$\xCO$/$\xRASC$/$\xDGA$]
+  \label{def:pomsets-down}
+  Update Def.~\ref{def:pomsets-ra} to:
+  \begin{enumerate}
+  \item[\ref{S4})]
+    $\aTr{\bEvs}{\bForm}$ implies $(\Qw{\aLoc}\limplies\aExp{\neq}\aVal)\land \DS{\aLoc}{\amode}\bForm[\aExp/\aLoc]$,
+  \item[\ref{S5})]
+    $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qw{\aLoc}\land \DS{\aLoc}{\amode}\bForm[\aExp/\aLoc]$.
+  \item[\ref{L5})]
+    $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode} \land((\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm)$,
+  \item[\ref{L6})]
+    $\aTr{\dEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode} \land\bForm$.
+  \end{enumerate}
+\end{definition}
+
+\subsection{Read-Read dependencies ($\xRRD$)}
+
+We use the uninterpreted symbol $\RW$.
+
+\begin{definition}[$\xRRD$]
+  \label{def:pomsets-rr}
+  Update Def.~\ref{def:pomsets-trans} to:
+  \begin{enumerate}
+  \item[\ref{T3})]
+    $\labelingForm(\aEv)$ implies
+    $\labelingForm_1(\aEv)[\TRUE/\RW]$ if $\labelingAct_1(\aEv)$ is a write,
+    \\
+    $\labelingForm(\aEv)$ implies
+    $\labelingForm_1(\aEv)[\FALSE/\RW]$ otherwise.
+  \end{enumerate}
+  \begin{enumerate}
+  \item[\ref{L5})]
+    $\aTr{\cEvs}{\bForm}$ implies $\RW\limplies(\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm$,
+  \end{enumerate}  
+\end{definition}
+
+We repeat all of the rules for preconditions and transformers,
+Only \ref{L5} and \ref{T3} are changed from Def.~\ref{def:pomsets-down}.
+We repeat \ref{L4}, although it is unchanged from Def.~\ref{def:pomsets-trans}.
+
+\begin{definition}[$\xCO$/$\xRASC$/$\xDGA$/$\xRRD$]
+  \label{def:pomsets-down-rr}
+  Update Def.~\ref{def:pomsets-trans} to:
+  \begin{enumerate}
+  \item[\ref{S3})]
+    $\labelingForm(\aEv)$ implies $\QS{\aLoc}{\amode}\land\aExp{=}\aVal$,
+  \item[\ref{L3})]
+    $\labelingForm(\aEv)$ implies $\QL{\aLoc}{\amode}$.
+  \item[\ref{T3})]
+    $\labelingForm(\aEv)$ implies
+    $\labelingForm_1(\aEv)[\TRUE/\Q{}][\TRUE/\RW]$ if $\labelingAct_1(\aEv)$ is a write,
+    \\
+    $\labelingForm(\aEv)$ implies
+    $\labelingForm_1(\aEv)[\TRUE/\Q{}][\FALSE/\RW]$ otherwise.
+  \end{enumerate}
+  \begin{enumerate}
+  \item[\ref{S4})]
+    $\aTr{\bEvs}{\bForm}$ implies $(\Qw{\aLoc}\limplies\aExp{\neq}\aVal)\land \DS{\aLoc}{\amode}\bForm[\aExp/\aLoc]$,
+  \item[\ref{S5})]
+    $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qw{\aLoc}\land \DS{\aLoc}{\amode}\bForm[\aExp/\aLoc]$.
+  \item[\ref{L4})]
+    $\aTr{\bEvs}{\bForm}$ implies $\aVal{=}\aReg\limplies\bForm$, 
+  \item[\ref{L5})]
+    $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode} \land(\RW\limplies(\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm)$,
+  \item[\ref{L6})]
+    $\aTr{\dEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode} \land\bForm$.
+  \end{enumerate}  
+\end{definition}
+
+[Control dependencies into reads as in MP with release on right and control
+dependency on left.]
+
+% \begin{definition}
+%   Let $[\aForm/\RW{}]$ be the substitution that replaces $\RW{}$ by $\aForm$.
+% \end{definition}
+
+\subsection{Proof Strategy}
 
@@
 
+\begin{comment}
   Internal visibility:
@@
   \item
-  $\bEv:\DRP{x}{0}\xfr\aEv:\DWP{x}{1}$
+    $\bEv:\DRP{x}{0}\xfr\aEv:\DWP{x}{1}$\\
+    $\bEv:\DRP{x}{0}\xrfxinv\DWP{x}{0}\xco\aEv:\DWP{x}{1}$
   \item %$\bEv\mathrel{\xfr\xrfx}\aEv$:
-  $\bEv:\DRP{x}{0}\xfr\DWP{x}{1}\xrfx\aEv:\DRP{x}{1}$
+    $\bEv:\DRP{x}{0}\xfr\DWP{x}{1}\xrfx\aEv:\DRP{x}{1}$\\
+    $\bEv:\DRP{x}{0}\xrfxinv\DWP{x}{0}\xco\DWP{x}{1}\xrfx\aEv:\DRP{x}{1}$
   \item %$\bEv\mathrel{\xrfxinv\xrfx}\aEv$:
@@
   \end{itemize}
+\end{comment}
+
+\endinput
 
@@
 % Let $G$ be an execution graph satisfying the ARM consistency 
-%requirements.
+% requirements.
 Given an execution graph $G$, we say that $\aEv$ is an \emph{internal read} if $\aEv\in\fcodom({\rpox}\cap {\rrfx})$.
@@
 % Use these to refer to the rules in text:
-%\ref{rf-comp} 
-%\ref{co-tot}
-%\ref{sc-per-loc}
-%\ref{external}
+% \ref{rf-comp} 
+% \ref{co-tot}
+% \ref{sc-per-loc}
+% \ref{external}
 
@@
 \item $\labelingForm(\aEv)=\TRUE$,
-%\item ${\le} = {\rob}$, and
+  % \item ${\le} = {\rob}$, and
 \item ${\gtN} = ({\rob} \cup {\reco})^*$

### 2021-01-23 paper/arm.tex
Updates

@@
 \label{sec:arm}
+
+We discuss \armeight{} using \emph{external global completion} (\EGC{})
+\cite{alglave-git-alternate,armed-cats}, which is very close to our model;
+see also \cite[\textsection B2.3.6]{arm-reference-manual}.
 
@@
 
-We use the uninterpreted symbols $\Dx{\aLoc}$.
+\begin{example}
+The following execution is allowed by \armeight{}, but disallowed by
+\refdef{def:pomsets-ra}.  The coherence order between the writes can be
+witnessed by a separate thread, which we have elided.
+\begin{gather*}
+  \PW{x}{2}\SEMI 
+  \PR[\mRA]{x}{r}\SEMI
+  \PW{y}{1} \PAR
+  \PW{y}{2}\SEMI
+  \PW[\mRA]{x}{1}
+  \\
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a}{\DW{x}{2}}{}
+      \raevent{b}{\DR[\mRA]{x}{2}}{right=of a}
+      \event{c}{\DW{y}{1}}{right=of b}
+      \event{d}{\DW{y}{2}}{right=2.5em of c}
+      \raevent{e}{\DW[\mRA]{x}{1}}{right=of d}
+      \rf{a}{b}
+      \sync{b}{c}
+      \wk{c}{d}
+      \sync{d}{e}
+      \wk[out=-165,in=-15]{e}{a}
+    \end{tikzinline}}
+\end{gather*}
+Under \EGC{}, this is explained by dropping the order
+$\DRP[\mRA]{x}{2}\xsync\DWP{y}{1}$, because $\DRP[\mRA]{x}{2}$ is fulfilled
+by a relaxed write in the same thread.
+\begin{gather*}
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a}{\DW{x}{2}}{}
+      \raevent{b}{\DR[\mRA]{x}{2}}{right=of a}
+      \event{c}{\DW{y}{1}}{right=of b}
+      \event{d}{\DW{y}{2}}{right=2.5em of c}
+      \raevent{e}{\DW[\mRA]{x}{1}}{right=of d}
+      \rf{a}{b}
+      %\sync{b}{c}
+      \wk{c}{d}
+      \sync{d}{e}
+      \wk[out=-165,in=-15]{e}{a}
+    \end{tikzinline}}  
+\end{gather*}
+
+More generally, this can be understood as a compiler optimization that
+downgrades a read from $\mRA$ to $\mRLX$ when it read can be fulfilled by a
+relaxed write in the same thread.
+\end{example}
+
+To model such \emph{downgraded acquires}, we use the uninterpreted symbols
+$\Dx{\aLoc}$.
 
@@
   \label{def:pomsets-down}
-  Update Def.~\ref{def:pomsets-ra} to:
+  Update \refdef{def:pomsets-ra} to:
   \begin{enumerate}
@@
 
-\subsection{Read-Read dependencies ($\xRRD$)}
+% The following execution is allowed both by \armeight{} and \tso{}, but
+% disallowed by \refdef{def:pomsets-ra}.
+% \begin{gather*}
+%   \PW{x}{2}\SEMI 
+%   \PR[\mRA]{x}{r}\SEMI
+%   \PR{y}{s} \PAR
+%   \PW{y}{2}\SEMI
+%   \PW[\mRA]{x}{1}
+%   \\
+%   \hbox{\begin{tikzinline}[node distance=1.5em]
+%       \event{a}{\DW{x}{2}}{}
+%       \raevent{b}{\DR[\mRA]{x}{2}}{right=of a}
+%       \event{c}{\DR{y}{0}}{right=of b}
+%       \event{d}{\DW{y}{2}}{right=2.5em of c}
+%       \raevent{e}{\DW[\mRA]{x}{1}}{right=of d}
+%       \rf{a}{b}
+%       \sync{b}{c}
+%       \wk{c}{d}
+%       \sync{d}{e}
+%       \wk[out=-165,in=-15]{e}{a}
+%     \end{tikzinline}}
+% \end{gather*}
 
-We use the uninterpreted symbol $\RW$.
+\begin{example}
+One might worry that our model is too permissive for $\mSC$ access, but
+\armeight{} allows some counterintuitive results for SC access.  In the
+following execution we elide the initializing write $\DWP{y}{0}$.
+\begin{gather*}
+  \IF{\PR{x}{}}\THEN \PW{x}{2} \FI\SEMI
+  \PR[\mSC]{x}{r}\SEMI
+  \PR[\mSC]{y}{s} \PAR
+  \PW[\mSC]{y}{2}\SEMI
+  \PW[\mSC]{x}{1}
+  \\
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a}{\DR{x}{1}}{}
+      \event{b}{\DW{x}{2}}{right=of a}
+      \scevent{c}{\DR[\mSC]{x}{2}}{right=of b}
+      \scevent{d}{\DR[\mSC]{y}{0}}{right=of c}
+      \scevent{e}{\DW[\mSC]{y}{2}}{right=2.5em of d}
+      \scevent{f}{\DW[\mSC]{x}{1}}{right=of e}
+      \po{a}{b}
+      \rf{b}{c}
+      %\sync{c}{d}
+      \sync{e}{f}
+      \wk{d}{e}
+      \rf[out=-165,in=-15]{f}{a}
+    \end{tikzinline}}
+\end{gather*}
+Under \EGC{}, this is explained by dropping the order
+$\DRP[\mSC]{x}{2}\xsync\DRP{y}{0}$, because $\DRP[\mSC]{x}{2}$ is fulfilled
+by a relaxed write in the same thread.
+\end{example}
+
+\subsection{Removing Read-Read dependencies ($\xRRD$)}
+
+\begin{example}
+The following execution is allowed by \armeight{}, but disallowed by
+\refdef{def:pomsets-ra}.  
+\begin{gather*}
+  \begin{gathered}
+  x\GETS0\SEMI 
+  x\GETS 1\SEMI y^\mRA\GETS1 \PAR r\GETS y\SEMI \IF{r}\THEN s\GETS x\FI
+  \\[-.4ex]
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{wx0}{\DW{x}{0}}{}
+      \event{wx1}{\DW{x}{1}}{right=of wx0}
+      \raevent{wy1}{\DW[\mRA]{y}{1}}{right=of wx1}
+      \event{ry1}{\DR{y}{1}}{right=2.5em of wy1}
+      \event{rx0}{\DR{x}{0}}{right=of ry1}
+      \sync{wx1}{wy1}
+      \po{ry1}{rx0}
+      \rf{wy1}{ry1}
+      \wk[in=-15,out=-165]{rx0}{wx1}
+      \wk{wx0}{wx1}
+    \end{tikzinline}}
+\end{gathered}
+\end{gather*}
+Under \EGC{}, this is explained by dropping the order
+$\DRP{y}{1}\xpo\DRP{x}{0}$, because \armeight{} does not include control
+dependencies between reads in the locally-ordered-before relation.
+\begin{gather*}
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{wx0}{\DW{x}{0}}{}
+      \event{wx1}{\DW{x}{1}}{right=of wx0}
+      \raevent{wy1}{\DWRel{y}{1}}{right=of wx1}
+      \event{ry1}{\DR{y}{1}}{right=2.5em of wy1}
+      \event{rx0}{\DR{x}{0}}{right=of ry1}
+      \sync{wx1}{wy1}
+      %\po{ry1}{rx0}
+      \rf{wy1}{ry1}
+      \wk[in=-15,out=-165]{rx0}{wx1}
+      \wk{wx0}{wx1}
+    \end{tikzinline}}  
+\end{gather*}
+\end{example}
+
+Since we do not distinguish control dependencies from other dependencies, we
+are forced to drop all dependencies between reads.  In order to do so, we use
+the uninterpreted symbol $\RW$.
 
@@
   \label{def:pomsets-rr}
-  Update Def.~\ref{def:pomsets-trans} to:
+  Update \refdef{def:pomsets-trans} to:
   \begin{enumerate}
@@
 We repeat all of the rules for preconditions and transformers,
-Only \ref{L5} and \ref{T3} are changed from Def.~\ref{def:pomsets-down}.
-We repeat \ref{L4}, although it is unchanged from Def.~\ref{def:pomsets-trans}.
+Only \ref{L5} and \ref{T3} are changed from \refdef{def:pomsets-down}.
+We repeat \ref{L4}, although it is unchanged from \refdef{def:pomsets-trans}.
 
@@
   \label{def:pomsets-down-rr}
-  Update Def.~\ref{def:pomsets-trans} to:
+  Update \refdef{def:pomsets-trans} to:
   \begin{enumerate}
@@
 
-[Control dependencies into reads as in MP with release on right and control
-dependency on left.]
-
 % \begin{definition}
@@
 
-\subsection{Proof Strategy}
-
-Alternate characterization of ARM:
-\cite[\textsection B2.3.6]{arm-reference-manual}
-\cite{alglave-git-alternate}
-\cite{armed-cats}
+%\subsection{Proof Strategy}
 

### 2021-01-23 paper/intro.tex
edits
More or less feature complete in the paper.
First draft of Sec 1.1

--- 2020-11-21 paper/intro.tex
@@
 \section{Introduction}
+
+This paper is about the interaction of two of the fundamental building
+blocks of computing: memory and sequential composition. One would like
+to think that these are well-worn topics, where every issue has been
+settled, but this is sadly not the case.
+
+\subsection{Memory}
+
+For single-threaded programs, memory can be thought of as you might
+expect: programs write to, and read from, memory references.
+This can be thought of as a total order of reads and writes,
+where each read has a matching \emph{fulfilling} write,
+for example:
+  \begin{gather*}
+    \THREAD{x\GETS0\SEMI x\GETS1\SEMI y\GETS2\SEMI
+    r\GETS y\SEMI s\GETS x}
+    \\[-.4ex]
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{wx0}{\DW{x}{0}}{}
+        \event{wx1}{\DW{x}{1}}{right=of wx0}
+        \event{wy2}{\DW{y}{2}}{right=of wx1}
+        \event{ry2}{\DR{y}{2}}{right=of wy2}
+        \event{rx1}{\DR{x}{1}}{right=of ry2}
+        \rf[out=20,in=160]{wy2}{ry2}
+        \rf[out=20,in=160]{wx1}{rx1}
+        \wk{wx0}{wx1}
+        \wk{wx1}{wy2}
+        \wk{wy2}{ry2}
+        \wk{ry2}{rx1}
+      \end{tikzinline}}
+  \end{gather*}
+This model naturally extends to the case of shared-memory concurrency
+in a natural way, leading to a \emph{sequentially consistent}
+semantics, in which \emph{program order} inside a thread implies
+a total \emph{causal order} between read and write events, for example:
+  \begin{gather*}
+    \THREAD{x\GETS0\SEMI x\GETS1\SEMI y\GETS2}
+    \PAR
+    \THREAD{r\GETS y\SEMI s\GETS x}
+    \\[-.4ex]
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{wx0}{\DW{x}{0}}{}
+        \event{wx1}{\DW{x}{1}}{right=of wx0}
+        \event{wy2}{\DW{y}{2}}{right=of wx1}
+        \event{ry2}{\DR{y}{2}}{right=of wy2}
+        \event{rx1}{\DR{x}{1}}{right=of ry2}
+        \rf[out=20,in=160]{wy2}{ry2}
+        \rf[out=20,in=160]{wx1}{rx1}
+        \wk{wx0}{wx1}
+        \wk{wx1}{wy2}
+        \wk{ry2}{rx1}
+      \end{tikzinline}}
+  \end{gather*}
+Unfortunately, this model does not compile efficiently to commodity
+hardware, resulting in a X\% increase in CPU time~\cite{???}, and
+hence power consumption.  Developers of software and compilers have
+therefore been faced with a difficult trade-off, between an elegant
+model of memory, and its impact on resource usage (such as size of
+data centres, electricity bills and carbon footprint). Unsurprisingly,
+many have chosen to prioritize efficiency over elegance.
+
+This has led to \emph{relaxed memory models}, in which the requirement of
+sequential consistency is weakened to only apply \emph{per-location} and not globally
+over the whole progfram. This allows executions which
+are inconsistent with program order, such as:
+  \begin{gather*}
+    \THREAD{x\GETS0\SEMI x\GETS1\SEMI y\GETS2}
+    \PAR
+    \THREAD{r\GETS y\SEMI s\GETS x}
+    \\[-.4ex]
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{wx0}{\DW{x}{0}}{}
+        \event{wx1}{\DW{x}{1}}{right=of wx0}
+        \event{wy2}{\DW{y}{2}}{right=of wx1}
+        \event{ry2}{\DR{y}{2}}{right=of wy2}
+        \event{rx0}{\DR{x}{0}}{right=of ry2}
+        \rf[out=20,in=160]{wy2}{ry2}
+        \rf[out=15,in=165]{wx0}{rx0}
+        \wk{wx0}{wx1}
+        \wk[out=200,in=340]{rx0}{wx1}
+      \end{tikzinline}}
+  \end{gather*}
+In such models, the causal order between events is important,
+and includes control and data dependencies, to avoid
+paradoxical ``out of thin air'' examples such as:
+  \begin{gather*}
+    \THREAD{r\GETS x\SEMI \IF r \THEN y\GETS1 \FI}
+    \PAR
+    \THREAD{s\GETS y\SEMI x\GETS s}
+    \\[-.4ex]
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{rx1}{\DR{x}{1}}{}
+        \event{wy1}{\DW{y}{1}}{right=of rx1}
+        \event{ry1}{\DR{y}{1}}{right=of wy1}
+        \event{wx1}{\DW{x}{1}}{right=of ry1}
+        \rf[out=20,in=160]{wy1}{ry1}
+        \rf[out=160,in=20]{wx1}{rx1}
+        \po{rx1}{wy1}
+        \po{ry1}{wx1}
+      \end{tikzinline}}
+  \end{gather*}
+This candidate execution forms a cycle in causal order, so is disallowed,
+but this depends crucially on the control dependency
+from $(\DR{x}{1})$ to $(\DW{y}{1})$, and the data dependency
+from $(\DR{y}{1})$ to $(\DW{x}{1})$. If either is missing, then this execution
+is acyclic and hence allowed. For example dropping the control dependency
+results in:
+  \begin{gather*}
+    \THREAD{r\GETS x\SEMI y\GETS1}
+    \PAR
+    \THREAD{s\GETS y\SEMI x\GETS s}
+    \\[-.4ex]
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{rx1}{\DR{x}{1}}{}
+        \event{wy1}{\DW{y}{1}}{right=of rx1}
+        \event{ry1}{\DR{y}{1}}{right=of wy1}
+        \event{wx1}{\DW{x}{1}}{right=of ry1}
+        \rf[out=20,in=160]{wy1}{ry1}
+        \rf[out=160,in=20]{wx1}{rx1}
+        \po{ry1}{wx1}
+      \end{tikzinline}}
+  \end{gather*}
+Unfortunately, while a simple syntactic approach to dependency calculation
+suffices for hardware models, it is not preserved by common compiler
+optimizarions. For example, if we calculate control dependencies syntactically,
+then there is a dependency from $(\DR{x}{1})$ to $(\DW{y}{1})$, and therefore a cycle in, the candidate execution:
+  \begin{gather*}
+    \THREAD{r\GETS x\SEMI \IF r \THEN y\GETS1 \ELSE y\GETS1 \FI}
+    \PAR
+    \THREAD{s\GETS y\SEMI x\GETS s}
+    \\[-.4ex]
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{rx1}{\DR{x}{1}}{}
+        \event{wy1}{\DW{y}{1}}{right=of rx1}
+        \event{ry1}{\DR{y}{1}}{right=of wy1}
+        \event{wx1}{\DW{x}{1}}{right=of ry1}
+        \rf[out=20,in=160]{wy1}{ry1}
+        \rf[out=160,in=20]{wx1}{rx1}
+        \po{rx1}{wy1}
+        \po{ry1}{wx1}
+      \end{tikzinline}}
+  \end{gather*}
+An optimizing compiler might lift the assignment $y\GETS1$ out of the conditional,
+thus removing the control dependency.
+
+Prominent solutions to the problem of dependency calculation include:
+\begin{itemize}
+
+\item \emph{syntactic} methods used in hardware models
+  such as ARM~\cite{???} or x86-TSO~\cite{???},
+\item \emph{alternate universe} methods (which give a semantics based on multiple speculative executions
+  of the same program) such as the Java Memory Model~\cite{???},
+  the speculative operational semantics of~\cite{???},
+  the promising semantics of~\cite{???},
+  or the event structures semantics of~\cite{???},
+\item \emph{rewriting} methods, which give an operational model
+  up to syntactic rewrites, such as~\cite{???}, and
+\item \emph{logical} methods, such as the pomsets with preconditions
+  model of~\cite{???}.
+
+\end{itemize}
+In this paper, we will focus on logical models, as those are compositional,
+and align well with existing models of sequential composition.
+
+\subsection{Sequential composition}
+
+%% Program logics and semantics tell us that when executing ((S1; S2), state0),
+%% we execute (S1, state0) to arrive at state1, then execute (S2, state1) to
+%% arrive at the final state2.
+
+%% This is a delightfully simple story that can be explained to children.  It is
+%% also a lie.
+
+%% Processors execute instructions out of order, due to pipelines and caches.
+%% Compilers reorder programs even more dramatically.  All of this reordering is
+%% meant to be unobservable in single-threaded code.  In multi-threaded code,
+%% however, all bets are off.  A formal attempt to understand the resulting mess
+%% is known as a ``relaxed memory model.''
+
+%% Most of the relaxed memory models that have been proposed are designed to
+%% help us understand whole program execution: they have no compositionality
+%% properties whatsoever.  Recently, denotation models have appeared that treat
+%% \emph{concurrent} execution compositionally.  One such model is ``Pomsets with
+%% Preconditions''.  It remains an open question, however, whether it is
+%% possible to treat \emph{sequential} execution compositionally in such a model,
+%% without overly restricting processors and compilers.
+
+%% We propose adding families of predicate transformers to Pomsets with
+%% Preconditions.  The resulting model is denotational, supporting both parallel
+%% and sequential composition.  When composing (S1;S2), the predicate
+%% transformer used to validate the precondition of an event in S2 is chosen
+%% based on the dependency order from S1 into this event.  As usual in work on
+%% relaxed memory, we have not handled loops or recursion.
+
+%% Happily, most of the results expected of a relaxed memory model can be
+%% established by appeal to prior work.  So here we are able to concentrate on
+%% the model itself.  The model is formalized in Agda, where we have established
+%% associativity for sequential composition.
+
+%% For the memory-model specialist, we retain the good properties of the prior
+%% work on Pomsets with Preconditions, fixing some errors along the way.  These
+%% properties include efficient implementation on ARMv8, support for compiler
+%% optimizations, support for logics that prove the absence of thin-air
+%% behaviors, and a local data race freedom theorem.
+
+% \begin{figure*}
+%   \input{fig-sp}
+%   \caption{Weakest Precondition and Strongest Postcondition \cite{}}
+% \end{figure*}
+
+Our approach follows that of weakest precondition semantics of
+\citet{DBLP:journals/cacm/Dijkstra75}, which provides an alternative
+characterization of Hoare logic \citep{Hoare:1969:ABC:363235.363259} by
+mapping postconditions to preconditions.
+
+% Our transformers $\aTr$ map preconditions to postconditions, in order to
+% apply to preconditions.
+% \begin{align*}
+%   \hoare{\aForm}{\bCmd}{\bForm}
+%   \;\;\Leftrightarrow\;\;
+%   \hoare{\aTr(\aCmd,\aForm)}{\aCmd\SEMI\bCmd}{\bForm}
+% \end{align*}
+% For example
+% \begin{gather*}
+%   \hoare{r{=}1}{\PW{y}{r}}{y{=}1}
+%   \\
+%   \hoare{x{=}1}{\PR{x}{r}\SEMI\PW{y}{r}}{y{=}1}
+% \end{gather*}
+
+\endinput
 

### 2021-01-23 paper/intro.tex
Fleshing out Sec 1.1

@@
 then there is a dependency from $(\DR{x}{1})$ to $(\DW{y}{1})$, and therefore a cycle in, the candidate execution:
-  \begin{gather*}
+`  \begin{gather*}
     \THREAD{r\GETS x\SEMI \IF r \THEN y\GETS1 \ELSE y\GETS1 \FI}
@@
 and align well with existing models of sequential composition.
-
+The heart of the model of~\cite{???} is to add logical preconditions
+to events, which are introduced by store (modelling data dependencies)
+or conditionals (modelling control dependencies):
+  \begin{gather*}
+    \IF{q \le r} \THEN z\GETS s \FI
+    \\
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{wz1}{(q {\le} r) \land (s{=}1) \mid \DW{z}{1}}{}
+      \end{tikzinline}}
+  \end{gather*}
+and are discharged by being ordered with a read event:
+  \begin{gather*}
+    r\GETS x\SEMI s\GETS y\SEMI \IF{r\le r} \THEN z\GETS s \FI
+    \\
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{rx1}{\DR{x}{1}}{}
+        \event{ry1}{\DR{y}{1}}{right=of rx1 }
+        \event{wz1}{(s{=}1) \limplies (r {\le} r) \land (s{=}1) \mid \DW{z}{1}}{right=of ry1}
+        \po{ry1}{wz1}
+      \end{tikzinline}}
+  \end{gather*}
+Note that there is dependency order from $(\DR{y}{1})$ to $(\DW{z}{1})$
+so the precondition for $(\DW{z}{1})$ only has to be satisfied assuming the hypothesis
+$(s=1)$. There is no matching order from $(\DR{x}{1})$ to $(\DW{z}{1})$
+which is why we do not assume the hypothesis $(r=1)$. The precondition on
+$(\DW{z}{1})$ is a tautology, and so can be elided:
+  \begin{gather*}
+    r\GETS x\SEMI s\GETS y\SEMI \IF{r\le r} \THEN z\GETS s \FI
+    \\
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{rx1}{\DR{x}{1}}{}
+        \event{ry1}{\DR{y}{1}}{right=of rx1 }
+        \event{wz1}{\DW{z}{1}}{right=of ry1}
+        \po{ry1}{wz1}
+      \end{tikzinline}}
+  \end{gather*}
 \subsection{Sequential composition}

### 2021-01-23 paper/intro.tex
Fleshing out Sec 1.1

@@
   \end{gather*}
+While existing models of relaxed memory have detailed treatmenmts of parallel composition,
+they pften give sequential composition little attention, either ignoring it altogether,
+or treating it operationally with its usual small-step semantics. This paper
+investigates how existing models of sequential composition interact with relaxed memory.
+
 \subsection{Sequential composition}

### 2021-01-23 paper/intro.tex
Fleshing out Sec 1.1

@@
   \begin{gather*}
-    \IF{q \le r} \THEN z\GETS s \FI
+    \IF{s{<}1} \THEN z\GETS r\&s \FI
     \\
@@
     \hbox{\begin{tikzinline}[node distance=1.5em]
-        \event{wz1}{(q {\le} r) \land (s{=}1) \mid \DW{z}{1}}{}
+        \event{wz0}{(s {<} 1) \land (r\&s){=}0 \mid \DW{z}{0}}{}
       \end{tikzinline}}
@@
   \begin{gather*}
-    r\GETS x\SEMI s\GETS y\SEMI \IF{r\le r} \THEN z\GETS s \FI
+    r\GETS x\SEMI s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r\&s \FI
     \\
@@
     \hbox{\begin{tikzinline}[node distance=1.5em]
-        \event{rx1}{\DR{x}{1}}{}
-        \event{ry1}{\DR{y}{1}}{right=of rx1 }
-        \event{wz1}{(s{=}1) \limplies (r {\le} r) \land (s{=}1) \mid \DW{z}{1}}{right=of ry1}
-        \po{ry1}{wz1}
+        \event{rx0}{\DR{x}{0}}{}
+        \event{ry0}{\DR{y}{0}}{right=of rx0}
+        \event{wz0}{(s{=}0) \limplies (s {<} 1) \land (r\&s){=}0 \mid \DW{z}{0}}{right=of ry0}
+        \po{ry0}{wz0}
       \end{tikzinline}}
   \end{gather*}
-Note that there is dependency order from $(\DR{y}{1})$ to $(\DW{z}{1})$
-so the precondition for $(\DW{z}{1})$ only has to be satisfied assuming the hypothesis
-$(s=1)$. There is no matching order from $(\DR{x}{1})$ to $(\DW{z}{1})$
-which is why we do not assume the hypothesis $(r=1)$. The precondition on
-$(\DW{z}{1})$ is a tautology, and so can be elided:
+Note that there is dependency order from $(\DR{y}{0})$ to $(\DW{z}{0})$
+so the precondition for $(\DW{z}{0})$ only has to be satisfied assuming the hypothesis
+$(s=0)$. There is no matching order from $(\DR{x}{0})$ to $(\DW{z}{0})$
+which is why we do not assume the hypothesis $(r=0)$. The precondition on
+$(\DW{z}{0})$ is a tautology, and so can be elided:
   \begin{gather*}
-    r\GETS x\SEMI s\GETS y\SEMI \IF{r\le r} \THEN z\GETS s \FI
+    r\GETS x\SEMI s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r\&s \FI
     \\
@@
     \hbox{\begin{tikzinline}[node distance=1.5em]
-        \event{rx1}{\DR{x}{1}}{}
-        \event{ry1}{\DR{y}{1}}{right=of rx1 }
-        \event{wz1}{\DW{z}{1}}{right=of ry1}
-        \po{ry1}{wz1}
+        \event{rx0}{\DR{x}{0}}{}
+        \event{ry0}{\DR{y}{0}}{right=of rx0}
+        \event{wz0}{\DW{z}{0}}{right=of ry0}
+        \po{ry0}{wz0}
       \end{tikzinline}}

### 2021-01-23 paper/arm.tex
Egits

@@
 
-\subsection{Down-Grading Acquires ($\xDGA$)}
+\subsection{Downgraded Reads ($\xDGR$)}
 \label{sec:internal}
@@
         \rf{a}{b}
-      %\sync{b}{c}
+        % \sync{b}{c}
         \wk{c}{d}
@@
   More generally, this can be understood as a compiler optimization that
-downgrades a read from $\mRA$ to $\mRLX$ when it read can be fulfilled by a
+  downgrades a read from $\mRA$ to $\mRLX$ when it can be fulfilled by a
   relaxed write in the same thread.
@@
 
-To model such \emph{downgraded acquires}, we use the uninterpreted symbols
+To model such \emph{downgraded reads}, we use the uninterpreted symbols
 $\Dx{\aLoc}$.
@@
 
-\begin{definition}[$\xCO$/$\xRASC$/$\xDGA$]
+\begin{definition}[$\xCO$/$\xRASC$/$\xDGR$]
   \label{def:pomsets-down}
@@
   One might worry that our model is too permissive for $\mSC$ access, but
-\armeight{} allows some counterintuitive results for SC access.  In the
-following execution we elide the initializing write $\DWP{y}{0}$.
+  \armeight{} itself allows some very counterintuitive results for $\mSC$
+  access.  In the following execution we elide the initializing write
+  $\DWP{y}{0}$.
   \begin{gather*}
@@
         \rf{b}{c}
-      %\sync{c}{d}
+        % \sync{c}{d}
         \sync{e}{f}
@@
 \begin{example}
+  \label{ex:rrd1}
   The following execution is allowed by \armeight{}, but disallowed by
@@
     \begin{gathered}
-  x\GETS0\SEMI 
-  x\GETS 1\SEMI y^\mRA\GETS1 \PAR r\GETS y\SEMI \IF{r}\THEN s\GETS x\FI
+      x\GETS0\SEMI x\GETS 1\SEMI y^\mRA\GETS1
+      \PAR
+      r\GETS y\SEMI \IF{r}\THEN s\GETS x\FI
       \\[-.4ex]
@@
         \sync{wx1}{wy1}
-      %\po{ry1}{rx0}
+        % \po{ry1}{rx0}
         \rf{wy1}{ry1}
@@
 
-We repeat all of the rules for preconditions and transformers,
-Only \ref{L5} and \ref{T3} are changed from \refdef{def:pomsets-down}.
-We repeat \ref{L4}, although it is unchanged from \refdef{def:pomsets-trans}.
+\begin{example}
+  Revisiting \refex{ex:rrd1}
+  \begin{align*}
+    \begin{gathered}
+      r\GETS y
+      \\[-.4ex]
+      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
+          \event{a}{\DR{y}{1}}{}
+          \xform{xd}{1{=}r\limplies\bForm}{below=of a}
+          \xform{xi}{(y{=}r\lor1{=}r)\limplies\bForm}{below=of xd}
+          \xo{a}{xd}
+        \end{tikzinline}}
+    \end{gathered}
+    &&
+    \begin{gathered}
+      \IF{r}\THEN s\GETS x\FI
+      \\[-.4ex]
+      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
+          \event{b}{r\mid\DR{x}{0}}{}
+        \end{tikzinline}}
+    \end{gathered}
+  \end{align*}
+\end{example}
 
-\begin{definition}[$\xCO$/$\xRASC$/$\xDGA$/$\xRRD$]
+\refdef{def:pomsets-down-rr} combines all of the features of
+\textsection\ref{sec:q}--\ref{sec:arm}:
+\begin{definition}[$\xCO$/$\xRASC$/$\xDGR$/$\xRRD$]
   \label{def:pomsets-down-rr}
@@
 \end{definition}
+Here, we have repeated all of the rules for preconditions and transformers,
+including \ref{L4}, which is unchanged from \refdef{def:pomsets-trans}.
+Only \ref{L5} and \ref{T3} are changed from \refdef{def:pomsets-down}.
 
@@
 
-%\subsection{Proof Strategy}
+% \subsection{Proof Strategy}
 

### 2021-01-23 paper/intro.tex
Egits

@@
 Unfortunately, this model does not compile efficiently to commodity
-hardware, resulting in a X\% increase in CPU time~\cite{???}, and
+hardware, resulting in a 37--73\% increase in CPU time~\cite{Liu:2019:ASC:3314221.3314611} on ARM, and
 hence power consumption.  Developers of software and compilers have
@@
 model of memory, and its impact on resource usage (such as size of
-data centres, electricity bills and carbon footprint). Unsurprisingly,
+data centers, electricity bills and carbon footprint). Unsurprisingly,
 many have chosen to prioritize efficiency over elegance.
@@
 sequential consistency is weakened to only apply \emph{per-location} and not globally
-over the whole progfram. This allows executions which
+over the whole program. This allows executions which
 are inconsistent with program order, such as:
@@
 suffices for hardware models, it is not preserved by common compiler
-optimizarions. For example, if we calculate control dependencies syntactically,
+optimizations. For example, if we calculate control dependencies syntactically,
 then there is a dependency from $(\DR{x}{1})$ to $(\DW{y}{1})$, and therefore a cycle in, the candidate execution:
-`  \begin{gather*}
+  \begin{gather*}
     \THREAD{r\GETS x\SEMI \IF r \THEN y\GETS1 \ELSE y\GETS1 \FI}
@@
 \item \emph{syntactic} methods used in hardware models
-  such as ARM~\cite{???} or x86-TSO~\cite{???},
+  such as ARM or x86-TSO \cite{alglave},
 \item \emph{alternate universe} methods (which give a semantics based on multiple speculative executions
-  of the same program) such as the Java Memory Model~\cite{???},
-  the speculative operational semantics of~\cite{???},
-  the promising semantics of~\cite{???},
-  or the event structures semantics of~\cite{???},
+  of the same program) such as the Java Memory Model~\cite{Manson:2005:JMM:1047659.1040336} 
+  and related models \cite{Jagadeesan:2010:GOS:2175486.2175503, DBLP:conf/popl/KangHLVD17, DBLP:journals/pacmpl/ChakrabortyV19},
+  % the speculative operational semantics of~\cite{Jagadeesan:2010:GOS:2175486.2175503},
+  % the promising semantics of~\cite{DBLP:conf/popl/KangHLVD17},
+  % or the event structures semantics of~\cite{DBLP:journals/pacmpl/ChakrabortyV19},
 \item \emph{rewriting} methods, which give an operational model
-  up to syntactic rewrites, such as~\cite{???}, and
+  up to syntactic rewrites, such as~\cite{Pichon-Pharabod:2016:CSR:2837614.2837616}, and
 \item \emph{logical} methods, such as the pomsets with preconditions
-  model of~\cite{???}.
+  model of~\cite{DBLP:journals/pacmpl/JagadeesanJR20}.
 
@@
 and align well with existing models of sequential composition.
-The heart of the model of~\cite{???} is to add logical preconditions
-to events, which are introduced by store (modelling data dependencies)
-or conditionals (modelling control dependencies):
+The heart of the model of~\cite{DBLP:journals/pacmpl/JagadeesanJR20} is to add logical preconditions
+to events, which are introduced by store actions (modeling data dependencies)
+and conditionals (modeling control dependencies):
   \begin{gather*}
@@
   \end{gather*}
-and are discharged by being ordered with a read event:
+Preconditions are discharged by being ordered after a read:
   \begin{gather*}
@@
 so the precondition for $(\DW{z}{0})$ only has to be satisfied assuming the hypothesis
-$(s=0)$. There is no matching order from $(\DR{x}{0})$ to $(\DW{z}{0})$
-which is why we do not assume the hypothesis $(r=0)$. The precondition on
-$(\DW{z}{0})$ is a tautology, and so can be elided:
+$(s{=}0)$. There is no matching order from $(\DR{x}{0})$ to $(\DW{z}{0})$
+which is why we do not assume the hypothesis $(r{=}0)$. The precondition on
+$(\DW{z}{0})$ is a tautology, and so can be elided in the diagram:
   \begin{gather*}
@@
 While existing models of relaxed memory have detailed treatmenmts of parallel composition,
-they pften give sequential composition little attention, either ignoring it altogether,
+they often give sequential composition little attention, either ignoring it altogether,
 or treating it operationally with its usual small-step semantics. This paper

### 2021-01-23 paper/intro.tex
Egits

@@
   such as ARM or x86-TSO \cite{alglave},
-\item \emph{alternate universe} methods (which give a semantics based on multiple speculative executions
+\item \emph{speculative execution} methods (which give a semantics based on multiple executions
   of the same program) such as the Java Memory Model~\cite{Manson:2005:JMM:1047659.1040336} 
@@
   \begin{gather*}
-    \IF{s{<}1} \THEN z\GETS r\&s \FI
+    \IF{s{<}1} \THEN z\GETS r{*}s \FI
     \\
@@
     \hbox{\begin{tikzinline}[node distance=1.5em]
-        \event{wz0}{(s {<} 1) \land (r\&s){=}0 \mid \DW{z}{0}}{}
+        \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
       \end{tikzinline}}
@@
   \begin{gather*}
-    r\GETS x\SEMI s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r\&s \FI
+    r\GETS x\SEMI s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r{*}s \FI
     \\
@@
         \event{ry0}{\DR{y}{0}}{right=of rx0}
-        \event{wz0}{(s{=}0) \limplies (s {<} 1) \land (r\&s){=}0 \mid \DW{z}{0}}{right=of ry0}
+        \event{wz0}{(s{=}0) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
         \po{ry0}{wz0}
@@
 $(s{=}0)$. There is no matching order from $(\DR{x}{0})$ to $(\DW{z}{0})$
-which is why we do not assume the hypothesis $(r{=}0)$. The precondition on
+which is why we do not assume the hypothesis $(r{=}0)$. Nonetheless, the precondition on
 $(\DW{z}{0})$ is a tautology, and so can be elided in the diagram:
   \begin{gather*}
-    r\GETS x\SEMI s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r\&s \FI
-    \\
+    % r\GETS x\SEMI s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r{*}s \FI
+    % \\
     \nonumber
@@
   \end{gather*}
-While existing models of relaxed memory have detailed treatmenmts of parallel composition,
+
+While existing models of relaxed memory have detailed treatments of parallel composition,
 they often give sequential composition little attention, either ignoring it altogether,

### 2021-01-24 paper/intro.tex
First draft of Sec 1.2

--- 2021-01-23 paper/intro.tex
@@
 
-% \begin{figure*}
-%   \input{fig-sp}
-%   \caption{Weakest Precondition and Strongest Postcondition \cite{}}
-% \end{figure*}
+\begin{figure*}
+  \input{fig-sp}
+  \caption{Weakest precondition semantics}
+  \label{fig:sp}
+\end{figure*}
 
@@
 characterization of Hoare logic \citep{Hoare:1969:ABC:363235.363259} by
-mapping postconditions to preconditions.
+mapping postconditions to preconditions. We recall the definition of
+$\fwp{\aCmd}{\bForm}$ in Figure~\ref{fig:sp}. The rule we are most interested
+in is the one for sequential composition, which maps seqential composition of programs
+to function composition of predicate transformers.
 
-% Our transformers $\aTr$ map preconditions to postconditions, in order to
-% apply to preconditions.
-% \begin{align*}
-%   \hoare{\aForm}{\bCmd}{\bForm}
-%   \;\;\Leftrightarrow\;\;
-%   \hoare{\aTr(\aCmd,\aForm)}{\aCmd\SEMI\bCmd}{\bForm}
-% \end{align*}
-% For example
-% \begin{gather*}
-%   \hoare{r{=}1}{\PW{y}{r}}{y{=}1}
-%   \\
-%   \hoare{x{=}1}{\PR{x}{r}\SEMI\PW{y}{r}}{y{=}1}
-% \end{gather*}
+Predicate transformers are a good fit to logical models of dependency calculation,
+since both are concerned with preconditions, and how they are transformed by
+sequential composition. Our first attempt is to associate a predicate transformer
+with each pomset. We visualize this in diagrams by showing how $\bForm$ is transformed,
+for example:
+  \begin{align*}
+    \begin{gathered}
+      r \GETS x
+      \\
+      \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
+          \event{rx0}{\DR{x}{0}}{}
+          \xform{rx0d}{(r{=}0) \limplies \bForm}{below=of rx0}
+          \xo{rx0}{rx0d}
+        \end{tikzinline}}
+    \end{gathered}
+    &&
+    \begin{gathered}
+      s \GETS y
+      \\
+      \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
+          \event{ry0}{\DR{y}{0}}{}
+          \xform{ry0d}{(s{=}0) \limplies \bForm}{below=of ry0}
+          \xo{ry0}{ry0d}
+        \end{tikzinline}}
+    \end{gathered}
+    &&
+    \begin{gathered}
+      \IF{s{<}1} \THEN z\GETS r\&s \FI
+      \\
+      \hbox{\begin{tikzinline}[node distance=1ex and1.5em]
+          \event{wz0}{(s {<} 1) \land (r\&s){=}0 \mid \DW{z}{0}}{}
+          \xform{wz0d}{\bForm}{below=of wz0}
+          \xo{wz0}{wz0d}
+      \end{tikzinline}}
+    \end{gathered}
+  \end{align*}
+For the sequentially consistent semantics, sequential composition composition
+is straightforward: we apply each predicate transformer to the preconditions
+of subsequent events, and compose the predicate transformers:
+  \begin{gather*}
+    r\GETS x\SEMI s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r\&s \FI
+    \\
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
+        \event{rx0}{\DR{x}{0}}{}
+        \event{ry0}{\DR{y}{0}}{right=of rx0}
+        \event{wz0}{(r{=}0) \limplies (s{=}0) \limplies (s {<} 1) \land (r\&s){=}0 \mid \DW{z}{0}}{right=of ry0}
+        \xform{rx0ry0d}{(r{=}0) \limplies(s{=}0) \limplies \bForm}{below=of ry0}
+        \po{rx0}{ry0}
+        \po{ry0}{wz0}
+        \xo{rx0}{rx0ry0d}
+        \xo{ry0}{rx0ry0d}
+        \xo{wz0}{rx0ry0d}
+      \end{tikzinline}}
+  \end{gather*}
+This model works for the sequentially consistent case, but needs to be
+weakened for the relaxed case. The key observation of this paper is
+that rather than working with one predicate transformer, we should
+work with a \emph{family} of predicate transformers, indexed by sets
+of events.
 
-\endinput
+For example, for single-event pomsets, there are two predicate
+transformers, since there are two subsets of any one-element set.
+We call the predicate transfomer for $\emptyset$ the \emph{independent}
+transformer, and the one indexed by $\{e\}$ the \emph{dependent} transformer.
+We visualize this by including more than one transformed predicate,
+with an edge leading to the dependent one. For example:
+  \begin{align*}
+    \begin{gathered}
+      r \GETS x
+      \\
+      \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
+          \event{rx0}{\DR{x}{0}}{}
+          \xform{rx0i}{\bForm}{above=of rx0}
+          \xform{rx0d}{(r{=}0) \limplies \bForm}{below=of rx0}
+          \xo{rx0}{rx0d}
+        \end{tikzinline}}
+    \end{gathered}
+    &&
+    \begin{gathered}
+      s \GETS y
+      \\
+      \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
+          \event{ry0}{\DR{y}{0}}{}
+          \xform{ry0i}{\bForm}{above=of ry0}
+          \xform{ry0d}{(s{=}0) \limplies \bForm}{below=of ry0}
+          \xo{ry0}{ry0d}
+        \end{tikzinline}}
+    \end{gathered}
+    &&
+    \begin{gathered}
+      \IF{s{<}1} \THEN z\GETS r\&s \FI
+      \\
+      \hbox{\begin{tikzinline}[node distance=1ex and1.5em]
+          \event{wz0}{(s {<} 1) \land (r\&s){=}0 \mid \DW{z}{0}}{}
+          \xform{wz0i}{\bForm}{above=of wz0}
+          \xform{wz0d}{\bForm}{below=of wz0}
+          \xo{wz0}{wz0d}
+      \end{tikzinline}}
+    \end{gathered}
+  \end{align*}
+The model of sequential composition then picks which
+predicate transformer to apply to an event's precondition by picking
+the one indexed by all the events before it in causal order.
 
-This paper builds on
-\cite{2019-sp} and
-\cite{10.1145/3428262}.
-
-Pomsets are a nice model of concurrent computation.
-Predicate transformers are a nice model of sequential computation.
-We show how these interact.
-
-We do this by attaching predicate transformers to \emph{configurations}.
-
-Thus the amount of transformation depends on which read events are seen in
-the configuration.
-
-We develop the theory of pomsets with predicate transformers, then apply the
-theory to model relaxed memory.
-
-Other things we could model include Map/Reduce and Fork/Join.
+For example, we can recover the expected semantics
+for the above example by choosing
+the predicate transformer which is independent of $(\DR x0)$
+but dependent on $(\DR y0)$, which is the transformer
+which maps $\bForm$ to $(s{=}0) \limplies \bForm$.
+  \begin{gather*}
+    r\GETS x\SEMI s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r\&s \FI
+    \\
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
+        \event{rx0}{\DR{x}{0}}{}
+        \event{ry0}{\DR{y}{0}}{right=of rx0}
+        \event{wz0}{(s{=}0) \limplies (s {<} 1) \land (r\&s){=}0 \mid \DW{z}{0}}{right=of ry0}
+        \xform{rx0ry0d}{(r{=}0) \limplies(s{=}0) \limplies \bForm}{below=of ry0}
+        \xform{rx0d}{(r{=}0) \limplies \bForm}{above=of rx0}
+        \xform{ry0d}{(s{=}0) \limplies \bForm}{right=of rx0d}
+        \xform{wz0i}{\bForm}{above=of wz0}
+        \po{ry0}{wz0}
+        \xo{rx0}{rx0d}
+        \xo{ry0}{ry0d}
+        \xo{rx0}{rx0ry0d}
+        \xo{ry0}{rx0ry0d}
+      \end{tikzinline}}
+  \end{gather*}

### 2021-01-24 paper/intro.tex
Fleshing out Sec 1.2

@@
   \end{gather*}
+As a sanity check, we can see that sequential composition is
+associative in this case, since it does not matter whether we
+associate to the left, with intermediate step:
+  \begin{gather*}
+    r\GETS x\SEMI s\GETS y
+    \\
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
+        \event{rx0}{\DR{x}{0}}{}
+        \event{ry0}{\DR{y}{0}}{right=of rx0}
+        \xform{rx0ry0d}{(r{=}0) \limplies(s{=}0) \limplies \bForm}{below=of ry0}
+        \xform{rx0d}{(r{=}0) \limplies \bForm}{above=of rx0}
+        \xform{ry0d}{(s{=}0) \limplies \bForm}{right=of rx0d}
+        \xform{wz0i}{\bForm}{right=of ry0d}
+        \xo{rx0}{rx0d}
+        \xo{ry0}{ry0d}
+        \xo{rx0}{rx0ry0d}
+        \xo{ry0}{rx0ry0d}
+      \end{tikzinline}}
+  \end{gather*}
+or to the right, with intermediate step:
+    \begin{gather*}
+    s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r\&s \FI
+    \\
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
+        \event{ry0}{\DR{y}{0}}{}
+        \event{wz0}{(s{=}0) \limplies (s {<} 1) \land (r\&s){=}0 \mid \DW{z}{0}}{right=of ry0}
+        \xform{ry0d}{(s{=}0) \limplies \bForm}{above=of ry0}
+        \xform{wz0i}{\bForm}{above=of wz0}
+        \po{ry0}{wz0}
+        \xo{ry0}{ry0d}
+      \end{tikzinline}}
+  \end{gather*}
+This is an instance of a general result that sequential composition forms a monoid,
+as one would hope.

### 2021-01-24 paper/intro.tex
First draft of Sec 1.3

@@
 as one would hope.
+
+\subsection{Contributions}
+
+This paper is the first model of relaxed memory with a compositional
+semantics for sequential composition.  It shows how the pomsets with
+preconditions model of~\cite{DBLP:journals/pacmpl/JagadeesanJR20} can be combined
+with the predicate transformer model of~\cite{???}.
+\begin{itemize}
+\item \S\ref{sec:model} presents the basic model, with few features
+  required of the logic of preconditions, but a resulting laack of fidelity
+  to exiting models,
+\item \S\ref{sec:q} adds a model of \emph{quiescence} to the logic,
+  required to model coherence (accessing $x$ has a precondition that $x$ is quiescent)
+  and synchronization (a releasing write requires all locations to be quiescent),
+\item \S\ref{sec:arm} adds the features required for efficient compilation
+  to modern architectures: downgrading some synchronized accesses to relaxed,
+  and removing read-read dependencies,
+\item \S\ref{sec:complications} show how to address common litmus tests, and
+\item \S\ref{sec:discussion} is a discussion of the design space.
+\end{itemize}
+% Acknowledgements go here, once we're not double-blinded.

### 2021-01-24 paper/abstract.tex
Trimmed abstract

@@
+This paper presents the first semantics for relaxed memory with a
+compositional definition of sequential composition.  Previous
+definitions of relaxed memory have given detailed treatments of
+parallel composition, but have given sequential composition less
+attention, often relegating it to a (sometimes speculative)
+operational semantics of single-threaded programs. In this paper we show how
+sequential composition can be restored to a first-class citizen, by giving
+it a denotational semantics in a model of pomsets with preconditions,
+extended with a family of predicate transformers. Previous work has shown that
+pomsets with preconditions are a model of concurrent composition,
+and that predicate transformers are a model of sequential composition.
+This is the first paper to show how they can be combined.
+
 % Program logics and semantics tell us that in order to derive
@@
 % \end{math}
-Program logics and semantics tell us that when executing ((S1; S2), state0),
-we execute (S1, state0) to arrive at state1, then execute (S2, state1) to
-arrive at the final state2.
+%% Program logics and semantics tell us that when executing ((S1; S2), state0),
+%% we execute (S1, state0) to arrive at state1, then execute (S2, state1) to
+%% arrive at the final state2.
 
-This is a delightfully simple story that can be explained to children.  It is
-also a lie.
+%% This is a delightfully simple story that can be explained to children.  It is
+%% also a lie.
 
-Processors execute instructions out of order, due to pipelines and caches.
-Compilers reorder programs even more dramatically.  All of this reordering is
-meant to be unobservable in single-threaded code.  In multi-threaded code,
-however, all bets are off.  A formal attempt to understand the resulting mess
-is known as a ``relaxed memory model.''
+%% Processors execute instructions out of order, due to pipelines and caches.
+%% Compilers reorder programs even more dramatically.  All of this reordering is
+%% meant to be unobservable in single-threaded code.  In multi-threaded code,
+%% however, all bets are off.  A formal attempt to understand the resulting mess
+%% is known as a ``relaxed memory model.''
 
-Most of the relaxed memory models that have been proposed are designed to
-help us understand whole program execution: they have no compositionality
-properties whatsoever.  Recently, denotation models have appeared that treat
-\emph{concurrent} execution compositionally.  One such model is ``Pomsets with
-Preconditions''.  It remains an open question, however, whether it is
-possible to treat \emph{sequential} execution compositionally in such a model,
-without overly restricting processors and compilers.
+%% Most of the relaxed memory models that have been proposed are designed to
+%% help us understand whole program execution: they have no compositionality
+%% properties whatsoever.  Recently, denotation models have appeared that treat
+%% \emph{concurrent} execution compositionally.  One such model is ``Pomsets with
+%% Preconditions''.  It remains an open question, however, whether it is
+%% possible to treat \emph{sequential} execution compositionally in such a model,
+%% without overly restricting processors and compilers.
 
-We propose adding families of predicate transformers to Pomsets with
-Preconditions.  The resulting model is denotational, supporting both parallel
-and sequential composition.  When composing (S1;S2), the predicate
-transformer used to validate the precondition of an event in S2 is chosen
-based on the dependency order from S1 into this event.  As usual in work on
-relaxed memory, we have not handled loops or recursion.
+%% We propose adding families of predicate transformers to Pomsets with
+%% Preconditions.  The resulting model is denotational, supporting both parallel
+%% and sequential composition.  When composing (S1;S2), the predicate
+%% transformer used to validate the precondition of an event in S2 is chosen
+%% based on the dependency order from S1 into this event.  As usual in work on
+%% relaxed memory, we have not handled loops or recursion.
 
-Happily, most of the results expected of a relaxed memory model can be
-established by appeal to prior work.  So here we are able to concentrate on
-the model itself.  The model is formalized in Agda, where we have established
-associativity for sequential composition.
+%% Happily, most of the results expected of a relaxed memory model can be
+%% established by appeal to prior work.  So here we are able to concentrate on
+%% the model itself.  The model is formalized in Agda, where we have established
+%% associativity for sequential composition.
 
-For the memory-model specialist, we retain the good properties of the prior
-work on Pomsets with Preconditions, fixing some errors along the way.  These
-properties include efficient implementation on ARMv8, support for compiler
-optimizations, support for logics that prove the absence of thin-air
-behaviors, and a local data race freedom theorem.
+%% For the memory-model specialist, we retain the good properties of the prior
+%% work on Pomsets with Preconditions, fixing some errors along the way.  These
+%% properties include efficient implementation on ARMv8, support for compiler
+%% optimizations, support for logics that prove the absence of thin-air
+%% behaviors, and a local data race freedom theorem.

### 2021-01-24 paper/intro.tex
Running example now uses multiplication
Pass through intro

@@
 settled, but this is sadly not the case.
+
+Let $\aReg$--$\bReg$ range over thread-local registers and $\aLoc$-$\cLoc$
+range over shared memory references.
 
@@
         \rf[out=20,in=160]{wx1}{rx1}
-        \wk{wx0}{wx1}
-        \wk{wx1}{wy2}
-        \wk{wy2}{ry2}
-        \wk{ry2}{rx1}
+        \po{wx0}{wx1}
+        \po{wx1}{wy2}
+        \po{wy2}{ry2}
+        \po{ry2}{rx1}
       \end{tikzinline}}
   \end{gather*}
+
 This model naturally extends to the case of shared-memory concurrency
@@
         \rf[out=20,in=160]{wx1}{rx1}
-        \wk{wx0}{wx1}
-        \wk{wx1}{wy2}
-        \wk{ry2}{rx1}
+        \po{wx0}{wx1}
+        \po{wx1}{wy2}
+        \po{ry2}{rx1}
       \end{tikzinline}}
   \end{gather*}
+
 Unfortunately, this model does not compile efficiently to commodity
@@
         \rf[out=15,in=165]{wx0}{rx0}
-        \wk{wx0}{wx1}
-        \wk[out=200,in=340]{rx0}{wx1}
+        \po{wx0}{wx1}
+        \po[out=200,in=340]{rx0}{wx1}
       \end{tikzinline}}
   \end{gather*}
+
 In such models, the causal order between events is important,
@@
   \end{gather*}
+
 Unfortunately, while a simple syntactic approach to dependency calculation
@@
 \end{itemize}
+
 In this paper, we will focus on logical models, as those are compositional,
@@
 
-\begin{figure*}
-  \input{fig-sp}
-  \caption{Weakest precondition semantics}
-  \label{fig:sp}
-\end{figure*}
+% \begin{figure*}
+%   \input{fig-sp}
+%   \caption{Weakest precondition semantics}
+%   \label{fig:sp}
+% \end{figure*}
 
@@
 mapping postconditions to preconditions. We recall the definition of
-$\fwp{\aCmd}{\bForm}$ in Figure~\ref{fig:sp}. The rule we are most interested
-in is the one for sequential composition, which maps seqential composition of programs
+$\fwp{\aCmd}{\bForm}$ for loop-free code below. % in Figure~\ref{fig:sp}
+\begin{itemize}
+\item
+  \begin{math}
+    \fwp{\SKIP}{\bForm} = \bForm
+  \end{math}
+\item 
+  \begin{math}
+    \fwp{\ABORT}{\bForm} = \FALSE
+  \end{math}
+\item
+  \begin{math}
+    \fwp{r\GETS \aExp}{\bForm} = \bForm[\aExp/r]
+  \end{math}
+\item
+  \begin{math}
+    \fwp{\aCmd_1;\aCmd_2}{\bForm} = \fwp{\aCmd_1}{\fwp{\aCmd_2}{\bForm}}
+  \end{math}
+\item
+  \begin{math}
+    \fwp{\IF{\aExp}\THEN \aCmd_1\ELSE \aCmd_2\FI}{\bForm}= {}
+  \end{math}
+  \\
+  \begin{math}
+    ((\aExp{\ne}0) \limplies \fwp{\aCmd_1}{\bForm}) \land
+    ((\aExp{=}0) \limplies \fwp{\aCmd_2}{\bForm})
+  \end{math}
+\end{itemize}
+The rule we are most interested
+in is the one for sequential composition, which maps sequential composition of programs
 to function composition of predicate transformers.
@@
     \begin{gathered}
-      \IF{s{<}1} \THEN z\GETS r\&s \FI
+      \IF{s{<}1} \THEN z\GETS r{*}s \FI
       \\
       \hbox{\begin{tikzinline}[node distance=1ex and1.5em]
-          \event{wz0}{(s {<} 1) \land (r\&s){=}0 \mid \DW{z}{0}}{}
+          \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
           \xform{wz0d}{\bForm}{below=of wz0}
@@
   \end{align*}
-For the sequentially consistent semantics, sequential composition composition
+In the rightmost program above, the write to $z$ affects the shared store, not the
+local state of the thread, therefore we assign it the identity transformer.
+
+For the sequentially consistent semantics, sequential composition 
 is straightforward: we apply each predicate transformer to the preconditions
@@
         \event{ry0}{\DR{y}{0}}{right=of rx0}
-        \event{wz0}{(r{=}0) \limplies (s{=}0) \limplies (s {<} 1) \land (r\&s){=}0 \mid \DW{z}{0}}{right=of ry0}
+        \event{wz0}{(r{=}0) \limplies (s{=}0) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
         \xform{rx0ry0d}{(r{=}0) \limplies(s{=}0) \limplies \bForm}{below=of ry0}
@@
 transformers, since there are two subsets of any one-element set.
-We call the predicate transfomer for $\emptyset$ the \emph{independent}
+We call the predicate transformer for $\emptyset$ the \emph{independent}
 transformer, and the one indexed by $\{e\}$ the \emph{dependent} transformer.
@@
     \begin{gathered}
-      \IF{s{<}1} \THEN z\GETS r\&s \FI
+      \IF{s{<}1} \THEN z\GETS r{*}s \FI
       \\
       \hbox{\begin{tikzinline}[node distance=1ex and1.5em]
-          \event{wz0}{(s {<} 1) \land (r\&s){=}0 \mid \DW{z}{0}}{}
+          \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
           \xform{wz0i}{\bForm}{above=of wz0}
@@
         \event{ry0}{\DR{y}{0}}{right=of rx0}
-        \event{wz0}{(s{=}0) \limplies (s {<} 1) \land (r\&s){=}0 \mid \DW{z}{0}}{right=of ry0}
+        \event{wz0}{(s{=}0) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
         \xform{rx0ry0d}{(r{=}0) \limplies(s{=}0) \limplies \bForm}{below=of ry0}
@@
     \begin{gather*}
-    s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r\&s \FI
+    s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r{*}s \FI
     \\
@@
         \event{ry0}{\DR{y}{0}}{}
-        \event{wz0}{(s{=}0) \limplies (s {<} 1) \land (r\&s){=}0 \mid \DW{z}{0}}{right=of ry0}
+        \event{wz0}{(s{=}0) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
         \xform{ry0d}{(s{=}0) \limplies \bForm}{above=of ry0}
@@
 preconditions model of~\cite{DBLP:journals/pacmpl/JagadeesanJR20} can be combined
-with the predicate transformer model of~\cite{???}.
+with the predicate transformer model of~\cite{DBLP:journals/cacm/Dijkstra75}.
 \begin{itemize}
 \item \S\ref{sec:model} presents the basic model, with few features
-  required of the logic of preconditions, but a resulting laack of fidelity
+  required of the logic of preconditions, but a resulting lack of fidelity
   to exiting models,

### 2021-01-24 paper/arm.tex
Ripped x out of examples

@@
   \item[\ref{S4})]
-    $\aTr{\bEvs}{\bForm}$ implies $(\Qw{\aLoc}\limplies\aExp{\neq}\aVal)\land \DS{\aLoc}{\amode}\bForm[\aExp/\aLoc]$,
+    $\aTr{\bEvs}{\bForm}$ implies $(\Qw{\aLoc}\limplies\aExp{\neq}\aVal)\land \DS{\aLoc}{\amode}\bForm\noSUB{[\aExp/\aLoc]}$,
   \item[\ref{S5})]
-    $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qw{\aLoc}\land \DS{\aLoc}{\amode}\bForm[\aExp/\aLoc]$.
+    $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qw{\aLoc}\land \DS{\aLoc}{\amode}\bForm\noSUB{[\aExp/\aLoc]}$.
   \item[\ref{L5})]
-    $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode} \land((\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm)$,
-  \item[\ref{L6})]
+  %   $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode} \land((\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm)$,
+  % \item[\ref{L6})]
     $\aTr{\dEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode} \land\bForm$.
@@
   \item[\ref{L5})]
-    $\aTr{\cEvs}{\bForm}$ implies $\RW\limplies(\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm$,
+    %$\aTr{\cEvs}{\bForm}$ implies $\RW\limplies(\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm$,
+    $\aTr{\cEvs}{\bForm}$ implies $\RW\limplies\bForm$,
   \end{enumerate}  
@@
 
-\refdef{def:pomsets-down-rr} combines all of the features of
-\textsection\ref{sec:q}--\ref{sec:arm}:
+\subsection{Full semantics for ARM}
+
+\refdef{def:pomsets-arm} combines all of the features of
+\textsection\ref{sec:q}--\ref{sec:arm}.  For completeness, we repeat
+\ref{L4}, which is unchanged from \refdef{def:pomsets-trans}.
+%Only \ref{L5} and \ref{T3} are changed from \refdef{def:pomsets-down}.
+
 \begin{definition}[$\xCO$/$\xRASC$/$\xDGR$/$\xRRD$]
-  \label{def:pomsets-down-rr}
+  \label{def:pomsets-arm}
   Update \refdef{def:pomsets-trans} to:
@@
   \item[\ref{S4})]
-    $\aTr{\bEvs}{\bForm}$ implies $(\Qw{\aLoc}\limplies\aExp{\neq}\aVal)\land \DS{\aLoc}{\amode}\bForm[\aExp/\aLoc]$,
+    $\aTr{\bEvs}{\bForm}$ implies $(\Qw{\aLoc}\limplies\aExp{\neq}\aVal)\land \DS{\aLoc}{\amode}\bForm\noSUB{[\aExp/\aLoc]}$,
   \item[\ref{S5})]
-    $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qw{\aLoc}\land \DS{\aLoc}{\amode}\bForm[\aExp/\aLoc]$.
+    $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qw{\aLoc}\land \DS{\aLoc}{\amode}\bForm\noSUB{[\aExp/\aLoc]}$.
   \item[\ref{L4})]
@@
   \item[\ref{L5})]
-    $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode} \land(\RW\limplies(\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm)$,
-  \item[\ref{L6})]
-    $\aTr{\dEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode} \land\bForm$.
+  %   $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode} \land(\RW\limplies(\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm)$,
+  % \item[\ref{L6})]
+    $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode}
+    \land (\RW\limplies\bForm)$.
   \end{enumerate}  
 \end{definition}
-Here, we have repeated all of the rules for preconditions and transformers,
-including \ref{L4}, which is unchanged from \refdef{def:pomsets-trans}.
-Only \ref{L5} and \ref{T3} are changed from \refdef{def:pomsets-down}.
+\begin{proposition}
+  Every \armeight{} execution is allowed by \refdef{def:pomsets-arm}.
+  \begin{proof}[Proof Sketch]
+    By \EGC{}, an \armeight{} execution linearization of local-order and
+    per-location program order.  Every such linearization is also a valid
+    pomset under \refdef{def:pomsets-arm}.
+  \end{proof}
+\end{proposition}
 

### 2021-01-24 paper/intro.tex
Added paragraph to intro

@@
 This paper is the first model of relaxed memory with a compositional
-semantics for sequential composition.  It shows how the pomsets with
-preconditions model of~\cite{DBLP:journals/pacmpl/JagadeesanJR20} can be combined
-with the predicate transformer model of~\cite{DBLP:journals/cacm/Dijkstra75}.
+semantics for sequential composition.  It shows how pomsets with
+preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} can be combined
+with predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75}.
 \begin{itemize}
@@
 % Acknowledgements go here, once we're not double-blinded.
+The definitions in this paper have been formalized in Agda.
+
+Because it is closely related, we expect that the memory-model results of
+\cite{DBLP:journals/pacmpl/JagadeesanJR20} apply to our model, including
+compositional reasoning for temporal safety properties and {local} \drfsc.
+In \textsection\ref{sec:arm}, we provide an alternative proof strategy for
+efficient compilation to \armeight{}, which improves upon that of
+\cite{DBLP:journals/pacmpl/JagadeesanJR20} by using a recent alternative
+characterization of \armeight{}.

### 2021-01-24 paper/arm.tex
FOrward reference in 2.1
Related work paragraph

@@
 We discuss \armeight{} using \emph{external global completion} (\EGC{})
-\cite{alglave-git-alternate,armed-cats}, which is very close to our model;
-see also \cite[\textsection B2.3.6]{arm-reference-manual}.
+\cite{alglave-git-alternate} %,armed-cats},
+\cite[\textsection B2.3.6]{arm-reference-manual}
+which is very close to our model.
 
@@
 \begin{example}
+  \label{ex:dgr1}
   The following execution is allowed by \armeight{}, but disallowed by
@@
 To model such \emph{downgraded reads}, we use the uninterpreted symbols
-$\Dx{\aLoc}$.
+$\Dx{\aLoc}$.  Load actions that requiring downgrading introduce $\Dx{\aLoc}$.
+Relaxed stores on $\aLoc$ substitute true for $\Dx{\aLoc}$, whereas
+synchronizing stores substitute false for $\Dx{\aLoc}$.
 
@@
 \end{definition}
+
+\begin{example}
+  Revisiting \refex{ex:dgr1}.
+  \begin{gather*}
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{a}{\DW{x}{2}}{}
+        \raevent{b}{\DR[\mRA]{x}{2}}{right=of a}
+        \event{c}{\DW{y}{1}}{right=of b}
+        \event{d}{\DW{y}{2}}{right=2.5em of c}
+        \raevent{e}{\DW[\mRA]{x}{1}}{right=of d}
+        \rf{a}{b}
+        % \sync{b}{c}
+        \wk{c}{d}
+        \sync{d}{e}
+        \wk[out=-165,in=-15]{e}{a}
+      \end{tikzinline}}  
+  \end{gather*}  
+\end{example}
 
@@
   \item[\ref{L5})]
-    %$\aTr{\cEvs}{\bForm}$ implies $\RW\limplies(\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm$,
+    % $\aTr{\cEvs}{\bForm}$ implies $\RW\limplies(\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm$,
     $\aTr{\cEvs}{\bForm}$ implies $\RW\limplies\bForm$,
@@
 \ref{L4}, which is unchanged from \refdef{def:pomsets-trans}.
-%Only \ref{L5} and \ref{T3} are changed from \refdef{def:pomsets-down}.
+% Only \ref{L5} and \ref{T3} are changed from \refdef{def:pomsets-down}.
 
@@
 \end{definition}
-\begin{proposition}
-  Every \armeight{} execution is allowed by \refdef{def:pomsets-arm}.
-  \begin{proof}[Proof Sketch]
-    By \EGC{}, an \armeight{} execution linearization of local-order and
-    per-location program order.  Every such linearization is also a valid
-    pomset under \refdef{def:pomsets-arm}.
-  \end{proof}
-\end{proposition}
+
+Every \armeight{} execution is allowed by \refdef{def:pomsets-arm}.  The
+proof of this fact is simplified by the recent characterization of
+\armeight{} in terms of \EGC{} \cite[\textsection
+B2.3.6]{arm-reference-manual}.  Under \EGC{}, an \armeight{} execution is a
+linearization of per-location program order and a subset of local-order.
+Every such linearization is also a valid pomset under
+\refdef{def:pomsets-arm}.
 

### 2021-01-24 paper/arm.tex
downgrade
x
edits

@@
   for every $\aLoc$.
-  Let $\DS{\aLoc}{\amode}$ and $\DL{\aLoc}{\amode}$ be defined:
+  Let $\DS{\aLoc}{\amode}{}$ and $\DL{\aLoc}{\amode}{}$ be defined:
   \begin{align*}
     \DS{\aLoc}{\mRLX}{\bForm}&=\bForm[\TRUE/\Dx{\aLoc}] 
-    &\DL{\aLoc}{\mRLX}&=\TRUE
+    &\DL{\aLoc}{\mRLX}{\bForm}&=\bForm%\TRUE
     \\
     \DS{\aLoc}{\mRA}{\bForm}&=\bForm[\FALSE/\D]
-    &\DL{\aLoc}{\mRA}&=\Dx{\aLoc}
+    &\DL{\aLoc}{\mRA}{\bForm}&=\bForm\land\Dx{\aLoc}
     \\
     \DS{\aLoc}{\mSC}{\bForm}&=\bForm[\FALSE/\D]
-    &\DL{\aLoc}{\mSC}&=\Dx{\aLoc}
+    &\DL{\aLoc}{\mSC}{\bForm}&=\bForm\land\Dx{\aLoc}
   \end{align*}
@@
   \begin{enumerate}
+  % \item[\ref{S4})]
+  %   $\aTr{\bEvs}{\bForm}$ implies $(\Qw{\aLoc}\limplies\aExp{\neq}\aVal)\land \DS{\aLoc}{\amode}\bForm\noSUB{[\aExp/\aLoc]}$,
+  % \item[\ref{S5})]
+  %   $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qw{\aLoc}\land \DS{\aLoc}{\amode}\bForm\noSUB{[\aExp/\aLoc]}$.
+  % \item[\ref{L5})]
+  %   % $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode}{} \land((\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm)$,
+  %   % \item[\ref{L6})]
+  %   $\aTr{\dEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode}{} \land\bForm$.
   \item[\ref{S4})]
-    $\aTr{\bEvs}{\bForm}$ implies $(\Qw{\aLoc}\limplies\aExp{\neq}\aVal)\land \DS{\aLoc}{\amode}\bForm\noSUB{[\aExp/\aLoc]}$,
+    $\aTr{\bEvs}{\bForm}$ implies $\DS{\aLoc}{\amode}{\bForm[(\Qw{\aLoc}\land\aExp{=}\aVal)/\Qw{\aLoc}]}$,
   \item[\ref{S5})]
-    $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qw{\aLoc}\land \DS{\aLoc}{\amode}\bForm\noSUB{[\aExp/\aLoc]}$.
+    $\aTr{\cEvs}{\bForm}$ implies $\DS{\aLoc}{\amode}{\bForm[\FALSE/\Qw{\aLoc}]}$.
   \item[\ref{L5})]
-    % $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode} \land((\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm)$,
-    % \item[\ref{L6})]
-    $\aTr{\dEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode} \land\bForm$.
+    $\aTr{\dEvs}{\bForm}$ implies $\DL{\aLoc}{\amode}{\bForm[\FALSE/\Qx{\aLoc}]}$.
   \end{enumerate}
@@
 \begin{example}
-  Revisiting \refex{ex:dgr1}.
-  \begin{gather*}
-    \hbox{\begin{tikzinline}[node distance=1.5em]
+  Revisiting \refex{ex:dgr1} and eliding irrelevant transformers:
+  \begin{align*}
+    \begin{gathered}[t]
+      \PW{x}{2}
+      \\
+      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
+          \event{a}{\DW{x}{2}}{}
+          \xform{x1d}{\bForm[\TRUE/\Dx{\aLoc}]}{below=of a}
+          % \xform{x1i}{\lnot\Qw{x}\land\bForm[\TRUE/\Dx{x}]}{below=of x1d}
+          \xo{a}{x1d}
+        \end{tikzinline}}  
+    \end{gathered}  
+    &&
+    \begin{gathered}[t]
+      \PR[\mRA]{x}{r}
+      \\
+      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
+          \raevent{b}{\DR[\mRA]{x}{2}}{}
+          % \xform{x1d}{2{=}r\limplies\bForm}{below=of b}
+          \xform{x1i}{\Dx{x} \land\bForm[\FALSE/\Qx{x}]}{below=of b} %x1d}
+          % \xo{b}{x1d}
+        \end{tikzinline}}  
+    \end{gathered}  
+    &&
+    \begin{gathered}[t]
+      \PW{y}{1}
+      \\
+      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
+          \event{c}{\DW{y}{1}}{}
+          % \xform{x1d}{\bForm[\TRUE/\Dx{\aLoc}]}{below=of c}
+          % \xform{x1i}{\lnot\Qw{y}\land\bForm[\TRUE/\Dx{y}]}{below=of x1d}
+          % \xo{c}{x1d}
+        \end{tikzinline}}  
+    \end{gathered}  
+  \end{align*}
+  Associating right:
+  \begin{align*}
+    \begin{gathered}[t]
+      \PW{x}{2}
+      \\
+      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
+          \event{a}{\DW{x}{2}}{}
+          \xform{x1d}{\bForm[\TRUE/\Dx{\aLoc}]}{right=.7em of a}
+          % \xform{x1i}{\lnot\Qw{x}\land\bForm[\TRUE/\Dx{x}]}{below=of x1d}
+          \xo{a}{x1d}
+        \end{tikzinline}}  
+    \end{gathered}  
+    &&
+    \begin{gathered}[t]
+      \PR[\mRA]{x}{r}
+      \SEMI
+      \PW{y}{1}
+      \\
+      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
+          \raevent{b}{\DR[\mRA]{x}{2}}{}
+          \event{c}{\Dx{x}\mid\DW{y}{1}}{right=.7em of b}
+        \end{tikzinline}}  
+    \end{gathered}  
+  \end{align*}
+  Composing, we have, as desired:
+  \begin{align*}
+    \begin{gathered}[t]
+      \PW{x}{2}
+      \SEMI
+      \PR[\mRA]{x}{r}
+      \SEMI
+      \PW{y}{1}
+      \\
+      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
           \event{a}{\DW{x}{2}}{}
@@
           \event{c}{\DW{y}{1}}{right=of b}
-        \event{d}{\DW{y}{2}}{right=2.5em of c}
-        \raevent{e}{\DW[\mRA]{x}{1}}{right=of d}
           \rf{a}{b}
-        % \sync{b}{c}
-        \wk{c}{d}
-        \sync{d}{e}
-        \wk[out=-165,in=-15]{e}{a}
         \end{tikzinline}}  
-  \end{gather*}  
+    \end{gathered}  
+  \end{align*}
 \end{example}
@@
 \begin{example}
-  Revisiting \refex{ex:rrd1}
+  Revisiting \refex{ex:rrd1} and eliding irrelevant transformers:
   \begin{align*}
-    \begin{gathered}
-      r\GETS y
-      \\[-.4ex]
-      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-          \event{a}{\DR{y}{1}}{}
-          \xform{xd}{1{=}r\limplies\bForm}{below=of a}
-          \xform{xi}{(y{=}r\lor1{=}r)\limplies\bForm}{below=of xd}
-          \xo{a}{xd}
+    \begin{gathered}[t]
+      \PR{y}{r}
+      \\
+      \hbox{\begin{tikzinline}[node distance=0.5em and .5em]
+          \event{b}{\DR{y}{1}}{}
+          \xform{x1i}{\RW\limplies\bForm}{right=of b} %x1d}
         \end{tikzinline}}  
@@
     &&
-    \begin{gathered}
+    \begin{gathered}[t]
       \IF{r}\THEN s\GETS x\FI
-      \\[-.4ex]
+      \\
       \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-          \event{b}{r\mid\DR{x}{0}}{}
+          \event{c}{r\mid\DW{x}{0}}{}
+          % \xform{x1d}{\bForm[\TRUE/\Dx{\aLoc}]}{below=of c}
+          % \xform{x1i}{\lnot\Qw{y}\land\bForm[\TRUE/\Dx{y}]}{below=of x1d}
+          % \xo{c}{x1d}
+        \end{tikzinline}}  
+    \end{gathered}  
+  \end{align*}
+  Composing sequentially:
+  \begin{align*}
+    \begin{gathered}[t]
+      \PR{y}{r}
+      \SEMI
+      \IF{r}\THEN s\GETS x\FI
+      \\
+      \hbox{\begin{tikzinline}[node distance=0.5em and .5em]
+          \event{b}{\DR{y}{1}}{}
+          \event{c}{\RW\limplies r\mid\DW{x}{0}}{right=of b}
+          %\xform{x1i}{\RW\limplies\bForm}{right=of b} %x1d}
+        \end{tikzinline}}  
+    \end{gathered}  
+  \end{align*}
+  Embedding the thread in thread group, we have, as desired:
+  \begin{align*}
+    \begin{gathered}[t]
+      \PR{y}{r}
+      \SEMI
+      \IF{r}\THEN s\GETS x\FI
+      \\
+      \hbox{\begin{tikzinline}[node distance=0.5em and .5em]
+          \event{b}{\DR{y}{1}}{}
+          \event{c}{\FALSE\limplies r\mid\DW{x}{0}}{right=of b}
+          %\xform{x1i}{\RW\limplies\bForm}{right=of b} %x1d}
         \end{tikzinline}}  
@@
     % \item[\ref{L6})]
-    $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode}
+    $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode}{}
     \land (\RW\limplies\bForm)$.

### 2021-01-25 paper/intro.tex
Related work paragraph
Wordsmithing

@@
 settled, but this is sadly not the case.
-
-Let $\aReg$--$\bReg$ range over thread-local registers and $\aLoc$-$\cLoc$
-range over shared memory references.
 
@@
   \end{gather*}
+(In examples, $\aReg$--$\bReg$ range over thread-local registers and $\aLoc$-$\cLoc$
+range over shared memory references.)
 
@@
 characterization of \armeight{}.
+
+As far as we are aware, there are no previous attempts to provide a
+compositional semantics of sequential composition in a relaxed memory model.
+For a discussion of related work for relaxed memory models in general, see
+\cite{DBLP:journals/pacmpl/JagadeesanJR20}.

### 2021-01-25 paper/arm.tex
Grrr

@@
 To model such \emph{downgraded reads}, we use the uninterpreted symbols
-$\Dx{\aLoc}$.  Load actions that requiring downgrading introduce $\Dx{\aLoc}$.
-Relaxed stores on $\aLoc$ substitute true for $\Dx{\aLoc}$, whereas
-synchronizing stores substitute false for $\Dx{\aLoc}$.
+$\Dx{\aLoc}$.
 
@@
   \label{def:DS}
-  Let $[\FALSE/\D]$ be the substitution that performs $[\FALSE/\Dx{\aLoc}]$
-  for every $\aLoc$.
-  Let $\DS{\aLoc}{\amode}{}$ and $\DL{\aLoc}{\amode}{}$ be defined:
+  % Let $[\FALSE/\D]$ be the substitution that performs $[\FALSE/\Dx{\aLoc}]$
+  % for every $\aLoc$.
+  Let $[\aForm/\Dx{*}]$ substitute $\aForm$ for every $\Dx{\bLoc}$.\\
+  Let formula $\DL{\aLoc}{\amode}$ and substitution $\DS{\aLoc}{\amode}$ be defined:
   \begin{align*}
-    \DS{\aLoc}{\mRLX}{\bForm}&=\bForm[\TRUE/\Dx{\aLoc}] 
-    &\DL{\aLoc}{\mRLX}{\bForm}&=\bForm%\TRUE
+    \DL{\aLoc}{\mRLX}&=\TRUE
+    &\DS{\aLoc}{\mRLX}&=[\TRUE/\Dx{\aLoc}] 
     \\
-    \DS{\aLoc}{\mRA}{\bForm}&=\bForm[\FALSE/\D]
-    &\DL{\aLoc}{\mRA}{\bForm}&=\bForm\land\Dx{\aLoc}
+    \DL{\aLoc}{\mRA}&=\Dx{\aLoc}%\land\D{}
+    &\DS{\aLoc}{\mRA}&=[\FALSE/\D]
     \\
-    \DS{\aLoc}{\mSC}{\bForm}&=\bForm[\FALSE/\D]
-    &\DL{\aLoc}{\mSC}{\bForm}&=\bForm\land\Dx{\aLoc}
+    \DL{\aLoc}{\mSC}&=\Dx{\aLoc}%\land\D{}
+    &\DS{\aLoc}{\mSC}&=[\FALSE/\D]
   \end{align*}
 \end{definition}
-
 \begin{definition}[$\xCO$/$\xRASC$/$\xDGR$]
   \label{def:pomsets-down}
-  Update \refdef{def:pomsets-ra} to:
+  Update \refdef{def:pomsets-ra} to (\ref{L4} unchanged):
   \begin{enumerate}
-  % \item[\ref{S4})]
-  %   $\aTr{\bEvs}{\bForm}$ implies $(\Qw{\aLoc}\limplies\aExp{\neq}\aVal)\land \DS{\aLoc}{\amode}\bForm\noSUB{[\aExp/\aLoc]}$,
-  % \item[\ref{S5})]
-  %   $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qw{\aLoc}\land \DS{\aLoc}{\amode}\bForm\noSUB{[\aExp/\aLoc]}$.
-  % \item[\ref{L5})]
-  %   % $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode}{} \land((\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm)$,
-  %   % \item[\ref{L6})]
-  %   $\aTr{\dEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode}{} \land\bForm$.
   \item[\ref{S4})]
-    $\aTr{\bEvs}{\bForm}$ implies $\DS{\aLoc}{\amode}{\bForm[(\Qw{\aLoc}\land\aExp{=}\aVal)/\Qw{\aLoc}]}$,
+    $\aTr{\bEvs}{\bForm}$ implies $\bForm[(\Qw{\aLoc}\land\aExp{=}\aVal)/\Qw{\aLoc}]\DS{\aLoc}{\amode}$,
   \item[\ref{S5})]
-    $\aTr{\cEvs}{\bForm}$ implies $\DS{\aLoc}{\amode}{\bForm[\FALSE/\Qw{\aLoc}]}$.
+    $\aTr{\cEvs}{\bForm}$ implies $\bForm[\FALSE/\QS{\aLoc}{\amode}]\DS{\aLoc}{\amode}$,
+  \item[\ref{L4})]
+    $\aTr{\bEvs}{\bForm}$ implies $\aVal{=}\aReg\limplies\bForm$, 
   \item[\ref{L5})]
-    $\aTr{\dEvs}{\bForm}$ implies $\DL{\aLoc}{\amode}{\bForm[\FALSE/\Qx{\aLoc}]}$.
+    $\aTr{\cEvs}{\bForm}$ implies $\DL{\aLoc}{\amode}\land \bForm[\FALSE/\QL{\aLoc}{\amode}]$.
   \end{enumerate}
 \end{definition}
+Load actions that require downgrading introduce $\Dx{\aLoc}$.
+Relaxed stores on $\aLoc$ substitute true for $\Dx{\aLoc}$, whereas
+synchronizing stores substitute false for $\Dx{\aLoc}$.
 
@@
           % \xform{x1d}{2{=}r\limplies\bForm}{below=of b}
-          \xform{x1i}{\Dx{x} \land\bForm[\FALSE/\Qx{x}]}{below=of b} %x1d}
+          \xform{x1i}{\Dx{x} \land\bForm[\FALSE/\Qr{x}]}{below=of b} %x1d}
           % \xo{b}{x1d}
@@
 \refdef{def:pomsets-arm} combines all of the features of
-\textsection\ref{sec:q}--\ref{sec:arm}.  For completeness, we repeat
-\ref{L4}, which is unchanged from \refdef{def:pomsets-trans}.
+\textsection\ref{sec:q}--\ref{sec:arm}.  
 % Only \ref{L5} and \ref{T3} are changed from \refdef{def:pomsets-down}.
@@
   \label{def:pomsets-arm}
-  Update \refdef{def:pomsets-trans} to:
+  Update \refdef{def:pomsets-trans} to (\ref{L4} unchanged):
   \begin{enumerate}

### 2021-01-25 paper/arm.tex
Grrr

@@
   \item[\ref{S4})]
-    $\aTr{\bEvs}{\bForm}$ implies $\bForm[(\Qw{\aLoc}\land\aExp{=}\aVal)/\Qw{\aLoc}]\DS{\aLoc}{\amode}$,
+    $\aTr{\bEvs}{\bForm}$ implies
+    \begin{math}
+      \bForm
+      \DS{\aLoc}{\amode}
+      [(\Qw{\aLoc}\land\aExp{=}\aVal)/\Qw{\aLoc}],
+    \end{math}
   \item[\ref{S5})]
-    $\aTr{\cEvs}{\bForm}$ implies $\bForm[\FALSE/\QS{\aLoc}{\amode}]\DS{\aLoc}{\amode}$,
+    $\aTr{\cEvs}{\bForm}$ implies
+    \begin{math}
+      \bForm
+      \DS{\aLoc}{\amode}
+      [\FALSE/\QS{\aLoc}{\amode}],
+    \end{math}
   \item[\ref{L4})]
-    $\aTr{\bEvs}{\bForm}$ implies $\aVal{=}\aReg\limplies\bForm$, 
+    $\aTr{\bEvs}{\bForm}$ implies
+    \begin{math}
+      \aVal{=}\aReg\limplies\bForm,
+    \end{math}
   \item[\ref{L5})]
-    $\aTr{\cEvs}{\bForm}$ implies $\DL{\aLoc}{\amode}\land \bForm[\FALSE/\QL{\aLoc}{\amode}]$.
+    $\aTr{\cEvs}{\bForm}$ implies
+    \begin{math}
+      \DL{\aLoc}{\amode}\land \bForm[\FALSE/\QL{\aLoc}{\amode}].
+    \end{math}
   \end{enumerate}
@@
       \\
-      \hbox{\begin{tikzinline}[node distance=0.5em and .5em]
+      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
           \event{b}{\DR{y}{1}}{}
@@
   \end{align*}
-  Embedding the thread in thread group, we have, as desired:
+  Embedding the thread in thread group, \ref{T3} yields:
   \begin{align*}
     \begin{gathered}[t]
-      \PR{y}{r}
-      \SEMI
-      \IF{r}\THEN s\GETS x\FI
-      \\
-      \hbox{\begin{tikzinline}[node distance=0.5em and .5em]
+      % \PR{y}{r}
+      % \SEMI
+      % \IF{r}\THEN s\GETS x\FI
+      % \\
+      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
           \event{b}{\DR{y}{1}}{}
@@
   \end{align*}
+  The precondition of $\DWP{x}{0}$ is a tautology, as required.
 \end{example}
@@
   \item[\ref{S4})]
-    $\aTr{\bEvs}{\bForm}$ implies $(\Qw{\aLoc}\limplies\aExp{\neq}\aVal)\land \DS{\aLoc}{\amode}\bForm\noSUB{[\aExp/\aLoc]}$,
+    $\aTr{\bEvs}{\bForm}$ implies
+    \begin{math}
+      %(\Qw{\aLoc}\limplies\aExp{\neq}\aVal)\land
+      \bForm %\noSUB{[\aExp/\aLoc]},
+      [(\Qw{\aLoc}\land\aExp{=}\aVal)/\Qw{\aLoc}]
+      \DS{\aLoc}{\amode}
+    \end{math}
   \item[\ref{S5})]
-    $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qw{\aLoc}\land \DS{\aLoc}{\amode}\bForm\noSUB{[\aExp/\aLoc]}$.
+    $\aTr{\cEvs}{\bForm}$ implies
+    \begin{math}
+      %\lnot\Qw{\aLoc}\land 
+      \bForm %\noSUB{[\aExp/\aLoc]}.
+      [\FALSE/\QS{\aLoc}{\amode}]
+      \DS{\aLoc}{\amode}
+    \end{math}
   \item[\ref{L4})]
@@
     % \item[\ref{L6})]
-    $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode}{}
-    \land (\RW\limplies\bForm)$.
+    $\aTr{\cEvs}{\bForm}$ implies
+    \begin{math}
+      % \lnot\Qx{\aLoc} \land
+      \DL{\aLoc}{\amode}{} \land (\RW\limplies\bForm[\FALSE/\QL{\aLoc}{\amode}]).
+    \end{math}
   \end{enumerate}  
@@
 
+\begin{example}
+  $\xRRD$ does not adversely affect subscription (\refex{ex:subscription}).
+  \begin{align*}
+    \begin{gathered}[t]
+      \PR[\mRA]{y}{r}
+      \\
+      \hbox{\begin{tikzinline}[node distance=0.5em and .5em]
+          \raevent{b}{\QL{y}{\mRA}\mid\DR[\mRA]{y}{1}}{}
+          \xform{x1i}{\Dx{y}\land(\RW\limplies\bForm[\FALSE/\QL{\aLoc}{\mRA}])}{right=of b} %x1d}
+        \end{tikzinline}}  
+    \end{gathered}  
+    &&
+    \begin{gathered}[t]
+      \PR{x}{s}
+      \\
+      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
+          \event{c}{\QL{y}{\mRA}\mid\DW{x}{0}}{}
+        \end{tikzinline}}  
+    \end{gathered}  
+  \end{align*}
+  Since $\QL{y}{\mRA}[\FALSE/\QL{\aLoc}{\mRA}]$ is false, we have:
+  \begin{align*}
+    \begin{gathered}[t]
+      \PR[\mRA]{y}{r}
+      \SEMI
+      \PR{x}{s}
+      \\
+      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
+          \raevent{b}{\QL{y}{\mRA}\mid\DR[\mRA]{y}{1}}{}
+          \event{c}{\Dx{y}\land(\RW\limplies\FALSE)\mid\DW{x}{0}}{right=of b}
+        \end{tikzinline}}  
+    \end{gathered}  
+  \end{align*}
+  If this is the complete thread, we apply \ref{T3} to yield:
+  \begin{align*}
+    \begin{gathered}[t]
+      % \PR[\mRA]{y}{r}
+      % \SEMI
+      % \PR{x}{s}
+      % \\
+      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
+          \raevent{b}{\QL{y}{\mRA}\mid\DR[\mRA]{y}{1}}{}
+          \event{c}{\Dx{y}\land(\FALSE\limplies\FALSE)\mid\DW{x}{0}}{right=of b}
+        \end{tikzinline}}  
+    \end{gathered}  
+  \end{align*}
+  Although $(\FALSE\limplies\FALSE)$ is a tautology, $\Dx{y}$ is not.  This
+  pomset can only contribute to a top-level pomset if the thread is preceded
+  by a relaxed write to $y$, allowing $\DRP[\mRA]{y}{1}$ to be downgraded to
+  a relaxed read.
+\end{example}
+
 Every \armeight{} execution is allowed by \refdef{def:pomsets-arm}.  The
 proof of this fact is simplified by the recent characterization of
-\armeight{} in terms of \EGC{} \cite[\textsection
-B2.3.6]{arm-reference-manual}.  Under \EGC{}, an \armeight{} execution is a
-linearization of per-location program order and a subset of local-order.
-Every such linearization is also a valid pomset under
+\armeight{} in terms of \emph{external global completion} (\EGC{})
+\cite[\textsection B2.3.6]{arm-reference-manual}.  Under \EGC{}, an
+\armeight{} execution is a linearization of per-location program order and a
+subset of local-order.  Every such linearization is also a valid pomset under
 \refdef{def:pomsets-arm}.

### 2021-01-25 paper/outro.tex
rc1

\section{Conclusions}

We have presented the first model of relaxed memory that treats sequential
composition as a first-class citizen. The model builds directly on \jjr{}.

For sequential composition, parallel composition and the conditional, we
believe that the definition is \emph{natural}, even \emph{canonical}.
For stores and loads, instead, the definition in Figure~\ref{fig:full} is a
Frankenstein's monster of features.  This complexity is \emph{essential},
however, not an accident of our poor choices.  Relaxed memory models must
please many audiences: compiler writers want one thing, hardware designers
another, and programmers yet another still.  The result is inevitably full of
compromise.

Given that \emph{complexity} cannot be eliminated from relaxed memory models, the
best one can do is attempt to understand its causes.  We have broken the
problem into eight manageable pieces, discussed throughout
\textsection\ref{sec:co}--\ref{sec:complications}.

{Logic} is the thread that sews these features together.

%A unique feature of our model is the centrality of logic:

% study each feature in isolation, as a small delta to the base definition
% given in \textsection\ref{sec:model}.
% In we studied eight such
% features.

% seem staggeringly complex, we argue that this complexity is unavoidable if
% one wants all the features that it embodies.  By breaking the definition into
% its constituent parts, we have shown how each of eight

### 2021-01-25 paper/intro.tex
Revised Sec 2.4
Fixed dangling ref

@@
 \section{Introduction}
+\label{sec:intro}
 
@@
   to modern architectures: downgrading some synchronized accesses to relaxed,
-  and removing read-read dependencies,
-\item \S\ref{sec:complications} show how to address common litmus tests, and
-\item \S\ref{sec:discussion} is a discussion of the design space.
+  and removing read-read dependencies, and
+\item \S\ref{sec:complications} show how to address common litmus tests.
 \end{itemize}

### 2021-01-26 paper/arm.tex
Extra polish

@@
       \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-          \event{c}{r\mid\DW{x}{0}}{}
+          \event{c}{r\mid\DR{x}{0}}{}
           % \xform{x1d}{\bForm[\TRUE/\Dx{\aLoc}]}{below=of c}
@@
           \event{b}{\DR{y}{1}}{}
-          \event{c}{\RW\limplies r\mid\DW{x}{0}}{right=of b}
+          \event{c}{\RW\limplies r\mid\DR{x}{0}}{right=of b}
           %\xform{x1i}{\RW\limplies\bForm}{right=of b} %x1d}
@@
           \event{b}{\DR{y}{1}}{}
-          \event{c}{\FALSE\limplies r\mid\DW{x}{0}}{right=of b}
+          \event{c}{\FALSE\limplies r\mid\DR{x}{0}}{right=of b}
           %\xform{x1i}{\RW\limplies\bForm}{right=of b} %x1d}
@@
   \end{align*}
-  The precondition of $\DWP{x}{0}$ is a tautology, as required.
+  The precondition of $\DRP{x}{0}$ is a tautology, as required.
 \end{example}
@@
   \label{def:pomsets-arm}
-  Update \refdef{def:pomsets-trans} to (\ref{L4} unchanged):
+  Update \refdef{def:pomsets-trans} to: % (\ref{L4} unchanged):
   \begin{enumerate}

### 2021-01-26 paper/intro.tex
Extra polish

@@
 
-This model naturally extends to the case of shared-memory concurrency
-in a natural way, leading to a \emph{sequentially consistent}
-semantics, in which \emph{program order} inside a thread implies
+This model naturally extends to the case of shared-memory concurrency, leading to a \emph{sequentially consistent}
+semantics \cite{Lamport:1979:MMC:1311099.1311750}, in which \emph{program order} inside a thread implies
 a total \emph{causal order} between read and write events, for example:
@@
 Unfortunately, this model does not compile efficiently to commodity
-hardware, resulting in a 37--73\% increase in CPU time~\cite{Liu:2019:ASC:3314221.3314611} on ARM, and
-hence power consumption.  Developers of software and compilers have
+hardware, resulting in a 37--73\% increase in CPU time on ARM~\cite{Liu:2019:ASC:3314221.3314611} and,
+hence, in power consumption.  Developers of software and compilers have
 therefore been faced with a difficult trade-off, between an elegant
@@
 transformers, since there are two subsets of any one-element set.
-We call the predicate transformer for $\emptyset$ the \emph{independent}
-transformer, and the one indexed by $\{e\}$ the \emph{dependent} transformer.
+The \emph{independent}
+transformer is indexed by the empty set, whereas the \emph{dependent}
+transformer is indexed by the singleton.
 We visualize this by including more than one transformed predicate,
@@
         \event{wz0}{(s{=}0) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
-        \xform{rx0ry0d}{(r{=}0) \limplies(s{=}0) \limplies \bForm}{below=of ry0}
-        \xform{rx0d}{(r{=}0) \limplies \bForm}{above=of rx0}
+        \xform{rx0ry0d}{(r{=}0) \limplies(s{=}0) \limplies \bForm}{above=of ry0}
+        \xform{rx0d}{(r{=}0) \limplies \bForm}{below=of rx0}
         \xform{ry0d}{(s{=}0) \limplies \bForm}{right=of rx0d}
@@
         \event{ry0}{\DR{y}{0}}{right=of rx0}
-        \xform{rx0ry0d}{(r{=}0) \limplies(s{=}0) \limplies \bForm}{below=of ry0}
-        \xform{rx0d}{(r{=}0) \limplies \bForm}{above=of rx0}
+        \xform{rx0ry0d}{(r{=}0) \limplies(s{=}0) \limplies \bForm}{above=of ry0}
+        \xform{rx0d}{(r{=}0) \limplies \bForm}{below=of rx0}
         \xform{ry0d}{(s{=}0) \limplies \bForm}{right=of rx0d}
-        \xform{wz0i}{\bForm}{right=of ry0d}
+        \xform{wz0i}{\bForm}{right=of rx0ry0d}
         \xo{rx0}{rx0d}
-        \xo{ry0}{ry0d}
+        \xo{ry0}{ry0d}the \emph{independent}
+transformer
         \xo{rx0}{rx0ry0d}

### 2021-01-29 paper/appendix.tex
Some tikz of predicate transformers.
Updates to address and read-read

--- 2021-01-08 paper/appendix.tex
@@
-\appendix
-% \appendices
-% \section{}
+\documentclass[conference,compsoc]{IEEEtran}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Missing includes
+\usepackage{graphicx}
+\usepackage[prologue]{xcolor}
+\definecolor[named]{ACMBlue}{cmyk}{1,0.1,0,0.1}
+\definecolor[named]{ACMYellow}{cmyk}{0,0.16,1,0}
+\definecolor[named]{ACMOrange}{cmyk}{0,0.42,1,0.01}
+\definecolor[named]{ACMRed}{cmyk}{0,0.90,0.86,0}
+\definecolor[named]{ACMLightBlue}{cmyk}{0.49,0.01,0,0}
+\definecolor[named]{ACMGreen}{cmyk}{0.20,0,1,0.19}
+\definecolor[named]{ACMPurple}{cmyk}{0.55,1,0,0.15}
+\definecolor[named]{ACMDarkBlue}{cmyk}{1,0.58,0,0.21}
+\usepackage[bookmarksnumbered,unicode]{hyperref}
+\hypersetup{colorlinks,
+  linkcolor=ACMPurple,
+  citecolor=ACMPurple,
+  urlcolor=ACMDarkBlue,
+  filecolor=ACMDarkBlue}
+\usepackage{amsmath}
+\usepackage{amsthm}
+\theoremstyle{plain}
+\newtheorem{theorem}{Theorem}
+%\newtheorem{conjecture}{Conjecture}
+\newtheorem{proposition}{Prop}
+\newtheorem{lemma}{Lemma}
+\theoremstyle{definition}
+\newtheorem{definition}{Def}
+\newtheorem{example}[definition]{Ex}
+\theoremstyle{remark}
+\newtheorem{remark}{Remark}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-This paper addresses several errors in \citep{10.1145/3428262}, which we
-henceforth refer to as \jjr{}.
+% \author{\IEEEauthorblockN{%
+%     Alan Jeffrey\IEEEauthorrefmark{1} and
+%     James Riely\IEEEauthorrefmark{2}
+% }
+% \IEEEauthorblockA{\IEEEauthorrefmark{1}Roblox}
+% \IEEEauthorblockA{\IEEEauthorrefmark{2}DePaul University}}
+\author{Anonymous\\Anonymous Institution}
 
-In \jjr{\textsection2.4}, parallel composition is defined allowing coalescing
-of events.  Here we have forbidden coalescing.  This difference appears to be
-arbitrary.  In \jjr{}, however, there is a mistake in the handling of
-termination actions.  The predicates should be joined using $\land$, not
-$\lor$.
+\title{Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers}
 
-In \jjr{\textsection2.6} the semantics of read is defined as follows:
-\begin{align*}
-  \sem{\aReg\GETS\aLoc^\amode\SEMI \aCmd} & \eqdef \textstyle\bigcup_\aVal\;
-  (\DRmode\aLoc\aVal) \prefix \sem{\aCmd} [\aLoc/\aReg]
-\end{align*}
-The definition of prefixing$((\aForm \mid \aAct) \prefix \aPSS)$ has several clauses.
-The most relevant are as follows, where $\bEv$ is the new event labeled with
-$(\aForm \mid \aAct)$ and $\aEv$ is an event from $\aPSS$:
-\begin{description}
-\item[{\labeltextsc[P4c]{(P4c)}{4c}}]
-  If $\bEv$ reads $\aVal$ from $\aLoc$ then either $\aEv=\bEv$ or
-  $\labelingForm'(\aEv)$ implies $\labelingForm(\aEv)[\aVal/\aLoc]$.
-\item[{\labeltextsc[P5a]{(P5a)}{5a}}]\labeltextsc[P5]{}{5}%
-  If $\bEv$ reads and $\aEv$ writes then either $\labelingForm'(\aEv)$
-  implies $\labelingForm(\aEv)$ or $\bEv\le'\aEv$.
-% \item[{\labeltextsc[P5b]{(P5b)}{5b}}]
-%   If $\bEv$ and $\aEv$ are in conflict then $\bEv\le'\aEv$.
-\end{description}
+\usepackage{macros}
+\usepackage[sectionbib,numbers,square]{natbib}
 
-We have discovered two issues with this definition.
+\usepackage{xr}
+\newcounter{pomsetPreParCount}\setcounter{pomsetPreParCount}{8}
+\newcounter{pomsetXSemiCount}\setcounter{pomsetXSemiCount}{8}
+\externaldocument{paper}
+\begin{document} 
+\pagestyle{plain}
+\appendices
+\input{discussion}
+\bibliographystyle{abbrvnat}
+\bibliography{bib}
 
-The first issue concerns the substitution $[\aLoc/\aReg]$.  It should be
-$[\aReg/\aLoc]$.  We noticed this error while developing the alternative
-characterization presented here.  The error causes redundant read elimination
-to fail in \jjr{}.  As a result, common subexpression elimination also fails.
-The problem can be seen in \ref{TC2}.
-\begin{gather*}
-  \taglabel{TC2}
-  r\GETS x\SEMI
-  s\GETS x\SEMI
-  \IF{r{=}s}\THEN y\GETS 1\FI
-  \PAR
-  x\GETS y
-\end{gather*}
-% In \jjr{\textsection3.1},
-We claimed that \ref{TC2} allowed the following
-execution:
-\begin{gather*}
-  \hbox{\begin{tikzinline}[node distance=1.5em]
-  \event{a1}{\DR{x}{1}}{}
-  \event{a2}{\DR{x}{1}}{right=of a1}
-  \event{a3}{\DW{y}{1}}{right=of a2}
-  % \po{a2}{a3}
-  % \po[out=15,in=165]{a1}{a3}
-  \event{b1}{\DR{y}{1}}{right=3em of a3}
-  \event{b2}{\DW{x}{1}}{right=of b1}
-  \rf{a3}{b1}
-  \po{b1}{b2}
-  \rf[out=169,in=11]{b2}{a2}
-  \rf[out=169,in=11]{b2}{a1}
-    \end{tikzinline}}
-\end{gather*}
-But this execution is not possible using the semantics of \jjr{}:
-$\DWP{y}{1}$ has precondition $r{=}s$ in
-\begin{math}
-  \sem{\IF{r{=}s}\THEN y\GETS 1\FI}.
-\end{math}
-Given the lack of order in the execution, the precondition of $\DWP{y}{1}$
-must entail $r{=}1\land r{=}x$ in 
-\begin{math}
-  \sem{s\GETS x\SEMI
-  \IF{r{=}s}\THEN y\GETS 1\FI}.
-\end{math}
-\ref{4c} imposes $r{=}1$, and \ref{5a} imposes $r{=}x$.  Adding the second
-read, the precondition of $\DWP{y}{1}$ must entail both $1{=}1\land 1{=}x$
-and also $x{=}1\land x{=}x$.  This can be simplified to $x{=}1$.  This leaves
-a requirement that must be satisfied by a preceding write.  Since the
-preceding write is the initialization to $0$, the requirement cannot be
-satisfied, and the execution is impossible.\footnote{In \jjr{} we ignore the
-  middle terms, mistakenly simplifying this to $1{=}1\land x{=}x$.
-  Correcting the error, the attempted execution is:
-\begin{gather*}
-  \hbox{\begin{tikzinline}[node distance=1.5em]
-  \event{a1}{\DR{x}{1}}{}
-  \event{a2}{\DR{x}{1}}{right=of a1}
-  \event{a3}{\DW{y}{1}}{right=of a2}
-  \po{a2}{a3}
-  \po[out=-20,in=-160]{a1}{a3}
-  \event{b1}{\DR{y}{1}}{right=3em of a3}
-  \event{b2}{\DW{x}{1}}{right=of b1}
-  \rf{a3}{b1}
-  \po{b1}{b2}
-  \rf[out=169,in=11]{b2}{a2}
-  \rf[out=169,in=11]{b2}{a1}
-    \end{tikzinline}}
-\end{gather*}}
+\end{document}
 
-The substitution $[\aLoc/\aReg]$ leaves the obligation on $\aLoc$ to be
-fulfilled by the preceding write.  Thus, the read does not update the
-\emph{value} of $\aLoc$ in subsequent predicates.  The substitution
-$[\aReg/\aLoc]$, instead, does update the value of $\aLoc$, thus removing any
-obligation on $\aLoc$ for preceding code.
-
-In order to write this, we must update the definition of prefixing reads to
-include the register.  Then \ref{4c} becomes:
-\begin{description}
-\item[\textsc{(p4c)}] If $\bEv$ reads $\aVal$ from $\aLoc$ then either
-  $\aEv=\bEv$ or $\labelingForm'(\aEv)$ implies
-  $\labelingForm(\aEv)[\aVal/\aReg]$.
-\end{description}
-
-We can then reason with \ref{TC2} as follows: $\DWP{y}{1}$ has precondition
-$r{=}s$ in
-\begin{math}
-  \sem{\IF{r{=}s}\THEN y\GETS 1\FI}.
-\end{math}
-To avoid introducing order in the execution, the precondition of $\DWP{y}{1}$
-must entail $r{=}1\land r{=}s$ in 
-\begin{math}
-  \sem{s\GETS x\SEMI
-  \IF{r{=}s}\THEN y\GETS 1\FI}.
-\end{math}
-\ref{4c} imposes $r{=}1$, and \ref{5a} imposes $r{=}x$.  Adding the second
-read, the precondition of $\DWP{y}{1}$ must entail both $1{=}1\land 1{=}x$
-and also $x{=}1\land x{=}x$.  This can be simplified to $x{=}1$.  This leaves
-a requirement that must be satisfied by a preceding write.
-
-With read elimination, the rule for relaxed reads is as follows:
-\begin{align*}
-  \sem{\PR{\aLoc}{\aReg} \SEMI \aCmd} &\eqdef
-  \sem{\aCmd}[\aLoc/\aReg]
-  \cup
-  \textstyle\bigcup_\aVal\;
-  \DRP{\aLoc}{\aVal} \prefix_{\aReg} %\Rdis{\aLoc}{\aVal}
-  \sem{\aCmd}[\aReg/\aLoc]
-\end{align*}
-It is interesting to note that the substitution is $[\aLoc/\aReg]$ on
-eliminated reads, and $[\aReg/\aLoc]$ on non-eliminated reads.  Intuitively,
-the subsequent value of $\aLoc$ is fixed by an explicit read, but not for an
-eliminated read.  In the latter case, the value is fixed by some preceding
-action.  The preceding action may itself be a read. This gives rise to some
-fear that we might introduce thin-air reads, since we do not enforce
-read-read coherence.  But this is not the case.  Consider the following example:
-\begin{gather*}
-  r\GETS x\SEMI
-  s\GETS x\SEMI
-  y\GETS s
-  \PAR
-  x\GETS y
-  \\
-  \hbox{\begin{tikzinline}[node distance=1.5em]
-  \event{a1}{\DR{x}{1}}{}
-  \event{a2}{\DR{x}{1}}{right=of a1}
-  \event{a3}{\DW{y}{1}}{right=of a2}
-  %\po{a2}{a3}
-  \po[out=-20,in=-160]{a1}{a3}
-  \event{b1}{\DR{y}{1}}{right=3em of a3}
-  \event{b2}{\DW{x}{1}}{right=of b1}
-  \rf{a3}{b1}
-  \po{b1}{b2}
-  \rf[out=169,in=11]{b2}{a2}
-  \rf[out=169,in=11]{b2}{a1}
-    \end{tikzinline}}
-  \\
-  \hbox{\begin{tikzinline}[node distance=1.5em]
-  \event{a1}{\DR{x}{1}}{}
-  \internal{a2}{\DR{x}{1}}{right=of a1}
-  \event{a3}{\DW{y}{1}}{right=of a2}
-  %\po{a2}{a3}
-  \po[out=-20,in=-160]{a1}{a3}
-  \event{b1}{\DR{y}{1}}{right=3em of a3}
-  \event{b2}{\DW{x}{1}}{right=of b1}
-  \rf{a3}{b1}
-  \po{b1}{b2}
-  %\rf[out=169,in=11]{b2}{a2}
-  \rf[out=169,in=11]{b2}{a1}
-    \end{tikzinline}}
-\end{gather*}
-But this is not a problem, since fulfillment requires that $\DWP{x}{1}$
-precede both reads of $x$.
-
-The second issue concerns acquiring reads.  Shortly after publication,
-\citet{anton} noticed a shortcoming of the implementation on \armeight{} in
-\jjr{\textsection 7}.  The proof given there assumes that all internal reads
-can be dropped.  However, this is not the case for acquiring reds.  For
-example, \jjr{} disallows the following execution, which is allowed by
-\armeight{} and \tso{}.
-\begin{gather*}
-  \PW{x}{2}\SEMI 
-  \PR[\mRA]{x}{r}\SEMI
-  \PR{y}{s} \PAR
-  \PW{y}{2}\SEMI
-  \PW[\mRA]{x}{1}
-  \\
-  \hbox{\begin{tikzinline}[node distance=1.5em]
-      \event{a}{\DW{x}{2}}{}
-      \event{b}{\DR[\mRA]{x}{2}}{right=of a}
-      \event{c}{\DR{y}{0}}{right=of b}
-      \event{d}{\DW{y}{2}}{right=2.5em of c}
-      \event{e}{\DW[\mRA]{x}{1}}{right=of d}
-      \rf{a}{b}
-      \sync{b}{c}
-      \wk{c}{d}
-      \sync{d}{e}
-      \wk[out=-165,in=-15]{e}{a}
-      % \rfi{a}{b}
-      % \bob{b}{c}
-      % \fre{c}{d}
-      % \bob{d}{e}
-      % \coe[out=-165,in=-15]{e}{a}
-    \end{tikzinline}}
-\end{gather*}
-The solution we have adopted is to allow an acquiring read to be downgraded
-to a relaxed read when it is preceded (sequentially) by a relaxed write that
-could fulfill it.  Back-porting this solution to \jjr{} requires that we add
-access predicates to the logic and allow
-
-In addition, the \armeight{} implementation result given here does not rely
-on read elimination.  Instead we use a recent alternative characterization of
-\armeight{} \citep{alglave-git-alternate,arm-reference-manual,armed-cats}.
+% Local Variables:
+% mode: latex
+% TeX-master: t
+% End:

### 2021-01-31 paper/arm.tex
added submission
Updates to address and read-read
Work in appendix

@@
 
-\subsection{Downgraded Reads ($\xDGR$)}
+\subsection{Downgraded Reads (\xDGR)}
 \label{sec:internal}
@@
 \end{definition}
-\begin{definition}[$\xCO$/$\xRASC$/$\xDGR$]
+\begin{definition}[\xCO/\xRASC/\xDGR]
   \label{def:pomsets-down}
@@
 
-\subsection{Removing Read-Read dependencies ($\xRRD$)}
+\subsection{Removing Read-Read dependencies (\xRRD)}
+\label{sec:rrd}
 
@@
 
-\begin{definition}[$\xRRD$]
+\begin{definition}[\xRRD]
   \label{def:pomsets-rr}
@@
   \begin{enumerate}
+  \item[\ref{L4})]
+    $\aTr{\bEvs}{\bForm}$ implies $\aVal{=}\aReg\limplies\bForm$, 
   \item[\ref{L5})]
     % $\aTr{\cEvs}{\bForm}$ implies $\RW\limplies(\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm$,
-    $\aTr{\cEvs}{\bForm}$ implies $\RW\limplies\bForm$,
+    $\aTr{\cEvs}{\bForm}$ implies $(\aVal{=}\aReg\lor\RW)\limplies\bForm$,
   \end{enumerate}  
 \end{definition}
+\ref{L5} requires $\aTr{\cEvs}{\bForm}$ implies $\bForm$ for writes and
+$\aTr{\bEvs}{\bForm}$ implies $\aVal{=}\aReg\limplies\bForm$ for reads.  For
+reads, the dependent and independent predicate transformers are identical.
 
@@
           \event{b}{\DR{y}{1}}{}
-          \xform{x1i}{\RW\limplies\bForm}{right=of b} %x1d}
+          \xform{x1i}{(1{=}r\lor\RW)\limplies\bForm}{right=of b} %x1d}
         \end{tikzinline}}  
@@
           \event{b}{\DR{y}{1}}{}
-          \event{c}{\RW\limplies r\mid\DR{x}{0}}{right=of b}
+          \event{c}{(1{=}r\lor\RW)\limplies r\mid\DR{x}{0}}{right=of b}
           %\xform{x1i}{\RW\limplies\bForm}{right=of b} %x1d}
@@
           \event{b}{\DR{y}{1}}{}
-          \event{c}{\FALSE\limplies r\mid\DR{x}{0}}{right=of b}
+          \event{c}{(1{=}r\lor\FALSE)\limplies r\mid\DR{x}{0}}{right=of b}
           %\xform{x1i}{\RW\limplies\bForm}{right=of b} %x1d}
@@
 
-\begin{definition}[$\xCO$/$\xRASC$/$\xDGR$/$\xRRD$]
+\begin{definition}[\xCO/\xRASC/\xDGR/\xRRD]
   \label{def:pomsets-arm}
@@
 \begin{example}
-  $\xRRD$ does not adversely affect subscription (\refex{ex:subscription}).
+  \xRRD{} does not adversely affect subscription (\refex{ex:subscription}).
   \begin{align*}

### 2021-02-05 paper/arm.tex
hacking at address dependencies
Working on appendices

@@
   \label{def:pomsets-down}
-  Update \refdef{def:pomsets-ra} to (\ref{L4} unchanged):
+  Update \refdef{def:pomsets-ra} to: % (\ref{L4} unchanged):
   \begin{enumerate}
@@
 \end{definition}
-\ref{L5} requires $\aTr{\cEvs}{\bForm}$ implies $\bForm$ for writes and
-$\aTr{\bEvs}{\bForm}$ implies $\aVal{=}\aReg\limplies\bForm$ for reads.  For
-reads, the dependent and independent predicate transformers are identical.
+For reads, the dependent and independent predicate transformers are identical
+(since $\RW$ is $\FALSE$).  For writes, \ref{L5} requires
+$\aTr{\cEvs}{\bForm}$ implies $\bForm$ for the independent case (since $\RW$
+is $\TRUE$).
 

### 2021-02-11 paper/outro.tex
submission
Draft of Fork-Join

@@
 believe that the definition is \emph{natural}, even \emph{canonical}.
-For stores and loads, instead, the definition in Figure~\ref{fig:full} is a
+For stores and loads, instead, the definition in \reffig{fig:full} is a
 Frankenstein's monster of features.  This complexity is \emph{essential},
-however, not an accident of our poor choices.  Relaxed memory models must
+however, not just an accident of our poor choices.  Relaxed memory models must
 please many audiences: compiler writers want one thing, hardware designers
@@
 
-Given that \emph{complexity} cannot be eliminated from relaxed memory models, the
-best one can do is attempt to understand its causes.  We have broken the
+Given that \emph{complexity} cannot be eliminated from relaxed memory models,
+the best one can do is attempt to understand its causes.  We have broken the
 problem into eight manageable pieces, discussed throughout
-\textsection\ref{sec:co}--\ref{sec:complications}.
+\textsection\ref{sec:q}--\ref{sec:complications}.  \refdef{def:pomsets-arm}
+summarizes all the features necessary for efficient implementation on
+\armeight{}.
 

### 2021-02-11 paper/arm.tex
Draft of Fork-Join
Sorted out bib

@@
   \label{ex:dgr1}
-  The following execution is allowed by \armeight{}, but disallowed by
-  \refdef{def:pomsets-ra}.  The coherence order between the writes can be
-  witnessed by a separate thread, which we have elided.
+  The following example is from \citeauthor{anton} \cite{anton}.  It is
+  allowed by \armeight{}, but disallowed by \refdef{def:pomsets-ra}.  The
+  coherence order between the writes can be witnessed by a separate thread,
+  which we have elided.
   \begin{gather*}
@@
   \label{def:pomsets-arm}
+  Let $[\aForm/\Q{}]$ be the substitution $[\aForm/\Qr{*}][\aForm/\Qw{*}][\aForm/\Qsc]$.  
   Update \refdef{def:pomsets-trans} to: % (\ref{L4} unchanged):

### 2021-02-11 paper/intro.tex
Sorted out bib

@@
 \cite{DBLP:journals/pacmpl/JagadeesanJR20} apply to our model, including
-compositional reasoning for temporal safety properties and {local} \drfsc.
+compositional reasoning for temporal safety properties and {local} \drfsc{}
+as in \cite{Dolan:2018:BDR:3192366.3192421,DBLP:conf/ppopp/DongolJR19}.
 In \textsection\ref{sec:arm}, we provide an alternative proof strategy for

### 2021-02-25 paper/arm.tex
Updated defs

@@
 \subsection{Downgraded Reads (\xDGR)}
-\label{sec:internal}
+\label{sec:downgrade}
 
@@
           % \xform{x1i}{\lnot\Qw{x}\land\bForm[\TRUE/\Dx{x}]}{below=of x1d}
-          \xo{a}{x1d}
+          \xos{a}{x1d}
         \end{tikzinline}}  
@@
   Update \refdef{def:pomsets-trans} to:
-  \begin{enumerate}
-  \item[\ref{T3})]
-    $\labelingForm(\aEv)$ implies
-    $\labelingForm_1(\aEv)[\TRUE/\RW]$ if $\labelingAct_1(\aEv)$ is a write,
-    \\
-    $\labelingForm(\aEv)$ implies
-    $\labelingForm_1(\aEv)[\FALSE/\RW]$ otherwise.
-  \end{enumerate}
+  % \begin{enumerate}
+  % \item[\ref{T3})]
+  %   $\labelingForm(\aEv)$ implies
+  %   $\labelingForm_1(\aEv)[\TRUE/\RW]$ if $\labelingAct_1(\aEv)$ is a write,
+  %   \\
+  %   $\labelingForm(\aEv)$ implies
+  %   $\labelingForm_1(\aEv)[\FALSE/\RW]$ otherwise.
+  % \end{enumerate}
   \begin{enumerate}
@@
     % $\aTr{\cEvs}{\bForm}$ implies $\RW\limplies(\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm$,
-    $\aTr{\cEvs}{\bForm}$ implies $(\aVal{=}\aReg\lor\RW)\limplies\bForm$,
+    $\aTr{\cEvs}{\bForm}$ implies $(\aVal{=}\aReg\lor\RW)\limplies\bForm$, when $\aEvs\neq\emptyset$,
+  \item[{\labeltext[L6]{L6)}{L6}}] 
+    $\aTr{\dEvs}{\bForm}$ implies $\bForm$, when $\aEvs=\emptyset$.
   \end{enumerate}  
 \end{definition}
-For reads, the dependent and independent predicate transformers are identical
-(since $\RW$ is $\FALSE$).  For writes, \ref{L5} requires
-$\aTr{\cEvs}{\bForm}$ implies $\bForm$ for the independent case (since $\RW$
-is $\TRUE$).
+\ref{L4} is the \emph{dependent} case,
+\ref{L5} the \emph{independent} case, and
+\ref{L6} the \emph{empty} case. 
+ Recall the substitutions performed by $\ref{sTOP}$
+(\refdef{def:pomsets-top}).
+For writes, \ref{L5} is the same as \ref{L6} (since $\RW$
+is $\TRUE$).  For reads, \ref{L5} is the
+same as \ref{L4} (since $\RW$ is $\FALSE$).
+% For reads, the dependent and independent predicate transformers are identical
+% .  For writes, \ref{L5} requires
+% $\aTr{\cEvs}{\bForm}$ implies $\bForm$ for the independent case.
+
+One reading of \ref{L5} is that when satisfying a precondition $\aForm$ it is
+safe to use the value of the read, as long as the action is not a write.  If
+the pomset is empty, however, there is no value $\aVal$ to use.  Therefore
+the best we can do is to emulate skip, as in \ref{L6}.  % In order to
+% eventually arrive at a top-level pomset, this means that subsequent code must
+% be independent of $\aReg$.
 
@@
   \end{align*}
-  Embedding the thread in thread group, \ref{T3} yields:
+  At top-level, $\ref{sTOP}$ yields:
   \begin{align*}
@@
     $\labelingForm(\aEv)$ implies $\QL{\aLoc}{\amode}$.
-  \item[\ref{T3})]
-    $\labelingForm(\aEv)$ implies
-    $\labelingForm_1(\aEv)[\TRUE/\Q{}][\TRUE/\RW]$ if $\labelingAct_1(\aEv)$ is a write,
-    \\
-    $\labelingForm(\aEv)$ implies
-    $\labelingForm_1(\aEv)[\TRUE/\Q{}][\FALSE/\RW]$ otherwise.
+  % \item[\ref{T3})]
+  %   $\labelingForm(\aEv)$ implies
+  %   $\labelingForm_1(\aEv)[\TRUE/\Q{}][\TRUE/\RW]$ if $\labelingAct_1(\aEv)$ is a write,
+  %   \\
+  %   $\labelingForm(\aEv)$ implies
+  %   $\labelingForm_1(\aEv)[\TRUE/\Q{}][\FALSE/\RW]$ otherwise.
   \end{enumerate}
@@
       % \lnot\Qx{\aLoc} \land
-      \DL{\aLoc}{\amode}{} \land (\RW\limplies\bForm[\FALSE/\QL{\aLoc}{\amode}]).
+      \DL{\aLoc}{\amode}{} \land ((\aVal{=}\aReg\lor\RW)\limplies\bForm[\FALSE/\QL{\aLoc}{\amode}]).
     \end{math}
@@
   \end{align*}
-  If this is the complete thread, we apply \ref{T3} to yield:
+  Applying $\ref{sTOP}$ yields:
   \begin{align*}
@@
       \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-          \raevent{b}{\QL{y}{\mRA}\mid\DR[\mRA]{y}{1}}{}
+          \raevent{b}{\TRUE\mid\DR[\mRA]{y}{1}}{}
           \event{c}{\Dx{y}\land(\FALSE\limplies\FALSE)\mid\DW{x}{0}}{right=of b}
@@
   \end{align*}
-  Although $(\FALSE\limplies\FALSE)$ is a tautology, $\Dx{y}$ is not.  This
-  pomset can only contribute to a top-level pomset if the thread is preceded
-  by a relaxed write to $y$, allowing $\DRP[\mRA]{y}{1}$ to be downgraded to
-  a relaxed read.
+  Although $(\FALSE\limplies\FALSE)$ is a tautology, $\Dx{y}$ is not.  Thus,
+  the pomset can only contribute to a top-level pomset if the thread is
+  preceded by a relaxed write to $y$, allowing $\DRP[\mRA]{y}{1}$ to be
+  downgraded to a relaxed read.
 \end{example}

### 2021-02-25 paper/outro.tex
Minor edits
Updated defs

@@
 believe that the definition is \emph{natural}, even \emph{canonical}.
-For stores and loads, instead, the definition in \reffig{fig:full} is a
+For stores and loads, instead, the definition in \reffig{fig:no-addr} is a
 Frankenstein's monster of features.  This complexity is \emph{essential},
@@
 the best one can do is attempt to understand its causes.  We have broken the
-problem into eight manageable pieces, discussed throughout
+problem into seven manageable pieces, discussed throughout
 \textsection\ref{sec:q}--\ref{sec:complications}.  \refdef{def:pomsets-arm}
 summarizes all the features necessary for efficient implementation on
-\armeight{}.
+\armeight{}.  We discuss address calculation, read-modify-write operations
+and fences in the appendix.
 

### 2021-03-16 paper/arm.tex
Changed S4 to include M=v rather than [(Q /\ M=v)/Q]
Dealing with bad example 5.1.

@@
 which is very close to our model.
+
+\begin{example}
+  Bad example:
+  \begin{gather*}
+    \PR{x}{r}\SEMI 
+    \PW{x}{2}\SEMI
+    \PR{x}{s}\SEMI
+    \PW{y}{s{-}1} \PAR
+    \PR{y}{r}\SEMI
+    \PW{x}{r}
+    \\
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{a}{\DR{x}{1}}{}
+        \event{b}{\DW{x}{2}}{right=of a}
+        \event{c}{\DR{x}{2}}{right=of b}
+        \event{d}{\DW{y}{1}}{right=of c}
+        \event{e}{\DR{y}{1}}{right=of d}
+        \event{f}{\DW{x}{1}}{right=of e}
+        \wk{a}{b}
+        \rf{b}{c}
+        \po{c}{d}
+        \rf{d}{e}
+        \po{e}{f}
+        \rf[out=-165,in=-15]{f}{a}
+      \end{tikzinline}}
+  \end{gather*}
+\end{example}
+
+\begin{example}
+  Should not wipe out local state only when there is a release-acquire pair,
+  as in the definition of \jjr{}, which allows the following execution.  Even
+  though there is a blocker $\DWP{x}{3}$, here we are using $\DWP{x}{0}$ to
+  justify the lack of dependence $\DRP{x}{4}\xpo\DWP{w}{4}$.
+  \begin{gather*}
+    \PW{x}{0}\SEMI
+    \IF{x}\THEN\PW{y}{2}\FI\SEMI
+    \IF{\PR[\mRA]{z}{}}\THEN
+      \PR{x}{s}\SEMI
+      \IF{s\;\mathsf{even}}\THEN\PW{w}{4}\FI
+    \FI
+    \PAR
+    \PW{x}{2}\SEMI
+    \IF{\PR[\mRA]{y}{}}\THEN
+      \PW{x}{3}\SEMI
+      \PW[\mRA]{z}{1}
+    \FI
+    \SEMI
+    \PW{x}{\PR{w}{}}
+    \\
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{2a}{\DW{x}{0}}{}
+        \event{2b}{\DR{x}{2}}{right=of 2a}
+        \event{2c}{\DW{y}{1}}{right=of 2b}
+        \raevent{2d}{\DR[\mRA]{z}{1}}{right=of 2c}
+        \event{2e}{\DR{x}{4}}{right=of 2d}
+        \event{2f}{\DW{w}{4}}{right=of 2e}
+        %
+        \event{3a}{\DW{x}{2}}{below=of 2a}
+        \raevent{3b}{\DR[\mRA]{y}{1}}{right=of 3a}
+        \event{3c}{\DW{x}{3}}{right=of 3b}
+        \raevent{3d}{\DW[\mRA]{z}{1}}{right=of 3c}
+        \event{3e}{\DR{w}{4}}{right=of 3d}
+        \event{3f}{\DW{x}{4}}{right=of 3e}
+        \wk{2a}{2b}
+        \po{2b}{2c}
+        \sync{2d}{2e}
+        \sync[out=20,in=160]{2d}{2f}
+        \sync{3b}{3c}
+        \sync{3c}{3d}
+        \po{3e}{3f}
+        \rf{3a}{2b}
+        \rf{2c}{3b}
+        \rf{3d}{2d}
+        \rf{3f}{2e}
+        \rf{2f}{3e}
+        \wk{2a}{3a}
+        %\wk[out=-20,in=-160]{3c}{3f}
+        %\wk[out=20,in=160]{2b}{2e}
+      \end{tikzinline}}
+  \end{gather*}
+  (The read-clobbering semantics from \textsection\ref{sec:substitutions}
+  does not change this.)
+\end{example}
+
+It does seem to me like a non-local blocker for program-order must include a
+read of $x$ or a release (to get order from the write) and an acquire (to get
+the order to the read).
+
+What if an acquire preceded by a release or non-local read on $x$ wiped the
+local state of $x$?  The problem with this kind of solution (and P6 in ooplsa
+specifically) is that it destroys roach-motel.
 
@@
   \begin{align*}
-    \DL{\aLoc}{\mRLX}&=\TRUE
-    &\DS{\aLoc}{\mRLX}&=[\TRUE/\Dx{\aLoc}] 
-    \\
-    \DL{\aLoc}{\mRA}&=\Dx{\aLoc}%\land\D{}
-    &\DS{\aLoc}{\mRA}&=[\FALSE/\D]
-    \\
-    \DL{\aLoc}{\mSC}&=\Dx{\aLoc}%\land\D{}
-    &\DS{\aLoc}{\mSC}&=[\FALSE/\D]
-  \end{align*}
+    \DL{\aLoc}{\amode}&=
+    \begin{cases}
+      \Dx{\aLoc} & \textif \amode\neq\mRLX
+      \\
+      \TRUE& \textotherwise
+    \end{cases}
+    &
+    \DS{\aLoc}{\amode}&=
+    \begin{cases}
+      [\TRUE/\Dx{\aLoc}]  & \textif \amode=\mRLX
+      \\
+      [\TRUE/\Dx{\aLoc}][\FALSE/\D] & \textotherwise
+    \end{cases}
+  \end{align*}
+  % \begin{align*}
+  %   \DS{\aLoc}{\mRLX}&=[\TRUE/\Dx{\aLoc}] 
+  %   &\DL{\aLoc}{\mRLX}&=\TRUE
+  %   \\
+  %   \DS{\aLoc}{\mRA}&=[\FALSE/\D]
+  %   &\DL{\aLoc}{\mRA}&=\Dx{\aLoc}%\land\D{}
+  %   \\
+  %   \DS{\aLoc}{\mSC}&=[\FALSE/\D]
+  %   &\DL{\aLoc}{\mSC}&=\Dx{\aLoc}%\land\D{}
+  % \end{align*}
 \end{definition}
@@
       \bForm
-      \DS{\aLoc}{\amode}
-      [(\Qw{\aLoc}\land\aExp{=}\aVal)/\Qw{\aLoc}],
+      \DS{\aLoc}{\amode} \land\aExp{=}\aVal,
+      %[(\Qw{\aLoc}\land\aExp{=}\aVal)/\Qw{\aLoc}],
     \end{math}
@@
     \end{math}
-  \item[\ref{L4})]
-    $\aTr{\bEvs}{\bForm}$ implies
-    \begin{math}
-      \aVal{=}\aReg\limplies\bForm,
-    \end{math}
+  % \item[\ref{L4})]
+  %   $\aTr{\bEvs}{\bForm}$ implies
+  %   \begin{math}
+  %     \aVal{=}\aReg\limplies\bForm,
+  %   \end{math}
   \item[\ref{L5})]
@@
 \begin{example}
+  \label{ex:dgr2}
   Revisiting \refex{ex:dgr1} and eliding irrelevant transformers:
@@
       %(\Qw{\aLoc}\limplies\aExp{\neq}\aVal)\land
-      \bForm %\noSUB{[\aExp/\aLoc]},
-      [(\Qw{\aLoc}\land\aExp{=}\aVal)/\Qw{\aLoc}]
+      \bForm 
       \DS{\aLoc}{\amode}
+      \land\aExp{=}\aVal %\noSUB{[\aExp/\aLoc]},
     \end{math}
@@
       \bForm %\noSUB{[\aExp/\aLoc]}.
+      \DS{\aLoc}{\amode}
       [\FALSE/\QS{\aLoc}{\amode}]
-      \DS{\aLoc}{\amode}
     \end{math}
@@
       % \lnot\Qx{\aLoc} \land
-      \DL{\aLoc}{\amode}{} \land ((\aVal{=}\aReg\lor\RW)\limplies\bForm[\FALSE/\QL{\aLoc}{\amode}]).
+      \DL{\aLoc}{\amode}{} \land
+      ((\aVal{=}\aReg\lor\RW)\limplies\bForm[\FALSE/\QL{\aLoc}{\amode}]),
+      %\textwhen \aEvs\neq\emptyset
     \end{math}
+  \item[\ref{L6})] 
+    $\aTr{\dEvs}{\bForm}$ implies
+    \begin{math}
+      \DL{\aLoc}{\amode}{} \land
+      \bForm,
+    \end{math}
+    when $\aEvs=\emptyset$.
   \end{enumerate}  

### 2021-03-29 paper/1SUB/egc.tex
saved old version

\section{The ARMv8 EGC Model}
Copyright info:
\begin{verbatim}
(*
 * The ARMv8 Application Level Memory Model.
 *
 * See section B2.3 of the ARMv8 ARM:
 *   https://developer.arm.com/docs/ddi0487/latest/arm-architecture-reference-manual-armv8-for-armv8-a-architecture-profile
 *
 * Author: Will Deacon <will.deacon@arm.com>
 * Author: Jade Alglave <jade.alglave@arm.com>
 *
 * Copyright (C) 2016-2020, ARM Ltd.
 * All rights reserved.
 *
 * Redistribution and use in source and binary forms, with or without
 * modification, are permitted provided that the following conditions are
 * met:
 *
 *     * Redistributions of source code must retain the above copyright
 *       notice, this list of conditions and the following disclaimer.
 *     * Redistributions in binary form must reproduce the above copyright
 *       notice, this list of conditions and the following disclaimer in
 *       the documentation and/or other materials provided with the
 *       distribution.
 *     * Neither the name of ARM nor the names of its contributors may be
 *       used to endorse or promote products derived from this software
 *       without specific prior written permission.
 *
 * THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS "AS
 * IS" AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED
 * TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A
 * PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT
 * HOLDER OR CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
 * SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT LIMITED
 * TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE, DATA, OR
 * PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY THEORY OF
 * LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING
 * NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF THIS
 * SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
 *)

\end{verbatim}

\texttt{armv8-util.cat}:
\begin{verbatim}
"ARMv8 AArch64"

(*
 * Include the cos.cat file shipped with herd.
 * This builds the co relation as a total order over writes to the same
 * location and then consequently defines the fr relation using co and
 * rf.
 *)
include "cos.cat"

(*
 * Include aarch64fences.cat so that barriers show up in generated diagrams.
 *)
include "aarch64fences.cat"

(* Helper functions *)
procedure included(r1, r2) =
  empty r1 \ r2
end

procedure equal(r1, r2) =
  call included(r1, r2)
  call included(r2, r1)
end

let intervening-write(r) = r & (_ * W); r

(*
 * As a restriction of the model, all observers are limited to the same
 * inner-shareable domain. Consequently, the ISH, OSH and SY barrier
 * options are all equivalent to each other.
 *)
let dsb.full = DSB.ISH | DSB.OSH | DSB.SY
let dsb.ld = DSB.ISHLD | DSB.OSHLD | DSB.LD
let dsb.st = DSB.ISHST | DSB.OSHST | DSB.ST

(*
 * A further restriction is that standard litmus tests are unable to
 * distinguish between DMB and DSB instructions, so the model treats
 * them as equivalent to each other.
 *)
let dmb.full = DMB.ISH | DMB.OSH | DMB.SY | dsb.full
let dmb.ld = DMB.ISHLD | DMB.OSHLD | DMB.LD | dsb.ld
let dmb.st = DMB.ISHST | DMB.OSHST | DMB.ST | dsb.st

(* Flag any use of shareability options, due to the restrictions above. *)
flag ~empty (dmb.full | dmb.ld | dmb.st) \
	    (DMB.SY | DMB.LD | DMB.ST | DSB.SY | DSB.LD | DSB.ST)
	    as Assuming-common-inner-shareable-domain

let IM0 = loc & ((IW * (M\IW)) | ((W\FW) * FW))

let intrinsic = (iico_data | iico_ctrl)+
\end{verbatim}

\texttt{armv8-common.cat}:
\begin{verbatim}
"ARMv8 AArch64"

include "armv8-util.cat"

(* Coherence-after *)
let ca = fr | co

(* Local read successor *)
let lrs = [W]; po-loc \ intervening-write(po-loc); [R]

(* Local write successor *)
let lws = po-loc; [W]

(* Observed-by *)
let obs = rfe | fre | coe

(* Read-modify-write *)
let rmw = lxsx | amo

(* Dependency-ordered-before *)
let dob = addr | data
	| ctrl; [W]
	| (ctrl | (addr; po)); [ISB]; po; [R]
	| addr; po; [W]
	| (addr | data); lrs

(* Atomic-ordered-before *)
let aob = rmw
	| [W & range(rmw)]; lrs; [A | Q]

(* Barrier-ordered-before *)
let bob = po; [dmb.full]; po
        | po; ([A];amo;[L]); po
	| [L]; po; [A]
	| [R]; po; [dmb.ld]; po
	| [A | Q]; po
	| [W]; po; [dmb.st]; po; [W]
	| po; [L]

(* Tag-ordered-before *)
let tob = [R & T]; intrinsic; [M \ T]

(* Locally-ordered-before *)
let rec lob = lws
        | dob
	| aob
	| bob
        | tob
	| lob; lob

(* Internal visibility requirement *)
acyclic po-loc | ca | rf as internal

(* Atomic: Basic LDXR/STXR constraint to forbid intervening writes. *)
empty rmw & (fre; coe) as atomic
\end{verbatim}

\texttt{egc.cat}:
\begin{verbatim}
"Armv8 External global completion requirement, not mixed-size"

include "armv8-common.cat"

let gc-req = (W * _) | (R * _) & ((range(rfe) * _) | (rfi^-1; lob))
let preorder-gcb = IM0 | lob & gc-req

with gcb from linearisations(M, preorder-gcb)
~empty gcb

let gcbl = gcb & loc
let rf-gcb = (W * R) & gcbl \ intervening-write(gcbl)
let co-gcb = (W * W) & gcbl

call equal(rf, rf-gcb)
call equal(co, co-gcb)
\end{verbatim}

### 2021-03-29 paper/1SUB/intro.tex
saved old version

--- 2021-02-11 paper/intro.tex
@@
           \xform{wz0d}{\bForm}{below=of wz0}
-          \xo{wz0}{wz0d}
+          \xo[out=-30]{wz0}{wz0d}
       \end{tikzinline}}
@@
         \po{ry0}{wz0}
-        \xo{rx0}{rx0ry0d}
-        \xo{ry0}{rx0ry0d}
-        \xo{wz0}{rx0ry0d}
+        \xos{rx0}{rx0ry0d}
+        \xos{ry0}{rx0ry0d}
+        \xos{wz0}{rx0ry0d}
       \end{tikzinline}}
@@
         \po{ry0}{wz0}
-        \xo{rx0}{rx0d}
-        \xo{ry0}{ry0d}
-        \xo{rx0}{rx0ry0d}
-        \xo{ry0}{rx0ry0d}
+        \xos{rx0}{rx0d}
+        \xos{ry0}{ry0d}
+        \xos{rx0}{rx0ry0d}
+        \xos{ry0}{rx0ry0d}
       \end{tikzinline}}
@@
         \event{rx0}{\DR{x}{0}}{}
-        \event{ry0}{\DR{y}{0}}{right=of rx0}
-        \xform{rx0ry0d}{(r{=}0) \limplies(s{=}0) \limplies \bForm}{above=of ry0}
-        \xform{rx0d}{(r{=}0) \limplies \bForm}{below=of rx0}
-        \xform{ry0d}{(s{=}0) \limplies \bForm}{right=of rx0d}
+        \event{ry0}{\DR{y}{0}}{right=5em of rx0}
+        \xform{rx0ry0d}{(r{=}0) \limplies(s{=}0) \limplies \bForm}{above
+          right=1ex and -2.5em of rx0}
+        \xform{rx0d}{(r{=}0) \limplies \bForm}{left=of rx0}
+        \xform{ry0d}{(s{=}0) \limplies \bForm}{right=of ry0}
         \xform{wz0i}{\bForm}{right=of rx0ry0d}
-        \xo{rx0}{rx0d}
-        \xo{ry0}{ry0d}the \emph{independent}
+        \xos{rx0}{rx0d}
+        \xos{ry0}{ry0d}the \emph{independent}
 transformer
-        \xo{rx0}{rx0ry0d}
-        \xo{ry0}{rx0ry0d}
+        \xos{rx0}{rx0ry0d}
+        \xos{ry0}{rx0ry0d}
       \end{tikzinline}}
@@
         \po{ry0}{wz0}
-        \xo{ry0}{ry0d}
+        \xo[xleft]{ry0}{ry0d}
       \end{tikzinline}}
@@
 
-This paper is the first model of relaxed memory with a compositional
-semantics for sequential composition.  It shows how pomsets with
-preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} can be combined
-with predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75}.
+In this paper, we show how pomsets with
+preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} can be combined with
+predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75} to create a
+compositional semantics for sequential composition.
 \begin{itemize}
@@
 
-As far as we are aware, there are no previous attempts to provide a
-compositional semantics of sequential composition in a relaxed memory model.
-For a discussion of related work for relaxed memory models in general, see
-\cite{DBLP:journals/pacmpl/JagadeesanJR20}.
+\section{Related Work}
+
+\begin{itemize}
+\item Models with strong dependencies that disallow compiler optimizations
+  and efficient implementation on \armeight{}:
+  \cite{Lamport:1979:MMC:1311099.1311750,
+    DBLP:conf/pldi/LahavVKHD17,
+    Dolan:2018:BDR:3192366.3192421,
+    DBLP:conf/pldi/LahavVKHD17,
+    DBLP:conf/lics/JeffreyR16,
+    Boehm:2014:OGA:2618128.2618134,
+    DBLP:journals/corr/abs-1804-04214}.
+\item Models with weak dependencies that allow some forms of thin-air
+  reasoning:
+  \cite{Manson:2005:JMM:1047659.1040336,
+    DBLP:conf/esop/JagadeesanPR10,
+    DBLP:conf/popl/KangHLVD17,
+    DBLP:journals/pacmpl/ChakrabortyV19}
+\item Models that do not calculate dependencies:
+  \cite{Batty:2011:MCC:1926385.1926394,
+    DBLP:journals/pacmpl/WattRP19,
+    DBLP:conf/pldi/WattPPBDFPG20,
+    goMM}.
+\end{itemize}
+Like us, \citet{DBLP:journals/corr/abs-1804-04214} use pomsets to create a
+compositional account of sequential composition.  However, their model
+requires a fence after every read on \armeight{}, and uses syntactic
+dependencies, thus invalidating many compiler optimizations.

### 2021-03-30 paper/intro.tex
Updated defs
added .tex everywhere for latexdiff

@@
 % \begin{figure*}
-%   \input{fig-sp}
+%   \input{fig-sp.tex}
 %   \caption{Weakest precondition semantics}

### 2021-03-30 paper/arm.tex
add .tex to input statements for latexdiff
update macros to distinguish REL and ACQ

@@
 
-We discuss \armeight{} using \emph{external global completion} (\EGC{})
+We discuss \armeight{} using \emph{external completion} (\EC{})
 \cite{alglave-git-alternate} %,armed-cats},
@@
     \IF{x}\THEN\PW{y}{2}\FI\SEMI
-    \IF{\PR[\mRA]{z}{}}\THEN
+    \IF{\PR[\mACQ]{z}{}}\THEN
       \PR{x}{s}\SEMI
@@
     \PW{x}{2}\SEMI
-    \IF{\PR[\mRA]{y}{}}\THEN
+    \IF{\PR[\mACQ]{y}{}}\THEN
       \PW{x}{3}\SEMI
-      \PW[\mRA]{z}{1}
+      \PW[\mREL]{z}{1}
     \FI
@@
         \event{2c}{\DW{y}{1}}{right=of 2b}
-        \raevent{2d}{\DR[\mRA]{z}{1}}{right=of 2c}
+        \raevent{2d}{\DR[\mACQ]{z}{1}}{right=of 2c}
         \event{2e}{\DR{x}{4}}{right=of 2d}
@@
         \event{3a}{\DW{x}{2}}{below=of 2a}
-        \raevent{3b}{\DR[\mRA]{y}{1}}{right=of 3a}
+        \raevent{3b}{\DR[\mACQ]{y}{1}}{right=of 3a}
         \event{3c}{\DW{x}{3}}{right=of 3b}
-        \raevent{3d}{\DW[\mRA]{z}{1}}{right=of 3c}
+        \raevent{3d}{\DW[\mREL]{z}{1}}{right=of 3c}
         \event{3e}{\DR{w}{4}}{right=of 3d}
@@
     \PW{x}{2}\SEMI 
-    \PR[\mRA]{x}{r}\SEMI
+    \PR[\mACQ]{x}{r}\SEMI
     \PW{y}{1} \PAR
     \PW{y}{2}\SEMI
-    \PW[\mRA]{x}{1}
+    \PW[\mREL]{x}{1}
     \\
@@
         \event{a}{\DW{x}{2}}{}
-        \raevent{b}{\DR[\mRA]{x}{2}}{right=of a}
+        \raevent{b}{\DR[\mACQ]{x}{2}}{right=of a}
         \event{c}{\DW{y}{1}}{right=of b}
         \event{d}{\DW{y}{2}}{right=2.5em of c}
-        \raevent{e}{\DW[\mRA]{x}{1}}{right=of d}
+        \raevent{e}{\DW[\mREL]{x}{1}}{right=of d}
         \rf{a}{b}
@@
   Under \EGC{}, this is explained by dropping the order
-  $\DRP[\mRA]{x}{2}\xsync\DWP{y}{1}$, because $\DRP[\mRA]{x}{2}$ is fulfilled
+  $\DRP[\mACQ]{x}{2}\xsync\DWP{y}{1}$, because $\DRP[\mACQ]{x}{2}$ is fulfilled
   by a relaxed write in the same thread.
@@
     \begin{gathered}[t]
-      \PR[\mRA]{x}{r}
-      \\
-      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-          \raevent{b}{\DR[\mRA]{x}{2}}{}
+      \PR[\mACQ]{x}{r}
+      \\
+      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
+          \raevent{b}{\DR[\mACQ]{x}{2}}{}
           % \xform{x1d}{2{=}r\limplies\bForm}{below=of b}
@@
     \begin{gathered}[t]
-      \PR[\mRA]{x}{r}
+      \PR[\mACQ]{x}{r}
       \SEMI
@@
       \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-          \raevent{b}{\DR[\mRA]{x}{2}}{}
+          \raevent{b}{\DR[\mACQ]{x}{2}}{}
           \event{c}{\Dx{x}\mid\DW{y}{1}}{right=.7em of b}
@@
       \SEMI
-      \PR[\mRA]{x}{r}
+      \PR[\mACQ]{x}{r}
       \SEMI
@@
           \event{a}{\DW{x}{2}}{}
-          \raevent{b}{\DR[\mRA]{x}{2}}{right=of a}
+          \raevent{b}{\DR[\mACQ]{x}{2}}{right=of a}
           \event{c}{\DW{y}{1}}{right=of b}
@@
 %   \PW{x}{2}\SEMI 
-%   \PR[\mRA]{x}{r}\SEMI
+%   \PR[\mACQ]{x}{r}\SEMI
 %   \PR{y}{s} \PAR
 %   \PW{y}{2}\SEMI
-%   \PW[\mRA]{x}{1}
+%   \PW[\mREL]{x}{1}
 %   \\
@@
 %     \event{a}{\DW{x}{2}}{}
-%     \raevent{b}{\DR[\mRA]{x}{2}}{right=of a}
+%     \raevent{b}{\DR[\mACQ]{x}{2}}{right=of a}
 %     \event{c}{\DR{y}{0}}{right=of b}
 %     \event{d}{\DW{y}{2}}{right=2.5em of c}
-%     \raevent{e}{\DW[\mRA]{x}{1}}{right=of d}
+%     \raevent{e}{\DW[\mREL]{x}{1}}{right=of d}
 %     \rf{a}{b}
@@
     \begin{gathered}
-      x\GETS0\SEMI x\GETS 1\SEMI y^\mRA\GETS1
+      \PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW[\mREL]{y}{1}
       \PAR
-      r\GETS y\SEMI \IF{r}\THEN s\GETS x\FI
+      \PR{y}{r}\SEMI \IF{r}\THEN \PR{x}{s}\FI
       \\[-.4ex]
@@
           \event{wx1}{\DW{x}{1}}{right=of wx0}
-          \raevent{wy1}{\DW[\mRA]{y}{1}}{right=of wx1}
+          \raevent{wy1}{\DW[\mREL]{y}{1}}{right=of wx1}
           \event{ry1}{\DR{y}{1}}{right=2.5em of wy1}
@@
     \begin{gathered}[t]
-      \IF{r}\THEN s\GETS x\FI
+      \IF{r}\THEN \PR{x}{s}\FI
       \\
@@
       \SEMI
-      \IF{r}\THEN s\GETS x\FI
+      \IF{r}\THEN \PR{x}{s}\FI
       \\
@@
       % \SEMI
-      % \IF{r}\THEN s\GETS x\FI
+      % \IF{r}\THEN \PR{x}{s}\FI
       % \\
@@
     \begin{gathered}[t]
-      \PR[\mRA]{y}{r}
+      \PR[\mACQ]{y}{r}
       \\
       \hbox{\begin{tikzinline}[node distance=0.5em and .5em]
-          \raevent{b}{\QL{y}{\mRA}\mid\DR[\mRA]{y}{1}}{}
+          \raevent{b}{\QL{y}{\mRA}\mid\DR[\mACQ]{y}{1}}{}
           \xform{x1i}{\Dx{y}\land(\RW\limplies\bForm[\FALSE/\QL{\aLoc}{\mRA}])}{right=of b} %x1d}
@@
     \begin{gathered}[t]
-      \PR[\mRA]{y}{r}
+      \PR[\mACQ]{y}{r}
       \SEMI
@@
       \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-          \raevent{b}{\QL{y}{\mRA}\mid\DR[\mRA]{y}{1}}{}
+          \raevent{b}{\QL{y}{\mRA}\mid\DR[\mACQ]{y}{1}}{}
           \event{c}{\Dx{y}\land(\RW\limplies\FALSE)\mid\DW{x}{0}}{right=of b}
@@
     \begin{gathered}[t]
-      % \PR[\mRA]{y}{r}
+      % \PR[\mACQ]{y}{r}
       % \SEMI
@@
       \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-          \raevent{b}{\TRUE\mid\DR[\mRA]{y}{1}}{}
+          \raevent{b}{\TRUE\mid\DR[\mACQ]{y}{1}}{}
           \event{c}{\Dx{y}\land(\FALSE\limplies\FALSE)\mid\DW{x}{0}}{right=of b}
@@
   the pomset can only contribute to a top-level pomset if the thread is
-  preceded by a relaxed write to $y$, allowing $\DRP[\mRA]{y}{1}$ to be
+  preceded by a relaxed write to $y$, allowing $\DRP[\mACQ]{y}{1}$ to be
   downgraded to a relaxed read.

### 2021-04-14 paper/abstract.tex
Extra polish
intro

--- 2021-01-22 paper/abstract.tex
@@
+% This paper presents the first compositional definition of sequential
+% composition that applies to a relaxed memory model weak enough to allow
+% efficient implementation on Arm.  We extend the denotational model of pomsets
+% with preconditions with predicate transformers. Previous work has shown that
+% pomsets with preconditions are a model of concurrent composition, and that
+% predicate transformers are a model of sequential composition.  This paper
+% show how they can be combined.
+
 % Program logics and semantics tell us that in order to derive
@@
 we execute (S1, state0) to arrive at state1, then execute (S2, state1) to
-arrive at the final state2.
-
-This is a delightfully simple story that can be explained to children.  It is
-also a lie.
-
+arrive at the final state2.  This is a delightfully simple story that can be
+explained to children.  It is also a lie.
+%
 Processors execute instructions out of order, due to pipelines and caches.
@@
 however, all bets are off.  A formal attempt to understand the resulting mess
-is known as a ``relaxed memory model.''
+is known as a ``relaxed memory model.''  The relaxed memory models that have
+been proposed to date either fail to address sequential composition, or
+overly restrict processors and compilers.
 
-Most of the relaxed memory models that have been proposed are designed to
-help us understand whole program execution: they have no compositionality
-properties whatsoever.  Recently, denotation models have appeared that treat
-\emph{concurrent} execution compositionally.  One such model is ``Pomsets with
-Preconditions''.  It remains an open question, however, whether it is
-possible to treat \emph{sequential} execution compositionally in such a model,
-without overly restricting processors and compilers.
-
-We propose adding families of predicate transformers to Pomsets with
-Preconditions.  The resulting model is denotational, supporting both parallel
-and sequential composition.  When composing (S1;S2), the predicate
+To support sequential composition, we propose adding families of predicate
+transformers to the existing model of ``Pomsets with Preconditions,'' which
+already supports parallel composition.  When composing (S1;S2), the predicate
 transformer used to validate the precondition of an event in S2 is chosen
-based on the dependency order from S1 into this event.  As usual in work on
-relaxed memory, we have not handled loops or recursion.
-
-Happily, most of the results expected of a relaxed memory model can be
-established by appeal to prior work.  So here we are able to concentrate on
-the model itself.  The model is formalized in Agda, where we have established
-associativity for sequential composition.
-
-For the memory-model specialist, we retain the good properties of the prior
-work on Pomsets with Preconditions, fixing some errors along the way.  These
-properties include efficient implementation on ARMv8, support for compiler
-optimizations, support for logics that prove the absence of thin-air
-behaviors, and a local data race freedom theorem.
+based on the semantic dependencies from S1 into this event.  Our model
+retains the good properties of the prior work, including efficient
+implementation on Arm8, support for compiler optimizations, support for
+logics that prove the absence of thin-air behaviors, and a local data race
+freedom theorem.

### 2021-04-14 paper/arm.tex
intro

--- 2021-02-25 paper/arm.tex
@@
-\section{Efficient Implementation on ARMv8}
-\label{sec:arm}
+\section{Arm}
+For simplicity, we restrict to top level parallel composition and ignore
+fences\footnote{Fences are not actions in \armeight{}, which complicates the
+  theorem statements.}.
 
-We discuss \armeight{} using \emph{external global completion} (\EGC{})
-\cite{alglave-git-alternate} %,armed-cats},
-\cite[\textsection B2.3.6]{arm-reference-manual}
-which is very close to our model.
+\subsection{Arm executions}
+\begin{definition}
+  An \emph{\armeight{} execution graph}, $\aEG$, is tuple
+  $(\aEvs, \labeling, {\rpoloc}, {\rlob})$ such that
+  \begin{enumerate}[,label=(\textsc{a}\arabic*),ref=\textsc{a}\arabic*]
+  \item $\Event\subseteq\AllEvents$ is a set of {events},
+  \item $\labeling: \Event \fun \Act$ defines a {label} for each event,
+  \item ${\rpoloc} : \Event\times\Event$, is a per-thread, per-location total
+    order, capturing \emph{per-location program order},
+  \item ${\rlob} : \Event\times\Event$, is a per-thread partial order capturing
+    \emph{locally-ordered-before}, such that
+    \begin{enumerate}
+    \item \label{arm-lob-poloc}
+      ${\rpoloc} \cup {\rlob}$ is acyclic.
+    \end{enumerate}
+  \end{enumerate}
+\end{definition}
 
-\subsection{Downgraded Reads (\xDGR)}
-\label{sec:downgrade}
+The definition of $\rlob$ is complex.  Comparing with our definition of
+sequential composition, it is sufficient to note that $\rlob$ includes
+\begin{enumerate}[label=(\textsc{l}\arabic*),ref=\textsc{l}\arabic*]
+\item read-write dependencies, required by \ref{seq-kappa},
+\item synchronization delay of ${\reorderra}$, required by \ref{seq-le-delays},
+\item $\mSC$ access delay of ${\eqreordersc}$, required by \ref{seq-le-delays},
+\item write-write and read-to-write coherence delay of ${\reorderco}$, required by \ref{seq-le-delays},
+\end{enumerate}
+and that $\rlob$ does \emph{not} include
+\begin{enumerate}[resume,label=(\textsc{l}\arabic*),ref=\textsc{l}\arabic*]
+\item \label{lob-rr} \labeltext[above]{}{lob-le}
+  read-read control dependencies, required by \ref{seq-kappa},
+\item \label{lob-rf}
+  write-to-read order of $\rrfx$, required by \ref{seq-le-rf},
+\item \label{lob-wr}
+  write-to-read coherence delay of ${\reorderco}$, required by \ref{seq-le-delays}.
+\end{enumerate}
+
+\begin{definition}
+  Execution $\aEG$ is
+  \emph{$({\rco}, {\rrfx}, {\rgcb})$-valid}, under \emph{External Global
+    Consistency} (\EGC{}) if
+  \begin{enumerate}[label=(\textsc{a}\arabic*),ref=\textsc{a}\arabic*]
+    \setcounter{enumi}{4}
+  %\item[\eqref{arm-co}] and \eqref{arm-rf}, as for \EC,
+  \item \label{arm-co}
+    ${\rco} : \Event\times\Event$, is a per-location total order on
+    writes, capturing \emph{coherence}, 
+  \item \label{arm-rf}
+    ${\rrfx} : \Event\times\Event$, is a surjective and injective
+    relation on reads, capturing \emph{reads-from}, such that
+    \begin{enumerate}
+    \item \label{arm-match}
+      if $\bEv\xrfx\aEv$ then $\labeling(\bEv) \rmatches \labeling(\aEv)$,      
+    \item \label{arm-local}
+      ${\rpoloc} \cup {\rco} \cup {\rrfx} \cup {\rfr}$ is acyclic,
+      where $\aEv\xfr\cEv$ if %$(\exists\bEv)$
+      $\aEv\xrfxinv\bEv\xco\cEv$, for some $\bEv$,
+      % \stepcounter{enumi}
+      % \item[] \labeltext[\textsc{a}2]{}{arm-cb}
+    \end{enumerate}
+    % \item ${\rpoloc} \cup {\rco} \cup {\rrfx} \cup {\rfr}$ is acyclic, where
+    %   $\bEv\xfr\aEv$ if $(\exists\cEv)$ $\bEv\xrfxinv\cEv\xco\aEv$,
+  \item \label{arm-gcb}
+    ${\rgcb}\supseteq\PBR{{\rco}\cup{\rrfx}}$ is a linear order %, capturing \emph{globally completes before}, %$\rcb : % \Event\times\Event$
+    such that 
+    \begin{enumerate}%[leftmargin=0pt]
+      % \item if $\bEv\xco\aEv$ then $\bEv\xgcb\aEv$,
+      % \item if $\bEv\xrfx\aEv$ then $\bEv\xgcb\aEv$,
+    \item \label{arm-gcb-blocks}
+      if $\bEv\xrfx\aEv$ and $\labeling(\cEv) \rblocks \labeling(\aEv)$ then either $\cEv\xgcb\bEv$ or $\aEv\xgcb\cEv$, 
+    \item \label{arm-gcb-lob}
+      if $\aEv\xlob\cEv$ then either $\aEv\xgcb\cEv$ or $(\exists\bEv)$
+      $\bEv\xrfx\aEv$ and $\bEv\xpoloc\aEv$ but not $\bEv\xlob\cEv$.
+    \end{enumerate}
+  \end{enumerate}
+
+  Execution $\aEG$ is
+  \emph{$({\rco}, {\rrfx}, {\rcb})$-valid} under \emph{External Consistency} (\EC{}) if
+  \begin{enumerate}[resume,label=(\textsc{a}\arabic*),ref=\textsc{a}\arabic*]
+  \item[\eqref{arm-co}] and \eqref{arm-rf}, as for \EGC,
+  \item \label{arm-cb}
+    ${\rcb}\supseteq\PBR{{\rco}\cup{\rlob}}$ is a linear order %, capturing \emph{completes before}, %$\rcb : % \Event\times\Event$
+    such that if $\bEv\xrfx\aEv$ then either
+    \begin{enumerate}%[leftmargin=0pt]
+    \item \label{arm-rfe}
+      % $\bEv\xcb\aEv$ and $\cEv\rblocks\aEv$ then either $\cEv\xcb\bEv$ or $\aEv\xcb\cEv$, or 
+      $\bEv\xcb\aEv$ and if $\labeling(\cEv) \rblocks \labeling(\aEv)$ then either $\cEv\xcb\bEv$ or $\aEv\xcb\cEv$, or
+      % $\bEv\xcb\aEv$ and $(\not\exists\cEv)$ such that $\bEv\xcb\cEv\xcb\aEv$ and $\cEv\rblocks\aEv$, or 
+    \item \label{arm-rfi}
+      $\bEv\xcbinv\aEv$ and $\bEv\xpoloc\aEv$ and $(\not\exists\cEv)$ $\labeling(\cEv) \rblocks \labeling(\aEv)$ and $\bEv\xpoloc\cEv\xpoloc\aEv$.
+      % $\bEv\xpoloc\aEv$ and $(\not\exists\cEv)$ such that $\bEv\xpoloc\cEv\xpoloc\aEv$ and $\cEv\rblocks\aEv$.
+      % \item if $\bEv\xco\aEv$ then $\bEv\xgcb\aEv$,
+      % \item if $\bEv\xrfx\aEv$ then $\bEv\xgcb\aEv$,
+      % \item if $\bEv\xrfx\aEv$ and $\cEv\rblocks\aEv$ then either $\cEv\xgcb\bEv$ or $\aEv\xgcb\cEv$,
+      % \item if $\bEv\xlob\aEv$ then either $\bEv\xgcb\aEv$ or $(\exists\cEv)$
+      %   $\cEv\xrfx\bEv$ and $\cEv\xpoloc\bEv$ but not $\cEv\xlob\aEv$.
+    \end{enumerate}
+  \end{enumerate}
+\end{definition}
+\cite{armed} explain \EGC{} and \EC{} using the following example.\footnote{We have
+changed an address dependency in the first thread to a data dependency.}
+\begin{gather*}
+  \PW{x}{1}\SEMI 
+  \PR{x}{r}\SEMI
+  \PW{y}{r} \PAR
+  \PR[\mACQ]{y}{1}\SEMI
+  \PR{x}{s}
+  \\
+  \tag{\cmark\armeight}
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a}{\DW{x}{1}}{}
+      \event{b}{\DR{x}{1}}{right=of a}
+      \event{c}{\DW{y}{1}}{right=of b}
+      \raevent{d}{\DR[\mACQ]{y}{1}}{right=2.5em of c}
+      \event{e}{\DR{x}{0}}{right=of d}
+      \rfx{a}{b}
+      \lob{b}{c}
+      \rfx{c}{d}
+      \lob{d}{e}
+      \co[out=-165,in=-15]{e}[above right]{a}
+    \end{tikzinline}}
+\end{gather*}
+\EGC{} drops $\rlob$-order in the first thread using \ref{arm-gcb-lob}, since
+$\DWP{x}{1}$ is not $\rlob$-ordered before $\DWP{y}{1}$.
+\begin{gather*}
+  \tag{$\rgcb$}
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a}{\DW{x}{1}}{}
+      \event{b}{\DR{x}{1}}{right=of a}
+      \event{c}{\DW{y}{1}}{right=of b}
+      \raevent{d}{\DR[\mACQ]{y}{1}}{right=2.5em of c}
+      \event{e}{\DR{x}{0}}{right=of d}
+      \gcbz{a}{b}
+      \gcbz{c}{d}
+      \gcbz{d}{e}
+      \gcbz[out=-165,in=-15]{e}{a}
+      % \rfx{a}{b}
+      % %\lob{b}{c}
+      % \rfx{c}{d}
+      % \lob{d}{e}
+      % \co[out=-165,in=-15]{e}[above right]{a}
+    \end{tikzinline}}
+\end{gather*}
+\EC{} drops $\rrfx$-order in the first thread using \ref{arm-rfi}.
+\begin{gather*}
+  \tag{$\rcb$}
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a}{\DW{x}{1}}{}
+      \event{b}{\DR{x}{1}}{right=of a}
+      \event{c}{\DW{y}{1}}{right=of b}
+      \raevent{d}{\DR[\mACQ]{y}{1}}{right=2.5em of c}
+      \event{e}{\DR{x}{0}}{right=of d}
+      \cbz{b}{c}
+      \cbz{c}{d}
+      \cbz{d}{e}
+      \cbz[out=-165,in=-15]{e}{a}
+      % %\rfx{a}{b}
+      % \lob{b}{c}
+      % \rfx{c}{d}
+      % \lob{d}{e}
+      % \co[out=-165,in=-15]{e}[above right]{a}
+    \end{tikzinline}}
+\end{gather*}
+
+\subsection{Arm Compilation 1}
+
+\citet{DBLP:journals/pacmpl/PodkopaevLV19} lowers to \armeight{} as follows:
+Relaxed access is implemented using \LDR/\STR, non-relaxed access using
+\LDAR/\STLR.  %, acquire and other fences using \DMBLD/\DMBSY.
+In this section, we consider a suboptimal strategy, which lowers non-relaxed
+reads to $(\DMBSY\SEMI\LDAR)$.
+
+We do not distinguish control dependencies from address dependencies, and
+therefore \ref{lob-rr} forces us to drop all dependencies between reads.  To
+achieve this, we modify the definition of $\labelingForm'_2$ in
+\reffig{fig:sem}.
+\begin{definition}
+  \label{def:semrr}
+  Let $\semrr{}$ be defined as in \reffig{fig:sem}, replacing the definition
+  of $\labelingForm'_2$ with:
+  \begin{displaymath}
+    \labelingForm'_2(\aEv)=
+    \begin{cases}
+      \aTr[1]{}{\labelingForm_2(\aEv)} & \text{if}\; \labeling(\aEv) \;\text{is a read}
+      \\
+      \aTr[1]{\Cdown{\aEv}}{\labelingForm_2(\aEv)} & \text{otherwise, where}\; \Cdown{\aEv}=\{ \cEv \mid \cEv \lt \aEv \}
+    \end{cases}
+    % $\Cdown{\aEv}=\{ \cEv \mid \cEv \lt \aEv \}$ if $\labeling(\aEv)$ is a write, and $\Cdown{\aEv}=\aEvs_1$, otherwise
+    % \Cdown{\aEv}=
+    % \begin{cases}
+    %   \{ \cEv \mid \cEv \lt \aEv \} & \textif \labeling(\aEv) \;\text{is a write}
+    %   \\
+    %   \aEvs_1 & \textotherwise
+    % \end{cases}
+  \end{displaymath}
+\end{definition}
+
+\begin{theorem}
+  \label{thm:ec}
+  Suppose $\aEG_1$ is $({\rco_1}, {\rrfx_1}, {\rgcb_1})$-valid for $\aCmd$
+  under the suboptimal lowering that maps non-relaxed reads to
+  $(\DMBSY\SEMI\LDAR)$.  Then there is a top-level pomset
+  $\aPS_2\in\semrr{\aCmd}$ such that $\aEvs_2=\aEvs_1$,
+  $\labeling_2=\labeling_1$, ${\rrfx_2}={\rrfx_1}$, and ${\le_2}={\rgcb_1}$.
+
+  \vspace{-.5\baselineskip}
+  \begin{proof}
+    First, we establish some lemmas about \armeight.
+
+    \vspace{-.5\baselineskip}
+    \begin{lemma}
+      \label{lemma:fr}
+      Suppose $\aEG$ is $({\rco}, {\rrfx}, {\rgcb})$-valid.  Then
+      ${\rgcb}\supseteq{\rfr}$.
+
+      \vspace{-.5\baselineskip}
+      \begin{proof}
+        Using the definition of ${\rfr}$ from \ref{arm-local}, we have
+        $\aEv\xrfxinv\bEv\xco\cEv$, and therefore $\labeling(\cEv)$ blocks
+        $\labeling(\aEv)$.    
+        Applying \ref{arm-gcb-blocks}, we have that either $\cEv\xgcb\bEv$ or $\aEv\xgcb\cEv$.
+        Since $\rgcb$ includes $\rco$, we have $\bEv\xgcb\cEv$, and therefore it
+        must be that $\aEv\xgcb\cEv$.
+      \end{proof}
+    \end{lemma}
+
+    \begin{lemma}
+      \label{lemma:wr}
+      Suppose $\aEG$ is $({\rco}, {\rrfx}, {\rgcb})$-valid and
+      $\cEv\xpoloc\aEv$, where $\labeling(\cEv)\rblocks\labeling(\aEv)$.
+      Then $\cEv\xgcb\aEv$.
+      %$\labeling(\cEv)=\DWP{x}{}$ and $\labeling(\aEv)=\DRP{x}{}$.
+
+      \vspace{-.5\baselineskip}
+      \begin{proof}
+        By way of contradiction, assume $\aEv\xgcb\cEv$.  If $\cEv\xrfx\aEv$
+        then by \ref{arm-gcb} we must also have $\cEv\xgcb\aEv$,
+        contradicting the assumption that $\rgcb$ is a total order.
+        %
+        Otherwise that there is some $\bEv\neq\cEv$ such that
+        $\bEv\xrfx\aEv$, and therefore $\bEv\xgcb\aEv$.  By transitivity,
+        $\bEv\xgcb\cEv$.  By the definition of $\rfr$, we have
+        $\aEv\xfr\cEv$.  But this contradicts \ref{arm-local}, since
+        $\cEv\xpoloc\aEv$.
+        %Applying \ref{arm-gcb-blocks}, we have that either $\cEv\xgcb\bEv$ or $\aEv\xgcb\cEv$.
+      \end{proof}
+    \end{lemma}
+    We show that all the order required in the pomset is also required by
+    \armeight{}.  \ref{rf-block} holds since ${\rcb_1}$ is consistent with
+    ${\rco_1}$ and ${\rfr_1}$.  As noted \ref{lob-le}, $\rlob$ includes the order
+    required by \ref{seq-kappa} and \ref{seq-le-delays}.  We need only show
+    that the order removed from \ref{arm-gcb-lob} can also be removed from
+    the pomset.  In order for \label{arm-gcb-lob} to remove order from $\aEv$
+    to $\cEv$, we must have $\bEv\xrfx\aEv$ and $\bEv\xpoloc\aEv$ but not
+    $\bEv\xlob\cEv$.  Because of our suboptimal lowering, it must be that
+    $\aEv$ is a relaxed read; otherwise the $\DMBSY$ would require
+    $\bEv\xlob\cEv$.  Thus we know that \ref{seq-le-delays} does not require
+    order from $\aEv$ to $\cEv$.  By chaining \ref{read-tau-ind} and
+    \ref{write-tau}, any dependence on the read can by satisfied without
+    introducing order in \ref{seq-kappa}.
+  \end{proof}  
+\end{theorem}
+
+% This model compiles correctly to arm using the lowering: Relaxed access is
+% implemented using \texttt{ldr}/\texttt{str}, non-relaxed reads using
+% \texttt{dmb st}\SEMI\texttt{ldar}, non-relaxed writes using \texttt{stlr},
+% acquire fences using \texttt{dmb}.\texttt{ld} and other fences using 
+% \texttt{dmb}.\texttt{sy}.
+% \begin{align*}
+%   \low{\PW[\gemode\mREL]{\REF{\aReg}}{\bReg}} &= \texttt{stlr}\;\bReg,\REF{\aReg}
+%   &
+%   \low{\PR[\mRLX]{\REF{\aReg}}{\bReg}} &= \texttt{ldr}\;\bReg,\REF{\aReg}
+%   \\
+%   \low{\PW[\mRLX]{\REF{\aReg}}{\bReg}} &= \texttt{str}\;\bReg,\REF{\aReg}
+%   &
+%   \low{\PR[\gemode\mACQ]{\REF{\aReg}}{\bReg}} &= \texttt{dmb st}\SEMI\texttt{ldar}\;\bReg,\REF{\aReg}
+% \end{align*}
+
+Downgrading messes up publication:
+\begin{gather*}
+  \PW{x}{\PR{x}{}+1}\SEMI
+  \PW[\mREL]{y}{1}
+  \PAR
+  \PW{x}{1}\SEMI
+  \IF{\PR[\mACQ]{y}{}\AND\PR[\mACQ]{x}{}}\THEN
+  \PR{z}{s}
+  \FI
+  \PAR
+  \PW{z}{1}\SEMI
+  \PW[\mREL]{x}{1}
+  \\
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{b1}{\DR{x}{1}}{}
+      \event{b2}{\DW{x}{2}}{right=of b1}
+      \raevent{b3}{\DW[\mREL]{y}{1}}{right=of b2}
+      \po{b1}{b2}
+      \sync{b2}{b3}
+      \event{a1}{\DW{x}{1}}{right=3em of b3}
+      \raevent{a2}{\DR[\mACQ]{y}{1}}{right=of a1}
+      \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
+      \event{a4}{\DR{z}{0}}{right=of a3}
+      \sync{a2}{a3}
+      \sync{a3}{a4}
+      \event{c1}{\DW{z}{1}}{right=3em of a4}
+      \raevent{c2}{\DW[\mREL]{x}{1}}{right=of c1}
+      \sync{c1}{c2}
+      \rf[out=-165,in=-15]{a1}{b1}
+      \rf[out=15,in=165]{b3}{a2}
+      \rf[out=-165,in=-15]{c2}{a3}
+      \wk{a4}{c1}
+    \end{tikzinline}}
+  \\
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{b1}{\DR{x}{1}}{}
+      \event{b2}{\DW{x}{2}}{right=of b1}
+      \raevent{b3}{\DW[\mREL]{y}{1}}{right=of b2}
+      \po{b1}{b2}
+      \sync{b2}{b3}
+      \event{a1}{\DW{x}{1}}{right=3em of b3}
+      \raevent{a2}{\DR[\mACQ]{y}{1}}{right=of a1}
+      \event{a3}{\DR{x}{1}}{right=of a2}
+      % \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
+      \event{a4}{\DR{z}{0}}{right=of a3}
+      \sync{a2}{a3}
+      \sync[out=15,in=165]{a2}{a4}
+      \event{c1}{\DW{z}{1}}{right=3em of a4}
+      \raevent{c2}{\DW[\mREL]{x}{1}}{right=of c1}
+      \sync{c1}{c2}
+      \rf[out=-165,in=-15]{a1}{b1}
+      \rf[out=15,in=165]{b3}{a2}
+      \rf[out=-165,in=-15]{c2}{a3}
+      \wk{a4}{c1}
+    \end{tikzinline}}
+\end{gather*}
+
+\subsection{Arm Compilation 2}
+
+\begin{definition}
+  \label{def:sem:frf}
+  Let $\frf{\semrr{}}$ be defined as for $\semrr{}$ in \refdef{def:semrr} and
+  \reffig{fig:sem}, changing \ref{seq-le-rf} and \ref{seq-le-delays}:
+  \begin{itemize}
+  \item[{\labeltext[\frf{\textsc{s}7b}]{(\frf{\textsc{s}7b})}{seq-le-rf-rf}}]
+    if $\labeling_1(\cEv) \rblocks \labeling_2(\aEv)$ then $\bEv\xrfx\aEv$
+    implies $\cEv\le\bEv$,
+  \item[{\labeltext[\frf{\textsc{s}7c}]{(\frf{\textsc{s}7c})}{seq-le-delays-rf}}]
+    if $\labeling_1(\bEv) \rdelaysp \labeling_2(\aEv)$ then $\bEv\le\aEv$,\\
+    where $\rdelayspdef$ replaces $\eqreorderco$ in \refdef{def:actions} of
+    $\rdelaysdef$ by
+    \begin{math}
+      {\reorderlws}
+      =
+      \{(\DW{\aLoc}{}, \DW{\aLoc}{}),\;(\DR{\aLoc}{}, \DW{\aLoc}{})\}
+    \end{math}.
+  \end{itemize}  
+\end{definition}
+The acronym $\textsf{lws}$ is adopted from \armeight.  It stands for
+\emph{Local Write Successor}.
+
+Note that \reflem{lem:rf:implies:le} fails for $\frf{\semrr{}}$, since
+$\bEv\xrfx\aEv$ may not imply $\bEv\le\aEv$ when $\bEv$ and $\aEv$ come from
+different sides of a sequential composition.  This means that $\rrfx$ must be
+verified during pomset construction, rather than post-hoc.  If one wants a
+post-hoc verification technique for $\rrfx$, it is possible to include
+program order ($\rpox$) in the pomset.
 
 \begin{example}
-  \label{ex:dgr1}
-  The following example is from \citeauthor{anton} \cite{anton}.  It is
-  allowed by \armeight{}, but disallowed by \refdef{def:pomsets-ra}.  The
-  coherence order between the writes can be witnessed by a separate thread,
-  which we have elided.
+  The obvious definition of $\rpox$ may be cyclic, due to the conditional. 
+\end{example}
+\begin{lemma}
+  $\aPS$ is top-level iff $\bEv\xrfx\aEv$ implies either
+  \begin{itemize}
+  \item external fulfillment: $\bEv\le\aEv$ and if $\labeling(\cEv)$ blocks
+    $\labeling(\aEv)$ then either $\cEv\le\bEv$ or $\aEv\le\cEv$, or
+  \item internal fulfillment: $\bEv\xpox\aEv$ and $(\not\exists\cEv)$
+    $\labeling(\cEv) \rblocks \labeling(\aEv)$ and $\bEv\xpox\cEv\xpox\aEv$.
+  \end{itemize}
+\end{lemma}
+% \begin{enumerate}
+% \item[(\textsc{m}7d)]
+%   if $\labeling(\cEv) \rblocks \labeling(\aEv)$
+%   then $\bEv\xrfx\aEv$ implies $\cEv\le\bEv$.
+%   %   if $\bEv\xrfx\aEv$ and $\labeling(\cEv) \rblocks \labeling(\aEv)$ then not $\bEv\le\cEv\xpox\aEv$.
+% \end{enumerate}
+
+% It follows from these that 
+% \begin{itemize}
+% \item if $\bEv\xfr\aEv$ then $\bEv\xgcb\aEv$, where $\bEv\xfr\aEv$ if
+%   $(\exists\cEv)$ $\bEv\xrfxinv\cEv\xco\aEv$.
+% \end{itemize}
+% And therefore
+% \begin{itemize}
+% \item if $\bEv\xeco\aEv$ then $\bEv\xgcb\aEv$.
+% \end{itemize}
+
+\begin{theorem}
+  \label{thm:ec}
+  Suppose $\aEG_1$ %$(\aEvs_1, \labeling_1, {\rpoloc_1}, {\rlob_1})$
+  is \EC-valid for $\aCmd$ via $({\rco_1}, {\rrfx_1}, {\rcb_1})$ and that
+  ${\rcb_1}\supseteq{\rfr_1}$.  Then there is a top-level pomset
+  $\aPS_2\in\frf{\semrr{\aCmd}}$ such that $\aEvs_2=\aEvs_1$,
+  $\labeling_2=\labeling_1$, ${\rrfx_2}={\rrfx_1}$, and ${\le_2}={\rcb_1}$.
+
+  \vspace{-.5\baselineskip}
+  \begin{proof}
+    We show that all the order required in the pomset is also required by
+    \armeight{}.  \ref{rf-block} holds since ${\rcb_1}$ is consistent with
+    ${\rco_1}$ and ${\rfr_1}$.  \ref{seq-le-rf-rf} follows from \ref{arm-rfi}.
+    As noted \ref{lob-le}, $\rlob$ includes the order required by
+    \ref{seq-kappa} and \ref{seq-le-delays-rf}.  
+  \end{proof}
+\end{theorem}
+
+The generality of \refthm{thm:ec} is not limited by the assumption that
+${\rcb_1}\supseteq{\rfr_1}$:
+\begin{lemma}
+  \label{lemma:fr}
+  Suppose $\aEG$ is \EC-valid via $({\rco}, {\rrfx}, {\rcb})$.  Then there a
+  permutation ${\rcbp}$ of ${\rcb}$ such that $\aEG$ is \EC-valid via
+  $({\rco}, {\rrfx}, {\rcbp})$ and ${\rcbp}\supseteq{\rfr}$, where ${\rfr}$
+  is defined in \ref{arm-local}.
+
+  \vspace{-.5\baselineskip}
+  \begin{proof}
+    We show that any ${\rcb}$ order that contradicts ${\rfr}$ is incidental.
+
+    By definition of $\rfr$, %$(\exists\bEv)$
+    $\aEv\xrfxinv\bEv\xco\cEv$, for some $\bEv$.
+    Since ${\rcb}\supseteq{\rco}$, we know that $\bEv\xco\cEv$.
+
+    If \ref{arm-rfe} applies to $\bEv\xrfx\aEv$, then $\aEv\xcb\cEv$, since
+    it cannot be that $\cEv\xco\bEv$.
+
+    Suppose \ref{arm-rfi} applies to $\bEv\xrfx\aEv$ and $\cEv$ is from a
+    different thread.  Because it is a different thread, we cannot have
+    $\aEv\xlob\cEv$, and thus the order in $\rcb$ is incidental.
+
+    Suppose \ref{arm-rfi} applies to $\bEv\xrfx\aEv$ and $\cEv$ is from the
+    same thread.  Since $\cEv\xco\bEv$, it cannot be that $\cEv\xpoloc\bEv$,
+    using \ref{arm-local}.  It also cannot be that
+    $\bEv\xpoloc\cEv\xpoloc\aEv$.  It must be that $\aEv\xpoloc\cEv$.  By
+    \ref{arm-lob-poloc}, we cannot have $\aEv\xlob\cEv$, and thus the order
+    in $\rcb$ is incidental.
+  \end{proof}
+\end{lemma}
+
+% Lemma: ${\rpoloc}\cup{\rpre}$ is acyclic.
+
+% Theorem: per-thread essential ${\le}$ $\subseteq$ ${\rpoloc}\cup{\rpre}$.
+
+Bad example:
 \begin{gather*}
+  \PEXCHG{x}{r}{2}\SEMI 
+  \PR{x}{s}\SEMI
+  \PW{y}{s{-}1} \PAR
+  \PR{y}{r}\SEMI
+  \PW{x}{r}
+  \\
+  \tag{\cmark\armeight}
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a}{\DR{x}{1}}{}
+      \event{b}{\DW{x}{2}}{right=of a}
+      \event{c}{\DR{x}{2}}{right=of b}
+      \event{d}{\DW{y}{1}}{right=of c}
+      \event{e}{\DR{y}{1}}{right=3em of d}
+      \event{f}{\DW{x}{1}}{right=of e}
+      \pre{a}{b}
+      \rf{b}{c}
+      \lob{c}{d}
+      \rf{d}{e}
+      \pre{e}{f}
+      \rf[out=-165,in=-15]{f}{a}
+    \end{tikzinline}}
+  % \hbox{\begin{tikzinline}[node distance=1.5em]
+  %   \event{a}{\DR{x}{1}}{}
+  %   \event{b}{\DW{x}{2}}{right=of a}
+  %   \event{c}{\DR{x}{2}}{right=of b}
+  %   \event{d}{\DW{y}{1}}{right=of c}
+  %   \event{e}{\DR{y}{1}}{right=3em of d}
+  %   \event{f}{\DW{x}{1}}{right=of e}
+  %   \rmw{a}{b}
+  %   \rfi{b}{c}
+  %   \dob{c}{d}
+  %   \rfe{d}{e}
+  %   \dob{e}{f}
+  %   \rfe[out=-165,in=-15]{f}{a}
+  % \end{tikzinline}}
+  \\
+  \tag{$\le$}
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a}{\DR{x}{1}}{}
+      \event{b}{\DW{x}{2}}{right=of a}
+      \event{c}{\DR{x}{2}}{right=of b}
+      \event{d}{\DW{y}{1}}{right=of c}
+      \event{e}{\DR{y}{1}}{right=3em of d}
+      \event{f}{\DW{x}{1}}{right=of e}
+      \rmw{a}{b}
+      \rf{b}{c}
+      % \po{c}{d}
+      \rf{d}{e}
+      \po{e}{f}
+      \rf[out=-165,in=-15]{f}{a}
+    \end{tikzinline}}
+\end{gather*}
+Anton example 1 \texttt{[rfi-coe-coe]}
+\begin{gather*}
+  \taglabel{rfi-coe-coe}
   \PW{x}{2}\SEMI 
-    \PR[\mRA]{x}{r}\SEMI
+  \PR[\mACQ]{x}{r}\SEMI
   \PW{y}{1} \PAR
   \PW{y}{2}\SEMI
-    \PW[\mRA]{x}{1}
+  \PW[\mREL]{x}{1}
+  \\
+  \tag{\cmark\armeight}
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a}{\DW{x}{2}}{}
+      \raevent{b}{\DR[\mACQ]{x}{2}}{right=of a}
+      \event{c}{\DW{y}{1}}{right=of b}
+      \event{d}{\DW{y}{2}}{right=2.5em of c}
+      \raevent{e}{\DW[\mREL]{x}{1}}{right=of d}
+      \rfi{a}{b}
+      \bob{b}{c}
+      \coe{c}{d}
+      \bob{d}{e}
+      \coe[out=-165,in=-15]{e}{a}
+    \end{tikzinline}}
   \\
@@
       \event{a}{\DW{x}{2}}{}
-        \raevent{b}{\DR[\mRA]{x}{2}}{right=of a}
+      \event{b}{\DR{x}{2}}{right=of a}
       \event{c}{\DW{y}{1}}{right=of b}
       \event{d}{\DW{y}{2}}{right=2.5em of c}
-        \raevent{e}{\DW[\mRA]{x}{1}}{right=of d}
+      \raevent{e}{\DW[\mREL]{x}{1}}{right=of d}
       \rf{a}{b}
-        \sync{b}{c}
       \wk{c}{d}
@@
 \end{gather*}
-  Under \EGC{}, this is explained by dropping the order
-  $\DRP[\mRA]{x}{2}\xsync\DWP{y}{1}$, because $\DRP[\mRA]{x}{2}$ is fulfilled
-  by a relaxed write in the same thread.
-  \begin{gather*}
-    \hbox{\begin{tikzinline}[node distance=1.5em]
-        \event{a}{\DW{x}{2}}{}
-        \raevent{b}{\DR[\mRA]{x}{2}}{right=of a}
-        \event{c}{\DW{y}{1}}{right=of b}
-        \event{d}{\DW{y}{2}}{right=2.5em of c}
-        \raevent{e}{\DW[\mRA]{x}{1}}{right=of d}
-        \rf{a}{b}
-        % \sync{b}{c}
-        \wk{c}{d}
-        \sync{d}{e}
-        \wk[out=-165,in=-15]{e}{a}
-      \end{tikzinline}}  
-  \end{gather*}
-
-  More generally, this can be understood as a compiler optimization that
-  downgrades a read from $\mRA$ to $\mRLX$ when it can be fulfilled by a
-  relaxed write in the same thread.
-\end{example}
-
-To model such \emph{downgraded reads}, we use the uninterpreted symbols
-$\Dx{\aLoc}$.
-
-\begin{definition}
-  \label{def:DS}
-  % Let $[\FALSE/\D]$ be the substitution that performs $[\FALSE/\Dx{\aLoc}]$
-  % for every $\aLoc$.
-  Let $[\aForm/\Dx{*}]$ substitute $\aForm$ for every $\Dx{\bLoc}$.\\
-  Let formula $\DL{\aLoc}{\amode}$ and substitution $\DS{\aLoc}{\amode}$ be defined:
-  \begin{align*}
-    \DL{\aLoc}{\mRLX}&=\TRUE
-    &\DS{\aLoc}{\mRLX}&=[\TRUE/\Dx{\aLoc}] 
-    \\
-    \DL{\aLoc}{\mRA}&=\Dx{\aLoc}%\land\D{}
-    &\DS{\aLoc}{\mRA}&=[\FALSE/\D]
-    \\
-    \DL{\aLoc}{\mSC}&=\Dx{\aLoc}%\land\D{}
-    &\DS{\aLoc}{\mSC}&=[\FALSE/\D]
-  \end{align*}
-\end{definition}
-\begin{definition}[\xCO/\xRASC/\xDGR]
-  \label{def:pomsets-down}
-  Update \refdef{def:pomsets-ra} to: % (\ref{L4} unchanged):
-  \begin{enumerate}
-  \item[\ref{S4})]
-    $\aTr{\bEvs}{\bForm}$ implies
-    \begin{math}
-      \bForm
-      \DS{\aLoc}{\amode}
-      [(\Qw{\aLoc}\land\aExp{=}\aVal)/\Qw{\aLoc}],
-    \end{math}
-  \item[\ref{S5})]
-    $\aTr{\cEvs}{\bForm}$ implies
-    \begin{math}
-      \bForm
-      \DS{\aLoc}{\amode}
-      [\FALSE/\QS{\aLoc}{\amode}],
-    \end{math}
-  \item[\ref{L4})]
-    $\aTr{\bEvs}{\bForm}$ implies
-    \begin{math}
-      \aVal{=}\aReg\limplies\bForm,
-    \end{math}
-  \item[\ref{L5})]
-    $\aTr{\cEvs}{\bForm}$ implies
-    \begin{math}
-      \DL{\aLoc}{\amode}\land \bForm[\FALSE/\QL{\aLoc}{\amode}].
-    \end{math}
-  \end{enumerate}
-\end{definition}
-Load actions that require downgrading introduce $\Dx{\aLoc}$.
-Relaxed stores on $\aLoc$ substitute true for $\Dx{\aLoc}$, whereas
-synchronizing stores substitute false for $\Dx{\aLoc}$.
-
-\begin{example}
-  Revisiting \refex{ex:dgr1} and eliding irrelevant transformers:
-  \begin{align*}
-    \begin{gathered}[t]
-      \PW{x}{2}
-      \\
-      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-          \event{a}{\DW{x}{2}}{}
-          \xform{x1d}{\bForm[\TRUE/\Dx{\aLoc}]}{below=of a}
-          % \xform{x1i}{\lnot\Qw{x}\land\bForm[\TRUE/\Dx{x}]}{below=of x1d}
-          \xo{a}{x1d}
-        \end{tikzinline}}  
-    \end{gathered}  
-    &&
-    \begin{gathered}[t]
-      \PR[\mRA]{x}{r}
-      \\
-      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-          \raevent{b}{\DR[\mRA]{x}{2}}{}
-          % \xform{x1d}{2{=}r\limplies\bForm}{below=of b}
-          \xform{x1i}{\Dx{x} \land\bForm[\FALSE/\Qr{x}]}{below=of b} %x1d}
-          % \xo{b}{x1d}
-        \end{tikzinline}}  
-    \end{gathered}  
-    &&
-    \begin{gathered}[t]
-      \PW{y}{1}
-      \\
-      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-          \event{c}{\DW{y}{1}}{}
-          % \xform{x1d}{\bForm[\TRUE/\Dx{\aLoc}]}{below=of c}
-          % \xform{x1i}{\lnot\Qw{y}\land\bForm[\TRUE/\Dx{y}]}{below=of x1d}
-          % \xo{c}{x1d}
-        \end{tikzinline}}  
-    \end{gathered}  
-  \end{align*}
-  Associating right:
-  \begin{align*}
-    \begin{gathered}[t]
-      \PW{x}{2}
-      \\
-      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-          \event{a}{\DW{x}{2}}{}
-          \xform{x1d}{\bForm[\TRUE/\Dx{\aLoc}]}{right=.7em of a}
-          % \xform{x1i}{\lnot\Qw{x}\land\bForm[\TRUE/\Dx{x}]}{below=of x1d}
-          \xos{a}{x1d}
-        \end{tikzinline}}  
-    \end{gathered}  
-    &&
-    \begin{gathered}[t]
-      \PR[\mRA]{x}{r}
-      \SEMI
-      \PW{y}{1}
-      \\
-      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-          \raevent{b}{\DR[\mRA]{x}{2}}{}
-          \event{c}{\Dx{x}\mid\DW{y}{1}}{right=.7em of b}
-        \end{tikzinline}}  
-    \end{gathered}  
-  \end{align*}
-  Composing, we have, as desired:
-  \begin{align*}
-    \begin{gathered}[t]
-      \PW{x}{2}
-      \SEMI
-      \PR[\mRA]{x}{r}
-      \SEMI
-      \PW{y}{1}
-      \\
-      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-          \event{a}{\DW{x}{2}}{}
-          \raevent{b}{\DR[\mRA]{x}{2}}{right=of a}
-          \event{c}{\DW{y}{1}}{right=of b}
-          \rf{a}{b}
-        \end{tikzinline}}  
-    \end{gathered}  
-  \end{align*}
-\end{example}
-
-% The following execution is allowed both by \armeight{} and \tso{}, but
-% disallowed by \refdef{def:pomsets-ra}.
-% \begin{gather*}
-%   \PW{x}{2}\SEMI 
-%   \PR[\mRA]{x}{r}\SEMI
-%   \PR{y}{s} \PAR
-%   \PW{y}{2}\SEMI
-%   \PW[\mRA]{x}{1}
-%   \\
-%   \hbox{\begin{tikzinline}[node distance=1.5em]
-%     \event{a}{\DW{x}{2}}{}
-%     \raevent{b}{\DR[\mRA]{x}{2}}{right=of a}
-%     \event{c}{\DR{y}{0}}{right=of b}
-%     \event{d}{\DW{y}{2}}{right=2.5em of c}
-%     \raevent{e}{\DW[\mRA]{x}{1}}{right=of d}
-%     \rf{a}{b}
-%     \sync{b}{c}
-%     \wk{c}{d}
-%     \sync{d}{e}
-%     \wk[out=-165,in=-15]{e}{a}
-%   \end{tikzinline}}
-% \end{gather*}
-
-\begin{example}
-  One might worry that our model is too permissive for $\mSC$ access, but
-  \armeight{} itself allows some very counterintuitive results for $\mSC$
-  access.  In the following execution we elide the initializing write
-  $\DWP{y}{0}$.
-  \begin{gather*}
-    \IF{\PR{x}{}}\THEN \PW{x}{2} \FI\SEMI
-    \PR[\mSC]{x}{r}\SEMI
-    \PR[\mSC]{y}{s} \PAR
-    \PW[\mSC]{y}{2}\SEMI
-    \PW[\mSC]{x}{1}
-    \\
-    \hbox{\begin{tikzinline}[node distance=1.5em]
-        \event{a}{\DR{x}{1}}{}
-        \event{b}{\DW{x}{2}}{right=of a}
-        \scevent{c}{\DR[\mSC]{x}{2}}{right=of b}
-        \scevent{d}{\DR[\mSC]{y}{0}}{right=of c}
-        \scevent{e}{\DW[\mSC]{y}{2}}{right=2.5em of d}
-        \scevent{f}{\DW[\mSC]{x}{1}}{right=of e}
-        \po{a}{b}
-        \rf{b}{c}
-        % \sync{c}{d}
-        \sync{e}{f}
-        \wk{d}{e}
-        \rf[out=-165,in=-15]{f}{a}
-      \end{tikzinline}}
-  \end{gather*}
-  Under \EGC{}, this is explained by dropping the order
-  $\DRP[\mSC]{x}{2}\xsync\DRP{y}{0}$, because $\DRP[\mSC]{x}{2}$ is fulfilled
-  by a relaxed write in the same thread.
-\end{example}
-
-\subsection{Removing Read-Read dependencies (\xRRD)}
-\label{sec:rrd}
-
-\begin{example}
-  \label{ex:rrd1}
-  The following execution is allowed by \armeight{}, but disallowed by
-  \refdef{def:pomsets-ra}.  
-  \begin{gather*}
-    \begin{gathered}
-      x\GETS0\SEMI x\GETS 1\SEMI y^\mRA\GETS1
-      \PAR
-      r\GETS y\SEMI \IF{r}\THEN s\GETS x\FI
-      \\[-.4ex]
-      \hbox{\begin{tikzinline}[node distance=1.5em]
-          \event{wx0}{\DW{x}{0}}{}
-          \event{wx1}{\DW{x}{1}}{right=of wx0}
-          \raevent{wy1}{\DW[\mRA]{y}{1}}{right=of wx1}
-          \event{ry1}{\DR{y}{1}}{right=2.5em of wy1}
-          \event{rx0}{\DR{x}{0}}{right=of ry1}
-          \sync{wx1}{wy1}
-          \po{ry1}{rx0}
-          \rf{wy1}{ry1}
-          \wk[in=-15,out=-165]{rx0}{wx1}
-          \wk{wx0}{wx1}
-        \end{tikzinline}}
-    \end{gathered}
-  \end{gather*}
-  Under \EGC{}, this is explained by dropping the order
-  $\DRP{y}{1}\xpo\DRP{x}{0}$, because \armeight{} does not include control
-  dependencies between reads in the locally-ordered-before relation.
-  \begin{gather*}
-    \hbox{\begin{tikzinline}[node distance=1.5em]
-        \event{wx0}{\DW{x}{0}}{}
-        \event{wx1}{\DW{x}{1}}{right=of wx0}
-        \raevent{wy1}{\DWRel{y}{1}}{right=of wx1}
-        \event{ry1}{\DR{y}{1}}{right=2.5em of wy1}
-        \event{rx0}{\DR{x}{0}}{right=of ry1}
-        \sync{wx1}{wy1}
-        % \po{ry1}{rx0}
-        \rf{wy1}{ry1}
-        \wk[in=-15,out=-165]{rx0}{wx1}
-        \wk{wx0}{wx1}
-      \end{tikzinline}}  
-  \end{gather*}
-\end{example}
-
-Since we do not distinguish control dependencies from other dependencies, we
-are forced to drop all dependencies between reads.  In order to do so, we use
-the uninterpreted symbol $\RW$.
-
-\begin{definition}[\xRRD]
-  \label{def:pomsets-rr}
-  Update \refdef{def:pomsets-trans} to:
-  % \begin{enumerate}
-  % \item[\ref{T3})]
-  %   $\labelingForm(\aEv)$ implies
-  %   $\labelingForm_1(\aEv)[\TRUE/\RW]$ if $\labelingAct_1(\aEv)$ is a write,
-  %   \\
-  %   $\labelingForm(\aEv)$ implies
-  %   $\labelingForm_1(\aEv)[\FALSE/\RW]$ otherwise.
-  % \end{enumerate}
-  \begin{enumerate}
-  \item[\ref{L4})]
-    $\aTr{\bEvs}{\bForm}$ implies $\aVal{=}\aReg\limplies\bForm$, 
-  \item[\ref{L5})]
-    % $\aTr{\cEvs}{\bForm}$ implies $\RW\limplies(\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm$,
-    $\aTr{\cEvs}{\bForm}$ implies $(\aVal{=}\aReg\lor\RW)\limplies\bForm$, when $\aEvs\neq\emptyset$,
-  \item[{\labeltext[L6]{L6)}{L6}}] 
-    $\aTr{\dEvs}{\bForm}$ implies $\bForm$, when $\aEvs=\emptyset$.
-  \end{enumerate}  
-\end{definition}
-\ref{L4} is the \emph{dependent} case,
-\ref{L5} the \emph{independent} case, and
-\ref{L6} the \emph{empty} case. 
- Recall the substitutions performed by $\ref{sTOP}$
-(\refdef{def:pomsets-top}).
-For writes, \ref{L5} is the same as \ref{L6} (since $\RW$
-is $\TRUE$).  For reads, \ref{L5} is the
-same as \ref{L4} (since $\RW$ is $\FALSE$).
-% For reads, the dependent and independent predicate transformers are identical
-% .  For writes, \ref{L5} requires
-% $\aTr{\cEvs}{\bForm}$ implies $\bForm$ for the independent case.
-
-One reading of \ref{L5} is that when satisfying a precondition $\aForm$ it is
-safe to use the value of the read, as long as the action is not a write.  If
-the pomset is empty, however, there is no value $\aVal$ to use.  Therefore
-the best we can do is to emulate skip, as in \ref{L6}.  % In order to
-% eventually arrive at a top-level pomset, this means that subsequent code must
-% be independent of $\aReg$.
-
-\begin{example}
-  Revisiting \refex{ex:rrd1} and eliding irrelevant transformers:
-  \begin{align*}
-    \begin{gathered}[t]
-      \PR{y}{r}
-      \\
-      \hbox{\begin{tikzinline}[node distance=0.5em and .5em]
-          \event{b}{\DR{y}{1}}{}
-          \xform{x1i}{(1{=}r\lor\RW)\limplies\bForm}{right=of b} %x1d}
-        \end{tikzinline}}  
-    \end{gathered}  
-    &&
-    \begin{gathered}[t]
-      \IF{r}\THEN s\GETS x\FI
-      \\
-      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-          \event{c}{r\mid\DR{x}{0}}{}
-          % \xform{x1d}{\bForm[\TRUE/\Dx{\aLoc}]}{below=of c}
-          % \xform{x1i}{\lnot\Qw{y}\land\bForm[\TRUE/\Dx{y}]}{below=of x1d}
-          % \xo{c}{x1d}
-        \end{tikzinline}}  
-    \end{gathered}  
-  \end{align*}
-  Composing sequentially:
-  \begin{align*}
-    \begin{gathered}[t]
-      \PR{y}{r}
-      \SEMI
-      \IF{r}\THEN s\GETS x\FI
-      \\
-      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-          \event{b}{\DR{y}{1}}{}
-          \event{c}{(1{=}r\lor\RW)\limplies r\mid\DR{x}{0}}{right=of b}
-          %\xform{x1i}{\RW\limplies\bForm}{right=of b} %x1d}
-        \end{tikzinline}}  
-    \end{gathered}  
-  \end{align*}
-  At top-level, $\ref{sTOP}$ yields:
-  \begin{align*}
-    \begin{gathered}[t]
-      % \PR{y}{r}
-      % \SEMI
-      % \IF{r}\THEN s\GETS x\FI
-      % \\
-      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-          \event{b}{\DR{y}{1}}{}
-          \event{c}{(1{=}r\lor\FALSE)\limplies r\mid\DR{x}{0}}{right=of b}
-          %\xform{x1i}{\RW\limplies\bForm}{right=of b} %x1d}
-        \end{tikzinline}}  
-    \end{gathered}  
-  \end{align*}
-  The precondition of $\DRP{x}{0}$ is a tautology, as required.
-\end{example}
-
-\subsection{Full semantics for ARM}
-
-\refdef{def:pomsets-arm} combines all of the features of
-\textsection\ref{sec:q}--\ref{sec:arm}.  
-% Only \ref{L5} and \ref{T3} are changed from \refdef{def:pomsets-down}.
-
-\begin{definition}[\xCO/\xRASC/\xDGR/\xRRD]
-  \label{def:pomsets-arm}
-  Let $[\aForm/\Q{}]$ be the substitution $[\aForm/\Qr{*}][\aForm/\Qw{*}][\aForm/\Qsc]$.  
-  Update \refdef{def:pomsets-trans} to: % (\ref{L4} unchanged):
-  \begin{enumerate}
-  \item[\ref{S3})]
-    $\labelingForm(\aEv)$ implies $\QS{\aLoc}{\amode}\land\aExp{=}\aVal$,
-  \item[\ref{L3})]
-    $\labelingForm(\aEv)$ implies $\QL{\aLoc}{\amode}$.
-  % \item[\ref{T3})]
-  %   $\labelingForm(\aEv)$ implies
-  %   $\labelingForm_1(\aEv)[\TRUE/\Q{}][\TRUE/\RW]$ if $\labelingAct_1(\aEv)$ is a write,
-  %   \\
-  %   $\labelingForm(\aEv)$ implies
-  %   $\labelingForm_1(\aEv)[\TRUE/\Q{}][\FALSE/\RW]$ otherwise.
-  \end{enumerate}
-  \begin{enumerate}
-  \item[\ref{S4})]
-    $\aTr{\bEvs}{\bForm}$ implies
-    \begin{math}
-      %(\Qw{\aLoc}\limplies\aExp{\neq}\aVal)\land
-      \bForm %\noSUB{[\aExp/\aLoc]},
-      [(\Qw{\aLoc}\land\aExp{=}\aVal)/\Qw{\aLoc}]
-      \DS{\aLoc}{\amode}
-    \end{math}
-  \item[\ref{S5})]
-    $\aTr{\cEvs}{\bForm}$ implies
-    \begin{math}
-      %\lnot\Qw{\aLoc}\land 
-      \bForm %\noSUB{[\aExp/\aLoc]}.
-      [\FALSE/\QS{\aLoc}{\amode}]
-      \DS{\aLoc}{\amode}
-    \end{math}
-  \item[\ref{L4})]
-    $\aTr{\bEvs}{\bForm}$ implies $\aVal{=}\aReg\limplies\bForm$, 
-  \item[\ref{L5})]
-    % $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode} \land(\RW\limplies(\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm)$,
-    % \item[\ref{L6})]
-    $\aTr{\cEvs}{\bForm}$ implies
-    \begin{math}
-      % \lnot\Qx{\aLoc} \land
-      \DL{\aLoc}{\amode}{} \land ((\aVal{=}\aReg\lor\RW)\limplies\bForm[\FALSE/\QL{\aLoc}{\amode}]).
-    \end{math}
-  \end{enumerate}  
-\end{definition}
-
-\begin{example}
-  \xRRD{} does not adversely affect subscription (\refex{ex:subscription}).
-  \begin{align*}
-    \begin{gathered}[t]
-      \PR[\mRA]{y}{r}
-      \\
-      \hbox{\begin{tikzinline}[node distance=0.5em and .5em]
-          \raevent{b}{\QL{y}{\mRA}\mid\DR[\mRA]{y}{1}}{}
-          \xform{x1i}{\Dx{y}\land(\RW\limplies\bForm[\FALSE/\QL{\aLoc}{\mRA}])}{right=of b} %x1d}
-        \end{tikzinline}}  
-    \end{gathered}  
-    &&
-    \begin{gathered}[t]
-      \PR{x}{s}
-      \\
-      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-          \event{c}{\QL{y}{\mRA}\mid\DW{x}{0}}{}
-        \end{tikzinline}}  
-    \end{gathered}  
-  \end{align*}
-  Since $\QL{y}{\mRA}[\FALSE/\QL{\aLoc}{\mRA}]$ is false, we have:
-  \begin{align*}
-    \begin{gathered}[t]
-      \PR[\mRA]{y}{r}
-      \SEMI
-      \PR{x}{s}
-      \\
-      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-          \raevent{b}{\QL{y}{\mRA}\mid\DR[\mRA]{y}{1}}{}
-          \event{c}{\Dx{y}\land(\RW\limplies\FALSE)\mid\DW{x}{0}}{right=of b}
-        \end{tikzinline}}  
-    \end{gathered}  
-  \end{align*}
-  Applying $\ref{sTOP}$ yields:
-  \begin{align*}
-    \begin{gathered}[t]
-      % \PR[\mRA]{y}{r}
-      % \SEMI
-      % \PR{x}{s}
-      % \\
-      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-          \raevent{b}{\TRUE\mid\DR[\mRA]{y}{1}}{}
-          \event{c}{\Dx{y}\land(\FALSE\limplies\FALSE)\mid\DW{x}{0}}{right=of b}
-        \end{tikzinline}}  
-    \end{gathered}  
-  \end{align*}
-  Although $(\FALSE\limplies\FALSE)$ is a tautology, $\Dx{y}$ is not.  Thus,
-  the pomset can only contribute to a top-level pomset if the thread is
-  preceded by a relaxed write to $y$, allowing $\DRP[\mRA]{y}{1}$ to be
-  downgraded to a relaxed read.
-\end{example}
-
-Every \armeight{} execution is allowed by \refdef{def:pomsets-arm}.  The
-proof of this fact is simplified by the recent characterization of
-\armeight{} in terms of \emph{external global completion} (\EGC{})
-\cite[\textsection B2.3.6]{arm-reference-manual}.  Under \EGC{}, an
-\armeight{} execution is a linearization of per-location program order and a
-subset of local-order.  Every such linearization is also a valid pomset under
-\refdef{def:pomsets-arm}.
-
-% \begin{definition}
-%   Let $[\aForm/\RW{}]$ be the substitution that replaces $\RW{}$ by $\aForm$.
-% \end{definition}
-
-% \subsection{Proof Strategy}
-
-\begin{comment}
-  Internal visibility:
-  If $\bEv\xpoloc\aEv$ then either
-  \begin{itemize}
-  \item %$\bEv\xca\aEv$:
-    $\bEv:\DWP{x}{1}\xco\aEv:\DWP{x}{2}$ 
-  \item %$\bEv\xrfx\aEv$:
-    $\bEv:\DWP{x}{1}\xrfx\aEv:\DRP{x}{1}$
-  \item %$\bEv\mathrel{\xco\xrfx}\aEv$:
-    $\bEv:\DWP{x}{1}\xco\DWP{x}{2}\xrfx\aEv:\DRP{x}{2}$
-  \item
-    $\bEv:\DRP{x}{0}\xfr\aEv:\DWP{x}{1}$\\
-    $\bEv:\DRP{x}{0}\xrfxinv\DWP{x}{0}\xco\aEv:\DWP{x}{1}$
-  \item %$\bEv\mathrel{\xfr\xrfx}\aEv$:
-    $\bEv:\DRP{x}{0}\xfr\DWP{x}{1}\xrfx\aEv:\DRP{x}{1}$\\
-    $\bEv:\DRP{x}{0}\xrfxinv\DWP{x}{0}\xco\DWP{x}{1}\xrfx\aEv:\DRP{x}{1}$
-  \item %$\bEv\mathrel{\xrfxinv\xrfx}\aEv$:
-    $\bEv:\DRP{x}{1}\xrfxinv\DWP{x}{1}\xrfx\aEv:\DRP{x}{1}$
-    % \begin{tikzinline}
-    %   \node(a){$\DWP{x}{1}$};
-    %   \node(b)[right=1.5em of a]{$\DRP{x}{1}$};
-    %   \node(c)[right=1.5em of b]{$\DRP{x}{1}$};
-    %   \rfx{a}{b}
-    %   \rfx[out=15,in=165]{a}{c}
-    % \end{tikzinline}
-  \end{itemize}
-  If $\bEv\xpoloc\aEv$ then the following are impossible:
-  \begin{itemize}
-  \item $\bEv:\DWP{x}{2}\xcoinv\aEv:\DWP{x}{1}$ 
-  \item $\bEv:\DWP{x}{1}\xfrinv\aEv:\DRP{x}{0}$ 
-  \item $\bEv:\DRP{x}{1}\xrfxinv\aEv:\DWP{x}{1}$ 
-  \item $\bEv:\DRP{x}{2}\xrfxinv\DWP{x}{2}\xcoinv\aEv:\DWP{x}{1}$
-  \item $\bEv:\DRP{x}{1}\xrfxinv\DWP{x}{1}\xfrinv\aEv:\DRP{x}{0}$
-  \end{itemize}
-\end{comment}
-
-\endinput
-
-\begin{figure*}
-\begin{verbatim}
-let IM0 = loc & ((IW * (M\IW)) | ((W\FW) * FW))
-
-(* Local read successor *)
-let lrs = [W]; po-loc \ intervening-write(po-loc); [R]
-
-(* Local write successor *)
-let lws = po-loc; [W]
-
-(* Dependency-ordered-before *)
-let dob =  addr | data               
-        | (addr | data); lrs
-        | addr; po; [W]         |  addr; po; [ISB]; po; [R]
-        | ctrl;     [W]         |  ctrl;     [ISB]; po; [R]
-our dob = addr; [W]
-        | data; [W]
-        | ctrl; [W]
-
-(* Barrier-ordered-before *)
-let bob = [A | Q]; po            | po; [dmb.full]; po          | po; ([A];amo;[L]); po
-        | po; [L]                | [R]; po; [dmb.ld]; po
-        | [L]; po; [A]           | [W]; po; [dmb.st]; po; [W]
-
-(* Locally-ordered-before *)                let aob = rmw 
-let rec lob = lws | dob | bob | lob; lob            | [W & range(rmw)]; lrs; [A | Q]
-
-(* Internal visibility requirement *)
-acyclic po-loc | fr | co | rf as internal
-
-(* External global completion requirement *)
-let gc-req = (W * _) | (R * _) & ((range(rfe) * _) | (rfi^-1; lob))
-let preorder-gcb = IM0 | lob & gc-req
-
-with gcb from linearisations(M, preorder-gcb)
-~empty gcb
-
-let gcbl = gcb & loc
-let rf-gcb = (W * R) & gcbl \ intervening-write(gcbl)
-let co-gcb = (W * W) & gcbl
-
-call equal(rf, rf-gcb)
-call equal(co, co-gcb)
-
-James:
-let po-conflict = po-loc \ (R * R)
-Claim we can add po-conflict to preorder-gcb without changing things
-let preorder-gcb = IM0 | lob & gc-req | po-conflict
-\end{verbatim}
-  \caption{ARM}
-  \label{fig:arm}
-\end{figure*}
-
-\textcolor{red}{[This is out of date, from an old submission]}
-
-In this section, we consider the fragment of our language without
-restriction.  For simplicity, we allow release and acquire synchronization
-but ban fences.  We assume that all memory locations are initialized to $0$
-and parallel-composition occurs only at top level.  We take the set of memory
-locations to be finite.  In other words, we assume that programs have the
-form
-\begin{displaymath}
-  {\aLoc_1}\GETS{0}\SEMI
-  \cdots\SEMI
-  {\aLoc_m}\GETS{0}\SEMI
-  (\aCmd^1 \PAR \cdots \PAR \aCmd^n)
-\end{displaymath}
-where $\aCmd^1$, \ldots, $\aCmd^n$ do not include composition, restriction or
-fence operations.
-
-Our language can be translated to ARM using \texttt{ldr} for relaxed read,
-\texttt{ldar} for acquiring read, \texttt{str} for relaxed write, and
-\texttt{stlr} for releasing write.  Relative to the ARM specification, we
-have removed loops and read-modify-write (RMW) operations, in addition to
-fences\footnote{We leave out fences for simplicity.  Following
-  \citet{DBLP:journals/pacmpl/PodkopaevLV19}, our $\FENCE$ instruction can be
-  translated to \texttt{dmb.sy}, since it has release-acquire semantics.
-  Acquire fences map to \texttt{dmb.ld}, and release fences to
-  \texttt{dmb.sy} --- \texttt{dmb.st} does not provide order to prior
-  reads.}.
-
-We show that any ARM-consistent execution graph for this sublanguage can be
-considered an execution of our semantics.  
-% 
-% Syntactically, we drop the superscript \textsf{rlx} on relaxed reads and
-% writes; in addition, we use structured conditionals rather than the more
-% general \textsf{goto}.  We refer to this sublanguage as $\muIMM$.
-% (Because the source language lacks RMW operations, the ``is
-% exclusive'' flag on every read will be \textsf{not-ex} and the RMW mode on
-% every write will be \textsf{normal}.)
-% 
-Due to space limitations, we do not include a full description ARM
-consistency in the main text .
-Here we provide a birds eye view of the details, drawing on the intuitions gleaned from~\citep{DBLP:journals/pacmpl/PulteFDFSS18}.  
-Interested readers should see \textsection\ref{sec:arm:proof}
-for further details.
-
-An ARM execution graph $G$ defines many relations, including program order
-($\rpox$), reads-from ($\rrfx$), coherence ($\rco$) and several dependency
-orders.  From these are derived:
-\begin{itemize}
-\item ${\rpoloc}$, which is the subrelation of $\rpox$ that only relates
-  actions on the same location,
-\item ${\rob}$, which is required to be acyclic (\ref{external}), and
-\item $\reco$, with the requirement that ${\rpoloc}\cup{\reco}$ be acyclic (\ref{sc-per-loc}).
-\end{itemize}
-% Let $G$ be an execution graph satisfying the ARM consistency 
-% requirements.
-Given an execution graph $G$, we say that $\aEv$ is an \emph{internal read} if $\aEv\in\fcodom({\rpox}\cap {\rrfx})$.
-
-The ${\rob}$ order is an acyclic global order on events, agreed upon by all threads, reflecting the progress of time in an \armeight\ execution.
-The cross thread component of the ordering is induced by the ordering on conflicting actions on the same location from different threads.
-The intra thread component of the ordering is induced by barrier ordering and data ordering.
-Notably, these dependencies  are determined syntactically.
-In particular, $\rob$
-may not necessarily include the intra thread component of $\rpoloc$ ordering.
-
-This motivates the translation of an \armeight\ execution into our setting.
-In our setting, the progress of time is given by $\lt$.
-We accommodate intra-thread reordering by internal read actions, thus excusing us from the obligation of placing them on the global $\lt$-timeline.
-
-Formally, from $G$ we construct a candidate pomset $\aPS$ as follows:
-\begin{itemize}
-\item $\Event= \textsf{E}$,
-\item $\labelingAct(\aEv)=\tau \mathsf{lab}(e)$, if $\aEv$ is a relaxed
-  internal read, 
-\item $\labelingAct(\aEv)=\mathsf{lab}(e)$, if $\aEv$ is not a relaxed
-  internal read,
-\item $\labelingForm(\aEv)=\TRUE$,
-  % \item ${\le} = {\rob}^?$, where $?$ denotes reflexive closure, and
-\item ${\gtN} = ({\rob} \cup {\reco})^*$, where $*$ denotes reflexive and transitive closure.
-\end{itemize}
-
-\begin{theorem}
-  If $G$ is ARM consistent, the constructed candidate satisfies the
-  requirements for a top-level memory model pomset.
-\end{theorem}
-Any $\lt$-ordering imposed in our model
-is enforced by \armeight, since our notion of semantic dependency is more
-permissive than \armeight 's syntactic dependency.
-So, the heart of the proof is showing the acyclicity of $({\rob} \cup {\reco})^*$ for the events under consideration.
-Since the cross thread portion of $\reco$ ($\rcoe,\rfre,\rrfe$) is included in $\rob$, this result is really about the influence of $\reco \cap \rpox$.
-Our translation of \armeight 's \rrfi\  as silent internal actions removes them from order considerations.
-Consequently, we only have to consider the suborder of ${\rob}$ derived without ever using  $\rrfi$ for the following key property demonstrated in \textsection\ref{sec:arm:proof}.
-\begin{lemma}\label{extendob}
-  Let $\aEv, \bEv$ be distinct events and $\bEv'\ (\xob\cap \xpox) \ \bEv\ ((\xeco\cap \xpox) \setminus \xrfi) \  \aEv\ (\xob \cap \xpox)  \ \aEv'$.  Then $\bEv' \xob \aEv'$.
-\end{lemma}
-
-\begin{remark}[Proof for TSO]
-  The proof for compilation into \tso\ is very similar.  In particular the facts listed above hold for \tso, where $\rob$ is replaced by (the
-  transitive closure of) the propagation relation defined for \tso\ 
-  \citep{alglave}.
-\end{remark}
-
-\section{Proof of compilation for ARMv8}
-\label{sec:arm:proof}
-\textcolor{red}{[This is out of date, from an old submission]}
-
-In this section, we develop the proof of correctness of compilation to \armeight.  In order to ease readability, we reproduce the definitions from the main text.
-
-Given a relation $R$, $R^?$ denotes reflexive closure, $R^+$ denotes
-transitive closure and $R^*$ denotes reflexive and transitive closure.  Given relations $R$ and $S$, $R;S$ denotes composition.
-
-The ARMv8 model is described using the following relations.
-\begin{itemize}
-\item $\IDR$, $\IDW$, $\IDAcq$, $\IDRel$: identity on reads, writes, acquires
-  and releases.
-  % \item $\IDR$ identity on reads
-  % \item $\IDW$: identity on writes
-  % \item $\IDAcq$: identity on acquires
-  % \item $\IDRel$: identity on releases
-\item $\IDLoc$: relates any two events that touch the same location.
-\item $\rpox$: program order.
-\item $\rdata$, $\rctrl$, $\raddr$: data, control and address dependencies.
-\item $\rrfx$: reads-from. $\rrfx^{-1}$ relates each read to a matching write
-  on the same location.
-\item $\rco$: coherence, which is a total order on the writes to a single
-  location.
-\item ${\rfr}\eqdef{\rco};\rrfx^{-1}$: from-read, which relates reads to
-  subsequent writes.
-\end{itemize}
-For any relation, the cross-thread subrelation is denoted by appending $e$;
-the intra-thread subrelation is denoted by appending $i$.  For example,
-${\rrfe}\eqdef{\rrfx}\setminus{\rpox}$ and ${\rrfi}\eqdef{\rrfx}\cap{\rpox}$.
-The subrelation restriction attention to actions on the same location is
-given by appending $\mathsf{loc}$.  For example, ${\rpoloc}\eqdef{\rpox}\cap{\IDLoc}$.
-
-The ARMv8 model defines the following relations.
-In our presentation, we have elided rules concerning fences and RMW operations.
-\begin{align*}
-  \tag{Extended coherence}
-  {\reco} &\eqdef {\rrf} \cup {\rfr} \cup {\rco}
-  \\
-  \tag{Observed externally}
-  {\robs} &\eqdef \smash{
-    {\rrfe} \cup {\rfre} \cup {\rcoe}
-  }
-  \\
-  \tag{Dependency order}
-  {\rdob} &\eqdef\smash{
-    ({\raddr}\cup{\rdata}); {\rrfi}^?
-    \cup ({\rctrl}\cup{\rdata}); {\IDW}; {\rcoi}^?
-    \cup {\raddr}; {\rpox}; {\IDW}
-  }
-  \\
-  \tag{Barrier order}
-  {\rbob} &\eqdef\smash{
-    {\IDAcq}; {\rpox}
-    \cup {\rpox};{\IDRel}; {\rcoi}^?
-  }
-  \\
-  \tag{Acyclic order}
-  {\rob} &\eqdef\smash{
-    ({\robs} \cup {\rdob} \cup {\rbob})^+
-  }
-\end{align*}
-\begin{definition}
-  An RMW-free and fence-free execution is \emph{ARM-consistent} if
-  \begin{align*}&
-    \tag{\textsc{$\rrfx$-completeness}}\label{rf-comp}
-    \fcodom(\rrfx)=\fdom(\rreads)
-    \\[-1ex]&
-    \tag{\textsc{$\rco$-totality}}\label{co-tot}
-    \text{For every location $\aLoc$, $\rco$ totally orders the writes of $\aLoc$}  
-    \\[-1ex]&
-    \tag{\textsc{sc-per-loc}}\label{sc-per-loc}
-    {\rpoloc} \cup {\rrfx} \cup {\rfr} \cup {\rco}\;\text{is acyclic}
-    \\[-1ex]&
-    \tag{\textsc{external}}\label{external}
-    {\rob}\;\text{is acyclic}
-  \end{align*}
-\end{definition}
-
-% Use these to refer to the rules in text:
-% \ref{rf-comp} 
-% \ref{co-tot}
-% \ref{sc-per-loc}
-% \ref{external}
-
-Given an execution graph $G$, we say that $\aEv$ is an \emph{internal read} if
-$\aEv\in\fcodom(\mathsf{po}\cap \mathsf{rf})$.    We are going to translate internal reads of execution graphs into internal reads of the semantics.
-
-From $G$ we construct a candidate pomset $\aPS$ as follows:
-\begin{itemize}
-\item $\Event= \textsf{E}$,
-\item $\labelingAct(\aEv)=\tau \mathsf{lab}(e)$, if $\aEv$ is a relaxed
-  internal read, 
-\item $\labelingAct(\aEv)=\mathsf{lab}(e)$, if $\aEv$ is not a relaxed
-  internal read,
-\item $\labelingForm(\aEv)=\TRUE$,
-  % \item ${\le} = {\rob}$, and
-\item ${\gtN} = ({\rob} \cup {\reco})^*$
-\end{itemize}
-To reempphasize, in this candidate pomset, $\rob$ is calculated by considering the definition of $\rob$ without $\rrfi$, ie.:
-\begin{align*}
-  \tag{Dependency order}
-  {\rdob} &\eqdef\smash{
-    ({\raddr}\cup{\rdata});
-    \cup ({\rctrl}\cup{\rdata}); {\IDW}; {\rcoi}^?
-    \cup {\raddr}; {\rpox}; {\IDW}
-  }
-  \\
-  \tag{Barrier order}
-  {\rbob} &\eqdef\smash{
-    {\IDAcq}; {\rpox}
-    \cup {\rpox};{\IDRel}; {\rcoi}^?
-  }
-  \\
-  \tag{Acyclic order}
-  {\rob} &\eqdef\smash{
-    ({\robs} \cup {\rdob} \cup {\rbob})^+
-  }
-\end{align*}
-
-We show that $\aPS$ is a top-level pomset, reasoning as follows.
-% We establish the criteria for a top-level memory-model pomset:
-\begin{itemize}
-\item ${\le}$ is a partial order.  This holds since $G.{\rar}$ is acyclic.
-  % \item If $\bEv \le \aEv$ then $\bEv \gtN \aEv$.  By construction.
-  % \item If $\bEv \le \aEv$ and $\aEv \gtN \bEv$ then $\bEv = \aEv$.  Proved below.
-  % \item If $\cEv \le \bEv \gtN \aEv$ or $\cEv \gtN \bEv \le \aEv$ then
-  %   $\cEv \gtN \aEv$. By construction.
-  % \end{itemize}
-
-  % Next, we establish the criteria for a 3-valued pomset with preconditions (Definition~\ref{def:3pre}).
-  % \begin{itemize}
-\item $\labelingForm(\aEv)$ implies $\labelingForm(\bEv)$ whenever
-  $\bEv\le\aEv$.   Trivial, since every formula is $\TRUE$.
-  % \item $\aPS$ is $\aLoc$-coherent; that is, when restricted to events that
-  %   read or write $\aLoc$, $\gtN$ forms a partial order.
-  % \end{itemize}
-
-  % Finally, we establish the criteria for a top-level pomset
-  % (Definition~\ref{def:x-closed}).
-  % \begin{itemize}
-\item $\aEv$ is location independent. Trivial, since every formula is $\TRUE$.
-\item If $\aEv$ reads $\aVal$ from $\aLoc$, then there is some $\bEv$ such that
-  \begin{itemize}
-  \item $\bEv \lt \aEv$,  
-  \item $\bEv$ writes $\aVal$ to $\aLoc$, and
-  \item if $\cEv$ writes to $\aLoc$
-    then either $\cEv \gtN \bEv$ or $\aEv \gtN \cEv$.
-  \end{itemize}    
-\end{itemize}
-
-\subsection{Proof that  $({\rob} \cup {\reco})^*$ is irreflexive. }
-
-\paragraph*{Proof of lemma~\ref{extendob}. }
-
-Let $\aEv, \bEv$ be distinct events and $\bEv'\ (\xob\cap \xpox) \ \bEv\ ((\xeco\cap \xpox) \setminus \xrfi) \  \aEv\ (\xob \cap \xpox)  \ \aEv'$.  Then $\bEv' \xob \aEv'$.
-
-\begin{proof}
-  If $\bEv'$ is an acquire,  or $\aEv$ is an release, or $\aEv'$ is a release, result is immediate.
-
-  We next consider the case where $\aEv$ is a read.
-In this case,  $\bEv$ is a write.
-Since $\bEv\ ((\xeco\cap \xpox) \setminus \xrfi) \  \aEv$, there is a write $\bEv_1$ such that $ \bEv \xcoe\ \bEv_1 \ \xrfe\ \aEv' $.
-So, $\bEv \xob \aEv$ and result follows in this case.
-
-  So, it suffices to prove the following assuming that $\bEv'$ is not an acquire and $\aEv'$ is not a release and $\aEv$ is not a release or a read and $\aEv, \bEv$ are distinct.
-  \begin{itemize}
-  \item If $\bEv'\ (\xob\cap \xpox)  \ \bEv(\xeco\cap\xpox)\aEv$ then $\bEv'\xob\aEv$.
-  \item If $\bEv\ (\xeco\cap\xpox) \ \aEv(\xob\cap\xpox)\aEv'$ then $\bEv\xob\aEv'$.
-  \end{itemize}
-
-  We first prove that if $\bEv'\ (\xob \cap \xpox) \ \bEv\ (\xeco \cap \xpox) \ \aEv$ then $\bEv'\xob\aEv$.   Proof proceeds by cases on the witness for $\bEv'\ (\xob\cap \xpox) \ \bEv$. 
-  \begin{itemize}
-  \item  If $\bEv' \xbob  \bEv$, then: 
-    \[ \bEv'\ (\smash{
-        {\IDAcq}; {\rpox}
-        \cup {\rpox};{\IDRel}; {\rcoi}^?) \ 
-      }
-      \bEv
-    \]
-    Since $\bEv'$ is not an acquire, $\bEv' ({\rpox};{\IDRel}; {\rcoi}^?) \bEv$, so $\bEv$ is a write.  Since $\aEv$ is not a read,  $\bEv \xcoi\ \aEv$. Thus, result follows.
-
-  \item If $\bEv' \xdob  \bEv$, then: 
-    \[ \bEv'\ 
-      \smash{
-        ( ({\rctrl}\cup{\rdata}); {\IDW}; {\rcoi}^?
-        \cup {\raddr}; {\rpox}; {\IDW}
-      } \
-      \bEv
-    \]
-    So, $\bEv$ is a write.  Since $\aEv$ is also a write, we deduce that 
-    \[ \bEv'\ 
-      \smash{
-        ( ({\rctrl}\cup{\rdata}); {\IDW}; {\rcoi}^?
-        \cup {\raddr}; {\rpox}; {\IDW}
-      } \
-      \aEv
-    \]
-  \end{itemize}
-
-  We next prove  that if $\bEv\ (\xeco \cap \xpox) \ \aEv\ (\xob\cap \xpox) \ \aEv'$ then $\bEv\xob\aEv'$, under the assumptions that  $\aEv'$ is not a release and $\aEv$ is not a release or a read and $\aEv, \bEv$ are distinct.
-
-  Proof proceeds by cases on the witness for $\aEv (\xob\cap \xpox) \aEv'$.
-
-  \begin{itemize}
-  \item  If $\aEv \xbob  \aEv'$, then: 
-    \[ \aEv\ (\smash{
-        {\IDAcq}; {\rpox}
-        \cup {\rpox};{\IDRel}; {\rcoi}^?) \ 
-      }
-      \aEv'
-    \]
-    Since $\aEv$ is not a read, $\aEv ({\rpox};{\IDRel}; {\rcoi}^?) \aEv'$.  Result follows since  $\bEv \xpox\ \aEv$.
-
-  \item If $\aEv \xdob  \aEv'$, then $\aEv$ is a read.  \qedhere
-
-  \end{itemize}
-\end{proof}
-
-\begin{lemma}\label{obeco1}
-  If $\bEv\xob\aEv$ then $\lnot(\aEv\xeco\bEv)$.
-
-  \begin{proof}
-    Proof by contradiction.  Let 
-    \[ \aEv \xob \aEv' \xeco \bEv' \xob \bEv \xob \cEv \xob \cEv' \xob \aEv \]
-    where $\aEv' \xpox \bEv'$.
-
-    By lemma~\ref{extendob}, if $\aEv \not=\aEv'$, we deduce $\aEv \xob \bEv'$, and thus $\aEv \xob \bEv$.  If $\bEv \not=\bEv'$, we deduce $\aEv' \xob \bEv$ and thus $\aEv \xob \bEv$.
-
-    Thus, if $\aEv \not=\aEv'$ or $\bEv \not=\bEv'$, then there is a cycle $\aEv \xob \bEv \xob \cEv \xob \cEv' \xob \aEv$.
-
-    So we can assume that  $\aEv' = \aEv$, $\bEv' = \bEv$ and 
-    \[ \aEv  \xeco \bEv \xob \cEv \xob \cEv' \xeco \aEv \]
-    where all of $\aEv, \bEv, \cEv, \cEv'$ access the same location and at least one of $\aEv,\bEv$ is a write, at least one of $\aEv,\cEv'$ is a write, and at least one of $\bEv,\cEv$ is a write.
-
-    We reason by cases.
-    \begin{itemize}
-    \item If $\cEv'$ is a write or both $(\aEv, \bEv)$ are writes.
-
-      We deduce that $\bEv \xeco \cEv' \xeco \aEv$ and thus $\bEv \xeco \aEv$.
-
-    \item $\cEv'$ is a read.  $\aEv$ is a write.  $\bEv$ is a read.
-
-      In this case $\cEv$ is a write.  From $\cEv \xob \aEv$, we deduce $\cEv \xeco \aEv$. Combining with $\bEv \xeco \cEv$, we deduce that $\bEv \xeco \aEv$.
-
-    \end{itemize}
-    In either case, there is a contradiction $\aEv \xeco \bEv \xeco \aEv$.
-  \end{proof}
-\end{lemma}
-
-\begin{lemma}\label{obeco2}
-  $({\rob} \cup {\reco})^*$ is irrreflexive.
-\end{lemma}
-\begin{proof}
-  The simple case that $\rob; \reco$ is irreflexive is proved above.  The full proof by contradiction.
-
-  Let $n \geq 1$ be the minimum such that:
-  \begin{align*} 
-    &\aEv^0_0 \xob \aEv^0_1 \xeco \bEv^0_0 \xob \bEv^0_1  \\
-    (\xeco \cap \xob) &  \   \aEv^1_0 \xob \aEv^1_1 \xeco \bEv^1_0 \xob \bEv^1_1 \\
-    (\xeco \cap \xob) & \ \ldots \\
-    & \ldots \bEv^n_1 \\
-    (\xeco \cap \xob) & \  \aEv^0_0
-  \end{align*}
-  where  for all $i$, we have:
-  \[ \aEv^i_0 \xpox \aEv^i_1 (\xeco \cap \xpox) \bEv^i_0 \xpox \bEv^i_1\] and 
-  \[ \neg (\bEv^i_1 \xpox (\aEv^{(i+1) \mod n}_0 \]
-
-  For any $i$, if $\aEv^i_0 \not= \aEv^i_1$ or $\bEv^i_0 \xpox \bEv^i_1$, via lemma~\ref{extendob}, we deduce that $\aEv^i_0  \xob \bEv^i_1$, contradicting minimality of $n$.
-
-  So, we can assume that $n \geq 1$ is such that:
-  \begin{align*} 
-    &\ \aEv^0 \xeco  \bEv^0 \\
-    (\xeco \cap \xob) &  \   \aEv^1  \xeco \bEv^1 \\
-    (\xeco \cap \xob) & \ \ldots \\
-    & \ldots \bEv^n \\
-    (\xeco \cap \xob) & \ \aEv^0
-  \end{align*}
-  which is a contradiction since it is a cycle in $\xeco$ and since at least one of $\aEv^i ,\bEv^i$ is a write for all $i$. 
-\end{proof}

### 2021-04-14 paper/intro.tex
update macros to distinguish REL and ACQ
intro

@@
 
-This paper is about the interaction of two of the fundamental building
-blocks of computing: memory and sequential composition. One would like
-to think that these are well-worn topics, where every issue has been
-settled, but this is sadly not the case.
+This paper is about the interaction of two of the fundamental building blocks
+of computing: sequential composition and mutable state. One would like to
+think that these are well-worn topics, where every issue has been settled,
+but this is not the case.
 
-\subsection{Memory}
+\subsection{Sequential Composition} %Predicate Transformers}
 
-For single-threaded programs, memory can be thought of as you might
-expect: programs write to, and read from, memory references.
-This can be thought of as a total order of reads and writes,
-where each read has a matching \emph{fulfilling} write,
-for example:
-  \begin{gather*}
-    \THREAD{x\GETS0\SEMI x\GETS1\SEMI y\GETS2\SEMI
-    r\GETS y\SEMI s\GETS x}
-    \\[-.4ex]
-    \nonumber
-    \hbox{\begin{tikzinline}[node distance=1.5em]
-        \event{wx0}{\DW{x}{0}}{}
-        \event{wx1}{\DW{x}{1}}{right=of wx0}
-        \event{wy2}{\DW{y}{2}}{right=of wx1}
-        \event{ry2}{\DR{y}{2}}{right=of wy2}
-        \event{rx1}{\DR{x}{1}}{right=of ry2}
-        \rf[out=20,in=160]{wy2}{ry2}
-        \rf[out=20,in=160]{wx1}{rx1}
-        \po{wx0}{wx1}
-        \po{wx1}{wy2}
-        \po{wy2}{ry2}
-        \po{ry2}{rx1}
-      \end{tikzinline}}
-  \end{gather*}
-(In examples, $\aReg$--$\bReg$ range over thread-local registers and $\aLoc$-$\cLoc$
-range over shared memory references.)
-
-This model naturally extends to the case of shared-memory concurrency, leading to a \emph{sequentially consistent}
-semantics \cite{Lamport:1979:MMC:1311099.1311750}, in which \emph{program order} inside a thread implies
-a total \emph{causal order} between read and write events, for example:
-  \begin{gather*}
-    \THREAD{x\GETS0\SEMI x\GETS1\SEMI y\GETS2}
-    \PAR
-    \THREAD{r\GETS y\SEMI s\GETS x}
-    \\[-.4ex]
-    \nonumber
-    \hbox{\begin{tikzinline}[node distance=1.5em]
-        \event{wx0}{\DW{x}{0}}{}
-        \event{wx1}{\DW{x}{1}}{right=of wx0}
-        \event{wy2}{\DW{y}{2}}{right=of wx1}
-        \event{ry2}{\DR{y}{2}}{right=of wy2}
-        \event{rx1}{\DR{x}{1}}{right=of ry2}
-        \rf[out=20,in=160]{wy2}{ry2}
-        \rf[out=20,in=160]{wx1}{rx1}
-        \po{wx0}{wx1}
-        \po{wx1}{wy2}
-        \po{ry2}{rx1}
-      \end{tikzinline}}
-  \end{gather*}
-
-Unfortunately, this model does not compile efficiently to commodity
-hardware, resulting in a 37--73\% increase in CPU time on ARM~\cite{Liu:2019:ASC:3314221.3314611} and,
-hence, in power consumption.  Developers of software and compilers have
-therefore been faced with a difficult trade-off, between an elegant
-model of memory, and its impact on resource usage (such as size of
-data centers, electricity bills and carbon footprint). Unsurprisingly,
-many have chosen to prioritize efficiency over elegance.
-
-This has led to \emph{relaxed memory models}, in which the requirement of
-sequential consistency is weakened to only apply \emph{per-location} and not globally
-over the whole program. This allows executions which
-are inconsistent with program order, such as:
-  \begin{gather*}
-    \THREAD{x\GETS0\SEMI x\GETS1\SEMI y\GETS2}
-    \PAR
-    \THREAD{r\GETS y\SEMI s\GETS x}
-    \\[-.4ex]
-    \nonumber
-    \hbox{\begin{tikzinline}[node distance=1.5em]
-        \event{wx0}{\DW{x}{0}}{}
-        \event{wx1}{\DW{x}{1}}{right=of wx0}
-        \event{wy2}{\DW{y}{2}}{right=of wx1}
-        \event{ry2}{\DR{y}{2}}{right=of wy2}
-        \event{rx0}{\DR{x}{0}}{right=of ry2}
-        \rf[out=20,in=160]{wy2}{ry2}
-        \rf[out=15,in=165]{wx0}{rx0}
-        \po{wx0}{wx1}
-        \po[out=200,in=340]{rx0}{wx1}
-      \end{tikzinline}}
-  \end{gather*}
-
-In such models, the causal order between events is important,
-and includes control and data dependencies, to avoid
-paradoxical ``out of thin air'' examples such as:
-  \begin{gather*}
-    \THREAD{r\GETS x\SEMI \IF r \THEN y\GETS1 \FI}
-    \PAR
-    \THREAD{s\GETS y\SEMI x\GETS s}
-    \\[-.4ex]
-    \nonumber
-    \hbox{\begin{tikzinline}[node distance=1.5em]
-        \event{rx1}{\DR{x}{1}}{}
-        \event{wy1}{\DW{y}{1}}{right=of rx1}
-        \event{ry1}{\DR{y}{1}}{right=of wy1}
-        \event{wx1}{\DW{x}{1}}{right=of ry1}
-        \rf[out=20,in=160]{wy1}{ry1}
-        \rf[out=160,in=20]{wx1}{rx1}
-        \po{rx1}{wy1}
-        \po{ry1}{wx1}
-      \end{tikzinline}}
-  \end{gather*}
-This candidate execution forms a cycle in causal order, so is disallowed,
-but this depends crucially on the control dependency
-from $(\DR{x}{1})$ to $(\DW{y}{1})$, and the data dependency
-from $(\DR{y}{1})$ to $(\DW{x}{1})$. If either is missing, then this execution
-is acyclic and hence allowed. For example dropping the control dependency
-results in:
-  \begin{gather*}
-    \THREAD{r\GETS x\SEMI y\GETS1}
-    \PAR
-    \THREAD{s\GETS y\SEMI x\GETS s}
-    \\[-.4ex]
-    \nonumber
-    \hbox{\begin{tikzinline}[node distance=1.5em]
-        \event{rx1}{\DR{x}{1}}{}
-        \event{wy1}{\DW{y}{1}}{right=of rx1}
-        \event{ry1}{\DR{y}{1}}{right=of wy1}
-        \event{wx1}{\DW{x}{1}}{right=of ry1}
-        \rf[out=20,in=160]{wy1}{ry1}
-        \rf[out=160,in=20]{wx1}{rx1}
-        \po{ry1}{wx1}
-      \end{tikzinline}}
-  \end{gather*}
-
-Unfortunately, while a simple syntactic approach to dependency calculation
-suffices for hardware models, it is not preserved by common compiler
-optimizations. For example, if we calculate control dependencies syntactically,
-then there is a dependency from $(\DR{x}{1})$ to $(\DW{y}{1})$, and therefore a cycle in, the candidate execution:
-  \begin{gather*}
-    \THREAD{r\GETS x\SEMI \IF r \THEN y\GETS1 \ELSE y\GETS1 \FI}
-    \PAR
-    \THREAD{s\GETS y\SEMI x\GETS s}
-    \\[-.4ex]
-    \nonumber
-    \hbox{\begin{tikzinline}[node distance=1.5em]
-        \event{rx1}{\DR{x}{1}}{}
-        \event{wy1}{\DW{y}{1}}{right=of rx1}
-        \event{ry1}{\DR{y}{1}}{right=of wy1}
-        \event{wx1}{\DW{x}{1}}{right=of ry1}
-        \rf[out=20,in=160]{wy1}{ry1}
-        \rf[out=160,in=20]{wx1}{rx1}
-        \po{rx1}{wy1}
-        \po{ry1}{wx1}
-      \end{tikzinline}}
-  \end{gather*}
-An optimizing compiler might lift the assignment $y\GETS1$ out of the conditional,
-thus removing the control dependency.
-
-Prominent solutions to the problem of dependency calculation include:
-\begin{itemize}
-
-\item \emph{syntactic} methods used in hardware models
-  such as ARM or x86-TSO \cite{alglave},
-\item \emph{speculative execution} methods (which give a semantics based on multiple executions
-  of the same program) such as the Java Memory Model~\cite{Manson:2005:JMM:1047659.1040336} 
-  and related models \cite{Jagadeesan:2010:GOS:2175486.2175503, DBLP:conf/popl/KangHLVD17, DBLP:journals/pacmpl/ChakrabortyV19},
-  % the speculative operational semantics of~\cite{Jagadeesan:2010:GOS:2175486.2175503},
-  % the promising semantics of~\cite{DBLP:conf/popl/KangHLVD17},
-  % or the event structures semantics of~\cite{DBLP:journals/pacmpl/ChakrabortyV19},
-\item \emph{rewriting} methods, which give an operational model
-  up to syntactic rewrites, such as~\cite{Pichon-Pharabod:2016:CSR:2837614.2837616}, and
-\item \emph{logical} methods, such as the pomsets with preconditions
-  model of~\cite{DBLP:journals/pacmpl/JagadeesanJR20}.
-
-\end{itemize}
-
-In this paper, we will focus on logical models, as those are compositional,
-and align well with existing models of sequential composition.
-The heart of the model of~\cite{DBLP:journals/pacmpl/JagadeesanJR20} is to add logical preconditions
-to events, which are introduced by store actions (modeling data dependencies)
-and conditionals (modeling control dependencies):
-  \begin{gather*}
-    \IF{s{<}1} \THEN z\GETS r{*}s \FI
-    \\
-    \nonumber
-    \hbox{\begin{tikzinline}[node distance=1.5em]
-        \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
-      \end{tikzinline}}
-  \end{gather*}
-Preconditions are discharged by being ordered after a read:
-  \begin{gather*}
-    r\GETS x\SEMI s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r{*}s \FI
-    \\
-    \nonumber
-    \hbox{\begin{tikzinline}[node distance=1.5em]
-        \event{rx0}{\DR{x}{0}}{}
-        \event{ry0}{\DR{y}{0}}{right=of rx0}
-        \event{wz0}{(s{=}0) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
-        \po{ry0}{wz0}
-      \end{tikzinline}}
-  \end{gather*}
-Note that there is dependency order from $(\DR{y}{0})$ to $(\DW{z}{0})$
-so the precondition for $(\DW{z}{0})$ only has to be satisfied assuming the hypothesis
-$(s{=}0)$. There is no matching order from $(\DR{x}{0})$ to $(\DW{z}{0})$
-which is why we do not assume the hypothesis $(r{=}0)$. Nonetheless, the precondition on
-$(\DW{z}{0})$ is a tautology, and so can be elided in the diagram:
-  \begin{gather*}
-    % r\GETS x\SEMI s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r{*}s \FI
-    % \\
-    \nonumber
-    \hbox{\begin{tikzinline}[node distance=1.5em]
-        \event{rx0}{\DR{x}{0}}{}
-        \event{ry0}{\DR{y}{0}}{right=of rx0}
-        \event{wz0}{\DW{z}{0}}{right=of ry0}
-        \po{ry0}{wz0}
-      \end{tikzinline}}
-  \end{gather*}
-
-While existing models of relaxed memory have detailed treatments of parallel composition,
-they often give sequential composition little attention, either ignoring it altogether,
-or treating it operationally with its usual small-step semantics. This paper
-investigates how existing models of sequential composition interact with relaxed memory.
-
-\subsection{Sequential composition}
-
-%% Program logics and semantics tell us that when executing ((S1; S2), state0),
-%% we execute (S1, state0) to arrive at state1, then execute (S2, state1) to
-%% arrive at the final state2.
-
-%% This is a delightfully simple story that can be explained to children.  It is
-%% also a lie.
-
-%% Processors execute instructions out of order, due to pipelines and caches.
-%% Compilers reorder programs even more dramatically.  All of this reordering is
-%% meant to be unobservable in single-threaded code.  In multi-threaded code,
-%% however, all bets are off.  A formal attempt to understand the resulting mess
-%% is known as a ``relaxed memory model.''
-
-%% Most of the relaxed memory models that have been proposed are designed to
-%% help us understand whole program execution: they have no compositionality
-%% properties whatsoever.  Recently, denotation models have appeared that treat
-%% \emph{concurrent} execution compositionally.  One such model is ``Pomsets with
-%% Preconditions''.  It remains an open question, however, whether it is
-%% possible to treat \emph{sequential} execution compositionally in such a model,
-%% without overly restricting processors and compilers.
-
-%% We propose adding families of predicate transformers to Pomsets with
-%% Preconditions.  The resulting model is denotational, supporting both parallel
-%% and sequential composition.  When composing (S1;S2), the predicate
-%% transformer used to validate the precondition of an event in S2 is chosen
-%% based on the dependency order from S1 into this event.  As usual in work on
-%% relaxed memory, we have not handled loops or recursion.
-
-%% Happily, most of the results expected of a relaxed memory model can be
-%% established by appeal to prior work.  So here we are able to concentrate on
-%% the model itself.  The model is formalized in Agda, where we have established
-%% associativity for sequential composition.
-
-%% For the memory-model specialist, we retain the good properties of the prior
-%% work on Pomsets with Preconditions, fixing some errors along the way.  These
-%% properties include efficient implementation on ARMv8, support for compiler
-%% optimizations, support for logics that prove the absence of thin-air
-%% behaviors, and a local data race freedom theorem.
-
-% \begin{figure*}
-%   \input{fig-sp.tex}
-%   \caption{Weakest precondition semantics}
-%   \label{fig:sp}
-% \end{figure*}
-
-Our approach follows that of weakest precondition semantics of
-\citet{DBLP:journals/cacm/Dijkstra75}, which provides an alternative
-characterization of Hoare logic \citep{Hoare:1969:ABC:363235.363259} by
-mapping postconditions to preconditions. We recall the definition of
-$\fwp{\aCmd}{\bForm}$ for loop-free code below. % in Figure~\ref{fig:sp}
-\begin{itemize}
+Introductory programmers are taught \emph{sequential abstraction}: that the
+program $\aCmd_1\SEMI \aCmd_2$ executes $\aCmd_1$ before $\aCmd_2$.  Since
+the late 60s, we've been able to explain this using logic
+\citep{Hoare:1969:ABC:363235.363259}.  In
+\citeauthor{DBLP:journals/cacm/Dijkstra75}'s
+[\citeyear{DBLP:journals/cacm/Dijkstra75}] formulation, we think of programs
+as \emph{predicate transformers}, where predicates describe the state of
+memory in the system.  In the calculus of weakest preconditions, programs map
+postconditions to preconditions.  We recall the definition of
+$\fwp{\aCmd}{\bForm}$ for loop-free code below. 
+\begin{multicols}{2}
+  \begin{enumerate}[,label=(\textsc{d}\arabic*),ref=\textsc{d}\arabic*]
   \item
@@
     \end{math}
-\item 
-  \begin{math}
-    \fwp{\ABORT}{\bForm} = \FALSE
-  \end{math}
-\item
-  \begin{math}
-    \fwp{r\GETS \aExp}{\bForm} = \bForm[\aExp/r]
-  \end{math}
+    % \item 
+    %   \begin{math}
+    %     \fwp{\ABORT}{\bForm} = \FALSE
+    %   \end{math}
+    \stepcounter{enumi}
+  \item[] \labeltext[\textsc{d}2]{}{wp-assign}
+    \begin{enumerate}[leftmargin=0pt]
+    \item \label{wp-write}
+      $\fwp{\PW{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
+    \item \label{wp-let}
+      $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
+    \item \label{wp-read}
+      $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \aLoc{=}\aReg\limplies\bForm$ %\bForm[\aLoc/\aReg]$ %
+    \end{enumerate}
+    % \item 
+    %   \begin{math}
+    %     \fwp{\LET{x}{\aExp}}{\bForm} = \bForm[\aExp/x]
+    %   \end{math}
   \item
@@
     \end{math}
-\end{itemize}
-The rule we are most interested
-in is the one for sequential composition, which maps sequential composition of programs
-to function composition of predicate transformers.
+  \end{enumerate}
+\end{multicols}
+For this language, the Hoare triple
+$\hoare{\aForm}{\aCmd}{\bForm}$ holds exactly when $\aForm \limplies
+\fwp{\aCmd}{\bForm}$.  We have split
+\citeauthor{DBLP:journals/cacm/Dijkstra75}'s rule for assignment
+\eqref{wp-assign} into three cases. In our notation,
+$\aReg$--$\bReg$ range over thread-local registers, which may be assigned at
+most once,
+$\aLoc$--$\cLoc$ range over shared memory references, and
+$\aExp$--$\bExp$ range over thread-local expressions, which do \emph{not}
+include
+$\aLoc$--$\cLoc$.\footnote{We wrote \ref{wp-read} using implication rather
+  than substitution; the two formulations are equivalent under the assumption
+  that each register is assigned at most once.}
+
+This is quite a pretty explanation of sequential computation in a sequential
+context.  In a concurrent context, however, \ref{wp-read} is unsound!
+\ref{wp-read} assumes that a read from
+$x$ must be fulfilled by a preceding write to
+$x$.  In a concurrent context, writes may come from other threads.
+
+Existing approaches to sequential composition in the concurrent context
+either assume exclusive access, as in concurrent separation logic
+\cite{OHearn:2007:RCL:1235896.1236121}, or abandon the logical approach
+altogether, as in the pomset model of
+\citet{DBLP:journals/corr/abs-1804-04214}---this model uses syntactic
+dependencies and thus dramatically limits compiler optimization.  This leaves
+open the question of how to apply logic to programs with data races.
+To understand the solution, one must first understand the constraints imposed
+by hardware and compilers.
+
+% Assuming all variables are initialized to $0$,
+% $\PW{x}{1}\SEMI\PW{y}{1}$ can be distinguished from $\PW{y}{1}\SEMI\PW{x}{1}$
+% by the concurrent observer $\PR{x}{r}\SEMI\PR{y}{s}$
+
+% Microprocessors and compilers do not execute this way, but they attempt to
+% preserve the abstraction, at least for sequential code.  For concurrent code,
+% all bets are off.
+
+% \citet{DBLP:conf/snapl/MarinoMMNS15} argue that the ``silently shifting
+% semicolon'' is problematic for programmers, and thus concurrent languages
+% should guarantee sequential abstraction.  But processor and language
+% implementors balk at the prospect, due to significant costs.
+
+% These leaves programmers in a bit of pickle: What does that semicolon mean?
+
+% The last decade has seen a slew of research in ``relaxed memory models,''
+% which attempt to explain concurrent execution.  The models are mostly
+% operational, sometimes with an additional series of axioms to eliminate
+% ``bad'' executions.  These models have limited modularity properties,
+% typically requiring whole program analysis.
+
+% In this paper, we attempt to
+
+\subsection{Memory Models}
+
+For single-threaded programs, memory can be thought of as you might
+expect: programs write to, and read from, memory references.
+This can be thought of as a total order of reads and writes,
+where each read has a matching \emph{fulfilling} write,
+for example:
+  \begin{gather*}
+    \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}\SEMI
+    \PR{y}{r}\SEMI \PR{x}{s}}
+    \\[-.4ex]
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{wx0}{\DW{x}{0}}{}
+        \event{wx1}{\DW{x}{1}}{right=of wx0}
+        \event{wy2}{\DW{y}{2}}{right=of wx1}
+        \event{ry2}{\DR{y}{2}}{right=3em of wy2}
+        \event{rx1}{\DR{x}{1}}{right=of ry2}
+        \rf[out=15,in=165]{wy2}{ry2}
+        \rf[out=15,in=165]{wx1}{rx1}
+        \po{wx0}{wx1}
+        \po{wx1}{wy2}
+        \po{wy2}{ry2}
+        \po{ry2}{rx1}
+      \end{tikzinline}}
+  \end{gather*}
+% (In examples, $\aReg$--$\bReg$ range over thread-local registers and $\aLoc$-$\cLoc$
+% range over shared memory references.)
+
+This model naturally extends to the case of shared-memory concurrency, leading to a \emph{sequentially consistent}
+semantics \cite{Lamport:1979:MMC:1311099.1311750}, in which \emph{program order} inside a thread implies
+a total \emph{causal order} between read and write events, for example:
+  \begin{gather*}
+    \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}}
+    \PAR
+    \THREAD{\PR{y}{r}\SEMI \PR{x}{s}}
+    \\[-.4ex]
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{wx0}{\DW{x}{0}}{}
+        \event{wx1}{\DW{x}{1}}{right=of wx0}
+        \event{wy2}{\DW{y}{2}}{right=of wx1}
+        \event{ry2}{\DR{y}{2}}{right=3em of wy2}
+        \event{rx1}{\DR{x}{1}}{right=of ry2}
+        \rf[out=15,in=165]{wy2}{ry2}
+        \rf[out=15,in=165]{wx1}{rx1}
+        \po{wx0}{wx1}
+        \po{wx1}{wy2}
+        \po{ry2}{rx1}
+      \end{tikzinline}}
+  \end{gather*}
+
+Unfortunately, this model does not compile efficiently to commodity
+hardware, resulting in a 37--73\% increase in CPU time on Arm8~\cite{Liu:2019:ASC:3314221.3314611} and,
+hence, in power consumption.  Developers of software and compilers have
+therefore been faced with a difficult trade-off, between an elegant
+model of memory, and its impact on resource usage (such as size of
+data centers, electricity bills and carbon footprint). Unsurprisingly,
+many have chosen to prioritize efficiency over elegance.
+
+This has led to \emph{relaxed memory models}, in which the requirement of
+sequential consistency is weakened to only apply \emph{per-location} and not globally
+over the whole program. This allows executions which
+are inconsistent with program order, such as:
+  \begin{gather*}
+    \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}}
+    \PAR
+    \THREAD{\PR{y}{r}\SEMI \PR{x}{s}}
+    \\[-.4ex]
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{wx0}{\DW{x}{0}}{}
+        \event{wx1}{\DW{x}{1}}{right=of wx0}
+        \event{wy2}{\DW{y}{2}}{right=of wx1}
+        \event{ry2}{\DR{y}{2}}{right=3em of wy2}
+        \event{rx0}{\DR{x}{0}}{right=of ry2}
+        \rf[out=15,in=165]{wy2}{ry2}
+        \rf[out=15,in=165]{wx0}{rx0}
+        \po{wx0}{wx1}
+        \po[out=-165,in=-15]{rx0}{wx1}
+      \end{tikzinline}}
+  \end{gather*}
+
+In such models, the causal order between events is important,
+and includes control and data dependencies, to avoid
+paradoxical ``out of thin air'' examples such as:
+  \begin{gather*}
+    \THREAD{\PR{x}{r}\SEMI \IF r \THEN \PW{y}{1} \FI}
+    \PAR
+    \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
+    \\[-.4ex]
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{rx1}{\DR{x}{1}}{}
+        \event{wy1}{\DW{y}{1}}{right=of rx1}
+        \event{ry1}{\DR{y}{1}}{right=3em of wy1}
+        \event{wx1}{\DW{x}{1}}{right=of ry1}
+        \rf[out=15,in=165]{wy1}{ry1}
+        \rf[out=165,in=15]{wx1}{rx1}
+        \po{rx1}{wy1}
+        \po{ry1}{wx1}
+      \end{tikzinline}}
+  \end{gather*}
+This candidate execution forms a cycle in causal order, so is disallowed,
+but this depends crucially on the control dependency
+from $(\DR{x}{1})$ to $(\DW{y}{1})$, and the data dependency
+from $(\DR{y}{1})$ to $(\DW{x}{1})$. If either is missing, then this execution
+is acyclic and hence allowed. For example dropping the control dependency
+results in:
+  \begin{gather*}
+    \THREAD{\PR{x}{r}\SEMI \PW{y}{1}}
+    \PAR
+    \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
+    \\[-.4ex]
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{rx1}{\DR{x}{1}}{}
+        \event{wy1}{\DW{y}{1}}{right=of rx1}
+        \event{ry1}{\DR{y}{1}}{right=3em of wy1}
+        \event{wx1}{\DW{x}{1}}{right=of ry1}
+        \rf[out=15,in=165]{wy1}{ry1}
+        \rf[out=165,in=15]{wx1}{rx1}
+        \po{ry1}{wx1}
+      \end{tikzinline}}
+  \end{gather*}
+
+Unfortunately, while a simple syntactic approach to dependency calculation
+suffices for hardware models, it is not preserved by common compiler
+optimizations. For example, if we calculate control dependencies syntactically,
+then there is a dependency from $(\DR{x}{1})$ to $(\DW{y}{1})$, and therefore a cycle in, the candidate execution:
+  \begin{gather*}
+    \THREAD{\PR{x}{r}\SEMI \IF r \THEN \PW{y}{1} \ELSE \PW{y}{1} \FI}
+    \PAR
+    \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
+    \\[-.4ex]
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{rx1}{\DR{x}{1}}{}
+        \event{wy1}{\DW{y}{1}}{right=of rx1}
+        \event{ry1}{\DR{y}{1}}{right=3em of wy1}
+        \event{wx1}{\DW{x}{1}}{right=of ry1}
+        \rf[out=15,in=165]{wy1}{ry1}
+        \rf[out=165,in=15]{wx1}{rx1}
+        \po{rx1}{wy1}
+        \po{ry1}{wx1}
+      \end{tikzinline}}
+  \end{gather*}
+An optimizing compiler might lift the assignment $\PW{y}{1}$ out of the conditional,
+thus removing the control dependency.
+
+To address this, \citet{DBLP:journals/pacmpl/JagadeesanJR20} introduced
+\emph{Pomsets with Preconditions}, where events are labeled with logical
+formulae.  Nontrivial preconditions are introduced by store actions (modeling
+data dependencies) and conditionals (modeling control dependencies):
+  \begin{gather*}
+    \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
+    \\
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
+      \end{tikzinline}}
+  \end{gather*}
+Preconditions are discharged by being ordered after a read:
+  \begin{gather*}
+    \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
+    \\
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{rx0}{\DR{x}{0}}{}
+        \event{ry0}{\DR{y}{0}}{right=of rx0}
+        \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
+        \po{ry0}{wz0}
+      \end{tikzinline}}
+  \end{gather*}
+Note that there is dependency order from $(\DR{y}{0})$ to $(\DW{z}{0})$
+so the precondition for $(\DW{z}{0})$ only has to be satisfied assuming the hypothesis
+$(0{=}s)$. There is no matching order from $(\DR{x}{0})$ to $(\DW{z}{0})$
+which is why we do not assume the hypothesis $(0{=}r)$. Nonetheless, the precondition on
+$(\DW{z}{0})$ is a tautology, and so can be elided in the diagram:
+\begin{gather*}
+  % \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
+  % \\
+  \nonumber
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{rx0}{\DR{x}{0}}{}
+      \event{ry0}{\DR{y}{0}}{right=of rx0}
+      \event{wz0}{\DW{z}{0}}{right=of ry0}
+      \po{ry0}{wz0}
+    \end{tikzinline}}
+\end{gather*}
+
+\subsection{Predicate Transformers For Relaxed Memory}
+
+Pomsets with Preconditions show how the logical approach to sequential
+dependency calculation can be mixed into a relaxed memory model.  However,
+\citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} do not provide a model of
+sequential composition.  Instead, their model uses \emph{prefixing}, which
+requires that the model is built from right to left: events are prepended one
+at a time, with perfect knowledge of the future.  This makes reasoning about
+sequential program fragments difficult.  For example,
+\citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} state the equivalence
+allowing reordering independent writes as follows,
+\begin{align*}
+  \sem{\aLoc \GETS \aExp \SEMI \bLoc  \GETS \bExp\SEMI\aCmd} &=
+  \sem{\bLoc  \GETS \bExp\SEMI \aLoc \GETS \aExp\SEMI\aCmd} \;\;\text{if}\; \aLoc\neq\bLoc
+\end{align*}
+where $\aCmd$ is the entire future computation!  By formalizing sequential
+composition, we can show:
+\begin{align*}
+  \sem{\aLoc \GETS \aExp \SEMI \bLoc  \GETS \bExp} &=
+  \sem{\bLoc  \GETS \bExp\SEMI \aLoc \GETS \aExp} \;\;\text{if}\; \aLoc\neq\bLoc
+\end{align*}
+Then the equivalence holds in any (sequential) context.
+
+% While existing models of relaxed memory have detailed treatments of parallel composition,
+% they often give sequential composition little attention, either ignoring it altogether,
+% or treating it operationally with its usual small-step semantics. This paper
+% investigates how existing models of sequential composition interact with relaxed memory.
+
+% \begin{figure*}
+%   \input{fig-sp.tex}
+%   \caption{Weakest precondition semantics}
+%   \label{fig:sp}
+% \end{figure*}
+
+% Our approach follows that of weakest precondition semantics of
+% \citet{DBLP:journals/cacm/Dijkstra75}, which provides an alternative
+% characterization of Hoare logic \citep{Hoare:1969:ABC:363235.363259} by
+% mapping postconditions to preconditions. We recall the definition of
+% $\fwp{\aCmd}{\bForm}$ for loop-free code below. % in Figure~\ref{fig:sp}
+% \begin{itemize}
+% \item
+%   \begin{math}
+%     \fwp{\SKIP}{\bForm} = \bForm
+%   \end{math}
+% % \item 
+% %   \begin{math}
+% %     \fwp{\ABORT}{\bForm} = \FALSE
+% %   \end{math}
+% \item
+%   \begin{math}
+%     \fwp{\LET{r}{\aExp}}{\bForm} = \bForm[\aExp/r]
+%   \end{math}
+% \item
+%   \begin{math}
+%     \fwp{\aCmd_1;\aCmd_2}{\bForm} = \fwp{\aCmd_1}{\fwp{\aCmd_2}{\bForm}}
+%   \end{math}
+% \item
+%   \begin{math}
+%     \fwp{\IF{\aExp}\THEN \aCmd_1\ELSE \aCmd_2\FI}{\bForm}= {}
+%   \end{math}
+%   \\
+%   \begin{math}
+%     ((\aExp{\ne}0) \limplies \fwp{\aCmd_1}{\bForm}) \land
+%     ((\aExp{=}0) \limplies \fwp{\aCmd_2}{\bForm})
+%   \end{math}
+% \end{itemize}
+% The rule we are most interested
+% in is the one for sequential composition, which maps sequential composition of programs
+% to function composition of predicate transformers.
 
@@
   \begin{gathered}
-      r \GETS x
+    \PR{x}{r}
     \\
@@
         \event{rx0}{\DR{x}{0}}{}
-          \xform{rx0d}{(r{=}0) \limplies \bForm}{below=of rx0}
+        \xform{rx0d}{(0{=}r) \limplies \bForm}{right=of rx0}
         \xo{rx0}{rx0d}
@@
   \begin{gathered}
-      s \GETS y
+    \PR{y}{s}
     \\
@@
         \event{ry0}{\DR{y}{0}}{}
-          \xform{ry0d}{(s{=}0) \limplies \bForm}{below=of ry0}
+        \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
         \xo{ry0}{ry0d}
@@
   \begin{gathered}
-      \IF{s{<}1} \THEN z\GETS r{*}s \FI
+    \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
     \\
-      \hbox{\begin{tikzinline}[node distance=1ex and1.5em]
+    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
         \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
-          \xform{wz0d}{\bForm}{below=of wz0}
-          \xo[out=-30]{wz0}{wz0d}
+        \xform{wz0d}{\bForm[r{*}s/z]}{right=of wz0}
+        \xo{wz0}{wz0d}
       \end{tikzinline}}
@@
 \end{align*}
-In the rightmost program above, the write to $z$ affects the shared store, not the
-local state of the thread, therefore we assign it the identity transformer.
+The predicate transformer from the write matches
+\citeauthor{DBLP:journals/cacm/Dijkstra75}.  For the reads, however,
+\ref{wp-read} defines the transformer of $\PR{x}{r}$ to be
+$(x{=}r) \limplies \bForm$.  Instead, we use $(0{=}r) \limplies \bForm$,
+reflecting the fact that $0$ may come from a concurrent write.  The
+obligation to find a matching write is moved from the sequential semantics of
+\emph{substitution} and \emph{implication} to the concurrent semantics of
+\emph{fulfillment}.
 
-For the sequentially consistent semantics, sequential composition 
-is straightforward: we apply each predicate transformer to the preconditions
-of subsequent events, and compose the predicate transformers:
+% In the rightmost program above, the write to $z$ affects the shared store, not the
+% local state of the thread, therefore we assign it the identity transformer.
+
+For the sequentially consistent semantics, sequential composition is
+straightforward: we apply each predicate transformer to the preconditions of
+subsequent events, composing the predicate transformers.  (In subsequent
+diagrams, we only show predicate transformers for reads.)
 \begin{gather*}
-    r\GETS x\SEMI s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r{*}s \FI
-    \\
+  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
+  \\[-1.5ex]
   \nonumber
@@
       \event{ry0}{\DR{y}{0}}{right=of rx0}
-        \event{wz0}{(r{=}0) \limplies (s{=}0) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
-        \xform{rx0ry0d}{(r{=}0) \limplies(s{=}0) \limplies \bForm}{below=of ry0}
+      \event{wz0}{(0{=}r) \limplies (0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
+      \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{left=of rx0}
       \po{rx0}{ry0}
       \po{ry0}{wz0}
-        \xos{rx0}{rx0ry0d}
-        \xos{ry0}{rx0ry0d}
-        \xos{wz0}{rx0ry0d}
+      \xo{rx0}{rx0ry0d}
+      \xo[out=155,in=15]{ry0}{rx0ry0d}
+      % \xo{wz0}{rx0ry0d}
     \end{tikzinline}}
@@
     \begin{gathered}
-      r \GETS x
+      \PR{x}{r}
       \\
@@
           \event{rx0}{\DR{x}{0}}{}
-          \xform{rx0i}{\bForm}{above=of rx0}
-          \xform{rx0d}{(r{=}0) \limplies \bForm}{below=of rx0}
+          \xform{rx0i}{\bForm}{left=.4em of rx0}
+          \xform{rx0d}{(0{=}r) \limplies \bForm}{right=of rx0}
           \xo{rx0}{rx0d}
@@
     \begin{gathered}
-      s \GETS y
+      \PR{y}{s}
       \\
@@
           \event{ry0}{\DR{y}{0}}{}
-          \xform{ry0i}{\bForm}{above=of ry0}
-          \xform{ry0d}{(s{=}0) \limplies \bForm}{below=of ry0}
+          \xform{ry0i}{\bForm}{left=.4em of ry0}
+          \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
           \xo{ry0}{ry0d}
@@
     \end{gathered}
-    &&
-    \begin{gathered}
-      \IF{s{<}1} \THEN z\GETS r{*}s \FI
-      \\
-      \hbox{\begin{tikzinline}[node distance=1ex and1.5em]
-          \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
-          \xform{wz0i}{\bForm}{above=of wz0}
-          \xform{wz0d}{\bForm}{below=of wz0}
-          \xo[out=-30]{wz0}{wz0d}
-      \end{tikzinline}}
-    \end{gathered}
+    % &&
+    % \begin{gathered}
+    %   \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
+    %   \\
+    %   \hbox{\begin{tikzinline}[node distance=1ex and1.5em]
+    %       \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
+    %       \xform{wz0i}{\bForm}{left=.4em of wz0}
+    %       \xform{wz0d}{\bForm}{right=of wz0}
+    %       \xo{wz0}{wz0d}
+    %   \end{tikzinline}}
+    % \end{gathered}
   \end{align*}
@@
 but dependent on $(\DR y0)$, which is the transformer
-which maps $\bForm$ to $(s{=}0) \limplies \bForm$.
+which maps $\bForm$ to $(0{=}s) \limplies \bForm$.
   \begin{gather*}
-    r\GETS x\SEMI s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r{*}s \FI
-    \\
+    \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
+    \\[-2ex]
     \nonumber
-    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
-        \event{rx0}{\DR{x}{0}}{}
-        \event{ry0}{\DR{y}{0}}{right=of rx0}
-        \event{wz0}{(s{=}0) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
-        \xform{rx0ry0d}{(r{=}0) \limplies(s{=}0) \limplies \bForm}{above=of ry0}
-        \xform{rx0d}{(r{=}0) \limplies \bForm}{below=of rx0}
-        \xform{ry0d}{(s{=}0) \limplies \bForm}{right=of rx0d}
-        \xform{wz0i}{\bForm}{above=of wz0}
-        \po{ry0}{wz0}
-        \xos{rx0}{rx0d}
-        \xos{ry0}{ry0d}
-        \xos{rx0}{rx0ry0d}
-        \xos{ry0}{rx0ry0d}
+    \hbox{\begin{tikzinline}[node distance=1ex and 1.2em]
+        \xform{rx0d}{(0{=}r) \limplies \bForm}{}
+        \event{rx0}{\DR{x}{0}}{right=of rx0d}
+        \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{right=of rx0}
+        \event{ry0}{\DR{y}{0}}{right=of rx0ry0d}
+        \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
+        \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0d}
+        \xform{wz0i}{\bForm}{left=of rx0d}
+        \po[out=20,in=170]{ry0}{wz0}
+        \xo{rx0}{rx0d}
+        \xo{ry0}{ry0d}
+        \xo{rx0}{rx0ry0d}
+        \xo{ry0}{rx0ry0d}
       \end{tikzinline}}
@@
 \begin{gather*}
-    r\GETS x\SEMI s\GETS y
+  \PR{x}{r}\SEMI \PR{y}{s}
   \\
@@
   \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
-        \event{rx0}{\DR{x}{0}}{}
-        \event{ry0}{\DR{y}{0}}{right=5em of rx0}
-        \xform{rx0ry0d}{(r{=}0) \limplies(s{=}0) \limplies \bForm}{above
-          right=1ex and -2.5em of rx0}
-        \xform{rx0d}{(r{=}0) \limplies \bForm}{left=of rx0}
-        \xform{ry0d}{(s{=}0) \limplies \bForm}{right=of ry0}
-        \xform{wz0i}{\bForm}{right=of rx0ry0d}
-        \xos{rx0}{rx0d}
-        \xos{ry0}{ry0d}the \emph{independent}
-transformer
-        \xos{rx0}{rx0ry0d}
-        \xos{ry0}{rx0ry0d}
+      \xform{rx0d}{(0{=}r) \limplies \bForm}{}
+      \event{rx0}{\DR{x}{0}}{right=of rx0d}
+      \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{right=of rx0}
+      \event{ry0}{\DR{y}{0}}{right=of rx0ry0d}
+      \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
+      \xform{wz0i}{\bForm}{left=of rx0d}
+      \xo{rx0}{rx0d}
+      \xo{ry0}{ry0d}
+      \xo{rx0}{rx0ry0d}
+      \xo{ry0}{rx0ry0d}
     \end{tikzinline}}
@@
 \begin{gather*}
-    s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r{*}s \FI
+  \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
   \\
@@
       \event{ry0}{\DR{y}{0}}{}
-        \event{wz0}{(s{=}0) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
-        \xform{ry0d}{(s{=}0) \limplies \bForm}{above=of ry0}
-        \xform{wz0i}{\bForm}{above=of wz0}
+      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
+      \xform{ry0d}{(0{=}s) \limplies \bForm}{left=of ry0}
+      \xform{wz0i}{\bForm}{left=of ry0d}
       \po{ry0}{wz0}
-        \xo[xleft]{ry0}{ry0d}
+      \xo{ry0}{ry0d}
     \end{tikzinline}}
 \end{gather*}
-This is an instance of a general result that sequential composition forms a monoid,
-as one would hope.
+This is an instance of the general result that sequential composition forms a monoid.
 
@@
 
+Prominent solutions to the problem of dependency calculation include:
+\begin{itemize}
+
+\item \emph{syntactic} methods used in hardware models
+  such as ARM or x86-TSO \cite{alglave},
+\item \emph{speculative execution} methods (which give a semantics based on multiple executions
+  of the same program) such as the Java Memory Model~\cite{Manson:2005:JMM:1047659.1040336} 
+  and related models \cite{Jagadeesan:2010:GOS:2175486.2175503, DBLP:conf/popl/KangHLVD17, DBLP:journals/pacmpl/ChakrabortyV19},
+  % the speculative operational semantics of~\cite{Jagadeesan:2010:GOS:2175486.2175503},
+  % the promising semantics of~\cite{DBLP:conf/popl/KangHLVD17},
+  % or the event structures semantics of~\cite{DBLP:journals/pacmpl/ChakrabortyV19},
+\item \emph{rewriting} methods, which give an operational model
+  up to syntactic rewrites, such as~\cite{Pichon-Pharabod:2016:CSR:2837614.2837616}, and
+\item \emph{logical} methods, such as the pomsets with preconditions
+  model of~\cite{DBLP:journals/pacmpl/JagadeesanJR20}.
+
+\end{itemize}
 \begin{itemize}
@@
 dependencies, thus invalidating many compiler optimizations.
+
+\citet{DBLP:conf/snapl/MarinoMMNS15} argue that the ``silently shifting
+semicolon'' is problematic for programmers, and thus concurrent languages
+should guarantee sequential abstraction.  But processor and language
+implementors balk at the prospect, due to significant costs.

### 2021-04-14 paper/intro.tex
related

@@
 \begin{itemize}
-\item \S\ref{sec:model} presents the basic model, with few features
-  required of the logic of preconditions, but a resulting lack of fidelity
-  to exiting models,
-\item \S\ref{sec:q} adds a model of \emph{quiescence} to the logic,
-  required to model coherence (accessing $x$ has a precondition that $x$ is quiescent)
-  and synchronization (a releasing write requires all locations to be quiescent),
-\item \S\ref{sec:arm} adds the features required for efficient compilation
-  to modern architectures: downgrading some synchronized accesses to relaxed,
-  and removing read-read dependencies, and
-\item \S\ref{sec:complications} show how to address common litmus tests.
+\item \S\ref{sec:model} presents the basic model, which satisfies many
+  desiderata, but not all.
+\item \S\ref{sec:arm} shows two approaches for efficient implementation on
+  Arm.  The first uses a suboptimal lowering for acquiring reads.  The second
+  uses an optimal lowering, but requires changes to the definitions of
+  parallel and sequential composition.
+\item \S\ref{sec:variants} generalizes the basic semantics of read and write
+  to validate compiler optimizations.
 \end{itemize}
 % Acknowledgements go here, once we're not double-blinded.
-The definitions in this paper have been formalized in Agda.
-
+%The definitions in this paper have been formalized in Agda.
 Because it is closely related, we expect that the memory-model results of
@@
 compositional reasoning for temporal safety properties and {local} \drfsc{}
-as in \cite{Dolan:2018:BDR:3192366.3192421,DBLP:conf/ppopp/DongolJR19}.
-In \textsection\ref{sec:arm}, we provide an alternative proof strategy for
-efficient compilation to \armeight{}, which improves upon that of
-\cite{DBLP:journals/pacmpl/JagadeesanJR20} by using a recent alternative
-characterization of \armeight{}.
+as in \cite{Dolan:2018:BDR:3192366.3192421,DBLP:conf/ppopp/DongolJR19,promising-ldrf}.
+% In \textsection\ref{sec:arm}, we provide an alternative proof strategy for
+% efficient compilation to \armeight{}, which improves upon that of
+% \cite{DBLP:journals/pacmpl/JagadeesanJR20} by using a recent alternative
+% characterization of \armeight{}.
 
@@
 
-Prominent solutions to the problem of dependency calculation include:
-\begin{itemize}
+\citet{DBLP:conf/snapl/MarinoMMNS15} argue that the ``silently shifting
+semicolon'' sufficiently problematic for programmers that concurrent
+languages should guarantee sequential abstraction, despite the performance
+penalties.  In this paper, we have take the opposite approach.  We have
+attempted to find the most intellectually tractable model that encompasses
+all of the messiness of relaxed memory.
 
-\item \emph{syntactic} methods used in hardware models
-  such as ARM or x86-TSO \cite{alglave},
-\item \emph{speculative execution} methods (which give a semantics based on multiple executions
-  of the same program) such as the Java Memory Model~\cite{Manson:2005:JMM:1047659.1040336} 
-  and related models \cite{Jagadeesan:2010:GOS:2175486.2175503, DBLP:conf/popl/KangHLVD17, DBLP:journals/pacmpl/ChakrabortyV19},
-  % the speculative operational semantics of~\cite{Jagadeesan:2010:GOS:2175486.2175503},
-  % the promising semantics of~\cite{DBLP:conf/popl/KangHLVD17},
-  % or the event structures semantics of~\cite{DBLP:journals/pacmpl/ChakrabortyV19},
-\item \emph{rewriting} methods, which give an operational model
-  up to syntactic rewrites, such as~\cite{Pichon-Pharabod:2016:CSR:2837614.2837616}, and
-\item \emph{logical} methods, such as the pomsets with preconditions
-  model of~\cite{DBLP:journals/pacmpl/JagadeesanJR20}.
+There are prior studies of relaxed memory that include sequential
+composition.  \citet{DBLP:journals/corr/abs-1804-04214} define a semantics
+using pomsets without preconditions. Instead, their model uses syntactic
+dependencies, thus invalidating many compiler optimizations.  They also
+require a fence after every relaxed read on \armeight{}.
 
-\end{itemize}
-\begin{itemize}
-\item Models with strong dependencies that disallow compiler optimizations
-  and efficient implementation on \armeight{}:
-  \cite{Lamport:1979:MMC:1311099.1311750,
-    DBLP:conf/pldi/LahavVKHD17,
-    Dolan:2018:BDR:3192366.3192421,
-    DBLP:conf/pldi/LahavVKHD17,
-    DBLP:conf/lics/JeffreyR16,
-    Boehm:2014:OGA:2618128.2618134,
+\citet{DBLP:conf/esop/PaviottiCPWOB20} give a denotational semantics,
+calculating dependencies using event structures rather than logic.  They give
+the semantics of sequential composition in continuation passing style,
+whereas we give it in direct style.
+
+\citet{Pichon-Pharabod:2016:CSR:2837614.2837616} also use event structures to
+calculate dependencies, combined with an operational semantics that
+incorporates program transformations.  This approach seems to require
+whole-program analysis.
+
+Other studies of relaxed memory can be categorized by their approach to
+dependency calculation.  Hardware models use syntactic dependencies
+\cite{alglave}.  Many software models do not bother with dependencies at all
+\cite{Batty:2011:MCC:1926385.1926394, DBLP:journals/pacmpl/WattRP19,
+  DBLP:conf/pldi/WattPPBDFPG20, goMM}.  Others have strong dependencies that
+disallow compiler optimizations and efficient implementation, typically
+requiring fences for every relaxed read on Arm
+\cite{Lamport:1979:MMC:1311099.1311750, DBLP:conf/pldi/LahavVKHD17,
+  Dolan:2018:BDR:3192366.3192421, DBLP:conf/pldi/LahavVKHD17,
+  DBLP:conf/lics/JeffreyR16, Boehm:2014:OGA:2618128.2618134,
   DBLP:journals/corr/abs-1804-04214}.
-\item Models with weak dependencies that allow some forms of thin-air
-  reasoning:
-  \cite{Manson:2005:JMM:1047659.1040336,
-    DBLP:conf/esop/JagadeesanPR10,
-    DBLP:conf/popl/KangHLVD17,
-    DBLP:journals/pacmpl/ChakrabortyV19}
-\item Models that do not calculate dependencies:
-  \cite{Batty:2011:MCC:1926385.1926394,
-    DBLP:journals/pacmpl/WattRP19,
-    DBLP:conf/pldi/WattPPBDFPG20,
-    goMM}.
-\end{itemize}
-Like us, \citet{DBLP:journals/corr/abs-1804-04214} use pomsets to create a
-compositional account of sequential composition.  However, their model
-requires a fence after every read on \armeight{}, and uses syntactic
-dependencies, thus invalidating many compiler optimizations.
 
-\citet{DBLP:conf/snapl/MarinoMMNS15} argue that the ``silently shifting
-semicolon'' is problematic for programmers, and thus concurrent languages
-should guarantee sequential abstraction.  But processor and language
-implementors balk at the prospect, due to significant costs.
+Many of the most prominent models are based on speculative execution
+\cite{Manson:2005:JMM:1047659.1040336, DBLP:conf/esop/JagadeesanPR10,
+  DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}.
+In their introduction, \citet{DBLP:journals/pacmpl/JagadeesanJR20} note that
+these models fail to validate compositional reasoning of temporal
+properties---see their examples OOTA4 and OOTA5 (from
+\citep{DBLP:journals/toplas/Lochbihler13}).  The difference with our model
+can be understood in terms of the valid program transformations.  The
+speculative models allow reads to be introduced, with subsequent case
+analysis on the value read---effectively, this can turn one read into two,
+with different choices taken for the values read.  Our model invalidates this
+transformation.  In return, our model enjoys compositionality for temporal
+safety properties.

### 2021-04-14 paper/intro.tex
second pass over intro/related

@@
 include
-$\aLoc$--$\cLoc$.\footnote{We wrote \ref{wp-read} using implication rather
-  than substitution; the two formulations are equivalent under the assumption
-  that each register is assigned at most once.}
+$\aLoc$--$\cLoc$.\footnote{Under these assumptions, \eqref{wp-read} is
+  equivalent to $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \bForm[\aLoc/\aReg]$.}
 
@@
 dependencies and thus dramatically limits compiler optimization.  This leaves
-open the question of how to apply logic to programs with data races.
-To understand the solution, one must first understand the constraints imposed
-by hardware and compilers.
+open the question of how to apply logic to racy programs without
+overstraining the implementation.  To understand the solution, one must first
+understand the constraints imposed by hardware and compilers.
 
@@
 
-Unfortunately, while a simple syntactic approach to dependency calculation
-suffices for hardware models, it is not preserved by common compiler
-optimizations. For example, if we calculate control dependencies syntactically,
-then there is a dependency from $(\DR{x}{1})$ to $(\DW{y}{1})$, and therefore a cycle in, the candidate execution:
+While syntactic dependency calculation suffices for hardware models, it is
+not preserved by common compiler optimizations. For example, if we calculate
+control dependencies syntactically, then there is a dependency from
+$(\DR{x}{1})$ to $(\DW{y}{1})$, and therefore a cycle in, the candidate
+execution:
 \begin{gather*}
@@
 \end{gather*}
-An optimizing compiler might lift the assignment $\PW{y}{1}$ out of the conditional,
-thus removing the control dependency.
+A compiler may lift the assignment $\PW{y}{1}$ out of the conditional,
+thus removing the dependency.
 
@@
 \begin{gather*}
+  \tag{$\dagger$}\label{eq1}
   \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
@@
 
-Predicate transformers are a good fit to logical models of dependency calculation,
-since both are concerned with preconditions, and how they are transformed by
-sequential composition. Our first attempt is to associate a predicate transformer
-with each pomset. We visualize this in diagrams by showing how $\bForm$ is transformed,
-for example:
+Predicate transformers are a good fit for logical models of dependency
+calculation, since both are concerned with preconditions and how they are
+transformed by sequential composition. Our first attempt is to associate a
+predicate transformer with each pomset. We visualize this in diagrams by
+showing how $\bForm$ is transformed, for example:
 \begin{align*}
@@
 
-For the sequentially consistent semantics, sequential composition is
+For a sequentially consistent semantics, sequential composition is
 straightforward: we apply each predicate transformer to the preconditions of
@@
 For example, we can recover the expected semantics
-for the above example by choosing
+for \eqref{eq1} by choosing
 the predicate transformer which is independent of $(\DR x0)$
@@
 
-In this paper, we show how pomsets with
-preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} can be combined with
-predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75} to create a
+We show how predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75} can
+be added to pomsets with
+preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} to create a
 compositional semantics for sequential composition.
@@
 % Acknowledgements go here, once we're not double-blinded.
-%The definitions in this paper have been formalized in Agda.
+% The definitions in this paper have been formalized in Agda.
 Because it is closely related, we expect that the memory-model results of
@@
 the semantics of sequential composition in continuation passing style,
-whereas we give it in direct style.
+whereas we give it in direct style.  The use step-indexing to account for
+loops; we expect that the same approach could be applied here.
 

### 2021-04-14 paper/intro.tex
second pass over intro/related
second pass related

@@
 There are prior studies of relaxed memory that include sequential
-composition.  \citet{DBLP:journals/corr/abs-1804-04214} define a semantics
+composition and/or precise calculation of semantic dependencies.
+\citet{DBLP:conf/esop/PaviottiCPWOB20} give a denotational semantics,
+calculating dependencies using event structures rather than logic.  They give
+the semantics of sequential composition in continuation passing style,
+whereas we give it in direct style.
+They use step-indexing to account for
+loops; we expect that the same approach could be applied here.
+%
+\citet{DBLP:journals/corr/abs-1804-04214} define a semantics
 using pomsets without preconditions. Instead, their model uses syntactic
@@
 require a fence after every relaxed read on \armeight{}.
-
-\citet{DBLP:conf/esop/PaviottiCPWOB20} give a denotational semantics,
-calculating dependencies using event structures rather than logic.  They give
-the semantics of sequential composition in continuation passing style,
-whereas we give it in direct style.  The use step-indexing to account for
-loops; we expect that the same approach could be applied here.
-
-\citet{Pichon-Pharabod:2016:CSR:2837614.2837616} also use event structures to
+%
+\citet{Pichon-Pharabod:2016:CSR:2837614.2837616} use event structures to
 calculate dependencies, combined with an operational semantics that
@@
 \cite{Lamport:1979:MMC:1311099.1311750, DBLP:conf/pldi/LahavVKHD17,
-  Dolan:2018:BDR:3192366.3192421, DBLP:conf/pldi/LahavVKHD17,
-  DBLP:conf/lics/JeffreyR16, Boehm:2014:OGA:2618128.2618134,
-  DBLP:journals/corr/abs-1804-04214}.
+  Dolan:2018:BDR:3192366.3192421, DBLP:conf/lics/JeffreyR16,
+  Boehm:2014:OGA:2618128.2618134}. %, DBLP:journals/corr/abs-1804-04214}.
 
@@
 analysis on the value read---effectively, this can turn one read into two,
-with different choices taken for the values read.  Our model invalidates this
-transformation.  In return, our model enjoys compositionality for temporal
-safety properties.
+with different conditional branches taken for the two copies of the read.
+Our model invalidates this transformation.  In return, our model enjoys
+compositionality for temporal safety properties.

### 2021-04-14 paper/arm.tex
pass over discussion and TOC

@@
 \section{Arm}
+\label{sec:arm}
 For simplicity, we restrict to top level parallel composition and ignore

### 2021-04-15 paper/intro.tex
finished through termination

@@
 expect: programs write to, and read from, memory references.
-This can be thought of as a total order of reads and writes,
-where each read has a matching \emph{fulfilling} write,
+This can be thought of as a total order of reads and writes (black arrows),
+where each read has a matching \emph{fulfilling} write (green arrows),
 for example:
@@
 \end{gather*}
+
 % (In examples, $\aReg$--$\bReg$ range over thread-local registers and $\aLoc$-$\cLoc$
@@
       \po{wx0}{wx1}
-      \po[out=-165,in=-15]{rx0}{wx1}
+      \wk[out=-165,in=-15]{rx0}{wx1}
     \end{tikzinline}}
@@
 
-There are prior studies of relaxed memory that include sequential
+There are few prior studies of relaxed memory that include sequential
 composition and/or precise calculation of semantic dependencies.
@@
   Boehm:2014:OGA:2618128.2618134}. %, DBLP:journals/corr/abs-1804-04214}.
+%
+Many of the most prominent models are operational, whole-program models based
+on speculative execution \cite{Manson:2005:JMM:1047659.1040336,
+  DBLP:conf/esop/JagadeesanPR10,
+  DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}.
 
-Many of the most prominent models are based on speculative execution
-\cite{Manson:2005:JMM:1047659.1040336, DBLP:conf/esop/JagadeesanPR10,
-  DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}.
-In their introduction, \citet{DBLP:journals/pacmpl/JagadeesanJR20} note that
-these models fail to validate compositional reasoning of temporal
-properties---see their examples OOTA4 and OOTA5 (from
-\citep{DBLP:journals/toplas/Lochbihler13}).  The difference with our model
-can be understood in terms of the valid program transformations.  The
-speculative models allow reads to be introduced, with subsequent case
-analysis on the value read---effectively, this can turn one read into two,
-with different conditional branches taken for the two copies of the read.
-Our model invalidates this transformation.  In return, our model enjoys
-compositionality for temporal safety properties.
+\citet{DBLP:journals/pacmpl/JagadeesanJR20} note that the speculative models
+listed above all, including \cite{DBLP:conf/popl/KangHLVD17}, fail to
+validate compositional reasoning of temporal properties---see their examples
+OOTA4 and OOTA5 (from \citep{DBLP:journals/toplas/Lochbihler13}).  The
+difference with our model can be understood in terms of the valid program
+transformations.  The speculative models allow reads to be introduced, with
+subsequent case analysis on the value read---effectively, this can turn one
+read into two, with different conditional branches taken for the two copies
+of the read.  Our model invalidates this transformation.  In return, our
+model enjoys compositionality for temporal safety properties.

### 2021-04-16 paper/abstract.tex
second pass over intro/related
Wordsmithing abstract

@@
 % \end{math}
-Program logics and semantics tell us that when executing ((S1; S2), state0),
-we execute (S1, state0) to arrive at state1, then execute (S2, state1) to
-arrive at the final state2.  This is a delightfully simple story that can be
-explained to children.  It is also a lie.
-%
-Processors execute instructions out of order, due to pipelines and caches.
-Compilers reorder programs even more dramatically.  All of this reordering is
-meant to be unobservable in single-threaded code.  In multi-threaded code,
-however, all bets are off.  A formal attempt to understand the resulting mess
-is known as a ``relaxed memory model.''  The relaxed memory models that have
-been proposed to date either fail to address sequential composition, or
-overly restrict processors and compilers.
+Program logics and semantics tell us that when executing $(S_1; S_2)$ starting in state $s_0$,
+we execute $S_1$ in $s_0$ to arrive at $s_1$, then execute $S_2$ in $s_1$ to
+arrive at the final state $s_2$.  This is, of course, an abstraction.  Processors
+execute instructions out of order, due to pipelines and caches, and compilers
+reorder programs even more dramatically.  All of this reordering is meant to
+be unobservable in single-threaded code, but is observable in multi-threaded code.
+A formal attempt to understand the resulting mess is known
+as a ``relaxed memory model.''  The relaxed memory models that have been
+proposed to date either fail to address sequential composition, or overly
+restrict processors and compilers.
 
-To support sequential composition, we propose adding families of predicate
+To support sequential composition while targeing modern hardware, we propose adding families of predicate
 transformers to the existing model of ``Pomsets with Preconditions,'' which
-already supports parallel composition.  When composing (S1;S2), the predicate
-transformer used to validate the precondition of an event in S2 is chosen
-based on the semantic dependencies from S1 into this event.  Our model
+already supports parallel composition.  When composing $(S_1;S_2)$, the predicate
+transformer used to validate the precondition of an event in $S_2$ is chosen
+based on the semantic dependencies from events in $S_1$ into this event in $S_2$.  Our model
 retains the good properties of the prior work, including efficient

### 2021-04-16 paper/intro.tex
Moved realted work
Wordsmithing intro

@@
 program $\aCmd_1\SEMI \aCmd_2$ executes $\aCmd_1$ before $\aCmd_2$.  Since
-the late 60s, we've been able to explain this using logic
+the late 1960s, we've been able to explain this using logic
 \citep{Hoare:1969:ABC:363235.363259}.  In
@@
 postconditions to preconditions.  We recall the definition of
-$\fwp{\aCmd}{\bForm}$ for loop-free code below. 
+$\fwp{\aCmd}{\bForm}$ for loop-free code below (where $\aReg$ ranges over thread-local \emph{registers}
+and $\aExp$ ranges over side-effect free \emph{expressions}). 
 \begin{multicols}{2}
@@
     \stepcounter{enumi}
-  \item[] \labeltext[\textsc{d}2]{}{wp-assign}
-    \begin{enumerate}[leftmargin=0pt]
-    \item \label{wp-write}
-      $\fwp{\PW{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
+  %\item[] \labeltext[\textsc{d}2]{}{wp-assign}
+  %  \begin{enumerate}[leftmargin=0pt]
+  %  \item \label{wp-write}
+  %    $\fwp{\PW{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
     \item \label{wp-let}
       $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
-    \item \label{wp-read}
-      $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \aLoc{=}\aReg\limplies\bForm$ %\bForm[\aLoc/\aReg]$ %
-    \end{enumerate}
+  %  \item \label{wp-read}
+  %    $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \aLoc{=}\aReg\limplies\bForm$ %\bForm[\aLoc/\aReg]$ %
+  %  \end{enumerate}
     % \item 
@@
 $\hoare{\aForm}{\aCmd}{\bForm}$ holds exactly when $\aForm \limplies
-\fwp{\aCmd}{\bForm}$.  We have split
-\citeauthor{DBLP:journals/cacm/Dijkstra75}'s rule for assignment
-\eqref{wp-assign} into three cases. In our notation,
-$\aReg$--$\bReg$ range over thread-local registers, which may be assigned at
-most once,
-$\aLoc$--$\cLoc$ range over shared memory references, and
-$\aExp$--$\bExp$ range over thread-local expressions, which do \emph{not}
-include
-$\aLoc$--$\cLoc$.\footnote{Under these assumptions, \eqref{wp-read} is
-  equivalent to $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \bForm[\aLoc/\aReg]$.}
-
-This is quite a pretty explanation of sequential computation in a sequential
-context.  In a concurrent context, however, \ref{wp-read} is unsound!
-\ref{wp-read} assumes that a read from
-$x$ must be fulfilled by a preceding write to
-$x$.  In a concurrent context, writes may come from other threads.
+\fwp{\aCmd}{\bForm}$.
+%% We have split
+%% \citeauthor{DBLP:journals/cacm/Dijkstra75}'s rule for assignment
+%% \eqref{wp-assign} into three cases. In our notation,
+%% $\aReg$--$\bReg$ range over thread-local registers, which may be assigned at
+%% most once,
+%% $\aLoc$--$\cLoc$ range over shared memory references, and
+%% $\aExp$--$\bExp$ range over thread-local expressions, which do \emph{not}
+%% include
+%% $\aLoc$--$\cLoc$.\footnote{Under these assumptions, \eqref{wp-read} is
+%%   equivalent to $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \bForm[\aLoc/\aReg]$.}
+%%
+This is an elegant explanation of sequential computation in a sequential
+context.  In a concurrent context, however, generalizing
+\ref{wp-let} from thread-local registers to cross-thread memory locations
+is unsound! \ref{wp-let} is sound because a read from a thread-local
+register must be fulfilled by a preceding write
+in the same thread.  In a concurrent context, aread from a cross-thread
+memory location may be fulfilled by a write
+in another thread.
 
@@
 open the question of how to apply logic to racy programs without
-overstraining the implementation.  To understand the solution, one must first
+overconstraining the implementation.  To understand the solution, one must first
 understand the constraints imposed by hardware and compilers.
@@
 \end{align*}
-Then the equivalence holds in any (sequential) context.
+Then the equivalence holds in any
+%% (sequential)
+context.
 
@@
 
-\subsection{Contributions}
-
-We show how predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75} can
-be added to pomsets with
-preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} to create a
-compositional semantics for sequential composition.
-\begin{itemize}
-\item \S\ref{sec:model} presents the basic model, which satisfies many
-  desiderata, but not all.
-\item \S\ref{sec:arm} shows two approaches for efficient implementation on
-  Arm.  The first uses a suboptimal lowering for acquiring reads.  The second
-  uses an optimal lowering, but requires changes to the definitions of
-  parallel and sequential composition.
-\item \S\ref{sec:variants} generalizes the basic semantics of read and write
-  to validate compiler optimizations.
-\end{itemize}
-% Acknowledgements go here, once we're not double-blinded.
-% The definitions in this paper have been formalized in Agda.
-Because it is closely related, we expect that the memory-model results of
-\cite{DBLP:journals/pacmpl/JagadeesanJR20} apply to our model, including
-compositional reasoning for temporal safety properties and {local} \drfsc{}
-as in \cite{Dolan:2018:BDR:3192366.3192421,DBLP:conf/ppopp/DongolJR19,promising-ldrf}.
-% In \textsection\ref{sec:arm}, we provide an alternative proof strategy for
-% efficient compilation to \armeight{}, which improves upon that of
-% \cite{DBLP:journals/pacmpl/JagadeesanJR20} by using a recent alternative
-% characterization of \armeight{}.
-
-\section{Related Work}
+\subsection{Related Work}
 
@@
 model enjoys compositionality for temporal safety properties.
+
+\subsection{Contributions}
+
+We show how predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75} can
+be added to pomsets with
+preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} to create a
+compositional semantics for sequential composition.
+\begin{itemize}
+\item \S\ref{sec:model} presents the basic model, which satisfies many
+  desiderata, but not all.
+\item \S\ref{sec:arm} shows two approaches for efficient implementation on
+  Arm.  The first uses a suboptimal lowering for acquiring reads.  The second
+  uses an optimal lowering, but requires changes to the definitions of
+  parallel and sequential composition.
+\item \S\ref{sec:variants} generalizes the basic semantics of read and write
+  to validate compiler optimizations.
+\end{itemize}
+% Acknowledgements go here, once we're not double-blinded.
+% The definitions in this paper have been formalized in Agda.
+Because it is closely related, we expect that the memory-model results of
+\cite{DBLP:journals/pacmpl/JagadeesanJR20} apply to our model, including
+compositional reasoning for temporal safety properties and {local} \drfsc{}
+as in \cite{Dolan:2018:BDR:3192366.3192421,DBLP:conf/ppopp/DongolJR19,promising-ldrf}.
+% In \textsection\ref{sec:arm}, we provide an alternative proof strategy for
+% efficient compilation to \armeight{}, which improves upon that of
+% \cite{DBLP:journals/pacmpl/JagadeesanJR20} by using a recent alternative
+% characterization of \armeight{}.

### 2021-04-16 paper/intro.tex
intro
repair damage of wordsmithing intro

@@
 postconditions to preconditions.  We recall the definition of
-$\fwp{\aCmd}{\bForm}$ for loop-free code below (where $\aReg$ ranges over thread-local \emph{registers}
-and $\aExp$ ranges over side-effect free \emph{expressions}). 
-\begin{multicols}{2}
+$\fwp{\aCmd}{\bForm}$ for loop-free code below (where $\aReg$--$\bReg$ range over thread-local \emph{registers}
+and $\aExp$--$\bExp$ range over side-effect free \emph{expressions}). 
+%\begin{multicols}{2}
   \begin{enumerate}[,label=(\textsc{d}\arabic*),ref=\textsc{d}\arabic*]
@@
     \end{math}
+  \item \label{wp-let}
+    $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
     % \item 
@@
     %   \end{math}
-    \stepcounter{enumi}
-  %\item[] \labeltext[\textsc{d}2]{}{wp-assign}
+    %   \stepcounter{enumi}
+    % \item[] \labeltext[\textsc{d}2]{}{wp-assign}
     %   \begin{enumerate}[leftmargin=0pt]
@@
     %     $\fwp{\PW{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
-    \item \label{wp-let}
-      $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
+    %   \item \label{wp-let}
+    %     $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
     %   \item \label{wp-read}
@@
     \end{math}
-    \\
     \begin{math}
@@
   \end{enumerate}
-\end{multicols}
+%\end{multicols}
 For this language, the Hoare triple
@@
 This is an elegant explanation of sequential computation in a sequential
-context.  In a concurrent context, however, generalizing
-\ref{wp-let} from thread-local registers to cross-thread memory locations
-is unsound! \ref{wp-let} is sound because a read from a thread-local
-register must be fulfilled by a preceding write
-in the same thread.  In a concurrent context, aread from a cross-thread
-memory location may be fulfilled by a write
-in another thread.
+context. Note that \ref{wp-let} is sound because a read from a thread-local
+register must be fulfilled by a preceding write in the same thread.
+In a concurrent context, with shared variables
+($\aLoc$--$\cLoc$), the obvious generalizations
+\begin{multicols}{2}
+\begin{itemize}
+\item[{\labeltext[\textsc{d}2b]{(\textsc{d}2b)}{wp-write}}] $\fwp{\LET{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
+\item[{\labeltext[\textsc{d}2c]{(\textsc{d}2c)}{wp-read}}] $\fwp{\LET{\aReg}{\aLoc}}{\bForm} = \bForm[\aLoc/\aReg]$
+\end{itemize}
+\end{multicols}
+\noindent
+are unsound!  In particular, a read from a shared memory location may be fulfilled by a write
+in another thread, invalidating \ref{wp-read}.  (We assume that expressions
+% $\aExp$--$\bExp$
+do \emph{not} include shared variables.) % $\aLoc$--$\cLoc$.)
 
@@
 The predicate transformer from the write matches
-\citeauthor{DBLP:journals/cacm/Dijkstra75}.  For the reads, however,
-\ref{wp-read} defines the transformer of $\PR{x}{r}$ to be
-$(x{=}r) \limplies \bForm$.  Instead, we use $(0{=}r) \limplies \bForm$,
-reflecting the fact that $0$ may come from a concurrent write.  The
-obligation to find a matching write is moved from the sequential semantics of
-\emph{substitution} and \emph{implication} to the concurrent semantics of
-\emph{fulfillment}.
+\citeauthor{DBLP:journals/cacm/Dijkstra75}'s \ref{wp-write}.  For the reads,
+however, \ref{wp-read} defines the transformer of $\PR{x}{r}$ to be
+$\bForm[x/r]$, which is equivalent to $(x{=}r) \limplies \bForm$ under the
+assumption that registers are assigned at most once.
+Instead, we use $(0{=}r) \limplies \bForm$, reflecting the fact that $0$ may
+come from a concurrent write.  The obligation to find a matching write is
+moved from the sequential semantics of \emph{substitution} and
+\emph{implication} to the concurrent semantics of \emph{fulfillment}.
 
@@
   uses an optimal lowering, but requires changes to the definitions of
-  parallel and sequential composition.
+  sequential composition.
 \item \S\ref{sec:variants} generalizes the basic semantics of read and write

### 2021-04-16 paper/abstract.tex
Nore wordsmithing

@@
 
-To support sequential composition while targeing modern hardware, we propose adding families of predicate
+To support sequential composition while targeting modern hardware, we propose adding families of predicate
 transformers to the existing model of ``Pomsets with Preconditions,'' which
 already supports parallel composition.  When composing $(S_1;S_2)$, the predicate
-transformer used to validate the precondition of an event in $S_2$ is chosen
-based on the semantic dependencies from events in $S_1$ into this event in $S_2$.  Our model
+transformers used to validate the preconditions of events in $S_2$ are chosen
+based on the semantic dependencies from events in $S_1$ to events in $S_2$.  Our model
 retains the good properties of the prior work, including efficient

### 2021-04-16 paper/intro.tex
stuff

@@
 \citet{DBLP:conf/snapl/MarinoMMNS15} argue that the ``silently shifting
-semicolon'' sufficiently problematic for programmers that concurrent
+semicolon'' is sufficiently problematic for programmers that concurrent
 languages should guarantee sequential abstraction, despite the performance
@@
 whereas we give it in direct style.
-They use step-indexing to account for
-loops; we expect that the same approach could be applied here.
 %
@@
 validate compositional reasoning of temporal properties---see their examples
-OOTA4 and OOTA5 (from \citep{DBLP:journals/toplas/Lochbihler13}).  The
+OOTA4, OOTA5, and \citep[Fig.~8]{DBLP:journals/toplas/Lochbihler13}).  The
 difference with our model can be understood in terms of the valid program
@@
 model enjoys compositionality for temporal safety properties.
+
+We provide a detailed comparison with
+\cite{DBLP:journals/pacmpl/JagadeesanJR20} in \textsection\ref{sec:diff}.
 
@@
   Arm.  The first uses a suboptimal lowering for acquiring reads.  The second
-  uses an optimal lowering, but requires changes to the definitions of
-  sequential composition.
-\item \S\ref{sec:variants} generalizes the basic semantics of read and write
+  uses an optimal lowering, but requires a nontrivial change to the
+  definition of sequential composition.
+\item \S\ref{sec:additional} generalizes the basic semantics of read and write
   to validate compiler optimizations.

### 2021-04-16 paper/arm.tex
finished through termination
updates

@@
 \end{definition}
-\cite{armed} explain \EGC{} and \EC{} using the following example.\footnote{We have
-changed an address dependency in the first thread to a data dependency.}
+\citet{armed} explain \EGC{} and \EC{} using the following example, which is
+allowed by Arm.\footnote{We have changed an address dependency in the first
+  thread to a data dependency.}
 \begin{gather*}
@@
   \\
-  \tag{\cmark\armeight}
+  %\tag{\cmark\armeight}
   \hbox{\begin{tikzinline}[node distance=1.5em]
@@
 
-\citet{DBLP:journals/pacmpl/PodkopaevLV19} lowers to \armeight{} as follows:
-Relaxed access is implemented using \LDR/\STR, non-relaxed access using
-\LDAR/\STLR.  %, acquire and other fences using \DMBLD/\DMBSY.
-In this section, we consider a suboptimal strategy, which lowers non-relaxed
-reads to $(\DMBSY\SEMI\LDAR)$.
-
-We do not distinguish control dependencies from address dependencies, and
+We do not distinguish control dependencies from other dependencies, and
 therefore \ref{lob-rr} forces us to drop all dependencies between reads.  To
@@
 \end{definition}
+
+Even with this small change, the optimal lowering for \armeight{} is unsound
+for our semantics.  The optimal lowering maps relaxed access to \LDR/\STR{} and
+non-relaxed access to \LDAR/\STLR{} \citep{DBLP:journals/pacmpl/PodkopaevLV19}.
+In this section, we consider a suboptimal strategy, which lowers non-relaxed
+reads to $(\DMBSY\SEMI\LDAR)$.  Significantly, we retain the optimal lowering
+for relaxed access.  In the next section we recover the optimal lowering by
+adopting an alternative semantics for \ref{seq-le}.
+
+To see why the optimal lowering fails, consider the following attempted
+execution, where the final values of both $x$ and $y$ are $2$.
+\begin{gather*}
+  %\taglabel{rfi-coe-coe}
+  \PW{x}{2}\SEMI 
+  \PR[\mACQ]{x}{r}\SEMI
+  \PW{y}{r{-}1} \PAR
+  \PW{y}{2}\SEMI
+  \PW[\mREL]{x}{1}
+  %\\
+  %\tag{\cmark\armeight}
+  % \hbox{\begin{tikzinline}[node distance=1.5em]
+  %     \event{a}{\DW{x}{2}}{}
+  %     \raevent{b}{\DR[\mACQ]{x}{2}}{right=of a}
+  %     \event{c}{\DW{y}{1}}{right=of b}
+  %     \event{d}{\DW{y}{2}}{right=2.5em of c}
+  %     \raevent{e}{\DW[\mREL]{x}{1}}{right=of d}
+  %     \rfi{a}{b}
+  %     \bob{b}{c}
+  %     \coe{c}{d}
+  %     \bob{d}{e}
+  %     \coe[out=-165,in=-15]{e}{a}
+  %   \end{tikzinline}}
+  \\
+  \tag{$\rgcb$}
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a}{\DW{x}{2}}{}
+      \raevent{b}{\DR[\mACQ]{x}{2}}{right=of a}
+      \event{c}{\DW{y}{1}}{right=of b}
+      \event{d}{\DW{y}{2}}{right=2.5em of c}
+      \raevent{e}{\DW[\mREL]{x}{1}}{right=of d}
+      \gcbz{a}{b}
+      %\sync{b}{c}
+      \gcbz{c}{d}
+      \gcbz{d}{e}
+      \gcbz[out=-165,in=-15]{e}{a}
+    \end{tikzinline}}
+  \\
+  \tag{$\le$}
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a}{\DW{x}{2}}{}
+      \raevent{b}{\DR[\mACQ]{x}{2}}{right=of a}
+      \event{c}{\DW{y}{1}}{right=of b}
+      \event{d}{\DW{y}{2}}{right=2.5em of c}
+      \raevent{e}{\DW[\mREL]{x}{1}}{right=of d}
+      \rf{a}{b}
+      \sync{b}{c}
+      \wk{c}{d}
+      \sync{d}{e}
+      \wk[out=-165,in=-15]{e}{a}
+    \end{tikzinline}}
+\end{gather*}
+This attempted execution is allowed by \armeight, but disallowed by our
+semantics.
+
+If the read of $x$ in the execution above is changed from acquiring to
+relaxed, then our semantics allows the execution, using the independent case
+for the read and satisfying the precondition of $\DWP{y}{1}$ by prepending
+$\DWP{x}{2}$.  It may be tempting, therefore to adopt a strategy of
+\emph{downgrading} acquires in certain cases.  Unfortunately, it is not
+possible to do this locally without violating important idioms such as
+publication.  For example, consider that $\DRP[\mRA]{x}{1}$ \emph{not} possible for
+the second thread in the following attempted execution, due to publication of
+$\DWP{x}{2}$ via $y$:
+\begin{gather*}
+  \PW{x}{\PR{x}{}+1}\SEMI
+  \PW[\mREL]{y}{1}
+  \PAR
+  \PW{x}{1}\SEMI
+  \IF{\PR[\mACQ]{y}{}\AND\PR[\mACQ]{x}{}}\THEN
+  \PR{z}{s}
+  \FI
+  \PAR
+  \PW{z}{1}\SEMI
+  \PW[\mREL]{x}{1}
+  \\
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{b1}{\DR{x}{1}}{}
+      \event{b2}{\DW{x}{2}}{right=of b1}
+      \raevent{b3}{\DW[\mREL]{y}{1}}{right=of b2}
+      \po{b1}{b2}
+      \sync{b2}{b3}
+      \event{a1}{\DW{x}{1}}{right=3em of b3}
+      \raevent{a2}{\DR[\mACQ]{y}{1}}{right=of a1}
+      \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
+      \event{a4}{\DR{z}{0}}{right=of a3}
+      \sync{a2}{a3}
+      \sync{a3}{a4}
+      \event{c1}{\DW{z}{1}}{right=3em of a4}
+      \raevent{c2}{\DW[\mREL]{x}{1}}{right=of c1}
+      \sync{c1}{c2}
+      \rf[out=-165,in=-15]{a1}{b1}
+      \rf[out=15,in=165]{b3}{a2}
+      \rf[out=-165,in=-15]{c2}{a3}
+      \wk{a4}{c1}
+    \end{tikzinline}}
+\end{gather*}
+Instead, if the read of $x$ is relaxed, then the publication via $y$ fails,
+and $\DRP{x}{1}$ in the second thread is possible.
+\begin{gather*}
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{b1}{\DR{x}{1}}{}
+      \event{b2}{\DW{x}{2}}{right=of b1}
+      \raevent{b3}{\DW[\mREL]{y}{1}}{right=of b2}
+      \po{b1}{b2}
+      \sync{b2}{b3}
+      \event{a1}{\DW{x}{1}}{right=3em of b3}
+      \raevent{a2}{\DR[\mACQ]{y}{1}}{right=of a1}
+      \event{a3}{\DR{x}{1}}{right=of a2}
+      % \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
+      \event{a4}{\DR{z}{0}}{right=of a3}
+      \sync{a2}{a3}
+      \sync[out=15,in=165]{a2}{a4}
+      \event{c1}{\DW{z}{1}}{right=3em of a4}
+      \raevent{c2}{\DW[\mREL]{x}{1}}{right=of c1}
+      \sync{c1}{c2}
+      \rf[out=-165,in=-15]{a1}{b1}
+      \rf[out=15,in=165]{b3}{a2}
+      \rf[out=-165,in=-15]{c2}{a3}
+      \wk{a4}{c1}
+    \end{tikzinline}}
+\end{gather*}
+
+Using the suboptimal lowering for acquiring reads, our semantics is sound for
+Arm.  The proof uses the characterization of Arm using \EGC{}.
 
@@
 
-Downgrading messes up publication:
+\subsection{Arm Compilation 2}
+\label{sec:arm2}
+
+We can achieve optimal lowering for Arm by weakening the semantics of
+sequential composition slightly.  In particular, we must lose
+\reflem{lem:rf:implies:le}, which states that $\bEv\xrfx\aEv$ implies
+$\bEv\le\aEv$.  Revisiting the example in the last subsection, we essentially
+mimic the \EC{} characterization:
 \begin{gather*}
-  \PW{x}{\PR{x}{}+1}\SEMI
-  \PW[\mREL]{y}{1}
-  \PAR
-  \PW{x}{1}\SEMI
-  \IF{\PR[\mACQ]{y}{}\AND\PR[\mACQ]{x}{}}\THEN
-  \PR{z}{s}
-  \FI
-  \PAR
-  \PW{z}{1}\SEMI
+  \PW{x}{2}\SEMI 
+  \PR[\mACQ]{x}{r}\SEMI
+  \PW{y}{r{-}1} \PAR
+  \PW{y}{2}\SEMI
   \PW[\mREL]{x}{1}
   \\
+  \tag{$\rcb$}
   \hbox{\begin{tikzinline}[node distance=1.5em]
-      \event{b1}{\DR{x}{1}}{}
-      \event{b2}{\DW{x}{2}}{right=of b1}
-      \raevent{b3}{\DW[\mREL]{y}{1}}{right=of b2}
-      \po{b1}{b2}
-      \sync{b2}{b3}
-      \event{a1}{\DW{x}{1}}{right=3em of b3}
-      \raevent{a2}{\DR[\mACQ]{y}{1}}{right=of a1}
-      \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
-      \event{a4}{\DR{z}{0}}{right=of a3}
-      \sync{a2}{a3}
-      \sync{a3}{a4}
-      \event{c1}{\DW{z}{1}}{right=3em of a4}
-      \raevent{c2}{\DW[\mREL]{x}{1}}{right=of c1}
-      \sync{c1}{c2}
-      \rf[out=-165,in=-15]{a1}{b1}
-      \rf[out=15,in=165]{b3}{a2}
-      \rf[out=-165,in=-15]{c2}{a3}
-      \wk{a4}{c1}
-    \end{tikzinline}}
-  \\
-  \hbox{\begin{tikzinline}[node distance=1.5em]
-      \event{b1}{\DR{x}{1}}{}
-      \event{b2}{\DW{x}{2}}{right=of b1}
-      \raevent{b3}{\DW[\mREL]{y}{1}}{right=of b2}
-      \po{b1}{b2}
-      \sync{b2}{b3}
-      \event{a1}{\DW{x}{1}}{right=3em of b3}
-      \raevent{a2}{\DR[\mACQ]{y}{1}}{right=of a1}
-      \event{a3}{\DR{x}{1}}{right=of a2}
-      % \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
-      \event{a4}{\DR{z}{0}}{right=of a3}
-      \sync{a2}{a3}
-      \sync[out=15,in=165]{a2}{a4}
-      \event{c1}{\DW{z}{1}}{right=3em of a4}
-      \raevent{c2}{\DW[\mREL]{x}{1}}{right=of c1}
-      \sync{c1}{c2}
-      \rf[out=-165,in=-15]{a1}{b1}
-      \rf[out=15,in=165]{b3}{a2}
-      \rf[out=-165,in=-15]{c2}{a3}
-      \wk{a4}{c1}
+      \event{a}{\DW{x}{2}}{}
+      \raevent{b}{\DR[\mACQ]{x}{2}}{right=of a}
+      \event{c}{\DW{y}{1}}{right=of b}
+      \event{d}{\DW{y}{2}}{right=2.5em of c}
+      \raevent{e}{\DW[\mREL]{x}{1}}{right=of d}
+      \rfint{a}{b}
+      \cbz{b}{c}
+      \cbz{c}{d}
+      \cbz{d}{e}
+      \cbz[out=-165,in=-15]{e}{a}
     \end{tikzinline}}
 \end{gather*}
+Here the $\rrfx$ relation \emph{contradicts} order!  We have both
+$\DWP{x}{2}\xrfint\DRP[\mACQ]{x}{2}$ and
+$\DWP{x}{2}\xcbinv\DRP[\mACQ]{x}{2}$.
 
-\subsection{Arm Compilation 2}
-
+The change to the semantics is small: we weaken relationship between $\rrfx$
+and $\le$ in \ref{seq-le-rf}.  Rather than ensuring that there is no
+\emph{global} blocker for a sequentially fulfilled read \ref{seq-le-rf}, we
+require only that there is no \emph{thread-local} blocker \ref{seq-le-rf-rf}.
+This change both allows and requires us to weaken the definition of
+\emph{delays} to drop write-to-read order from $\eqreorderco$.
 \begin{definition}
   \label{def:sem:frf}
-  Let $\frf{\semrr{}}$ be defined as for $\semrr{}$ in \refdef{def:semrr} and
-  \reffig{fig:sem}, changing \ref{seq-le-rf} and \ref{seq-le-delays}:
+  Let $\frf{\semrr{}}$ be defined as for $\semrr{}$ in
+  \refdef{def:semrr}/\reffig{fig:sem}, changing \ref{seq-le-rf} and
+  \ref{seq-le-delays}:
   \begin{itemize}
@@
 
-Note that \reflem{lem:rf:implies:le} fails for $\frf{\semrr{}}$, since
-$\bEv\xrfx\aEv$ may not imply $\bEv\le\aEv$ when $\bEv$ and $\aEv$ come from
-different sides of a sequential composition.  This means that $\rrfx$ must be
-verified during pomset construction, rather than post-hoc.  If one wants a
-post-hoc verification technique for $\rrfx$, it is possible to include
-program order ($\rpox$) in the pomset.
+With the weakening of \ref{seq-le-rf-rf}, we must be careful not to allow
+spurious pairs to be added to the $\rrfx$ relation.  The use of
+$\rextendsdef{}{}$ in \ref{if-rf-extends} does this, ensuring that no new
+$\rrfx$ is introduced between events in $\aEvs_1\cap\aEvs_2$ when coalescing.
+This is necessary to ensure that
+\begin{math}
+  \frf{\semrr{\IF{b}\THEN\PR{x}{r}\PAR\PW{x}{1}\ELSE\PR{x}{r}\SEMI\PW{x}{1}\FI}}
+\end{math}
+does not include 
+\begin{math}
+  \smash{\hbox{\begin{tikzinlinesmall}[node distance=1.5em]
+        \event{a}{\DR{x}{1}}{}
+        \event{b}{\DW{x}{1}}{right=of a}
+        \rfint[out=165,in=15]{b}{a}
+        \wk{a}{b}
+      \end{tikzinlinesmall}}}
+\end{math}, taking $\rrfx$ from the left and $\le$ from the right.
 
-\begin{example}
-  The obvious definition of $\rpox$ may be cyclic, due to the conditional. 
-\end{example}
+We emphasize that \reflem{lem:rf:implies:le} fails for $\frf{\semrr{}}$,
+since $\bEv\xrfx\aEv$ may not imply $\bEv\le\aEv$ when $\bEv$ and $\aEv$ come
+from different sides of a sequential composition.  This means that $\rrfx$
+must be verified during pomset construction, rather than post-hoc.  The
+following lemma gives a post-hoc verification technique for $\rrfx$, using
+program order ($\rpox$).\footnote{It is obvious how to enhance the semantics
+  of most operators to define $\rpox$.  When combining pomsets using the
+  conditional, the obvious definition of $\rpox$ may result in cycles, since
+  $\rpox$-ordered events may coalesce.  In this case we include a separate
+  pomset for each way of breaking these $\rpox$ cycles.}
+% \begin{example}
+%   The obvious definition of $\rpox$ may be cyclic, due to the conditional. 
+% \end{example}
 \begin{lemma}
-  $\aPS$ is top-level iff $\bEv\xrfx\aEv$ implies either
+  Any $\aPS$ in the image of $\frf{\semrr{}}$ is top-level iff
+  for every $\bEv\xrfx\aEv$ either
   \begin{itemize}
@@
 
-Bad example:
-\begin{gather*}
-  \PEXCHG{x}{r}{2}\SEMI 
-  \PR{x}{s}\SEMI
-  \PW{y}{s{-}1} \PAR
-  \PR{y}{r}\SEMI
-  \PW{x}{r}
-  \\
-  \tag{\cmark\armeight}
-  \hbox{\begin{tikzinline}[node distance=1.5em]
-      \event{a}{\DR{x}{1}}{}
-      \event{b}{\DW{x}{2}}{right=of a}
-      \event{c}{\DR{x}{2}}{right=of b}
-      \event{d}{\DW{y}{1}}{right=of c}
-      \event{e}{\DR{y}{1}}{right=3em of d}
-      \event{f}{\DW{x}{1}}{right=of e}
-      \pre{a}{b}
-      \rf{b}{c}
-      \lob{c}{d}
-      \rf{d}{e}
-      \pre{e}{f}
-      \rf[out=-165,in=-15]{f}{a}
-    \end{tikzinline}}
+% Bad example:
+% \begin{gather*}
+%   \PEXCHG{x}{r}{2}\SEMI 
+%   \PR{x}{s}\SEMI
+%   \PW{y}{s{-}1} \PAR
+%   \PR{y}{r}\SEMI
+%   \PW{x}{r}
+%   \\
+%   \tag{\cmark\armeight}
+%   \hbox{\begin{tikzinline}[node distance=1.5em]
+%       \event{a}{\DR{x}{1}}{}
+%       \event{b}{\DW{x}{2}}{right=of a}
+%       \event{c}{\DR{x}{2}}{right=of b}
+%       \event{d}{\DW{y}{1}}{right=of c}
+%       \event{e}{\DR{y}{1}}{right=3em of d}
+%       \event{f}{\DW{x}{1}}{right=of e}
+%       \pre{a}{b}
+%       \rf{b}{c}
+%       \lob{c}{d}
+%       \rf{d}{e}
+%       \pre{e}{f}
+%       \rf[out=-165,in=-15]{f}{a}
+%     \end{tikzinline}}
+%   % \hbox{\begin{tikzinline}[node distance=1.5em]
+%   %   \event{a}{\DR{x}{1}}{}
+%   %   \event{b}{\DW{x}{2}}{right=of a}
+%   %   \event{c}{\DR{x}{2}}{right=of b}
+%   %   \event{d}{\DW{y}{1}}{right=of c}
+%   %   \event{e}{\DR{y}{1}}{right=3em of d}
+%   %   \event{f}{\DW{x}{1}}{right=of e}
+%   %   \rmw{a}{b}
+%   %   \rfi{b}{c}
+%   %   \dob{c}{d}
+%   %   \rfe{d}{e}
+%   %   \dob{e}{f}
+%   %   \rfe[out=-165,in=-15]{f}{a}
+%   % \end{tikzinline}}
+%   \\
+%   \tag{$\le$}
 %   \hbox{\begin{tikzinline}[node distance=1.5em]
@@
 %       \rmw{a}{b}
-  %   \rfi{b}{c}
-  %   \dob{c}{d}
-  %   \rfe{d}{e}
-  %   \dob{e}{f}
-  %   \rfe[out=-165,in=-15]{f}{a}
+%       \rf{b}{c}
+%       % \po{c}{d}
+%       \rf{d}{e}
+%       \po{e}{f}
+%       \rf[out=-165,in=-15]{f}{a}
 %     \end{tikzinline}}
-  \\
-  \tag{$\le$}
-  \hbox{\begin{tikzinline}[node distance=1.5em]
-      \event{a}{\DR{x}{1}}{}
-      \event{b}{\DW{x}{2}}{right=of a}
-      \event{c}{\DR{x}{2}}{right=of b}
-      \event{d}{\DW{y}{1}}{right=of c}
-      \event{e}{\DR{y}{1}}{right=3em of d}
-      \event{f}{\DW{x}{1}}{right=of e}
-      \rmw{a}{b}
-      \rf{b}{c}
-      % \po{c}{d}
-      \rf{d}{e}
-      \po{e}{f}
-      \rf[out=-165,in=-15]{f}{a}
-    \end{tikzinline}}
-\end{gather*}
-Anton example 1 \texttt{[rfi-coe-coe]}
-\begin{gather*}
-  \taglabel{rfi-coe-coe}
-  \PW{x}{2}\SEMI 
-  \PR[\mACQ]{x}{r}\SEMI
-  \PW{y}{1} \PAR
-  \PW{y}{2}\SEMI
-  \PW[\mREL]{x}{1}
-  \\
-  \tag{\cmark\armeight}
-  \hbox{\begin{tikzinline}[node distance=1.5em]
-      \event{a}{\DW{x}{2}}{}
-      \raevent{b}{\DR[\mACQ]{x}{2}}{right=of a}
-      \event{c}{\DW{y}{1}}{right=of b}
-      \event{d}{\DW{y}{2}}{right=2.5em of c}
-      \raevent{e}{\DW[\mREL]{x}{1}}{right=of d}
-      \rfi{a}{b}
-      \bob{b}{c}
-      \coe{c}{d}
-      \bob{d}{e}
-      \coe[out=-165,in=-15]{e}{a}
-    \end{tikzinline}}
-  \\
-  \hbox{\begin{tikzinline}[node distance=1.5em]
-      \event{a}{\DW{x}{2}}{}
-      \event{b}{\DR{x}{2}}{right=of a}
-      \event{c}{\DW{y}{1}}{right=of b}
-      \event{d}{\DW{y}{2}}{right=2.5em of c}
-      \raevent{e}{\DW[\mREL]{x}{1}}{right=of d}
-      \rf{a}{b}
-      \wk{c}{d}
-      \sync{d}{e}
-      \wk[out=-165,in=-15]{e}{a}
-    \end{tikzinline}}
-\end{gather*}
+% \end{gather*}

### 2021-04-16 paper/outro.tex
semantics in figure
updates

--- 2021-01-25 paper/outro.tex
@@
-\section{Conclusions}
+\section{Future Work}
 
-We have presented the first model of relaxed memory that treats sequential
-composition as a first-class citizen. The model builds directly on \jjr{}.
+\citet{DBLP:conf/esop/PaviottiCPWOB20} use step-indexing to account for
+loops; we expect that the same approach could be applied here.
 
-For sequential composition, parallel composition and the conditional, we
-believe that the definition is \emph{natural}, even \emph{canonical}.
-For stores and loads, instead, the definition in Figure~\ref{fig:full} is a
-Frankenstein's monster of features.  This complexity is \emph{essential},
-however, not an accident of our poor choices.  Relaxed memory models must
-please many audiences: compiler writers want one thing, hardware designers
-another, and programmers yet another still.  The result is inevitably full of
-compromise.
+% We have presented the first model of relaxed memory that treats sequential
+% composition as a first-class citizen. The model builds directly on \jjr{}.
 
-Given that \emph{complexity} cannot be eliminated from relaxed memory models, the
-best one can do is attempt to understand its causes.  We have broken the
-problem into eight manageable pieces, discussed throughout
-\textsection\ref{sec:co}--\ref{sec:complications}.
+% For sequential composition, parallel composition and the conditional, we
+% believe that the definition is \emph{natural}, even \emph{canonical}.
+% For stores and loads, instead, the definition in \reffig{fig:no-addr} is a
+% Frankenstein's monster of features.  This complexity is \emph{essential},
+% however, not just an accident of our poor choices.  Relaxed memory models must
+% please many audiences: compiler writers want one thing, hardware designers
+% another, and programmers yet another still.  The result is inevitably full of
+% compromise.
 
-{Logic} is the thread that sews these features together.
+% Given that \emph{complexity} cannot be eliminated from relaxed memory models,
+% the best one can do is attempt to understand its causes.  We have broken the
+% problem into seven manageable pieces, discussed throughout
+% \textsection\ref{sec:q}--\ref{sec:complications}.  \refdef{def:pomsets-arm}
+% summarizes all the features necessary for efficient implementation on
+% \armeight{}.  We discuss address calculation, read-modify-write operations
+% and fences in the appendix.
 
-%A unique feature of our model is the centrality of logic:
+% {Logic} is the thread that sews these features together.
 
-% study each feature in isolation, as a small delta to the base definition
-% given in \textsection\ref{sec:model}.
-% In we studied eight such
-% features.
+% %A unique feature of our model is the centrality of logic:
 
-% seem staggeringly complex, we argue that this complexity is unavoidable if
-% one wants all the features that it embodies.  By breaking the definition into
-% its constituent parts, we have shown how each of eight
+% % study each feature in isolation, as a small delta to the base definition
+% % given in \textsection\ref{sec:model}.
+% % In we studied eight such
+% % features.
+
+% % seem staggeringly complex, we argue that this complexity is unavoidable if
+% % one wants all the features that it embodies.  By breaking the definition into
+% % its constituent parts, we have shown how each of eight
+
+% \subsection*{Acknowledgements}
+% % This paper has been greatly improved by the comments of the anonymous reviewers.
+% Riely was supported by the National Science Foundation under
+% grant No.~CCR-1617175.
+
+% % It is based upon work supported by the National Science Foundation under
+% % Grant No. CCR-1617175. Any opinions, findings, and conclusions or
+% % recommendations expressed in this material are those of the author and do
+% % not necessarily reflect the views of the NSF.}

### 2021-04-17 paper/arm.tex
no dangling refs

@@
 \begin{theorem}
-  \label{thm:ec}
+  \label{thm:egc}
   Suppose $\aEG_1$ is $({\rco_1}, {\rrfx_1}, {\rgcb_1})$-valid for $\aCmd$
@@
     \begin{lemma}
-      \label{lemma:fr}
+      \label{lemma:fr1}
       Suppose $\aEG$ is $({\rco}, {\rrfx}, {\rgcb})$-valid.  Then
@@
     that the order removed from \ref{arm-gcb-lob} can also be removed from
-    the pomset.  In order for \label{arm-gcb-lob} to remove order from $\aEv$
+    the pomset.  In order for \ref{arm-gcb-lob} to remove order from $\aEv$
     to $\cEv$, we must have $\bEv\xrfx\aEv$ and $\bEv\xpoloc\aEv$ but not
@@
 \begin{lemma}
-  \label{lemma:fr}
+  \label{lemma:fr2}
   Suppose $\aEG$ is \EC-valid via $({\rco}, {\rrfx}, {\rcb})$.  Then there a

### 2021-04-17 paper/oopsla.tex
aligning tex and agda
no dangling refs

--- 2021-01-08 paper/appendix.tex
@@
-\appendix
-% \appendices
-% \section{}
+\section{Differences with ``Pomsets with Preconditions''}
+\label{sec:diff}
 
-This paper addresses several errors in \citep{10.1145/3428262}, which we
-henceforth refer to as \jjr{}.
+\myparagraph{Substitution}
+
+\jjr{} uses substitution rather than Skolemizing.  Indeed our use of
+Skolemization is motivated by disjunction closure for predicate transformers,
+which do not appear in \jjr{}.  In \reffig{fig:sem}, 
+we gave the semantics of read for nonempty pomsets as:
+\begin{enumerate}
+\item[{\labeltext[\textsc{r}4a]{(\textsc{r}4a)}{read-tau-dep-oopsla}}]
+  if $(\aEvs\cap\bEvs)\neq\emptyset$ then
+  \begin{math}
+    \aTr{\bEvs}{\bForm} \rimplies
+    \aVal{=}\aReg
+    \limplies \bForm
+  \end{math},    
+\item[{\labeltext[\textsc{r}4b]{(\textsc{r}4b)}{read-tau-ind-oopsla}}]
+  if $(\aEvs\cap\bEvs)=\emptyset$ then
+  \begin{math}
+    \aTr{\bEvs}{\bForm} \rimplies
+    \PBR{\aVal{=}\aReg \lor \aLoc{=}\aReg} \limplies
+    \bForm.
+  \end{math}
+\end{enumerate}
+In \jjr{}, the definition is roughly as follows:
+% (adding the case for $\ref{L6}$, which was missing):
+\begin{enumerate}
+\item[{\labeltext[\textsc{r}4a$'$]{(\textsc{r}4a$'$)}{read-tau-dep-oopsla-sub}}]
+  if $(\aEvs\cap\bEvs)\neq\emptyset$ then
+  \begin{math}
+    \aTr{\bEvs}{\bForm} \rimplies
+    \bForm[\aVal/\aReg][\aVal/\aLoc]
+    % \aVal{=}\aReg
+    % \limplies \bForm[\aReg/\aLoc]
+  \end{math},    
+\item[{\labeltext[\textsc{r}4b$'$]{(\textsc{r}4b$'$)}{read-tau-ind-oopsla-sub}}]
+  if $(\aEvs\cap\bEvs)=\emptyset$ then
+  \begin{math}
+    \aTr{\bEvs}{\bForm} \rimplies
+    \bForm[\aVal/\aReg][\aVal/\aLoc]\land\bForm[\aLoc/\aReg]
+  \end{math}
+\end{enumerate}
+The use of conjunction in \ref{read-tau-ind-oopsla-sub} causes disjunction closure to fail
+because the predicate transformer
+% $\aTr{}{\bForm}=\bForm[\aVal/\aReg][\aVal/\aLoc]\land\bForm[\aLoc/\aReg]$ does not distribute through
+% disjunction:
+% \begin{math}
+%   \aTr{}{\bForm_1\lor \bForm_2}=
+%   (\bForm_1\lor \bForm_2)[\aVal/\aReg][\aVal/\aLoc]\land(\bForm_1\lor \bForm_2)[\aLoc/\aReg]
+%   \neq
+%   (\bForm_1[\aVal/\aReg][\aVal/\aLoc]\land\bForm_1[\aLoc/\aReg]) \lor
+%   (\bForm_2[\aVal/\aReg][\aVal/\aLoc]\land\bForm_2[\aLoc/\aReg])
+%   = \aTr{}{\bForm_1} \lor \aTr{}{\bForm_2}
+% \end{math}
+$\aTr{}{\bForm}=\bForm'\land\bForm''$ does not distribute through
+disjunction, even assuming that the prime operations do:\footnote{%
+  \begin{math}
+    (\bForm_1\lor \bForm_2)'=(\bForm_1'\lor \bForm_2')
+  \end{math}
+  and
+  \begin{math}
+    (\bForm_1\lor \bForm_2)''=(\bForm_1''\lor \bForm_2'')
+  \end{math}.
+}
+\begin{math}
+  \aTr{}{\bForm_1\lor \bForm_2}=
+  \href{https://www.wolframalpha.com/input/?i=\%28a+or+b\%29+and+\%28c+or+d\%29}{(\bForm_1'\lor \bForm_2')\land(\bForm_1''\lor \bForm_2'')}
+  \neq
+  \href{https://www.wolframalpha.com/input/?i=\%28a+and+c\%29+or+\%28b+and+d\%29}{(\bForm_1'\land\bForm_1'') \lor (\bForm_2'\land\bForm_2'')}
+  = \aTr{}{\bForm_1} \lor \aTr{}{\bForm_2}
+\end{math}.
+% \begin{math}
+%   (\bForm_{1}^{1}\lor \bForm_{1}^{2}) \land (\bForm_{2}^{1}\lor \bForm_{2}^{2})
+%   \neq
+%   (\bForm_{1}^{1}\land\bForm_{2}^{1}) \lor (\bForm_{1}^{2}\land\bForm_{1}^{2}).
+% \end{math}
+See also \textsection\ref{sec:ex:assoc}.
+
+The substitutions collapse $\aLoc$ and $\aReg$, allowing local invariant
+reasoning, causality test case 1, discussed at the end of
+\textsection\ref{sec:ex:control}.  Without Skolemizing it is necessary to
+substitute $[\aLoc/\aReg]$, since the reverse substitution $[\aReg/\aLoc]$ is
+useless when $\aReg$ is bound---compare with
+\textsection\ref{sec:substitutions}.  As discussed below (\ref{p:downset}),
+including this substitution affects the interaction of \xLIR{} and downset
+closure.
+
+Removing the substitution of $[x/r]$ in the independent case has a technical
+advantage: we no longer require \emph{extended} expressions (which include
+memory references), since substitutions no longer introduce memory
+references.
+
+\begin{scope}
+  The substitution $[x/r]$ does not work with Skolemization, even for the
+  dependent case, since we lose the unique marker for each read.  In effect,
+  this forces the reads to the same values.
+  % To be concrete, the candidate
+  % definition would modify \ref{L4} to be:
+  % \begin{enumerate}
+  % \item[\ref{L4})]
+  %   $\aTr{\bEvs}{\bForm} \rimplies \aVal{=}\aLoc\limplies\bForm[\aLoc/\aReg]$.
+  %   % \item[\ref{L5})]
+  %   %   $\aTr{\cEvs}{\bForm} \rimplies (\aVal{=}\aLoc\lor\TRUE)\limplies\bForm[\aLoc/\aReg]$. %, when $\aEvs\neq\emptyset$,
+  %   % \item[\ref{L6})] 
+  %   %   $\aTr{\dEvs}{\bForm}\; \rimplies \bForm$, when $\aEvs=\emptyset$.
+  % \end{enumerate}
+  Using this definition, consider the following:
+  \begin{gather*}
+    \PR{x}{r}\SEMI
+    \PR{x}{s}\SEMI
+    \IF{r{<}s}\THEN \PW{y}{1}\FI 
+    \\[-1ex]
+    \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
+        \event{a1}{\DR{x}{1}}{}
+        \event{a2}{\DR{x}{2}}{right=of a1}
+        \event{a3}{1{=}x\limplies 2{=}x\limplies x{<} x\mid\DW{y}{1}}{right=of a2}
+        \po[out=20,in=160]{a1}{a3}
+        \po{a2}{a3}
+      \end{tikzinline}}
+  \end{gather*}
+  Although the execution seems reasonable, the precondition on the write is
+  not a tautology.
+\end{scope}
+
+% There, item \ref{loadpre-kappa2}  of $\sLOADPRE{}{}{}$ is written 
+% \begin{enumerate}
+% \item[] %[\ref{loadpre-kappa2})]
+%   if $\aEv\in\aEvs_2\setminus\aEvs_1$ then either \\
+%   $\labelingForm(\aEv) \rimplies \labelingForm_2(\aEv)[\aLoc/\aReg][\aVal/\aLoc]$ and $(\exists\bEv\in\aEvs_1)\bEv{<}\aEv$, or \\
+%   $\labelingForm(\aEv) \rimplies \labelingForm_2(\aEv)[\aLoc/\aReg][\aVal/\aLoc] \land \labelingForm_2(\aEv)[\aLoc/\aReg]$.
+% \end{enumerate}
+
+% [Skolemization ensures disjunction closure, which is necessary
+% for associativity. Show example.]
+
+\myparagraph[p:downset]{Downset closure}
+
+\jjr{} enforces downset closure in the prefixing rule.  Even without this,
+downset closure would be different for the two semantics, due to the use of
+substitution in \jjr{}.  Consider the final pomset in the last example of
+\textsection\ref{sec:downset} under the semantics of this paper, which elides
+the middle read event:
+\begin{align*}
+  \begin{gathered}[t]
+    \PW{x}{0} 
+    \SEMI\PR{x}{r} 
+    \SEMI\IF{r{\geq}0}\THEN \PW{y}{1} \FI
+    \\
+    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
+        \event{a0}{\DW{x}{0}}{}
+        % \event{a1}{\DR{x}{1}}{right=of a0}
+        \event{a2}{r{\geq}0\mid\DW{y}{1}}{right=3em of a1}      
+        % \wk{a0}{a1}
+      \end{tikzinline}}    
+  \end{gathered}
+\end{align*}
+In \jjr{}, the substitution $[x/r]$ is performed by the middle read
+regardless of whether it is included in the pomset, with the subsequent
+substitution of $[0/x]$ by the preceding write, we have $[x/r][0/x]$, which
+is $[0/r][0/x]$, resulting in:
+\begin{align*}
+  \begin{gathered}[t]
+    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
+        \event{a0}{\DW{x}{0}}{}
+        % \event{a1}{\DR{x}{1}}{right=of a0}
+        \event{a2}{0{\geq}0\mid\DW{y}{1}}{right=3em of a1}      
+        % \wk{a0}{a1}
+      \end{tikzinline}}    
+  \end{gathered}
+\end{align*}
+
+\myparagraph{Consistency}
+\jjr{} imposes \emph{consistency}, which requires that for every pomset
+$\aPS$, $\bigwedge_{\aEv}\labelingForm(\aEv)$ is satisfiable.  
+\begin{scope}
+  Associativity requires that we allow pomsets with inconsistent
+  preconditions.  Consider a variant of the example from \textsection\ref{sec:semca}.
+  \begin{scope}
+    \footnotesize
+    \begin{align*}
+      \begin{gathered}
+        \IF{\aExp}\THEN\PW{x}{1}\FI
+        \\
+        \hbox{\begin{tikzinline}[node distance=1em]
+            \event{a}{\aExp\mid\DW{x}{1}}{}
+          \end{tikzinline}}
+      \end{gathered}
+      &&
+      \begin{gathered}
+        \IF{\BANG\aExp}\THEN\PW{x}{1}\FI
+        \\
+        \hbox{\begin{tikzinline}[node distance=1em]
+            \event{a}{\lnot\aExp\mid\DW{x}{1}}{}
+          \end{tikzinline}}
+      \end{gathered}
+      &&
+      \begin{gathered}
+        \IF{\aExp}\THEN\PW{y}{1}\FI
+        \\
+        \hbox{\begin{tikzinline}[node distance=1em]
+            \event{a}{\aExp\mid\DW{y}{1}}{}
+          \end{tikzinline}}
+      \end{gathered}
+      &&
+      \begin{gathered}
+        \IF{\BANG\aExp}\THEN\PW{y}{1}\FI
+        \\
+        \hbox{\begin{tikzinline}[node distance=1em]
+            \event{a}{\lnot\aExp\mid\DW{y}{1}}{}
+          \end{tikzinline}}
+      \end{gathered}
+    \end{align*}
+  \end{scope}
+  Associating left and right, we have:
+  \begin{scope}
+    \footnotesize
+    \begin{align*}
+      \begin{gathered}
+        \IF{\aExp}\THEN\PW{x}{1}\FI
+        \SEMI
+        \IF{\BANG\aExp}\THEN\PW{x}{1}\FI
+        \\
+        \hbox{\begin{tikzinline}[node distance=1em]
+            \event{a}{\DW{x}{1}}{}
+          \end{tikzinline}}
+      \end{gathered}
+      &&
+      \begin{gathered}
+        \IF{\aExp}\THEN\PW{y}{1}\FI
+        \SEMI
+        \IF{\BANG\aExp}\THEN\PW{y}{1}\FI
+        \\
+        \hbox{\begin{tikzinline}[node distance=1em]
+            \event{a}{\DW{y}{1}}{}
+          \end{tikzinline}}
+      \end{gathered}
+    \end{align*}
+  \end{scope}  
+  Associating into the middle, instead, we require:
+  \begin{scope}
+    \footnotesize
+    \begin{align*}
+      \begin{gathered}
+        \IF{\aExp}\THEN\PW{x}{1}\FI
+        \\
+        \hbox{\begin{tikzinline}[node distance=1em]
+            \event{a}{\aExp\mid\DW{x}{1}}{}
+          \end{tikzinline}}
+      \end{gathered}
+      &&
+      \begin{gathered}
+        \IF{\BANG\aExp}\THEN\PW{x}{1}\FI
+        \SEMI
+        \IF{\aExp}\THEN\PW{y}{1}\FI
+        \\
+        \hbox{\begin{tikzinline}[node distance=1em]
+            \event{a}{\lnot\aExp\mid\DW{x}{1}}{}
+            \event{b}{\aExp\mid\DW{y}{1}}{right=of a}
+          \end{tikzinline}}
+      \end{gathered}
+      &&
+      \begin{gathered}
+        \IF{\BANG\aExp}\THEN\PW{y}{1}\FI
+        \\
+        \hbox{\begin{tikzinline}[node distance=1em]
+            \event{a}{\lnot\aExp\mid\DW{y}{1}}{}
+          \end{tikzinline}}
+      \end{gathered}
+    \end{align*}
+  \end{scope}
+  Joining left and right, we have:
+  \begin{scope}
+    \footnotesize
+    \begin{align*}
+      \begin{gathered}
+        \IF{\aExp}\THEN\PW{x}{1}\FI
+        \SEMI
+        \IF{\BANG\aExp}\THEN\PW{x}{1}\FI
+        \SEMI
+        \IF{\aExp}\THEN\PW{y}{1}\FI
+        \SEMI
+        \IF{\BANG\aExp}\THEN\PW{y}{1}\FI
+        \\
+        \hbox{\begin{tikzinline}[node distance=1em]
+            \event{a}{\DW{x}{1}}{}
+            \event{b}{\DW{y}{1}}{right=of a}
+          \end{tikzinline}}
+      \end{gathered}
+    \end{align*}
+  \end{scope}  
+\end{scope}
+
+\myparagraph{Causal Strengthening}
+\labeltext[]{Causal Strengthening}{xCausal}
+\jjr{} imposes \emph{causal strengthening}, which requires for every pomset
+$\aPS$, if $\bEv\le\aEv$ then $\labelingForm(\aEv) \rimplies \labelingForm(\bEv)$. 
+\begin{scope}
+  Associativity requires that we allow pomsets without causal strengthening.
+  Consider the following.
+  \begin{align*}
+    \begin{gathered}
+      \IF{\aExp}\THEN\PR{x}{r}\FI
+      \\
+      \hbox{\begin{tikzinline}[node distance=1em]
+          \event{a}{\aExp\mid\DR{x}{1}}{}
+        \end{tikzinline}}
+    \end{gathered}
+    &&
+    \begin{gathered}
+      \PW{y}{r}
+      \\
+      \hbox{\begin{tikzinline}[node distance=1em]
+          \event{a}{r{=}1\mid\DW{y}{1}}{}
+        \end{tikzinline}}
+    \end{gathered}
+    &&
+    \begin{gathered}
+      \IF{\BANG\aExp}\THEN\PR{x}{s}\FI
+      \\
+      \hbox{\begin{tikzinline}[node distance=1em]
+          \event{a}{\lnot\aExp\mid\DR{x}{1}}{}
+        \end{tikzinline}}
+    \end{gathered}
+  \end{align*}
+  Associating left, with causal strengthening:
+  \begin{align*}
+    \begin{gathered}
+      \IF{\aExp}\THEN\PR{x}{r}\FI
+      \SEMI
+      \PW{y}{r}
+      \\
+      \hbox{\begin{tikzinline}[node distance=1em]
+          \event{a}{\aExp\mid\DR{x}{1}}{}
+          \event{b}{\aExp\mid\DW{y}{1}}{right=of a}
+          \po{a}{b}
+        \end{tikzinline}}
+    \end{gathered}
+    &&
+    \begin{gathered}
+      \IF{\BANG\aExp}\THEN\PR{x}{s}\FI
+      \\
+      \hbox{\begin{tikzinline}[node distance=1em]
+          \event{a}{\lnot\aExp\mid\DR{x}{1}}{}
+        \end{tikzinline}}
+    \end{gathered}
+  \end{align*}
+  Finally, merging:
+  \begin{align*}
+    \begin{gathered}
+      \IF{\aExp}\THEN\PR{x}{r}\FI
+      \SEMI
+      \PW{y}{r}
+      \SEMI
+      \IF{\BANG\aExp}\THEN\PR{x}{s}\FI
+      \\
+      \hbox{\begin{tikzinline}[node distance=1em]
+          \event{a}{\DR{x}{1}}{}
+          \event{b}{\aExp\mid\DW{y}{1}}{right=of a}
+          \po{a}{b}
+        \end{tikzinline}}
+    \end{gathered}
+  \end{align*}
+  Instead, associating right:
+  \begin{align*}
+    \begin{gathered}
+      \IF{\aExp}\THEN\PR{x}{r}\FI
+      \\
+      \hbox{\begin{tikzinline}[node distance=1em]
+          \event{a}{\aExp\mid\DR{x}{1}}{}
+        \end{tikzinline}}
+    \end{gathered}
+    &&
+    \begin{gathered}
+      \PW{y}{r}
+      \SEMI
+      \IF{\BANG\aExp}\THEN\PR{x}{s}\FI
+      \\
+      \hbox{\begin{tikzinline}[node distance=1em]
+          \event{a}{\lnot\aExp\mid\DR{x}{1}}{}
+          \event{b}{r{=}1\mid\DW{y}{1}}{left=of a}
+        \end{tikzinline}}
+    \end{gathered}
+  \end{align*}
+  Merging:
+  \begin{align*}
+    \begin{gathered}
+      \IF{\aExp}\THEN\PR{x}{r}\FI
+      \SEMI
+      \PW{y}{r}
+      \SEMI
+      \IF{\BANG\aExp}\THEN\PR{x}{s}\FI
+      \\
+      \hbox{\begin{tikzinline}[node distance=1em]
+          \event{a}{\DR{x}{1}}{}
+          \event{b}{\DW{y}{1}}{right=of a}
+          \po{a}{b}
+        \end{tikzinline}}
+    \end{gathered}
+  \end{align*}
+  With causal strengthening, the precondition of $\DW{y}{1}$ depends upon how
+  we associate.  This is not an issue in \jjr{}, which always associates to
+  the right.
+\end{scope}
+
+% \myparagraph{Causal Strengthening and Address Dependencies}
+% \labeltext[]{Causal Strengthening and Address Dependencies}{xADDRxRRD}
+
+\begin{scope}  
+  One use of causal strengthening is to ensure that address dependencies do
+  not introduce thin air reads.  Associating to the right, the intermediate
+  state of the example in \textsection\ref{sec:addr} is:
+  \begin{align*}
+    \begin{gathered}[t]
+      \PR{\REF{r}}{s}
+      \SEMI
+      \PW{x}{s}
+      \\
+      \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
+          \event{a2}{r\EQ2\mid\DR{\REF{2}}{1}}{}
+          \event{a3}{(r\EQ2\limplies 1\EQ s) \limplies s\EQ1\mid\DW{x}{1}}{right=of a2}
+          \po{a2}{a3}
+        \end{tikzinline}}
+    \end{gathered}
+  \end{align*}
+  In \jjr{}, we have, instead:
+  \begin{gather*}
+    % \begin{gathered}[t]
+    %   \PW{x}{s}
+    %   \\
+    %   \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
+    %     \event{b}{s\EQ1\mid\DW{x}{1}}{}
+    %   \end{tikzinline}}
+    % \end{gathered}
+    % \\
+    \begin{gathered}
+      % \PR{y}{r}\SEMI
+      \PR{\REF{r}}{s}\SEMI \PW{x}{s}
+      \\
+      \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
+          % \event{a1}{\DR{y}{2}}{}
+          \event{a2}{r\EQ2\mid\DR{\REF{2}}{1}}{}%right=of a1}
+          \event{a3}{r\EQ2\land\REF{2}\EQ1\mid\DW{x}{1}}{right=of a2}
+          \po{a2}{a3}
+        \end{tikzinline}}
+    \end{gathered}
+  \end{gather*}
+  Without causal strengthening, the precondition of $\DWP{x}{1}$ would be
+  simply $\REF{2}\EQ1$.  The treatment in this paper, using implication
+  rather than conjunction, is more precise.
+\end{scope}
+
+\myparagraph{Parallel Composition}
 
@@
 
+\myparagraph{Internal Acquiring Reads}
+
+The proof of compilation to Arm in \jjr assumes that all internal reads can
+be eliminated.
+% Shortly after publication, \citet{anton}
+% noticed a shortcoming of the implementation on \armeight{} in
+% \jjr{\textsection 7}.  The proof given there assumes that all internal reads
+% can be dropped.
+However, this is not the case for acquiring reds.  For example, \jjr{}
+disallows the following execution, which is allowed by \armeight{} and
+\tso{}.
+\begin{gather*}
+  \PW{x}{2}\SEMI 
+  \PR[\mACQ]{x}{r}\SEMI
+  \PR{y}{s} \PAR
+  \PW{y}{2}\SEMI
+  \PW[\mREL]{x}{1}
+  \\
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a}{\DW{x}{2}}{}
+      \raevent{b}{\DR[\mACQ]{x}{2}}{right=of a}
+      \event{c}{\DR{y}{0}}{right=of b}
+      \event{d}{\DW{y}{2}}{right=2.5em of c}
+      \raevent{e}{\DW[\mREL]{x}{1}}{right=of d}
+      \rf{a}{b}
+      \sync{b}{c}
+      \wk{c}{d}
+      \sync{d}{e}
+      \wk[out=-165,in=-15]{e}{a}
+      % \rfi{a}{b}
+      % \bob{b}{c}
+      % \fre{c}{d}
+      % \bob{d}{e}
+      % \coe[out=-165,in=-15]{e}{a}
+    \end{tikzinline}}
+\end{gather*}
+We discussed our solution in \textsection\ref{sec:arm}.
+% The solution we have adopted is to allow an acquiring read to be downgraded
+% to a relaxed read when it is preceded (sequentially) by a relaxed write that
+% could fulfill it.  This solution allows executions that are not allowed under
+% \armeight{} since we do not insist that the local relaxed write is actually
+% read from.  This may seem counterintuitive, but we don't see a local way to
+% be more precise.
+
+% As a result, we use a different proof strategy for \armeight{}
+% implementation, which does not rely on read elimination.  The proof idea uses
+% a recent alternative characterization of \armeight{}
+% \citep{alglave-git-alternate,arm-reference-manual}. %,armed-cats}.
+
+\myparagraph{Redundant Read Elimination}
+
+Contrary to the claim, redundant read elimination fails for \jjr{}.
+We discussed redundant read elimination in \textsection\ref{sec:semreg}.
+Consider JMM Causality Test Case 2, which we discussed there.
+\begin{gather*}
+  \PR{x}{r}\SEMI
+  \PR{x}{s}\SEMI
+  \IF{r{=}s}\THEN \PW{y}{1}\FI
+  \PAR
+  \PW{x}{y}
+  \\
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a1}{\DR{x}{1}}{}
+      \event{a2}{\DR{x}{1}}{right=of a1}
+      \event{a3}{\DW{y}{1}}{right=of a2}
+      \event{b1}{\DR{y}{1}}{right=3em of a3}
+      \event{b2}{\DW{x}{1}}{right=of b1}
+      \rf{a3}{b1}
+      \po{b1}{b2}
+      \rf[out=169,in=11]{b2}{a2}
+      \rf[out=169,in=11]{b2}{a1}
+    \end{tikzinline}}
+\end{gather*}
+Under the semantics of \jjr{}, we have
+\begin{gather*}
+  \PR{x}{r}\SEMI
+  \PR{x}{s}\SEMI
+  \IF{r{=}s}\THEN \PW{y}{1}\FI
+  \\
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a1}{\DR{x}{1}}{}
+      \event{a2}{\DR{x}{1}}{right=of a1}
+      \event{a3}{1\EQ1\land1\EQ x \land x\EQ1 \land x=x\mid\DW{y}{1}}{right=of a2}
+    \end{tikzinline}}
+\end{gather*}
+The precondition of $\DWP{y}{1}$ is \emph{not} a tautology, and therefore
+redundant read elimination fails.
+(It is a tautology in
+\begin{math}
+  \PR{x}{r}\SEMI
+  \LET{s}{r}\SEMI
+  \IF{r{=}s}\THEN \PW{y}{1}\FI
+\end{math}.)
+In \jjr{\textsection3.1}, we incorrectly stated that the precondition of
+$\DWP{y}{1}$ was $1\EQ1\land x\EQ x$.
+
+\begin{comment}
+  Precondition of $\DWP{y}{1}$ is $(r{=}s)$ in
+  \begin{math}
+    \sem{\IF{r{=}s}\THEN \PW{y}{1}\FI}.
+  \end{math}
+  Predicate transformers for $\emptyset$ in $\sem{\PR{x}{r}}$ and $\sem{\PR{x}{s}}$ are
+  \begin{align*}
+    \PREDP{(r{=}1 \lor r{=}x)\limplies\bForm[r/x]},
+    \\
+    \PREDP{(s{=}1 \lor s{=}x)\limplies\bForm[s/x]}.
+  \end{align*}
+  Combining the transformers, we have
+  \begin{displaymath}
+    \PREDP{(r{=}1 \lor r{=}x)\limplies(s{=}1 \lor s{=}r)\limplies\bForm[s/x]}.
+  \end{displaymath}
+  Applying this to $(r{=}s)$, we have
+  \begin{displaymath}
+    \PREDP{(r{=}1 \lor r{=}x)\limplies (s{=}1 \lor s{=}r)\limplies (r{=}s)},
+  \end{displaymath}
+  which is not a tautology.
+
+  Same problem occurs \jjr{}, where we have:
+  \begin{align*}
+    \PREDP{\bForm[v/x,r] \land \bForm[x/r]},
+    \\
+    \PREDP{\bForm[v/x,s] \land \bForm[x/s]}.
+  \end{align*}
+  Combining the transformers, we have
+  \begin{displaymath}
+    \PREDP{\bForm[v/x,r,s] \land \bForm [v/x,r][x/s] \land \bForm[x/r][v/x,s] \land \bForm[x/r,s]}.
+  \end{displaymath}
+  Applying this to $(r{=}s)$, we have
+  \begin{displaymath}
+    \PREDP{v{=}v \land v{=}x \land x{=}v \land x{=}x},
+  \end{displaymath}
+  which is not a tautology.
+
+  The semantics here allows this by coalescing:
+  \begin{gather*}
+    \PR{x}{r}\SEMI
+    \PR{x}{s}\SEMI
+    \IF{r{=}s}\THEN \PW{y}{1}\FI
+    \PAR
+    \PW{x}{y}
+    \\
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{a1}{\DR{x}{1}}{}
+        \event{a3}{\DW{y}{1}}{right=of a1}
+        \event{b1}{\DR{y}{1}}{right=3em of a3}
+        \event{b2}{\DW{x}{1}}{right=of b1}
+        \rf{a3}{b1}
+        \po{b1}{b2}
+        \rf[out=169,in=11]{b2}{a1}
+      \end{tikzinline}}
+  \end{gather*}
+
   In \jjr{\textsection2.6} the semantics of read is defined as follows:
   \begin{align*}
-  \sem{\aReg\GETS\aLoc^\amode\SEMI \aCmd} & \eqdef \textstyle\bigcup_\aVal\;
+    \sem{\PR[\amode]{\aLoc}{\aReg}\SEMI \aCmd} & \eqdef \textstyle\bigcup_\aVal\;
     (\DRmode\aLoc\aVal) \prefix \sem{\aCmd} [\aLoc/\aReg]
@@
     If $\bEv$ reads $\aVal$ from $\aLoc$ then either $\aEv=\bEv$ or
-  $\labelingForm'(\aEv)$ implies $\labelingForm(\aEv)[\aVal/\aLoc]$.
+    $\labelingForm'(\aEv) \rimplies \labelingForm(\aEv)[\aVal/\aLoc]$.
   \item[{\labeltextsc[P5a]{(P5a)}{5a}}]\labeltextsc[P5]{}{5}%
-  If $\bEv$ reads and $\aEv$ writes then either $\labelingForm'(\aEv)$
-  implies $\labelingForm(\aEv)$ or $\bEv\le'\aEv$.
+    If $\bEv$ reads and $\aEv$ writes then either $\labelingForm'(\aEv) \rimplies \labelingForm(\aEv)$ or $\bEv\le'\aEv$.
     % \item[{\labeltextsc[P5b]{(P5b)}{5b}}]
@@
     \taglabel{TC2}
-  r\GETS x\SEMI
-  s\GETS x\SEMI
-  \IF{r{=}s}\THEN y\GETS 1\FI
+    \PR{x}{r}\SEMI
+    \PR{x}{s}\SEMI
+    \IF{r{=}s}\THEN \PW{y}{1}\FI
     \PAR
-  x\GETS y
+    \PW{x}{y}
   \end{gather*}
@@
   \begin{math}
-  \sem{\IF{r{=}s}\THEN y\GETS 1\FI}.
+    \sem{\IF{r{=}s}\THEN \PW{y}{1}\FI}.
   \end{math}
@@
   \begin{math}
-  \sem{s\GETS x\SEMI
-  \IF{r{=}s}\THEN y\GETS 1\FI}.
+    \sem{\PR{x}{s}\SEMI
+      \IF{r{=}s}\THEN \PW{y}{1}\FI}.
   \end{math}
@@
   \item[\textsc{(p4c)}] If $\bEv$ reads $\aVal$ from $\aLoc$ then either
-  $\aEv=\bEv$ or $\labelingForm'(\aEv)$ implies
-  $\labelingForm(\aEv)[\aVal/\aReg]$.
+    $\aEv=\bEv$ or $\labelingForm'(\aEv) \rimplies \labelingForm(\aEv)[\aVal/\aReg]$.
   \end{description}
@@
   \begin{gather*}
-  r\GETS x\SEMI
-  s\GETS x\SEMI
-  y\GETS s
+    \PR{x}{r}\SEMI
+    \PR{x}{s}\SEMI
+    \PW{y}{s}
     \PAR
-  x\GETS y
+    \PW{x}{y}
     \\
@@
         \event{a3}{\DW{y}{1}}{right=of a2}
-  %\po{a2}{a3}
+        % \po{a2}{a3}
         \po[out=-20,in=-160]{a1}{a3}
@@
         \po{b1}{b2}
-  %\rf[out=169,in=11]{b2}{a2}
+        % \rf[out=169,in=11]{b2}{a2}
         \rf[out=169,in=11]{b2}{a1}
@@
   precede both reads of $x$.
+\end{comment}
 
-The second issue concerns acquiring reads.  Shortly after publication,
-\citet{anton} noticed a shortcoming of the implementation on \armeight{} in
-\jjr{\textsection 7}.  The proof given there assumes that all internal reads
-can be dropped.  However, this is not the case for acquiring reds.  For
-example, \jjr{} disallows the following execution, which is allowed by
-\armeight{} and \tso{}.
-\begin{gather*}
-  \PW{x}{2}\SEMI 
-  \PR[\mRA]{x}{r}\SEMI
-  \PR{y}{s} \PAR
-  \PW{y}{2}\SEMI
-  \PW[\mRA]{x}{1}
-  \\
-  \hbox{\begin{tikzinline}[node distance=1.5em]
-      \event{a}{\DW{x}{2}}{}
-      \event{b}{\DR[\mRA]{x}{2}}{right=of a}
-      \event{c}{\DR{y}{0}}{right=of b}
-      \event{d}{\DW{y}{2}}{right=2.5em of c}
-      \event{e}{\DW[\mRA]{x}{1}}{right=of d}
-      \rf{a}{b}
-      \sync{b}{c}
-      \wk{c}{d}
-      \sync{d}{e}
-      \wk[out=-165,in=-15]{e}{a}
-      % \rfi{a}{b}
-      % \bob{b}{c}
-      % \fre{c}{d}
-      % \bob{d}{e}
-      % \coe[out=-165,in=-15]{e}{a}
-    \end{tikzinline}}
-\end{gather*}
-The solution we have adopted is to allow an acquiring read to be downgraded
-to a relaxed read when it is preceded (sequentially) by a relaxed write that
-could fulfill it.  Back-porting this solution to \jjr{} requires that we add
-access predicates to the logic and allow
+\endinput
 
-In addition, the \armeight{} implementation result given here does not rely
-on read elimination.  Instead we use a recent alternative characterization of
-\armeight{} \citep{alglave-git-alternate,arm-reference-manual,armed-cats}.
+\myparagraph{Read-Modify-Write Actions}
+
+In \jjr{}, the atomicity axioms \ref{pom-rmw-atomic} erroneously applies only to
+overlapping writes, not overlapping reads.  The difficulty can be seen in
+\refex{ex:rmw-33}.
+
+\jjr{} does not specify the calculation of dependency for \RMW{}s, as
+discussed in \refex{ex:rmw-dep}.
+
+\myparagraph{Stupid Bug in LDRF}
+Definition of race is wrong.
+Should say that at least one is relaxed.

### 2021-04-17 paper/outro.tex
fleshed out merge
Fleshed out outro

@@
 
-\citet{DBLP:conf/esop/PaviottiCPWOB20} use step-indexing to account for
-loops; we expect that the same approach could be applied here.
+This paper is the first to present a compositional semantics for
+sequential composition in a relaxed memory model which can be
+efficiently compiled to modern CPUs. There is, as usual, more research
+to be done.
+
+We have not treated loops in this model, though we expect that the usual
+approach of showing continuity for all the semantic operations with respect to set inclusion
+would go through. \citet{DBLP:conf/esop/PaviottiCPWOB20} use step-indexing to account for
+loops; a similar approach could be applied here.
+
+In \S\ref{sec:arm1} we presented a compilation strategy to \armeight{}
+for a simplified model, but which introduces fences to acquiring
+reads. These fences are not required in \S\ref{sec:arm2}, but at the
+cost of model complexity. It would be illuminating to find out what
+the performance penalty is for these fences.
+
+An earlier version of this paper has been mechanized in Agda; it would be
+reassuring to update the mechanization to bring it in line with the current state.
 

### 2021-04-17 paper/abstract.tex
anonymizing appendix

@@
 as a ``relaxed memory model.''  The relaxed memory models that have been
-proposed to date either fail to address sequential composition, or overly
+proposed to date either fail to address sequential composition directly, or overly
 restrict processors and compilers.
@@
 freedom theorem.
+
+\endinput
+
+Program logics and semantics tell us that when executing (C1; C2) starting in
+state s0, we execute C1 in s0 to arrive at s1, then execute C2 in s1 to
+arrive at the final state s2. This is, of course, an abstraction. Processors
+execute instructions out of order, due to pipelines and caches, and compilers
+reorder programs even more dramatically. All of this reordering is meant to
+be unobservable in single-threaded code, but is observable in multi-threaded
+code. A formal attempt to understand the resulting mess is known as a
+"relaxed memory model." The relaxed memory models that have been proposed to
+date either fail to address sequential composition, or overly restrict
+processors and compilers.
+
+To support sequential composition while targeting modern hardware, we propose
+adding families of predicate transformers to the existing model of "Pomsets
+with Preconditions," which already supports parallel composition.  When
+composing (C1;C2), the predicate transformers used to validate the
+preconditions of events in $C2$ are chosen based on the semantic
+dependencies from events in $C1$ to events in $C2$.  Our model retains the
+good properties of the prior work, including efficient implementation on
+Arm8, support for compiler optimizations, support for logics that prove the
+absence of thin-air behaviors, and a local data race freedom theorem.

### 2021-04-17 paper/oopsla.tex
anonymizing appendix
goodnight again

@@
 
-\myparagraph{Parallel Composition}
-
-In \jjr{\textsection2.4}, parallel composition is defined allowing coalescing
-of events.  Here we have forbidden coalescing.  This difference appears to be
-arbitrary.  In \jjr{}, however, there is a mistake in the handling of
-termination actions.  The predicates should be joined using $\land$, not
-$\lor$.
-
 \myparagraph{Internal Acquiring Reads}
 
-The proof of compilation to Arm in \jjr assumes that all internal reads can
+The proof of compilation to Arm in \jjr{} assumes that all internal reads can
 be eliminated.
@@
 However, this is not the case for acquiring reds.  For example, \jjr{}
-disallows the following execution, which is allowed by \armeight{} and
+disallows the following execution, where the final values of $x$ is $2$ and
+the final value of $y$ is $2$.  This execution is allowed by \armeight{} and
 \tso{}.
@@
 \end{gather*}
-We discussed our solution in \textsection\ref{sec:arm}.
+We discussed two approaches to this problem in \textsection\ref{sec:arm}.
 % The solution we have adopted is to allow an acquiring read to be downgraded
@@
 \end{math}.)
-In \jjr{\textsection3.1}, we incorrectly stated that the precondition of
+\jjr{\textsection3.1}, incorrectly stated that the precondition of
 $\DWP{y}{1}$ was $1\EQ1\land x\EQ x$.
+
+\myparagraph{Parallel Composition}
+
+In \jjr{\textsection2.4}, parallel composition is defined allowing coalescing
+of events.  Here we have forbidden coalescing.  This difference appears to be
+arbitrary.  In \jjr{}, however, there is a mistake in the handling of
+termination actions.  The predicates should be joined using $\land$, not
+$\lor$.
 

### 2021-04-21 paper/arm.tex
Fleshed out outro
typos

@@
 \item $\mSC$ access delay of ${\eqreordersc}$, required by \ref{seq-le-delays},
-\item write-write and read-to-write coherence delay of ${\reorderco}$, required by \ref{seq-le-delays},
+\item write-write and read-to-write coherence delay of ${\eqreorderco}$, required by \ref{seq-le-delays},
 \end{enumerate}
@@
 \item \label{lob-wr}
-  write-to-read coherence delay of ${\reorderco}$, required by \ref{seq-le-delays}.
+  write-to-read coherence delay of ${\eqreorderco}$, required by \ref{seq-le-delays}.
 \end{enumerate}
@@
 \end{definition}
-\citet{armed} explain \EGC{} and \EC{} using the following example, which is
-allowed by Arm.\footnote{We have changed an address dependency in the first
-  thread to a data dependency.}
+\citet{armed} show that \EGC{} and \EC{} are both equivalent to the standard
+definition of Arm8.  They explain \EGC{} and \EC{} using the following
+example, which is allowed by Arm8.\footnote{We have changed an address
+  dependency in the first thread to a data dependency.}
 \begin{gather*}
@@
       \lob{d}{e}
-      \co[out=-165,in=-15]{e}[above right]{a}
+      \co[out=-165,in=-15]{e}[above,pos=.375]{a}
     \end{tikzinline}}
@@
 \subsection{Arm Compilation 1}
+\label{sec:arm1}
 
@@
 
-The change to the semantics is small: we weaken relationship between $\rrfx$
+The change to the semantics is small: we weaken the relationship between $\rrfx$
 and $\le$ in \ref{seq-le-rf}.  Rather than ensuring that there is no
-\emph{global} blocker for a sequentially fulfilled read \ref{seq-le-rf}, we
-require only that there is no \emph{thread-local} blocker \ref{seq-le-rf-rf}.
+\emph{global} blocker for a sequentially fulfilled read \eqref{seq-le-rf}, we
+require only that there is no \emph{thread-local} blocker \eqref{seq-le-rf-rf}.
 This change both allows and requires us to weaken the definition of
@@
 spurious pairs to be added to the $\rrfx$ relation.  The use of
-$\rextendsdef{}{}$ in \ref{if-rf-extends} does this, ensuring that no new
-$\rrfx$ is introduced between events in $\aEvs_1\cap\aEvs_2$ when coalescing.
+$\rextendsdef{}{}$ in \ref{if-rf-extends} does this, ensuring that new
+$\rrfx$ is not introduced between events in $\aEvs_1\cap\aEvs_2$ when coalescing.
 This is necessary to ensure that
@@
         \rfint[out=165,in=15]{b}{a}
-        \wk{a}{b}
+        \wki{a}{b}
       \end{tikzinlinesmall}}}

### 2021-04-21 paper/oopsla.tex
typos

@@
   dependent case, since we lose the unique marker for each read.  In effect,
-  this forces the reads to the same values.
+  this forces all reads of a location to see the same values.
   % To be concrete, the candidate
@@
 \myparagraph{Causal Strengthening}
-\labeltext[]{Causal Strengthening}{xCausal}
+%\labeltext[]{Causal Strengthening}{xCausal}
 \jjr{} imposes \emph{causal strengthening}, which requires for every pomset
@@
 \end{math}.)
-\jjr{\textsection3.1}, incorrectly stated that the precondition of
+\jjr{\textsection3.1} incorrectly stated that the precondition of
 $\DWP{y}{1}$ was $1\EQ1\land x\EQ x$.

### 2021-04-22 paper/arm.tex
added second arm example in appendix

@@
       \lob{d}{e}
-      \co[out=-165,in=-15]{e}[above,pos=.375]{a}
+      \fr[out=-165,in=-15]{e}[above,pos=.375]{a}
     \end{tikzinline}}
@@
 If the read of $x$ in the execution above is changed from acquiring to
-relaxed, then our semantics allows the execution, using the independent case
+relaxed, then our semantics allows the $\rgcb$ execution, using the independent case
 for the read and satisfying the precondition of $\DWP{y}{1}$ by prepending

### 2021-04-27 paper/arm.tex
generlize RMW

@@
 \emph{downgrading} acquires in certain cases.  Unfortunately, it is not
-possible to do this locally without violating important idioms such as
-publication.  For example, consider that $\DRP[\mRA]{x}{1}$ \emph{not} possible for
+possible to do this locally without invalidating important idioms such as
+publication.  For example, consider that $\DRP[\mRA]{x}{1}$ is \emph{not} possible for
 the second thread in the following attempted execution, due to publication of

### 2021-04-29 paper/oopsla.tex
uncovered some examples

@@
 The substitutions collapse $\aLoc$ and $\aReg$, allowing local invariant
-reasoning, causality test case 1, discussed at the end of
+reasoning (\xLIR{}), as required by causality test case 1, discussed at the end of
 \textsection\ref{sec:ex:control}.  Without Skolemizing it is necessary to
@@
 \myparagraph{Causal Strengthening}
-%\labeltext[]{Causal Strengthening}{xCausal}
+% \labeltext[]{Causal Strengthening}{xCausal}
 \jjr{} imposes \emph{causal strengthening}, which requires for every pomset
@@
 
-\begin{comment}
+\myparagraph{Read-Modify-Write Actions}
+
+In \jjr{}, the atomicity axioms \ref{pom-rmw-atomic} erroneously applies only to
+overlapping writes, not overlapping reads.  The difficulty can be seen in
+\refex{ex:rmw-33}.
+
+In addition, \jjr{} uses $\sLOAD{}{}$ instead of $\sLOADP{}{}$ when
+calculating of dependency for \RMW{}s.  For a discussion, see the example at
+the end of \textsection\ref{sec:rmw}.
+
+\section{A Note on Mixed-Mode Data Races}
+
+In preparing this paper, we came across the following example, which appears
+to invalidate Theorem 4.1 of \cite{DBLP:conf/ppopp/DongolJR19}.
+\begin{gather}
+  \nonumber
+  \PW{x}{1}\SEMI
+  \PW[\mREL]{y}{1}\SEMI
+  \PR[\mACQ]{x}{r}
+  \PAR
+  \IF{\PR[\mACQ]{y}{}}\THEN \PW[\mREL]{x}{2}\FI
+  \\
+  \tag{\P}
+  \label{mix1}
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a1}{\DW{x}{1}}{}
+      \raevent{a2}{\DW[\mREL]{y}{1}}{right=of a1}
+      \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
+      \raevent{b1}{\DR[\mACQ]{y}{1}}{right=3em of a3}
+      % \raevent{b1}{\DR[\mACQ]{y}{1}}{below=of a1}
+      \raevent{b2}{\DW[\mREL]{x}{2}}{right=of b1}
+      \sync{a1}{a2}
+      \rf[out=20,in=160]{a1}{a3}
+      \rf[out=20,in=160]{a2}{b1}
+      \wk[out=-20,in=-160]{a3}{b2}
+      \sync{b1}{b2}
+      % \node(ai)[left=3em of a1]{};
+      % \bgoval[yellow!50]{(ai)}{P}
+      % \bgoval[pink!50]{(a1)(a2)(b1)(b2)}{P'\setminus P}
+      % \bgoval[green!10]{(a3)}{P'''\setminus P'}
+    \end{tikzinline}}
+  \\
+  \nonumber
+  %\label{mix2}
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a1}{\DW{x}{1}}{}
+      \raevent{a2}{\DW[\mREL]{y}{1}}{right=of a1}
+      \raevent{a3}{\DR[\mACQ]{x}{2}}{right=of a2}
+      \raevent{b1}{\DR[\mACQ]{y}{1}}{right=3em of a3}
+      \raevent{b2}{\DW[\mREL]{x}{2}}{right=of b1}
+      \sync{a1}{a2}
+      \rf[out=20,in=160]{a2}{b1}
+      \rf[out=160,in=20]{b2}{a3}
+      \sync{b1}{b2}
+    \end{tikzinline}}
+\end{gather}
+The program is data-race free.  The two executions shown are the only
+top-level executions that include $\DWP[\mREL]{x}{2}$.
+
+Theorem 4.1 of \cite{DBLP:conf/ppopp/DongolJR19} is stated by extending
+execution sequences.  In the terminology of
+\cite{DBLP:conf/ppopp/DongolJR19}, a read is \emph{$L$-weak} if it is
+sequentially stale.  Let
+\begin{math}
+  \rho=\DWP{x}{1}\allowbreak
+  \DWP[\mREL]{y}{1}\allowbreak
+  \DRP[\mACQ]{y}{1}\allowbreak
+  \DWP[\mREL]{x}{2}
+\end{math}
+be a sequence and
+\begin{math}
+  \alpha=\DRP[\mACQ]{x}{1}.
+\end{math}
+$\rho$ is $L$-sequential and $\alpha$ is $L$-weak in $\rho\alpha$.  But there
+is no execution of this program that includes a data race, contradicting the
+theorem.  The error seems to be in Lemma A.4 of
+\cite{DBLP:conf/ppopp/DongolJR19}, which states that if $\alpha$ is $L$-weak
+after an $L$-sequential $\rho$, then $\alpha$ must be in a data race.  That
+is clearly false here, since $\DRP[\mACQ]{x}{1}$ is stale, but the program is
+data race free.
+
+In proving the SC-LDRF result in \jjr{\textsection8}, we noted that our proof
+technique is more robust than that of \cite{DBLP:conf/ppopp/DongolJR19},
+because it limits the prefixes that must be considered.  In \eqref{mix1}, the
+induction hypothesis requires that we add $\DRP[\mACQ]{x}{1}$ before
+$\DWP[\mREL]{x}{2}$ since $\DRP[\mACQ]{x}{1}\xwk\DWP[\mREL]{x}{2}$.  In
+particular,
+\begin{gather*}
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a1}{\DW{x}{1}}{}
+      \raevent{a2}{\DW[\mREL]{y}{1}}{right=of a1}
+      % \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
+      \raevent{b1}{\DR[\mACQ]{y}{1}}{right=3em of a3}
+      % \raevent{b1}{\DR[\mACQ]{y}{1}}{below=of a1}
+      \raevent{b2}{\DW[\mREL]{x}{2}}{right=of b1}
+      \sync{a1}{a2}
+      % \rf[out=20,in=160]{a1}{a3}
+      \rf[out=20,in=160]{a2}{b1}
+      % \wk[out=-20,in=-160]{a3}{b2}
+      \sync{b1}{b2}
+      % \node(ai)[left=3em of a1]{};
+      % \bgoval[yellow!50]{(ai)}{P}
+      % \bgoval[pink!50]{(a1)(a2)(b1)(b2)}{P'\setminus P}
+      % \bgoval[green!10]{(a3)}{P'''\setminus P'}
+    \end{tikzinline}}
+\end{gather*}
+is not a downset of \eqref{mix1}, because
+$\DRP[\mACQ]{x}{1}\xwk\DWP[\mREL]{x}{2}$.  As we noted in \jjr{\textsection8},
+this affects the inductive order in which we move across pomsets, but does
+not affect the set of pomsets that are considered.  In particular,
+\begin{gather*}
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a1}{\DW{x}{1}}{}
+      \raevent{a2}{\DW[\mREL]{y}{1}}{right=of a1}
+      % \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
+      \raevent{b1}{\DR[\mACQ]{y}{1}}{right=3em of a3}
+      % \raevent{b1}{\DR[\mACQ]{y}{1}}{below=of a1}
+      % \raevent{b2}{\DW[\mREL]{x}{2}}{right=of b1}
+      \sync{a1}{a2}
+      % \rf[out=20,in=160]{a1}{a3}
+      \rf[out=20,in=160]{a2}{b1}
+      % \wk[out=-20,in=-160]{a3}{b2}
+      % \sync{b1}{b2}
+      % \node(ai)[left=3em of a1]{};
+      % \bgoval[yellow!50]{(ai)}{P}
+      % \bgoval[pink!50]{(a1)(a2)(b1)(b2)}{P'\setminus P}
+      % \bgoval[green!10]{(a3)}{P'''\setminus P'}
+    \end{tikzinline}}
+\end{gather*}
+is a downset of \eqref{mix1}.
+
+\section{Additional RMW Examples}
+\begin{example}
+  This definition ensures atomicity, disallowing executions such as
+  \cite[Ex.~3.2]{DBLP:journals/pacmpl/PodkopaevLV19}:
+  \begin{gather*}
+    % \taglabel{RMW1}
+    \begin{gathered}
+      \PW{x}{0}\SEMI \PFADD[\mRLX][\mRLX]{x}{s}{1}
+      \PAR
+      \PW{x}{2}\SEMI \PR{x}{s}
+      \\
+      \hbox{\begin{tikzinline}[node distance=1.5em]
+          \event{a2}{\DR{x}{0}}{}
+          \event{a1}{\DW{x}{0}}{left=of a2}
+          \rf{a1}{a2}
+          \event{a3}{\DW{x}{2}}{right=of a2}
+          \wk{a2}{a3}
+          \event{b2}{\DW{x}{1}}{right=of a3}
+          \event{b3}{\DR{x}{1}}{right=of b2}
+          \rmw[out=-15,in=-165]{a2}[below]{b2}
+          \wk{a3}{b2}
+          \rf{b2}{b3}
+          \liftrmw[out=165,in=15]{a3}{a2}
+        \end{tikzinline}}
+    \end{gathered}
+  \end{gather*}
+  By \ref{pom-rmw-atomic1}, since $\DWP{x}{2}\xwk\DWP{x}{1}$, it must be that
+  $\DWP{x}{2}\xwk\DRP{x}{0}$, creating a cycle.
+\end{example}
+
+\begin{example}
+  \label{ex:rmw-33}
+  Two successful \RMW{}s cannot see the same write:
+  \begin{gather*}
+    \begin{gathered}
+      \PW{x}{0}\SEMI (\FADD^{\mRLX,\mRLX}(\aLoc,1)\PAR \FADD^{\mRLX,\mRLX}(\aLoc,1))
+      \\
+      \hbox{\begin{tikzinline}[node distance=1.5em]
+          \event{i}{\DW{x}{0}}{}
+          \event{a1}{a{:}\DR{x}{0}}{right=3em of i}
+          \event{a2}{b{:}\DW{x}{1}}{right=of a1}
+          \event{b1}{c{:}\DR{x}{0}}{right=3em of a2}
+          \event{b2}{d{:}\DW{x}{1}}{right=of b1}
+          \rmw{a1}{a2}
+          \rmw{b1}{b2}
+          \rf{i}{a1}
+          \rf[out=15,in=165]{i}{b1}
+          \wk[out=-15,in=-165]{a1}{b2}
+          \liftrmw[out=-15,in=-165]{a2}{b1}
+          % \wk{a1}{b2}
+          \wk{b1}{a2}
+        \end{tikzinline}}
+    \end{gathered}
+  \end{gather*}
+  The order from read-to-write is required by fulfillment.  
+  Apply \ref{pom-rmw-atomic1} of the second \RMW{} to $a\xwk d$, we have that $a\xwk c$.  Subsequently
+  applying \ref{pom-rmw-atomic2} of the first \RMW{}, we have $b \xwk c$, creating a cycle.
+\end{example}
+
+\begin{example}
+  By using two actions rather than one, the definition allows examples such as the
+  following, which is allowed by \armeight{} 
+  \cite[Ex.~3.10]{DBLP:journals/pacmpl/PodkopaevLV19}:
+  \begin{gather*}
+    % \taglabel{RMW2}
+    \begin{gathered}
+      \PR{y}{r}\SEMI
+      \PW{z}{r}
+      \PAR
+      \PR{z}{r}\SEMI
+      \PW{x}{0}\SEMI
+      \PFADD[\mRLX][\mREL]{x}{s}{1} \SEMI
+      \PW{y}{s}{+}1
+      \\[-1ex]
+      \hbox{\begin{tikzinline}[node distance=1.5em]
+          \event{a1}{\DR{y}{1}}{}
+          \event{a2}{\DW{z}{1}}{right=of a1}
+          \po{a1}{a2}
+          \event{b1}{\DR{z}{1}}{right=3em of a2}
+          \rf{a2}{b1}
+          \event{b2}{\DW{x}{0}}{right=of b1}
+          \event{b3}{\DR{x}{0}}{right=of b2}
+          \rf{b2}{b3}
+          \event{b4}{\DWRel{x}{1}}{right=2em of b3}
+          \rmw{b3}{b4}
+          \event{b5}{\DW{y}{1}}{right=of b4}
+          \sync[out=-15,in=-165]{b1}{b4}
+          \po[out=-20,in=-160]{b3}{b5}
+          \rf[out=170,in=10]{b5}{a1}
+        \end{tikzinline}}
+    \end{gathered}
+  \end{gather*}
+\end{example}
+
+\begin{example}
+  Consider the \textsc{cdrf} example from \cite{DBLP:conf/pldi/LeeCPCHLV20}:
+  \begin{gather*}
+    \begin{gathered}
+      \begin{aligned}
+        &\PFADD[\mACQ][\mREL]{x}{r}{1}\SEMI \IF{r{=}0}\THEN \PW{y}{1} \FI
+        \\\PAR\;\;&
+        \PFADD[\mACQ][\mREL]{x}{r}{1}\SEMI \IF{r{=}0}\THEN \IF{y}\THEN \PW{x}{0} \FI \FI
+      \end{aligned}
+      \\
+      \hbox{\footnotesize\begin{tikzinline}[node distance=1.5em]
+          \raevent{a1}{\DR[\mACQ]{x}{0}}{}
+          \raevent{a1b}{\DW[\mREL]{x}{1}}{right=of a1}
+          \event{a2}{\DW{y}{1}}{right=of a1b}
+          \raevent{b0}{\DR[\mACQ]{x}{0}}{right=3em of a2}
+          \raevent{b0b}{\DW[\mREL]{x}{1}}{right=of b0}
+          \event{b1}{\DR{y}{1}}{right=of b0b}
+          \event{b2}{\DW{x}{0}}{right=of b1}
+          \rmw{a1}{a1b}
+          \rmw{b0}{b0b}
+          \rf[out=-13,in=-163]{a2}{b1}
+          \sync[out=20,in=160]{a1}{a2}
+          \sync[out=20,in=160]{b0}{b1}
+          \po{b1}{b2}
+          \rf[out=-165,in=-12]{b2}{a1}
+        \end{tikzinline}}
+    \end{gathered}
+  \end{gather*}
+\end{example}
+\begin{example}
+  Consider this example from \cite[\textsection C]{DBLP:conf/pldi/LeeCPCHLV20}:
+  \begin{gather*}
+    \begin{gathered}
+      \begin{aligned}
+        &\PCAS[\mRLX][\mRLX]{x}{r}{0}{1}\SEMI \IF{r{\leq}1}\THEN \PW{y}{1} \FI
+        \\\PAR\;\;&
+        \PCAS[\mRLX][\mRLX]{x}{r}{0}{2}\SEMI \IF{r{=}0}\THEN \IF{y}\THEN \PW{x}{0} \FI \FI
+      \end{aligned}
+      \\
+      \hbox{\footnotesize\begin{tikzinline}[node distance=1.5em]
+          \event{a1}{\DR{x}{0}}{}
+          \event{a1b}{\DW{x}{1}}{right=of a1}
+          \event{a2}{\DW{y}{1}}{right=of a1b}
+          \event{b0}{\DR{x}{0}}{right=3em of a2}
+          \event{b0b}{\DW{x}{2}}{right=of b0}
+          \event{b1}{\DR{y}{1}}{right=of b0b}
+          \event{b2}{\DW{x}{0}}{right=of b1}
+          \rmw{a1}{a1b}
+          \rmw{b0}{b0b}
+          \rf[out=-13,in=-163]{a2}{b1}
+          \po[out=20,in=160]{a1}{a2}
+          \po[out=20,in=160]{b0}{b1}
+          \po{b1}{b2}
+          \rf[out=-165,in=-12]{b2}{a1}
+        \end{tikzinline}}
+    \end{gathered}
+  \end{gather*}
+\end{example}
+
+\endinput
+
+\myparagraph{Stupid Bug in LDRF}
+Definition of race is wrong.
+Should say that at least one is relaxed.
+
 Precondition of $\DWP{y}{1}$ is $(r{=}s)$ in
@@
 precede both reads of $x$.
-\end{comment}
-
-\endinput
-
-\myparagraph{Read-Modify-Write Actions}
-
-In \jjr{}, the atomicity axioms \ref{pom-rmw-atomic} erroneously applies only to
-overlapping writes, not overlapping reads.  The difficulty can be seen in
-\refex{ex:rmw-33}.
-
-\jjr{} does not specify the calculation of dependency for \RMW{}s, as
-discussed in \refex{ex:rmw-dep}.
-
-\myparagraph{Stupid Bug in LDRF}
-Definition of race is wrong.
-Should say that at least one is relaxed.

### 2021-05-26 paper/arm.tex
generlize RMW
Added m5a

@@
 for relaxed access.  In the next section we recover the optimal lowering by
-adopting an alternative semantics for \ref{seq-le}.
+adopting an alternative semantics for \ref{seq-le-rf} and \ref{seq-le-delays}.
 
@@
     We show that all the order required in the pomset is also required by
-    \armeight{}.  \ref{rf-block} holds since ${\rcb_1}$ is consistent with
+    \armeight{}.  \ref{pom-le-block} holds since ${\rcb_1}$ is consistent with
     ${\rco_1}$ and ${\rfr_1}$.  As noted \ref{lob-le}, $\rlob$ includes the order
@@
     We show that all the order required in the pomset is also required by
-    \armeight{}.  \ref{rf-block} holds since ${\rcb_1}$ is consistent with
+    \armeight{}.  \ref{pom-le-block} holds since ${\rcb_1}$ is consistent with
     ${\rco_1}$ and ${\rfr_1}$.  \ref{seq-le-rf-rf} follows from \ref{arm-rfi}.

### 2021-05-26 paper/oopsla.tex
uncovered some examples
Added m5a

@@
 % can be dropped.
-However, this is not the case for acquiring reds.  For example, \jjr{}
+However, this is not the case for acquiring reads.  For example, \jjr{}
 disallows the following execution, where the final values of $x$ is $2$ and
@@
 the end of \textsection\ref{sec:rmw}.
+
+\myparagraph{Data Race Freedom}
+
+The definition of data race is wrong in \jjr{}.  It should require that that
+at least one action is relaxed.
+
+Not that the definition of \emph{$L$-stable} applies in the case that
+conflicting writes are totally ordered.  This gives a result more in the
+spirit of \cite{Dolan:2018:BDR:3192366.3192421}.  In particular, this special
+case of the theorem clarifies the discussion of \ref{past}.
 
@@
 
-\section{Additional RMW Examples}
-\begin{example}
-  This definition ensures atomicity, disallowing executions such as
-  \cite[Ex.~3.2]{DBLP:journals/pacmpl/PodkopaevLV19}:
-  \begin{gather*}
-    % \taglabel{RMW1}
-    \begin{gathered}
-      \PW{x}{0}\SEMI \PFADD[\mRLX][\mRLX]{x}{s}{1}
-      \PAR
-      \PW{x}{2}\SEMI \PR{x}{s}
-      \\
-      \hbox{\begin{tikzinline}[node distance=1.5em]
-          \event{a2}{\DR{x}{0}}{}
-          \event{a1}{\DW{x}{0}}{left=of a2}
-          \rf{a1}{a2}
-          \event{a3}{\DW{x}{2}}{right=of a2}
-          \wk{a2}{a3}
-          \event{b2}{\DW{x}{1}}{right=of a3}
-          \event{b3}{\DR{x}{1}}{right=of b2}
-          \rmw[out=-15,in=-165]{a2}[below]{b2}
-          \wk{a3}{b2}
-          \rf{b2}{b3}
-          \liftrmw[out=165,in=15]{a3}{a2}
-        \end{tikzinline}}
-    \end{gathered}
-  \end{gather*}
-  By \ref{pom-rmw-atomic1}, since $\DWP{x}{2}\xwk\DWP{x}{1}$, it must be that
-  $\DWP{x}{2}\xwk\DRP{x}{0}$, creating a cycle.
-\end{example}
-
-\begin{example}
-  \label{ex:rmw-33}
-  Two successful \RMW{}s cannot see the same write:
-  \begin{gather*}
-    \begin{gathered}
-      \PW{x}{0}\SEMI (\FADD^{\mRLX,\mRLX}(\aLoc,1)\PAR \FADD^{\mRLX,\mRLX}(\aLoc,1))
-      \\
-      \hbox{\begin{tikzinline}[node distance=1.5em]
-          \event{i}{\DW{x}{0}}{}
-          \event{a1}{a{:}\DR{x}{0}}{right=3em of i}
-          \event{a2}{b{:}\DW{x}{1}}{right=of a1}
-          \event{b1}{c{:}\DR{x}{0}}{right=3em of a2}
-          \event{b2}{d{:}\DW{x}{1}}{right=of b1}
-          \rmw{a1}{a2}
-          \rmw{b1}{b2}
-          \rf{i}{a1}
-          \rf[out=15,in=165]{i}{b1}
-          \wk[out=-15,in=-165]{a1}{b2}
-          \liftrmw[out=-15,in=-165]{a2}{b1}
-          % \wk{a1}{b2}
-          \wk{b1}{a2}
-        \end{tikzinline}}
-    \end{gathered}
-  \end{gather*}
-  The order from read-to-write is required by fulfillment.  
-  Apply \ref{pom-rmw-atomic1} of the second \RMW{} to $a\xwk d$, we have that $a\xwk c$.  Subsequently
-  applying \ref{pom-rmw-atomic2} of the first \RMW{}, we have $b \xwk c$, creating a cycle.
-\end{example}
-
-\begin{example}
-  By using two actions rather than one, the definition allows examples such as the
-  following, which is allowed by \armeight{} 
-  \cite[Ex.~3.10]{DBLP:journals/pacmpl/PodkopaevLV19}:
-  \begin{gather*}
-    % \taglabel{RMW2}
-    \begin{gathered}
-      \PR{y}{r}\SEMI
-      \PW{z}{r}
-      \PAR
-      \PR{z}{r}\SEMI
-      \PW{x}{0}\SEMI
-      \PFADD[\mRLX][\mREL]{x}{s}{1} \SEMI
-      \PW{y}{s}{+}1
-      \\[-1ex]
-      \hbox{\begin{tikzinline}[node distance=1.5em]
-          \event{a1}{\DR{y}{1}}{}
-          \event{a2}{\DW{z}{1}}{right=of a1}
-          \po{a1}{a2}
-          \event{b1}{\DR{z}{1}}{right=3em of a2}
-          \rf{a2}{b1}
-          \event{b2}{\DW{x}{0}}{right=of b1}
-          \event{b3}{\DR{x}{0}}{right=of b2}
-          \rf{b2}{b3}
-          \event{b4}{\DWRel{x}{1}}{right=2em of b3}
-          \rmw{b3}{b4}
-          \event{b5}{\DW{y}{1}}{right=of b4}
-          \sync[out=-15,in=-165]{b1}{b4}
-          \po[out=-20,in=-160]{b3}{b5}
-          \rf[out=170,in=10]{b5}{a1}
-        \end{tikzinline}}
-    \end{gathered}
-  \end{gather*}
-\end{example}
-
-\begin{example}
-  Consider the \textsc{cdrf} example from \cite{DBLP:conf/pldi/LeeCPCHLV20}:
-  \begin{gather*}
-    \begin{gathered}
-      \begin{aligned}
-        &\PFADD[\mACQ][\mREL]{x}{r}{1}\SEMI \IF{r{=}0}\THEN \PW{y}{1} \FI
-        \\\PAR\;\;&
-        \PFADD[\mACQ][\mREL]{x}{r}{1}\SEMI \IF{r{=}0}\THEN \IF{y}\THEN \PW{x}{0} \FI \FI
-      \end{aligned}
-      \\
-      \hbox{\footnotesize\begin{tikzinline}[node distance=1.5em]
-          \raevent{a1}{\DR[\mACQ]{x}{0}}{}
-          \raevent{a1b}{\DW[\mREL]{x}{1}}{right=of a1}
-          \event{a2}{\DW{y}{1}}{right=of a1b}
-          \raevent{b0}{\DR[\mACQ]{x}{0}}{right=3em of a2}
-          \raevent{b0b}{\DW[\mREL]{x}{1}}{right=of b0}
-          \event{b1}{\DR{y}{1}}{right=of b0b}
-          \event{b2}{\DW{x}{0}}{right=of b1}
-          \rmw{a1}{a1b}
-          \rmw{b0}{b0b}
-          \rf[out=-13,in=-163]{a2}{b1}
-          \sync[out=20,in=160]{a1}{a2}
-          \sync[out=20,in=160]{b0}{b1}
-          \po{b1}{b2}
-          \rf[out=-165,in=-12]{b2}{a1}
-        \end{tikzinline}}
-    \end{gathered}
-  \end{gather*}
-\end{example}
-\begin{example}
-  Consider this example from \cite[\textsection C]{DBLP:conf/pldi/LeeCPCHLV20}:
-  \begin{gather*}
-    \begin{gathered}
-      \begin{aligned}
-        &\PCAS[\mRLX][\mRLX]{x}{r}{0}{1}\SEMI \IF{r{\leq}1}\THEN \PW{y}{1} \FI
-        \\\PAR\;\;&
-        \PCAS[\mRLX][\mRLX]{x}{r}{0}{2}\SEMI \IF{r{=}0}\THEN \IF{y}\THEN \PW{x}{0} \FI \FI
-      \end{aligned}
-      \\
-      \hbox{\footnotesize\begin{tikzinline}[node distance=1.5em]
-          \event{a1}{\DR{x}{0}}{}
-          \event{a1b}{\DW{x}{1}}{right=of a1}
-          \event{a2}{\DW{y}{1}}{right=of a1b}
-          \event{b0}{\DR{x}{0}}{right=3em of a2}
-          \event{b0b}{\DW{x}{2}}{right=of b0}
-          \event{b1}{\DR{y}{1}}{right=of b0b}
-          \event{b2}{\DW{x}{0}}{right=of b1}
-          \rmw{a1}{a1b}
-          \rmw{b0}{b0b}
-          \rf[out=-13,in=-163]{a2}{b1}
-          \po[out=20,in=160]{a1}{a2}
-          \po[out=20,in=160]{b0}{b1}
-          \po{b1}{b2}
-          \rf[out=-165,in=-12]{b2}{a1}
-        \end{tikzinline}}
-    \end{gathered}
-  \end{gather*}
-\end{example}
-
 \endinput
-
-\myparagraph{Stupid Bug in LDRF}
-Definition of race is wrong.
-Should say that at least one is relaxed.
 

### 2021-06-09 paper/arm.tex
Weakest preconditions

@@
   \begin{enumerate}[,label=(\textsc{a}\arabic*),ref=\textsc{a}\arabic*]
-  \item $\Event\subseteq\AllEvents$ is a set of {events},
-  \item $\labeling: \Event \fun \Act$ defines a {label} for each event,
-  \item ${\rpoloc} : \Event\times\Event$, is a per-thread, per-location total
+  \item $\aEvs\subseteq\AllEvents$ is a set of {events},
+  \item $\labeling: \aEvs \fun \Act$ defines a {label} for each event,
+  \item ${\rpoloc} : \aEvs\times\aEvs$, is a per-thread, per-location total
     order, capturing \emph{per-location program order},
-  \item ${\rlob} : \Event\times\Event$, is a per-thread partial order capturing
+  \item ${\rlob} : \aEvs\times\aEvs$, is a per-thread partial order capturing
     \emph{locally-ordered-before}, such that
@@
   \item \label{arm-co}
-    ${\rco} : \Event\times\Event$, is a per-location total order on
+    ${\rco} : \aEvs\times\aEvs$, is a per-location total order on
     writes, capturing \emph{coherence}, 
   \item \label{arm-rf}
-    ${\rrfx} : \Event\times\Event$, is a surjective and injective
+    ${\rrfx} : \aEvs\times\aEvs$, is a surjective and injective
     relation on reads, capturing \emph{reads-from}, such that
@@
   \item \label{arm-gcb}
-    ${\rgcb}\supseteq\PBR{{\rco}\cup{\rrfx}}$ is a linear order %, capturing \emph{globally completes before}, %$\rcb : % \Event\times\Event$
+    ${\rgcb}\supseteq\PBR{{\rco}\cup{\rrfx}}$ is a linear order %, capturing \emph{globally completes before}, %$\rcb : % \aEvs\times\aEvs$
     such that 
@@
   \item \label{arm-cb}
-    ${\rcb}\supseteq\PBR{{\rco}\cup{\rlob}}$ is a linear order %, capturing \emph{completes before}, %$\rcb : % \Event\times\Event$
+    ${\rcb}\supseteq\PBR{{\rco}\cup{\rlob}}$ is a linear order %, capturing \emph{completes before}, %$\rcb : % \aEvs\times\aEvs$
     such that if $\bEv\xrfx\aEv$ then either

### 2021-06-09 paper/outro.tex
Weakest preconditions

@@
 
-This paper is the first to present a compositional semantics for
+This paper is the first to present a direct compositional semantics for
 sequential composition in a relaxed memory model which can be
@@
 the performance penalty is for these fences.
+
+Dissatisfying thing: read elimination is a mess.  Can't do the nice thing and
+use
+\begin{math}
+  \aTr{\bEvs}{\bForm} \riff \aLoc{=}\aReg \limplies \bForm
+\end{math}
+for \ref{read-tau-empty} because there may be a release-acquire pair between
+the read and the matching write.
 

### 2021-06-15 paper/intro.tex
generlize RMW
added referee reports

@@
 $\fwp{\aCmd}{\bForm}$ for loop-free code below (where $\aReg$--$\bReg$ range over thread-local \emph{registers}
-and $\aExp$--$\bExp$ range over side-effect free \emph{expressions}). 
+and $\aExp$--$\bExp$ range over side-effect-free \emph{expressions}). 
 %\begin{multicols}{2}
@@
 altogether, as in the pomset model of
-\citet{DBLP:journals/corr/abs-1804-04214}---this model uses syntactic
-dependencies and thus dramatically limits compiler optimization.  This leaves
-open the question of how to apply logic to racy programs without
-overconstraining the implementation.  To understand the solution, one must first
-understand the constraints imposed by hardware and compilers.
+\citet{DBLP:journals/corr/abs-1804-04214}, which enforces syntactic
+dependencies.  This leaves open the question of how to apply logic to racy
+programs without overconstraining the implementation.  To understand the
+solution, one must first understand the constraints imposed by hardware and
+compilers.
 
@@
 semantics \cite{Lamport:1979:MMC:1311099.1311750}, in which \emph{program order} inside a thread implies
-a total \emph{causal order} between read and write events, for example:
+a total \emph{causal order} between read and write events, for example
+(where $\SEMI\!$ has higher precedence than $\PAR$):
 \begin{gather*}
@@
 sequential consistency is weakened to only apply \emph{per-location} and not globally
-over the whole program. This allows executions which
+over the whole program. This allows executions that
 are inconsistent with program order, such as:
@@
 \end{gather*}
-Preconditions are discharged by being ordered after a read:
+Preconditions are discharged by being ordered after a read (we assume the
+usual precedence for logical operators---$\lnot$, $\land$, $\lor$, $\limplies$): 
 \begin{gather*}
@@
 languages should guarantee sequential abstraction, despite the performance
-penalties.  In this paper, we have take the opposite approach.  We have
+penalties.  In this paper, we take the opposite approach.  We have
 attempted to find the most intellectually tractable model that encompasses

### 2021-06-17 paper/2SUB/definitions.tex
reorg

\section{The Basic Model}
\label{sec:model}

After some preliminaries (\textsection\ref{sec:prelim}--\ref{sec:actions}),
we define the basic model and establish some basic properties
(\textsection\ref{sec:pomsets} and \reffig{fig:sem}).  We then explain the
model using examples (\textsection\ref{sec:ex:pomset}--\ref{sec:ex:last}).
We encourage readers to skim the definitions and then skip to
\textsection\ref{sec:ex:pomset}, coming back as needed.

%% Batty suggest example where dependencies are added and also go away, perhaps
%% by store forwarding. Something like:
%% \texttt{(r=x; y=1); (s=y; z=s+r)}

% In this section, we present the mathematical preliminaries for the
% model (which can be skipped on first reading). We then present the
% model incrementally, starting with a model built using
% \emph{partially ordered multisets}
% (\emph{pomsets})~\cite{GISCHER1988199,Plotkin:1997:TSP:266557.266600},
% and then adding preconditions and finally predicate transformers.

% In later sections, we will discuss extensions to the logic, and to the
% semantics of load, store and thread initialization, in order to model
% relaxed memory more faithfully. We stress that these features do
% \emph{not} change any of the structures of the language: conditionals,
% parallel composition, and sequential composition are as defined in this section.

\subsection{Preliminaries}
\label{sec:prelim}
The syntax is built from
\begin{itemize}
\item a set of \emph{values} $\Val$, ranged over by
  $\aVal$, $\bVal$, $\cVal$, $\dVal$,
\item a set of \emph{registers} $\Reg$, ranged over by
  $\aReg$, $\bReg$,
\item a set of \emph{expressions} $\Exp$, ranged over by
  $\aExp$, $\bExp$,  $\cExp$.
\end{itemize}

\emph{Memory references} are tagged values, written $\REF{\cVal}$.  Let $\Loc$
be the set of memory references, ranged over by $\aLoc$, $\bLoc$, $\cLoc$.
% 
We require that
\begin{itemize}
\item values and registers are disjoint, 
\item values include at least the constants $0$ and $1$,  
\item expressions include at least registers and values, 
\item expressions do \emph{not} include references: $\aExp[\bExp/\aLoc]=\aExp$.
\end{itemize}

We model the following language.
\begin{gather*}
  \begin{aligned}
    \amode,\bmode \BNFDEF& \mRLX
    \BNFSEP \mREL
    \BNFSEP \mACQ
    \BNFSEP \mSC
    % \amode \BNFDEF& \mRLX
    % \BNFSEP \mRA 
    % \BNFSEP \mSC
    % &\mkern100mu
    % \fmode \BNFDEF& \fACQ 
    % \BNFSEP \fREL
    % \BNFSEP \fSC
  \end{aligned}
  \\
  \aCmd
  \BNFDEF \LET{\aReg}{\aExp}
  \BNFSEP \PR[\amode]{\REF{\cExp}}[\ascope]{\aReg}
  \BNFSEP \PW[\amode]{\REF{\cExp}}[\ascope]{\aExp}
  \BNFSEP \PF[\ascope]{\fmode}
  \BNFSEP \SKIP
  \BNFSEP \aCmd_1 \SEMI \aCmd_2
  \BNFSEP \IF{\aExp} \THEN \aCmd_1 \ELSE \aCmd_2 \FI
  \BNFSEP \aCmd_1 \LPAR[\bThrd] \aCmd_2
  % \\[-.5ex]
  % \BNFSEP& \PCAS[\amode_1][\amode_2]{\REF{\cExp}}[\ascope]{\aReg}{\aExp}{\bExp}
  % \BNFSEP \PFADD[\amode_1][\amode_2]{\REF{\cExp}}[\ascope]{\aReg}{\aExp}
  % \BNFSEP \PEXCHG[\amode_1][\amode_2]{\REF{\cExp}}[\ascope]{\aReg}{\aExp}
\end{gather*}

% \emph{Memory modes}, $\amode$, are {relaxed} ($\mRLX$), {release-acquire}
% ($\mRA$), and {sequentially consistent} ($\mSC$).  Relaxed mode is the
% default; we regularly elide it from examples.  $\mRA$/$\mSC$ accesses are
% collectively known as \emph{synchronized accesses}.

% \emph{Fence modes}, $\bmode$, are {acquire} ($\fACQ$), {release} ($\fREL$), 
% and {acquire-release} ($\fSC$).

\emph{Access modes}, $\amode$, are {relaxed} ($\mRLX$),
{release} ($\mREL$), {acquire} ($\mACQ$), and
{sequentially consistent} ($\mSC$).
%
Let expressions ($\LET{\aReg}{\aExp}$) only affect thread-local state and
thus do not have a mode.
%
Reads ($\PR[\amode]{\REF{\cExp}}[\ascope]{\aReg}$) support
%$\mWK$,
$\mRLX$,
%$\mREL$,
$\mACQ$,
%$\mRA$, 
$\mSC$. 
Writes ($\PW[\amode]{\REF{\cExp}}[\ascope]{\aReg}$) support
%$\mWK$,
$\mRLX$,
$\mREL$,
%$\mACQ$,
%$\mRA$, 
$\mSC$. 
Fences ($\PF[\ascope]{\fmode}$) support
%$\mWK$,
%$\mRLX$,
$\mREL$,
$\mACQ$,
%$\mRA$,
$\mSC$.

\emph{Commands}, aka \emph{statements}, $\aCmd$, include memory accesses at a
given mode, as well as the usual structural constructs. Following
\cite{DBLP:conf/icfp/FerreiraHJ96}, $\LPAR$ denotes parallel composition,
preserving thread state on the left after a join.  In examples and
sublanguages without join, we use the symmetric $\PAR$ operator.

We use common syntax sugar, such as \emph{extended expressions}, $\aEExp$,
which include memory locations.  For example, if $\aEExp$ includes a single
occurrence of $\aLoc$, then $\bLoc\GETS\aEExp\SEMI \aCmd$ is shorthand for
$\aReg\GETS\aLoc\SEMI\bLoc\GETS\aEExp[\aReg/\aLoc]\SEMI \aCmd$.  Each
occurrence of $\aLoc$ in an extended expression corresponds to an separate
read.  We also write $\IF{\aExp} \THEN \aCmd \FI$ as shorthand for
$\IF{\aExp} \THEN \aCmd\ELSE \SKIP\FI$.

Throughout \textsection\ref{sec:intro}--\ref{sec:arm} we 
require that
\begin{itemize}
\item each register is assigned at most once in a program.
  % \end{itemize}
  % In \textsection\ref{sec:complications} and following, we
  % require instead that
  % \begin{itemize}
  % \end{itemize}
\end{itemize}
In \textsection\ref{sec:additional}, we drop this restriction, requiring
instead that
\begin{itemize}
\item there are registers
  $\uRegs{\AllEvents}=\{\uReg{\aEv}\mid\aEv\in\AllEvents\}$, that do not
  appear in programs: $\aCmd[\bExp/\uReg{\aEv}]=\aCmd$.
\end{itemize}
% In contexts that make no use of $\uRegs{\AllEvents}$, we make the first
% assumption.

The semantics is built from the following.
\begin{itemize}
\item a set of \emph{events} $\AllEvents$, ranged over by $\aEv$, $\bEv$,
  $\cEv$, %$\dEv$,
  and subsets ranged over by $\aEvs$, $\bEvs$, $\cEvs$,  
\item a set of \emph{logical formulae} $\Formulae$, ranged over by $\aForm$,
  $\bForm$, $\cForm$,
\item a set of \emph{actions} $\Act$, ranged over by $\aAct$, $\bAct$.
\end{itemize}
% $\dEvs$.

We require that
\begin{itemize}
\item formulae include $\TRUE$, $\FALSE$ and the equalities $(\aExp{=}\bExp)$ and $(\aLoc{=}\aExp)$,
\item formulae are closed under $\lnot$, $\land$, $\lor$, $\limplies$, and
  substitutions $[\aExp/\aReg]$, $[\aExp/\aLoc]$,
\item there is a relation $\rimpliesdef$ between
  formulae, capturing entailment, %\subseteq(\Formulae\times\Formulae)$ %
\item $\rimpliesdef$ has the expected semantics for $=$, $\lnot$,
  $\land$, $\lor$, $\limplies$ and substitutions $[\aExp/\aReg]$, $[\aExp/\aLoc]$,
\item there are three binary relations over $\Act\times\Act$:
  $\rmatchesdef$, $\rblocksdef$, and $\rdelaysdef$,
\item there are two subsets of $\Act$, distinguishing
  $\sreaddef$ and $\sreleasedef$ actions.
\end{itemize}

Logical formulae include equations over registers and memory references, such as
$(\aReg{=}\bReg{+}1)$ and $(\aLoc{=}1)$.
% For use in \textsection\ref{sec:tc1}, we also include equations over memory references, such as $(\aLoc{=}1)$.
% I would like to drop this, an be careful about program vs logical syntax
We use expressions as formulae, coercing $\aExp$ to $\aExp{\neq}0$.
% Equations have precedence over logical operators; thus
% $\aReg{=}\aVal\limplies\bReg{>}\bVal$ is read
% $(\aReg{=}\aVal)\limplies(\bReg{>}\bVal)$.  As usual, implication associates to the
% right; thus $\aForm\limplies\bForm\limplies\cForm$ is read
% $\aForm\limplies(\bForm\limplies\cForm)$.
% As usual, implication associates to the right; thus
% $\aReg{=}\aVal\limplies\bReg{>}\bVal\limplies\bForm$ is read
% $(\aReg{=}\aVal)\limplies((\bReg{>}\bVal)\limplies\bForm)$.

% Formulae are subject to substitutions; % of the form $[\aExp/\aReg]$ and
% % $[\aExp/\aLoc]$;
% actions are not.

% \begin{definition}
%   \label{def:independent}
%   We say $\aForm$ is \emph{independent of $\aLoc$} if, for every
%   $\aVal$, $\aForm \vDash \aForm[\aVal/\aLoc] \vDash \aForm$; it is
%   \emph{dependent} otherwise.
%   We say $\aForm$ is \emph{location independent} if it is independent of
%   every location.
%   We say
%   $\aForm$ \emph{implies} $\bForm$ if $\aForm\vDash\bForm$.
%   We say that
We write $\aForm\riff\bForm$ when $\aForm\rimplies\bForm$ and $\bForm\rimplies\aForm$.
We say
$\aForm$ is a \emph{tautology} if $\TRUE \rimplies \aForm$.
% We say that
We say
$\aForm$ is \emph{unsatisfiable} if $\aForm \rimplies \FALSE$, and
\emph{satisfiable} otherwise.
% \end{definition}

\subsection{Actions in This Paper}
\label{sec:actions}
In this paper, we let actions be reads and writes and fences:
\begin{displaymath}
  \aAct,\bAct \BNFDEF \DW[\amode]{\aLoc}[\ascope]{\aVal}[\aThrd]
  \BNFSEP \DR[\amode]{\aLoc}[\ascope]{\aVal}[\aThrd]
  \BNFSEP \DF[\ascope]{\fmode}[\aThrd]
\end{displaymath}
% % writes $\DWP[\amode]{\aLoc}[\ascope]{\aVal}[\aThrd]$,
% reads $\DRP[\amode]{\aLoc}[\ascope]{\aVal}[\aThrd]$,
% and fences $\DFP[\ascope]{\fmode}[\aThrd]$.

We use shorthand when referring to actions.  In definitions, we drop elements
of actions that are existentially quantified.  In examples, we drop elements
of actions, using defaults.
% We write $\DXP[\amode]{}[\ascope]{}[\aThrd]$ to
% stand for $\DWP[\amode]{}[\ascope]{}[\aThrd]$ or
% $\DRP[\amode]{}[\ascope]{}[\aThrd]$.
% 
Let $\lemode$ be the smallest order over access and fence modes such that
%$\mRLX\lemode\mRA\lemode\mSC$ and $\fREL\lemode\fSC$ and $\fACQ\lemode\fSC$.
$\mRLX\lemode\mREL\lemode\mSC$ and $\mRLX\lemode\mACQ\lemode\mSC$.
We write $\DWP[\gemode\mREL]{}{}$ to stand for either $\DWP[\mREL]{}{}$ or
$\DWP[\mSC]{}{}$, and similarly for the other actions and modes.

% We also define shorthand for sets of
% actions using an order on access and fence modes:  
% \begin{align*}
%   \begin{tikzcenter}
%     \node (rlx) at (0, 0) {$\mathstrut\mRLX$};
%     \node (ra)  at (1, 0) {$\mathstrut\mRA$};
%     \node (sc)  at (2, 0) {$\mathstrut\mSC$};
%     \draw[->](rlx)to(ra);
%     \draw[->](ra)to(sc);
%   \end{tikzcenter}
%   &&
%   \begin{tikzcenter}
%     \node (fsc) at (3, 0) {$\mathstrut\fSC$};
%     \node (rel) at (2, -0.2) {$\mathstrut\fREL$};
%     \node (acq) at (2,  0.2) {$\mathstrut\fACQ$};
%     \draw[->](rel)to(fsc);
%     \draw[->](acq)to(fsc);
%   \end{tikzcenter}
% \end{align*}
% We write $\amode\lemode\bmode$ for this order.
% Let $\amode\lubmode\bmode$ denote the least upper bound of $\amode$ and $\bmode$.

% or $\DFP[\ascope]{\amode}[\aThrd]$.
% We write $\DWP[\gemode\mREL]{}{}$ to stand for either
% $\DWP[\mREL]{}{}$ or $\DWP[\mSC]{}{}$, and similarly for other actions and
% modes.

\begin{definition}
  \label{def:actions}
  Actions $\DRP{}{}$ are $\sreaddef$ actions.
  Actions $\DWP[\gemode\mREL]{}{}$ and $\DFP{\gemode\fREL}$ are
  $\sreleasedef$ actions.

  We say $\aAct \rmatchesdef \bAct$ if $\aAct=\DWP{\aLoc}{\aVal}$ and $\bAct=\DRP{\aLoc}{\aVal}$.
  % We say $\DWP{\aLoc}{\aVal} \rmatchesdef \DRP{\aLoc}{\bVal}$ when $\aVal=\bVal$.
  % Action $\DWP{\aLoc}{\aVal} \rmatchesdef \DRP{\aLoc}{\bVal}$ when $\aVal=\bVal$.

  We say $\aAct \rblocksdef \bAct$ if $\aAct=\DWP{\aLoc}{}$ and $\bAct=\DRP{\aLoc}{}$, regardless of value.
  % We say $\DWP{\aLoc}{\aVal} \rblocksdef \DRP{\aLoc}{\bVal}$, for any $\aVal$, $\bVal$.
  % Action $\DWP{\aLoc}{\aVal} \rblocksdef \DRP{\aLoc}{\bVal}$, for any $\aVal$, $\bVal$.

  % Let two actions \emph{overlap} if they access the same location.
  % We say $\aAct \roverlapsdef \bAct$ if they access the same location.

  Let ${\eqreorderco}$ capture write-write, read-write coherence:
  \begin{math}
    {\eqreorderco}
    =
    \{(\DW{\aLoc}{}\Cb \DW{\aLoc}{})\Cc(\DR{\aLoc}{}\Cb \DW{\aLoc}{})\Cc(\DW{\aLoc}{}\Cb \DR{\aLoc}{})\}
  \end{math}.

  Let ${\reorderra}$ capture conflict due to synchronization:
  \begin{math}
    {\reorderra}
    =
    \{(\aAct\Cb             \DW[\gemode\mREL]{}{}     )\Cc
    (\aAct\Cb               \DF{\gemode\fREL}        )\Cc
    (\DR{}{}\Cb             \DF{\gemode\fACQ}        )\Cc
    % (\DR{\aLoc}{}\Cb        \DR[\gemode\mACQ]{\aLoc}{})\Cc
    (\DR[\gemode\mACQ]{}{}\Cb\aAct                    )\Cc
    (\DF{\gemode\fACQ}\Cb   \aAct                    )\Cc
    (\DF{\gemode\fREL}\Cb   \DW{}{}                  )\Cc
    (\DW[\gemode\mREL]{\aLoc}{}\Cb\DW{\aLoc}{})\}
  \end{math}.

  Let ${\eqreordersc}$ capture conflict due to $\mSC$ access:
  \begin{math}
    {\eqreordersc}
    =
    % \{(\DX[\mSC]{}{}\Cb\DX[\mSC]{}{})\}
    \{(\DW[\mSC]{}{}\Cb \DW[\mSC]{}{})\Cc(\DR[\mSC]{}{}\Cb \DW[\mSC]{}{}) \Cc(\DW[\mSC]{}{}\Cb \DR[\mSC]{}{})\Cc(\DR[\mSC]{}{}\Cb \DR[\mSC]{}{})\}
  \end{math}.

  We say $\aAct \rdelaysdef \bAct$ if $\aAct\eqreorderco\bAct$ or $\aAct\reorderra\bAct$ or $\aAct\eqreordersc\bAct$.
\end{definition}

\subsection{The Semantic Domain} %: Pomsets with Predicate Transformers}
\label{sec:pomsets}

\emph{Predicate transformers} are functions on formulae that preserve
logical structure, providing a natural model of sequential composition.
The definition comes from \citet{DBLP:journals/cacm/Dijkstra75}:

\begin{definition}
  \label{def:trans}
  A \emph{predicate transformer} is a %monotone
  function
  $\aTr{}{}:\Formulae\fun\Formulae$ such that
  \begin{multicols}{2}
    \begin{enumerate}[,label=(\textsc{x}\arabic*),ref=\textsc{x}\arabic*]
    \item \label{tr-false}
      $\aTr{}{\FALSE}$ is $\FALSE$,    
    \item \label{tr-and}
      $\aTr{}{\bForm_1\land\bForm_2}$ is $\aTr{}{\bForm_1}\land\aTr{}{\bForm_2}$,    
    \item \label{tr-or}
      $\aTr{}{\bForm_1\lor\bForm_2}$ is $\aTr{}{\bForm_1}\lor\aTr{}{\bForm_2}$, 
    \item \label{tr-implies}
      if $\aForm \rimplies \bForm$, then $\aTr{}{\aForm} \rimplies
      \aTr{}{\bForm}$.
    \end{enumerate}
  \end{multicols}
\end{definition}
\noindent
We consistently use $\bForm$ as the parameter of predicate transformers.
Note that substitutions ($\bForm[\aExp/\aReg]$ and $\bForm[\aExp/\aLoc]$) and
implications on the right ($\aForm\limplies\bForm$) are predicate
transformers.

As discussed in \S\ref{sec:intro}, predicate transformers suffice for
sequentially consistent models, but not relaxed models, where dependency
calculation is crucial.  For dependency calculation, we use a \emph{family}
of predicate transformers, indexed by sets of events. In sequential
composition, we will use $\aTr{\Cdown{\aEv}}{}$ as the predicate transformer
applied to event $\aEv$ where $\bEv\in(\Cdown{\aEv})$ if $\bEv<\aEv$.

\begin{definition}
  \label{def:family}
  A \emph{family of predicate transformers} over $\aEvs$ consists of a
  predicate transformer $\aTr{\bEvs}{}$ for each $\bEvs\subseteq\AllEvents$,
  such that if $\cEvs \cap \aEvs \subseteq \bEvs$ then $\aTr{\cEvs}{\bForm}
  \rimplies \aTr{\bEvs}{\bForm}$.

  We write $\aTr{}{\bForm}$ as an abbreviation of $\aTr{\aEvs}{\bForm}$.
\end{definition}

\begin{definition}
  \label{def:pomset}
  A \emph{pomset with predicate transformers} over $\Act$
  is a tuple $(\aEvs, \labeling, \labelingForm, \aTr{}{}, \aTerm, {\rrfx}, {\le})$ where
  \begin{enumerate}[,label=(\textsc{m}\arabic*),ref=\textsc{m}\arabic*]
    \makecounter{BE}
  \item \label{pom-E} \makecounter{E}
    $\aEvs\subseteq\AllEvents$ is a set of \emph{events},
    \makecounter{Blambda}
  \item \label{pom-lambda} \makecounter{lambda}
    $\labeling: \aEvs \fun \Act$ defines a \emph{label} for each event,
    \makecounter{Bkappa}
  \item \label{pom-kappa} \makecounter{kappa}
    $\labelingForm:\aEvs\fun\Formulae$ defines a \emph{precondition} for each event, such that
    \begin{enumerate}
    \item \label{pom-kappa-sat}
      $\labelingForm(\aEv)$ is satisfiable,
    \end{enumerate}
    \makecounter{Btau}
  \item \label{pom-tau} \makecounter{tau}
    $\aTr{}{}:2^{\AllEvents}\fun\Formulae \fun\Formulae$ is a \emph{family of predicate transformers} over $\aEvs$, 
    \makecounter{Bterm}
  \item \label{pom-term} \makecounter{term}
    $\aTerm:\Formulae$ is a \emph{termination condition}, such that 
    \begin{enumerate}
    \item \label{pom-term-tau}
      $\aTerm\rimplies\aTr{}{\TRUE}$,
    \end{enumerate}
    \makecounter{Brf}
  \item \label{pom-rf} \makecounter{rf}
    ${\rrfx} \subseteq \aEvs\times\aEvs$ is an injective relation capturing
    \emph{reads-from}, such that 
    % \end{enumerate}
    % A pomset is a \emph{candidate} if there is an injective relation
    % ${\rrfx} : \aEvs\times\aEvs$, capturing \emph{reads-from}, such that
    \begin{enumerate}
      % \begin{enumerate}[,label=(\textsc{i}\arabic*),ref=\textsc{i}\arabic*]
      % \item \label{rf-injective}
      %   if $\bEv\xrfx\aEv$ and $\cEv\xrfx\aEv$ then $\bEv=\cEv$, that is, ${\rrfx}$ is injective,
    \item \label{pom-rf-match}
      if $\bEv\xrfx\aEv$ then $\labeling(\bEv) \rmatches \labeling(\aEv)$,
      % \item \label{rf-block}
      %   if $\bEv\xrfx\aEv$ and $\labeling(\cEv) \rblocks \labeling(\aEv)$ then either $\cEv\le\bEv$ or $\aEv\le\cEv$.
      % \item \label{rf-le}
      %   if $\bEv\xrfx\aEv$ then $\bEv\le\aEv$.
    \end{enumerate}
    \makecounter{Ble}
  \item \label{pom-le} \makecounter{le}
    ${\le} \subseteq \aEvs\times\aEvs$, is a partial order capturing
    \emph{causality}, such that
    \begin{enumerate}
    \item \label{pom-le-match}
      if $\bEv\xrfx\aEv$ then either $\bEv\le\aEv$ or $\aEv\le\bEv$,
    \item \label{pom-le-block}
      if $\bEv\xrfx\aEv$ and $\labeling(\cEv) \rblocks \labeling(\aEv)$ then either $\cEv\le\bEv$ or $\aEv\le\cEv$.
    \end{enumerate}
  \end{enumerate}
  A pomset is \emph{complete} if
  % \begin{multicols}{2}
  \begin{enumerate}[,label=(\textsc{c}\arabic*),ref=\textsc{c}\arabic*]
    \setcounter{enumi}{\value{Blambda}}
  \item \label{top-rf}
    if $\labeling(\aEv)$ is a $\sread$ then there is some $\bEv\xrfx\aEv$,
    \setcounter{enumi}{\value{Bkappa}}
  \item \label{top-kappa}
    $\labelingForm(\aEv)$ is a tautology (for every $\aEv\in\aEvs$),
    \setcounter{enumi}{\value{Bterm}}
  \item \label{top-term}
    $\aTerm$ is a tautology.
  \end{enumerate}
  % \end{multicols}
  % \labeltext[\textsc{t}1]{(\textsc{t}1)}{top-term} $\aTerm$ is a tautology and \labeltext[\textsc{t}2]{(\textsc{t}2)}{top-ev} for every $\aEv\in\aEvs$,
  % % $\labelingForm(\aEv)$ is a tautology and
  % % if $\labeling(\aEv)$ is a $\sread$ then there is some $\bEv\xrfx\aEv$.
  % % \begin{multicols}{2}
  % \begin{enumerate}[label=(\textsc{t}\arabic*),ref=\textsc{t}\arabic*]
  %   \setcounter{enumi}{2}
  % \item[]
  %   \begin{enumerate}[leftmargin=0pt]
  %   \item \label{top-kappa}
  %     $\labelingForm(\aEv)$ is a tautology,    
  %   \item \label{top-rf}
  %     if $\labeling(\aEv)$ is a $\sread$ then there is some $\bEv\xrfx\aEv$.
  %   \end{enumerate}
  % \end{enumerate}
  % % \end{multicols}
\end{definition}
\input{fig-sem.tex}

We give the semantics of programs $\sembase{}$ in \reffig{fig:sem}.

Let $\aPS$ range over pomsets, and $\aPSS$ over sets of pomsets.
% Let $\Pom$ be the set of all pomsets.

The model has seven components, which can be daunting at first glance.  To
aid the reader, we use consistent numbering throughout. For example, item $7$
always refers to the order relation.

The core of the model is a pomset, which includes a set of events
\eqref{pom-E}, a labeling \eqref{pom-lambda}, and an order \eqref{pom-le}.
We also include the \emph{reads-from} relation explicitly in the model
\eqref{pom-rf}.

On top of this basic structure, \ref{pom-kappa}--\ref{pom-term} add a layer
of logic.  For each pomset, \ref{pom-term} provides a termination condition.
For each event in a pomset, \ref{pom-kappa} provides a precondition.  For
each set of events in a pomset, \ref{pom-tau} provides a predicate
transformer.  Sequential dependency is calculated by $\labelingForm'_2$ in
the semantics of sequential composition.

Before discussing the details of the model, we note that the semantics
satisfies the expected monoid laws and is closed with respect to
\emph{augmentation}.  Augments include more order and stronger formulae; in
examples, we typically consider pomsets that are augment-minimal.  One
intuitive reading of augment closure is that adding order can only cause
preconditions to weaken.
\begin{lemma}
  \label{lem:monoid}
  \begin{enumerate*}[label=(\alph*),ref=\alph*]
  \item \label{monoid:unit}
    \begin{math}
      \aPSS
      =
      \xLPARP{\aPSS}{\sSKIP}
      =
      \xSEMIP{\aPSS}{\sSKIP}
      =
      \xSEMIP{\sSKIP}{\aPSS}.
    \end{math}

  \\\item \label{monoid:par}
    \begin{math}
      \xLPAR{\xLPARP{\aPSS_1}{\aPSS_2}}{\aPSS_3}
      =
      \xLPAR{\aPSS_1}{\xLPARP{\aPSS_2}{\aPSS_3}}.
    \end{math}

  \\\item \label{monoid:seq}
    \begin{math}
      \xSEMI{\xSEMIP{\aPSS_1}{\aPSS_2}}{\aPSS_3}
      =
      \xSEMI{\aPSS_1}{\xSEMIP{\aPSS_2}{\aPSS_3}}.
    \end{math}

  \\\item
    \begin{math} \label{monoid:ifelse:if:if}
      \xIFTHEN{\aForm}{\aPSS_1}{\aPSS_2}
      =
      \xSEMI{
        \xIFTHEN{\aForm}{\aPSS_1}{}
      }{
        \xIFTHEN{\lnot\aForm}{\aPSS_2}{}
      }
      =
      \xSEMI{
        \xIFTHEN{\lnot\aForm}{\aPSS_2}{}
      }{
        \xIFTHEN{\aForm}{\aPSS_1}{}
      }.
    \end{math}

  \\\item \label{monoid:if:dead}
    \begin{math}
      \xIFTHEN{\aForm}{\aPSS_1}{\aPSS_2}
      =
      \aPSS_1
    \end{math}
    if $\aForm$ is a tautology.

  \\\item \label{monoid:if:if}
    \begin{math}
      \xIFTHEN{\aForm}{
        \xIFTHEN{\bForm}{\aPSS}{}
      }{}
      =
      \xIFTHEN{\aForm\land\bForm}{\aPSS}{}.
    \end{math}

  \\\item \label{monoid:if:seq}
    \begin{math}
      \xIFTHEN{\aForm}{
        \xSEMI{\aPSS_1}{\aPSS_3}
      }{
        \xSEMI{\aPSS_2}{\aPSS_3}
      }
      \supseteq
      \xSEMI{
        \xIFTHEN{\aForm}{\aPSS_1}{\aPSS_2}
      }{
        \aPSS_3
      }.
    \end{math}

  \\\item \label{monoid:seq:if}
    \begin{math}
      \xIFTHEN{\aForm}{
        \xSEMI{\aPSS_1}{\aPSS_2}
      }{
        \xSEMI{\aPSS_1}{\aPSS_3}
      }
      \supseteq
      \xSEMI{
        \aPSS_1
      }{
        \xIFTHEN{\aForm}{\aPSS_2}{\aPSS_3}
      }.
    \end{math}

  \\\item \label{monoid:if:elim}
    \begin{math}
      \xIFTHEN{\aForm}{\aPSS}{\aPSS}
      \supseteq
      \aPSS.
    \end{math}
  \end{enumerate*}      
  \vspace{-.5\baselineskip}
  \begin{proof}
    Straightforward calculation.
    \eqref{monoid:unit} requires \ref{pom-term-tau} for
    the termination condition in $\xSEMIP{\aPSS}{\sSKIP}$.

    \eqref{monoid:seq} requires both
    conjunction closure (\ref{tr-and}, for the termination condition) and disjunction
    closure (\ref{tr-or}, for the predicate transformers themselves).

    \eqref{monoid:ifelse:if:if} requires \ref{seq-le-delays} not impose order
    when $\labelingForm_1(\bEv) \land \labelingForm_2(\aEv)$ is
    unsatisfiable, which in turn requires that $\labelingForm$ calculates
    \emph{weakest} preconditions, rather than simple preconditions (see
    \textsection\ref{sec:if:seq}).

    \eqref{monoid:if:dead} requires \ref{pom-kappa-sat}.

    In \textsection\ref{sec:semca}, we refine the semantics to validate the
    reverse inclusions for \eqref{monoid:if:seq}, \eqref{monoid:seq:if}, and
    \eqref{monoid:if:elim}.
  \end{proof}
\end{lemma}
\begin{definition}
  \label{def:augment}
  $\aPS_2$ is an \emph{augment} of $\aPS_1$ if
  \begin{multicols}{4}
    \begin{enumerate}
    \item $\aEvs_2=\aEvs_1$,
    \item $\labelingAct_2(\aEv)=\labelingAct_1(\aEv)$,
    \item $\labelingForm_2(\aEv) \riff \labelingForm_1(\aEv)$,
    \item $\aTr[2]{\bEvs}{\bForm}\riff \aTr[1]{\bEvs}{\bForm}$,
    \item $\aTerm[2] \riff \aTerm[1]$,
    \item ${\rrfx_2}\supseteq{\rrfx_1}$,
    \item ${\le_2}\supseteq{\le_1}$.
    \end{enumerate}
  \end{multicols}
\end{definition}
\begin{lemma}
  % Suppose $\aPS_1\in\sembase{\aCmd}$.
  If $\aPS_1\in\sembase{\aCmd}$ and $\aPS_2$  augments $\aPS_1$ then $\aPS_2\in\sembase{\aCmd}$.
  % \item If $\aPS_2$ is a downset of $\aPS_1$ then $\aPS_2\in\sembase{\aCmd}$.
  % \end{lemma}

  \vspace{-.5\baselineskip}
  \begin{proof}
    Induction on the definition of $\sembase{}$.
  \end{proof}
\end{lemma}

% \begin{definition}
%   Suppose $\aRel_1:\aEvs_1\times\aEvs_1$ and $\aRel_2:\aEvs_2\times\aEvs_2$.
%   We say $\aRel \rextendsdef{\aRel_1}{\aRel_2}$ if
%   $\aRel\supseteq (\aRel_1\cup \aRel_2)$ and
%   $\aRel\cap(\aEvs_1\times \aEvs_1) = \aRel_1$ and
%   $\aRel\cap(\aEvs_2\times \aEvs_2) = \aRel_2$.
% \end{definition}

% We include empty sets as prep for adding while loops.

% In diagrams, we use different colors for arrows.  We distinguish
% $\rrfx$ edges that are included in order from those that are not.
% \begin{itemize}
% \item \makebox{$\aEv\xrf\bEv$} arises from $\rrfx$, where $\aEv\le\bEv$,
% \item \makebox{$\aEv\xrfint\bEv$} arises from $\rrfx$, where $\lnot(\aEv\le\bEv)$.
% \end{itemize}
% To help the reader understand why order is included, we also different colors
% for arrows induced by order.  We adopt the following conventions:
% \begin{itemize}  
%   % \item relaxed accesses are blue, with a single border,
%   % \item synchronized accesses are red, with a double border,
%   % \item \makebox{$\aEv\xrf\bEv$} arises from fulfillment, where $\aEv$ \emph{matches} $\bEv$,
%   % \item \makebox{$\aEv\xwk\bEv$} arises either from fulfillment, where $\aEv$
%   %   \emph{blocks} $\bEv$, or from prefixing, where $\aEv$ was prefixed before
%   %   $\bEv$ and their actions \emph{conflict},
% \item \makebox{$\aEv\xwk\bEv$} arises from \emph{fulfillment},
% \item \makebox{$\aEv\xpo\bEv$} arises from control/data/address \emph{dependency},
% \item \makebox{$\aEv\xsync\bEv$} arises from \emph{synchronized access}.
% \end{itemize}

\subsection{Pomsets}
\label{sec:ex:pomset}

We first explain the core of model, ignoring the logic (rules 3--5).
%We defer the discussion of $\sLET{}{}$ to \textsection\ref{sec:ex:data}.
We defer discussion of $\sIF{}$ to \textsection\ref{sec:ex:control}.

Reads, writes, and fences map to pomsets with at most one event.  $\SKIP$
maps to the empty pomset.  Ignoring the logic, the definitions are
straightforward.  Note only that $\sembase{\PW{x}{1}}$ can write any value
$\aVal$; the fact that $\aVal$ must be $1$ is captured in the logic (see
\textsection\ref{sec:ex:term}).

The structural rules combine pomsets:  Parallel composition is disjoint
union, inheriting labeling, order and $\rrfx$ from the two sides.  Any
$\rrfx$ edges added between the two sides must also be added to the order
(\ref{par-le-rf1} and \ref{par-le-rf2}).
% 
Sequential composition is similar, with two changes: \ref{seq-E} does not
require disjointness (see \textsection\ref{sec:ex:term}), and
\ref{seq-le-delays} may require order (see example \ref{pub}, below).

Note that reads-from implies order.
\begin{lemma}
  \label{lem:rf:implies:le}
  For any $\aPS$ in the range of $\sembase{}$, $\bEv\xrfx\aEv$ implies
  $\bEv\le\aEv$.

  \vspace{-.5\baselineskip}
  \begin{proof}
    Induction on the definition of $\sembase{}$, using \ref{par-le-rf1}, \ref{par-le-rf2}, and \ref{seq-le-rf}.
  \end{proof}
\end{lemma}
In top-level pomsets, every read must have a matching write in $\rrfx$
\eqref{top-rf}.  Together with \ref{pom-le-match} and \ref{pom-le-block}, the lemma
guarantees that reads are \emph{fulfilled} at top-level, as in
\cite[\textsection 2.7]{DBLP:journals/pacmpl/JagadeesanJR20}.\footnote{The
  basic model would be the same if we move $\rrfx$ from the model itself to
  be existentially quantified in the definition of top-level pomset, along
  with \ref{pom-le-match} and \ref{pom-le-block}.  This was the approach of
  \citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} We include $\rrfx$
  explicitly for use in \textsection\ref{sec:arm2}, where we introduce a
  variant semantics $\frf{\semrr{}}$ for which \reflem{lem:rf:implies:le}
  fails.}

From \refdef{def:actions}, recall that $\aAct \rdelaysdef \bAct$ if
$\aAct\eqreorderco\bAct$ or $\aAct\reorderra\bAct$ or
$\aAct\eqreordersc\bAct$.  \ref{seq-le-delays} guarantees that sequential
order is enforced between conflicting accesses of the same location
($\eqreorderco$), into a release and out of an acquire ($\reorderra$), and
between SC accesses ($\eqreordersc$).  Combined with the fulfillment
requirements (\ref{pom-le-match}, \ref{pom-le-block} and \reflem{lem:rf:implies:le}),
these ensure coherence, publication, subscription and other idioms.  For
example, consider the following:\footnote{We use different colors for arrows representing order:
  \begin{itemize}  
  \item \makebox{$\bEv\xpo\aEv$} arises from control/data/address \emph{dependency} (\ref{seq-kappa}, definition of $\labelingForm'_2(\bEv)$),
  \item \makebox{$\bEv\xsync\aEv$} arises from $\reorderra$ or $\eqreordersc$ \eqref{seq-le-delays},
  \item \makebox{$\bEv\xwki\aEv$} arises from $\eqreorderco$ \eqref{seq-le-delays},
  \item \makebox{$\bEv\xrf\aEv$} arises from \emph{reads-from} \eqref{pom-le-match},
  %\item \makebox{$\bEv\xsyncsc\aEv$} arises from \emph{strong fencing} \eqref{cand-lesync-sc},
  \item \makebox{$\bEv\xwk\aEv$} arises from \emph{blocking} \eqref{pom-le-block}.
  \end{itemize}
  In \textsection\ref{sec:arm2}, it is possible for $\rrf$ to contradict
  $\le$.  In this case, we use a dotted arrow for $\rrf$: $\bEv\xrfint\aEv$
  indicates that $\aEv\le\bEv$.}
\begin{gather*}
  \taglabel{pub}
  \begin{gathered}    
    \PW{x}{0}\SEMI 
    \PW{x}{1}\SEMI \PW[\mREL]{y}{1} \PAR \PR[\mACQ]{y}{r}\SEMI \PR{x}{s}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{wx1}{\DW{x}{1}}{right=of wx0}
        \event{wy1}{\DW[\mREL]{y}{1}}{right=of wx1}
        \event{ry1}{\DR[\mACQ]{y}{1}}{right=2.5em of wy1}
        \event{rx0}{\DR{x}{0}}{right=of ry1}
        \sync{wx1}{wy1}
        \sync{ry1}{rx0}
        \rf{wy1}{ry1}
        \wk[out=170,in=10]{rx0}{wx1}
        \wki{wx0}{wx1}
      \end{tikzinline}}
  \end{gathered}
\end{gather*}
The execution is disallowed due to the cycle.  All of the order shown is
required at top-level: The intra-thread order comes from \ref{seq-le-delays}:
$\DWP{x}{0}\xwki \DWP{x}{1}$ is required by $\eqreorderco$.
$\DWP{x}{1}\xsync \DWP[\mREL]{y}{1}$ and $\DRP[\mACQ]{y}{1}\xsync \DRP{x}{0}$ are
required by $\reorderra$.  The cross-thread order is required by fulfillment:
\ref{top-rf} requires that all top-level reads are in the image of $\xrfx$.
\ref{pom-le-match} ensures that $\DWP[\mREL]{y}{1}\xrfx \DRP[\mACQ]{y}{1}$, and
\reflem{lem:rf:implies:le} subsequently ensures that
$\DWP[\mREL]{y}{1}\le \DRP[\mACQ]{y}{1}$.  The \emph{antidependency}
$\DRP{x}{0}\xwk \DWP{x}{1}$ is required by \ref{pom-le-block}.  (Alternatively,
we could have $\DWP{x}{1}\xwk \DWP{x}{0}$, again resulting in a cycle.)

The semantics gives the expected results for store buffering and load buffering,
as well as litmus tests involving fences and SC access.  The model of
coherence is weaker than C11, in order to support common subexpression
elimination, and stronger than Java, in order to support local reasoning
about data races.  See
\cite[\textsection3.1]{DBLP:journals/pacmpl/JagadeesanJR20} for a discussion.

In the structural rules $\sSEMI{}{}$ and $\sIF{}$, we say that
$\bEv\in\aEvs_1$ and $\aEv\in\aEvs_2$ \emph{coalesce} if $\bEv=\aEv$.

\ref{seq-E} allows \emph{mumbling} \cite{DBLP:journals/iandc/Brookes96} by
coalescing events.  For example
\begin{math}
  \sembase{\PW{x}{1} \SEMI \PW{x}{1}}
\end{math}
includes the singleton pomset
\begin{tikzinlinesmall}[node distance=.5em and 1.5em]
  \event{a1}{\DW{x}{1}}{}
\end{tikzinlinesmall}.  From this it is easy to see that
\begin{math}
  \sembase{\PW{x}{1}\SEMI\PW{x}{1}}
  \supseteq
  \sembase{\PW{x}{1}}
\end{math}
is a valid refinement.  It is equally obvious that
\begin{math}
  \sem{\PW{\aLoc}{1}} 
  \not\supseteq
  \sem{\PW{\aLoc}{1} \SEMI \PW{\aLoc}{1}}
\end{math}
is not a valid refinement, since the latter includes a two-element pomset,
but the former does not.\footnote{These are distinguished by the context:
  \begin{math}
    \hole{} \PAR
    \PR{x}{r} \SEMI
    \PW{x}{2} \SEMI
    \PR{x}{s}\SEMI
    \IF{\aReg {=} \bReg} \THEN \PW{\cLoc}{1} \FI.
  \end{math}}

\subsection{Termination}
\label{sec:ex:term}

In top-level pomsets, \ref{top-term} requires that $\aTerm$ is a tautology,
capturing termination.  Terminated pomsets are often called \emph{complete},
whereas nonterminated pomsets are \emph{incomplete}.

Ignoring predicate transformers, the structural rules, \ref{par-term} and
\ref{seq-term}, take $\aTerm$ to be $\aTerm_1\land\aTerm_2$.  This is as
expected: the program terminates if both subprograms terminate.

The interesting rules are $\sLOAD{}{}$, $\sFENCE{}$, and $\sSTORE{}{}$.

In $\sLOAD{}{}$, there is no restriction on $\aTerm$ for relaxed reads.  From
this, it is easy to see that
\begin{math}
  \sembase{\PR{x}{r}}
  \supseteq
  \sembase{\SKIP}
\end{math}
is a valid refinement (where the default mode is $\mRLX$).

In $\sFENCE{}$, instead, \ref{fence-term} ensures that all fences are included at
top-level.  This also ensures
$\sembase{\PF{\amode}}\not\supseteq\sembase{\IF{\aExp}\THEN\PF{\amode}\FI}$,
since $\sembase{\IF{\aExp}\THEN\PF{\amode}\FI}$ includes the empty set with
termination condition $\lnot\aExp$, but $\sembase{\PF{\amode}}$ can only
include the empty set with termination condition $\FALSE$.

In $\sSTORE{}{}$, \ref{write-term-empty} is similar.  In addition,
\ref{write-term-nonempty} ensures that top-level pomsets do not include bogus
writes.  Suppose $\aPS\in\sembase{\PW{x}{1}}$.  As we noted above, $\aPS$ can
include $(1{=}\aVal\mid\DW{x}{\aVal})$, for any value $\aVal$.  At top-level,
however, \ref{write-term-nonempty} requires that $\aTerm$ implies
$1{=}\aVal$.  In this case, \ref{pom-kappa-sat} would filter the pomset,
since preconditions must be satisfiable.  However, unsatisfiable writes can be become satisfiable via
merging:
% Nonetheless, it is important that
% \ref{write-term-nonempty} specifies $\aTerm\riff\aExp{=}\aVal$ rather than
% $\aTerm\riff\TRUE$, since 
\begin{align*}
  \begin{gathered}
    \PW{x}{1}
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \event{a}{\DW{x}{1}}{}      
      \end{tikzinline}}    
  \end{gathered}
  &&
  \begin{gathered}
    \PW{x}{2}
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \event{b}{2{=}3\mid\DW{x}{3}}{}      
      \end{tikzinline}}    
  \end{gathered}
  &&
  \begin{gathered}
    \IF{\aExp}\THEN\PW{x}{3}\FI
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \event{c}{\aExp\mid\DW{x}{3}}{}      
      \end{tikzinline}}    
  \end{gathered}
\end{align*}
By merging, the semantics allows the following:
\begin{align*}
  \begin{gathered}
    \PW{x}{1}
    \SEMI
    \PW{x}{2}
    \SEMI
    \IF{\aExp}\THEN\PW{x}{3}\FI
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \event{a}{\DW{x}{1}}{}      
        \event{c}{\aExp\mid\DW{x}{3}}{right=of a}
        \wki{a}{c}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}
This pomset is incomplete, however, since $\aTerm\riff2{=}3$.

\subsection{Data Dependencies, Preconditions, and Predicate Transformers}
\label{sec:ex:data}

% Predicate transformers affect preconditions in the definition of
% $\labelingForm'_2$, which is used in \ref{seq-kappa2} and \ref{seq-kappa12}
% to calculate dependencies.  In this subsection we ignore predicate
% transformers, taking $\labelingForm'_2=\labelingForm_2$.

In top-level pomsets, \ref{top-kappa} requires that every precondition
$\labelingForm(\aEv)$ is a tautology.

Preconditions are discharged during sequential composition by applying
predicate transformers $\aTr[1]{}{}$ from the left to preconditions
$\labelingForm_2(\aEv)$ on the right.  The specific rules are
\ref{seq-kappa2} and \ref{seq-kappa12}, which use the transformed predicate
$\labelingForm'_2(\aEv)=\aTr[1]{\Cdown{\aEv}}{\labelingForm_2(\aEv})$, where
$\Cdown{\aEv}=\{ \cEv \mid \cEv \lt \aEv \}$ is the set of events that
precede $\aEv$ in causal order.  We call $\Cdown{\aEv}$ the \emph{dependent
  set} for $\aEv$.  Then $\aEvs\setminus(\Cdown{\aEv})$ is the
\emph{independent set}. % for $\aEv$.

% Sequential composition uses the usual rule for composition of predicate
% transformers (but preserving the indexing set). For the pomset, we take the
% union of their events, preserving actions, but crucially in
% cases~\ref{seq-kappa2} and~\ref{seq-kappa12} we apply a predicate transformer
% $\aTr[1]{\cEvs}{}$ from the left-hand side to a precondition
% $\labelingForm_2(\aEv)$ from the right-hand side to build the precondition
% $\labelingForm'_2(\aEv)$.  The indexing set $\cEvs$ for the predicate
% transformer is $\{ \cEv\mid \cEv<\aEv \}$, so can depend on the causal order.

Before looking at the details, it is useful to have a high-level view of how
nontrivial preconditions and predicate transformers are introduced.  (We
discuss address dependencies in \textsection\ref{sec:addr}.)
\begin{multicols}{2}
  Preconditions are introduced in: 
  \begin{itemize}
  \item[\eqref{seq-kappa-release}] for release actions,
  \item[\eqref{if-kappa}] for control dependencies, 
  \item[\eqref{write-kappa}] for data dependencies on writes.
  \end{itemize}

  Predicate transformers are introduced in:
  \begin{itemize}
  \item[\eqref{read-tau-dep}] for reads in the dependent set,
  \item[\eqref{read-tau-ind}] for reads in the independent set,
  \item[\eqref{write-tau}] for writes.
  \end{itemize}
  % \item These are glued together in \ref{seq-kappa}.
  %   \begin{itemize}
  %   \item \ref{seq-kappa2}  handles non-coalescing events,
  %   \item \ref{seq-kappa12} handles coalescing events.
  %   \end{itemize}
\end{multicols}

The rules track dependencies.  We discuss data dependencies \eqref{write-kappa}
here and control dependencies \eqref{if-kappa} in
\textsection\ref{sec:ex:control}.  Unless otherwise noted, we assume pomsets
are \emph{complete} and \emph{augment-minimal}.  We do not discuss
\ref{seq-kappa-release} further.  It simply ensures that all writes are
present before a release, even for incomplete pomsets (see
\textsection\ref{sec:ex:term}).

A simple example of a data dependency is a pomset
$\aPS\in\sem{\PR{x}{r}\SEMI \PW{y}{r}}$.  If $\aPS$ is complete, it must have
two events. Then $\sSEMI{}{}$ requires that there are %$\aVal,\bVal\in\Val$ and
$\aPS_1\in \sem{\PR{x}{r}}$ and $\aPS_2\in \sem{\PW{y}{r}}$ of the form:
\begin{align*}
  \tag{$\ddagger$}\label{eq2}
  \begin{gathered}
    \PR{x}{r} 
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \xform{xd}{\aVal{=}r\limplies\bForm}{}
        \eventr{\bEv}{a1}{\DR{x}{\aVal}}{left=of xd}
        \xform{xi}{(x{=}r\lor\aVal{=}r)\limplies\bForm}{left=.5em of a1}
        \xo{a1}{xd}
      \end{tikzinline}}    
  \end{gathered}
  &&
  \begin{gathered}
    \PW{y}{r}
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \xform{xd}{\bForm[r/y]}{}
        \eventr{\aEv}{a2}{r{=}\bVal\mid\DW{y}{\bVal}}{left=of xd}      
        \xform{xi}{\bForm[r/y]}{left=.5em of a2}
        \xo{a2}{xd}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}
First we consider the case that $\aVal=\bVal$.  For example if $\aVal=\bVal=1$, we have:
\begin{align*}
  \begin{gathered}
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \xform{xd}{1{=}r\limplies\bForm}{}
        \eventr{\bEv}{a1}{\DR{x}{1}}{left=of xd}
        \xform{xi}{(x{=}r\lor1{=}r)\limplies\bForm}{left=.5em of a1}
        \xo{a1}{xd}
      \end{tikzinline}}    
  \end{gathered}
  &&
  \begin{gathered}
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \xform{xd}{\bForm[r/y]}{}
        \eventr{\aEv}{a2}{r{=}1\mid\DW{y}{1}}{left=of xd}      
        \xform{xi}{\bForm[r/y]}{left=.5em of a2}
        \xo{a2}{xd}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}
For the read, the dependent transformer $\smash{\aTr[1]{\{\bEv\}}{}}$ is
$1{=}r\limplies\bForm$; the independent transformer $\aTr[1]{\emptyset}{}$ is
$(x{=}r\lor1{=}r)\limplies\bForm$.  These are determined by
\ref{read-tau-dep} and \ref{read-tau-ind}, respectively.  For the write, both
% the dependent transformer
$\smash{\aTr[2]{\{\aEv\}}{}}$ and
% the independent transformer
$\aTr[2]{\emptyset}{}$ are $\bForm[r/y]$, as are determined by
\ref{write-tau}.
%
Combining these into a single pomset, we have:
\begin{align*}
  \begin{gathered}
    \PR{x}{r}\SEMI \PW{y}{r}
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \xform{xd}{1{=}r\limplies\bForm[r/y]}{}
        \eventr{\bEv}{a1}{\DR{x}{1}}{left=of xd}
        \xform{xi}{(x{=}r\lor1{=}r)\limplies\bForm[r/y]}{left=.5em of a1}
        \eventr{\aEv}{a2}{\aForm\mid\DW{y}{1}}{right=of xd}      
        \xo{a1}{xd}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}
By \ref{seq-tau}, predicate transformers are determined by composition; thus
$\aTr{\bEvs}{\bForm}$ is $\aTr[1]{\bEvs}{\aTr[2]{\bEvs}{\bForm}}$.  Since the
transformer does not depend on whether the write is included, we do not draw
dependencies for the write in the diagram.

Turning to the precondition $\aForm$ on the write, recall that in order for
$\aEv$ to participate in a top-level pomset, the precondition $\aForm$ must
be a tautology at top-level.  There are two possibilities.
\begin{itemize}
\item If $\bEv\le\aEv$ then we apply the dependent transformer and
  $\aForm=(1{=}r\limplies r{=}1)$, a tautology.
\item If $\bEv\not\le\aEv$ then we apply the independent transformer and
  $\aForm=((x{=}r\lor1{=}r)\limplies r{=}1)$.  Under the assumption that $r$
  is bound, this is logically equivalent to $(x{=}1)$. (We make this more
  precise in \textsection\ref{sec:semreg}.)
\end{itemize}
Eliding transformers, the two outcomes are:
\begin{align*}
  \begin{gathered}
    \PR{x}{r}\SEMI \PW{y}{r}
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \eventr{\bEv}{a1}{\DR{x}{1}}{}
        \eventr{\aEv}{a2}{\DW{y}{1}}{right=of a1}
        \po{a1}{a2}
      \end{tikzinline}}    
  \end{gathered}
  &&
  \begin{gathered}
    \PR{x}{r}\SEMI \PW{y}{r}
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \eventr{\bEv}{a1}{\DR{x}{1}}{}
        \eventr{\aEv}{a2}{x{=}1\mid\DW{y}{1}}{right=of a1}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}
The independent case on the right can only participate in a top-level pomset
if the precondition $(x{=}1)$ is discharged.  To do so, we must prepend a
pomset $\aPS_0$ that writes $1$ to $x$:
\begin{align*}
  \begin{gathered}
    \PW{x}{1}
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \xform{xd}{\bForm[1/x]}{}
        \eventr{\cEv}{a2}{1{=}1\mid\DW{x}{1}}{left=of xd}      
        \xform{xi}{\bForm[1/x]}{left=.5em of a2}
        \xo{a2}{xd}
      \end{tikzinline}}    
  \end{gathered}
  &&
  \begin{gathered}
    \PW{x}{1}\SEMI \PR{x}{r}\SEMI \PW{y}{r}
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \eventr{\cEv}{a0}{1{=}1\mid\DW{x}{1}}{}
        \eventr{\bEv}{a1}{\DR{x}{1}}{right=of a0}
        \eventr{\aEv}{a2}{1{=}1\mid\DW{y}{1}}{right=of a1}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}
Here we apply the predicate transformer $\aTr[0]{\emptyset}{}$ to $(x{=}1)$,
resulting in the tautology $(1{=}1)$.

Now suppose that $\aVal\neq\bVal$ in \eqref{eq2}.  Again there are two
possibilities, where we take $\aVal=0$ and $\bVal=1$:
\begin{align*}
  \begin{gathered}
    \PR{x}{r}\SEMI \PW{y}{r}
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \eventr{\bEv}{a1}{\DR{x}{0}}{}
        \eventr{\aEv}{a2}{0{=}r\limplies r{=}1\mid\DW{y}{1}}{right=of a1}
        \po{a1}{a2}
      \end{tikzinline}}    
  \end{gathered}
  &&
  \begin{gathered}
    \PR{x}{r}\SEMI \PW{y}{r}
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \eventr{\bEv}{a1}{\DR{x}{0}}{}
        \eventr{\aEv}{a2}{(x{=}r\lor0{=}r)\limplies r{=}1\mid\DW{y}{1}}{right=of a1}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}
Assuming that $r$ is bound, both preconditions on $\aEv$ are unsatisfiable.

If a write is independent of a read, then clearly no order is imposed between
them.  For example, the precondition of $\aEv$ is a tautology in:
\begin{align*}
  \begin{gathered}
    \PR{x}{r}\SEMI \PW{y}{1}
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \xform{xd}{0{=}r\limplies\bForm[r/y]}{}
        \eventr{\bEv}{a1}{\DR{x}{0}}{left=of xd}
        \xform{xi}{(x{=}r\lor0{=}r)\limplies\bForm[r/y]}{left=.5em of a1}
        \eventr {\aEv}{a2}{(x{=}r\lor0{=}r)\limplies 1{=}1\mid\DW{y}{1}}{right=of xd}      
        \xo{a1}{xd}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}
% \begin{gather*}
%   \PR{x}{r}\SEMI \PW{y}{1}
%   \\
%   \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
%       \event{rx0}{\DR{x}{0}}{}
%       \event{wy1}{1{=}1 \mid \DW{y}{1}}{right=of rx0}
%     \end{tikzinline}}
% \end{gather*}

\subsection{Control Dependencies}
\label{sec:ex:control}

In $\sIF{\aForm}\sTHEN\aPSS_1\sELSE\aPSS_2\sFI$, the predicate transformer
\eqref{if-tau} is
$(\aForm\land\aTr[1]{\bEvs}{\bForm})\lor(\neg\aForm\land\aTr[2]{\bEvs}{\bForm})$,
%$\aTerm \rimplies (\aForm\land\aTerm[1])\lor(\neg\aForm\land\aTerm[2])$,
which is the disjunctive equivalent of \citeauthor{DBLP:journals/cacm/Dijkstra75}'s
conjunctive formulation: $(\aForm\limplies\aTr[1]{\bEvs}{\bForm})\land(\neg\aForm\limplies\aTr[2]{\bEvs}{\bForm})$.
% $\aTerm \rimplies (\aForm\limplies\aTerm[1])\land(\neg\aForm\limplies\aTerm[2])$.
%  \citet{DBLP:journals/cacm/Dijkstra75} used conjunctive normal form.
%  Recall that
%  \href{https://www.wolframalpha.com/input/?i=\%28a+and+b\%29+or+\%28not+a+and+c\%29}{$(\aForm\land\cForm_1)\lor(\neg\aForm\land\cForm_2)$}
%  is logically equivalent to
%  \href{https://www.wolframalpha.com/input/?i=\%28a+implies+b\%29+and+\%28\%28not+a\%29+implies+c\%29}{$(\aForm\limplies\cForm_1)\land(\neg\aForm\limplies\cForm_2)$}.

This semantics validates dead code elimination:
if $\aExp{\neq}0$ is a tautology then
\begin{math}
  \sem{\IF{\aExp}\THEN\aCmd_1\ELSE\aCmd_2\FI} \supseteq
  \sem{\aCmd_1}.
\end{math}
The reverse inclusion does not hold.

For events from $\aEvs_1$, \ref{if-kappa1} requires
$\aForm\land\labelingForm_1(\aEv)$. For events from $\aEvs_2$,
\ref{if-kappa2} requires $\lnot\aForm\land\labelingForm_2(\aEv)$.  For
coalescing events in $\aEvs_1\cap\aEvs_2$, \ref{if-kappa12} requires
$(\aForm\land\labelingForm_1(\aEv)) \lor
(\lnot\aForm\land\labelingForm_2(\aEv))$.  This semantics allows common code
to be lifted out of a conditional, validating the transformation
\begin{math}
  \sem{\IF{\aExp}\THEN\aCmd\ELSE\aCmd\FI} \supseteq
  \sem{\aCmd}.
\end{math}
%
The use of $\rextendsdef{}{}$ in \ref{if-le-extends} and \ref{if-rf-extends}
ensures that no new order is introduced between events in
$\aEvs_1\cap\aEvs_2$ when coalescing; see \textsection\ref{sec:arm2}.

By allowing events to coalesce, \ref{if-kappa12} ensures that control
dependencies are calculated semantically.  For example, consider
$\aPS\in\sem{\IF {r{=}1} \THEN \PW{y}{r} \ELSE \PW{y}{1}\FI}$, which is build
from $\aPS_1\in\sem{\PW{y}{r}}$ and $\aPS_2\in\sem{\PW{y}{1}}$ such as:
\begin{align*}
  \begin{gathered}
    \PW{y}{r}
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \eventr{e}{wy1}{r{=}1 \mid \DW{y}{1}}{}
      \end{tikzinline}}
  \end{gathered}
  &&
  \begin{gathered}
    \PW{y}{1}
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \eventr{e}{wy1}{1{=}1 \mid \DW{y}{1}}{}
      \end{tikzinline}}
  \end{gathered}
  &&
  \begin{gathered}
    \IF{r{=}1} \THEN \PW{y}{r} \ELSE \PW{y}{1}\FI
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \eventr{e}{wy1}{(r{=}1 \limplies r{=}1) \land (r{\ne}1 \limplies 1{=}1) \mid \DW{y}{1}}{}
      \end{tikzinline}}
  \end{gathered}  
\end{align*}
Here, the precondition in the combined pomset is a tautology, independent of $r$.

Control dependencies are eliminated in the same way as data dependencies.
For example:
\begin{align*}
  \begin{gathered}
    \PR{x}{r} 
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \xform{xd}{\aVal{=}r\limplies\bForm}{}
        \eventr{\bEv}{a1}{\DR{x}{\aVal}}{left=of xd}
        \xform{xi}{(x{=}r\lor\aVal{=}r)\limplies\bForm}{left=.5em of a1}
        \xo{a1}{xd}
      \end{tikzinline}}    
  \end{gathered}
  &&
  \begin{gathered}
    \IF{r{=}1}\THEN\PW{y}{1}\FI
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \xform{xd}{r{=}1\limplies\bForm[1/y]}{}
        \eventr{\aEv}{a2}{r{=}1\mid\DW{y}{\bVal}}{left=of xd}      
        \xform{xi}{r{=}1\limplies\bForm[1/y]}{left=.5em of a2}
        \xo{a2}{xd}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}
Reasoning as we did for \eqref{eq2} in \textsection\ref{sec:ex:data}, there are two possibilities:
\begin{align*}
  \begin{gathered}
    \PR{x}{r}\SEMI \IF{r{=}1}\THEN\PW{y}{1}\FI
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \eventr{\bEv}{a1}{\DR{x}{1}}{}
        \eventr{\aEv}{a2}{\DW{y}{1}}{right=of a1}
        \po{a1}{a2}
      \end{tikzinline}}    
  \end{gathered}
  &&
  \begin{gathered}
    \PR{x}{r}\SEMI \IF{r{=}1}\THEN\PW{y}{1}\FI
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \eventr{\bEv}{a1}{\DR{x}{1}}{}
        \eventr{\aEv}{a2}{x{=}1\mid\DW{y}{1}}{right=of a1}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}

As another example, consider JMM causality test case 1 \cite{PughWebsite}:
\begin{gather*}
  \begin{gathered}
    \PW{x}{0} \SEMI
    (\PR{x}{r}\SEMI\IF{r{\geq}0}\THEN \PW{y}{1} \FI
    \PAR
    \PW{x}{y})
    \\[-1ex]
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{rx1}{\DR{x}{1}}{right=3em of wx0}
        \event{wy1}{\aForm\mid\DW{y}{1}}{right=of rx1}
        \event{ry1}{\DR{y}{1}}{right=3em of wy1}
        \event{wx1}{\DW{x}{1}}{right=of ry1}
        \po{ry1}{wx1}
        \rf[out=-168,in=-12]{wx1}{rx1}
        \rf{wy1}{ry1}
        \wk[out=10,in=170]{wx0}{wx1}
        \wki{wx0}{rx1}
      \end{tikzinline}}
    % \hbox{\begin{tikzinline}[node distance=1.5em and 2em]
    %   \event{wy1}{0\geq0\mid\DW{y}{1}}{}
    %   \event{rx1}{\DR{x}{1}}{left=of wy1}
    %   \event{wx0}{\DW{x}{0}}{below left=-.2em and 2em of rx1}
    %   \event{ry1}{\DR{y}{1}}{below=of rx1}
    %   \event{wx1}{\DW{x}{1}}{right=of ry1}
    %   \po{ry1}{wx1}
    %   \rf[out=140,in=-20]{wx1}{rx1}
    %   \rf{wy1}{ry1}
    %   \wk[out=10,in=155]{wx0}{wx1}
    %   \wk{wx0}{rx1}
    % \end{tikzinline}}
  \end{gathered}
\end{gather*}
The precondition $\aForm$ is
\begin{math}
  \PBR{(1{=}r \lor x{=}r) \limplies r{\geq}0}[0/x]
\end{math}
which is 
\begin{math}
  \PBR{(1{=}r \lor 0{=}r) \limplies r{\geq}0}
\end{math}
which is a tautology.

\subsection{Reordering Transformations}
\label{sec:ex:valid}

% When $\sem{\aCmd} \supseteq \sem{\aCmd'}$, we say that $\aCmd'$ is a
% \emph{valid transformation} of $\aCmd$.
% In this subsection, we show the
% validity of specific optimizations.

The semantics validates many peephole optimizations.  Most apply only to
relaxed access.
\begin{align*}
  %\taglabel{RR}
  \sembase{\PR{\aLoc}{\aReg} \SEMI \PR{\bLoc}{\bReg}} &=
  \sembase{\PR{\bLoc}{\bReg}\SEMI \PR{\aLoc}{\aReg}} &&\text{if } \aReg\neq\bReg
  \\
  %\taglabel{WW}
  \sembase{\aLoc \GETS \aExp \SEMI \bLoc  \GETS \bExp} &=
  \sembase{\bLoc  \GETS \bExp\SEMI \aLoc \GETS \aExp} &&\text{if } \aLoc\neq\bLoc
  \\
  %\taglabel{RW}
  \sembase{\aLoc \GETS \aExp  \SEMI \PR{\bLoc}{\bReg}} &=
  \sembase{\PR{\bLoc}{\bReg} \SEMI\aLoc \GETS \aExp} &&\text{if }
  \aLoc\neq\bLoc \textand \bReg\not\in\free(\aExp)%\disjoint{{\free(\aLoc \GETS \aExp)}}{{\free(\PR{\bLoc}{\bReg})}}
\end{align*}
% \ref{WW}, \ref{RW} and \ref{RR} require that two sides of the semicolon
% have disjoint ids; for example, \ref{RW} requires $\disjoint{{\free(\aReg
% \GETS \aLoc)}}{{\free(\bLoc \GETS \bExp)}}$. 
% \ref{RR} requires either $\aReg\neq\bReg$ or
% $\aLoc=\bLoc$.  \ref{WW} and \ref{RW} require that two sides of the
% semicolon have disjoint ids; for example, \ref{RW} requires
% $\disjoint{{\free(\PR{\aLoc}{\aReg})}}{{\free(\bLoc \GETS \bExp)}}$.
%\ref{5} imposes no order between events in \ref{RR}--\ref{RW}.  %Note that \ref{RR} allows aliasing.
Here $\free(\aCmd)$ is the set of locations and registers that occur in $\aCmd$.
Using augmentation closure, the semantics also validates roach-motel reorderings \cite{SevcikThesis}.  For
example, on read/write pairs:
  \begin{align*}
    %\tag{\textsc{roach1}}\label{AcqW}
    \sembase{x^\amode \GETS \aExp \SEMI\PR{y}{\bReg}} &\supseteq
    \sembase{\PR{y}{\bReg}  \SEMI x^\amode\GETS \aExp} 
    &&\text{if }
    \aLoc\neq\bLoc \textand \bReg\not\in\free(\aExp)%\disjoint{{\free(\aLoc \GETS \aExp)}}{{\free(\PR{\bLoc}{\bReg})}}
    \\
    %\tag{\textsc{roach2}}\label{RelW}
    \sembase{x \GETS \aExp \SEMI\PR[\amode]{y}{\bReg}} &\supseteq
    \sembase{\PR[\amode]{y}{\bReg}  \SEMI x\GETS \aExp} 
    &&\text{if }
    \aLoc\neq\bLoc \textand \bReg\not\in\free(\aExp)%\disjoint{{\free(\aLoc \GETS \aExp)}}{{\free(\PR{\bLoc}{\bReg})}}
  \end{align*}

% As expected, %sequential and
% parallel composition commutes with conditionals and declarations, and
% conditionals and declarations commute with each other.  For example,
% we have \emph{scope extrusion}~\cite{Milner:1999:CMS:329902}:
% \begin{align*}
%   \taglabel{SE}
%   \sem{\aCmd\PAR \VAR\aLoc\SEMI\bCmd} &=
%   \sem{\VAR\aLoc\SEMI(\aCmd\PAR\bCmd)}
%   &&\text{if } \aLoc\not\in\free(\aCmd)
% \end{align*}

\subsection{Conditional and Coherence}
\label{sec:if:seq}

[This is out of date.]

\begin{lemma}
  \label{lem:if:seq}
  \begin{math}
    \xIFTHEN{\aForm}{\aPSS_1}{\aPSS_2}
    \supseteq
    \xSEMI{
      \xIFTHEN{\aForm}{\aPSS_1}{}
    }{
      \xIFTHEN{\lnot\aForm}{\aPSS_2}{}
    }
  \end{math}

  \begin{math}
    \xIFTHEN{\aForm}{\aPSS_1}{\aPSS_2}
    \supseteq
    \xSEMI{
      \xIFTHEN{\lnot\aForm}{\aPSS_2}{}
    }{
      \xIFTHEN{\aForm}{\aPSS_1}{}
    }
  \end{math}
\end{lemma}

Reverse direction does not hold, due to \ref{seq-le-delays}.
\begin{enumerate}
\item[\eqref{seq-le-delays}] if
  $\labeling_1(\bEv) \rdelays \labeling_2(\aEv)$ then $\bEv\le\aEv$.  
\end{enumerate}
 An alternate phrasing might be attractive:
\begin{enumerate}
\item[{\labeltext[\textsc{s}7c$'$]{(\textsc{s}7c$'$)}{seq-le-delays'}}] if
  $\labeling_1(\bEv) \rdelays \labeling_2(\aEv)$ and
  $\labelingForm(\bEv)\land\labelingForm(\aEv)$ is satisfiable then
  $\bEv\le\aEv$.
\end{enumerate}
But \ref{seq-le-delays'} is incompatible with the ability to strengthen
preconditions using augment closure.  Consider the following.
\begin{align*}
  \begin{gathered}[t]
    \IF{r}\THEN\PW{x}{2}\FI
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \event{a1}{r{\neq}0\mid\DW{x}{2}}{}
      \end{tikzinline}}    
  \end{gathered}
  &&
  \begin{gathered}[t]
    \PW{x}{1}
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \event{a2}{            \DW{x}{1}}{}
      \end{tikzinline}}    
  \end{gathered}
  &&
  \begin{gathered}[t]
    \PW{x}{2}
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \event{a3}{            \DW{x}{2}}{}
      \end{tikzinline}}    
  \end{gathered}
  &&
  \begin{gathered}[t]
    \IF{\BANG r}\THEN\PW{x}{1}\FI
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \event{a4}{r{=}0   \mid\DW{x}{1}}{}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}
% \begin{align*}
%   \begin{gathered}[t]
%     \IF{r}\THEN\PW{x}{2}\FI
%     \SEQ
%     \PW{x}{1}
%     \SEQ
%     \PW{x}{2}
%     \SEQ
%     \IF{\BANG r}\THEN\PW{x}{1}\FI
%     \\
%     \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
%         \event{a1}{r{\neq}0\mid\DW{x}{2}}{}
%         \event{a2}{            \DW{x}{1}}{right=of a1}
%         \event{a3}{            \DW{x}{2}}{right=of a2}
%         \event{a4}{r{=}0   \mid\DW{x}{1}}{right=of a3}
%       \end{tikzinline}}    
%   \end{gathered}
% \end{align*}
Augmenting the middle preconditions and then using sequential composition, we have:
\begin{align*}
  \begin{gathered}[t]
    \IF{r}\THEN\PW{x}{2}\FI
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \event{a1}{r{\neq}0\mid\DW{x}{2}}{}
      \end{tikzinline}}    
  \end{gathered}
  &&
  \begin{gathered}[t]
    \PW{x}{1}
    \SEQ
    \PW{x}{2}
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \event{a2}{r{\neq}0\mid\DW{x}{1}}{}
        \event{a3}{r{=}0   \mid\DW{x}{2}}{right=of a1}
      \end{tikzinline}}    
  \end{gathered}
  &&
  \begin{gathered}[t]
    \IF{\BANG r}\THEN\PW{x}{1}\FI
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \event{a4}{r{=}0   \mid\DW{x}{1}}{}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}
Note that \ref{seq-le-delays'} does not require any order between the two
writes of the middle pomset.
% \begin{align*}
%   \begin{gathered}[t]
%     \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
%         \event{a1}{r{\neq}0\mid\DW{x}{2}}{}
%         \event{a2}{r{=}0   \mid\DW{x}{1}}{right=of a1}
%         \event{a3}{r{\neq}0\mid\DW{x}{2}}{right=of a2}
%         \event{a4}{r{=}0   \mid\DW{x}{1}}{right=of a3}
%       \end{tikzinline}}    
%   \end{gathered}
% \end{align*}
Merging left and right, we have:
\begin{align*}
  \begin{gathered}[t]
    \IF{r}\THEN\PW{x}{2}\FI
    \SEQ
    \PW{x}{1}
    \SEQ
    \PW{x}{2}
    \SEQ
    \IF{\BANG r}\THEN\PW{x}{1}\FI
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \event{a1}{\DW{x}{2}}{}
        \event{a4}{\DW{x}{1}}{right=of a1}
        \wki{a1}{a4}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}
As shown by the following single-threaded code, allowing this outcome would violate \drfsc{}.
\begin{align*}
  \begin{gathered}[t]
    \PW{y}{1}
    \SEQ
    \PR{y}{r}
    \SEQ
    \IF{r}\THEN\PW{x}{2}\FI
    \SEQ
    \PW{x}{1}
    \SEQ
    \PW{x}{2}
    \SEQ
    \IF{\BANG r}\THEN\PW{x}{1}\FI
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \event{a1}{\DW{x}{2}}{}
        \event{a4}{\DW{x}{1}}{right=of a1}
        \wki{a1}{a4}
        \event{b2}{\DR{y}{1}}{left=of a1}
        \event{b1}{\DW{y}{1}}{left=of b2}
        \rf{b1}{b2}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}

To validate the reverse direction of \reflem{lem:if:seq}, it may be tempting
to define the semantics using \emph{weakest} preconditions, rather than
preconditions.  But in this case the notion of program refinement could not be
simple set inclusion---for example, in general we would \emph{not} have
\begin{math}
  \aPSS_1
  \supseteq
  \xIFTHEN{\aForm}{\aPSS_1}{}.
\end{math}

As a result, we leave \reflem{lem:if:seq} as an inequation.  The equational
form may be valid using some notion of \emph{observational} or
\emph{contextual} refinement, but we do not pursue that here.

\subsection{Associativity and Skolemization}
\label{sec:ex:assoc}

The predicate transformers we have chosen for \ref{read-tau-dep} and
\ref{read-tau-ind} are different from the ones used traditionally, which are
written using substitution \jjr{}.  Attempting to write \ref{read-tau-dep} in
this style we would have:
\begin{enumerate}[topsep=0pt]
\item[{\labeltext[\textsc{r}4a$'$]{(\textsc{r}4a$'$)}{read-tau-dep-p}}]
  if $(\aEvs\cap\bEvs)\neq\emptyset$ then
  \begin{math}
    \aTr{\bEvs}{\bForm} \riff    
    \bForm[\aVal/\aReg]
  \end{math},    
\end{enumerate}
Recall that \ref{read-tau-empty} says that $\bForm$ must be independent of
$\aReg$ in order to appear in a top-level pomset: if $\aEvs=\emptyset$ then
\begin{math}
  \aTr{\bEvs}{\bForm} \riff
  % \PBR{\aVal{=}\aReg \lor \aLoc{=}\aReg} \limplies
  \bForm.
\end{math}
This choice for \ref{read-tau-empty} is forced by \refdef{def:family}, which
states that the predicate transformer for a small subset of $\aEvs$ must
imply the transformer for a larger subset.

Sadly, this definition fails associativity.

Consider the following, eliding transformers for the writes (``$\BANG$''
represents logical negation):
\begin{align*}
  \begin{gathered}[t]
    \PR{y}{r}
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1em]
        \event{a}{\DR{y}{1}}{}
        \xform{xd}{1{=}r\limplies\bForm}{right=of a}
        \xform{xi}{(y{=}r\lor1{=}r)\limplies\bForm}{left=.5em of a}
        \xo{a1}{xd}
      \end{tikzinline}}    
  \end{gathered}
  &&
  \begin{gathered}[t]
    \PW{x}{\BANG r} 
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1em]
        \event{b}{r{=}0\mid\DW{x}{1}}{}
      \end{tikzinline}}    
  \end{gathered}
  &&
  \begin{gathered}[t]
    \PW{x}{\BANG\BANG r}
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1em]
        \event{b}{r{\neq}0\mid\DW{x}{1}}{}
      \end{tikzinline}}    
  \end{gathered}
  &&
  \begin{gathered}[t]
    \PW{x}{0} 
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1em]
        \event{b}{\DW{x}{0}}{}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}
Coalescing the writes and associating to the right,
we have the following, since $(r{=}0 \lor r{\neq}0)\riff\TRUE$:
\begin{align*}
  \begin{gathered}[t]
    \PR{y}{r}
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1em]
        \event{a}{\DR{y}{1}}{}
      \end{tikzinline}}    
  \end{gathered}
  &&
  \begin{gathered}[t]
    \PW{x}{\BANG r} 
    \SEMI \PW{x}{\BANG\BANG r} 
    \SEMI \PW{x}{0} 
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1em]
        \event{b}{\DW{x}{1}}{}
        \event{c}{\DW{x}{0}}{right=of b}
        \wki{b}{c}
      \end{tikzinline}}    
  \end{gathered}
  &&
% \end{align*}  
% The precondition of $\DWP{x}{1}$ is a tautology, thus we have:
% \begin{align*}
  \begin{gathered}[t]
    \PR{y}{r}
    \SEMI (\PW{x}{\BANG r} 
    \SEMI \PW{x}{\BANG\BANG r} 
    \SEMI \PW{x}{0} )
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1em]
        \event{a}{\DR{y}{1}}{}
        \event{b}{\DW{x}{1}}{right=of a}
        \event{c}{\DW{x}{0}}{right=of b}
        \wki{b}{c}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}
The precondition of $\DWP{x}{1}$ is a tautology.
Associating to the left and the coalescing, instead:
\begin{align*}
  \begin{gathered}[t]
    \PR{y}{r}
    \SEMI \PW{x}{\BANG r} 
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1em]
        \event{a}{\DR{y}{1}}{}
        \event{b}{(y{=}r\lor1{=}r)\limplies r{=}0\mid\DW{x}{1}}{right=of a}
      \end{tikzinline}}    
  \end{gathered}
  &&
  \begin{gathered}[t]
    \PW{x}{\BANG\BANG r} 
    \SEMI \PW{x}{0} 
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1em]
        \event{b}{r{\neq}0\mid\DW{x}{1}}{}
        \event{c}{\DW{x}{0}}{right=of b}
        \wki{b}{c}
      \end{tikzinline}}    
  \end{gathered}
  &&
% \end{align*}  
% Sequencing and merging:
% \begin{align*}
  \begin{gathered}[t]
    (\PR{y}{r}
    \SEMI \PW{x}{\BANG r} )
    \SEMI (\PW{x}{\BANG\BANG r} 
    \SEMI \PW{x}{0} )
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1em]
        \event{a}{\DR{y}{1}}{}
        \event{b}{\aForm\mid\DW{x}{1}}{right=of a}
        \event{c}{\DW{x}{0}}{right=of b}
        \wki{b}{c}
      \end{tikzinline}}    
  \end{gathered}
\end{align*}
where $\aForm=((y{=}r\lor1{=}r)\limplies r{=}0)\lor (r{\neq}0)$.  The
precondition $\aForm$ is not a tautology.  In a top-level pomset, this forces
dependency order from $\DRP{y}{1}$ to $\DWP{x}{1}$.

Our solution is to Skolemize, replacing 
uses of $\bForm[\aVal/\aReg]$
by $(\aReg{=}\aVal)\limplies\bForm$,
for uniquely chosen $\aReg$.
%  We have proven associativity of
%  \refdef{def:pomsets-trans} in Agda.
The proof of associativity requires that predicate
transformers distribute through disjunction (\refdef{def:trans}).  The
attempt to define predicate transformers using substitution fails for
\ref{read-tau-empty} because the predicate transformer
$\aTr{}{\bForm}=(\forall\aReg)\bForm$ does not distribute through
disjunction:
\begin{math}
  \aTr{}{\bForm_1\lor \bForm_2}=
  (\forall r)(\bForm_1\lor \bForm_2)
  \neq
  ((\forall r)(\bForm_1)) \lor ((\forall r)(\bForm_2))
  = \aTr{}{\bForm_1} \lor \aTr{}{\bForm_2}
\end{math}.  Since $\aTr{}\bForm=(\forall\aReg)\bForm$ does not distribute
through disjunction, we use $\aTr{}\bForm=\bForm$ instead (which trivially
distributes through disjunction).  This change means we cannot use
substitution, since $\bForm$ does not imply $\bForm[\aVal/\aReg]$.
Fortunately, Skolemizing solves this problem, since $\bForm$ implies
$(\aReg{=}\aVal)\limplies\bForm$.

\subsection{Comparison with Weakest Preconditions}

We compare traditional transformers to the dependent-case transformers of
\reffig{fig:sem}. %; thus we consider only totally ordered executions.
% Because
% we only consider the dependent case, we drop the superscript $\aEvs$ on
% $\aTr{\aEvs}{}$ throughout this section.  We also assume that each register
% appears at most once in a program, as we did throughout
% \textsection\ref{sec:model}--\ref{sec:arm}.

Because of augment closure, we are not interested in isolating the
\emph{weakest} precondition.  Thus we think of transformers as Hoare triples.
In addition, all programs in our language are strongly normalizing, so we
need not distinguish strong and weak correctness.  In this setting, the Hoare
triple $\hoare{\aForm}{\aCmd}{\bForm}$ holds exactly when
$\aForm \limplies \fwp{\aCmd}{\bForm}$.

Hoare triples do not distinguish thread-local variables from shared
variables.  Thus, the assignment rule applies to all types of storage. The
rules can be written as on the left below:
\begin{align*}
\begin{aligned}
  \fwp{\PW{\aLoc}{\aExp}}{\bForm} &= \bForm[\aExp/\aLoc]
  \\
  \fwp{\LET{\aReg}{\aExp}}{\bForm} &= \bForm[\aExp/\aReg]
  \\
  \fwp{\PR{\aLoc}{\aReg}}{\bForm} &= \aLoc{=}\aReg\limplies\bForm
\end{aligned}
&&
\begin{aligned}
  \trd{\PW{\aLoc}{\aExp}}{\bForm} &= \bForm[\aExp/\aLoc]
  \\
  \trd{\LET{\aReg}{\aExp}}{\bForm} &= \bForm[\aExp/\aReg]
  \\
  \trd{\PR{\aLoc}{\aReg}}{\bForm} &= \aVal{=}\aReg\limplies\bForm &&
  \textwhere \labelingAct(\aEv)=\DR{\aLoc}{\aVal}
\end{aligned}
\end{align*}
Here we have chosen an alternative formulation for the read rule, which is
equivalent to the more traditional $\bForm[\aLoc/\aReg]$, as long as registers
are assigned at most once in a program.  Our predicate transformers for the
dependent case are shown on the right above.  Only the read rule differs from
the traditional one.

For programs where every register is bound and every read is fulfilled, our
dependent transformers are the same as the traditional ones.  Thus, when
comparing to weakest preconditions, let us only consider totally-ordered
executions of our semantics where every read could be fulfilled by prepending
some writes.  For example, we ignore pomsets of $\PW{x}{2}\SEMI\PR{x}{r}$
that read $1$ for $x$.

For example, let $\aCmd_i$ be defined:
% as follows.
\begin{align*}
  \aCmd_1&=\PR{x}{s}\SEMI\PW{x}{s{+}r}
  &  
  \aCmd_2&=\PW{x}{t}\SEMI\aCmd_1
  &  
  \aCmd_3&=\LET{t}{2}\SEMI\LET{r}{5}\SEMI\aCmd_2
\end{align*}
% \begin{itemize}
% \item
%   \begin{math}
%     \fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]
%   \end{math}
% \item
%   \begin{math}
%     \fwp{\PR{\aLoc}{\aReg}\;\,}{\bForm} = %\bForm[x/r]
%     %     (\forall\bReg)
%     %     \aLoc{=}\bReg\limplies\bForm [\bReg/\aLoc]
%     \aLoc{=}\aReg\limplies\bForm
%   \end{math}
% \item
%   \begin{math}
%     \fwp{\PW{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]
%   \end{math}
% \end{itemize}
% General relation between Hoare triples and $\fwp{}{}$:
% \begin{itemize}
% \item $\hoare{\fwp{\aCmd}{\bForm}}{\aCmd}{\bForm}$,
% \item If $\hoare{\aForm}{\aCmd}{\bForm}$ and $\aCmd$ terminates when starting
%   in any state satisfying $\aForm$, then $\aForm \limplies \fwp{\aCmd}{\bForm}$.
% \end{itemize}
The following pomset appears in the semantics of $\aCmd_2$.  A pomset for
$\aCmd_3$ can be derived by substituting $[2/t,\allowbreak5/r]$.  A pomset
for $\aCmd_1$ can be derived by eliminating the initial write.
\begin{gather*}
  % \begin{gathered}[t]
  %   \PW{x}{3}
  %   \\
  %   \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
  %     \event{c}{\DW{x}{3}}{}
  %     \xform{xd}{\bForm}{below=of c}
  %     \xo[xright]{c}{xd}
  %   \end{tikzinline}}
  % \end{gathered}
  % \qquad\quad
  % \begin{gathered}[t]
  %   \PR{x}{s}\SEMI\PW{x}{s{+}r}
  %   \\
  %   \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
  %     \event{a}{\DR{x}{2}}{}
  %     \event{b}{2{=}s\limplies(s{+}r){=}7\mid\DW{x}{7}}{right=of a}%6.5em of a}
  %     \po{a}{b}
  %     \xform{xdd}{2{=}s \limplies \bForm[s{+}r/x]}{below right=.5em and -1em of a}
  %       %     \xform{xdd}{2{=}s \limplies \bForm[s{+}r/x]}{above=of a}
  %       %     \xform{xdi}{2{=}s \limplies \bForm[s{+}r/x]}{below=of a}
  %       %     \xform{xii}{(2{=}s\lor x{=}s)\limplies\bForm[s{+}r/x]}{above=of b}
  %       %     \xform{xid}{(2{=}s\lor x{=}s)\limplies\bForm[s{+}r/x]}{below=of b}
  %       %     \xo[xright]{a}{xdi}
  %       %     \xo[xright]{b}{xid}
  %     \xo[xright]{a}{xdd}
  %     \xo[xright]{b}{xdd}
  %   \end{tikzinline}}
  % \end{gathered}
  % \\[1ex]
  \begin{gathered}[t]
    % \LET{t}{2}\SEMI
    % \LET{r}{5}\SEMI
    \PW{x}{t}\SEMI
    \PR{x}{s}\SEMI\PW{x}{s{+}r}
    \\
    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
        \event{a}{\DR{x}{2}}{}
        \event{b}{2{=}s\limplies(s{+}r){=}7\mid\DW{x}{7}}{right=of a}
        \event{c}{t{=}2\mid\DW{x}{2}}{left=of a}
        \xform{xdd}{2{=}s \limplies \bForm[s{+}r/x]}{right=of b}%below right=.5em and -1em of a}
        %\xo{a}{xdd}
        \xo{b}{xdd}
        % \xo{c}{xdd}
        \po{a}{b}
        \rf{c}{a}
      \end{tikzinline}}
  \end{gathered}
\end{gather*}
The predicate transformers are:
% \begin{align*}
%   \fwp{\aCmd_1}{\bForm} &= x{=}s\limplies\bForm[s{+}r/x] 
%   \\
%   \fwp{\aCmd_2}{\bForm} &= t\,{=}s\limplies\bForm[s{+}r/x] 
%   \\
%   \fwp{\aCmd_3}{\bForm} &= 2{=}s\limplies\bForm[s{+}5/x] 
%   \\
%   \trd{\aCmd_1}{\bForm} = \trd{\aCmd_2}{\bForm} &= 2{=}s\limplies\bForm[s{+}r/x] 
%   \\
%   \trd{\aCmd_3}{\bForm} &= 2{=}s\limplies\bForm[s{+}5/x] 
% \end{align*}
\begin{scope}
  \small
  \begin{align*}
    \fwp{\aCmd_1}{\bForm} &= x{=}s\limplies\bForm[s{+}r/x] 
    &
    \trd{\aCmd_1}{\bForm} &= 2{=}s\limplies\bForm[s{+}r/x] 
    \\
    \fwp{\aCmd_2}{\bForm} &= t\,{=}s\limplies\bForm[s{+}r/x] 
    &
    \trd{\aCmd_2}{\bForm} &= 2{=}s\limplies\bForm[s{+}r/x] 
    \\
    \fwp{\aCmd_3}{\bForm} &= 2{=}s\limplies\bForm[s{+}5/x] 
    &
    \trd{\aCmd_3}{\bForm} &= 2{=}s\limplies\bForm[s{+}5/x] 
  \end{align*}
\end{scope}

% % Let $\rho:\Reg\fun\Val$ and $\chi:\Loc\fun\Val$ be substitutions.
% Let $\aState$ and $\rho$ range over substitutions $(\Reg\cup\Loc)\fun\Val$.
% Treating substitutions as states, the big-step operational semantics of
% programs can be defined as a relation $\bigstep{\aState}{\aCmd}{\bState}$.
% % Ie, $\aForm\aState$ implies $\bForm\bState$.
% \begin{align}
%   \label{wp1}
%   \bigstep{[5/r,2/x]}{\aCmd_1&}{[5/r,2/s,7/x]}
%   \\
%   \label{wp2}
%   \bigstep{[\NEG5/r,2/x]}{\aCmd_1&}{[\NEG5/r,2/s,\NEG3/x]}
% \end{align}

% Then the semantics of Hoare triples guarantees that if
% $\aForm\limplies\fwp{\aCmd}{\bForm}$, $\bigstep{\aState}{\aCmd}{\rho}$ and
% $\aForm\aState$ is a tautology then $\bForm\bState$ is a tautology.
% \begin{align*}
%   \fwp{\aCmd_1}{x{>}0} &= (x{+}r{>}0) 
% \end{align*}
% In \eqref{wp1}, the pre- and post-conditions are satisfied.
% In \eqref{wp2}, they are not.

% \begin{itemize}  
% \item Suppose $\bigstep{\aState}{\aCmd}{\rho}$ and $\aForm\limplies\fwp{\aCmd}{\bForm}$.\\
%   If $\aForm\aState$ is a tautology then $\bForm\bState$ is a tautology.\\
%   Ie, $\aForm\aState$ implies $\bForm\bState$.
% \item Suppose $\bigstep{\aState}{\aCmd}{\rho}$ and $\hoare{\aForm}{\aCmd}{\bForm}$.\\
%   If $\aForm\aState$ is a tautology then $\bForm\bState$ is a tautology.\\
%   Ie, $\aForm\aState$ implies $\bForm\bState$.
% \item Suppose $\bigstep{\aState}{\aCmd}{\rho}$ and $\aForm=\fwp{\aCmd}{\bForm}$.\\
%   $\aForm\aState$ is a tautology if and only if $\bForm\bState$ is a tautology.\\
%   Ie, $\aForm\aState$ iff $\bForm\bState$.
% % \item Weakest: If $\aForm'\aState$ is a tautology, then $\aForm$ implies $\aForm'$.
% \end{itemize}
% Weakest preconditions are \emph{sound} in that if $\aForm$ holds in the
% initial state $\aState$, then $\bForm$ holds in the final state $\bState$.
% Formally,

\begin{comment}
  If $\aPS\in\sem{\aCmd}$ is top-level and quiescent then 
  $\aTr{\aEvs}{\bForm}$ implies $\fwp{\aCmd}{\bForm}$.

  For any substitution $\aSub=[{v_1/r_1},\ldots, {v_n/r_n}]$ there is some
  $\aPS\in\sem{\aCmd}$ %that is top-level and quiescent
  such that all preconditions in $\aPS\aSub$ are tautologies then 
  $\fwp{\aCmd}{\bForm}\aSub$
\end{comment}

% For a language where all programs are
% terminating, we have for any statement $\aCmd$:
% \begin{align*}
%   \hoare{\aForm}{\aCmd}{\bForm} 
%   \;\;\Leftrightarrow\;\;
%   \aForm \textimplies \fwp{\aCmd}{\bForm}
% \end{align*}
% Interpretation is that if $\aState\models\fwp{\aCmd}{\bForm}$ and
% $\bigstep{\aState}{\aCmd}{\rho}$
% then $\bState\models\bForm$.

% Let $\aCmd_0$ be
% \begin{math}
%   \PW{\aLoc_1}{\aVal_1}\SEMI\cdots\SEMI \PW{\aLoc_n}{\aVal_n}, 
% \end{math}
% such that $\fwp{\aCmd_0}{\aForm}$ is a tautology, and $\aLoc_i=\aLoc_j$
% implies $i=j$.

% Let $\aSub_\aPS=[{\aVal_1/\aLoc_1},\ldots, {\aVal_n/\aLoc_n}]$ be the final
% state of $\aPS$.

% Let $\aState$ and $\rho$ range over substitutions $\Loc\fun\Exp$.
% If we leave the registers free, we have:
% \begin{align}
%   \label{wp1x}
%   \bigstep{[2/x]}{\aCmd&}{[6/x]}
% \end{align}

% Using \refdef{def:pomsets-trans}:
% \begin{align*}
%   \begin{gathered}[t]
%     %     \PR{x}{s}\SEMI\PW{x}{s{+}r}
%     \PR{x}{s}
%     \\
%     \hbox{\begin{tikzinline}[node distance=.5em and 1em]
%       \event{a}{\DR{x}{2}}{}
%       \xform{xi}{\bForm}{above=of a}
%       \xform{xd}{2{=}s \limplies \bForm}{below=of a}
%       \xo[xright]{a}{xd}
%     \end{tikzinline}}
%   \end{gathered}
%   &&
%   \begin{gathered}[t]
%     \PW{x}{s{+}r}
%     \\
%     \hbox{\begin{tikzinline}[node distance=.5em and 1em]
%       \event{a}{(s{+}r){=}7\mid\DW{x}{7}}{}
%       \xform{xi}{\bForm}{above=of a}
%       \xform{xd}{\bForm}{below=of a}
%       \xo[xright]{a}{xd}
%     \end{tikzinline}}
%   \end{gathered}
% \end{align*}
% Composing
% \begin{align*}
%   \begin{gathered}[t]
%     \PR{x}{s}\SEMI\PW{x}{s{+}r}
%     \\
%     \hbox{\begin{tikzinline}[node distance=.5em and 1em]
%       \event{a}{\DR{x}{2}}{}
%       \event{b}{(s{+}r){=}7\mid\DW{x}{7}}{right=of a}
%       \xform{xdd}{2{=}s \limplies \bForm}{above=of a}
%       \xform{xdi}{2{=}s \limplies \bForm}{below=of a}
%       \xform{xii}{\bForm}{above=of b}
%       \xform{xid}{\bForm}{below=of b}
%       \xo[xright]{a}{xdi}
%       \xo[xright]{b}{xid}
%       \xo[xright]{a}{xdd}
%       \xo[xright]{b}{xdd}
%     \end{tikzinline}}
%   \end{gathered}
% \end{align*}

% Using \refdef{def:pomsets-lir}:
% \begin{align*}
%   \begin{gathered}[t]
%     %     \PR{x}{s}\SEMI\PW{x}{s{+}r}
%     \PR{x}{s}
%     \\
%     \hbox{\begin{tikzinline}[node distance=.5em and 1em]
%       \event{a}{\DR{x}{2}}{}
%       \xform{xi}{(2{=}s\lor x{=}s)\limplies \bForm}{above=of a}
%       \xform{xd}{2{=}s \limplies \bForm}{below=of a}
%       \xo[xright]{a}{xd}
%     \end{tikzinline}}
%   \end{gathered}
%   &&
%   \begin{gathered}[t]
%     \PW{x}{s{+}r}
%     \\
%     \hbox{\begin{tikzinline}[node distance=.5em and 1em]
%       \event{a}{(s{+}r){=}7\mid\DW{x}{7}}{}
%       \xform{xi}{\bForm[s{+}r/x]}{above=of a}
%       \xform{xd}{\bForm[s{+}r/x]}{below=of a}
%       \xo[xright]{a}{xd}
%     \end{tikzinline}}
%   \end{gathered}
% \end{align*}
% Composing

% For example, let $\aCmd_1=\PR{x}{r}$ and $\aCmd_2=\PW{x}{r{+}1}$ and
% $\aCmd=\aCmd_1\SEMI \aCmd_2$.
% \begin{align*}
%   \fwp{\aCmd_2}{x{>}1}&=(r{+}1{>}1) = (r{>}0)
%   \\
%   \fwp{\aCmd_1}{r{>}0}=\fwp{\aCmd_0}{x{>}1}&=(x{>}0)
% \end{align*}
% Let $\aPS_i\in\sem{\aCmd_i}$.
% \begin{align*}
%   \aTr[2]{\aEvs_2}{x{>}1}&=(r{+}1{>}1) = (r{>}0)
%   \\
%   \aTr[0]{\aEvs_0}{x{>}1}&=(0{=}\aReg \limplies r{>}0)
%   \\
%   \aTr[0]{\aEvs_0}{x{>}1}&=(1{=}\aReg \limplies r{>}0)
%   \\
%   \aTr[0]{\aEvs_0}{x{>}1}&=(2{=}\aReg \limplies r{>}0)
% \end{align*}

% \begin{proposition}
%   If $\aPS\in\sem{\aCmd}$ is top-level and quiescent then 
%   $\aTr{\aEvs}{\aForm}$ implies $\fwp{\aCmd}{\aForm}$.

%   For any substitution $\aSub=[{\aVal_1/\aReg_1},\ldots, {\aVal_n/\aReg_n}]$ there is some
%   $\aPS\in\sem{\aCmd}$ %that is top-level and quiescent
%   such that all preconditions in $\aPS\aSub$ are tautologies then 
%   $\fwp{\aCmd}{\aForm}\aSub$
% \end{proposition}

% \subsection{Fences}
% \label{sec:fences}

\subsection{Substitutions}
\label{sec:substitutions}

In $\sLOAD{}{}$, it is also possible to collapse $\aLoc$ and $\aReg$ via substitution:
\begin{enumerate}
\item[{\labeltext[\textsc{r}4a$'$]{(\textsc{r}4a$'$)}{read-tau-dep-sub}}]
  if $(\aEvs\cap\bEvs)\neq\emptyset$ then
  \begin{math}
    \aTr{\bEvs}{\bForm} \riff
    \aVal{=}\aReg
    \limplies \bForm[\aReg/\aLoc]
  \end{math},    
\item[{\labeltext[\textsc{r}4b$'$]{(\textsc{r}4b$'$)}{read-tau-ind-sub}}]
  if $\aEvs\neq\emptyset$ and $(\aEvs\cap\bEvs)=\emptyset$ then
  \begin{math}
    \aTr{\bEvs}{\bForm} \riff
    \PBR{\aVal{=}\aReg \lor \aLoc{=}\aReg} \limplies
    \bForm[\aReg/\aLoc],
  \end{math}
\item[{\labeltext[\textsc{r}4c$'$]{(\textsc{r}4c$'$)}{read-tau-empty-sub}}]
  if $\aEvs=\emptyset$ then
  \begin{math}
    \aTr{\bEvs}{\bForm} \riff
    % \PBR{\aVal{=}\aReg \lor \aLoc{=}\aReg} \limplies
    \bForm[\aReg/\aLoc],
  \end{math}
\end{enumerate}
Perhaps surprisingly, this semantics is incomparable with that of
\reffig{fig:sem}.  Consider the following:
\begin{gather*}
  \IF{r\land s\;\mathsf{even}}\THEN \PW{y}{1}\FI\SEMI
  \IF{r\land s}\THEN \PW{z}{1}\FI
  \\
  \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
      \event{a3}{r\land s\;\mathsf{even}\mid\DW{y}{1}}{}
      \event{a4}{r\land s\mid\DW{z}{1}}{right=of a3}
    \end{tikzinline}}
\end{gather*}
Prepending $\PRP{x}{s}$, we get the same result regardless of whether we
substitute $[s/x]$, since $x$ does not occur in either precondition.  Here
we show the independent case:
\begin{gather*}
  \PR{x}{s}\SEMI
  \IF{r\land s\;\mathsf{even}}\THEN \PW{y}{1}\FI\SEMI
  \IF{r\land s}\THEN \PW{z}{1}\FI
  \\
  \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
      \event{a2}{\DR{x}{2}}{}
      \event{a3}{(2{=}s\lor x{=}s)\limplies (r\land s\;\mathsf{even})\mid\DW{y}{1}}{right=of a2}
      \event{a4}{(2{=}s\lor x{=}s)\limplies (r\land s)\mid\DW{z}{1}}{right=of a3}
    \end{tikzinline}}
\end{gather*}
Since the preconditions mention $x$, prepending $\PRP{x}{r}$, we now get
different results depending on whether we perform the substitution.  Without
any substitution, we have:
\begin{gather*}
  \PR{x}{r}\SEMI
  \PR{x}{s}\SEMI
  \IF{r\land s\;\mathsf{even}}\THEN \PW{y}{1}\FI\SEMI
  \IF{r\land s}\THEN \PW{z}{1}\FI
  \\[-2ex]
  \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
      \event{a1}{\DR{x}{1}}{}
      \event{a2}{\DR{x}{2}}{right=of a1}
      \event{a3}{1{=}r\limplies  (2{=}s\lor x{=}s)\limplies (r\land s\;\mathsf{even})\mid\DW{y}{1}}{right=of a2}
      \event{a4}{1{=}r\limplies  (2{=}s\lor x{=}s)\limplies (r\land s)\mid\DW{z}{1}}{right=of a3}
      \po[out=12,in=168]{a1}{a3}
      \po[out=10,in=170]{a1}{a4}
    \end{tikzinline}}
\end{gather*}
Prepending $\PWP{x}{0}$, which substitutes $[0/x]$, the precondition of
$\DWP{y}{1}$ becomes
$(1{=}r\limplies (2{=}s\lor0{=}s)\limplies (r\land s\;\mathsf{even}))$, which
is a tautology, whereas the precondition of $\DW{z}{1}$ becomes
$(1{=}r\limplies(2{=}s\lor0{=}s)\limplies (r\land s))$, which is not.  In
order to be top-level, $\DWP{z}{1}$ must be dependency ordered after
$\DRP{x}{2}$; in this case the precondition becomes
$(1{=}r\limplies2{=}s\limplies (r\land s))$, which is a tautology.
\begin{gather*}
  % \PW{x}{0}\SEMI
  % \PR{x}{r}\SEMI
  % \PR{x}{s}\SEMI
  % \IF{r\land s\;\mathsf{even}}\THEN \PW{y}{1}\FI\SEMI
  % \IF{r\land s}\THEN \PW{z}{1}\FI
  % \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a0}{\DW{x}{0}}{}
      \event{a1}{\DR{x}{1}}{right=of a0}
      \event{a2}{\DR{x}{2}}{right=of a1}
      \event{a3}{\DW{y}{1}}{right=of a2}
      \event{a4}{\DW{z}{1}}{right=of a3}
      % \wk{a0}{a1}
      % \wk[out=-20,in=-160]{a0}{a2}
      \po[out=20,in=160]{a1}{a3}
      \po[out=20,in=160]{a1}{a4}
      \po[out=-20,in=-160]{a2}{a4}
    \end{tikzinline}}
\end{gather*}
The situation reverses with the substitution $[r/x]$:
\begin{gather*}
  \PR{x}{r}\SEMI
  \PR{x}{s}\SEMI
  \IF{r\land s\;\mathsf{even}}\THEN \PW{y}{1}\FI\SEMI
  \IF{r\land s}\THEN \PW{z}{1}\FI
  \\[-2ex]
  \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
      \event{a1}{\DR{x}{1}}{}
      \event{a2}{\DR{x}{2}}{right=of a1}
      \event{a3}{1{=}r\limplies  (2{=}s\lor r{=}s)\limplies (r\land s\;\mathsf{even})\mid\DW{y}{1}}{right=of a2}
      \event{a4}{1{=}r\limplies  (2{=}s\lor r{=}s)\limplies (r\land s)\mid\DW{z}{1}}{right=of a3}
      \po[out=12,in=168]{a1}{a3}
      \po[out=10,in=170]{a1}{a4}
    \end{tikzinline}}
\end{gather*}
Prepending $\PWP{x}{0}$:
%\vspace{-.5\baselineskip}
\begin{gather*}
  % \PW{x}{0}\SEMI
  % \PR{x}{r}\SEMI
  % \PR{x}{s}\SEMI
  % \IF{r\land s\;\mathsf{even}}\THEN \PW{y}{1}\FI\SEMI
  % \IF{r\land s}\THEN \PW{z}{1}\FI
  % \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a0}{\DW{x}{0}}{}
      \event{a1}{\DR{x}{1}}{right=of a0}
      \event{a2}{\DR{x}{2}}{right=of a1}
      \event{a3}{\DW{y}{1}}{right=of a2}
      \event{a4}{\DW{z}{1}}{right=of a3}
      % \wk{a0}{a1}
      % \wk[out=-20,in=-160]{a0}{a2}
      \po[out=20,in=160]{a1}{a3}
      \po[out=20,in=160]{a1}{a4}
      \po{a2}{a3}
    \end{tikzinline}}
\end{gather*}
The dependency has changed from $\DRP{x}{2}\xpo\DWP{z}{1}$ to
$\DRP{x}{2}\xpo\DWP{y}{1}$.  The resulting sets of pomsets are
incomparable.

Thinking in terms of hardware, the difference is whether reads update the
cache, thus clobbering preceding writes.  With $[r/x]$, reads clobber the
cache, whereas without the substitution, they do not.  Since most caches work
this way, the model with $[r/x]$ is likely preferred for modeling hardware.
However, this substitution only makes sense in a model with read-read
coherence and read-read dependencies, which we will see is not the case for Arm.  By
leaving out the substitution, we also ensure that downgraded reads are
fulfilled by preceding writes, not reads.

\label{sec:ex:last}

### 2021-06-17 paper/2SUB/outro.tex
reorg

--- 2021-04-17 paper/outro.tex
@@
 
-This paper is the first to present a compositional semantics for
+This paper is the first to present a direct denotational semantics for
 sequential composition in a relaxed memory model which can be
@@
 reassuring to update the mechanization to bring it in line with the current state.
+
+We don't handle access elimination.
 

### 2021-06-21 paper/arm.tex
added referee reports
swap order of le and rf

@@
 \item \label{lob-rf}
-  write-to-read order of $\rrfx$, required by \ref{seq-le-rf},
+  write-to-read order of $\rrfx$, required by \ref{seq-rf-le},
 \item \label{lob-wr}
@@
 for relaxed access.  In the next section we recover the optimal lowering by
-adopting an alternative semantics for \ref{seq-le-rf} and \ref{seq-le-delays}.
+adopting an alternative semantics for \ref{seq-rf-le} and \ref{seq-le-delays}.
 
@@
 for the read and satisfying the precondition of $\DWP{y}{1}$ by prepending
-$\DWP{x}{2}$.  It may be tempting, therefore to adopt a strategy of
+$\DWP{x}{2}$.  It may be tempting, therefore, to adopt a strategy of
 \emph{downgrading} acquires in certain cases.  Unfortunately, it is not
@@
     We show that all the order required in the pomset is also required by
-    \armeight{}.  \ref{pom-le-block} holds since ${\rcb_1}$ is consistent with
+    \armeight{}.  \ref{pom-rf-block} holds since ${\rcb_1}$ is consistent with
     ${\rco_1}$ and ${\rfr_1}$.  As noted \ref{lob-le}, $\rlob$ includes the order
@@
 The change to the semantics is small: we weaken the relationship between $\rrfx$
-and $\le$ in \ref{seq-le-rf}.  Rather than ensuring that there is no
-\emph{global} blocker for a sequentially fulfilled read \eqref{seq-le-rf}, we
-require only that there is no \emph{thread-local} blocker \eqref{seq-le-rf-rf}.
+and $\le$ in \ref{seq-rf-le}.  Rather than ensuring that there is no
+\emph{global} blocker for a sequentially fulfilled read \eqref{seq-rf-le}, we
+require only that there is no \emph{thread-local} blocker \eqref{seq-rf-le-rf}.
 This change both allows and requires us to weaken the definition of
@@
   Let $\frf{\semrr{}}$ be defined as for $\semrr{}$ in
-  \refdef{def:semrr}/\reffig{fig:sem}, changing \ref{seq-le-rf} and
+  \refdef{def:semrr}/\reffig{fig:sem}, changing \ref{seq-rf-le} and
   \ref{seq-le-delays}:
   \begin{itemize}
-  \item[{\labeltext[\frf{\textsc{s}7b}]{(\frf{\textsc{s}7b})}{seq-le-rf-rf}}]
+  \item[{\labeltext[\frf{\textsc{s}7b}]{(\frf{\textsc{s}7b})}{seq-rf-le-rf}}]
     if $\labeling_1(\cEv) \rblocks \labeling_2(\aEv)$ then $\bEv\xrfx\aEv$
     implies $\cEv\le\bEv$,
-  \item[{\labeltext[\frf{\textsc{s}7c}]{(\frf{\textsc{s}7c})}{seq-le-delays-rf}}]
+  \item[{\labeltext[\frf{\textsc{s}6b}]{(\frf{\textsc{s}6b})}{seq-le-delays-rf}}]
     if $\labeling_1(\bEv) \rdelaysp \labeling_2(\aEv)$ then $\bEv\le\aEv$,\\
@@
 
-With the weakening of \ref{seq-le-rf-rf}, we must be careful not to allow
+With the weakening of \ref{seq-rf-le-rf}, we must be careful not to allow
 spurious pairs to be added to the $\rrfx$ relation.  The use of
@@
     We show that all the order required in the pomset is also required by
-    \armeight{}.  \ref{pom-le-block} holds since ${\rcb_1}$ is consistent with
-    ${\rco_1}$ and ${\rfr_1}$.  \ref{seq-le-rf-rf} follows from \ref{arm-rfi}.
+    \armeight{}.  \ref{pom-rf-block} holds since ${\rcb_1}$ is consistent with
+    ${\rco_1}$ and ${\rfr_1}$.  \ref{seq-rf-le-rf} follows from \ref{arm-rfi}.
     As noted \ref{lob-le}, $\rlob$ includes the order required by

### 2021-06-21 paper/oopsla.tex
Weakest preconditions
swap order of le and rf

@@
   \begin{math}
-    \aTr{\bEvs}{\bForm} \rimplies
+    \aTr{\bEvs}{\bForm} \riff
     \aVal{=}\aReg
@@
   \begin{math}
-    \aTr{\bEvs}{\bForm} \rimplies
+    \aTr{\bEvs}{\bForm} \riff
     \PBR{\aVal{=}\aReg \lor \aLoc{=}\aReg} \limplies
@@
   \begin{math}
-    \aTr{\bEvs}{\bForm} \rimplies
+    \aTr{\bEvs}{\bForm} \riff
     \bForm[\aVal/\aReg][\aVal/\aLoc]
@@
   \begin{math}
-    \aTr{\bEvs}{\bForm} \rimplies
+    \aTr{\bEvs}{\bForm} \riff
     \bForm[\aVal/\aReg][\aVal/\aLoc]\land\bForm[\aLoc/\aReg]
@@
   % \item[\ref{L4})]
-  %   $\aTr{\bEvs}{\bForm} \rimplies \aVal{=}\aLoc\limplies\bForm[\aLoc/\aReg]$.
+  %   $\aTr{\bEvs}{\bForm} \riff \aVal{=}\aLoc\limplies\bForm[\aLoc/\aReg]$.
   %   % \item[\ref{L5})]
-  %   %   $\aTr{\cEvs}{\bForm} \rimplies (\aVal{=}\aLoc\lor\TRUE)\limplies\bForm[\aLoc/\aReg]$. %, when $\aEvs\neq\emptyset$,
+  %   %   $\aTr{\cEvs}{\bForm} \riff (\aVal{=}\aLoc\lor\TRUE)\limplies\bForm[\aLoc/\aReg]$. %, when $\aEvs\neq\emptyset$,
   %   % \item[\ref{L6})] 
-  %   %   $\aTr{\dEvs}{\bForm}\; \rimplies \bForm$, when $\aEvs=\emptyset$.
+  %   %   $\aTr{\dEvs}{\bForm}\; \riff \bForm$, when $\aEvs=\emptyset$.
   % \end{enumerate}
@@
 %   if $\aEv\in\aEvs_2\setminus\aEvs_1$ then either \\
-%   $\labelingForm(\aEv) \rimplies \labelingForm_2(\aEv)[\aLoc/\aReg][\aVal/\aLoc]$ and $(\exists\bEv\in\aEvs_1)\bEv{<}\aEv$, or \\
-%   $\labelingForm(\aEv) \rimplies \labelingForm_2(\aEv)[\aLoc/\aReg][\aVal/\aLoc] \land \labelingForm_2(\aEv)[\aLoc/\aReg]$.
+%   $\labelingForm(\aEv) \riff \labelingForm_2(\aEv)[\aLoc/\aReg][\aVal/\aLoc]$ and $(\exists\bEv\in\aEvs_1)\bEv{<}\aEv$, or \\
+%   $\labelingForm(\aEv) \riff \labelingForm_2(\aEv)[\aLoc/\aReg][\aVal/\aLoc] \land \labelingForm_2(\aEv)[\aLoc/\aReg]$.
 % \end{enumerate}
@@
 
-Not that the definition of \emph{$L$-stable} applies in the case that
+Note that the definition of \emph{$L$-stable} applies in the case that
 conflicting writes are totally ordered.  This gives a result more in the
 spirit of \cite{Dolan:2018:BDR:3192366.3192421}.  In particular, this special
-case of the theorem clarifies the discussion of \ref{past}.
+case of the theorem clarifies the discussion of the \textsc{past} example
+in \jjr{};
 

### 2021-06-24 paper/abstract.tex
added referee reports
updates and fixes

--- 2021-04-16 paper/abstract.tex
@@
 as a ``relaxed memory model.''  The relaxed memory models that have been
-proposed to date either fail to address sequential composition, or overly
+proposed to date either fail to address sequential composition directly, or overly
 restrict processors and compilers.
 
-To support sequential composition while targeting modern hardware, we propose adding families of predicate
-transformers to the existing model of ``Pomsets with Preconditions,'' which
-already supports parallel composition.  When composing $(S_1;S_2)$, the predicate
-transformers used to validate the preconditions of events in $S_2$ are chosen
-based on the semantic dependencies from events in $S_1$ to events in $S_2$.  Our model
-retains the good properties of the prior work, including efficient
-implementation on Arm8, support for compiler optimizations, support for
-logics that prove the absence of thin-air behaviors, and a local data race
-freedom theorem.
+To support sequential composition while targeting modern hardware, we propose
+using preconditions and families of predicate transformers.  When composing
+$(S_1;S_2)$, the predicate transformers used to validate the preconditions of
+events in $S_2$ are chosen based on the semantic dependencies from events in
+$S_1$ to events in $S_2$.  We apply this approach to two existing memory
+models: ``Modular Relaxed Dependencies'' for C11 and ``Pomsets with
+Preconditions.''
+
+\endinput

### 2021-06-24 paper/drf.tex
updates and fixes

\section{Local Data Race Freedom and Sequential Consistency}
\label{sec:sc}

We adapt \citeauthor{Dolan:2018:BDR:3192366.3192421}'s
[\citeyear{Dolan:2018:BDR:3192366.3192421}] notion of \emph{Local Data Race
  Freedom (LDRF)} to our setting.

The result requires that locations are properly initialized.  We assume a
sufficient condition: that programs have the form
``$\aLoc_1\GETS\aVal_1\SEMI \cdots \aLoc_n\GETS\aVal_n\SEMI\aCmd$'' where
every location mentioned in $\aCmd$ is some $\aLoc_i$.

We make two further restrictions to simplify the exposition.  To simplify the
definition of \emph{happens-before}, we ban fences and \RMWs.  To simplify
the proof, we assume there are no local declarations of the form
$(\VAR\aLoc\SEMI \aCmd)$.

To state the theorem, we require several technical definitions.  The reader
unfamiliar with \citep{Dolan:2018:BDR:3192366.3192421} may prefer to skip to
the examples in the proof sketch, referring back as needed.

\bookmark{Definitions}
\paragraph{Data Race}
Data races are defined using \emph{program} order $(\rpox)$, not
\emph{pomset} order $(\le)$. %, and thus is stable with respect to augmentation.
In \ref{SB}, for example, $(\DR{x}{0})$ has an $x$-race with $(\DW{x}{1})$,
but not $(\DW{x}{0})$, which is $\rpox$-before it.

It is obvious how to enhance the semantics of prefixing and most other
operators to define $\rpox$.  When combining pomsets using the conditional,
the obvious definition may result in cycles, since $\rpox$-ordered reads may
coalesce---see the discussion of \ref{CA} in \textsection\ref{sec:refine}.  In
this case we include a separate pomset for each way of breaking these cycles.

Because we ignore the features of
\textsection\ref{sec:variants}, we can adopt the simplest definition of
\emph{synchronizes\hyp{}with}~($\rsw$): Let $\bEv\xsw\aEv$ exactly when
$\bEv$ fulfills $\aEv$, $\bEv$ is a release, $\aEv$ is an acquire, and
$\lnot(\bEv\xpox\aEv)$.

Let ${\rhb}=({\rpox}\cup{\rsw})^+$ be the \emph{happens-before} relation.  In
\ref{Pub1}, for example, $(\DW{x}{1})$ happens-before $(\DR{x}{0})$, but this
fails if either $\mRA$ access is relaxed.

Let $L\subseteq\Loc$ be a set of locations.  We say that $\bEv$ \emph{has an
  $L$-race with} $\aEv$ (notation $\bEv\lrace{L}\aEv$) when at least one is
relaxed, they \emph{conflict}
(Def.~\ref{def:rf}) at
some location in $L$, and they are unordered by $\rhb$: neither $\bEv\xhb\aEv$ nor
$\aEv\xhb\bEv$.

\paragraph{Generators}
We say that $\aPS'$ \emph{generates} $\aPS$ if either
$\aPS$ augments $\aPS'$ or $\aPS$ implies $\aPS'$.  For example, the
unordered pomset $(\DR{x}{1})$ $(\DW{y}{1})$ generates the ordered pomset
$(\DR{x}{1})\xpo(\aReg=1\mid\DW{y}{1})$.

We say that $\aPS$ is a \emph{generation-minimal} in $\aPSS$ if $\aPS\in\aPSS$ and
there is no $\aPS\neq\aPS'\in\aPSS$ that generates $\aPS$.

Let $\semmin{\aCmd}=\{\aPS\in\sem{\aCmd} \mid \aPS$ is \emph{top-level}
(Def.~\ref{def:top}) and generation-minimal in $\sem{\aCmd}\}$.

\paragraph{Extensions}

We say that $\aPS'$ \emph{$\aCmd$-extends} $\aPS$ if %$\aPS\in\semmin{\aCmd}$,
$\aPS\neq\aPS'\in\semmin{\aCmd}$ and $\aPS$ is a downset of $\aPS'$.

\paragraph{Similarity}
We say that \emph{$\aPS'$ is $\aEv$-similar to $\aPS$} if they differ at most
in (1) pomset order adjacent to $\aEv$ and (2) the value associated with
event $\aEv$, if it is a
read.  % We say they are \emph{similar} if they are $\aEv$-similar for some $\aEv$.
Formally: $\Event'=\Event$, $\labelingForm'=\labelingForm$,
$\restrict {\le'}{\Event\setminus\{\aEv\}}=\restrict {\le}{\Event\setminus\{\aEv\}}$,
if $\aEv$ is not a read then $\labelingAct'=\labelingAct$, and if $\aEv$ is a
read then
$\restrict{\labelingAct'}{\Event\setminus\{\aEv\}}=\restrict{\labelingAct}{\Event\setminus\{\aEv\}}$
and $\labelingAct'(\aEv) = \labelingAct(\aEv)[\aVal'/\aVal]$, for some
$\aVal'$, $\aVal$.

\paragraph{Stability}
We say that $\aPS$ is \emph{$L$-stable in $\aCmd$} if
(1) $\aPS\in\semmin{\aCmd}$, 
(2) $\aPS$ is $\rpox$-convex (nothing missing in program order),
(3) there is no $\aCmd$-extension of $\aPS$ with a \emph{crossing} $L$-race:
that is, there is no $\bEv\in\Event$, no $\aPS'$ $\aCmd$-extending
$\aPS$, and no $\aEv\in\Event'\setminus\Event$ such that $\bEv\lrace{L}\aEv$.
The empty pomset is $L$-stable.

\paragraph{Sequentiality}
Let ${\pole{L}}={\lt_L}\cup{\rpox}$, where $\lt_L$ is the restriction of $\lt$ to events that access locations in $L$.
We say that $\aPS'$ is \emph{$L$-sequential after $\aPS$} if 
$\aPS'$ is $\rpox$-convex and %
$\pole{L}$ is acyclic in $\Event'\setminus\Event$.

\bookmark{Theorem and Proof Sketch}
\begin{theorem}
  Let $\aPS$ be $L$-stable in $\aCmd$.  Let $\aPS'$ be a $\aCmd$-extension of
  $\aPS$ that is $L$-sequential after $\aPS$.  Let $\aPS''$ be a
  $\aCmd$-extension of $\aPS'$ that is $\rpox$-convex, such that no subset of
  $\Event''$ satisfies these criteria.
  Then either (1) $\aPS''$ is $L$-sequential after $\aPS$ or (2) there is
  some $\aCmd$-extension $\aPS'''$ of $\aPS'$ and some
  $\aEv\in(\Event''\setminus\Event')$ such that (a) $\aPS'''$ is
  $\aEv$-similar to $\aPS''$, (b) $\aPS'''$ is $L$-sequential after $\aPS$,
  and (c) $\bEv\lrace{L}\aEv$, for some $\bEv\in(\Event''\setminus\Event)$.
\end{theorem}
The theorem provides an inductive characterization of \emph{Sequential
  Consistency for Local
  Data-Race Freedom (SC-LDRF)}: Any extension of a $L$-stable pomset is either
$L$-sequential, or is $\aEv$-similar to a $L$-sequential extension that
includes a race involving $\aEv$.
\begin{proof}[Proof Sketch]
  In order to develop a technique to find $\aPS'''$ from $\aPS''$, we analyze
  pomset order in generation-minimal top-level pomsets.  First, we note that
  $\le_*$ (the transitive reduction $\le$) can be decomposed into three
  disjoint relations.  Let ${\rppo}=({\le_*}\cap{\rpox})$ denote
  \emph{preserved} program order, as required by prefixing (Def.~\ref{def:prefix}).  The other two relations are cross-thread subsets of
  $({\le_*}\setminus{\rpox})$, as required by fulfillment (Def.~\ref{def:rf}): $\rrfe$ orders writes before reads, satisfying fulfillment
  requirement \ref{rf3}; $\rxw$ orders read and write accesses before writes,
  satisfying requirement \ref{rf4}. ({Within a thread, \ref{rf3} and
    \ref{rf4} follow from prefixing requirement \ref{5b}, which is included
    in ${\rppo}$.})

  Using this decomposition, we can show the following.
  \begin{lemma}
    Suppose $\aPS''\in\semmin{\aCmd}$ has a read $\aEv$ that is maximal in
    $({\rppo}\cup{\rrfe})$ and such that every $\rpox$-following read is
    also $\le$-following ($\aEv\xpox\bEv$ implies $\aEv\le\bEv$, for every
    read $\bEv$).  Further, suppose there is an $\aEv$-similar $\aPS'''$
    that satisfies the requirements of fulfillment.  Then
    $\aPS'''\in\semmin{\aCmd}$.
  \end{lemma}
  The proof of the lemma follows an inductive construction of
  $\semmin{\aCmd}$, starting from a large set with little order, and
  pruning the set as order is added: We begin with all pomsets generated by
  the semantics without imposing the requirements of fulfillment (including
  only $\rppo$).  We then prune reads which cannot be fulfilled, starting
  with those that are minimally ordered.  This proof is simplified by
  precluding local declarations.

  We can prove a similar result for $({\rpox}\cup{\rrfe})$-maximal read
  and write accesses.

  Turning to the proof of the theorem, if $\aPS''$ is $L$-sequential after
  $\aPS$, then the result follows from (1).  Otherwise, there must be a
  $\pole{L}$ cycle in $\aPS''$ involving all of the actions in
  $(\Event''\setminus\Event')$: If there were no such cycle, then $\aPS''$
  would be $L$-sequential; if there were elements outside the cycle, then
  there would be a subset of $\Event''$ that satisfies these criteria.

  If there is a $({\rpox}\cup{\rrfe})$-maximal access, we select one of
  these as $\aEv$.  If $\aEv$ is a write, we reverse the outgoing order in
  $\rxw$; the ability to reverse this order witnesses the race.  If $\aEv$
  is a read, we switch its fulfilling write to a ``newer'' one, updating
  $\rxw$; the ability to switch witnesses the race.  For
  example, for $\aPS''$ on the left below, we choose the $\aPS'''$ on the
  right;  $\aEv$ is the read of $x$, which races with $(\DW{x}{1})$.  % Program order
  \begin{gather*}
    x\GETS 0 \SEMI y\GETS 0 \SEMI  (x \GETS 1  \SEMI y \GETS 1
    \PAR
    \IF{y}\THEN \aReg \GETS x \FI)
    \\[-.5ex]
    \hbox{\begin{tikzinline}[node distance=1.5em and 2em]
        \event{wy0}{\DW{y}{0}}{}
        \event{wx0}{\DW{x}{0}}{below=of wy0}
        \event{wx1}{\DW{x}{1}}{right=3em of wy0}
        \event{wy1}{\DW{y}{1}}{right=of wx1}
        \event{ry1}{\DR{y}{1}}{below=of wx1}
        \event{rx}{\DR{x}{0}}{below=of wy1}
        \rf[bend right]{wx0}{rx}
        \rf{wy1}{ry1}
        \wk[bend left]{wy0}{wy1}
        \pox{wx1}{wy1}
        \pox{ry1}[below]{rx}
        \wk{rx}{wx1}
        \node(ix)[left=of wx0]{};
        \node(iy)[left=of wy0]{};
        \bgoval[yellow!50]{(ix)(iy)}{P}
        \bgoval[pink!50]{(wx0)(wy0)}{P'\setminus P}
        \bgoval[green!10]{(ry1)(wx1)(rx)(wy1)}{P''\setminus P'}
        \pox{wx0}{wy0}
        \pox{wy0}{wx1}
        \pox{wy0}[below]{ry1}
      \end{tikzinline}}
    \qquad
    \hbox{\begin{tikzinline}[node distance=1.5em and 2em]
        \event{wy0}{\DW{y}{0}}{}
        \event{wx0}{\DW{x}{0}}{below=of wy0}
        \event{wx1}{\DW{x}{1}}{right=3em of wy0}
        \event{wy1}{\DW{y}{1}}{right=of wx1}
        \event{ry1}{\DR{y}{1}}{below=of wx1}
        \event{rx}{\DR{x}{1}}{below=of wy1}
        \rf{wx1}{rx}
        \rf{wy1}{ry1}
        \wk[bend left]{wy0}{wy1}
        \pox{wx1}{wy1}
        \pox{ry1}[below]{rx}
        \wk{wx0}{wx1}
        \node(ix)[left=of wx0]{};
        \node(iy)[left=of wy0]{};
        \bgoval[yellow!50]{(ix)(iy)}{P}
        \bgoval[pink!50]{(wx0)(wy0)}{P'\setminus P}
        \bgoval[green!10]{(ry1)(wx1)(rx)(wy1)}{P'''\setminus P'}
        \pox{wx0}{wy0}
        \pox{wy0}{wx1}
        \pox{wy0}[below]{ry1}
      \end{tikzinline}}
  \end{gather*}    
  It is important that $\aEv$ be $({\rpox}\cup{\rrfe})$-maximal, not just
  $({\rppo}\cup{\rrfe})$-maximal.  The latter criterion would allow us to
  choose $\aEv$ to be the read of $y$, but then there would be no
  $\aEv$-similar pomset: if an execution reads $0$ for $y$ then there is no
  read of $x$, due to the conditional.

  If there is no $({\rpox}\cup{\rrfe})$-maximal access, then all
  cross-thread order must be from $\rrfe$.  In this case, we select a
  $({\rppo}\cup{\rrfe})$-maximal read, switching its fulfilling write to an
  ``older'' one.  As an example, consider the following; once again,
  $\aEv$ is the read of $x$, which races with $(\DW{x}{1})$.
  \begin{gather*}
    x\GETS 0 \SEMI y\GETS 0 \SEMI (\aReg \GETS x  \SEMI y \GETS 1
    \PAR
    \bReg \GETS y \SEMI x \GETS \bReg)
    \\[-.5ex]
    \hbox{\begin{tikzinline}[node distance=1.5em and 2em]
        \event{wx0}{\DW{y}{0}}{}
        \event{ry}{\DR{x}{1}}{right=3em of wx0}
        \event{wx1}{\DW{y}{1}}{right=of ry}
        \event{wy0}{\DW{x}{0}}{below=of wx0}
        \event{rx1}{\DR{y}{1}}{right=3em of wy0}
        \event{wy1}{\DW{x}{1}}{right=of rx1}
        \rf{wx1}{rx1}
        \rf{wy1}{ry}
        \po{rx1}{wy1}
        \pox{ry}{wx1}
        \wk[bend left]{wx0}{wx1}
        \wk[bend right]{wy0}{wy1}
        \node(ix)[left=of wx0]{};
        \node(iy)[left=of wy0]{};
        \bgoval[yellow!50]{(ix)(iy)}{P}
        \bgoval[pink!50]{(wx0)(wy0)}{P'\setminus P}
        \bgoval[green!10]{(ry)(wx1)(rx1)(wy1)}{P''\setminus P'}
        \pox{wy0}{wx0}
        \pox{wx0}{ry}
        \pox{wx0}[below]{rx1}
      \end{tikzinline}}
    \qquad
    \hbox{\begin{tikzinline}[node distance=1.5em and 2em]
        \event{wx0}{\DW{y}{0}}{}
        \event{ry}{\DR{x}{0}}{right=3em of wx0}
        \event{wx1}{\DW{y}{1}}{right=of ry}
        \event{wy0}{\DW{x}{0}}{below=of wx0}
        \event{rx1}{\DR{y}{1}}{right=3em of wy0}
        \event{wy1}{\DW{x}{1}}{right=of rx1}
        \pox{ry}{wx1}
        \wk[bend left]{wx0}{wx1}
        \rf{wx1}{rx1}
        \rf{wy0}{ry}
        \po{rx1}{wy1}
        \wk{ry}{wy1}
        \node(ix)[left=of wx0]{};
        \node(iy)[left=of wy0]{};
        \bgoval[yellow!50]{(ix)(iy)}{P}
        \bgoval[pink!50]{(wx0)(wy0)}{P'\setminus P}
        \bgoval[green!10]{(ry)(wx1)(rx1)(wy1)}{P'''\setminus P'}
        \pox{wy0}{wx0}
        \pox{wx0}{ry}
        \pox{wx0}[below]{rx1}
      \end{tikzinline}}
  \end{gather*}
  This example requires $(\DW{x}{0})$.  Proper initialization ensures the
  existence of such ``older'' writes.
\end{proof}

\bookmark{Mixed Races}
The premises of the theorem allow us to avoid the complications caused by ``mixed races'' in
\cite{DBLP:conf/ppopp/DongolJR19}.  In the left pomset below, $\aPS''$ is not
an extension of $\aPS'$, since $\aPS'$ is not a downset of $\aPS''$.  
When considering this pomset, we must perform the decomposition on the right.
\begin{gather*}
  (x\GETS 0 \SEMI   x^\mRA \GETS 1)
  \PAR
  (\aReg\GETS x^\mRA)
  \\[-2ex]
  \hbox{\begin{tikzinline}[node distance=1.5em and 2em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DWRel{x}{1}}{right=of wx0}
      \event{rx}{\DRAcq{x}{0}}{below=of wx0}
      \rf{wx0}{rx}
      \pox{wx0}{wx1}
      \wk{rx}{wx1}
      \node(ix)[left=of wx0]{};
      \bgoval[yellow!50]{(ix)}{P}
      \bgoval[pink!50]{(wx0)(wx1)}{P'\setminus P}
      \bgovalright[green!10]{(rx)}{P''\setminus P'}
    \end{tikzinline}}
  \qquad
  \qquad
  \qquad
  \hbox{\begin{tikzinline}[node distance=1.5em and 2em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DWRel{x}{1}}{right=of wx0}
      \event{rx}{\DRAcq{x}{0}}{below=of wx0}
      \rf{wx0}{rx}
      \pox{wx0}{wx1}
      \wk{rx}{wx1}
      \node(ix)[left=of wx0]{};
      \bgoval[yellow!50]{(ix)}{P}
      \bgoval[pink!50]{(wx0)(rx)}{P'\setminus P}
      \bgoval[green!10]{(wx1)}{P''\setminus P'}
    \end{tikzinline}}
\end{gather*}
This affects the inductive order in which we move across pomsets, but does
not affect the set of pomsets that are considered.  This simplification is
enabled by denotational reasoning.

\bookmark{Comparison to Java}
In our language, past races are always resolved at a stable point, as in
\ref{Co3}.  As another example, consider the following, which is disallowed
here, but allowed by Java \cite[Ex.~2]{Dolan:2018:BDR:3192366.3192421}.  We
include an SC fence here to mimic the behavior of volatiles in the JMM.
\begin{gather*}
  \taglabel{past}
  \begin{gathered}
    (x\GETS 1 \SEMI   y^\mRA \GETS 1)
    \PAR
    (x\GETS 2 \SEMI \FENCE^\mSC \SEMI \IF{y^\mRA}\THEN r\GETS x \SEMI s\GETS x\FI)
    \\[-2ex]
    \hbox{\begin{tikzinline}[node distance=1.2em]
        \event{wx1}{\DW{x}{1}}{}
        \event{wy1}{\DWRel{y}{1}}{right=of wx0}
        \sync{wx1}{wy1}
        \event{wx2}{\DW{x}{2}}{right=3em of wy1}
        \event{f}{\DFS{\mSC}}{right=of wx2}
        \sync{wx2}{f}
        \event{ry1}{\DR[\mRA]{y}{1}}{right=of f}
        \sync{f}{ry1}
        \event{rx1}{\DR{x}{1}}{right=3em of ry1}
        \event{rx2}{\DR{x}{2}}{right=of rx1}
        \sync{ry1}{rx1}
        \sync[out=15,in=165]{ry1}{rx2}
        \rf[out=10,in=170]{wy1}{ry1}
        \wk[out=10,in=170]{wx1}{wx2}
        \wk[out=-170,in=-10]{rx1}{wx2}
        \bgellipsesmaller[yellow!50]{(wy1)(f)}{}
      \end{tikzinline}}
  \end{gathered}
\end{gather*}
The highlighted events are $L$-stable.  The order from $(\DR{x}{1})$ to
$(\DW{x}{2})$ is required by fulfillment, causing the cycle.  If the fence is
removed, there would be no order from $(\DW{x}{2})$ to
$(\DRAcq{y}{1})$, the highlighted events would no longer be $L$-stable, and
the execution would be allowed.  This more relaxed notion of ``past'' is not
expressible using \citeauthor{Dolan:2018:BDR:3192366.3192421}'s
synchronization primitives.

The notion of ``future'' is also richer here.  Consider \cite[Ex.~3]{Dolan:2018:BDR:3192366.3192421}:
\begin{gather*}
  \taglabel{future}
  \begin{gathered}
    (r\GETS 1 \SEMI \REF{r}\GETS 42\SEMI s\GETS \REF{r}\SEMI  x^\mRA\GETS r)
    \PAR
    (r\GETS x \SEMI \REF{r}\GETS 7)
    \\[-1ex]
    \hbox{\begin{tikzinline}[node distance=1.2em]
        \event{a1}{\DW{\REF{1}}{42}}{}
        \event{a2}{\DR{\REF{1}}{7}}{right=of a1}
        \wk{a1}{a2}
        \event{a4}{\DWRel{x}{1}}{right=of a2}
        \sync{a2}{a4}
        \event{b1}{\DR{x}{1}}{right=3em of a4}
        \event{b2}{\DW{\REF{1}}{7}}{right=of b1}
        \po{b1}{b2}
        \rf{a4}{b1}
        \wk[out=170,in=10]{b2}{a2}
      \end{tikzinline}}
  \end{gathered}
\end{gather*}
There is no interesting stable point here.  The execution is disallowed
because of a read from the causal future.  If we changed $x^\mRA$ to
$x^\mRLX$, then there would be no order from $(\DR{\REF{1}}{7})$ to
$(\DW[\mRLX]{x}{1})$, and the execution would be allowed.  The distinction
between ``causal future'' and ``temporal future'' is not expressible in
\citeauthor{Dolan:2018:BDR:3192366.3192421}'s operational semantics.

Our definition of $L$-sequentiality does not quite correspond to SC
executions, since actions may be elided by read/write elimination
(\textsection\ref{sec:refine}).  However, for any properly initialized
$L$-sequential pomset that uses elimination, there is larger $L$-sequential
pomset that does not use elimination. This can be shown inductively---in the
inductive step, writes that are introduced can be ignored by existing reads,
and reads that are introduced can be fulfilled, for some value, by some
preceding write.

### 2021-06-24 paper/intro.tex
swap order of le and rf
updates and fixes

@@
 \label{sec:intro}
+
+\emph{Sequentiality} is a \emph{leaky abstraction} \cite{leaky}.  For
+example, sequentiality tells us that when executing
+$(\PR{x}{r_1}\SEMI\PW{y}{r_2}),$ the assignment $\PR{x}{r_1}$ is executed
+before $\PW{y}{r_2}$.  Thus, one might reasonably expect that the final value
+of $r_1$ is independent of the initial value of $r_2$.  In most modern
+languages, however, this fails to hold when the program is run concurrently
+with $(\PR{y}{s}\SEMI\PW{x}{s})$,
+%$\PW{x}{\PR{y}{}}$,
+which copies $y$ to $x$.
+
+In certain cases it is possible to ban concurrent access using separation
+\cite{OHearn:2007:RCL:1235896.1236121}, or to accept inefficiency in order to
+obtain sequential consistency \cite{DBLP:conf/snapl/MarinoMMNS15}.  When
+these approaches are not available, however, we are left with an enormous gap
+in our understanding of one of the most basic elements of computing: the
+humble semicolon.
+Existing approaches either
+\begin{itemize}
+\item don't bother tracking dependencies, allowing ``thin air'' executions \cite{DBLP:phd/ethos/Batty15},
+\item use syntactic dependencies
+  \cite{alglave,DBLP:journals/corr/abs-1804-04214}, disabling compiler optimizations,
+\item use complex non-compositional operational models  \cite{Manson:2005:JMM:1047659.1040336,
+    DBLP:conf/esop/JagadeesanPR10,
+    DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}, 
+  %JMM, promising semantics, weakestMO, etc
+\item use prefixing \cite{DBLP:journals/pacmpl/JagadeesanJR20} or
+  continuation-passing \cite{DBLP:conf/esop/PaviottiCPWOB20} rather than
+  a direct presentation of sequential composition.
+\end{itemize}
+
+In this paper, we show that families of predicate transformers can be used to
+calculate semantic dependencies in a way that is \emph{compositional} and
+\emph{direct}: \emph{compositional} in that the denotation of
+$(\aCmd_1\SEMI \aCmd_2)$ can be computed from the denotation of $\aCmd_1$ and
+the denotation of $\aCmd_2$, and \emph{direct} in that these can be
+calculated independently.
+
+Our contributions:
+\begin{itemize}
+\item We define a denotational model for calculating dependencies.
+\item We apply the model to the \PwP{} semantics of
+  \cite{DBLP:journals/pacmpl/JagadeesanJR20} and to the \MRD{} semantics of
+  \cite{DBLP:conf/esop/PaviottiCPWOB20}.
+\item We provide a tool to execute litmus tests for both models.
+\item We prove \drfsc{} for \PwP{}; \MRD{} inherits the result from \rcXI.
+\item We prove a lowering result for \PwP{}.
+\item We extend the model to include many features.
+\end{itemize}
+
+\section{Overview}
+\label{sec:overview}
 
@@
 In a concurrent context, with shared variables
-($\aLoc$--$\cLoc$), the obvious generalizations
+($\aLoc$--$\cLoc$), the obvious generalization
 \begin{multicols}{2}
@@
 \noindent
-are unsound!  In particular, a read from a shared memory location may be fulfilled by a write
+is unsound!  In particular, a read from a shared memory location may be fulfilled by a write
 in another thread, invalidating \ref{wp-read}.  (We assume that expressions

### 2021-06-28 paper/arm.tex
updates and fixes
Reorg

@@
-\section{Arm}
+\section{Lowering PwT-MCA to Arm}
 \label{sec:arm}
@@
 
-\subsection{Arm Compilation 1}
+\subsection{PwT-MCA Lowering 1}
 \label{sec:arm1}
 
-We do not distinguish control dependencies from other dependencies, and
-therefore \ref{lob-rr} forces us to drop all dependencies between reads.  To
-achieve this, we modify the definition of $\labelingForm'_2$ in
-\reffig{fig:sem}.
-\begin{definition}
-  \label{def:semrr}
-  Let $\semrr{}$ be defined as in \reffig{fig:sem}, replacing the definition
-  of $\labelingForm'_2$ with:
-  \begin{displaymath}
-    \labelingForm'_2(\aEv)=
-    \begin{cases}
-      \aTr[1]{}{\labelingForm_2(\aEv)} & \text{if}\; \labeling(\aEv) \;\text{is a read}
-      \\
-      \aTr[1]{\Cdown{\aEv}}{\labelingForm_2(\aEv)} & \text{otherwise, where}\; \Cdown{\aEv}=\{ \cEv \mid \cEv \lt \aEv \}
-    \end{cases}
-    % $\Cdown{\aEv}=\{ \cEv \mid \cEv \lt \aEv \}$ if $\labeling(\aEv)$ is a write, and $\Cdown{\aEv}=\aEvs_1$, otherwise
-    % \Cdown{\aEv}=
-    % \begin{cases}
-    %   \{ \cEv \mid \cEv \lt \aEv \} & \textif \labeling(\aEv) \;\text{is a write}
-    %   \\
-    %   \aEvs_1 & \textotherwise
-    % \end{cases}
-  \end{displaymath}
-\end{definition}
-
-Even with this small change, the optimal lowering for \armeight{} is unsound
-for our semantics.  The optimal lowering maps relaxed access to \LDR/\STR{} and
+The optimal lowering for \armeight{} is unsound
+for \PwTmca.  The optimal lowering maps relaxed access to \LDR/\STR{} and
 non-relaxed access to \LDAR/\STLR{} \citep{DBLP:journals/pacmpl/PodkopaevLV19}.
@@
 
-\subsection{Arm Compilation 2}
+\subsection{PwT-MCA Lowering 2}
 \label{sec:arm2}
@@
     \end{math}.
+  \item \textcolor{red}{TODO: I think this should order W->R if there is no
+      rf the other way}
   \end{itemize}  

### 2021-06-28 paper/mca.tex
Reorg

\section{PwT-MCA}

\input{fig-sem.tex}
See \reffig{fig:seq}.

### 2021-06-28 paper/oopsla.tex
updates and fixes
Reorg

@@
-\section{Differences with ``Pomsets with Preconditions''}
+\subsection{Comparison to ``A Promising Semantics 2.1'' [POPL 2017]}
+\label{sec:promising}
+
+[Todo.]
+
+Case analysis gives very weak results when combined with thread inlining.
+See \cite[\textsection B.1]{DBLP:journals/pacmpl/ChakrabortyV19appendix}.
+These happen by performing transformations that: 
+(1) introduce conditionals,
+(2) inline two threads on both sides of the introduced conditional,
+(3) choose different orders for the two threads for the two sides of the conditional.
+
+Case analysis gives very weak results when combined with read introduction.
+See \cite{promising-ldrf}.
+These happen by performing transformations that: 
+(1) introduce reads,
+(2) introduce conditionals,
+(3) choose different values for the reads on the two sides of the conditional.
+
+The fact that the semantics is not verifiable a posteriori is something it
+shares with \weakestmo{}, where the justification relation must be built
+inductively.
+
+\weakestmo{} admits FADD, but \PS{} does not.
+\PS{} admits CohCYC, but \weakestmo{} does not.
+
+\subsection{Comparison to ``Pomsets with Preconditions'' [OOPSLA 2020]}
 \label{sec:diff}
+
+\PwTmca{} is closely related to \PwP{} model of
+\citep{DBLP:journals/pacmpl/JagadeesanJR20}.  The major difference is that
+\PwTmca{} supports sequential composition.  In the remainder of this section,
+we discuss other differences.  We also point out some errors in
+\cite{DBLP:journals/pacmpl/JagadeesanJR20}, all of which have been confirmed
+by the authors.
 
@@
 \end{align*}
+
+\myparagraph{Augmentation of Preconditions}
+\jjr{} allows augmentation of preconditions.
+As discussed in \textsection\ref{sec:delay}, this causes associativity to fail
+for $\rdelay$, at least when attempting to validate \reflem{lem:if}\eqref{lem:ifelse:if:if1}--\eqref{lem:ifelse:if:if2}
+Thus, we use \emph{weakest} preconditions, rather than general preconditions.
+As a result, we fail to validate the following
+refinement:
+\begin{math}
+  \aPSS_1
+  \not\supseteq
+  \xIFTHEN{\aForm}{\aPSS_1}{}.
+\end{math}
 
@@
 
-\section{A Note on Mixed-Mode Data Races}
-
-In preparing this paper, we came across the following example, which appears
-to invalidate Theorem 4.1 of \cite{DBLP:conf/ppopp/DongolJR19}.
-\begin{gather}
-  \nonumber
-  \PW{x}{1}\SEMI
-  \PW[\mREL]{y}{1}\SEMI
-  \PR[\mACQ]{x}{r}
-  \PAR
-  \IF{\PR[\mACQ]{y}{}}\THEN \PW[\mREL]{x}{2}\FI
-  \\
-  \tag{\P}
-  \label{mix1}
-  \hbox{\begin{tikzinline}[node distance=1.5em]
-      \event{a1}{\DW{x}{1}}{}
-      \raevent{a2}{\DW[\mREL]{y}{1}}{right=of a1}
-      \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
-      \raevent{b1}{\DR[\mACQ]{y}{1}}{right=3em of a3}
-      % \raevent{b1}{\DR[\mACQ]{y}{1}}{below=of a1}
-      \raevent{b2}{\DW[\mREL]{x}{2}}{right=of b1}
-      \sync{a1}{a2}
-      \rf[out=20,in=160]{a1}{a3}
-      \rf[out=20,in=160]{a2}{b1}
-      \wk[out=-20,in=-160]{a3}{b2}
-      \sync{b1}{b2}
-      % \node(ai)[left=3em of a1]{};
-      % \bgoval[yellow!50]{(ai)}{P}
-      % \bgoval[pink!50]{(a1)(a2)(b1)(b2)}{P'\setminus P}
-      % \bgoval[green!10]{(a3)}{P'''\setminus P'}
-    \end{tikzinline}}
-  \\
-  \nonumber
-  %\label{mix2}
-  \hbox{\begin{tikzinline}[node distance=1.5em]
-      \event{a1}{\DW{x}{1}}{}
-      \raevent{a2}{\DW[\mREL]{y}{1}}{right=of a1}
-      \raevent{a3}{\DR[\mACQ]{x}{2}}{right=of a2}
-      \raevent{b1}{\DR[\mACQ]{y}{1}}{right=3em of a3}
-      \raevent{b2}{\DW[\mREL]{x}{2}}{right=of b1}
-      \sync{a1}{a2}
-      \rf[out=20,in=160]{a2}{b1}
-      \rf[out=160,in=20]{b2}{a3}
-      \sync{b1}{b2}
-    \end{tikzinline}}
-\end{gather}
-The program is data-race free.  The two executions shown are the only
-top-level executions that include $\DWP[\mREL]{x}{2}$.
-
-Theorem 4.1 of \cite{DBLP:conf/ppopp/DongolJR19} is stated by extending
-execution sequences.  In the terminology of
-\cite{DBLP:conf/ppopp/DongolJR19}, a read is \emph{$L$-weak} if it is
-sequentially stale.  Let
-\begin{math}
-  \rho=\DWP{x}{1}\allowbreak
-  \DWP[\mREL]{y}{1}\allowbreak
-  \DRP[\mACQ]{y}{1}\allowbreak
-  \DWP[\mREL]{x}{2}
-\end{math}
-be a sequence and
-\begin{math}
-  \alpha=\DRP[\mACQ]{x}{1}.
-\end{math}
-$\rho$ is $L$-sequential and $\alpha$ is $L$-weak in $\rho\alpha$.  But there
-is no execution of this program that includes a data race, contradicting the
-theorem.  The error seems to be in Lemma A.4 of
-\cite{DBLP:conf/ppopp/DongolJR19}, which states that if $\alpha$ is $L$-weak
-after an $L$-sequential $\rho$, then $\alpha$ must be in a data race.  That
-is clearly false here, since $\DRP[\mACQ]{x}{1}$ is stale, but the program is
-data race free.
-
-In proving the SC-LDRF result in \jjr{\textsection8}, we noted that our proof
-technique is more robust than that of \cite{DBLP:conf/ppopp/DongolJR19},
-because it limits the prefixes that must be considered.  In \eqref{mix1}, the
-induction hypothesis requires that we add $\DRP[\mACQ]{x}{1}$ before
-$\DWP[\mREL]{x}{2}$ since $\DRP[\mACQ]{x}{1}\xwk\DWP[\mREL]{x}{2}$.  In
-particular,
-\begin{gather*}
-  \hbox{\begin{tikzinline}[node distance=1.5em]
-      \event{a1}{\DW{x}{1}}{}
-      \raevent{a2}{\DW[\mREL]{y}{1}}{right=of a1}
-      % \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
-      \raevent{b1}{\DR[\mACQ]{y}{1}}{right=3em of a3}
-      % \raevent{b1}{\DR[\mACQ]{y}{1}}{below=of a1}
-      \raevent{b2}{\DW[\mREL]{x}{2}}{right=of b1}
-      \sync{a1}{a2}
-      % \rf[out=20,in=160]{a1}{a3}
-      \rf[out=20,in=160]{a2}{b1}
-      % \wk[out=-20,in=-160]{a3}{b2}
-      \sync{b1}{b2}
-      % \node(ai)[left=3em of a1]{};
-      % \bgoval[yellow!50]{(ai)}{P}
-      % \bgoval[pink!50]{(a1)(a2)(b1)(b2)}{P'\setminus P}
-      % \bgoval[green!10]{(a3)}{P'''\setminus P'}
-    \end{tikzinline}}
-\end{gather*}
-is not a downset of \eqref{mix1}, because
-$\DRP[\mACQ]{x}{1}\xwk\DWP[\mREL]{x}{2}$.  As we noted in \jjr{\textsection8},
-this affects the inductive order in which we move across pomsets, but does
-not affect the set of pomsets that are considered.  In particular,
-\begin{gather*}
-  \hbox{\begin{tikzinline}[node distance=1.5em]
-      \event{a1}{\DW{x}{1}}{}
-      \raevent{a2}{\DW[\mREL]{y}{1}}{right=of a1}
-      % \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
-      \raevent{b1}{\DR[\mACQ]{y}{1}}{right=3em of a3}
-      % \raevent{b1}{\DR[\mACQ]{y}{1}}{below=of a1}
-      % \raevent{b2}{\DW[\mREL]{x}{2}}{right=of b1}
-      \sync{a1}{a2}
-      % \rf[out=20,in=160]{a1}{a3}
-      \rf[out=20,in=160]{a2}{b1}
-      % \wk[out=-20,in=-160]{a3}{b2}
-      % \sync{b1}{b2}
-      % \node(ai)[left=3em of a1]{};
-      % \bgoval[yellow!50]{(ai)}{P}
-      % \bgoval[pink!50]{(a1)(a2)(b1)(b2)}{P'\setminus P}
-      % \bgoval[green!10]{(a3)}{P'''\setminus P'}
-    \end{tikzinline}}
-\end{gather*}
-is a downset of \eqref{mix1}.
-
 \endinput

### 2021-06-28 paper/arm.tex
cleanup

@@
 \item \label{lob-rf}
-  write-to-read order of $\rrfx$, required by \ref{seq-rf-le},
+  write-to-read order of $\rrfx$, required by \ref{pom-rf-le},
 \item \label{lob-wr}
@@
 for relaxed access.  In the next section we recover the optimal lowering by
-adopting an alternative semantics for \ref{seq-rf-le} and \ref{seq-le-delays}.
+adopting an alternative semantics for \ref{pom-rf-le} and \ref{seq-le-delays}.
 
@@
 
-The change to the semantics is small: we weaken the relationship between $\rrfx$
-and $\le$ in \ref{seq-rf-le}.  Rather than ensuring that there is no
-\emph{global} blocker for a sequentially fulfilled read \eqref{seq-rf-le}, we
-require only that there is no \emph{thread-local} blocker \eqref{seq-rf-le-rf}.
-This change both allows and requires us to weaken the definition of
-\emph{delays} to drop write-to-read order from $\eqreorderco$.
-\begin{definition}
-  \label{def:sem:frf}
-  Let $\frf{\semrr{}}$ be defined as for $\semrr{}$ in
-  \refdef{def:semrr}/\reffig{fig:sem}, changing \ref{seq-rf-le} and
-  \ref{seq-le-delays}:
-  \begin{itemize}
-  \item[{\labeltext[\frf{\textsc{s}7b}]{(\frf{\textsc{s}7b})}{seq-rf-le-rf}}]
-    if $\labeling_1(\cEv) \rblocks \labeling_2(\aEv)$ then $\bEv\xrfx\aEv$
-    implies $\cEv\le\bEv$,
-  \item[{\labeltext[\frf{\textsc{s}6b}]{(\frf{\textsc{s}6b})}{seq-le-delays-rf}}]
-    if $\labeling_1(\bEv) \rdelaysp \labeling_2(\aEv)$ then $\bEv\le\aEv$,\\
-    where $\rdelayspdef$ replaces $\eqreorderco$ in \refdef{def:actions} of
-    $\rdelaysdef$ by
-    \begin{math}
-      {\reorderlws}
-      =
-      \{(\DW{\aLoc}{}, \DW{\aLoc}{}),\;(\DR{\aLoc}{}, \DW{\aLoc}{})\}
-    \end{math}.
-  \item \textcolor{red}{TODO: I think this should order W->R if there is no
-      rf the other way}
-  \end{itemize}  
-\end{definition}
-The acronym $\textsf{lws}$ is adopted from \armeight.  It stands for
-\emph{Local Write Successor}.
-
-With the weakening of \ref{seq-rf-le-rf}, we must be careful not to allow
-spurious pairs to be added to the $\rrfx$ relation.  The use of
-$\rextendsdef{}{}$ in \ref{if-rf-extends} does this, ensuring that new
-$\rrfx$ is not introduced between events in $\aEvs_1\cap\aEvs_2$ when coalescing.
-This is necessary to ensure that
-\begin{math}
-  \frf{\semrr{\IF{b}\THEN\PR{x}{r}\PAR\PW{x}{1}\ELSE\PR{x}{r}\SEMI\PW{x}{1}\FI}}
-\end{math}
-does not include 
-\begin{math}
-  \smash{\hbox{\begin{tikzinlinesmall}[node distance=1.5em]
-        \event{a}{\DR{x}{1}}{}
-        \event{b}{\DW{x}{1}}{right=of a}
-        \rfint[out=165,in=15]{b}{a}
-        \wki{a}{b}
-      \end{tikzinlinesmall}}}
-\end{math}, taking $\rrfx$ from the left and $\le$ from the right.
+% The change to the semantics is small: we weaken the relationship between $\rrfx$
+% and $\le$ in \ref{seq-rf-le}.  Rather than ensuring that there is no
+% \emph{global} blocker for a sequentially fulfilled read \eqref{seq-rf-le}, we
+% require only that there is no \emph{thread-local} blocker \eqref{seq-rf-le-rf}.
+% This change both allows and requires us to weaken the definition of
+% \emph{delays} to drop write-to-read order from $\eqreorderco$.
+% \begin{definition}
+%   \label{def:sem:frf}
+%   Let $\frf{\semrr{}}$ be defined as for $\semrr{}$ in
+%   \refdef{def:semrr}/\reffig{fig:sem}, changing \ref{seq-rf-le} and
+%   \ref{seq-le-delays}:
+%   \begin{itemize}
+%   \item[{\labeltext[\frf{\textsc{s}7b}]{(\frf{\textsc{s}7b})}{seq-rf-le-rf}}]
+%     if $\labeling_1(\cEv) \rblocks \labeling_2(\aEv)$ and $\bEv\xrfx\aEv$
+%     then $\cEv\le\bEv$,
+%   \item[{\labeltext[\frf{\textsc{s}6b}]{(\frf{\textsc{s}6b})}{seq-le-delays-rf}}]
+%     if $\labeling_1(\bEv) \rdelaysp \labeling_2(\aEv)$ then $\bEv\le\aEv$,\\
+%     where $\rdelayspdef$ replaces $\eqreorderco$ in \refdef{def:actions} of
+%     $\rdelaysdef$ by
+%     \begin{math}
+%       {\reorderlws}
+%       =
+%       \{(\DW{\aLoc}{}, \DW{\aLoc}{}),\;(\DR{\aLoc}{}, \DW{\aLoc}{})\}
+%     \end{math}.
+%   \item \textcolor{red}{TODO: I think this should order W->R if there is no
+%       rf the other way}
+%   \end{itemize}  
+% \end{definition}
+% The acronym $\textsf{lws}$ is adopted from \armeight.  It stands for
+% \emph{Local Write Successor}.
 

### 2021-06-28 paper/mca.tex
cleanup

@@
 
-\input{fig-sem.tex}
-See \reffig{fig:seq}.
+From the data model, we require an additional binary relation over
+$\rdelays\subseteq(\Act\times\Act)$.  For the actions in this paper, we
+define this as follows.
+
+\begin{scope}
+  Let ${\eqreorderco}$ capture write-write, read-write coherence:
+  \begin{math}
+    {\eqreorderco}
+    =
+    \{(\DW{\aLoc}{}\Cb \DW{\aLoc}{})\Cc(\DR{\aLoc}{}\Cb \DW{\aLoc}{})\Cc(\DW{\aLoc}{}\Cb \DR{\aLoc}{})\}
+  \end{math}.
+
+  Let ${\reorderra}$ capture conflict due to synchronization:
+  \begin{math}
+    {\reorderra}
+    =
+    \{(\aAct\Cb             \DW[\gemode\mREL]{}{}     )\Cc
+    (\aAct\Cb               \DF{\gemode\fREL}        )\Cc
+    (\DR{}{}\Cb             \DF{\gemode\fACQ}        )\Cc
+    % (\DR{\aLoc}{}\Cb        \DR[\gemode\mACQ]{\aLoc}{})\Cc
+    (\DR[\gemode\mACQ]{}{}\Cb\aAct                    )\Cc
+    (\DF{\gemode\fACQ}\Cb   \aAct                    )\Cc
+    (\DF{\gemode\fREL}\Cb   \DW{}{}                  )\Cc
+    (\DW[\gemode\mREL]{\aLoc}{}\Cb\DW{\aLoc}{})\}
+  \end{math}.
+
+  Let ${\eqreordersc}$ capture conflict due to $\mSC$ access:
+  \begin{math}
+    {\eqreordersc}
+    =
+    % \{(\DX[\mSC]{}{}\Cb\DX[\mSC]{}{})\}
+    \{(\DW[\mSC]{}{}\Cb \DW[\mSC]{}{})\Cc(\DR[\mSC]{}{}\Cb \DW[\mSC]{}{}) \Cc(\DW[\mSC]{}{}\Cb \DR[\mSC]{}{})\Cc(\DR[\mSC]{}{}\Cb \DR[\mSC]{}{})\}
+  \end{math}.
+
+  We say $\aAct \rdelaysdef \bAct$ if $\aAct\eqreorderco\bAct$ or $\aAct\reorderra\bAct$ or $\aAct\eqreordersc\bAct$.
+\end{scope}
+
+\begin{scope}
+  Add the following to the definition \emph{pomset} and \emph{complete
+    pomset} from \refdef{def:pomset}:
+  \begin{enumerate}[,label=(\textsc{m}\arabic*),ref=\textsc{m}\arabic*]
+    \setcounter{enumi}{\value{Brf}}
+  \item \label{pom-rf} \makecounter{rf} ${\rrfx} \subseteq \aEvs\times\aEvs$
+    is an injective relation capturing \emph{reads-from}, such that
+    % \end{enumerate}
+    % A pomset is a \emph{candidate} if there is an injective relation
+    % ${\rrfx} : \aEvs\times\aEvs$, capturing \emph{reads-from}, such that
+    \begin{enumerate}
+      % \begin{enumerate}[,label=(\textsc{i}\arabic*),ref=\textsc{i}\arabic*]
+      % \item \label{rf-injective}
+      %   if $\bEv\xrfx\aEv$ and $\cEv\xrfx\aEv$ then $\bEv=\cEv$, that is,
+      %   ${\rrfx}$ is injective,
+    \item \label{pom-rf-match} if $\bEv\xrfx\aEv$ then
+      $\labeling(\bEv) \rmatches \labeling(\aEv)$,
+    \item \label{pom-rf-block} if $\bEv\xrfx\aEv$ and
+      $\labeling(\cEv) \rblocks \labeling(\aEv)$ then either $\cEv\le\bEv$ or
+      $\aEv\le\cEv$,
+    \item \label{pom-rf-le} if $\bEv\xrfx\aEv$ then $\bEv\le\aEv$.
+    \end{enumerate}
+  \end{enumerate}
+  \begin{enumerate}[,label=(\textsc{c}\arabic*),ref=\textsc{c}\arabic*]
+    \setcounter{enumi}{\value{Brf}}
+  \item \label{top-rf}
+    if $\labeling(\aEv)$ is a $\sread$ then there is some $\bEv\xrfx\aEv$.
+  \end{enumerate}
+\end{scope}
+
+% To include \RMW{}s, also add
+% \begin{enumerate}
+% \item[(\textsc{p}10)]\;
+%   ${\rrmw}=\PBR{{\rrmw_1}\cup{\rrmw_2}}$,
+% \end{enumerate}
+
+\input{fig-mca1.tex}
+\input{fig-mca2.tex}
+The semantic rules are given in \reffig{fig:mca1}.  Let
+\begin{math}
+  \sembase{\aCmd_1 \LPAR[\bThrd] \aCmd_2} = \sLPAR{\sembase{\aCmd_1}}{\sembase{\aCmd_2}}.
+\end{math}
+
+In the second semantics, we weaken the relationship between $\rrfx$
+and $\le$ in \ref{pom-rf-le}.  Rather than ensuring that there is no
+\emph{global} blocker for a sequentially fulfilled read \eqref{pom-rf-le}, we
+require only that there is no \emph{thread-local} blocker
+\eqref{seq-rf-le-rf}.
+This change both allows and requires us to weaken the definition of
+\emph{delays} to drop write-to-read order from $\eqreorderco$.
+
+With the weakening of \ref{seq-rf-le-rf}, we must be careful not to allow
+spurious pairs to be added to the $\rrfx$ relation.  The use of
+$\rextendsdef{}{}$ in \ref{if-rf-extends} does this, ensuring that new
+$\rrfx$ is not introduced between events in $\aEvs_1\cap\aEvs_2$ when coalescing.
+This is necessary to ensure that
+\begin{math}
+  \frf{\semrr{\IF{b}\THEN\PR{x}{r}\PAR\PW{x}{1}\ELSE\PR{x}{r}\SEMI\PW{x}{1}\FI}}
+\end{math}
+does not include 
+\begin{math}
+  \smash{\hbox{\begin{tikzinlinesmall}[node distance=1.5em]
+        \event{a}{\DR{x}{1}}{}
+        \event{b}{\DW{x}{1}}{right=of a}
+        \rfint[out=165,in=15]{b}{a}
+        \wki{a}{b}
+      \end{tikzinlinesmall}}}
+\end{math}, taking $\rrfx$ from the left and $\le$ from the right.
+
+Replace \eqref{pom-rf-le} by
+\begin{enumerate}
+\item[{\labeltext[\textsc{m}7c$'$]{(\textsc{m}7c$'$)}{pom-rf-le'}}]
+ if $\bEv\xrfx\aEv$ then either $\bEv\le\aEv$ or $\aEv\le\bEv$.
+\end{enumerate}
+Derive $\rdelayspdef$ from $\rdelaysdef$ by replacing $\eqreorderco$ with
+\begin{math}
+  {\reorderlws}
+  =
+  \{(\DW{\aLoc}{}, \DW{\aLoc}{}),\;(\DR{\aLoc}{}, \DW{\aLoc}{})\}.
+\end{math}
+The acronym $\textsf{lws}$ is adopted from \armeight.  It stands for
+\emph{Local Write Successor}.
+
+The semantic rules are given in \reffig{fig:mca2}.
+
+A derived property of sequential composition:
+\begin{center}
+  if $\labeling_1(\cEv) \rblocks \labeling_2(\aEv)$ and $\bEv\xrfx\aEv$
+  then $\cEv\le\bEv$.  
+\end{center}

### 2021-06-28 paper/outro.tex
better lemma
cleanup

--- 2021-04-17 paper/outro.tex
@@
 
+The promising semantics (\PS{}) validates read introduction whereas \PwT{}
+does not.  As a result \PS{} admits behaviors that break reasoning about
+temporal safety properties (see \textsection\ref{sec:promising}).
+Nonetheless, read introduction is ubiquitous is some compilers.  It would be
+interesting to know if there is a performance penalty for banning read
+introduction.
+
 An earlier version of this paper has been mechanized in Agda; it would be
 reassuring to update the mechanization to bring it in line with the current state.
+
+We don't handle access elimination.
 

### 2021-06-29 paper/arm.tex
revert change to s6c

@@
 sequential composition slightly.  In particular, we must lose
-\reflem{lem:rf:implies:le}, which states that $\bEv\xrfx\aEv$ implies
+\ref{pom-rf-le}, which states that $\bEv\xrfx\aEv$ implies
 $\bEv\le\aEv$.  Revisiting the example in the last subsection, we essentially
@@
 
-We emphasize that \reflem{lem:rf:implies:le} fails for $\frf{\semrr{}}$,
-since $\bEv\xrfx\aEv$ may not imply $\bEv\le\aEv$ when $\bEv$ and $\aEv$ come
+We emphasize that \ref{pom-rf-le} does not hold for $\frf{\semrr{}}$:
+ $\bEv\xrfx\aEv$ may not imply $\bEv\le\aEv$ when $\bEv$ and $\aEv$ come
 from different sides of a sequential composition.  This means that $\rrfx$

### 2021-06-29 paper/arm.tex
a minor comment on PWT->Arm

@@
 fences\footnote{Fences are not actions in \armeight{}, which complicates the
-  theorem statements.}.
+  theorem statements.%
+  \app{Is it true? AFAIU, it isn't, they are explicitly used in definition of $\textsc{bob}$.
+    I've just found that Jade prefers to omit them in pictures.}}.
 

### 2021-06-29 paper/mca.tex
revert change to s6c
first draft of PwT-C11

@@
-\section{PwT-MCA}
+\section{\PwTmcaTITLE{}: Pomsets with Predicate Transformers for MCA}
 
-From the data model, we require an additional binary relation over
-$\rdelays\subseteq(\Act\times\Act)$.  For the actions in this paper, we
+We define two models for \mca{} architectures: \PwTmca{1} and \PwTmca{2}.
+We drop the subscript when referring to both models.
+
+\subsection{First Semantics}
+
+\begin{definition}
+  \label{def:pwt:mca1}
+  A \PwTmca{1} is a \PwT{} (\refdef{def:pomset}) equipped
+  with a relation $\rrfx$ such that 
+  \begin{enumerate}[,label=(\textsc{m}\arabic*),ref=\textsc{m}\arabic*]
+    \setcounter{enumi}{\value{Brf}}
+  \item \label{pom-rf} \makecounter{rf} ${\rrfx} \subseteq \aEvs\times\aEvs$
+    is an injective relation capturing \emph{reads-from}, such that
+    % \end{enumerate}
+    % A pomset is a \emph{candidate} if there is an injective relation
+    % ${\rrfx} : \aEvs\times\aEvs$, capturing \emph{reads-from}, such that
+    \begin{enumerate}
+      % \begin{enumerate}[,label=(\textsc{i}\arabic*),ref=\textsc{i}\arabic*]
+      % \item \label{rf-injective}
+      %   if $\bEv\xrfx\aEv$ and $\cEv\xrfx\aEv$ then $\bEv=\cEv$, that is,
+      %   ${\rrfx}$ is injective,
+    \item \label{pom-rf-match} if $\bEv\xrfx\aEv$ then
+      $\labeling(\bEv) \rmatches \labeling(\aEv)$,
+    \item \label{pom-rf-block} if $\bEv\xrfx\aEv$ and
+      $\labeling(\cEv) \rblocks \labeling(\aEv)$ then either $\cEv\le\bEv$ or
+      $\aEv\le\cEv$,
+    \item \label{pom-rf-le} if $\bEv\xrfx\aEv$ then $\bEv\le\aEv$.
+    \end{enumerate}
+  \end{enumerate}
+  A \PwTmca{} is \emph{complete} if it is a complete \PwT{} (\refdef{def:pomset}) and
+  \begin{enumerate}[,label=(\textsc{c}\arabic*),ref=\textsc{c}\arabic*]
+    \setcounter{enumi}{\value{Brf}}
+  \item \label{top-rf}
+    if $\labeling(\aEv)$ is a $\sread$ then there is some $\bEv\xrfx\aEv$.
+  \end{enumerate}
+\end{definition}
+
+% To include \RMW{}s, also add
+% \begin{enumerate}
+% \item[(\textsc{p}10)]\;
+%   ${\rrmw}=\PBR{{\rrmw_1}\cup{\rrmw_2}}$,
+% \end{enumerate}
+
+\input{fig-mca1.tex}
+\input{fig-mca2.tex}
+
+The semantic rules are given in \reffig{fig:mca1}.  Let
+\begin{math}
+  \sembase{\aCmd_1 \LPAR[\bThrd] \aCmd_2} = \sLPAR{\sembase{\aCmd_1}}{\sembase{\aCmd_2}}.
+\end{math}
+
+From the data model, we require an additional binary relation 
+${\rdelays}\subseteq(\Act\times\Act)$.  For the actions in this paper, we
 define this as follows.
@@
 
-\begin{scope}
-  Add the following to the definition \emph{pomset} and \emph{complete
-    pomset} from \refdef{def:pomset}:
-  \begin{enumerate}[,label=(\textsc{m}\arabic*),ref=\textsc{m}\arabic*]
-    \setcounter{enumi}{\value{Brf}}
-  \item \label{pom-rf} \makecounter{rf} ${\rrfx} \subseteq \aEvs\times\aEvs$
-    is an injective relation capturing \emph{reads-from}, such that
-    % \end{enumerate}
-    % A pomset is a \emph{candidate} if there is an injective relation
-    % ${\rrfx} : \aEvs\times\aEvs$, capturing \emph{reads-from}, such that
-    \begin{enumerate}
-      % \begin{enumerate}[,label=(\textsc{i}\arabic*),ref=\textsc{i}\arabic*]
-      % \item \label{rf-injective}
-      %   if $\bEv\xrfx\aEv$ and $\cEv\xrfx\aEv$ then $\bEv=\cEv$, that is,
-      %   ${\rrfx}$ is injective,
-    \item \label{pom-rf-match} if $\bEv\xrfx\aEv$ then
-      $\labeling(\bEv) \rmatches \labeling(\aEv)$,
-    \item \label{pom-rf-block} if $\bEv\xrfx\aEv$ and
-      $\labeling(\cEv) \rblocks \labeling(\aEv)$ then either $\cEv\le\bEv$ or
-      $\aEv\le\cEv$,
-    \item \label{pom-rf-le} if $\bEv\xrfx\aEv$ then $\bEv\le\aEv$.
-    \end{enumerate}
-  \end{enumerate}
-  \begin{enumerate}[,label=(\textsc{c}\arabic*),ref=\textsc{c}\arabic*]
-    \setcounter{enumi}{\value{Brf}}
-  \item \label{top-rf}
-    if $\labeling(\aEv)$ is a $\sread$ then there is some $\bEv\xrfx\aEv$.
-  \end{enumerate}
-\end{scope}
-
-% To include \RMW{}s, also add
-% \begin{enumerate}
-% \item[(\textsc{p}10)]\;
-%   ${\rrmw}=\PBR{{\rrmw_1}\cup{\rrmw_2}}$,
-% \end{enumerate}
-
-\input{fig-mca1.tex}
-\input{fig-mca2.tex}
-The semantic rules are given in \reffig{fig:mca1}.  Let
-\begin{math}
-  \sembase{\aCmd_1 \LPAR[\bThrd] \aCmd_2} = \sLPAR{\sembase{\aCmd_1}}{\sembase{\aCmd_2}}.
-\end{math}
+\subsection{Second Semantics}
 
@@
 
-Replace \eqref{pom-rf-le} by
+\begin{definition}
+  \label{def:pwt:mca2}
+  A \PwTmca{2} is a \PwT{} (\refdef{def:pomset}) equipped
+  with a relation $\rrfx$ such that 
+  \begin{enumerate}[,label=(\textsc{m}\arabic*),ref=\textsc{m}\arabic*]
+    \setcounter{enumi}{\value{Brf}}
+  \item \label{pom-rf'} ${\rrfx} \subseteq \aEvs\times\aEvs$
+    is an injective relation capturing \emph{reads-from}, such that
+    % \end{enumerate}
+    % A pomset is a \emph{candidate} if there is an injective relation
+    % ${\rrfx} : \aEvs\times\aEvs$, capturing \emph{reads-from}, such that
+    \begin{multicols}{2}    
       \begin{enumerate}
-\item[{\labeltext[\textsc{m}7c$'$]{(\textsc{m}7c$'$)}{pom-rf-le'}}]
+        % \begin{enumerate}[,label=(\textsc{i}\arabic*),ref=\textsc{i}\arabic*]
+        % \item \label{rf-injective}
+        %   if $\bEv\xrfx\aEv$ and $\cEv\xrfx\aEv$ then $\bEv=\cEv$, that is,
+        %   ${\rrfx}$ is injective,
+      \item[\eqref{pom-rf-match}]
+        \eqref{pom-rf-block}\;
+        as in \refdef{def:pwt:mca1},
+        \setcounter{enumii}{2}
+      \item \label{pom-rf-le'} 
         if $\bEv\xrfx\aEv$ then either $\bEv\le\aEv$ or $\aEv\le\bEv$.
       \end{enumerate}
+    \end{multicols}
+  \end{enumerate}
+\end{definition}
 Derive $\rdelayspdef$ from $\rdelaysdef$ by replacing $\eqreorderco$ with
@@
 The semantic rules are given in \reffig{fig:mca2}.
-
-A derived property of sequential composition:
-\begin{center}
-  if $\labeling_1(\cEv) \rblocks \labeling_2(\aEv)$ and $\bEv\xrfx\aEv$
-  then $\cEv\le\bEv$.  
-\end{center}

### 2021-06-29 paper/arm.tex
first draft of PwT-C11
accepted anton's edit

--- 2021-06-29 paper/arm.tex
@@
-\section{Lowering PwT-MCA to Arm}
+\section{Lowering \PwTmcaTITLE{} to Arm}
 \label{sec:arm}
-For simplicity, we restrict to top level parallel composition and ignore
-fences\footnote{Fences are not actions in \armeight{}, which complicates the
-  theorem statements.}.
+For simplicity, we restrict to top level parallel composition.
+% and ignore
+% fences\footnote{Fences are not actions in \armeight{}, which complicates the
+%   theorem statements.%
+%   \app{Is it true? AFAIU, it isn't, they are explicitly used in definition of $\textsc{bob}$.
+%     I've just found that Jade prefers to omit them in pictures.}}.
+% JWR: I didn't know that!  It's better with them...
 
@@
 
-\subsection{PwT-MCA Lowering 1}
+\subsection{Lowering \PwTmcaTITLE{1} to Arm}
 \label{sec:arm1}
@@
 The optimal lowering for \armeight{} is unsound
-for \PwTmca.  The optimal lowering maps relaxed access to \LDR/\STR{} and
+for \PwTmca{1}.  The optimal lowering maps relaxed access to \LDR/\STR{} and
 non-relaxed access to \LDAR/\STLR{} \citep{DBLP:journals/pacmpl/PodkopaevLV19}.
@@
 
-\subsection{PwT-MCA Lowering 2}
+\subsection{Lowering \PwTmcaTITLE{2} to Arm}
 \label{sec:arm2}

### 2021-06-29 paper/arm.tex
small cleanup of arm semantics

@@
 \label{sec:arm}
-For simplicity, we restrict to top level parallel composition.
+For simplicity, we restrict to top-level parallel composition.
 % and ignore
@@
   \item $\labeling: \aEvs \fun \Act$ defines a {label} for each event,
-  \item ${\rpoloc} : \aEvs\times\aEvs$, is a per-thread, per-location total
+  \item ${\rpoloc} \subseteq \aEvs{\times}\aEvs$, is a per-thread, per-location total
     order, capturing \emph{per-location program order},
-  \item ${\rlob} : \aEvs\times\aEvs$, is a per-thread partial order capturing
+  \item ${\rlob} \subseteq \aEvs\times\aEvs$, is a per-thread partial order capturing
     \emph{locally-ordered-before}, such that
@@
   \item \label{arm-co}
-    ${\rco} : \aEvs\times\aEvs$, is a per-location total order on
+    ${\rco} \subseteq \aEvs\times\aEvs$, is a per-location total order on
     writes, capturing \emph{coherence}, 
   \item \label{arm-rf}
-    ${\rrfx} : \aEvs\times\aEvs$, is a surjective and injective
-    relation on reads, capturing \emph{reads-from}, such that
+    ${\rrfx} \subseteq \aEvs\times\aEvs$, is a relation, capturing \emph{reads-from}, such that
     \begin{enumerate}
+    \item \label{arm-rf-reads}
+      ${\rrfx}$ is surjective and injective relation on $\{\aEv\in\aEvs\mid\labeling(\aEv)\;\text{is a}\;\sread\}$,
     \item \label{arm-match}

### 2021-06-30 paper/arm.tex
another stab at c11

@@
 %     \end{math}.
-%   \item \textcolor{red}{TODO: I think this should order W->R if there is no
+%   \item \todo{I think this should order W->R if there is no
 %       rf the other way}
@@
 
-We emphasize that \ref{pom-rf-le} does not hold for $\frf{\semrr{}}$:
- $\bEv\xrfx\aEv$ may not imply $\bEv\le\aEv$ when $\bEv$ and $\aEv$ come
-from different sides of a sequential composition.  This means that $\rrfx$
-must be verified during pomset construction, rather than post-hoc.  The
-following lemma gives a post-hoc verification technique for $\rrfx$, using
-program order ($\rpox$).\footnote{It is obvious how to enhance the semantics
-  of most operators to define $\rpox$.  When combining pomsets using the
-  conditional, the obvious definition of $\rpox$ may result in cycles, since
-  $\rpox$-ordered events may coalesce.  In this case we include a separate
-  pomset for each way of breaking these $\rpox$ cycles.}
-% \begin{example}
-%   The obvious definition of $\rpox$ may be cyclic, due to the conditional. 
-% \end{example}
+We first show that \EC-validity is unchanged if we assume
+${\rcb}\supseteq{\rfr}$:
 \begin{lemma}
-  Any $\aPS$ in the image of $\frf{\semrr{}}$ is top-level iff
-  for every $\bEv\xrfx\aEv$ either
-  \begin{itemize}
-  \item external fulfillment: $\bEv\le\aEv$ and if $\labeling(\cEv)$ blocks
-    $\labeling(\aEv)$ then either $\cEv\le\bEv$ or $\aEv\le\cEv$, or
-  \item internal fulfillment: $\bEv\xpox\aEv$ and $(\not\exists\cEv)$
-    $\labeling(\cEv) \rblocks \labeling(\aEv)$ and $\bEv\xpox\cEv\xpox\aEv$.
-  \end{itemize}
+  \label{lemma:fr2}
+  Suppose $\aEG$ is \EC-valid via $({\rco}, {\rrfx}, {\rcb})$.  Then there a
+  permutation ${\rcbp}$ of ${\rcb}$ such that $\aEG$ is \EC-valid via
+  $({\rco}, {\rrfx}, {\rcbp})$ and ${\rcbp}\supseteq{\rfr}$, where ${\rfr}$
+  is defined in \ref{arm-local}.
+
+  \vspace{-.5\baselineskip}
+  \begin{proof}
+    Suppose $\aEv\xfr\cEv$. By definition of $\rfr$, %$(\exists\bEv)$
+    $\aEv\xrfxinv\bEv\xco\cEv$, for some $\bEv$.
+    %Since ${\rco}\subseteq{\rcb}$, we know that $\bEv\xcb\cEv$.
+    We show that either (1) ${\aEv\xcb\cEv}$, or (2) ${\cEv\xcb\aEv}$ and we
+    can reverse the order in ${\rcbp}$ to satisfy the requirements.
+
+    If \ref{arm-rfe} applies to $\bEv\xrfx\aEv$, then $\aEv\xcb\cEv$, since
+    it cannot be that $\cEv\xco\bEv$.
+
+    Suppose \ref{arm-rfi} applies to $\bEv\xrfx\aEv$ and $\cEv$ is from a
+    different thread than $\aEv$.  Because it is a different thread, we cannot have
+    $\aEv\xlob\cEv$, and therefore we can choose ${\cEv\xcb\aEv}$ in $\rcbp$.
+
+    Suppose \ref{arm-rfi} applies to $\bEv\xrfx\aEv$ and $\cEv$ is from the same thread as $\aEv$.
+    Applying \ref{arm-local} to $\aEv\xfr\cEv$, it cannot be that $\cEv\xpoloc\aEv$.
+    %Using \ref{arm-local} and $\bEv\xco\cEv$, it cannot be that $\cEv\xpoloc\bEv$.
+    Since $\rpoloc$ is a per-thread-and-per-location total order, it must be that $\aEv\xpoloc\cEv$.
+    Applying \ref{arm-lob-poloc}, we cannot have $\aEv\xlob\cEv$, and therefore we can choose ${\cEv\xcb\aEv}$ in $\rcbp$.
+  \end{proof}
 \end{lemma}
-% \begin{enumerate}
-% \item[(\textsc{m}7d)]
-%   if $\labeling(\cEv) \rblocks \labeling(\aEv)$
-%   then $\bEv\xrfx\aEv$ implies $\cEv\le\bEv$.
-%   %   if $\bEv\xrfx\aEv$ and $\labeling(\cEv) \rblocks \labeling(\aEv)$ then not $\bEv\le\cEv\xpox\aEv$.
-% \end{enumerate}
+Here is a contradictory non-example illustrating the last case of the proof:
+\begin{align*}
+  \PW{x}{2}
+  \SEMI
+  \PR{x}{r}
+  \PAR
+  \PW{x}{1}
+  &&
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{c}{{c}:\DW{x}{2}}{}
+      \event{d}{{d}:\DW{x}{1}}{below right=of c}
+      \event{e}{{e}:\DR{x}{1}}{above right=of d}
+      \co{d}[pos=.2,left]{c}
+      \rfx{d}[pos=.2,right]{e}
+      \poloc[out=15,in=165]{c}{e}
+      \fr{e}[above]{c}
+    \end{tikzinline}}
+\end{align*}
 
@@
 
-The generality of \refthm{thm:ec} is not limited by the assumption that
-${\rcb_1}\supseteq{\rfr_1}$:
+We emphasize that \ref{pom-rf-le} does not hold for $\frf{\semrr{}}$:
+ $\bEv\xrfx\aEv$ may not imply $\bEv\le\aEv$ when $\bEv$ and $\aEv$ come
+from different sides of a sequential composition.  This means that $\rrfx$
+must be verified during pomset construction, rather than post-hoc.  The
+following lemma gives a post-hoc verification technique for $\rrfx$, using
+program order ($\rpox$) and phantom events ($\fmrg{}$).  The construction is
+discussed in \textsection\ref{sec:c11}.
 \begin{lemma}
-  \label{lemma:fr2}
-  Suppose $\aEG$ is \EC-valid via $({\rco}, {\rrfx}, {\rcb})$.  Then there a
-  permutation ${\rcbp}$ of ${\rcb}$ such that $\aEG$ is \EC-valid via
-  $({\rco}, {\rrfx}, {\rcbp})$ and ${\rcbp}\supseteq{\rfr}$, where ${\rfr}$
-  is defined in \ref{arm-local}.
-
-  \vspace{-.5\baselineskip}
-  \begin{proof}
-    We show that any ${\rcb}$ order that contradicts ${\rfr}$ is incidental.
-
-    By definition of $\rfr$, %$(\exists\bEv)$
-    $\aEv\xrfxinv\bEv\xco\cEv$, for some $\bEv$.
-    Since ${\rcb}\supseteq{\rco}$, we know that $\bEv\xco\cEv$.
-
-    If \ref{arm-rfe} applies to $\bEv\xrfx\aEv$, then $\aEv\xcb\cEv$, since
-    it cannot be that $\cEv\xco\bEv$.
-
-    Suppose \ref{arm-rfi} applies to $\bEv\xrfx\aEv$ and $\cEv$ is from a
-    different thread.  Because it is a different thread, we cannot have
-    $\aEv\xlob\cEv$, and thus the order in $\rcb$ is incidental.
-
-    Suppose \ref{arm-rfi} applies to $\bEv\xrfx\aEv$ and $\cEv$ is from the
-    same thread.  Since $\cEv\xco\bEv$, it cannot be that $\cEv\xpoloc\bEv$,
-    using \ref{arm-local}.  It also cannot be that
-    $\bEv\xpoloc\cEv\xpoloc\aEv$.  It must be that $\aEv\xpoloc\cEv$.  By
-    \ref{arm-lob-poloc}, we cannot have $\aEv\xlob\cEv$, and thus the order
-    in $\rcb$ is incidental.
-  \end{proof}
+  Any $\aPS$ in the image of $\frf{\semrr{}}$ is top-level iff
+  for every $\bEv\xrfx\aEv$ either
+  \begin{itemize}
+  \item external fulfillment:
+    $\bEv\le\aEv$ and if $\labeling(\cEv) \rblocks\labeling(\aEv)$ then either $\cEv\le\bEv$ or $\aEv\le\cEv$, or
+  \item internal fulfillment:
+    $(\exists\bEv'\mid\fmrg{\bEv'}=\bEv)$
+    $(\exists\aEv'\mid\fmrg{\aEv'}=\aEv)$
+\todo{The following is sketchy.}
+    $\bEv\xpox\aEv$ and $(\not\exists\cEv)$
+    $\labeling(\cEv) \rblocks \labeling(\aEv)$ and $\bEv\xpox\cEv\xpox\aEv$.
+  \end{itemize}
 \end{lemma}
+% \begin{enumerate}
+% \item[(\textsc{m}7d)]
+%   if $\labeling(\cEv) \rblocks \labeling(\aEv)$
+%   then $\bEv\xrfx\aEv$ implies $\cEv\le\bEv$.
+%   %   if $\bEv\xrfx\aEv$ and $\labeling(\cEv) \rblocks \labeling(\aEv)$ then not $\bEv\le\cEv\xpox\aEv$.
+% \end{enumerate}
 

### 2021-06-30 paper/mca.tex
put invariants into PwT-C11
another stab at c11

@@
 \section{\PwTmcaTITLE{}: Pomsets with Predicate Transformers for MCA}
-
+\label{sec:mca}
 We define two models for \mca{} architectures: \PwTmca{1} and \PwTmca{2}.
@@
 \subsection{First Semantics}
+\label{sec:mca1}
 
@@
   \end{enumerate}
-  A \PwTmca{} is \emph{complete} if it is a complete \PwT{} (\refdef{def:pomset}) and
+
+  A \PwTmca{} is \emph{complete} if %it is a complete \PwT{} (\refdef{def:pomset}) and
+  \begin{multicols}{2}
     \begin{enumerate}[,label=(\textsc{c}\arabic*),ref=\textsc{c}\arabic*]
+    \item[\eqref{top-kappa}]
+      \eqref{top-term}\;
+      as in \refdef{def:pomset},
       \setcounter{enumi}{\value{Brf}}
@@
     \end{enumerate}
+  \end{multicols}
 \end{definition}
@@
 \subsection{Second Semantics}
+\label{sec:mca2}
 
@@
 \eqref{seq-rf-le-rf}.
-This change both allows and requires us to weaken the definition of
-\emph{delays} to drop write-to-read order from $\eqreorderco$.
+% This change both allows and requires us to weaken the definition of
+% \emph{delays} to drop write-to-read order from $\eqreorderco$.
 
 With the weakening of \ref{seq-rf-le-rf}, we must be careful not to allow
-spurious pairs to be added to the $\rrfx$ relation.  The use of
-$\rextendsdef{}{}$ in \ref{if-rf-extends} does this, ensuring that new
-$\rrfx$ is not introduced between events in $\aEvs_1\cap\aEvs_2$ when coalescing.
-This is necessary to ensure that
+spurious pairs to be added to the $\rrfx$ relation.  Thus we add
+\ref{par-rf-extends},
+\ref{seq-rf-extends}, and
+\ref{if-rf-extends}.
+For example, \ref{if-rf-extends} ensure that
 \begin{math}
@@
 \end{definition}
-Derive $\rdelayspdef$ from $\rdelaysdef$ by replacing $\eqreorderco$ with
-\begin{math}
-  {\reorderlws}
-  =
-  \{(\DW{\aLoc}{}, \DW{\aLoc}{}),\;(\DR{\aLoc}{}, \DW{\aLoc}{})\}.
-\end{math}
-The acronym $\textsf{lws}$ is adopted from \armeight.  It stands for
-\emph{Local Write Successor}.
+% Derive $\rdelayspdef$ from $\rdelaysdef$ by replacing $\eqreorderco$ with
+% \begin{math}
+%   {\reorderlws}
+%   =
+%   \{(\DW{\aLoc}{}, \DW{\aLoc}{}),\;(\DR{\aLoc}{}, \DW{\aLoc}{})\}.
+% \end{math}
+% The acronym $\textsf{lws}$ is adopted from \armeight.  It stands for
+% \emph{Local Write Successor}.
 

### 2021-07-02 paper/intro.tex
Reorg
first draft of PwT-C11
finished with c11

@@
 \item We define a denotational model for calculating dependencies.
-\item We apply the model to the \PwP{} semantics of
-  \cite{DBLP:journals/pacmpl/JagadeesanJR20} and to the \MRD{} semantics of
-  \cite{DBLP:conf/esop/PaviottiCPWOB20}.
+\item We apply the model to the \cXI{} semantics of
+  \cite{DBLP:conf/esop/PaviottiCPWOB20} (\PwTc) and to the \mca{} semantics of
+  \cite{DBLP:journals/pacmpl/JagadeesanJR20} (\PwTmca{}).
 \item We provide a tool to execute litmus tests for both models.
@@
 composition and/or precise calculation of semantic dependencies.
+\citet{DBLP:journals/pacmpl/JagadeesanJR20} give a denotational semantics,
+using prefixing rather than sequential compositions.
 \citet{DBLP:conf/esop/PaviottiCPWOB20} give a denotational semantics,
@@
 the semantics of sequential composition in continuation passing style,
-whereas we give it in direct style.
-%
+whereas we give it in direct style.  This paper provides a general technique
+for computing sequential dependencies and applies it to these two approaches.
+We provide a detailed comparison with
+\cite{DBLP:journals/pacmpl/JagadeesanJR20} in \textsection\ref{sec:diff}.
+
 \citet{DBLP:journals/corr/abs-1804-04214} define a semantics
@@
   DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}.
-
-\citet{DBLP:journals/pacmpl/JagadeesanJR20} note that the speculative models
-listed above all, including \cite{DBLP:conf/popl/KangHLVD17}, fail to
-validate compositional reasoning of temporal properties---see their examples
-OOTA4, OOTA5, and \citep[Fig.~8]{DBLP:journals/toplas/Lochbihler13}).  The
-difference with our model can be understood in terms of the valid program
-transformations.  The speculative models allow reads to be introduced, with
-subsequent case analysis on the value read---effectively, this can turn one
-read into two, with different conditional branches taken for the two copies
-of the read.  Our model invalidates this transformation.  In return, our
-model enjoys compositionality for temporal safety properties.
-
-We provide a detailed comparison with
-\cite{DBLP:journals/pacmpl/JagadeesanJR20} in \textsection\ref{sec:diff}.
+We provide a detailed comparison with these approaches in \textsection\ref{sec:promising}.
 

### 2021-07-02 paper/mca1.tex
finished with c11

--- 2021-06-30 paper/mca.tex
@@
-\section{\PwTmcaTITLE{}: Pomsets with Predicate Transformers for MCA}
-\label{sec:mca}
+\section{\PwTmcaTITLE{1}: Pomsets with Predicate Transformers for MCA (Part 1)}
+\label{sec:mca1}
 We define two models for \mca{} architectures: \PwTmca{1} and \PwTmca{2}.
 We drop the subscript when referring to both models.
-
-\subsection{First Semantics}
-\label{sec:mca1}
 
@@
 
-The semantic rules are given in \reffig{fig:mca1}.  Let
+The semantic rules are given in \reffig{fig:mca1}.
+We write $\semmca{1}{}$ for the semantic function.
+Let
 \begin{math}
-  \sembase{\aCmd_1 \LPAR[\bThrd] \aCmd_2} = \sLPAR{\sembase{\aCmd_1}}{\sembase{\aCmd_2}}.
+  \semmca{1}{\aCmd_1 \LPAR[\bThrd] \aCmd_2} = \sLPAR{\semmca{1}{\aCmd_1}}{\semmca{1}{\aCmd_2}}.
 \end{math}
 
-From the data model, we require an additional binary relation 
-${\rdelays}\subseteq(\Act\times\Act)$.  For the actions in this paper, we
-define this as follows.
+% From the data model, we require an additional binary relation 
+% ${\rdelays}\subseteq(\Act\times\Act)$.  For the actions in this paper, we
+% define this as follows.
 
-\begin{scope}
-  Let ${\eqreorderco}$ capture write-write, read-write coherence:
-  \begin{math}
-    {\eqreorderco}
-    =
-    \{(\DW{\aLoc}{}\Cb \DW{\aLoc}{})\Cc(\DR{\aLoc}{}\Cb \DW{\aLoc}{})\Cc(\DW{\aLoc}{}\Cb \DR{\aLoc}{})\}
-  \end{math}.
-
-  Let ${\reorderra}$ capture conflict due to synchronization:
-  \begin{math}
-    {\reorderra}
-    =
-    \{(\aAct\Cb             \DW[\gemode\mREL]{}{}     )\Cc
-    (\aAct\Cb               \DF{\gemode\fREL}        )\Cc
-    (\DR{}{}\Cb             \DF{\gemode\fACQ}        )\Cc
-    % (\DR{\aLoc}{}\Cb        \DR[\gemode\mACQ]{\aLoc}{})\Cc
-    (\DR[\gemode\mACQ]{}{}\Cb\aAct                    )\Cc
-    (\DF{\gemode\fACQ}\Cb   \aAct                    )\Cc
-    (\DF{\gemode\fREL}\Cb   \DW{}{}                  )\Cc
-    (\DW[\gemode\mREL]{\aLoc}{}\Cb\DW{\aLoc}{})\}
-  \end{math}.
-
-  Let ${\eqreordersc}$ capture conflict due to $\mSC$ access:
-  \begin{math}
-    {\eqreordersc}
-    =
-    % \{(\DX[\mSC]{}{}\Cb\DX[\mSC]{}{})\}
-    \{(\DW[\mSC]{}{}\Cb \DW[\mSC]{}{})\Cc(\DR[\mSC]{}{}\Cb \DW[\mSC]{}{}) \Cc(\DW[\mSC]{}{}\Cb \DR[\mSC]{}{})\Cc(\DR[\mSC]{}{}\Cb \DR[\mSC]{}{})\}
-  \end{math}.
-
-  We say $\aAct \rdelaysdef \bAct$ if $\aAct\eqreorderco\bAct$ or $\aAct\reorderra\bAct$ or $\aAct\eqreordersc\bAct$.
-\end{scope}
-
-\subsection{Second Semantics}
-\label{sec:mca2}
-
-In the second semantics, we weaken the relationship between $\rrfx$
-and $\le$ in \ref{pom-rf-le}.  Rather than ensuring that there is no
-\emph{global} blocker for a sequentially fulfilled read \eqref{pom-rf-le}, we
-require only that there is no \emph{thread-local} blocker
-\eqref{seq-rf-le-rf}.
-% This change both allows and requires us to weaken the definition of
-% \emph{delays} to drop write-to-read order from $\eqreorderco$.
-
-With the weakening of \ref{seq-rf-le-rf}, we must be careful not to allow
-spurious pairs to be added to the $\rrfx$ relation.  Thus we add
-\ref{par-rf-extends},
-\ref{seq-rf-extends}, and
-\ref{if-rf-extends}.
-For example, \ref{if-rf-extends} ensure that
-\begin{math}
-  \frf{\semrr{\IF{b}\THEN\PR{x}{r}\PAR\PW{x}{1}\ELSE\PR{x}{r}\SEMI\PW{x}{1}\FI}}
-\end{math}
-does not include 
-\begin{math}
-  \smash{\hbox{\begin{tikzinlinesmall}[node distance=1.5em]
-        \event{a}{\DR{x}{1}}{}
-        \event{b}{\DW{x}{1}}{right=of a}
-        \rfint[out=165,in=15]{b}{a}
-        \wki{a}{b}
-      \end{tikzinlinesmall}}}
-\end{math}, taking $\rrfx$ from the left and $\le$ from the right.
-
-\begin{definition}
-  \label{def:pwt:mca2}
-  A \PwTmca{2} is a \PwT{} (\refdef{def:pomset}) equipped
-  with a relation $\rrfx$ such that 
-  \begin{enumerate}[,label=(\textsc{m}\arabic*),ref=\textsc{m}\arabic*]
-    \setcounter{enumi}{\value{Brf}}
-  \item \label{pom-rf'} ${\rrfx} \subseteq \aEvs\times\aEvs$
-    is an injective relation capturing \emph{reads-from}, such that
-    % \end{enumerate}
-    % A pomset is a \emph{candidate} if there is an injective relation
-    % ${\rrfx} : \aEvs\times\aEvs$, capturing \emph{reads-from}, such that
-    \begin{multicols}{2}    
-      \begin{enumerate}
-        % \begin{enumerate}[,label=(\textsc{i}\arabic*),ref=\textsc{i}\arabic*]
-        % \item \label{rf-injective}
-        %   if $\bEv\xrfx\aEv$ and $\cEv\xrfx\aEv$ then $\bEv=\cEv$, that is,
-        %   ${\rrfx}$ is injective,
-      \item[\eqref{pom-rf-match}]
-        \eqref{pom-rf-block}\;
-        as in \refdef{def:pwt:mca1},
-        \setcounter{enumii}{2}
-      \item \label{pom-rf-le'} 
-        if $\bEv\xrfx\aEv$ then either $\bEv\le\aEv$ or $\aEv\le\bEv$.
-      \end{enumerate}
-    \end{multicols}
-  \end{enumerate}
-\end{definition}
-% Derive $\rdelayspdef$ from $\rdelaysdef$ by replacing $\eqreorderco$ with
-% \begin{math}
-%   {\reorderlws}
-%   =
-%   \{(\DW{\aLoc}{}, \DW{\aLoc}{}),\;(\DR{\aLoc}{}, \DW{\aLoc}{})\}.
-% \end{math}
-% The acronym $\textsf{lws}$ is adopted from \armeight.  It stands for
-% \emph{Local Write Successor}.
-
-The semantic rules are given in \reffig{fig:mca2}.
+% \begin{scope}
+% \end{scope}

### 2021-07-02 paper/oopsla.tex
cleanup
another stab at c11
finished with c11

@@
 
-[Todo.]
+\citet{DBLP:journals/pacmpl/JagadeesanJR20} note that the speculative models
+listed above all, including \cite{DBLP:conf/popl/KangHLVD17}, fail to
+validate compositional reasoning of temporal properties---see their examples
+OOTA4, OOTA5, and \citep[Fig.~8]{DBLP:journals/toplas/Lochbihler13}).  The
+difference with our model can be understood in terms of the valid program
+transformations.  The speculative models allow reads to be introduced, with
+subsequent case analysis on the value read---effectively, this can turn one
+read into two, with different conditional branches taken for the two copies
+of the read.  Our model invalidates this transformation.  In return, our
+model enjoys compositionality for temporal safety properties.
+
+\todo{Write this.}
 
@@
 Skolemization is motivated by disjunction closure for predicate transformers,
-which do not appear in \jjr{}.  In \reffig{fig:sem}, 
+which do not appear in \jjr{}.  In \reffig{fig:seq}, 
 we gave the semantics of read for nonempty pomsets as:

### 2021-07-02 paper/drf.tex
finished with c11
changed example in C11

@@
 \label{sec:sc}
+
+Idea: merging reads can only make a difference is if there is a race.
+Let $\sempomca{i}{}$ be the semantic function generated by applying the
+construction of \reflem{lem:po} to $\semmca{i}{}$.
 
@@
 ``$\aLoc_1\GETS\aVal_1\SEMI \cdots \aLoc_n\GETS\aVal_n\SEMI\aCmd$'' where
-every location mentioned in $\aCmd$ is some $\aLoc_i$.
-
-We make two further restrictions to simplify the exposition.  To simplify the
-definition of \emph{happens-before}, we ban fences and \RMWs.  To simplify
-the proof, we assume there are no local declarations of the form
-$(\VAR\aLoc\SEMI \aCmd)$.
+every location mentioned in $\aCmd$ is some $\aLoc_i$.  To simplify the
+definition of \emph{happens-before}, we ban fences and \RMWs.
+% To simplify
+% the proof, we assume there are no local declarations of the form
+% $(\VAR\aLoc\SEMI \aCmd)$.
 
@@
     \PAR
-    (x\GETS 2 \SEMI \FENCE^\mSC \SEMI \IF{y^\mRA}\THEN r\GETS x \SEMI s\GETS x\FI)
+    (x\GETS 2 \SEMI \FENCE{\mSC} \SEMI \IF{y^\mRA}\THEN r\GETS x \SEMI s\GETS x\FI)
     \\[-2ex]

### 2021-07-02 paper/drf.tex
messing with c11 to make drf easier

@@
-\section{Local Data Race Freedom and Sequential Consistency}
+% \section{Local Data Race Freedom and Sequential Consistency}
+\section{LDRF-SC for \PwTmcaTITLE{}}
 \label{sec:sc}
 
-Idea: merging reads can only make a difference is if there is a race.
-Let $\sempomca{i}{}$ be the semantic function generated by applying the
-construction of \reflem{lem:po} to $\semmca{i}{}$.
+\cbstart
+In this appendix, we establish a \drfsc{} for \PwTmca{2}.  We prove an
+\emph{external} result, where the notion of \emph{data-race} is independent
+of the semantics itself.  Since every \PwTmca{2} is also a \PwTmca{1}, the
+result also applies there.  Our result is also \emph{local}.  Using
+\citeauthor{Dolan:2018:BDR:3192366.3192421}'s
+[\citeyear{Dolan:2018:BDR:3192366.3192421}] notion of \emph{Local Data Race
+  Freedom (LDRF)}.
 
-We adapt \citeauthor{Dolan:2018:BDR:3192366.3192421}'s
-[\citeyear{Dolan:2018:BDR:3192366.3192421}] notion of \emph{Local Data Race
-  Freedom (LDRF)} to our setting.
+We do not address \PwTc{}.  The existing \drfsc{} result for \cXI{}
+\cite{DBLP:phd/ethos/Batty15} does not rely on dependencies and thus applies
+to \PwTc{}.  However, \citeauthor{DBLP:phd/ethos/Batty15}'s result is
+\emph{internal}, where data-races are defined using the semantics of \cXI{}
+itself.  Using the notion of dependency defined here, it should be possible
+to prove an stronger external result for \cXI{}---we leave this as future
+work.
+
+\citet{DBLP:journals/pacmpl/JagadeesanJR20} prove \ldrfsc{} for Pomsets with
+Preconditions (\PwP{}).  \PwTmca{} generalizes \PwP{} to account for
+sequential composition.  Most of the machinery of \ldrfsc{}, however, has
+little to do with sequential semantics.  Thus, we have borrowed heavily from
+the text of \cite{DBLP:journals/pacmpl/JagadeesanJR20}; indeed, we have
+copied directly from the \LaTeX{} source, which is publicly available.  We
+indicate substantial changes or additions using a change-bar on the right.
+
+There are two main changes:
+\begin{itemize}
+\item \PwP{} assumes \ref{pom-rf-le} ($\rrfx$ implies $\le$)
+  and thus is similar to \PwTmca{1}.  \PwTmca{2} is a weaker model that is
+  new to this paper.  Hence we cannot use
+  \citeauthor{DBLP:journals/pacmpl/JagadeesanJR20}'s result directly.
+\item \citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} did not provide an
+  accurate account of program order for merged actions.  We correct this
+  deficiency.  Let $\sempomca{i}{}$ be the semantic function generated by
+  applying the construction of \reflem{lem:po} to $\semmca{i}{}$.  The proof
+  concerns only \emph{complete} pomsets.
+\end{itemize}
+
+\cbend
 
@@
 definition of \emph{happens-before}, we ban fences and \RMWs.
-% To simplify
-% the proof, we assume there are no local declarations of the form
-% $(\VAR\aLoc\SEMI \aCmd)$.
 
@@
 \emph{pomset} order $(\le)$. %, and thus is stable with respect to augmentation.
-In \ref{SB}, for example, $(\DR{x}{0})$ has an $x$-race with $(\DW{x}{1})$,
-but not $(\DW{x}{0})$, which is $\rpox$-before it.
 
-It is obvious how to enhance the semantics of prefixing and most other
-operators to define $\rpox$.  When combining pomsets using the conditional,
-the obvious definition may result in cycles, since $\rpox$-ordered reads may
-coalesce---see the discussion of \ref{CA} in \textsection\ref{sec:refine}.  In
-this case we include a separate pomset for each way of breaking these cycles.
-
-Because we ignore the features of
-\textsection\ref{sec:variants}, we can adopt the simplest definition of
+Because we ban fences and \RMWs, we can adopt the simplest definition of
 \emph{synchronizes\hyp{}with}~($\rsw$): Let $\bEv\xsw\aEv$ exactly when
@@
 
-Let ${\rhb}=({\rpox}\cup{\rsw})^+$ be the \emph{happens-before} relation.  In
-\ref{Pub1}, for example, $(\DW{x}{1})$ happens-before $(\DR{x}{0})$, but this
-fails if either $\mRA$ access is relaxed.
+Let ${\rhb}=({\rpox}\cup{\rsw})^+$ be the \emph{happens-before} relation.
 
 Let $L\subseteq\Loc$ be a set of locations.  We say that $\bEv$ \emph{has an
-  $L$-race with} $\aEv$ (notation $\bEv\lrace{L}\aEv$) when at least one is
-relaxed, they \emph{conflict}
-(Def.~\ref{def:rf}) at
-some location in $L$, and they are unordered by $\rhb$: neither $\bEv\xhb\aEv$ nor
+  $L$-race with} $\aEv$ (notation $\bEv\lrace{L}\aEv$) when
+\begin{enumerate*}
+\item at least one is relaxed, 
+\item at least one is a write,
+\item they access the same location in $L$, and
+\item they are unordered by $\rhb$: neither $\bEv\xhb\aEv$ nor
   $\aEv\xhb\bEv$.
+\end{enumerate*}
 
@@
   read of $x$, due to the conditional.
+
+  In the above argument, it is unimportant whether $\aEv$ reads-from an
+  internal or an external write; thus the argument applies to
+  \PwTmca{2} and \PwTmca{1} as it does for \PwTmca{1}.
 

### 2021-07-04 paper/drf.tex
a stab at DRF

@@
 
-\cbstart
+\begin{changed}
   In this appendix, we establish a \drfsc{} for \PwTmca{2}.  We prove an
@@
   We do not address \PwTc{}.  The existing \drfsc{} result for \cXI{}
-\cite{DBLP:phd/ethos/Batty15} does not rely on dependencies and thus applies
-to \PwTc{}.  However, \citeauthor{DBLP:phd/ethos/Batty15}'s result is
-\emph{internal}, where data-races are defined using the semantics of \cXI{}
-itself.  Using the notion of dependency defined here, it should be possible
-to prove an stronger external result for \cXI{}---we leave this as future
-work.
+  \cite{DBLP:phd/ethos/Batty15} does not rely on dependencies and thus
+  applies to \PwTc{}.  However, \citeauthor{DBLP:phd/ethos/Batty15}'s result
+  is \emph{internal}, where data-races are defined using the semantics of
+  \cXI{} itself.  Using the notion of dependency defined here, it should be
+  possible to prove an stronger external result for \cXI{}---we leave this as
+  future work.
 
-\citet{DBLP:journals/pacmpl/JagadeesanJR20} prove \ldrfsc{} for Pomsets with
-Preconditions (\PwP{}).  \PwTmca{} generalizes \PwP{} to account for
+  \citet{DBLP:journals/pacmpl/JagadeesanJR20} prove \ldrfsc{} for Pomsets
+  with Preconditions (\PwP{}).  \PwTmca{} generalizes \PwP{} to account for
   sequential composition.  Most of the machinery of \ldrfsc{}, however, has
-little to do with sequential semantics.  Thus, we have borrowed heavily from
-the text of \cite{DBLP:journals/pacmpl/JagadeesanJR20}; indeed, we have
-copied directly from the \LaTeX{} source, which is publicly available.  We
-indicate substantial changes or additions using a change-bar on the right.
+  little to do with sequential semantics.  Thus, we have borrowed heavily
+  from the text of \cite{DBLP:journals/pacmpl/JagadeesanJR20}; indeed, we
+  have copied directly from the \LaTeX{} source, which is publicly available.
+  We indicate substantial changes or additions using a change-bar on the
+  right.
 
-There are two main changes:
+  There are several changes:
   \begin{itemize}
-\item \PwP{} assumes \ref{pom-rf-le} ($\rrfx$ implies $\le$)
-  and thus is similar to \PwTmca{1}.  \PwTmca{2} is a weaker model that is
-  new to this paper.  Hence we cannot use
-  \citeauthor{DBLP:journals/pacmpl/JagadeesanJR20}'s result directly.
-\item \citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} did not provide an
-  accurate account of program order for merged actions.  We correct this
-  deficiency.  Let $\sempomca{i}{}$ be the semantic function generated by
-  applying the construction of \reflem{lem:po} to $\semmca{i}{}$.  The proof
-  concerns only \emph{complete} pomsets.
+  \item \PwP{} imposes several conditions that we have dropped:
+    \emph{consistency}, \emph{causal strengthening}, \emph{downset closure}
+    (see \textsection\ref{sec:diff}).
+  \item \PwP{} allows preconditions that are stronger than the weakest precondition.
+  \item \PwP{} imposes \ref{pom-rf-le} ($\rrfx$ implies $\le$) and thus is
+    similar to \PwTmca{1}.  \PwTmca{2} is a weaker model that is new to this
+    paper.  %Hence we cannot use the \PwP{} result directly.
+  \item \PwP{} did not provide an accurate account of program order for
+    merged actions.  We use \reflem{lem:po} to correct this deficiency.
   \end{itemize}
-
-\cbend
+  The first two items require us to define $\semmin{}$ differently, below.
+\end{changed}
 
@@
 \bookmark{Definitions}
+\begin{changed}
+  \paragraph{Program Order}
+  Let $\sempomca{2}{}$ be defined by applying the construction of
+  \reflem{lem:po} to $\semmca{2}{}$.  We consider only \emph{complete}
+  pomsets.  For these, we derive program order on compound events as follows.
+  By \reflem{lem:compound:phantom:exists}, if there is a compound event
+  $\aEv$, then there is a phantom event $\cEv\in\fmrginv{\aEv}$ such that
+  $\labelingForm(\cEv)$ is a tautology.  If there is exactly one tautology,
+  we identify $\aEv$ with $\cEv$ in program order.  If there is more than one
+  tautology, \reflem{lem:sequential:simple}, below, shows that it suffices to
+  pick an arbitrary one---we identify $\aEv$ with the $\cEv\in\fmrginv{\aEv}$
+  that is minimal in program order.
+\end{changed}
+For example, consider \jmm{} causality test case 2, with an added write to $z$:
+\begin{gather*}
+  \tag{$\ddagger\ddagger$}\label{eq6}
+  \begin{gathered}
+    \PR{x}{r}\SEMI
+    \PW{z}{1}\SEMI
+    \PR{x}{s}\SEMI
+    \IF{r{=}s}\THEN \PW{y}{1}\FI
+    \PAR
+    \PW{x}{y}
+    \\
+    \hbox{\begin{tikzinline}[node distance=1em and 1.5em]
+        \phevent{p1}{\DR{x}{1}}{}
+        \event{z}{\DW{z}{1}}{right=of p1}
+        \phevent{p2}{\DR{x}{0}}{right=of z}
+        \event{a3}{\DW{y}{1}}{right=of p2}
+        \event{b1}{\DR{y}{1}}{right=3em of a3}
+        \event{b2}{\DW{x}{1}}{right=of b1}
+        \rf{a3}{b1}
+        \po{b1}{b2}
+        \pox[out=-15,in=-170]{p1}{z}
+        \pox[out=-15,in=-170]{z}{p2}
+        \pox[out=-15,in=-170]{p2}{a3}
+        \pox[out=-15,in=-170]{b1}{b2}
+        \event{a1}{\DR{x}{1}}{above=of z}
+        \mrg{p1}[pos=.3]{a1}
+        \mrg{p2}[right]{a1}      
+        \rf{b2}{a1}
+        \pox{a1}{z}
+      \end{tikzinline}}
+  \end{gathered}
+\end{gather*}
+% For TC2, we have: 
+% \begin{gather*}
+%   \PR{x}{r}\SEMI
+%   \PR{x}{s}\SEMI
+%   \IF{r{=}s}\THEN \PW{y}{1}\FI
+%   \PAR
+%   \PW{x}{y}
+%   \\
+%   \hbox{\begin{tikzinline}[node distance=1.5em]
+%       \event{a1}{\DR{x}{1}}{}
+%       % \event{a2}{\DR{x}{1}}{right=of a1}
+%       \event{a3}{\DW{y}{1}}{right=of a1}
+%       % \po{a1}{a3}
+%       % \po[out=-20,in=-160]{a1}{a3}
+%       \event{b1}{\DR{y}{1}}{right=3em of a3}
+%       \event{b2}{\DW{x}{1}}{right=of b1}
+%       \rf{a3}{b1}
+%       \po{b1}{b2}
+%       % \rf[out=169,in=11]{b2}{a2}
+%       \rf[out=169,in=11]{b2}{a1}
+%       \phevent{p2}{\DR{x}{1}}{below=of a1}
+%       \phevent{p1}{\DR{x}{1}}{left=of p2}
+%       \mrg{p1}[pos=.3]{a1}
+%       \mrg{p2}[right]{a1}      
+%       \pox[out=-15,in=-170]{b1}[below]{b2}
+%       \pox[out=-15,in=-170]{p1}[below]{p2}
+%       \pox{p2}[below]{a3}
+%     \end{tikzinline}}
+%   \\
+%   \hbox{\begin{tikzinline}[node distance=1.5em]
+%       \event{a1}{\DR{x}{0}}{}
+%       % \event{a2}{\DR{x}{1}}{right=of a1}
+%       \event{a3}{\DW{y}{1}}{right=of a1}
+%       % \po{a1}{a3}
+%       % \po[out=-20,in=-160]{a1}{a3}
+%       \event{b1}{\DR{y}{1}}{right=3em of a3}
+%       \event{b2}{\DW{x}{1}}{right=of b1}
+%       \rf{a3}{b1}
+%       \po{b1}{b2}
+%       % \rf[out=169,in=11]{b2}{a2}
+%       %\rf[out=169,in=11]{b2}{a1}
+%       \phevent{p2}{\DR{x}{0}}{below=of a1}
+%       \phevent{p1}{\DR{x}{0}}{left=of p2}
+%       \mrg{p1}[pos=.3]{a1}
+%       \mrg{p2}[right]{a1}      
+%       \pox[out=-15,in=-170]{b1}[below]{b2}
+%       \pox[out=-15,in=-170]{p1}[below]{p2}
+%       \pox{p2}[below]{a3}
+%     \end{tikzinline}}
+% \end{gather*}
+
 \paragraph{Data Race}
@@
 
+\begin{changed}
+  Note that ${\rhb}$ is a preorder in general. It is a partial order on
+  \emph{simple} events (\refdef{def:po}).
+\end{changed}
+
 Let $L\subseteq\Loc$ be a set of locations.  We say that $\bEv$ \emph{has an
@@
 
+\begin{changed}
   \paragraph{Generators}
-We say that $\aPS'$ \emph{generates} $\aPS$ if either
-$\aPS$ augments $\aPS'$ or $\aPS$ implies $\aPS'$.  For example, the
-unordered pomset $(\DR{x}{1})$ $(\DW{y}{1})$ generates the ordered pomset
-$(\DR{x}{1})\xpo(\aReg=1\mid\DW{y}{1})$.
+  We say that $\aPS'\in\downclose{\aPSS}$ if there is some $\aPS\in\aPSS$ such that
+  $\aPS$ is \emph{complete} (\refdef{def:pwt:mca:complete}) and $\aPS'$ is a
+  \emph{downset} of $\aPS$ (\refdef{def:downset}).
 
-We say that $\aPS$ is a \emph{generation-minimal} in $\aPSS$ if $\aPS\in\aPSS$ and
-there is no $\aPS\neq\aPS'\in\aPSS$ that generates $\aPS$.
+  % We say that $\aPS'$ \emph{generates} $\aPS$ if
+  % $\aPS$ augments $\aPS'$ (\refdef{def:augment}).
 
-Let $\semmin{\aCmd}=\{\aPS\in\sem{\aCmd} \mid \aPS$ is \emph{top-level}
-(Def.~\ref{def:top}) and generation-minimal in $\sem{\aCmd}\}$.
+  Let $\aPS$ be \emph{augmentation-minimal} in $\aPSS$ if $\aPS\in\aPSS$ and
+  there is no $\aPS{\neq}\aPS'{\in}\aPSS$ such that $\aPS$ augments $\aPS'$.
+
+  Let $\semmin{\aCmd}=\{\aPS\in\downclose{}\sempomca{2}{\aCmd} \mid \aPS$ is
+  augmentation-minimal in $\downclose{}\sempomca{2}{\aCmd}\}$.
+\end{changed}
 
@@
 We say that \emph{$\aPS'$ is $\aEv$-similar to $\aPS$} if they differ at most
-in (1) pomset order adjacent to $\aEv$ and (2) the value associated with
-event $\aEv$, if it is a
-read.  % We say they are \emph{similar} if they are $\aEv$-similar for some $\aEv$.
+in
+\begin{enumerate*}
+\item pomset order adjacent to $\aEv$, and
+\item the value associated with event $\aEv$, if it is a read.
+\end{enumerate*}
+% We say they are \emph{similar} if they are $\aEv$-similar for some $\aEv$.
 Formally: $\Event'=\Event$, $\labelingForm'=\labelingForm$,
@@
 We say that $\aPS$ is \emph{$L$-stable in $\aCmd$} if
-(1) $\aPS\in\semmin{\aCmd}$, 
-(2) $\aPS$ is $\rpox$-convex (nothing missing in program order),
-(3) there is no $\aCmd$-extension of $\aPS$ with a \emph{crossing} $L$-race:
-that is, there is no $\bEv\in\Event$, no $\aPS'$ $\aCmd$-extending
-$\aPS$, and no $\aEv\in\Event'\setminus\Event$ such that $\bEv\lrace{L}\aEv$.
+\begin{enumerate*}  
+\item $\aPS\in\semmin{\aCmd}$, 
+\item $\aPS$ is $\rpox$-convex (nothing missing in program order),
+\item there is no $\aCmd$-extension of $\aPS$ with a \emph{crossing}
+  $L$-race: that is, there is no $\bEv\in\Event$, no $\aPS'$
+  $\aCmd$-extending $\aPS$, and no $\aEv\in\Event'\setminus\Event$ such that
+  $\bEv\lrace{L}\aEv$.
+\end{enumerate*}
 The empty pomset is $L$-stable.
@@
 We say that $\aPS'$ is \emph{$L$-sequential after $\aPS$} if 
-$\aPS'$ is $\rpox$-convex and %
-$\pole{L}$ is acyclic in $\Event'\setminus\Event$.
+\begin{enumerate*}
+\item $\aPS'$ is $\rpox$-convex,
+\item $\pole{L}$ is acyclic in $\Event'\setminus\Event$.
+\end{enumerate*}
+
+\begin{changed}
+  \paragraph{Simplicity}
+  We say that $\aPS'$ is \emph{$L$-simple after $\aPS$} if all of the events
+  in $\Event'\setminus\Event$ that access locations in $L$ are \emph{simple}
+  (\refdef{def:po}).
+
+  \begin{lemma}
+    \label{lem:sequential:simple}
+    Suppose $\aPS'\in\semmin{\aCmd}$ and $\aPS$ is {$L$-sequential after
+      $\aPS$}.
+    Let $\aPS''$ be the restriction of $\aPS'$ that is {$L$-simple after
+      $\aPS$} (throwing out compound $L$-events after $\aPS$).
+    Then $\aPS''\in\semmin{\aCmd}$.
+    % projection of $\aPS$ is a $\aPS'\in\semmin{\aCmd}$ such that
+    % \begin{enumerate*}
+    % \item $\aPS'$ is $L$-sequential and $L$-simple after $\aPS$, and
+    % \item  $\aPS'$ is  derived from  $\aPS$  by removing  every compound  event
+    %   $\aEv$ that  accesses a location in  $L$, thus changing the remaining
+    %   events in $\fmrginvp{\aEv}$ from phantom events to real ones.
+    % \end{enumerate*}
+  \end{lemma}
+  As a negative example, note that \eqref{eq6} is not $L$-sequential---in
+  fact there is no execution of the program that results in the simple events
+  of \eqref{eq6}: without merging the reads, there would be a dependency
+  $\DRP{x}{1} \xpo \DWP{y}{1}$.  $L$-sequential executions of this code must read
+  $0$ for $x$:
+  \begin{gather*}
+    \PR{x}{r}\SEMI
+    \PW{z}{1}\SEMI
+    \PR{x}{s}\SEMI
+    \IF{r{=}s}\THEN \PW{y}{1}\FI
+    \PAR
+    \PW{x}{y}
+    \\
+    \hbox{\begin{tikzinline}[node distance=1em and 1.5em]
+        \phevent{p1}{\DR{x}{0}}{}
+        \event{z}{\DW{z}{1}}{right=of p1}
+        \phevent{p2}{\DR{x}{0}}{right=of z}
+        \event{a3}{\DW{y}{1}}{right=of p2}
+        \event{b1}{\DR{y}{1}}{right=3em of a3}
+        \event{b2}{\DW{x}{1}}{right=of b1}
+        \rf{a3}{b1}
+        \po{b1}{b2}
+        \pox[out=-15,in=-170]{p1}{z}
+        \pox[out=-15,in=-170]{z}{p2}
+        \pox[out=-15,in=-170]{p2}{a3}
+        \pox[out=-15,in=-170]{b1}{b2}
+        \event{a1}{\DR{x}{0}}{above=of z}
+        \mrg{p1}[pos=.3]{a1}
+        \mrg{p2}[right]{a1}      
+        \pox{a1}{z}
+      \end{tikzinline}}
+  \end{gather*}
+\end{changed}
 
@@
   $\Event''$ satisfies these criteria.
-  Then either (1) $\aPS''$ is $L$-sequential after $\aPS$ or (2) there is
-  some $\aCmd$-extension $\aPS'''$ of $\aPS'$ and some
-  $\aEv\in(\Event''\setminus\Event')$ such that (a) $\aPS'''$ is
-  $\aEv$-similar to $\aPS''$, (b) $\aPS'''$ is $L$-sequential after $\aPS$,
-  and (c) $\bEv\lrace{L}\aEv$, for some $\bEv\in(\Event''\setminus\Event)$.
+  Then either (1) $\aPS''$ is $L$-sequential
+  \begin{changed}
+    and $L$-simple
+  \end{changed}
+  after $\aPS$ or (2) there is some $\aCmd$-extension $\aPS'''$ of $\aPS'$
+  and some $\aEv\in(\Event''\setminus\Event')$ such that (a) $\aPS'''$ is
+  $\aEv$-similar to $\aPS''$, (b) $\aPS'''$ is $L$-sequential
+  \begin{changed}
+    and $L$-simple
+  \end{changed}
+  after $\aPS$, and (c) $\bEv\lrace{L}\aEv$, for some
+  $\bEv\in(\Event''\setminus\Event)$.
 \end{theorem}
@@
 \begin{proof}[Proof Sketch]
+  \begin{changed}
+    We show $L$-sequentiality.  $L$-simplicity then follows from
+    \reflem{lem:sequential:simple}.
+  \end{changed}
+
   In order to develop a technique to find $\aPS'''$ from $\aPS''$, we analyze
@@
   disjoint relations.  Let ${\rppo}=({\le_*}\cap{\rpox})$ denote
-  \emph{preserved} program order, as required by prefixing (Def.~\ref{def:prefix}).  The other two relations are cross-thread subsets of
-  $({\le_*}\setminus{\rpox})$, as required by fulfillment (Def.~\ref{def:rf}): $\rrfe$ orders writes before reads, satisfying fulfillment
-  requirement \ref{rf3}; $\rxw$ orders read and write accesses before writes,
-  satisfying requirement \ref{rf4}. ({Within a thread, \ref{rf3} and
-    \ref{rf4} follow from prefixing requirement \ref{5b}, which is included
-    in ${\rppo}$.})
+  \emph{preserved} program order, as required by sequential composition and
+  conditional.  The other two relations are cross-thread subsets of
+  $({\le_*}\setminus{\rpox})$: $\rrfe$ (reads-from-external) orders writes
+  before reads, satisfying \ref{par-rf-le}; $\rcae$
+  (coherence-after-external) orders read and write accesses before writes,
+  satisfying \ref{pom-rf-block}. (Within a thread, \ref{seq-le-mca2} induces
+  order that is included in ${\rppo}$.)
 
@@
   only $\rppo$).  We then prune reads which cannot be fulfilled, starting
-  with those that are minimally ordered.  This proof is simplified by
-  precluding local declarations.
+  with those that are minimally ordered.
+  % This proof is simplified by
+  % precluding local declarations.
 
@@
   these as $\aEv$.  If $\aEv$ is a write, we reverse the outgoing order in
-  $\rxw$; the ability to reverse this order witnesses the race.  If $\aEv$
+  $\rcae$; the ability to reverse this order witnesses the race.  If $\aEv$
   is a read, we switch its fulfilling write to a ``newer'' one, updating
-  $\rxw$; the ability to switch witnesses the race.  For
+  $\rcae$; the ability to switch witnesses the race.  For
   example, for $\aPS''$ on the left below, we choose the $\aPS'''$ on the
@@
 
+  \begin{changed}
     In the above argument, it is unimportant whether $\aEv$ reads-from an
-  internal or an external write; thus the argument applies to
-  \PwTmca{2} and \PwTmca{1} as it does for \PwTmca{1}.
+    internal or an external write; thus the argument applies to \PwTmca{2}
+    and \PwTmca{1} as it does for \PwTmca{1}.
+  \end{changed}
 
@@
 \end{proof}
-
-\bookmark{Mixed Races}
-The premises of the theorem allow us to avoid the complications caused by ``mixed races'' in
-\cite{DBLP:conf/ppopp/DongolJR19}.  In the left pomset below, $\aPS''$ is not
-an extension of $\aPS'$, since $\aPS'$ is not a downset of $\aPS''$.  
-When considering this pomset, we must perform the decomposition on the right.
-\begin{gather*}
-  (x\GETS 0 \SEMI   x^\mRA \GETS 1)
-  \PAR
-  (\aReg\GETS x^\mRA)
-  \\[-2ex]
-  \hbox{\begin{tikzinline}[node distance=1.5em and 2em]
-      \event{wx0}{\DW{x}{0}}{}
-      \event{wx1}{\DWRel{x}{1}}{right=of wx0}
-      \event{rx}{\DRAcq{x}{0}}{below=of wx0}
-      \rf{wx0}{rx}
-      \pox{wx0}{wx1}
-      \wk{rx}{wx1}
-      \node(ix)[left=of wx0]{};
-      \bgoval[yellow!50]{(ix)}{P}
-      \bgoval[pink!50]{(wx0)(wx1)}{P'\setminus P}
-      \bgovalright[green!10]{(rx)}{P''\setminus P'}
-    \end{tikzinline}}
-  \qquad
-  \qquad
-  \qquad
-  \hbox{\begin{tikzinline}[node distance=1.5em and 2em]
-      \event{wx0}{\DW{x}{0}}{}
-      \event{wx1}{\DWRel{x}{1}}{right=of wx0}
-      \event{rx}{\DRAcq{x}{0}}{below=of wx0}
-      \rf{wx0}{rx}
-      \pox{wx0}{wx1}
-      \wk{rx}{wx1}
-      \node(ix)[left=of wx0]{};
-      \bgoval[yellow!50]{(ix)}{P}
-      \bgoval[pink!50]{(wx0)(rx)}{P'\setminus P}
-      \bgoval[green!10]{(wx1)}{P''\setminus P'}
-    \end{tikzinline}}
-\end{gather*}
-This affects the inductive order in which we move across pomsets, but does
-not affect the set of pomsets that are considered.  This simplification is
-enabled by denotational reasoning.
-
-\bookmark{Comparison to Java}
-In our language, past races are always resolved at a stable point, as in
-\ref{Co3}.  As another example, consider the following, which is disallowed
-here, but allowed by Java \cite[Ex.~2]{Dolan:2018:BDR:3192366.3192421}.  We
-include an SC fence here to mimic the behavior of volatiles in the JMM.
-\begin{gather*}
-  \taglabel{past}
-  \begin{gathered}
-    (x\GETS 1 \SEMI   y^\mRA \GETS 1)
-    \PAR
-    (x\GETS 2 \SEMI \FENCE{\mSC} \SEMI \IF{y^\mRA}\THEN r\GETS x \SEMI s\GETS x\FI)
-    \\[-2ex]
-    \hbox{\begin{tikzinline}[node distance=1.2em]
-        \event{wx1}{\DW{x}{1}}{}
-        \event{wy1}{\DWRel{y}{1}}{right=of wx0}
-        \sync{wx1}{wy1}
-        \event{wx2}{\DW{x}{2}}{right=3em of wy1}
-        \event{f}{\DFS{\mSC}}{right=of wx2}
-        \sync{wx2}{f}
-        \event{ry1}{\DR[\mRA]{y}{1}}{right=of f}
-        \sync{f}{ry1}
-        \event{rx1}{\DR{x}{1}}{right=3em of ry1}
-        \event{rx2}{\DR{x}{2}}{right=of rx1}
-        \sync{ry1}{rx1}
-        \sync[out=15,in=165]{ry1}{rx2}
-        \rf[out=10,in=170]{wy1}{ry1}
-        \wk[out=10,in=170]{wx1}{wx2}
-        \wk[out=-170,in=-10]{rx1}{wx2}
-        \bgellipsesmaller[yellow!50]{(wy1)(f)}{}
-      \end{tikzinline}}
-  \end{gathered}
-\end{gather*}
-The highlighted events are $L$-stable.  The order from $(\DR{x}{1})$ to
-$(\DW{x}{2})$ is required by fulfillment, causing the cycle.  If the fence is
-removed, there would be no order from $(\DW{x}{2})$ to
-$(\DRAcq{y}{1})$, the highlighted events would no longer be $L$-stable, and
-the execution would be allowed.  This more relaxed notion of ``past'' is not
-expressible using \citeauthor{Dolan:2018:BDR:3192366.3192421}'s
-synchronization primitives.
-
-The notion of ``future'' is also richer here.  Consider \cite[Ex.~3]{Dolan:2018:BDR:3192366.3192421}:
-\begin{gather*}
-  \taglabel{future}
-  \begin{gathered}
-    (r\GETS 1 \SEMI \REF{r}\GETS 42\SEMI s\GETS \REF{r}\SEMI  x^\mRA\GETS r)
-    \PAR
-    (r\GETS x \SEMI \REF{r}\GETS 7)
-    \\[-1ex]
-    \hbox{\begin{tikzinline}[node distance=1.2em]
-        \event{a1}{\DW{\REF{1}}{42}}{}
-        \event{a2}{\DR{\REF{1}}{7}}{right=of a1}
-        \wk{a1}{a2}
-        \event{a4}{\DWRel{x}{1}}{right=of a2}
-        \sync{a2}{a4}
-        \event{b1}{\DR{x}{1}}{right=3em of a4}
-        \event{b2}{\DW{\REF{1}}{7}}{right=of b1}
-        \po{b1}{b2}
-        \rf{a4}{b1}
-        \wk[out=170,in=10]{b2}{a2}
-      \end{tikzinline}}
-  \end{gathered}
-\end{gather*}
-There is no interesting stable point here.  The execution is disallowed
-because of a read from the causal future.  If we changed $x^\mRA$ to
-$x^\mRLX$, then there would be no order from $(\DR{\REF{1}}{7})$ to
-$(\DW[\mRLX]{x}{1})$, and the execution would be allowed.  The distinction
-between ``causal future'' and ``temporal future'' is not expressible in
-\citeauthor{Dolan:2018:BDR:3192366.3192421}'s operational semantics.
-
-Our definition of $L$-sequentiality does not quite correspond to SC
-executions, since actions may be elided by read/write elimination
-(\textsection\ref{sec:refine}).  However, for any properly initialized
-$L$-sequential pomset that uses elimination, there is larger $L$-sequential
-pomset that does not use elimination. This can be shown inductively---in the
-inductive step, writes that are introduced can be ignored by existing reads,
-and reads that are introduced can be fulfilled, for some value, by some
-preceding write.

### 2021-07-04 paper/mca1.tex
a stab at DRF

@@
 We drop the subscript when referring to both models.
+
+Note that reads-from implies order by \eqref{pom-rf-le}.
+% \begin{lemma}
+%   \label{lem:rf:implies:le}
+%   For any $\aPS$ in the range of $\sembase{}$, $\bEv\xrfx\aEv$ implies
+%   $\bEv\le\aEv$.
+
+%   \vspace{-.5\baselineskip}
+%   \begin{proof}
+%     Induction on the definition of $\sembase{}$, using \ref{par-rf-le1}, \ref{par-rf-le2}, and \ref{pom-rf-le}.
+%   \end{proof}
+% \end{lemma}
+In top-level pomsets, every read must have a matching write in $\rrfx$
+\eqref{top-rf}.  Together with \ref{pom-rf-match} and \ref{pom-rf-block}, the lemma
+guarantees that reads are \emph{fulfilled} at top-level, as in
+\cite[\textsection 2.7]{DBLP:journals/pacmpl/JagadeesanJR20}.\footnote{The
+  basic model would be the same if we move $\rrfx$ from the model itself to
+  be existentially quantified in the definition of top-level pomset, along
+  with \ref{pom-rf-match} and \ref{pom-rf-block}.  This was the approach of
+  \citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} We include $\rrfx$
+  explicitly for use in \textsection\ref{sec:arm2}, where we introduce a
+  variant semantics $\frf{\semrr{}}$ where \ref{pom-rf-le} is not required.}
+
+From \refdef{def:actions}, recall that $\aAct \rdelaysdef \bAct$ if
+$\aAct\eqreorderco\bAct$ or $\aAct\reorderra\bAct$ or
+$\aAct\eqreordersc\bAct$.  \ref{seq-le-delays} guarantees that sequential
+order is enforced between conflicting accesses of the same location
+($\eqreorderco$), into a release and out of an acquire ($\reorderra$), and
+between SC accesses ($\eqreordersc$).  Combined with the fulfillment
+requirements (\ref{pom-rf-match}, \ref{pom-rf-block} and \ref{pom-rf-le}),
+these ensure coherence, publication, subscription and other idioms.  For
+example, consider the following:\footnote{We use different colors for arrows representing order:
+  \begin{itemize}  
+  \item \makebox{$\bEv\xpo\aEv$} arises from control/data/address \emph{dependency} (\ref{seq-kappa}, definition of $\labelingForm'_2(\bEv)$),
+  \item \makebox{$\bEv\xsync\aEv$} arises from $\reorderra$ or $\eqreordersc$ \eqref{seq-le-delays},
+  \item \makebox{$\bEv\xwki\aEv$} arises from $\eqreorderco$ \eqref{seq-le-delays},
+  \item \makebox{$\bEv\xrf\aEv$} arises from \emph{reads-from} \eqref{pom-rf-match},
+  %\item \makebox{$\bEv\xsyncsc\aEv$} arises from \emph{strong fencing} \eqref{cand-lesync-sc},
+  \item \makebox{$\bEv\xwk\aEv$} arises from \emph{blocking} \eqref{pom-rf-block}.
+  \end{itemize}
+  In \textsection\ref{sec:arm2}, it is possible for $\rrf$ to contradict
+  $\le$.  In this case, we use a dotted arrow for $\rrf$: $\bEv\xrfint\aEv$
+  indicates that $\aEv\le\bEv$.}
+\begin{gather*}
+  \taglabel{pub}
+  \begin{gathered}    
+    \PW{x}{0}\SEMI 
+    \PW{x}{1}\SEMI \PW[\mREL]{y}{1} \PAR \PR[\mACQ]{y}{r}\SEMI \PR{x}{s}
+    \\[-.4ex]
+    \nonumber
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{wx0}{\DW{x}{0}}{}
+        \event{wx1}{\DW{x}{1}}{right=of wx0}
+        \event{wy1}{\DW[\mREL]{y}{1}}{right=of wx1}
+        \event{ry1}{\DR[\mACQ]{y}{1}}{right=2.5em of wy1}
+        \event{rx0}{\DR{x}{0}}{right=of ry1}
+        \sync{wx1}{wy1}
+        \sync{ry1}{rx0}
+        \rf{wy1}{ry1}
+        \wk[out=170,in=10]{rx0}{wx1}
+        \wki{wx0}{wx1}
+      \end{tikzinline}}
+  \end{gathered}
+\end{gather*}
+The execution is disallowed due to the cycle.  All of the order shown is
+required at top-level: The intra-thread order comes from \ref{seq-le-delays}:
+$\DWP{x}{0}\xwki \DWP{x}{1}$ is required by $\eqreorderco$.
+$\DWP{x}{1}\xsync \DWP[\mREL]{y}{1}$ and $\DRP[\mACQ]{y}{1}\xsync \DRP{x}{0}$ are
+required by $\reorderra$.  The cross-thread order is required by fulfillment:
+\ref{top-rf} requires that all top-level reads are in the image of $\xrfx$.
+\ref{pom-rf-match} ensures that $\DWP[\mREL]{y}{1}\xrfx \DRP[\mACQ]{y}{1}$, and
+\ref{pom-rf-le} subsequently ensures that
+$\DWP[\mREL]{y}{1}\le \DRP[\mACQ]{y}{1}$.  The \emph{antidependency}
+$\DRP{x}{0}\xwk \DWP{x}{1}$ is required by \ref{pom-rf-block}.  (Alternatively,
+we could have $\DWP{x}{1}\xwk \DWP{x}{0}$, again resulting in a cycle.)
+
+The semantics gives the expected results for store buffering and load buffering,
+as well as litmus tests involving fences and SC access.  The model of
+coherence is weaker than C11, in order to support common subexpression
+elimination, and stronger than Java, in order to support local reasoning
+about data races.  See
+\cite[\textsection3.1]{DBLP:journals/pacmpl/JagadeesanJR20} for a discussion.
 
@@
 
+  \label{def:pwt:mca:complete}
   A \PwTmca{} is \emph{complete} if %it is a complete \PwT{} (\refdef{def:pomset}) and

### 2021-07-04 paper/oopsla.tex
a stab at DRF

@@
 The substitutions collapse $\aLoc$ and $\aReg$, allowing local invariant
-reasoning (\xLIR{}), as required by causality test case 1, discussed at the end of
+reasoning (\xLIR{}), as required by \jmm{} causality test case 1, discussed at the end of
 \textsection\ref{sec:ex:control}.  Without Skolemizing it is necessary to

### 2021-07-04 paper/drf.tex
drf hacking

@@
 
-  We do not address \PwTc{}.  The existing \drfsc{} result for \cXI{}
+  We do not address \PwTc{}.  The internal \drfsc{} result for \cXI{}
   \cite{DBLP:phd/ethos/Batty15} does not rely on dependencies and thus
-  applies to \PwTc{}.  However, \citeauthor{DBLP:phd/ethos/Batty15}'s result
-  is \emph{internal}, where data-races are defined using the semantics of
-  \cXI{} itself.  Using the notion of dependency defined here, it should be
-  possible to prove an stronger external result for \cXI{}---we leave this as
-  future work.
+  applies to \PwTc{}.  In internal \drfsc{}, data-races are defined using the
+  semantics of the language itself.  Using the notion of dependency defined
+  here, it should be possible to prove an stronger external result for
+  \cXI{}, similar to that of \cite{DBLP:conf/pldi/LahavVKHD17}---we leave
+  this as future work.
 
@@
 
-\begin{changed}
-  Note that ${\rhb}$ is a preorder in general. It is a partial order on
-  \emph{simple} events (\refdef{def:po}).
-\end{changed}
+% \begin{changed}
+%   Note that ${\rhb}$ is a preorder in general. It is a partial order on
+%   \emph{simple} events (\refdef{def:po}).
+% \end{changed}
 
@@
 \begin{enumerate*}
-\item pomset order adjacent to $\aEv$, and
-\item the value associated with event $\aEv$, if it is a read.
+\item pomset order adjacent to $\aEv$, 
+\item the value associated with event $\aEv$, if it is a read, and
+\item the addition and removal of read events $\rpox$-after $\aEv$.
 \end{enumerate*}
 % We say they are \emph{similar} if they are $\aEv$-similar for some $\aEv$.
-Formally: $\Event'=\Event$, $\labelingForm'=\labelingForm$,
-$\restrict {\le'}{\Event\setminus\{\aEv\}}=\restrict {\le}{\Event\setminus\{\aEv\}}$,
-if $\aEv$ is not a read then $\labelingAct'=\labelingAct$, and if $\aEv$ is a
-read then
-$\restrict{\labelingAct'}{\Event\setminus\{\aEv\}}=\restrict{\labelingAct}{\Event\setminus\{\aEv\}}$
-and $\labelingAct'(\aEv) = \labelingAct(\aEv)[\aVal'/\aVal]$, for some
-$\aVal'$, $\aVal$.
+% Formally: $\Event'=\Event$, $\labelingForm'=\labelingForm$,
+% $\restrict {\le'}{\Event\setminus\{\aEv\}}=\restrict {\le}{\Event\setminus\{\aEv\}}$,
+% if $\aEv$ is not a read then $\labelingAct'=\labelingAct$, and if $\aEv$ is a
+% read then
+% $\restrict{\labelingAct'}{\Event\setminus\{\aEv\}}=\restrict{\labelingAct}{\Event\setminus\{\aEv\}}$
+% and $\labelingAct'(\aEv) = \labelingAct(\aEv)[\aVal'/\aVal]$, for some
+% $\aVal'$, $\aVal$.
 
@@
   \begin{lemma}
-    Suppose $\aPS''\in\semmin{\aCmd}$ has a read $\aEv$ that is maximal in
-    $({\rppo}\cup{\rrfe})$ and such that every $\rpox$-following read is
-    also $\le$-following ($\aEv\xpox\bEv$ implies $\aEv\le\bEv$, for every
-    read $\bEv$).  Further, suppose there is an $\aEv$-similar $\aPS'''$
-    that satisfies the requirements of fulfillment.  Then
-    $\aPS'''\in\semmin{\aCmd}$.
+    Suppose $\aPS''\in\semmin{\aCmd}$ has an external read $\bEv\xrfxpp\aEv$ that is
+    maximal in $({\rppo}\cup{\rrfe})$.  Further suppose that there another
+    write $\bEv'$ that could fulfill $\aEv$.
+    Then there exists an $\aEv$-similar $\aPS'''$ with $\bEv'\xrfxppp\aEv$
+    such that $\aPS'''\in\semmin{\aCmd}$.
+    % and such that every $\rpox$-following read is also $\le$-following
+    % ($\aEv\xpox\bEv$ implies $\aEv\le\bEv$, for every read $\bEv$).
+    % Further, suppose there is an $\aEv$-similar $\aPS'''$
+    % that satisfies the requirements of fulfillment.  Then
+    % $\aPS'''\in\semmin{\aCmd}$.
   \end{lemma}
@@
 
-  If there is no $({\rpox}\cup{\rrfe})$-maximal access, then all
-  cross-thread order must be from $\rrfe$.  In this case, we select a
+  If there is no $({\rpox}\cup{\rrfe})$-maximal access, then all cross-thread
+  order must be from $\rrfe$.  In this case, we select a
   $({\rppo}\cup{\rrfe})$-maximal read, switching its fulfilling write to an
-  ``older'' one.  As an example, consider the following; once again,
-  $\aEv$ is the read of $x$, which races with $(\DW{x}{1})$.
+  ``older'' one.  If there are several of these, we choose one that is
+  $\rpox$-minimal.  As an example, consider the following; once again, $\aEv$
+  is the read of $x$, which races with $(\DW{x}{1})$.
   \begin{gather*}

### 2021-07-05 paper/arm.tex
monday

@@
 \begin{lemma}
-  Any $\aPS$ in the image of $\frf{\semrr{}}$ is top-level iff
+  Any $\aPS$ in the image of $\frf{\semrr{}}$ is complete iff
   for every $\bEv\xrfx\aEv$ either
@@
     $(\exists\aEv'\mid\fmrg{\aEv'}=\aEv)$
-\todo{The following is sketchy.}
-    $\bEv\xpox\aEv$ and $(\not\exists\cEv)$
-    $\labeling(\cEv) \rblocks \labeling(\aEv)$ and $\bEv\xpox\cEv\xpox\aEv$.
+    $\bEv'\xpox\aEv'$ and $(\not\exists\cEv')$
+    $\labelingForm(\cEv)$ is a tautology and
+    $\labeling(\cEv) \rblocks \labeling(\aEv)$ and $\bEv'\xpox\cEv\xpox\aEv'$.
   \end{itemize}

### 2021-07-05 paper/intro.tex
monday

@@
 In certain cases it is possible to ban concurrent access using separation
-\cite{OHearn:2007:RCL:1235896.1236121}, or to accept inefficiency in order to
-obtain sequential consistency \cite{DBLP:conf/snapl/MarinoMMNS15}.  When
-these approaches are not available, however, we are left with an enormous gap
-in our understanding of one of the most basic elements of computing: the
-humble semicolon.
-Existing approaches either
+\cite{OHearn:2007:RCL:1235896.1236121}, or to accept inefficient
+implementation in order to obtain sequential consistency
+\cite{DBLP:conf/snapl/MarinoMMNS15}.  When these approaches are not
+available, however, we are left with an enormous gap in our understanding of
+one of the most basic elements of computing: the humble semicolon.  Until
+recently, existing approaches either
 \begin{itemize}
-\item don't bother tracking dependencies, allowing ``thin air'' executions \cite{DBLP:phd/ethos/Batty15},
-\item use syntactic dependencies
-  \cite{alglave,DBLP:journals/corr/abs-1804-04214}, disabling compiler optimizations,
-\item use complex non-compositional operational models  \cite{Manson:2005:JMM:1047659.1040336,
+\item didn't bother tracking dependencies, allowing ``thin air'' executions
+  \cite{Batty:2011:MCC:1926385.1926394},
+\item tracked dependencies conservatively, using syntax, requiring inefficient
+  implementation of relaxed access
+  \cite{DBLP:conf/pldi/LahavVKHD17,Boehm:2014:OGA:2618128.2618134,DBLP:conf/oopsla/VafeiadisN13,DBLP:journals/corr/abs-1804-04214}---while
+  this may be fine for C, it's a non-started for safe languages like Java,
+  which lack C's ``plain'' access,
+\item computed dependencies using non-compositional operational models over
+  alternate worlds \cite{Manson:2005:JMM:1047659.1040336,
     DBLP:conf/esop/JagadeesanPR10,
-    DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}, 
+    DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}---as
+  discussed in \textsection\ref{sec:promising}, these models fail to validate
+  temporal safety properties.
   %JMM, promising semantics, weakestMO, etc
-\item use prefixing \cite{DBLP:journals/pacmpl/JagadeesanJR20} or
-  continuation-passing \cite{DBLP:conf/esop/PaviottiCPWOB20} rather than
-  a direct presentation of sequential composition.
 \end{itemize}
 
-In this paper, we show that families of predicate transformers can be used to
-calculate semantic dependencies in a way that is \emph{compositional} and
-\emph{direct}: \emph{compositional} in that the denotation of
-$(\aCmd_1\SEMI \aCmd_2)$ can be computed from the denotation of $\aCmd_1$ and
-the denotation of $\aCmd_2$, and \emph{direct} in that these can be
-calculated independently.
+Recently, two denotational models have been proposed that compute sequential
+dependencies semantically. \citet{DBLP:conf/esop/PaviottiCPWOB20} defined
+Modular Relaxed Dependencies (\MRD), which use event structures to calculate
+dependencies for \cXI.  \citet{DBLP:journals/pacmpl/JagadeesanJR20} defined
+Pomsets with Preconditions (\PwP), which use preconditions and logic to
+calculate dependencies for a Java-like language on \mca{} hardware, such as
+\armeight{} \cite{DBLP:journals/pacmpl/PulteFDFSS18}.  However, neither paper
+treated sequential composition as a first-class citizen.  \MRD{} encoded
+sequential composition using continuation-passing, and \PwP{} used prefixing,
+adding one event at a time on the left.  In both cases, the approaches
+require perfect knowledge of the future.
 
-Our contributions:
+In this paper, we show that \PwP{} can be extended with \emph{families of
+  predicate transformers} to calculate sequential dependencies in a way that
+is \emph{compositional} and \emph{direct}: \emph{compositional} in that the
+denotation of $(\aCmd_1\SEMI \aCmd_2)$ can be computed from the denotation of
+$\aCmd_1$ and the denotation of $\aCmd_2$, and \emph{direct} in that these
+can be calculated independently.
+
+One of the main difficulties is to find a definition that is associative:
+the denotation of $((\aCmd_1\SEMI \aCmd_2)\SEMI \aCmd_3)$ should be the same as 
+the denotation of $(\aCmd_1\SEMI (\aCmd_2\SEMI \aCmd_3))$.
+
+We have layered the definitions, both to manage complexity and to make clear
+that the main contribution of this paper is broadly applicable.  After an
+overview, we define sequential dependencies in \textsection\ref{sec:model}.
+The next two sections add concurrency.  In \textsection\ref{sec:mca}, we
+define \PwTmca{}, which models Java for \mca{} hardware, similar to that of
+\citet{DBLP:journals/pacmpl/JagadeesanJR20}.  In \textsection\ref{sec:c11},
+we define \PwTc{}, which models \cXI, adapting the approach of
+\citet{DBLP:conf/esop/PaviottiCPWOB20}.
+
+We describe two variants of \PwTmca{}.  The first has a nicer semantics; the
+second has a more efficient lowering for acquiring reads on \armeight{}.
+Both variants have optimal lowering for relaxed access.
+
+\todo{flesh this out}
 \begin{itemize}
-\item We define a denotational model for calculating dependencies.
-\item We apply the model to the \cXI{} semantics of
-  \cite{DBLP:conf/esop/PaviottiCPWOB20} (\PwTc) and to the \mca{} semantics of
-  \cite{DBLP:journals/pacmpl/JagadeesanJR20} (\PwTmca{}).
 \item We provide a tool to execute litmus tests for both models.
-\item We prove \drfsc{} for \PwP{}; \MRD{} inherits the result from \rcXI.
-\item We prove a lowering result for \PwP{}.
-\item We extend the model to include many features.
+\item \textsection\ref{sec:drf} We prove \drfsc{} for \PwTmca{}; \MRD{} inherits the result from \rcXI.
+\item \textsection\ref{sec:arm} We prove a lowering result for \PwTmca{}.
+\item \textsection\ref{sec:additional} We extend the model to include many features.
 \end{itemize}
+
+% \subsection{Contributions}
+
+% \todo{}
+% We show how predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75} can
+% be added to pomsets with
+% preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} to create a
+% compositional semantics for sequential composition.
+\begin{itemize}
+\item \S\ref{sec:model} presents the basic model, which satisfies many
+  desiderata, but not all.
+\item \S\ref{sec:arm} shows two approaches for efficient implementation on
+  Arm.  The first uses a suboptimal lowering for acquiring reads.  The second
+  uses an optimal lowering, but requires a nontrivial change to the
+  definition of sequential composition.
+\item \S\ref{sec:additional} generalizes the basic semantics of read and write
+  to validate compiler optimizations.
+\end{itemize}
+% Acknowledgements go here, once we're not double-blinded.
+% The definitions in this paper have been formalized in Agda.
+Because it is closely related, we expect that the memory-model results of
+\cite{DBLP:journals/pacmpl/JagadeesanJR20} apply to our model, including
+compositional reasoning for temporal safety properties and {local} \drfsc{}
+as in \cite{Dolan:2018:BDR:3192366.3192421,DBLP:conf/ppopp/DongolJR19,promising-ldrf}.
+% In \textsection\ref{sec:arm}, we provide an alternative proof strategy for
+% efficient compilation to \armeight{}, which improves upon that of
+% \cite{DBLP:journals/pacmpl/JagadeesanJR20} by using a recent alternative
+% characterization of \armeight{}.
 
@@
 
-Existing approaches to sequential composition in the concurrent context
-either assume exclusive access, as in concurrent separation logic
-\cite{OHearn:2007:RCL:1235896.1236121}, or abandon the logical approach
-altogether, as in the pomset model of
-\citet{DBLP:journals/corr/abs-1804-04214}, which enforces syntactic
-dependencies.  This leaves open the question of how to apply logic to racy
-programs without overconstraining the implementation.  To understand the
-solution, one must first understand the constraints imposed by hardware and
-compilers.
+In this paper we answer the following question: what does sequential
+composition mean in a concurrent context?  An acceptable answer must satisfy
+several desiderata:
+\begin{enumerate}
+\item\label{too:cold} it should not impose too much order, overconstraining the implementation,
+\item\label{too:hot} it should not impose too little order, allowing bogus executions, and
+\item it should be \emph{compositional} and \emph{direct}, as described in \textsection\ref{sec:intro}.
+\end{enumerate}
+Memory models differ in how they navigate between desiderata \ref{too:cold}
+and \ref{too:hot}.  In one direction there are both more valid compiler
+optimizations and also more potentially dubious executions, in the other
+direction, less of both.  To understand the tradeoffs, one must first
+understand the underlying hardware and compilers.
+% Existing work studies program logics over non-composition operational or
+% axiomatic models
+% \cite{OHearn:2007:RCL:1235896.1236121,DBLP:conf/oopsla/VafeiadisN13,DBLP:conf/esop/SvendsenPDLV18}.
+%
+% Existing approaches either assume exclusive access, as in concurrent
+% separation logic \cite{OHearn:2007:RCL:1235896.1236121}, or abandon the
+% logical approach altogether, as in the pomset model of
+% \citet{DBLP:journals/corr/abs-1804-04214}, which enforces syntactic
+% dependencies.  This leaves open the question of how to apply logic to racy
+% programs without overconstraining the implementation.
 
@@
 To address this, \citet{DBLP:journals/pacmpl/JagadeesanJR20} introduced
-\emph{Pomsets with Preconditions}, where events are labeled with logical
+\emph{Pomsets with Preconditions (\PwP{})}, where events are labeled with logical
 formulae.  Nontrivial preconditions are introduced by store actions (modeling
@@
 We provide a detailed comparison with these approaches in \textsection\ref{sec:promising}.
-
-\subsection{Contributions}
-
-We show how predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75} can
-be added to pomsets with
-preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} to create a
-compositional semantics for sequential composition.
-\begin{itemize}
-\item \S\ref{sec:model} presents the basic model, which satisfies many
-  desiderata, but not all.
-\item \S\ref{sec:arm} shows two approaches for efficient implementation on
-  Arm.  The first uses a suboptimal lowering for acquiring reads.  The second
-  uses an optimal lowering, but requires a nontrivial change to the
-  definition of sequential composition.
-\item \S\ref{sec:additional} generalizes the basic semantics of read and write
-  to validate compiler optimizations.
-\end{itemize}
-% Acknowledgements go here, once we're not double-blinded.
-% The definitions in this paper have been formalized in Agda.
-Because it is closely related, we expect that the memory-model results of
-\cite{DBLP:journals/pacmpl/JagadeesanJR20} apply to our model, including
-compositional reasoning for temporal safety properties and {local} \drfsc{}
-as in \cite{Dolan:2018:BDR:3192366.3192421,DBLP:conf/ppopp/DongolJR19,promising-ldrf}.
-% In \textsection\ref{sec:arm}, we provide an alternative proof strategy for
-% efficient compilation to \armeight{}, which improves upon that of
-% \cite{DBLP:journals/pacmpl/JagadeesanJR20} by using a recent alternative
-% characterization of \armeight{}.

### 2021-07-06 paper/intro.tex
got the definitions to line up with using weakest preconditions.

@@
     DBLP:conf/esop/JagadeesanPR10,
-    DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}---as
-  discussed in \textsection\ref{sec:promising}, these models fail to validate
-  temporal safety properties.
-  %JMM, promising semantics, weakestMO, etc
+    DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}---these
+  models validate many compiler optimizations, but also
+  %allow some dubious executions 
+  fail to validate temporal safety properties
+  (see \textsection\ref{sec:promising}).
 \end{itemize}
@@
 In this paper, we show that \PwP{} can be extended with \emph{families of
-  predicate transformers} to calculate sequential dependencies in a way that
+  predicate transformers (\PwT{})} to calculate sequential dependencies in a way that
 is \emph{compositional} and \emph{direct}: \emph{compositional} in that the
@@
 $\aCmd_1$ and the denotation of $\aCmd_2$, and \emph{direct} in that these
-can be calculated independently.
+can be calculated independently.  The definition is associative: the
+denotation of $((\aCmd_1\SEMI \aCmd_2)\SEMI \aCmd_3)$ is the same as the
+denotation of $(\aCmd_1\SEMI (\aCmd_2\SEMI \aCmd_3))$.  It also validates
+expected laws concerning the interaction of sequencing and conditional
+execution.
 
-One of the main difficulties is to find a definition that is associative:
-the denotation of $((\aCmd_1\SEMI \aCmd_2)\SEMI \aCmd_3)$ should be the same as 
-the denotation of $(\aCmd_1\SEMI (\aCmd_2\SEMI \aCmd_3))$.
-
-We have layered the definitions, both to manage complexity and to make clear
-that the main contribution of this paper is broadly applicable.  After an
-overview, we define sequential dependencies in \textsection\ref{sec:model}.
-The next two sections add concurrency.  In \textsection\ref{sec:mca}, we
-define \PwTmca{}, which models Java for \mca{} hardware, similar to that of
+To manage complexity, we have layered the definitions.  After an overview, we
+define sequential dependencies in \textsection\ref{sec:model}.  The next two
+sections add concurrency.  In \textsection\ref{sec:mca}, we define \PwTmca{},
+which models Java for \mca{} hardware, similar to that of
 \citet{DBLP:journals/pacmpl/JagadeesanJR20}.  In \textsection\ref{sec:c11},
 we define \PwTc{}, which models \cXI, adapting the approach of
-\citet{DBLP:conf/esop/PaviottiCPWOB20}.
-
-We describe two variants of \PwTmca{}.  The first has a nicer semantics; the
-second has a more efficient lowering for acquiring reads on \armeight{}.
-Both variants have optimal lowering for relaxed access.
+\citet{DBLP:conf/esop/PaviottiCPWOB20}.  In \textsection\ref{sec:additional},
+we extend the semantics to include additional features, such as address calculation.
+In \textsection{sec:results}, we summarize the results.
 

### 2021-07-06 paper/mca1.tex
monday
got the definitions to line up with using weakest preconditions.

@@
-\section{\PwTmcaTITLE{1}: Pomsets with Predicate Transformers for MCA (Part 1)}
+\section{\PwTmcaTITLE{}: Pomsets with Predicate Transformers for MCA}
+\label{sec:mca}
+We define two models for \mca{} architectures: \PwTmca{1} and \PwTmca{2}.
+The first has a nicer semantics; the second has a more efficient lowering for
+acquiring reads on \armeight{}.  Both variants have optimal lowering for
+relaxed access.  We drop the subscript when referring to both models.
+
+\subsection{\PwTmcaTITLE{1}}
 \label{sec:mca1}
-We define two models for \mca{} architectures: \PwTmca{1} and \PwTmca{2}.
-We drop the subscript when referring to both models.
 

### 2021-07-07 paper/intro.tex
progress

@@
 we extend the semantics to include additional features, such as address calculation.
-In \textsection{sec:results}, we summarize the results.
+In \textsection\ref{sec:results}, we summarize the results.
 
@@
 \item We provide a tool to execute litmus tests for both models.
-\item \textsection\ref{sec:drf} We prove \drfsc{} for \PwTmca{}; \MRD{} inherits the result from \rcXI.
+\item \textsection\ref{sec:sc} We prove \drfsc{} for \PwTmca{}; \MRD{} inherits the result from \rcXI.
 \item \textsection\ref{sec:arm} We prove a lowering result for \PwTmca{}.

### 2021-07-07 paper/mca1.tex
progress

@@
 
+Concerning \reflem{lem:if}:
+For \PwTmca{}, \eqref{lem:ifelse:if:if1} and \eqref{lem:ifelse:if:if2} are
+inclusions rather than equations.  In \textsection\ref{sec:delay}, we refine
+the semantics to validate the reverse inclusions.
+
 \subsection{\PwTmcaTITLE{1}}
 \label{sec:mca1}
+
+Parallel composition is disjoint union.  Any $\rrfx$ edges added between the
+two sides must also be added to the order (\ref{par-rf-le1} and
+\ref{par-rf-le2}).
+
+Sequential composition is similar, with two changes: \ref{seq-E} does not
+require disjointness (see \textsection\ref{sec:ex:term}), and
+\ref{seq-le-delays} may require order (see example \ref{pub}, below).
+
+We include the \emph{reads-from} relation explicitly in the model
+\eqref{pom-rf}.
 

### 2021-07-07 paper/mca2.tex
progress

--- 2021-06-30 paper/mca.tex
@@
-\section{\PwTmcaTITLE{}: Pomsets with Predicate Transformers for MCA}
-\label{sec:mca}
-We define two models for \mca{} architectures: \PwTmca{1} and \PwTmca{2}.
-We drop the subscript when referring to both models.
+%\subsection{\PwTmcaTITLE{2}: Pomsets with Predicate Transformers for MCA (Part 2)}
+\subsection{\PwTmcaTITLE{2}}
+\label{sec:mca2}
 
-\subsection{First Semantics}
-\label{sec:mca1}
-
-\begin{definition}
-  \label{def:pwt:mca1}
-  A \PwTmca{1} is a \PwT{} (\refdef{def:pomset}) equipped
-  with a relation $\rrfx$ such that 
-  \begin{enumerate}[,label=(\textsc{m}\arabic*),ref=\textsc{m}\arabic*]
-    \setcounter{enumi}{\value{Brf}}
-  \item \label{pom-rf} \makecounter{rf} ${\rrfx} \subseteq \aEvs\times\aEvs$
-    is an injective relation capturing \emph{reads-from}, such that
-    % \end{enumerate}
-    % A pomset is a \emph{candidate} if there is an injective relation
-    % ${\rrfx} : \aEvs\times\aEvs$, capturing \emph{reads-from}, such that
-    \begin{enumerate}
-      % \begin{enumerate}[,label=(\textsc{i}\arabic*),ref=\textsc{i}\arabic*]
-      % \item \label{rf-injective}
-      %   if $\bEv\xrfx\aEv$ and $\cEv\xrfx\aEv$ then $\bEv=\cEv$, that is,
-      %   ${\rrfx}$ is injective,
-    \item \label{pom-rf-match} if $\bEv\xrfx\aEv$ then
-      $\labeling(\bEv) \rmatches \labeling(\aEv)$,
-    \item \label{pom-rf-block} if $\bEv\xrfx\aEv$ and
-      $\labeling(\cEv) \rblocks \labeling(\aEv)$ then either $\cEv\le\bEv$ or
-      $\aEv\le\cEv$,
-    \item \label{pom-rf-le} if $\bEv\xrfx\aEv$ then $\bEv\le\aEv$.
-    \end{enumerate}
-  \end{enumerate}
-
-  A \PwTmca{} is \emph{complete} if %it is a complete \PwT{} (\refdef{def:pomset}) and
-  \begin{multicols}{2}
-    \begin{enumerate}[,label=(\textsc{c}\arabic*),ref=\textsc{c}\arabic*]
-    \item[\eqref{top-kappa}]
-      \eqref{top-term}\;
-      as in \refdef{def:pomset},
-      \setcounter{enumi}{\value{Brf}}
-    \item \label{top-rf}
-      if $\labeling(\aEv)$ is a $\sread$ then there is some $\bEv\xrfx\aEv$.
-    \end{enumerate}
-  \end{multicols}
-\end{definition}
-
-% To include \RMW{}s, also add
-% \begin{enumerate}
-% \item[(\textsc{p}10)]\;
-%   ${\rrmw}=\PBR{{\rrmw_1}\cup{\rrmw_2}}$,
-% \end{enumerate}
-
-\input{fig-mca1.tex}
-\input{fig-mca2.tex}
-
-The semantic rules are given in \reffig{fig:mca1}.  Let
-\begin{math}
-  \sembase{\aCmd_1 \LPAR[\bThrd] \aCmd_2} = \sLPAR{\sembase{\aCmd_1}}{\sembase{\aCmd_2}}.
-\end{math}
-
-From the data model, we require an additional binary relation 
-${\rdelays}\subseteq(\Act\times\Act)$.  For the actions in this paper, we
-define this as follows.
-
-\begin{scope}
-  Let ${\eqreorderco}$ capture write-write, read-write coherence:
-  \begin{math}
-    {\eqreorderco}
-    =
-    \{(\DW{\aLoc}{}\Cb \DW{\aLoc}{})\Cc(\DR{\aLoc}{}\Cb \DW{\aLoc}{})\Cc(\DW{\aLoc}{}\Cb \DR{\aLoc}{})\}
-  \end{math}.
-
-  Let ${\reorderra}$ capture conflict due to synchronization:
-  \begin{math}
-    {\reorderra}
-    =
-    \{(\aAct\Cb             \DW[\gemode\mREL]{}{}     )\Cc
-    (\aAct\Cb               \DF{\gemode\fREL}        )\Cc
-    (\DR{}{}\Cb             \DF{\gemode\fACQ}        )\Cc
-    % (\DR{\aLoc}{}\Cb        \DR[\gemode\mACQ]{\aLoc}{})\Cc
-    (\DR[\gemode\mACQ]{}{}\Cb\aAct                    )\Cc
-    (\DF{\gemode\fACQ}\Cb   \aAct                    )\Cc
-    (\DF{\gemode\fREL}\Cb   \DW{}{}                  )\Cc
-    (\DW[\gemode\mREL]{\aLoc}{}\Cb\DW{\aLoc}{})\}
-  \end{math}.
-
-  Let ${\eqreordersc}$ capture conflict due to $\mSC$ access:
-  \begin{math}
-    {\eqreordersc}
-    =
-    % \{(\DX[\mSC]{}{}\Cb\DX[\mSC]{}{})\}
-    \{(\DW[\mSC]{}{}\Cb \DW[\mSC]{}{})\Cc(\DR[\mSC]{}{}\Cb \DW[\mSC]{}{}) \Cc(\DW[\mSC]{}{}\Cb \DR[\mSC]{}{})\Cc(\DR[\mSC]{}{}\Cb \DR[\mSC]{}{})\}
-  \end{math}.
-
-  We say $\aAct \rdelaysdef \bAct$ if $\aAct\eqreorderco\bAct$ or $\aAct\reorderra\bAct$ or $\aAct\eqreordersc\bAct$.
-\end{scope}
-
-\subsection{Second Semantics}
-\label{sec:mca2}
+%
+The use of $\rextendsdef{}{}$ in \ref{if-rf-extends}
+ensures that no $\rrfx$ is introduced between events in
+$\aEvs_1\cap\aEvs_2$ when coalescing.
 
@@
   \label{def:pwt:mca2}
-  A \PwTmca{2} is a \PwT{} (\refdef{def:pomset}) equipped
-  with a relation $\rrfx$ such that 
-  \begin{enumerate}[,label=(\textsc{m}\arabic*),ref=\textsc{m}\arabic*]
-    \setcounter{enumi}{\value{Brf}}
-  \item \label{pom-rf'} ${\rrfx} \subseteq \aEvs\times\aEvs$
-    is an injective relation capturing \emph{reads-from}, such that
+  A \PwTmca{2} is a \PwT{} (\refdef{def:pomset}) equipped with an injective
+  relation $\rrfx$ that satisfies requirements \ref{pom-rf-match} and
+  \ref{pom-rf-block} of \refdef{def:pwt:mca1}.
+
+  A A \PwTmca{2} is \emph{complete} if it satisfies
+  \ref{top-kappa}, \ref{top-term}, and \ref{top-rf}.
+  % \begin{enumerate}[,label=(\textsc{m}\arabic*),ref=\textsc{m}\arabic*]
+  %   \setcounter{enumi}{\value{Brf}}
+  % \item \label{pom-rf'} ${\rrfx} \subseteq \aEvs\times\aEvs$
+  %   is an injective relation capturing \emph{reads-from}, such that
+  %   % \end{enumerate}
+  %   % A pomset is a \emph{candidate} if there is an injective relation
+  %   % ${\rrfx} : \aEvs\times\aEvs$, capturing \emph{reads-from}, such that
+  %   \begin{multicols}{2}    
+  %     \begin{enumerate}
+  %       % \begin{enumerate}[,label=(\textsc{i}\arabic*),ref=\textsc{i}\arabic*]
+  %       % \item \label{rf-injective}
+  %       %   if $\bEv\xrfx\aEv$ and $\cEv\xrfx\aEv$ then $\bEv=\cEv$, that is,
+  %       %   ${\rrfx}$ is injective,
+  %     \item[\eqref{pom-rf-match}]
+  %       \eqref{pom-rf-block}\;
+  %       as in \refdef{def:pwt:mca1},
+  %       \setcounter{enumii}{2}
+  %     \item \label{pom-rf-le'} 
+  %       if $\bEv\xrfx\aEv$ then either $\bEv\le\aEv$ or $\aEv\le\bEv$.
   %     \end{enumerate}
-    % A pomset is a \emph{candidate} if there is an injective relation
-    % ${\rrfx} : \aEvs\times\aEvs$, capturing \emph{reads-from}, such that
-    \begin{multicols}{2}    
-      \begin{enumerate}
-        % \begin{enumerate}[,label=(\textsc{i}\arabic*),ref=\textsc{i}\arabic*]
-        % \item \label{rf-injective}
-        %   if $\bEv\xrfx\aEv$ and $\cEv\xrfx\aEv$ then $\bEv=\cEv$, that is,
-        %   ${\rrfx}$ is injective,
-      \item[\eqref{pom-rf-match}]
-        \eqref{pom-rf-block}\;
-        as in \refdef{def:pwt:mca1},
-        \setcounter{enumii}{2}
-      \item \label{pom-rf-le'} 
-        if $\bEv\xrfx\aEv$ then either $\bEv\le\aEv$ or $\aEv\le\bEv$.
-      \end{enumerate}
-    \end{multicols}
-  \end{enumerate}
+  %   \end{multicols}
+  % \end{enumerate}
 \end{definition}
+A \PwTmca{2} need not satisfy requirement \ref{pom-rf-le}, and thus we may
+have $\bEv\xrfx\aEv$ and $\aEv\le\bEv$.
+
 % Derive $\rdelayspdef$ from $\rdelaysdef$ by replacing $\eqreorderco$ with
@@
 The semantic rules are given in \reffig{fig:mca2}.
+We write $\semmca{2}{}$ for the semantic function.

### 2021-07-07 paper/results.tex
progress

--- 2021-06-17 paper/2SUB/definitions.tex
@@
-\section{The Basic Model}
-\label{sec:model}
+\section{Results}
+\label{sec:results}
 
-After some preliminaries (\textsection\ref{sec:prelim}--\ref{sec:actions}),
-we define the basic model and establish some basic properties
-(\textsection\ref{sec:pomsets} and \reffig{fig:sem}).  We then explain the
-model using examples (\textsection\ref{sec:ex:pomset}--\ref{sec:ex:last}).
-We encourage readers to skim the definitions and then skip to
-\textsection\ref{sec:ex:pomset}, coming back as needed.
-
-%% Batty suggest example where dependencies are added and also go away, perhaps
-%% by store forwarding. Something like:
-%% \texttt{(r=x; y=1); (s=y; z=s+r)}
-
-% In this section, we present the mathematical preliminaries for the
-% model (which can be skipped on first reading). We then present the
-% model incrementally, starting with a model built using
-% \emph{partially ordered multisets}
-% (\emph{pomsets})~\cite{GISCHER1988199,Plotkin:1997:TSP:266557.266600},
-% and then adding preconditions and finally predicate transformers.
-
-% In later sections, we will discuss extensions to the logic, and to the
-% semantics of load, store and thread initialization, in order to model
-% relaxed memory more faithfully. We stress that these features do
-% \emph{not} change any of the structures of the language: conditionals,
-% parallel composition, and sequential composition are as defined in this section.
-
-\subsection{Preliminaries}
-\label{sec:prelim}
-The syntax is built from
-\begin{itemize}
-\item a set of \emph{values} $\Val$, ranged over by
-  $\aVal$, $\bVal$, $\cVal$, $\dVal$,
-\item a set of \emph{registers} $\Reg$, ranged over by
-  $\aReg$, $\bReg$,
-\item a set of \emph{expressions} $\Exp$, ranged over by
-  $\aExp$, $\bExp$,  $\cExp$.
-\end{itemize}
-
-\emph{Memory references} are tagged values, written $\REF{\cVal}$.  Let $\Loc$
-be the set of memory references, ranged over by $\aLoc$, $\bLoc$, $\cLoc$.
-% 
-We require that
-\begin{itemize}
-\item values and registers are disjoint, 
-\item values include at least the constants $0$ and $1$,  
-\item expressions include at least registers and values, 
-\item expressions do \emph{not} include references: $\aExp[\bExp/\aLoc]=\aExp$.
-\end{itemize}
-
-We model the following language.
-\begin{gather*}
-  \begin{aligned}
-    \amode,\bmode \BNFDEF& \mRLX
-    \BNFSEP \mREL
-    \BNFSEP \mACQ
-    \BNFSEP \mSC
-    % \amode \BNFDEF& \mRLX
-    % \BNFSEP \mRA 
-    % \BNFSEP \mSC
-    % &\mkern100mu
-    % \fmode \BNFDEF& \fACQ 
-    % \BNFSEP \fREL
-    % \BNFSEP \fSC
-  \end{aligned}
-  \\
-  \aCmd
-  \BNFDEF \LET{\aReg}{\aExp}
-  \BNFSEP \PR[\amode]{\REF{\cExp}}[\ascope]{\aReg}
-  \BNFSEP \PW[\amode]{\REF{\cExp}}[\ascope]{\aExp}
-  \BNFSEP \PF[\ascope]{\fmode}
-  \BNFSEP \SKIP
-  \BNFSEP \aCmd_1 \SEMI \aCmd_2
-  \BNFSEP \IF{\aExp} \THEN \aCmd_1 \ELSE \aCmd_2 \FI
-  \BNFSEP \aCmd_1 \LPAR[\bThrd] \aCmd_2
-  % \\[-.5ex]
-  % \BNFSEP& \PCAS[\amode_1][\amode_2]{\REF{\cExp}}[\ascope]{\aReg}{\aExp}{\bExp}
-  % \BNFSEP \PFADD[\amode_1][\amode_2]{\REF{\cExp}}[\ascope]{\aReg}{\aExp}
-  % \BNFSEP \PEXCHG[\amode_1][\amode_2]{\REF{\cExp}}[\ascope]{\aReg}{\aExp}
-\end{gather*}
-
-% \emph{Memory modes}, $\amode$, are {relaxed} ($\mRLX$), {release-acquire}
-% ($\mRA$), and {sequentially consistent} ($\mSC$).  Relaxed mode is the
-% default; we regularly elide it from examples.  $\mRA$/$\mSC$ accesses are
-% collectively known as \emph{synchronized accesses}.
-
-% \emph{Fence modes}, $\bmode$, are {acquire} ($\fACQ$), {release} ($\fREL$), 
-% and {acquire-release} ($\fSC$).
-
-\emph{Access modes}, $\amode$, are {relaxed} ($\mRLX$),
-{release} ($\mREL$), {acquire} ($\mACQ$), and
-{sequentially consistent} ($\mSC$).
-%
-Let expressions ($\LET{\aReg}{\aExp}$) only affect thread-local state and
-thus do not have a mode.
-%
-Reads ($\PR[\amode]{\REF{\cExp}}[\ascope]{\aReg}$) support
-%$\mWK$,
-$\mRLX$,
-%$\mREL$,
-$\mACQ$,
-%$\mRA$, 
-$\mSC$. 
-Writes ($\PW[\amode]{\REF{\cExp}}[\ascope]{\aReg}$) support
-%$\mWK$,
-$\mRLX$,
-$\mREL$,
-%$\mACQ$,
-%$\mRA$, 
-$\mSC$. 
-Fences ($\PF[\ascope]{\fmode}$) support
-%$\mWK$,
-%$\mRLX$,
-$\mREL$,
-$\mACQ$,
-%$\mRA$,
-$\mSC$.
-
-\emph{Commands}, aka \emph{statements}, $\aCmd$, include memory accesses at a
-given mode, as well as the usual structural constructs. Following
-\cite{DBLP:conf/icfp/FerreiraHJ96}, $\LPAR$ denotes parallel composition,
-preserving thread state on the left after a join.  In examples and
-sublanguages without join, we use the symmetric $\PAR$ operator.
-
-We use common syntax sugar, such as \emph{extended expressions}, $\aEExp$,
-which include memory locations.  For example, if $\aEExp$ includes a single
-occurrence of $\aLoc$, then $\bLoc\GETS\aEExp\SEMI \aCmd$ is shorthand for
-$\aReg\GETS\aLoc\SEMI\bLoc\GETS\aEExp[\aReg/\aLoc]\SEMI \aCmd$.  Each
-occurrence of $\aLoc$ in an extended expression corresponds to an separate
-read.  We also write $\IF{\aExp} \THEN \aCmd \FI$ as shorthand for
-$\IF{\aExp} \THEN \aCmd\ELSE \SKIP\FI$.
-
-Throughout \textsection\ref{sec:intro}--\ref{sec:arm} we 
-require that
-\begin{itemize}
-\item each register is assigned at most once in a program.
-  % \end{itemize}
-  % In \textsection\ref{sec:complications} and following, we
-  % require instead that
-  % \begin{itemize}
-  % \end{itemize}
-\end{itemize}
-In \textsection\ref{sec:additional}, we drop this restriction, requiring
-instead that
-\begin{itemize}
-\item there are registers
-  $\uRegs{\AllEvents}=\{\uReg{\aEv}\mid\aEv\in\AllEvents\}$, that do not
-  appear in programs: $\aCmd[\bExp/\uReg{\aEv}]=\aCmd$.
-\end{itemize}
-% In contexts that make no use of $\uRegs{\AllEvents}$, we make the first
-% assumption.
-
-The semantics is built from the following.
-\begin{itemize}
-\item a set of \emph{events} $\AllEvents$, ranged over by $\aEv$, $\bEv$,
-  $\cEv$, %$\dEv$,
-  and subsets ranged over by $\aEvs$, $\bEvs$, $\cEvs$,  
-\item a set of \emph{logical formulae} $\Formulae$, ranged over by $\aForm$,
-  $\bForm$, $\cForm$,
-\item a set of \emph{actions} $\Act$, ranged over by $\aAct$, $\bAct$.
-\end{itemize}
-% $\dEvs$.
-
-We require that
-\begin{itemize}
-\item formulae include $\TRUE$, $\FALSE$ and the equalities $(\aExp{=}\bExp)$ and $(\aLoc{=}\aExp)$,
-\item formulae are closed under $\lnot$, $\land$, $\lor$, $\limplies$, and
-  substitutions $[\aExp/\aReg]$, $[\aExp/\aLoc]$,
-\item there is a relation $\rimpliesdef$ between
-  formulae, capturing entailment, %\subseteq(\Formulae\times\Formulae)$ %
-\item $\rimpliesdef$ has the expected semantics for $=$, $\lnot$,
-  $\land$, $\lor$, $\limplies$ and substitutions $[\aExp/\aReg]$, $[\aExp/\aLoc]$,
-\item there are three binary relations over $\Act\times\Act$:
-  $\rmatchesdef$, $\rblocksdef$, and $\rdelaysdef$,
-\item there are two subsets of $\Act$, distinguishing
-  $\sreaddef$ and $\sreleasedef$ actions.
-\end{itemize}
-
-Logical formulae include equations over registers and memory references, such as
-$(\aReg{=}\bReg{+}1)$ and $(\aLoc{=}1)$.
-% For use in \textsection\ref{sec:tc1}, we also include equations over memory references, such as $(\aLoc{=}1)$.
-% I would like to drop this, an be careful about program vs logical syntax
-We use expressions as formulae, coercing $\aExp$ to $\aExp{\neq}0$.
-% Equations have precedence over logical operators; thus
-% $\aReg{=}\aVal\limplies\bReg{>}\bVal$ is read
-% $(\aReg{=}\aVal)\limplies(\bReg{>}\bVal)$.  As usual, implication associates to the
-% right; thus $\aForm\limplies\bForm\limplies\cForm$ is read
-% $\aForm\limplies(\bForm\limplies\cForm)$.
-% As usual, implication associates to the right; thus
-% $\aReg{=}\aVal\limplies\bReg{>}\bVal\limplies\bForm$ is read
-% $(\aReg{=}\aVal)\limplies((\bReg{>}\bVal)\limplies\bForm)$.
-
-% Formulae are subject to substitutions; % of the form $[\aExp/\aReg]$ and
-% % $[\aExp/\aLoc]$;
-% actions are not.
-
-% \begin{definition}
-%   \label{def:independent}
-%   We say $\aForm$ is \emph{independent of $\aLoc$} if, for every
-%   $\aVal$, $\aForm \vDash \aForm[\aVal/\aLoc] \vDash \aForm$; it is
-%   \emph{dependent} otherwise.
-%   We say $\aForm$ is \emph{location independent} if it is independent of
-%   every location.
-%   We say
-%   $\aForm$ \emph{implies} $\bForm$ if $\aForm\vDash\bForm$.
-%   We say that
-We write $\aForm\riff\bForm$ when $\aForm\rimplies\bForm$ and $\bForm\rimplies\aForm$.
-We say
-$\aForm$ is a \emph{tautology} if $\TRUE \rimplies \aForm$.
-% We say that
-We say
-$\aForm$ is \emph{unsatisfiable} if $\aForm \rimplies \FALSE$, and
-\emph{satisfiable} otherwise.
-% \end{definition}
-
-\subsection{Actions in This Paper}
-\label{sec:actions}
-In this paper, we let actions be reads and writes and fences:
-\begin{displaymath}
-  \aAct,\bAct \BNFDEF \DW[\amode]{\aLoc}[\ascope]{\aVal}[\aThrd]
-  \BNFSEP \DR[\amode]{\aLoc}[\ascope]{\aVal}[\aThrd]
-  \BNFSEP \DF[\ascope]{\fmode}[\aThrd]
-\end{displaymath}
-% % writes $\DWP[\amode]{\aLoc}[\ascope]{\aVal}[\aThrd]$,
-% reads $\DRP[\amode]{\aLoc}[\ascope]{\aVal}[\aThrd]$,
-% and fences $\DFP[\ascope]{\fmode}[\aThrd]$.
-
-We use shorthand when referring to actions.  In definitions, we drop elements
-of actions that are existentially quantified.  In examples, we drop elements
-of actions, using defaults.
-% We write $\DXP[\amode]{}[\ascope]{}[\aThrd]$ to
-% stand for $\DWP[\amode]{}[\ascope]{}[\aThrd]$ or
-% $\DRP[\amode]{}[\ascope]{}[\aThrd]$.
-% 
-Let $\lemode$ be the smallest order over access and fence modes such that
-%$\mRLX\lemode\mRA\lemode\mSC$ and $\fREL\lemode\fSC$ and $\fACQ\lemode\fSC$.
-$\mRLX\lemode\mREL\lemode\mSC$ and $\mRLX\lemode\mACQ\lemode\mSC$.
-We write $\DWP[\gemode\mREL]{}{}$ to stand for either $\DWP[\mREL]{}{}$ or
-$\DWP[\mSC]{}{}$, and similarly for the other actions and modes.
-
-% We also define shorthand for sets of
-% actions using an order on access and fence modes:  
-% \begin{align*}
-%   \begin{tikzcenter}
-%     \node (rlx) at (0, 0) {$\mathstrut\mRLX$};
-%     \node (ra)  at (1, 0) {$\mathstrut\mRA$};
-%     \node (sc)  at (2, 0) {$\mathstrut\mSC$};
-%     \draw[->](rlx)to(ra);
-%     \draw[->](ra)to(sc);
-%   \end{tikzcenter}
-%   &&
-%   \begin{tikzcenter}
-%     \node (fsc) at (3, 0) {$\mathstrut\fSC$};
-%     \node (rel) at (2, -0.2) {$\mathstrut\fREL$};
-%     \node (acq) at (2,  0.2) {$\mathstrut\fACQ$};
-%     \draw[->](rel)to(fsc);
-%     \draw[->](acq)to(fsc);
-%   \end{tikzcenter}
-% \end{align*}
-% We write $\amode\lemode\bmode$ for this order.
-% Let $\amode\lubmode\bmode$ denote the least upper bound of $\amode$ and $\bmode$.
-
-% or $\DFP[\ascope]{\amode}[\aThrd]$.
-% We write $\DWP[\gemode\mREL]{}{}$ to stand for either
-% $\DWP[\mREL]{}{}$ or $\DWP[\mSC]{}{}$, and similarly for other actions and
-% modes.
-
-\begin{definition}
-  \label{def:actions}
-  Actions $\DRP{}{}$ are $\sreaddef$ actions.
-  Actions $\DWP[\gemode\mREL]{}{}$ and $\DFP{\gemode\fREL}$ are
-  $\sreleasedef$ actions.
-
-  We say $\aAct \rmatchesdef \bAct$ if $\aAct=\DWP{\aLoc}{\aVal}$ and $\bAct=\DRP{\aLoc}{\aVal}$.
-  % We say $\DWP{\aLoc}{\aVal} \rmatchesdef \DRP{\aLoc}{\bVal}$ when $\aVal=\bVal$.
-  % Action $\DWP{\aLoc}{\aVal} \rmatchesdef \DRP{\aLoc}{\bVal}$ when $\aVal=\bVal$.
-
-  We say $\aAct \rblocksdef \bAct$ if $\aAct=\DWP{\aLoc}{}$ and $\bAct=\DRP{\aLoc}{}$, regardless of value.
-  % We say $\DWP{\aLoc}{\aVal} \rblocksdef \DRP{\aLoc}{\bVal}$, for any $\aVal$, $\bVal$.
-  % Action $\DWP{\aLoc}{\aVal} \rblocksdef \DRP{\aLoc}{\bVal}$, for any $\aVal$, $\bVal$.
-
-  % Let two actions \emph{overlap} if they access the same location.
-  % We say $\aAct \roverlapsdef \bAct$ if they access the same location.
-
-  Let ${\eqreorderco}$ capture write-write, read-write coherence:
-  \begin{math}
-    {\eqreorderco}
-    =
-    \{(\DW{\aLoc}{}\Cb \DW{\aLoc}{})\Cc(\DR{\aLoc}{}\Cb \DW{\aLoc}{})\Cc(\DW{\aLoc}{}\Cb \DR{\aLoc}{})\}
-  \end{math}.
-
-  Let ${\reorderra}$ capture conflict due to synchronization:
-  \begin{math}
-    {\reorderra}
-    =
-    \{(\aAct\Cb             \DW[\gemode\mREL]{}{}     )\Cc
-    (\aAct\Cb               \DF{\gemode\fREL}        )\Cc
-    (\DR{}{}\Cb             \DF{\gemode\fACQ}        )\Cc
-    % (\DR{\aLoc}{}\Cb        \DR[\gemode\mACQ]{\aLoc}{})\Cc
-    (\DR[\gemode\mACQ]{}{}\Cb\aAct                    )\Cc
-    (\DF{\gemode\fACQ}\Cb   \aAct                    )\Cc
-    (\DF{\gemode\fREL}\Cb   \DW{}{}                  )\Cc
-    (\DW[\gemode\mREL]{\aLoc}{}\Cb\DW{\aLoc}{})\}
-  \end{math}.
-
-  Let ${\eqreordersc}$ capture conflict due to $\mSC$ access:
-  \begin{math}
-    {\eqreordersc}
-    =
-    % \{(\DX[\mSC]{}{}\Cb\DX[\mSC]{}{})\}
-    \{(\DW[\mSC]{}{}\Cb \DW[\mSC]{}{})\Cc(\DR[\mSC]{}{}\Cb \DW[\mSC]{}{}) \Cc(\DW[\mSC]{}{}\Cb \DR[\mSC]{}{})\Cc(\DR[\mSC]{}{}\Cb \DR[\mSC]{}{})\}
-  \end{math}.
-
-  We say $\aAct \rdelaysdef \bAct$ if $\aAct\eqreorderco\bAct$ or $\aAct\reorderra\bAct$ or $\aAct\eqreordersc\bAct$.
-\end{definition}
-
-\subsection{The Semantic Domain} %: Pomsets with Predicate Transformers}
-\label{sec:pomsets}
-
-\emph{Predicate transformers} are functions on formulae that preserve
-logical structure, providing a natural model of sequential composition.
-The definition comes from \citet{DBLP:journals/cacm/Dijkstra75}:
-
-\begin{definition}
-  \label{def:trans}
-  A \emph{predicate transformer} is a %monotone
-  function
-  $\aTr{}{}:\Formulae\fun\Formulae$ such that
-  \begin{multicols}{2}
-    \begin{enumerate}[,label=(\textsc{x}\arabic*),ref=\textsc{x}\arabic*]
-    \item \label{tr-false}
-      $\aTr{}{\FALSE}$ is $\FALSE$,    
-    \item \label{tr-and}
-      $\aTr{}{\bForm_1\land\bForm_2}$ is $\aTr{}{\bForm_1}\land\aTr{}{\bForm_2}$,    
-    \item \label{tr-or}
-      $\aTr{}{\bForm_1\lor\bForm_2}$ is $\aTr{}{\bForm_1}\lor\aTr{}{\bForm_2}$, 
-    \item \label{tr-implies}
-      if $\aForm \rimplies \bForm$, then $\aTr{}{\aForm} \rimplies
-      \aTr{}{\bForm}$.
-    \end{enumerate}
-  \end{multicols}
-\end{definition}
-\noindent
-We consistently use $\bForm$ as the parameter of predicate transformers.
-Note that substitutions ($\bForm[\aExp/\aReg]$ and $\bForm[\aExp/\aLoc]$) and
-implications on the right ($\aForm\limplies\bForm$) are predicate
-transformers.
-
-As discussed in \S\ref{sec:intro}, predicate transformers suffice for
-sequentially consistent models, but not relaxed models, where dependency
-calculation is crucial.  For dependency calculation, we use a \emph{family}
-of predicate transformers, indexed by sets of events. In sequential
-composition, we will use $\aTr{\Cdown{\aEv}}{}$ as the predicate transformer
-applied to event $\aEv$ where $\bEv\in(\Cdown{\aEv})$ if $\bEv<\aEv$.
-
-\begin{definition}
-  \label{def:family}
-  A \emph{family of predicate transformers} over $\aEvs$ consists of a
-  predicate transformer $\aTr{\bEvs}{}$ for each $\bEvs\subseteq\AllEvents$,
-  such that if $\cEvs \cap \aEvs \subseteq \bEvs$ then $\aTr{\cEvs}{\bForm}
-  \rimplies \aTr{\bEvs}{\bForm}$.
-
-  We write $\aTr{}{\bForm}$ as an abbreviation of $\aTr{\aEvs}{\bForm}$.
-\end{definition}
-
-\begin{definition}
-  \label{def:pomset}
-  A \emph{pomset with predicate transformers} over $\Act$
-  is a tuple $(\aEvs, \labeling, \labelingForm, \aTr{}{}, \aTerm, {\rrfx}, {\le})$ where
-  \begin{enumerate}[,label=(\textsc{m}\arabic*),ref=\textsc{m}\arabic*]
-    \makecounter{BE}
-  \item \label{pom-E} \makecounter{E}
-    $\aEvs\subseteq\AllEvents$ is a set of \emph{events},
-    \makecounter{Blambda}
-  \item \label{pom-lambda} \makecounter{lambda}
-    $\labeling: \aEvs \fun \Act$ defines a \emph{label} for each event,
-    \makecounter{Bkappa}
-  \item \label{pom-kappa} \makecounter{kappa}
-    $\labelingForm:\aEvs\fun\Formulae$ defines a \emph{precondition} for each event, such that
-    \begin{enumerate}
-    \item \label{pom-kappa-sat}
-      $\labelingForm(\aEv)$ is satisfiable,
-    \end{enumerate}
-    \makecounter{Btau}
-  \item \label{pom-tau} \makecounter{tau}
-    $\aTr{}{}:2^{\AllEvents}\fun\Formulae \fun\Formulae$ is a \emph{family of predicate transformers} over $\aEvs$, 
-    \makecounter{Bterm}
-  \item \label{pom-term} \makecounter{term}
-    $\aTerm:\Formulae$ is a \emph{termination condition}, such that 
-    \begin{enumerate}
-    \item \label{pom-term-tau}
-      $\aTerm\rimplies\aTr{}{\TRUE}$,
-    \end{enumerate}
-    \makecounter{Brf}
-  \item \label{pom-rf} \makecounter{rf}
-    ${\rrfx} \subseteq \aEvs\times\aEvs$ is an injective relation capturing
-    \emph{reads-from}, such that 
-    % \end{enumerate}
-    % A pomset is a \emph{candidate} if there is an injective relation
-    % ${\rrfx} : \aEvs\times\aEvs$, capturing \emph{reads-from}, such that
-    \begin{enumerate}
-      % \begin{enumerate}[,label=(\textsc{i}\arabic*),ref=\textsc{i}\arabic*]
-      % \item \label{rf-injective}
-      %   if $\bEv\xrfx\aEv$ and $\cEv\xrfx\aEv$ then $\bEv=\cEv$, that is, ${\rrfx}$ is injective,
-    \item \label{pom-rf-match}
-      if $\bEv\xrfx\aEv$ then $\labeling(\bEv) \rmatches \labeling(\aEv)$,
-      % \item \label{rf-block}
-      %   if $\bEv\xrfx\aEv$ and $\labeling(\cEv) \rblocks \labeling(\aEv)$ then either $\cEv\le\bEv$ or $\aEv\le\cEv$.
-      % \item \label{rf-le}
-      %   if $\bEv\xrfx\aEv$ then $\bEv\le\aEv$.
-    \end{enumerate}
-    \makecounter{Ble}
-  \item \label{pom-le} \makecounter{le}
-    ${\le} \subseteq \aEvs\times\aEvs$, is a partial order capturing
-    \emph{causality}, such that
-    \begin{enumerate}
-    \item \label{pom-le-match}
-      if $\bEv\xrfx\aEv$ then either $\bEv\le\aEv$ or $\aEv\le\bEv$,
-    \item \label{pom-le-block}
-      if $\bEv\xrfx\aEv$ and $\labeling(\cEv) \rblocks \labeling(\aEv)$ then either $\cEv\le\bEv$ or $\aEv\le\cEv$.
-    \end{enumerate}
-  \end{enumerate}
-  A pomset is \emph{complete} if
-  % \begin{multicols}{2}
-  \begin{enumerate}[,label=(\textsc{c}\arabic*),ref=\textsc{c}\arabic*]
-    \setcounter{enumi}{\value{Blambda}}
-  \item \label{top-rf}
-    if $\labeling(\aEv)$ is a $\sread$ then there is some $\bEv\xrfx\aEv$,
-    \setcounter{enumi}{\value{Bkappa}}
-  \item \label{top-kappa}
-    $\labelingForm(\aEv)$ is a tautology (for every $\aEv\in\aEvs$),
-    \setcounter{enumi}{\value{Bterm}}
-  \item \label{top-term}
-    $\aTerm$ is a tautology.
-  \end{enumerate}
-  % \end{multicols}
-  % \labeltext[\textsc{t}1]{(\textsc{t}1)}{top-term} $\aTerm$ is a tautology and \labeltext[\textsc{t}2]{(\textsc{t}2)}{top-ev} for every $\aEv\in\aEvs$,
-  % % $\labelingForm(\aEv)$ is a tautology and
-  % % if $\labeling(\aEv)$ is a $\sread$ then there is some $\bEv\xrfx\aEv$.
-  % % \begin{multicols}{2}
-  % \begin{enumerate}[label=(\textsc{t}\arabic*),ref=\textsc{t}\arabic*]
-  %   \setcounter{enumi}{2}
-  % \item[]
-  %   \begin{enumerate}[leftmargin=0pt]
-  %   \item \label{top-kappa}
-  %     $\labelingForm(\aEv)$ is a tautology,    
-  %   \item \label{top-rf}
-  %     if $\labeling(\aEv)$ is a $\sread$ then there is some $\bEv\xrfx\aEv$.
-  %   \end{enumerate}
-  % \end{enumerate}
-  % % \end{multicols}
-\end{definition}
-\input{fig-sem.tex}
-
-We give the semantics of programs $\sembase{}$ in \reffig{fig:sem}.
-
-Let $\aPS$ range over pomsets, and $\aPSS$ over sets of pomsets.
-% Let $\Pom$ be the set of all pomsets.
-
-The model has seven components, which can be daunting at first glance.  To
-aid the reader, we use consistent numbering throughout. For example, item $7$
-always refers to the order relation.
-
-The core of the model is a pomset, which includes a set of events
-\eqref{pom-E}, a labeling \eqref{pom-lambda}, and an order \eqref{pom-le}.
-We also include the \emph{reads-from} relation explicitly in the model
-\eqref{pom-rf}.
-
-On top of this basic structure, \ref{pom-kappa}--\ref{pom-term} add a layer
-of logic.  For each pomset, \ref{pom-term} provides a termination condition.
-For each event in a pomset, \ref{pom-kappa} provides a precondition.  For
-each set of events in a pomset, \ref{pom-tau} provides a predicate
-transformer.  Sequential dependency is calculated by $\labelingForm'_2$ in
-the semantics of sequential composition.
-
-Before discussing the details of the model, we note that the semantics
-satisfies the expected monoid laws and is closed with respect to
-\emph{augmentation}.  Augments include more order and stronger formulae; in
-examples, we typically consider pomsets that are augment-minimal.  One
-intuitive reading of augment closure is that adding order can only cause
-preconditions to weaken.
-\begin{lemma}
-  \label{lem:monoid}
-  \begin{enumerate*}[label=(\alph*),ref=\alph*]
-  \item \label{monoid:unit}
-    \begin{math}
-      \aPSS
-      =
-      \xLPARP{\aPSS}{\sSKIP}
-      =
-      \xSEMIP{\aPSS}{\sSKIP}
-      =
-      \xSEMIP{\sSKIP}{\aPSS}.
-    \end{math}
-
-  \\\item \label{monoid:par}
-    \begin{math}
-      \xLPAR{\xLPARP{\aPSS_1}{\aPSS_2}}{\aPSS_3}
-      =
-      \xLPAR{\aPSS_1}{\xLPARP{\aPSS_2}{\aPSS_3}}.
-    \end{math}
-
-  \\\item \label{monoid:seq}
-    \begin{math}
-      \xSEMI{\xSEMIP{\aPSS_1}{\aPSS_2}}{\aPSS_3}
-      =
-      \xSEMI{\aPSS_1}{\xSEMIP{\aPSS_2}{\aPSS_3}}.
-    \end{math}
-
-  \\\item
-    \begin{math} \label{monoid:ifelse:if:if}
-      \xIFTHEN{\aForm}{\aPSS_1}{\aPSS_2}
-      =
-      \xSEMI{
-        \xIFTHEN{\aForm}{\aPSS_1}{}
-      }{
-        \xIFTHEN{\lnot\aForm}{\aPSS_2}{}
-      }
-      =
-      \xSEMI{
-        \xIFTHEN{\lnot\aForm}{\aPSS_2}{}
-      }{
-        \xIFTHEN{\aForm}{\aPSS_1}{}
-      }.
-    \end{math}
-
-  \\\item \label{monoid:if:dead}
-    \begin{math}
-      \xIFTHEN{\aForm}{\aPSS_1}{\aPSS_2}
-      =
-      \aPSS_1
-    \end{math}
-    if $\aForm$ is a tautology.
-
-  \\\item \label{monoid:if:if}
-    \begin{math}
-      \xIFTHEN{\aForm}{
-        \xIFTHEN{\bForm}{\aPSS}{}
-      }{}
-      =
-      \xIFTHEN{\aForm\land\bForm}{\aPSS}{}.
-    \end{math}
-
-  \\\item \label{monoid:if:seq}
-    \begin{math}
-      \xIFTHEN{\aForm}{
-        \xSEMI{\aPSS_1}{\aPSS_3}
-      }{
-        \xSEMI{\aPSS_2}{\aPSS_3}
-      }
-      \supseteq
-      \xSEMI{
-        \xIFTHEN{\aForm}{\aPSS_1}{\aPSS_2}
-      }{
-        \aPSS_3
-      }.
-    \end{math}
-
-  \\\item \label{monoid:seq:if}
-    \begin{math}
-      \xIFTHEN{\aForm}{
-        \xSEMI{\aPSS_1}{\aPSS_2}
-      }{
-        \xSEMI{\aPSS_1}{\aPSS_3}
-      }
-      \supseteq
-      \xSEMI{
-        \aPSS_1
-      }{
-        \xIFTHEN{\aForm}{\aPSS_2}{\aPSS_3}
-      }.
-    \end{math}
-
-  \\\item \label{monoid:if:elim}
-    \begin{math}
-      \xIFTHEN{\aForm}{\aPSS}{\aPSS}
-      \supseteq
-      \aPSS.
-    \end{math}
-  \end{enumerate*}      
-  \vspace{-.5\baselineskip}
-  \begin{proof}
-    Straightforward calculation.
-    \eqref{monoid:unit} requires \ref{pom-term-tau} for
-    the termination condition in $\xSEMIP{\aPSS}{\sSKIP}$.
-
-    \eqref{monoid:seq} requires both
-    conjunction closure (\ref{tr-and}, for the termination condition) and disjunction
-    closure (\ref{tr-or}, for the predicate transformers themselves).
-
-    \eqref{monoid:ifelse:if:if} requires \ref{seq-le-delays} not impose order
-    when $\labelingForm_1(\bEv) \land \labelingForm_2(\aEv)$ is
-    unsatisfiable, which in turn requires that $\labelingForm$ calculates
-    \emph{weakest} preconditions, rather than simple preconditions (see
-    \textsection\ref{sec:if:seq}).
-
-    \eqref{monoid:if:dead} requires \ref{pom-kappa-sat}.
-
-    In \textsection\ref{sec:semca}, we refine the semantics to validate the
-    reverse inclusions for \eqref{monoid:if:seq}, \eqref{monoid:seq:if}, and
-    \eqref{monoid:if:elim}.
-  \end{proof}
-\end{lemma}
-\begin{definition}
-  \label{def:augment}
-  $\aPS_2$ is an \emph{augment} of $\aPS_1$ if
-  \begin{multicols}{4}
-    \begin{enumerate}
-    \item $\aEvs_2=\aEvs_1$,
-    \item $\labelingAct_2(\aEv)=\labelingAct_1(\aEv)$,
-    \item $\labelingForm_2(\aEv) \riff \labelingForm_1(\aEv)$,
-    \item $\aTr[2]{\bEvs}{\bForm}\riff \aTr[1]{\bEvs}{\bForm}$,
-    \item $\aTerm[2] \riff \aTerm[1]$,
-    \item ${\rrfx_2}\supseteq{\rrfx_1}$,
-    \item ${\le_2}\supseteq{\le_1}$.
-    \end{enumerate}
-  \end{multicols}
-\end{definition}
-\begin{lemma}
-  % Suppose $\aPS_1\in\sembase{\aCmd}$.
-  If $\aPS_1\in\sembase{\aCmd}$ and $\aPS_2$  augments $\aPS_1$ then $\aPS_2\in\sembase{\aCmd}$.
-  % \item If $\aPS_2$ is a downset of $\aPS_1$ then $\aPS_2\in\sembase{\aCmd}$.
-  % \end{lemma}
-
-  \vspace{-.5\baselineskip}
-  \begin{proof}
-    Induction on the definition of $\sembase{}$.
-  \end{proof}
-\end{lemma}
-
-% \begin{definition}
-%   Suppose $\aRel_1:\aEvs_1\times\aEvs_1$ and $\aRel_2:\aEvs_2\times\aEvs_2$.
-%   We say $\aRel \rextendsdef{\aRel_1}{\aRel_2}$ if
-%   $\aRel\supseteq (\aRel_1\cup \aRel_2)$ and
-%   $\aRel\cap(\aEvs_1\times \aEvs_1) = \aRel_1$ and
-%   $\aRel\cap(\aEvs_2\times \aEvs_2) = \aRel_2$.
-% \end{definition}
-
-% We include empty sets as prep for adding while loops.
-
-% In diagrams, we use different colors for arrows.  We distinguish
-% $\rrfx$ edges that are included in order from those that are not.
-% \begin{itemize}
-% \item \makebox{$\aEv\xrf\bEv$} arises from $\rrfx$, where $\aEv\le\bEv$,
-% \item \makebox{$\aEv\xrfint\bEv$} arises from $\rrfx$, where $\lnot(\aEv\le\bEv)$.
-% \end{itemize}
-% To help the reader understand why order is included, we also different colors
-% for arrows induced by order.  We adopt the following conventions:
-% \begin{itemize}  
-%   % \item relaxed accesses are blue, with a single border,
-%   % \item synchronized accesses are red, with a double border,
-%   % \item \makebox{$\aEv\xrf\bEv$} arises from fulfillment, where $\aEv$ \emph{matches} $\bEv$,
-%   % \item \makebox{$\aEv\xwk\bEv$} arises either from fulfillment, where $\aEv$
-%   %   \emph{blocks} $\bEv$, or from prefixing, where $\aEv$ was prefixed before
-%   %   $\bEv$ and their actions \emph{conflict},
-% \item \makebox{$\aEv\xwk\bEv$} arises from \emph{fulfillment},
-% \item \makebox{$\aEv\xpo\bEv$} arises from control/data/address \emph{dependency},
-% \item \makebox{$\aEv\xsync\bEv$} arises from \emph{synchronized access}.
-% \end{itemize}
-
-\subsection{Pomsets}
-\label{sec:ex:pomset}
-
-We first explain the core of model, ignoring the logic (rules 3--5).
-%We defer the discussion of $\sLET{}{}$ to \textsection\ref{sec:ex:data}.
-We defer discussion of $\sIF{}$ to \textsection\ref{sec:ex:control}.
-
-Reads, writes, and fences map to pomsets with at most one event.  $\SKIP$
-maps to the empty pomset.  Ignoring the logic, the definitions are
-straightforward.  Note only that $\sembase{\PW{x}{1}}$ can write any value
-$\aVal$; the fact that $\aVal$ must be $1$ is captured in the logic (see
-\textsection\ref{sec:ex:term}).
-
-The structural rules combine pomsets:  Parallel composition is disjoint
-union, inheriting labeling, order and $\rrfx$ from the two sides.  Any
-$\rrfx$ edges added between the two sides must also be added to the order
-(\ref{par-le-rf1} and \ref{par-le-rf2}).
-% 
-Sequential composition is similar, with two changes: \ref{seq-E} does not
-require disjointness (see \textsection\ref{sec:ex:term}), and
-\ref{seq-le-delays} may require order (see example \ref{pub}, below).
-
-Note that reads-from implies order.
-\begin{lemma}
-  \label{lem:rf:implies:le}
-  For any $\aPS$ in the range of $\sembase{}$, $\bEv\xrfx\aEv$ implies
-  $\bEv\le\aEv$.
-
-  \vspace{-.5\baselineskip}
-  \begin{proof}
-    Induction on the definition of $\sembase{}$, using \ref{par-le-rf1}, \ref{par-le-rf2}, and \ref{seq-le-rf}.
-  \end{proof}
-\end{lemma}
-In top-level pomsets, every read must have a matching write in $\rrfx$
-\eqref{top-rf}.  Together with \ref{pom-le-match} and \ref{pom-le-block}, the lemma
-guarantees that reads are \emph{fulfilled} at top-level, as in
-\cite[\textsection 2.7]{DBLP:journals/pacmpl/JagadeesanJR20}.\footnote{The
-  basic model would be the same if we move $\rrfx$ from the model itself to
-  be existentially quantified in the definition of top-level pomset, along
-  with \ref{pom-le-match} and \ref{pom-le-block}.  This was the approach of
-  \citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} We include $\rrfx$
-  explicitly for use in \textsection\ref{sec:arm2}, where we introduce a
-  variant semantics $\frf{\semrr{}}$ for which \reflem{lem:rf:implies:le}
-  fails.}
-
-From \refdef{def:actions}, recall that $\aAct \rdelaysdef \bAct$ if
-$\aAct\eqreorderco\bAct$ or $\aAct\reorderra\bAct$ or
-$\aAct\eqreordersc\bAct$.  \ref{seq-le-delays} guarantees that sequential
-order is enforced between conflicting accesses of the same location
-($\eqreorderco$), into a release and out of an acquire ($\reorderra$), and
-between SC accesses ($\eqreordersc$).  Combined with the fulfillment
-requirements (\ref{pom-le-match}, \ref{pom-le-block} and \reflem{lem:rf:implies:le}),
-these ensure coherence, publication, subscription and other idioms.  For
-example, consider the following:\footnote{We use different colors for arrows representing order:
-  \begin{itemize}  
-  \item \makebox{$\bEv\xpo\aEv$} arises from control/data/address \emph{dependency} (\ref{seq-kappa}, definition of $\labelingForm'_2(\bEv)$),
-  \item \makebox{$\bEv\xsync\aEv$} arises from $\reorderra$ or $\eqreordersc$ \eqref{seq-le-delays},
-  \item \makebox{$\bEv\xwki\aEv$} arises from $\eqreorderco$ \eqref{seq-le-delays},
-  \item \makebox{$\bEv\xrf\aEv$} arises from \emph{reads-from} \eqref{pom-le-match},
-  %\item \makebox{$\bEv\xsyncsc\aEv$} arises from \emph{strong fencing} \eqref{cand-lesync-sc},
-  \item \makebox{$\bEv\xwk\aEv$} arises from \emph{blocking} \eqref{pom-le-block}.
-  \end{itemize}
-  In \textsection\ref{sec:arm2}, it is possible for $\rrf$ to contradict
-  $\le$.  In this case, we use a dotted arrow for $\rrf$: $\bEv\xrfint\aEv$
-  indicates that $\aEv\le\bEv$.}
-\begin{gather*}
-  \taglabel{pub}
-  \begin{gathered}    
-    \PW{x}{0}\SEMI 
-    \PW{x}{1}\SEMI \PW[\mREL]{y}{1} \PAR \PR[\mACQ]{y}{r}\SEMI \PR{x}{s}
-    \\[-.4ex]
-    \nonumber
-    \hbox{\begin{tikzinline}[node distance=1.5em]
-        \event{wx0}{\DW{x}{0}}{}
-        \event{wx1}{\DW{x}{1}}{right=of wx0}
-        \event{wy1}{\DW[\mREL]{y}{1}}{right=of wx1}
-        \event{ry1}{\DR[\mACQ]{y}{1}}{right=2.5em of wy1}
-        \event{rx0}{\DR{x}{0}}{right=of ry1}
-        \sync{wx1}{wy1}
-        \sync{ry1}{rx0}
-        \rf{wy1}{ry1}
-        \wk[out=170,in=10]{rx0}{wx1}
-        \wki{wx0}{wx1}
-      \end{tikzinline}}
-  \end{gathered}
-\end{gather*}
-The execution is disallowed due to the cycle.  All of the order shown is
-required at top-level: The intra-thread order comes from \ref{seq-le-delays}:
-$\DWP{x}{0}\xwki \DWP{x}{1}$ is required by $\eqreorderco$.
-$\DWP{x}{1}\xsync \DWP[\mREL]{y}{1}$ and $\DRP[\mACQ]{y}{1}\xsync \DRP{x}{0}$ are
-required by $\reorderra$.  The cross-thread order is required by fulfillment:
-\ref{top-rf} requires that all top-level reads are in the image of $\xrfx$.
-\ref{pom-le-match} ensures that $\DWP[\mREL]{y}{1}\xrfx \DRP[\mACQ]{y}{1}$, and
-\reflem{lem:rf:implies:le} subsequently ensures that
-$\DWP[\mREL]{y}{1}\le \DRP[\mACQ]{y}{1}$.  The \emph{antidependency}
-$\DRP{x}{0}\xwk \DWP{x}{1}$ is required by \ref{pom-le-block}.  (Alternatively,
-we could have $\DWP{x}{1}\xwk \DWP{x}{0}$, again resulting in a cycle.)
-
-The semantics gives the expected results for store buffering and load buffering,
-as well as litmus tests involving fences and SC access.  The model of
-coherence is weaker than C11, in order to support common subexpression
-elimination, and stronger than Java, in order to support local reasoning
-about data races.  See
-\cite[\textsection3.1]{DBLP:journals/pacmpl/JagadeesanJR20} for a discussion.
-
-In the structural rules $\sSEMI{}{}$ and $\sIF{}$, we say that
-$\bEv\in\aEvs_1$ and $\aEv\in\aEvs_2$ \emph{coalesce} if $\bEv=\aEv$.
-
-\ref{seq-E} allows \emph{mumbling} \cite{DBLP:journals/iandc/Brookes96} by
-coalescing events.  For example
-\begin{math}
-  \sembase{\PW{x}{1} \SEMI \PW{x}{1}}
-\end{math}
-includes the singleton pomset
-\begin{tikzinlinesmall}[node distance=.5em and 1.5em]
-  \event{a1}{\DW{x}{1}}{}
-\end{tikzinlinesmall}.  From this it is easy to see that
-\begin{math}
-  \sembase{\PW{x}{1}\SEMI\PW{x}{1}}
-  \supseteq
-  \sembase{\PW{x}{1}}
-\end{math}
-is a valid refinement.  It is equally obvious that
-\begin{math}
-  \sem{\PW{\aLoc}{1}} 
-  \not\supseteq
-  \sem{\PW{\aLoc}{1} \SEMI \PW{\aLoc}{1}}
-\end{math}
-is not a valid refinement, since the latter includes a two-element pomset,
-but the former does not.\footnote{These are distinguished by the context:
-  \begin{math}
-    \hole{} \PAR
-    \PR{x}{r} \SEMI
-    \PW{x}{2} \SEMI
-    \PR{x}{s}\SEMI
-    \IF{\aReg {=} \bReg} \THEN \PW{\cLoc}{1} \FI.
-  \end{math}}
-
-\subsection{Termination}
-\label{sec:ex:term}
-
-In top-level pomsets, \ref{top-term} requires that $\aTerm$ is a tautology,
-capturing termination.  Terminated pomsets are often called \emph{complete},
-whereas nonterminated pomsets are \emph{incomplete}.
-
-Ignoring predicate transformers, the structural rules, \ref{par-term} and
-\ref{seq-term}, take $\aTerm$ to be $\aTerm_1\land\aTerm_2$.  This is as
-expected: the program terminates if both subprograms terminate.
-
-The interesting rules are $\sLOAD{}{}$, $\sFENCE{}$, and $\sSTORE{}{}$.
-
-In $\sLOAD{}{}$, there is no restriction on $\aTerm$ for relaxed reads.  From
-this, it is easy to see that
-\begin{math}
-  \sembase{\PR{x}{r}}
-  \supseteq
-  \sembase{\SKIP}
-\end{math}
-is a valid refinement (where the default mode is $\mRLX$).
-
-In $\sFENCE{}$, instead, \ref{fence-term} ensures that all fences are included at
-top-level.  This also ensures
-$\sembase{\PF{\amode}}\not\supseteq\sembase{\IF{\aExp}\THEN\PF{\amode}\FI}$,
-since $\sembase{\IF{\aExp}\THEN\PF{\amode}\FI}$ includes the empty set with
-termination condition $\lnot\aExp$, but $\sembase{\PF{\amode}}$ can only
-include the empty set with termination condition $\FALSE$.
-
-In $\sSTORE{}{}$, \ref{write-term-empty} is similar.  In addition,
-\ref{write-term-nonempty} ensures that top-level pomsets do not include bogus
-writes.  Suppose $\aPS\in\sembase{\PW{x}{1}}$.  As we noted above, $\aPS$ can
-include $(1{=}\aVal\mid\DW{x}{\aVal})$, for any value $\aVal$.  At top-level,
-however, \ref{write-term-nonempty} requires that $\aTerm$ implies
-$1{=}\aVal$.  In this case, \ref{pom-kappa-sat} would filter the pomset,
-since preconditions must be satisfiable.  However, unsatisfiable writes can be become satisfiable via
-merging:
-% Nonetheless, it is important that
-% \ref{write-term-nonempty} specifies $\aTerm\riff\aExp{=}\aVal$ rather than
-% $\aTerm\riff\TRUE$, since 
-\begin{align*}
-  \begin{gathered}
-    \PW{x}{1}
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \event{a}{\DW{x}{1}}{}      
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-  \begin{gathered}
-    \PW{x}{2}
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \event{b}{2{=}3\mid\DW{x}{3}}{}      
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-  \begin{gathered}
-    \IF{\aExp}\THEN\PW{x}{3}\FI
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \event{c}{\aExp\mid\DW{x}{3}}{}      
-      \end{tikzinline}}    
-  \end{gathered}
-\end{align*}
-By merging, the semantics allows the following:
-\begin{align*}
-  \begin{gathered}
-    \PW{x}{1}
-    \SEMI
-    \PW{x}{2}
-    \SEMI
-    \IF{\aExp}\THEN\PW{x}{3}\FI
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \event{a}{\DW{x}{1}}{}      
-        \event{c}{\aExp\mid\DW{x}{3}}{right=of a}
-        \wki{a}{c}
-      \end{tikzinline}}    
-  \end{gathered}
-\end{align*}
-This pomset is incomplete, however, since $\aTerm\riff2{=}3$.
-
-\subsection{Data Dependencies, Preconditions, and Predicate Transformers}
-\label{sec:ex:data}
-
-% Predicate transformers affect preconditions in the definition of
-% $\labelingForm'_2$, which is used in \ref{seq-kappa2} and \ref{seq-kappa12}
-% to calculate dependencies.  In this subsection we ignore predicate
-% transformers, taking $\labelingForm'_2=\labelingForm_2$.
-
-In top-level pomsets, \ref{top-kappa} requires that every precondition
-$\labelingForm(\aEv)$ is a tautology.
-
-Preconditions are discharged during sequential composition by applying
-predicate transformers $\aTr[1]{}{}$ from the left to preconditions
-$\labelingForm_2(\aEv)$ on the right.  The specific rules are
-\ref{seq-kappa2} and \ref{seq-kappa12}, which use the transformed predicate
-$\labelingForm'_2(\aEv)=\aTr[1]{\Cdown{\aEv}}{\labelingForm_2(\aEv})$, where
-$\Cdown{\aEv}=\{ \cEv \mid \cEv \lt \aEv \}$ is the set of events that
-precede $\aEv$ in causal order.  We call $\Cdown{\aEv}$ the \emph{dependent
-  set} for $\aEv$.  Then $\aEvs\setminus(\Cdown{\aEv})$ is the
-\emph{independent set}. % for $\aEv$.
-
-% Sequential composition uses the usual rule for composition of predicate
-% transformers (but preserving the indexing set). For the pomset, we take the
-% union of their events, preserving actions, but crucially in
-% cases~\ref{seq-kappa2} and~\ref{seq-kappa12} we apply a predicate transformer
-% $\aTr[1]{\cEvs}{}$ from the left-hand side to a precondition
-% $\labelingForm_2(\aEv)$ from the right-hand side to build the precondition
-% $\labelingForm'_2(\aEv)$.  The indexing set $\cEvs$ for the predicate
-% transformer is $\{ \cEv\mid \cEv<\aEv \}$, so can depend on the causal order.
-
-Before looking at the details, it is useful to have a high-level view of how
-nontrivial preconditions and predicate transformers are introduced.  (We
-discuss address dependencies in \textsection\ref{sec:addr}.)
-\begin{multicols}{2}
-  Preconditions are introduced in: 
-  \begin{itemize}
-  \item[\eqref{seq-kappa-release}] for release actions,
-  \item[\eqref{if-kappa}] for control dependencies, 
-  \item[\eqref{write-kappa}] for data dependencies on writes.
-  \end{itemize}
-
-  Predicate transformers are introduced in:
-  \begin{itemize}
-  \item[\eqref{read-tau-dep}] for reads in the dependent set,
-  \item[\eqref{read-tau-ind}] for reads in the independent set,
-  \item[\eqref{write-tau}] for writes.
-  \end{itemize}
-  % \item These are glued together in \ref{seq-kappa}.
-  %   \begin{itemize}
-  %   \item \ref{seq-kappa2}  handles non-coalescing events,
-  %   \item \ref{seq-kappa12} handles coalescing events.
-  %   \end{itemize}
-\end{multicols}
-
-The rules track dependencies.  We discuss data dependencies \eqref{write-kappa}
-here and control dependencies \eqref{if-kappa} in
-\textsection\ref{sec:ex:control}.  Unless otherwise noted, we assume pomsets
-are \emph{complete} and \emph{augment-minimal}.  We do not discuss
-\ref{seq-kappa-release} further.  It simply ensures that all writes are
-present before a release, even for incomplete pomsets (see
-\textsection\ref{sec:ex:term}).
-
-A simple example of a data dependency is a pomset
-$\aPS\in\sem{\PR{x}{r}\SEMI \PW{y}{r}}$.  If $\aPS$ is complete, it must have
-two events. Then $\sSEMI{}{}$ requires that there are %$\aVal,\bVal\in\Val$ and
-$\aPS_1\in \sem{\PR{x}{r}}$ and $\aPS_2\in \sem{\PW{y}{r}}$ of the form:
-\begin{align*}
-  \tag{$\ddagger$}\label{eq2}
-  \begin{gathered}
-    \PR{x}{r} 
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \xform{xd}{\aVal{=}r\limplies\bForm}{}
-        \eventr{\bEv}{a1}{\DR{x}{\aVal}}{left=of xd}
-        \xform{xi}{(x{=}r\lor\aVal{=}r)\limplies\bForm}{left=.5em of a1}
-        \xo{a1}{xd}
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-  \begin{gathered}
-    \PW{y}{r}
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \xform{xd}{\bForm[r/y]}{}
-        \eventr{\aEv}{a2}{r{=}\bVal\mid\DW{y}{\bVal}}{left=of xd}      
-        \xform{xi}{\bForm[r/y]}{left=.5em of a2}
-        \xo{a2}{xd}
-      \end{tikzinline}}    
-  \end{gathered}
-\end{align*}
-First we consider the case that $\aVal=\bVal$.  For example if $\aVal=\bVal=1$, we have:
-\begin{align*}
-  \begin{gathered}
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \xform{xd}{1{=}r\limplies\bForm}{}
-        \eventr{\bEv}{a1}{\DR{x}{1}}{left=of xd}
-        \xform{xi}{(x{=}r\lor1{=}r)\limplies\bForm}{left=.5em of a1}
-        \xo{a1}{xd}
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-  \begin{gathered}
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \xform{xd}{\bForm[r/y]}{}
-        \eventr{\aEv}{a2}{r{=}1\mid\DW{y}{1}}{left=of xd}      
-        \xform{xi}{\bForm[r/y]}{left=.5em of a2}
-        \xo{a2}{xd}
-      \end{tikzinline}}    
-  \end{gathered}
-\end{align*}
-For the read, the dependent transformer $\smash{\aTr[1]{\{\bEv\}}{}}$ is
-$1{=}r\limplies\bForm$; the independent transformer $\aTr[1]{\emptyset}{}$ is
-$(x{=}r\lor1{=}r)\limplies\bForm$.  These are determined by
-\ref{read-tau-dep} and \ref{read-tau-ind}, respectively.  For the write, both
-% the dependent transformer
-$\smash{\aTr[2]{\{\aEv\}}{}}$ and
-% the independent transformer
-$\aTr[2]{\emptyset}{}$ are $\bForm[r/y]$, as are determined by
-\ref{write-tau}.
-%
-Combining these into a single pomset, we have:
-\begin{align*}
-  \begin{gathered}
-    \PR{x}{r}\SEMI \PW{y}{r}
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \xform{xd}{1{=}r\limplies\bForm[r/y]}{}
-        \eventr{\bEv}{a1}{\DR{x}{1}}{left=of xd}
-        \xform{xi}{(x{=}r\lor1{=}r)\limplies\bForm[r/y]}{left=.5em of a1}
-        \eventr{\aEv}{a2}{\aForm\mid\DW{y}{1}}{right=of xd}      
-        \xo{a1}{xd}
-      \end{tikzinline}}    
-  \end{gathered}
-\end{align*}
-By \ref{seq-tau}, predicate transformers are determined by composition; thus
-$\aTr{\bEvs}{\bForm}$ is $\aTr[1]{\bEvs}{\aTr[2]{\bEvs}{\bForm}}$.  Since the
-transformer does not depend on whether the write is included, we do not draw
-dependencies for the write in the diagram.
-
-Turning to the precondition $\aForm$ on the write, recall that in order for
-$\aEv$ to participate in a top-level pomset, the precondition $\aForm$ must
-be a tautology at top-level.  There are two possibilities.
-\begin{itemize}
-\item If $\bEv\le\aEv$ then we apply the dependent transformer and
-  $\aForm=(1{=}r\limplies r{=}1)$, a tautology.
-\item If $\bEv\not\le\aEv$ then we apply the independent transformer and
-  $\aForm=((x{=}r\lor1{=}r)\limplies r{=}1)$.  Under the assumption that $r$
-  is bound, this is logically equivalent to $(x{=}1)$. (We make this more
-  precise in \textsection\ref{sec:semreg}.)
-\end{itemize}
-Eliding transformers, the two outcomes are:
-\begin{align*}
-  \begin{gathered}
-    \PR{x}{r}\SEMI \PW{y}{r}
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \eventr{\bEv}{a1}{\DR{x}{1}}{}
-        \eventr{\aEv}{a2}{\DW{y}{1}}{right=of a1}
-        \po{a1}{a2}
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-  \begin{gathered}
-    \PR{x}{r}\SEMI \PW{y}{r}
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \eventr{\bEv}{a1}{\DR{x}{1}}{}
-        \eventr{\aEv}{a2}{x{=}1\mid\DW{y}{1}}{right=of a1}
-      \end{tikzinline}}    
-  \end{gathered}
-\end{align*}
-The independent case on the right can only participate in a top-level pomset
-if the precondition $(x{=}1)$ is discharged.  To do so, we must prepend a
-pomset $\aPS_0$ that writes $1$ to $x$:
-\begin{align*}
-  \begin{gathered}
-    \PW{x}{1}
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \xform{xd}{\bForm[1/x]}{}
-        \eventr{\cEv}{a2}{1{=}1\mid\DW{x}{1}}{left=of xd}      
-        \xform{xi}{\bForm[1/x]}{left=.5em of a2}
-        \xo{a2}{xd}
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-  \begin{gathered}
-    \PW{x}{1}\SEMI \PR{x}{r}\SEMI \PW{y}{r}
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \eventr{\cEv}{a0}{1{=}1\mid\DW{x}{1}}{}
-        \eventr{\bEv}{a1}{\DR{x}{1}}{right=of a0}
-        \eventr{\aEv}{a2}{1{=}1\mid\DW{y}{1}}{right=of a1}
-      \end{tikzinline}}    
-  \end{gathered}
-\end{align*}
-Here we apply the predicate transformer $\aTr[0]{\emptyset}{}$ to $(x{=}1)$,
-resulting in the tautology $(1{=}1)$.
-
-Now suppose that $\aVal\neq\bVal$ in \eqref{eq2}.  Again there are two
-possibilities, where we take $\aVal=0$ and $\bVal=1$:
-\begin{align*}
-  \begin{gathered}
-    \PR{x}{r}\SEMI \PW{y}{r}
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \eventr{\bEv}{a1}{\DR{x}{0}}{}
-        \eventr{\aEv}{a2}{0{=}r\limplies r{=}1\mid\DW{y}{1}}{right=of a1}
-        \po{a1}{a2}
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-  \begin{gathered}
-    \PR{x}{r}\SEMI \PW{y}{r}
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \eventr{\bEv}{a1}{\DR{x}{0}}{}
-        \eventr{\aEv}{a2}{(x{=}r\lor0{=}r)\limplies r{=}1\mid\DW{y}{1}}{right=of a1}
-      \end{tikzinline}}    
-  \end{gathered}
-\end{align*}
-Assuming that $r$ is bound, both preconditions on $\aEv$ are unsatisfiable.
-
-If a write is independent of a read, then clearly no order is imposed between
-them.  For example, the precondition of $\aEv$ is a tautology in:
-\begin{align*}
-  \begin{gathered}
-    \PR{x}{r}\SEMI \PW{y}{1}
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \xform{xd}{0{=}r\limplies\bForm[r/y]}{}
-        \eventr{\bEv}{a1}{\DR{x}{0}}{left=of xd}
-        \xform{xi}{(x{=}r\lor0{=}r)\limplies\bForm[r/y]}{left=.5em of a1}
-        \eventr {\aEv}{a2}{(x{=}r\lor0{=}r)\limplies 1{=}1\mid\DW{y}{1}}{right=of xd}      
-        \xo{a1}{xd}
-      \end{tikzinline}}    
-  \end{gathered}
-\end{align*}
-% \begin{gather*}
-%   \PR{x}{r}\SEMI \PW{y}{1}
-%   \\
-%   \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
-%       \event{rx0}{\DR{x}{0}}{}
-%       \event{wy1}{1{=}1 \mid \DW{y}{1}}{right=of rx0}
-%     \end{tikzinline}}
-% \end{gather*}
-
-\subsection{Control Dependencies}
-\label{sec:ex:control}
-
-In $\sIF{\aForm}\sTHEN\aPSS_1\sELSE\aPSS_2\sFI$, the predicate transformer
-\eqref{if-tau} is
-$(\aForm\land\aTr[1]{\bEvs}{\bForm})\lor(\neg\aForm\land\aTr[2]{\bEvs}{\bForm})$,
-%$\aTerm \rimplies (\aForm\land\aTerm[1])\lor(\neg\aForm\land\aTerm[2])$,
-which is the disjunctive equivalent of \citeauthor{DBLP:journals/cacm/Dijkstra75}'s
-conjunctive formulation: $(\aForm\limplies\aTr[1]{\bEvs}{\bForm})\land(\neg\aForm\limplies\aTr[2]{\bEvs}{\bForm})$.
-% $\aTerm \rimplies (\aForm\limplies\aTerm[1])\land(\neg\aForm\limplies\aTerm[2])$.
-%  \citet{DBLP:journals/cacm/Dijkstra75} used conjunctive normal form.
-%  Recall that
-%  \href{https://www.wolframalpha.com/input/?i=\%28a+and+b\%29+or+\%28not+a+and+c\%29}{$(\aForm\land\cForm_1)\lor(\neg\aForm\land\cForm_2)$}
-%  is logically equivalent to
-%  \href{https://www.wolframalpha.com/input/?i=\%28a+implies+b\%29+and+\%28\%28not+a\%29+implies+c\%29}{$(\aForm\limplies\cForm_1)\land(\neg\aForm\limplies\cForm_2)$}.
-
-This semantics validates dead code elimination:
-if $\aExp{\neq}0$ is a tautology then
-\begin{math}
-  \sem{\IF{\aExp}\THEN\aCmd_1\ELSE\aCmd_2\FI} \supseteq
-  \sem{\aCmd_1}.
-\end{math}
-The reverse inclusion does not hold.
-
-For events from $\aEvs_1$, \ref{if-kappa1} requires
-$\aForm\land\labelingForm_1(\aEv)$. For events from $\aEvs_2$,
-\ref{if-kappa2} requires $\lnot\aForm\land\labelingForm_2(\aEv)$.  For
-coalescing events in $\aEvs_1\cap\aEvs_2$, \ref{if-kappa12} requires
-$(\aForm\land\labelingForm_1(\aEv)) \lor
-(\lnot\aForm\land\labelingForm_2(\aEv))$.  This semantics allows common code
-to be lifted out of a conditional, validating the transformation
-\begin{math}
-  \sem{\IF{\aExp}\THEN\aCmd\ELSE\aCmd\FI} \supseteq
-  \sem{\aCmd}.
-\end{math}
-%
-The use of $\rextendsdef{}{}$ in \ref{if-le-extends} and \ref{if-rf-extends}
-ensures that no new order is introduced between events in
-$\aEvs_1\cap\aEvs_2$ when coalescing; see \textsection\ref{sec:arm2}.
-
-By allowing events to coalesce, \ref{if-kappa12} ensures that control
-dependencies are calculated semantically.  For example, consider
-$\aPS\in\sem{\IF {r{=}1} \THEN \PW{y}{r} \ELSE \PW{y}{1}\FI}$, which is build
-from $\aPS_1\in\sem{\PW{y}{r}}$ and $\aPS_2\in\sem{\PW{y}{1}}$ such as:
-\begin{align*}
-  \begin{gathered}
-    \PW{y}{r}
-    \\
-    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
-        \eventr{e}{wy1}{r{=}1 \mid \DW{y}{1}}{}
-      \end{tikzinline}}
-  \end{gathered}
-  &&
-  \begin{gathered}
-    \PW{y}{1}
-    \\
-    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
-        \eventr{e}{wy1}{1{=}1 \mid \DW{y}{1}}{}
-      \end{tikzinline}}
-  \end{gathered}
-  &&
-  \begin{gathered}
-    \IF{r{=}1} \THEN \PW{y}{r} \ELSE \PW{y}{1}\FI
-    \\
-    \nonumber
-    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
-        \eventr{e}{wy1}{(r{=}1 \limplies r{=}1) \land (r{\ne}1 \limplies 1{=}1) \mid \DW{y}{1}}{}
-      \end{tikzinline}}
-  \end{gathered}  
-\end{align*}
-Here, the precondition in the combined pomset is a tautology, independent of $r$.
-
-Control dependencies are eliminated in the same way as data dependencies.
-For example:
-\begin{align*}
-  \begin{gathered}
-    \PR{x}{r} 
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \xform{xd}{\aVal{=}r\limplies\bForm}{}
-        \eventr{\bEv}{a1}{\DR{x}{\aVal}}{left=of xd}
-        \xform{xi}{(x{=}r\lor\aVal{=}r)\limplies\bForm}{left=.5em of a1}
-        \xo{a1}{xd}
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-  \begin{gathered}
-    \IF{r{=}1}\THEN\PW{y}{1}\FI
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \xform{xd}{r{=}1\limplies\bForm[1/y]}{}
-        \eventr{\aEv}{a2}{r{=}1\mid\DW{y}{\bVal}}{left=of xd}      
-        \xform{xi}{r{=}1\limplies\bForm[1/y]}{left=.5em of a2}
-        \xo{a2}{xd}
-      \end{tikzinline}}    
-  \end{gathered}
-\end{align*}
-Reasoning as we did for \eqref{eq2} in \textsection\ref{sec:ex:data}, there are two possibilities:
-\begin{align*}
-  \begin{gathered}
-    \PR{x}{r}\SEMI \IF{r{=}1}\THEN\PW{y}{1}\FI
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \eventr{\bEv}{a1}{\DR{x}{1}}{}
-        \eventr{\aEv}{a2}{\DW{y}{1}}{right=of a1}
-        \po{a1}{a2}
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-  \begin{gathered}
-    \PR{x}{r}\SEMI \IF{r{=}1}\THEN\PW{y}{1}\FI
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \eventr{\bEv}{a1}{\DR{x}{1}}{}
-        \eventr{\aEv}{a2}{x{=}1\mid\DW{y}{1}}{right=of a1}
-      \end{tikzinline}}    
-  \end{gathered}
-\end{align*}
-
-As another example, consider JMM causality test case 1 \cite{PughWebsite}:
-\begin{gather*}
-  \begin{gathered}
-    \PW{x}{0} \SEMI
-    (\PR{x}{r}\SEMI\IF{r{\geq}0}\THEN \PW{y}{1} \FI
-    \PAR
-    \PW{x}{y})
-    \\[-1ex]
-    \hbox{\begin{tikzinline}[node distance=1.5em]
-        \event{wx0}{\DW{x}{0}}{}
-        \event{rx1}{\DR{x}{1}}{right=3em of wx0}
-        \event{wy1}{\aForm\mid\DW{y}{1}}{right=of rx1}
-        \event{ry1}{\DR{y}{1}}{right=3em of wy1}
-        \event{wx1}{\DW{x}{1}}{right=of ry1}
-        \po{ry1}{wx1}
-        \rf[out=-168,in=-12]{wx1}{rx1}
-        \rf{wy1}{ry1}
-        \wk[out=10,in=170]{wx0}{wx1}
-        \wki{wx0}{rx1}
-      \end{tikzinline}}
-    % \hbox{\begin{tikzinline}[node distance=1.5em and 2em]
-    %   \event{wy1}{0\geq0\mid\DW{y}{1}}{}
-    %   \event{rx1}{\DR{x}{1}}{left=of wy1}
-    %   \event{wx0}{\DW{x}{0}}{below left=-.2em and 2em of rx1}
-    %   \event{ry1}{\DR{y}{1}}{below=of rx1}
-    %   \event{wx1}{\DW{x}{1}}{right=of ry1}
-    %   \po{ry1}{wx1}
-    %   \rf[out=140,in=-20]{wx1}{rx1}
-    %   \rf{wy1}{ry1}
-    %   \wk[out=10,in=155]{wx0}{wx1}
-    %   \wk{wx0}{rx1}
-    % \end{tikzinline}}
-  \end{gathered}
-\end{gather*}
-The precondition $\aForm$ is
-\begin{math}
-  \PBR{(1{=}r \lor x{=}r) \limplies r{\geq}0}[0/x]
-\end{math}
-which is 
-\begin{math}
-  \PBR{(1{=}r \lor 0{=}r) \limplies r{\geq}0}
-\end{math}
-which is a tautology.
-
-\subsection{Reordering Transformations}
-\label{sec:ex:valid}
+% \subsection{Reordering Transformations}
+% \label{sec:ex:valid}
 
@@
 % \end{align*}
-
-\subsection{Conditional and Coherence}
-\label{sec:if:seq}
-
-[This is out of date.]
-
-\begin{lemma}
-  \label{lem:if:seq}
-  \begin{math}
-    \xIFTHEN{\aForm}{\aPSS_1}{\aPSS_2}
-    \supseteq
-    \xSEMI{
-      \xIFTHEN{\aForm}{\aPSS_1}{}
-    }{
-      \xIFTHEN{\lnot\aForm}{\aPSS_2}{}
-    }
-  \end{math}
-
-  \begin{math}
-    \xIFTHEN{\aForm}{\aPSS_1}{\aPSS_2}
-    \supseteq
-    \xSEMI{
-      \xIFTHEN{\lnot\aForm}{\aPSS_2}{}
-    }{
-      \xIFTHEN{\aForm}{\aPSS_1}{}
-    }
-  \end{math}
-\end{lemma}
-
-Reverse direction does not hold, due to \ref{seq-le-delays}.
-\begin{enumerate}
-\item[\eqref{seq-le-delays}] if
-  $\labeling_1(\bEv) \rdelays \labeling_2(\aEv)$ then $\bEv\le\aEv$.  
-\end{enumerate}
- An alternate phrasing might be attractive:
-\begin{enumerate}
-\item[{\labeltext[\textsc{s}7c$'$]{(\textsc{s}7c$'$)}{seq-le-delays'}}] if
-  $\labeling_1(\bEv) \rdelays \labeling_2(\aEv)$ and
-  $\labelingForm(\bEv)\land\labelingForm(\aEv)$ is satisfiable then
-  $\bEv\le\aEv$.
-\end{enumerate}
-But \ref{seq-le-delays'} is incompatible with the ability to strengthen
-preconditions using augment closure.  Consider the following.
-\begin{align*}
-  \begin{gathered}[t]
-    \IF{r}\THEN\PW{x}{2}\FI
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \event{a1}{r{\neq}0\mid\DW{x}{2}}{}
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-  \begin{gathered}[t]
-    \PW{x}{1}
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \event{a2}{            \DW{x}{1}}{}
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-  \begin{gathered}[t]
-    \PW{x}{2}
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \event{a3}{            \DW{x}{2}}{}
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-  \begin{gathered}[t]
-    \IF{\BANG r}\THEN\PW{x}{1}\FI
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \event{a4}{r{=}0   \mid\DW{x}{1}}{}
-      \end{tikzinline}}    
-  \end{gathered}
-\end{align*}
-% \begin{align*}
-%   \begin{gathered}[t]
-%     \IF{r}\THEN\PW{x}{2}\FI
-%     \SEQ
-%     \PW{x}{1}
-%     \SEQ
-%     \PW{x}{2}
-%     \SEQ
-%     \IF{\BANG r}\THEN\PW{x}{1}\FI
-%     \\
-%     \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-%         \event{a1}{r{\neq}0\mid\DW{x}{2}}{}
-%         \event{a2}{            \DW{x}{1}}{right=of a1}
-%         \event{a3}{            \DW{x}{2}}{right=of a2}
-%         \event{a4}{r{=}0   \mid\DW{x}{1}}{right=of a3}
-%       \end{tikzinline}}    
-%   \end{gathered}
-% \end{align*}
-Augmenting the middle preconditions and then using sequential composition, we have:
-\begin{align*}
-  \begin{gathered}[t]
-    \IF{r}\THEN\PW{x}{2}\FI
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \event{a1}{r{\neq}0\mid\DW{x}{2}}{}
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-  \begin{gathered}[t]
-    \PW{x}{1}
-    \SEQ
-    \PW{x}{2}
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \event{a2}{r{\neq}0\mid\DW{x}{1}}{}
-        \event{a3}{r{=}0   \mid\DW{x}{2}}{right=of a1}
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-  \begin{gathered}[t]
-    \IF{\BANG r}\THEN\PW{x}{1}\FI
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \event{a4}{r{=}0   \mid\DW{x}{1}}{}
-      \end{tikzinline}}    
-  \end{gathered}
-\end{align*}
-Note that \ref{seq-le-delays'} does not require any order between the two
-writes of the middle pomset.
-% \begin{align*}
-%   \begin{gathered}[t]
-%     \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-%         \event{a1}{r{\neq}0\mid\DW{x}{2}}{}
-%         \event{a2}{r{=}0   \mid\DW{x}{1}}{right=of a1}
-%         \event{a3}{r{\neq}0\mid\DW{x}{2}}{right=of a2}
-%         \event{a4}{r{=}0   \mid\DW{x}{1}}{right=of a3}
-%       \end{tikzinline}}    
-%   \end{gathered}
-% \end{align*}
-Merging left and right, we have:
-\begin{align*}
-  \begin{gathered}[t]
-    \IF{r}\THEN\PW{x}{2}\FI
-    \SEQ
-    \PW{x}{1}
-    \SEQ
-    \PW{x}{2}
-    \SEQ
-    \IF{\BANG r}\THEN\PW{x}{1}\FI
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \event{a1}{\DW{x}{2}}{}
-        \event{a4}{\DW{x}{1}}{right=of a1}
-        \wki{a1}{a4}
-      \end{tikzinline}}    
-  \end{gathered}
-\end{align*}
-As shown by the following single-threaded code, allowing this outcome would violate \drfsc{}.
-\begin{align*}
-  \begin{gathered}[t]
-    \PW{y}{1}
-    \SEQ
-    \PR{y}{r}
-    \SEQ
-    \IF{r}\THEN\PW{x}{2}\FI
-    \SEQ
-    \PW{x}{1}
-    \SEQ
-    \PW{x}{2}
-    \SEQ
-    \IF{\BANG r}\THEN\PW{x}{1}\FI
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \event{a1}{\DW{x}{2}}{}
-        \event{a4}{\DW{x}{1}}{right=of a1}
-        \wki{a1}{a4}
-        \event{b2}{\DR{y}{1}}{left=of a1}
-        \event{b1}{\DW{y}{1}}{left=of b2}
-        \rf{b1}{b2}
-      \end{tikzinline}}    
-  \end{gathered}
-\end{align*}
-
-To validate the reverse direction of \reflem{lem:if:seq}, it may be tempting
-to define the semantics using \emph{weakest} preconditions, rather than
-preconditions.  But in this case the notion of program refinement could not be
-simple set inclusion---for example, in general we would \emph{not} have
-\begin{math}
-  \aPSS_1
-  \supseteq
-  \xIFTHEN{\aForm}{\aPSS_1}{}.
-\end{math}
-
-As a result, we leave \reflem{lem:if:seq} as an inequation.  The equational
-form may be valid using some notion of \emph{observational} or
-\emph{contextual} refinement, but we do not pursue that here.
-
-\subsection{Associativity and Skolemization}
-\label{sec:ex:assoc}
-
-The predicate transformers we have chosen for \ref{read-tau-dep} and
-\ref{read-tau-ind} are different from the ones used traditionally, which are
-written using substitution \jjr{}.  Attempting to write \ref{read-tau-dep} in
-this style we would have:
-\begin{enumerate}[topsep=0pt]
-\item[{\labeltext[\textsc{r}4a$'$]{(\textsc{r}4a$'$)}{read-tau-dep-p}}]
-  if $(\aEvs\cap\bEvs)\neq\emptyset$ then
-  \begin{math}
-    \aTr{\bEvs}{\bForm} \riff    
-    \bForm[\aVal/\aReg]
-  \end{math},    
-\end{enumerate}
-Recall that \ref{read-tau-empty} says that $\bForm$ must be independent of
-$\aReg$ in order to appear in a top-level pomset: if $\aEvs=\emptyset$ then
-\begin{math}
-  \aTr{\bEvs}{\bForm} \riff
-  % \PBR{\aVal{=}\aReg \lor \aLoc{=}\aReg} \limplies
-  \bForm.
-\end{math}
-This choice for \ref{read-tau-empty} is forced by \refdef{def:family}, which
-states that the predicate transformer for a small subset of $\aEvs$ must
-imply the transformer for a larger subset.
-
-Sadly, this definition fails associativity.
-
-Consider the following, eliding transformers for the writes (``$\BANG$''
-represents logical negation):
-\begin{align*}
-  \begin{gathered}[t]
-    \PR{y}{r}
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1em]
-        \event{a}{\DR{y}{1}}{}
-        \xform{xd}{1{=}r\limplies\bForm}{right=of a}
-        \xform{xi}{(y{=}r\lor1{=}r)\limplies\bForm}{left=.5em of a}
-        \xo{a1}{xd}
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-  \begin{gathered}[t]
-    \PW{x}{\BANG r} 
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1em]
-        \event{b}{r{=}0\mid\DW{x}{1}}{}
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-  \begin{gathered}[t]
-    \PW{x}{\BANG\BANG r}
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1em]
-        \event{b}{r{\neq}0\mid\DW{x}{1}}{}
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-  \begin{gathered}[t]
-    \PW{x}{0} 
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1em]
-        \event{b}{\DW{x}{0}}{}
-      \end{tikzinline}}    
-  \end{gathered}
-\end{align*}
-Coalescing the writes and associating to the right,
-we have the following, since $(r{=}0 \lor r{\neq}0)\riff\TRUE$:
-\begin{align*}
-  \begin{gathered}[t]
-    \PR{y}{r}
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1em]
-        \event{a}{\DR{y}{1}}{}
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-  \begin{gathered}[t]
-    \PW{x}{\BANG r} 
-    \SEMI \PW{x}{\BANG\BANG r} 
-    \SEMI \PW{x}{0} 
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1em]
-        \event{b}{\DW{x}{1}}{}
-        \event{c}{\DW{x}{0}}{right=of b}
-        \wki{b}{c}
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-% \end{align*}  
-% The precondition of $\DWP{x}{1}$ is a tautology, thus we have:
-% \begin{align*}
-  \begin{gathered}[t]
-    \PR{y}{r}
-    \SEMI (\PW{x}{\BANG r} 
-    \SEMI \PW{x}{\BANG\BANG r} 
-    \SEMI \PW{x}{0} )
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1em]
-        \event{a}{\DR{y}{1}}{}
-        \event{b}{\DW{x}{1}}{right=of a}
-        \event{c}{\DW{x}{0}}{right=of b}
-        \wki{b}{c}
-      \end{tikzinline}}    
-  \end{gathered}
-\end{align*}
-The precondition of $\DWP{x}{1}$ is a tautology.
-Associating to the left and the coalescing, instead:
-\begin{align*}
-  \begin{gathered}[t]
-    \PR{y}{r}
-    \SEMI \PW{x}{\BANG r} 
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1em]
-        \event{a}{\DR{y}{1}}{}
-        \event{b}{(y{=}r\lor1{=}r)\limplies r{=}0\mid\DW{x}{1}}{right=of a}
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-  \begin{gathered}[t]
-    \PW{x}{\BANG\BANG r} 
-    \SEMI \PW{x}{0} 
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1em]
-        \event{b}{r{\neq}0\mid\DW{x}{1}}{}
-        \event{c}{\DW{x}{0}}{right=of b}
-        \wki{b}{c}
-      \end{tikzinline}}    
-  \end{gathered}
-  &&
-% \end{align*}  
-% Sequencing and merging:
-% \begin{align*}
-  \begin{gathered}[t]
-    (\PR{y}{r}
-    \SEMI \PW{x}{\BANG r} )
-    \SEMI (\PW{x}{\BANG\BANG r} 
-    \SEMI \PW{x}{0} )
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1em]
-        \event{a}{\DR{y}{1}}{}
-        \event{b}{\aForm\mid\DW{x}{1}}{right=of a}
-        \event{c}{\DW{x}{0}}{right=of b}
-        \wki{b}{c}
-      \end{tikzinline}}    
-  \end{gathered}
-\end{align*}
-where $\aForm=((y{=}r\lor1{=}r)\limplies r{=}0)\lor (r{\neq}0)$.  The
-precondition $\aForm$ is not a tautology.  In a top-level pomset, this forces
-dependency order from $\DRP{y}{1}$ to $\DWP{x}{1}$.
-
-Our solution is to Skolemize, replacing 
-uses of $\bForm[\aVal/\aReg]$
-by $(\aReg{=}\aVal)\limplies\bForm$,
-for uniquely chosen $\aReg$.
-%  We have proven associativity of
-%  \refdef{def:pomsets-trans} in Agda.
-The proof of associativity requires that predicate
-transformers distribute through disjunction (\refdef{def:trans}).  The
-attempt to define predicate transformers using substitution fails for
-\ref{read-tau-empty} because the predicate transformer
-$\aTr{}{\bForm}=(\forall\aReg)\bForm$ does not distribute through
-disjunction:
-\begin{math}
-  \aTr{}{\bForm_1\lor \bForm_2}=
-  (\forall r)(\bForm_1\lor \bForm_2)
-  \neq
-  ((\forall r)(\bForm_1)) \lor ((\forall r)(\bForm_2))
-  = \aTr{}{\bForm_1} \lor \aTr{}{\bForm_2}
-\end{math}.  Since $\aTr{}\bForm=(\forall\aReg)\bForm$ does not distribute
-through disjunction, we use $\aTr{}\bForm=\bForm$ instead (which trivially
-distributes through disjunction).  This change means we cannot use
-substitution, since $\bForm$ does not imply $\bForm[\aVal/\aReg]$.
-Fortunately, Skolemizing solves this problem, since $\bForm$ implies
-$(\aReg{=}\aVal)\limplies\bForm$.
-
-\subsection{Comparison with Weakest Preconditions}
-
-We compare traditional transformers to the dependent-case transformers of
-\reffig{fig:sem}. %; thus we consider only totally ordered executions.
-% Because
-% we only consider the dependent case, we drop the superscript $\aEvs$ on
-% $\aTr{\aEvs}{}$ throughout this section.  We also assume that each register
-% appears at most once in a program, as we did throughout
-% \textsection\ref{sec:model}--\ref{sec:arm}.
-
-Because of augment closure, we are not interested in isolating the
-\emph{weakest} precondition.  Thus we think of transformers as Hoare triples.
-In addition, all programs in our language are strongly normalizing, so we
-need not distinguish strong and weak correctness.  In this setting, the Hoare
-triple $\hoare{\aForm}{\aCmd}{\bForm}$ holds exactly when
-$\aForm \limplies \fwp{\aCmd}{\bForm}$.
-
-Hoare triples do not distinguish thread-local variables from shared
-variables.  Thus, the assignment rule applies to all types of storage. The
-rules can be written as on the left below:
-\begin{align*}
-\begin{aligned}
-  \fwp{\PW{\aLoc}{\aExp}}{\bForm} &= \bForm[\aExp/\aLoc]
-  \\
-  \fwp{\LET{\aReg}{\aExp}}{\bForm} &= \bForm[\aExp/\aReg]
-  \\
-  \fwp{\PR{\aLoc}{\aReg}}{\bForm} &= \aLoc{=}\aReg\limplies\bForm
-\end{aligned}
-&&
-\begin{aligned}
-  \trd{\PW{\aLoc}{\aExp}}{\bForm} &= \bForm[\aExp/\aLoc]
-  \\
-  \trd{\LET{\aReg}{\aExp}}{\bForm} &= \bForm[\aExp/\aReg]
-  \\
-  \trd{\PR{\aLoc}{\aReg}}{\bForm} &= \aVal{=}\aReg\limplies\bForm &&
-  \textwhere \labelingAct(\aEv)=\DR{\aLoc}{\aVal}
-\end{aligned}
-\end{align*}
-Here we have chosen an alternative formulation for the read rule, which is
-equivalent to the more traditional $\bForm[\aLoc/\aReg]$, as long as registers
-are assigned at most once in a program.  Our predicate transformers for the
-dependent case are shown on the right above.  Only the read rule differs from
-the traditional one.
-
-For programs where every register is bound and every read is fulfilled, our
-dependent transformers are the same as the traditional ones.  Thus, when
-comparing to weakest preconditions, let us only consider totally-ordered
-executions of our semantics where every read could be fulfilled by prepending
-some writes.  For example, we ignore pomsets of $\PW{x}{2}\SEMI\PR{x}{r}$
-that read $1$ for $x$.
-
-For example, let $\aCmd_i$ be defined:
-% as follows.
-\begin{align*}
-  \aCmd_1&=\PR{x}{s}\SEMI\PW{x}{s{+}r}
-  &  
-  \aCmd_2&=\PW{x}{t}\SEMI\aCmd_1
-  &  
-  \aCmd_3&=\LET{t}{2}\SEMI\LET{r}{5}\SEMI\aCmd_2
-\end{align*}
-% \begin{itemize}
-% \item
-%   \begin{math}
-%     \fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]
-%   \end{math}
-% \item
-%   \begin{math}
-%     \fwp{\PR{\aLoc}{\aReg}\;\,}{\bForm} = %\bForm[x/r]
-%     %     (\forall\bReg)
-%     %     \aLoc{=}\bReg\limplies\bForm [\bReg/\aLoc]
-%     \aLoc{=}\aReg\limplies\bForm
-%   \end{math}
-% \item
-%   \begin{math}
-%     \fwp{\PW{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]
-%   \end{math}
-% \end{itemize}
-% General relation between Hoare triples and $\fwp{}{}$:
-% \begin{itemize}
-% \item $\hoare{\fwp{\aCmd}{\bForm}}{\aCmd}{\bForm}$,
-% \item If $\hoare{\aForm}{\aCmd}{\bForm}$ and $\aCmd$ terminates when starting
-%   in any state satisfying $\aForm$, then $\aForm \limplies \fwp{\aCmd}{\bForm}$.
-% \end{itemize}
-The following pomset appears in the semantics of $\aCmd_2$.  A pomset for
-$\aCmd_3$ can be derived by substituting $[2/t,\allowbreak5/r]$.  A pomset
-for $\aCmd_1$ can be derived by eliminating the initial write.
-\begin{gather*}
-  % \begin{gathered}[t]
-  %   \PW{x}{3}
-  %   \\
-  %   \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-  %     \event{c}{\DW{x}{3}}{}
-  %     \xform{xd}{\bForm}{below=of c}
-  %     \xo[xright]{c}{xd}
-  %   \end{tikzinline}}
-  % \end{gathered}
-  % \qquad\quad
-  % \begin{gathered}[t]
-  %   \PR{x}{s}\SEMI\PW{x}{s{+}r}
-  %   \\
-  %   \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-  %     \event{a}{\DR{x}{2}}{}
-  %     \event{b}{2{=}s\limplies(s{+}r){=}7\mid\DW{x}{7}}{right=of a}%6.5em of a}
-  %     \po{a}{b}
-  %     \xform{xdd}{2{=}s \limplies \bForm[s{+}r/x]}{below right=.5em and -1em of a}
-  %       %     \xform{xdd}{2{=}s \limplies \bForm[s{+}r/x]}{above=of a}
-  %       %     \xform{xdi}{2{=}s \limplies \bForm[s{+}r/x]}{below=of a}
-  %       %     \xform{xii}{(2{=}s\lor x{=}s)\limplies\bForm[s{+}r/x]}{above=of b}
-  %       %     \xform{xid}{(2{=}s\lor x{=}s)\limplies\bForm[s{+}r/x]}{below=of b}
-  %       %     \xo[xright]{a}{xdi}
-  %       %     \xo[xright]{b}{xid}
-  %     \xo[xright]{a}{xdd}
-  %     \xo[xright]{b}{xdd}
-  %   \end{tikzinline}}
-  % \end{gathered}
-  % \\[1ex]
-  \begin{gathered}[t]
-    % \LET{t}{2}\SEMI
-    % \LET{r}{5}\SEMI
-    \PW{x}{t}\SEMI
-    \PR{x}{s}\SEMI\PW{x}{s{+}r}
-    \\
-    \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-        \event{a}{\DR{x}{2}}{}
-        \event{b}{2{=}s\limplies(s{+}r){=}7\mid\DW{x}{7}}{right=of a}
-        \event{c}{t{=}2\mid\DW{x}{2}}{left=of a}
-        \xform{xdd}{2{=}s \limplies \bForm[s{+}r/x]}{right=of b}%below right=.5em and -1em of a}
-        %\xo{a}{xdd}
-        \xo{b}{xdd}
-        % \xo{c}{xdd}
-        \po{a}{b}
-        \rf{c}{a}
-      \end{tikzinline}}
-  \end{gathered}
-\end{gather*}
-The predicate transformers are:
-% \begin{align*}
-%   \fwp{\aCmd_1}{\bForm} &= x{=}s\limplies\bForm[s{+}r/x] 
-%   \\
-%   \fwp{\aCmd_2}{\bForm} &= t\,{=}s\limplies\bForm[s{+}r/x] 
-%   \\
-%   \fwp{\aCmd_3}{\bForm} &= 2{=}s\limplies\bForm[s{+}5/x] 
-%   \\
-%   \trd{\aCmd_1}{\bForm} = \trd{\aCmd_2}{\bForm} &= 2{=}s\limplies\bForm[s{+}r/x] 
-%   \\
-%   \trd{\aCmd_3}{\bForm} &= 2{=}s\limplies\bForm[s{+}5/x] 
-% \end{align*}
-\begin{scope}
-  \small
-  \begin{align*}
-    \fwp{\aCmd_1}{\bForm} &= x{=}s\limplies\bForm[s{+}r/x] 
-    &
-    \trd{\aCmd_1}{\bForm} &= 2{=}s\limplies\bForm[s{+}r/x] 
-    \\
-    \fwp{\aCmd_2}{\bForm} &= t\,{=}s\limplies\bForm[s{+}r/x] 
-    &
-    \trd{\aCmd_2}{\bForm} &= 2{=}s\limplies\bForm[s{+}r/x] 
-    \\
-    \fwp{\aCmd_3}{\bForm} &= 2{=}s\limplies\bForm[s{+}5/x] 
-    &
-    \trd{\aCmd_3}{\bForm} &= 2{=}s\limplies\bForm[s{+}5/x] 
-  \end{align*}
-\end{scope}
-
-% % Let $\rho:\Reg\fun\Val$ and $\chi:\Loc\fun\Val$ be substitutions.
-% Let $\aState$ and $\rho$ range over substitutions $(\Reg\cup\Loc)\fun\Val$.
-% Treating substitutions as states, the big-step operational semantics of
-% programs can be defined as a relation $\bigstep{\aState}{\aCmd}{\bState}$.
-% % Ie, $\aForm\aState$ implies $\bForm\bState$.
-% \begin{align}
-%   \label{wp1}
-%   \bigstep{[5/r,2/x]}{\aCmd_1&}{[5/r,2/s,7/x]}
-%   \\
-%   \label{wp2}
-%   \bigstep{[\NEG5/r,2/x]}{\aCmd_1&}{[\NEG5/r,2/s,\NEG3/x]}
-% \end{align}
-
-% Then the semantics of Hoare triples guarantees that if
-% $\aForm\limplies\fwp{\aCmd}{\bForm}$, $\bigstep{\aState}{\aCmd}{\rho}$ and
-% $\aForm\aState$ is a tautology then $\bForm\bState$ is a tautology.
-% \begin{align*}
-%   \fwp{\aCmd_1}{x{>}0} &= (x{+}r{>}0) 
-% \end{align*}
-% In \eqref{wp1}, the pre- and post-conditions are satisfied.
-% In \eqref{wp2}, they are not.
-
-% \begin{itemize}  
-% \item Suppose $\bigstep{\aState}{\aCmd}{\rho}$ and $\aForm\limplies\fwp{\aCmd}{\bForm}$.\\
-%   If $\aForm\aState$ is a tautology then $\bForm\bState$ is a tautology.\\
-%   Ie, $\aForm\aState$ implies $\bForm\bState$.
-% \item Suppose $\bigstep{\aState}{\aCmd}{\rho}$ and $\hoare{\aForm}{\aCmd}{\bForm}$.\\
-%   If $\aForm\aState$ is a tautology then $\bForm\bState$ is a tautology.\\
-%   Ie, $\aForm\aState$ implies $\bForm\bState$.
-% \item Suppose $\bigstep{\aState}{\aCmd}{\rho}$ and $\aForm=\fwp{\aCmd}{\bForm}$.\\
-%   $\aForm\aState$ is a tautology if and only if $\bForm\bState$ is a tautology.\\
-%   Ie, $\aForm\aState$ iff $\bForm\bState$.
-% % \item Weakest: If $\aForm'\aState$ is a tautology, then $\aForm$ implies $\aForm'$.
-% \end{itemize}
-% Weakest preconditions are \emph{sound} in that if $\aForm$ holds in the
-% initial state $\aState$, then $\bForm$ holds in the final state $\bState$.
-% Formally,
-
-\begin{comment}
-  If $\aPS\in\sem{\aCmd}$ is top-level and quiescent then 
-  $\aTr{\aEvs}{\bForm}$ implies $\fwp{\aCmd}{\bForm}$.
-
-  For any substitution $\aSub=[{v_1/r_1},\ldots, {v_n/r_n}]$ there is some
-  $\aPS\in\sem{\aCmd}$ %that is top-level and quiescent
-  such that all preconditions in $\aPS\aSub$ are tautologies then 
-  $\fwp{\aCmd}{\bForm}\aSub$
-\end{comment}
-
-% For a language where all programs are
-% terminating, we have for any statement $\aCmd$:
-% \begin{align*}
-%   \hoare{\aForm}{\aCmd}{\bForm} 
-%   \;\;\Leftrightarrow\;\;
-%   \aForm \textimplies \fwp{\aCmd}{\bForm}
-% \end{align*}
-% Interpretation is that if $\aState\models\fwp{\aCmd}{\bForm}$ and
-% $\bigstep{\aState}{\aCmd}{\rho}$
-% then $\bState\models\bForm$.
-
-% Let $\aCmd_0$ be
-% \begin{math}
-%   \PW{\aLoc_1}{\aVal_1}\SEMI\cdots\SEMI \PW{\aLoc_n}{\aVal_n}, 
-% \end{math}
-% such that $\fwp{\aCmd_0}{\aForm}$ is a tautology, and $\aLoc_i=\aLoc_j$
-% implies $i=j$.
-
-% Let $\aSub_\aPS=[{\aVal_1/\aLoc_1},\ldots, {\aVal_n/\aLoc_n}]$ be the final
-% state of $\aPS$.
-
-% Let $\aState$ and $\rho$ range over substitutions $\Loc\fun\Exp$.
-% If we leave the registers free, we have:
-% \begin{align}
-%   \label{wp1x}
-%   \bigstep{[2/x]}{\aCmd&}{[6/x]}
-% \end{align}
-
-% Using \refdef{def:pomsets-trans}:
-% \begin{align*}
-%   \begin{gathered}[t]
-%     %     \PR{x}{s}\SEMI\PW{x}{s{+}r}
-%     \PR{x}{s}
-%     \\
-%     \hbox{\begin{tikzinline}[node distance=.5em and 1em]
-%       \event{a}{\DR{x}{2}}{}
-%       \xform{xi}{\bForm}{above=of a}
-%       \xform{xd}{2{=}s \limplies \bForm}{below=of a}
-%       \xo[xright]{a}{xd}
-%     \end{tikzinline}}
-%   \end{gathered}
-%   &&
-%   \begin{gathered}[t]
-%     \PW{x}{s{+}r}
-%     \\
-%     \hbox{\begin{tikzinline}[node distance=.5em and 1em]
-%       \event{a}{(s{+}r){=}7\mid\DW{x}{7}}{}
-%       \xform{xi}{\bForm}{above=of a}
-%       \xform{xd}{\bForm}{below=of a}
-%       \xo[xright]{a}{xd}
-%     \end{tikzinline}}
-%   \end{gathered}
-% \end{align*}
-% Composing
-% \begin{align*}
-%   \begin{gathered}[t]
-%     \PR{x}{s}\SEMI\PW{x}{s{+}r}
-%     \\
-%     \hbox{\begin{tikzinline}[node distance=.5em and 1em]
-%       \event{a}{\DR{x}{2}}{}
-%       \event{b}{(s{+}r){=}7\mid\DW{x}{7}}{right=of a}
-%       \xform{xdd}{2{=}s \limplies \bForm}{above=of a}
-%       \xform{xdi}{2{=}s \limplies \bForm}{below=of a}
-%       \xform{xii}{\bForm}{above=of b}
-%       \xform{xid}{\bForm}{below=of b}
-%       \xo[xright]{a}{xdi}
-%       \xo[xright]{b}{xid}
-%       \xo[xright]{a}{xdd}
-%       \xo[xright]{b}{xdd}
-%     \end{tikzinline}}
-%   \end{gathered}
-% \end{align*}
-
-% Using \refdef{def:pomsets-lir}:
-% \begin{align*}
-%   \begin{gathered}[t]
-%     %     \PR{x}{s}\SEMI\PW{x}{s{+}r}
-%     \PR{x}{s}
-%     \\
-%     \hbox{\begin{tikzinline}[node distance=.5em and 1em]
-%       \event{a}{\DR{x}{2}}{}
-%       \xform{xi}{(2{=}s\lor x{=}s)\limplies \bForm}{above=of a}
-%       \xform{xd}{2{=}s \limplies \bForm}{below=of a}
-%       \xo[xright]{a}{xd}
-%     \end{tikzinline}}
-%   \end{gathered}
-%   &&
-%   \begin{gathered}[t]
-%     \PW{x}{s{+}r}
-%     \\
-%     \hbox{\begin{tikzinline}[node distance=.5em and 1em]
-%       \event{a}{(s{+}r){=}7\mid\DW{x}{7}}{}
-%       \xform{xi}{\bForm[s{+}r/x]}{above=of a}
-%       \xform{xd}{\bForm[s{+}r/x]}{below=of a}
-%       \xo[xright]{a}{xd}
-%     \end{tikzinline}}
-%   \end{gathered}
-% \end{align*}
-% Composing
-
-% For example, let $\aCmd_1=\PR{x}{r}$ and $\aCmd_2=\PW{x}{r{+}1}$ and
-% $\aCmd=\aCmd_1\SEMI \aCmd_2$.
-% \begin{align*}
-%   \fwp{\aCmd_2}{x{>}1}&=(r{+}1{>}1) = (r{>}0)
-%   \\
-%   \fwp{\aCmd_1}{r{>}0}=\fwp{\aCmd_0}{x{>}1}&=(x{>}0)
-% \end{align*}
-% Let $\aPS_i\in\sem{\aCmd_i}$.
-% \begin{align*}
-%   \aTr[2]{\aEvs_2}{x{>}1}&=(r{+}1{>}1) = (r{>}0)
-%   \\
-%   \aTr[0]{\aEvs_0}{x{>}1}&=(0{=}\aReg \limplies r{>}0)
-%   \\
-%   \aTr[0]{\aEvs_0}{x{>}1}&=(1{=}\aReg \limplies r{>}0)
-%   \\
-%   \aTr[0]{\aEvs_0}{x{>}1}&=(2{=}\aReg \limplies r{>}0)
-% \end{align*}
-
-% \begin{proposition}
-%   If $\aPS\in\sem{\aCmd}$ is top-level and quiescent then 
-%   $\aTr{\aEvs}{\aForm}$ implies $\fwp{\aCmd}{\aForm}$.
-
-%   For any substitution $\aSub=[{\aVal_1/\aReg_1},\ldots, {\aVal_n/\aReg_n}]$ there is some
-%   $\aPS\in\sem{\aCmd}$ %that is top-level and quiescent
-%   such that all preconditions in $\aPS\aSub$ are tautologies then 
-%   $\fwp{\aCmd}{\aForm}\aSub$
-% \end{proposition}
-
-% \subsection{Fences}
-% \label{sec:fences}
-
-\subsection{Substitutions}
-\label{sec:substitutions}
-
-In $\sLOAD{}{}$, it is also possible to collapse $\aLoc$ and $\aReg$ via substitution:
-\begin{enumerate}
-\item[{\labeltext[\textsc{r}4a$'$]{(\textsc{r}4a$'$)}{read-tau-dep-sub}}]
-  if $(\aEvs\cap\bEvs)\neq\emptyset$ then
-  \begin{math}
-    \aTr{\bEvs}{\bForm} \riff
-    \aVal{=}\aReg
-    \limplies \bForm[\aReg/\aLoc]
-  \end{math},    
-\item[{\labeltext[\textsc{r}4b$'$]{(\textsc{r}4b$'$)}{read-tau-ind-sub}}]
-  if $\aEvs\neq\emptyset$ and $(\aEvs\cap\bEvs)=\emptyset$ then
-  \begin{math}
-    \aTr{\bEvs}{\bForm} \riff
-    \PBR{\aVal{=}\aReg \lor \aLoc{=}\aReg} \limplies
-    \bForm[\aReg/\aLoc],
-  \end{math}
-\item[{\labeltext[\textsc{r}4c$'$]{(\textsc{r}4c$'$)}{read-tau-empty-sub}}]
-  if $\aEvs=\emptyset$ then
-  \begin{math}
-    \aTr{\bEvs}{\bForm} \riff
-    % \PBR{\aVal{=}\aReg \lor \aLoc{=}\aReg} \limplies
-    \bForm[\aReg/\aLoc],
-  \end{math}
-\end{enumerate}
-Perhaps surprisingly, this semantics is incomparable with that of
-\reffig{fig:sem}.  Consider the following:
-\begin{gather*}
-  \IF{r\land s\;\mathsf{even}}\THEN \PW{y}{1}\FI\SEMI
-  \IF{r\land s}\THEN \PW{z}{1}\FI
-  \\
-  \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-      \event{a3}{r\land s\;\mathsf{even}\mid\DW{y}{1}}{}
-      \event{a4}{r\land s\mid\DW{z}{1}}{right=of a3}
-    \end{tikzinline}}
-\end{gather*}
-Prepending $\PRP{x}{s}$, we get the same result regardless of whether we
-substitute $[s/x]$, since $x$ does not occur in either precondition.  Here
-we show the independent case:
-\begin{gather*}
-  \PR{x}{s}\SEMI
-  \IF{r\land s\;\mathsf{even}}\THEN \PW{y}{1}\FI\SEMI
-  \IF{r\land s}\THEN \PW{z}{1}\FI
-  \\
-  \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-      \event{a2}{\DR{x}{2}}{}
-      \event{a3}{(2{=}s\lor x{=}s)\limplies (r\land s\;\mathsf{even})\mid\DW{y}{1}}{right=of a2}
-      \event{a4}{(2{=}s\lor x{=}s)\limplies (r\land s)\mid\DW{z}{1}}{right=of a3}
-    \end{tikzinline}}
-\end{gather*}
-Since the preconditions mention $x$, prepending $\PRP{x}{r}$, we now get
-different results depending on whether we perform the substitution.  Without
-any substitution, we have:
-\begin{gather*}
-  \PR{x}{r}\SEMI
-  \PR{x}{s}\SEMI
-  \IF{r\land s\;\mathsf{even}}\THEN \PW{y}{1}\FI\SEMI
-  \IF{r\land s}\THEN \PW{z}{1}\FI
-  \\[-2ex]
-  \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-      \event{a1}{\DR{x}{1}}{}
-      \event{a2}{\DR{x}{2}}{right=of a1}
-      \event{a3}{1{=}r\limplies  (2{=}s\lor x{=}s)\limplies (r\land s\;\mathsf{even})\mid\DW{y}{1}}{right=of a2}
-      \event{a4}{1{=}r\limplies  (2{=}s\lor x{=}s)\limplies (r\land s)\mid\DW{z}{1}}{right=of a3}
-      \po[out=12,in=168]{a1}{a3}
-      \po[out=10,in=170]{a1}{a4}
-    \end{tikzinline}}
-\end{gather*}
-Prepending $\PWP{x}{0}$, which substitutes $[0/x]$, the precondition of
-$\DWP{y}{1}$ becomes
-$(1{=}r\limplies (2{=}s\lor0{=}s)\limplies (r\land s\;\mathsf{even}))$, which
-is a tautology, whereas the precondition of $\DW{z}{1}$ becomes
-$(1{=}r\limplies(2{=}s\lor0{=}s)\limplies (r\land s))$, which is not.  In
-order to be top-level, $\DWP{z}{1}$ must be dependency ordered after
-$\DRP{x}{2}$; in this case the precondition becomes
-$(1{=}r\limplies2{=}s\limplies (r\land s))$, which is a tautology.
-\begin{gather*}
-  % \PW{x}{0}\SEMI
-  % \PR{x}{r}\SEMI
-  % \PR{x}{s}\SEMI
-  % \IF{r\land s\;\mathsf{even}}\THEN \PW{y}{1}\FI\SEMI
-  % \IF{r\land s}\THEN \PW{z}{1}\FI
-  % \\
-  \hbox{\begin{tikzinline}[node distance=1.5em]
-      \event{a0}{\DW{x}{0}}{}
-      \event{a1}{\DR{x}{1}}{right=of a0}
-      \event{a2}{\DR{x}{2}}{right=of a1}
-      \event{a3}{\DW{y}{1}}{right=of a2}
-      \event{a4}{\DW{z}{1}}{right=of a3}
-      % \wk{a0}{a1}
-      % \wk[out=-20,in=-160]{a0}{a2}
-      \po[out=20,in=160]{a1}{a3}
-      \po[out=20,in=160]{a1}{a4}
-      \po[out=-20,in=-160]{a2}{a4}
-    \end{tikzinline}}
-\end{gather*}
-The situation reverses with the substitution $[r/x]$:
-\begin{gather*}
-  \PR{x}{r}\SEMI
-  \PR{x}{s}\SEMI
-  \IF{r\land s\;\mathsf{even}}\THEN \PW{y}{1}\FI\SEMI
-  \IF{r\land s}\THEN \PW{z}{1}\FI
-  \\[-2ex]
-  \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
-      \event{a1}{\DR{x}{1}}{}
-      \event{a2}{\DR{x}{2}}{right=of a1}
-      \event{a3}{1{=}r\limplies  (2{=}s\lor r{=}s)\limplies (r\land s\;\mathsf{even})\mid\DW{y}{1}}{right=of a2}
-      \event{a4}{1{=}r\limplies  (2{=}s\lor r{=}s)\limplies (r\land s)\mid\DW{z}{1}}{right=of a3}
-      \po[out=12,in=168]{a1}{a3}
-      \po[out=10,in=170]{a1}{a4}
-    \end{tikzinline}}
-\end{gather*}
-Prepending $\PWP{x}{0}$:
-%\vspace{-.5\baselineskip}
-\begin{gather*}
-  % \PW{x}{0}\SEMI
-  % \PR{x}{r}\SEMI
-  % \PR{x}{s}\SEMI
-  % \IF{r\land s\;\mathsf{even}}\THEN \PW{y}{1}\FI\SEMI
-  % \IF{r\land s}\THEN \PW{z}{1}\FI
-  % \\
-  \hbox{\begin{tikzinline}[node distance=1.5em]
-      \event{a0}{\DW{x}{0}}{}
-      \event{a1}{\DR{x}{1}}{right=of a0}
-      \event{a2}{\DR{x}{2}}{right=of a1}
-      \event{a3}{\DW{y}{1}}{right=of a2}
-      \event{a4}{\DW{z}{1}}{right=of a3}
-      % \wk{a0}{a1}
-      % \wk[out=-20,in=-160]{a0}{a2}
-      \po[out=20,in=160]{a1}{a3}
-      \po[out=20,in=160]{a1}{a4}
-      \po{a2}{a3}
-    \end{tikzinline}}
-\end{gather*}
-The dependency has changed from $\DRP{x}{2}\xpo\DWP{z}{1}$ to
-$\DRP{x}{2}\xpo\DWP{y}{1}$.  The resulting sets of pomsets are
-incomparable.
-
-Thinking in terms of hardware, the difference is whether reads update the
-cache, thus clobbering preceding writes.  With $[r/x]$, reads clobber the
-cache, whereas without the substitution, they do not.  Since most caches work
-this way, the model with $[r/x]$ is likely preferred for modeling hardware.
-However, this substitution only makes sense in a model with read-read
-coherence and read-read dependencies, which we will see is not the case for Arm.  By
-leaving out the substitution, we also ensure that downgraded reads are
-fulfilled by preceding writes, not reads.
-
-\label{sec:ex:last}

### 2021-07-07 paper/intro.tex
done through section3

--- 2021-07-06 paper/intro.tex
@@
 \begin{itemize}
-\item didn't bother tracking dependencies, allowing ``thin air'' executions
+\item did not bother tracking dependencies, allowing ``thin air'' executions
   \cite{Batty:2011:MCC:1926385.1926394},
@@
 sections add concurrency.  In \textsection\ref{sec:mca}, we define \PwTmca{},
-which models Java for \mca{} hardware, similar to that of
+which provides a Java-like model for \mca{} hardware, similar to that of
 \citet{DBLP:journals/pacmpl/JagadeesanJR20}.  In \textsection\ref{sec:c11},
@@
 \citet{DBLP:conf/esop/PaviottiCPWOB20}.  In \textsection\ref{sec:additional},
-we extend the semantics to include additional features, such as address calculation.
-In \textsection{sec:results}, we summarize the results.
+we extend the semantics to include additional features, such as address
+calculation.  In \textsection\ref{sec:results}, we summarize our results.
 
-\todo{flesh this out}
-\begin{itemize}
-\item We provide a tool to execute litmus tests for both models.
-\item \textsection\ref{sec:drf} We prove \drfsc{} for \PwTmca{}; \MRD{} inherits the result from \rcXI.
-\item \textsection\ref{sec:arm} We prove a lowering result for \PwTmca{}.
-\item \textsection\ref{sec:additional} We extend the model to include many features.
-\end{itemize}
+% \todo{flesh this out}
+% \begin{itemize}
+% \item We provide a tool to execute litmus tests for both models.
+% \item \textsection\ref{sec:sc} We prove \drfsc{} for \PwTmca{}; \MRD{} inherits the result from \rcXI.
+% \item \textsection\ref{sec:arm} We prove a lowering result for \PwTmca{}.
+% \item \textsection\ref{sec:additional} We extend the model to include many features.
+% \end{itemize}
 
-% \subsection{Contributions}
+% % \subsection{Contributions}
 
-% \todo{}
-% We show how predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75} can
-% be added to pomsets with
-% preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} to create a
-% compositional semantics for sequential composition.
-\begin{itemize}
-\item \S\ref{sec:model} presents the basic model, which satisfies many
-  desiderata, but not all.
-\item \S\ref{sec:arm} shows two approaches for efficient implementation on
-  Arm.  The first uses a suboptimal lowering for acquiring reads.  The second
-  uses an optimal lowering, but requires a nontrivial change to the
-  definition of sequential composition.
-\item \S\ref{sec:additional} generalizes the basic semantics of read and write
-  to validate compiler optimizations.
-\end{itemize}
-% Acknowledgements go here, once we're not double-blinded.
-% The definitions in this paper have been formalized in Agda.
-Because it is closely related, we expect that the memory-model results of
-\cite{DBLP:journals/pacmpl/JagadeesanJR20} apply to our model, including
-compositional reasoning for temporal safety properties and {local} \drfsc{}
-as in \cite{Dolan:2018:BDR:3192366.3192421,DBLP:conf/ppopp/DongolJR19,promising-ldrf}.
-% In \textsection\ref{sec:arm}, we provide an alternative proof strategy for
-% efficient compilation to \armeight{}, which improves upon that of
-% \cite{DBLP:journals/pacmpl/JagadeesanJR20} by using a recent alternative
-% characterization of \armeight{}.
+% % \todo{}
+% % We show how predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75} can
+% % be added to pomsets with
+% % preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} to create a
+% % compositional semantics for sequential composition.
+% \begin{itemize}
+% \item \S\ref{sec:model} presents the basic model, which satisfies many
+%   desiderata, but not all.
+% \item \S\ref{sec:arm} shows two approaches for efficient implementation on
+%   Arm.  The first uses a suboptimal lowering for acquiring reads.  The second
+%   uses an optimal lowering, but requires a nontrivial change to the
+%   definition of sequential composition.
+% \item \S\ref{sec:additional} generalizes the basic semantics of read and write
+%   to validate compiler optimizations.
+% \end{itemize}
+% % Acknowledgements go here, once we're not double-blinded.
+% % The definitions in this paper have been formalized in Agda.
+% Because it is closely related, we expect that the memory-model results of
+% \cite{DBLP:journals/pacmpl/JagadeesanJR20} apply to our model, including
+% compositional reasoning for temporal safety properties and {local} \drfsc{}
+% as in \cite{Dolan:2018:BDR:3192366.3192421,DBLP:conf/ppopp/DongolJR19,promising-ldrf}.
+% % In \textsection\ref{sec:arm}, we provide an alternative proof strategy for
+% % efficient compilation to \armeight{}, which improves upon that of
+% % \cite{DBLP:journals/pacmpl/JagadeesanJR20} by using a recent alternative
+% % characterization of \armeight{}.
 

### 2021-07-07 paper/tool.tex
First stab at tool section

\newcommand{\pwter}{{\sc PwTer}}

\section{\pwter: automatic litmus test evaluator}
\label{sec:tool}

\pwter automatically and exhaustively calculates the allowed outcomes of litmus tests for the PwT, PwT-po and PwT-C11 models. It is built in OCaml, and uses Z3~\cite{???} to judge the truth of predicates constructed by the models. \pwter obviates the need for error-prone hand evaluation.

\pwter allows several modes of evaluation: it can evaluate the rules of Fig.~\ref{fig:seq}, implementing PwT; it can generate program order according to Section~\ref{sec:c11}, implementing PwwT-po; and similar to \MRD~\cite{DBLP:conf/esop/PaviottiCPWOB20}, it can construct C11-style pre-executions and filter them according to the rules of C11.
Our choice of C11 rules is careful: we take the rules of RC11~\cite{DBLP:conf/pldi/LahavVKHD17}, but we replace the overly strong No-Thin-Air restriction $\text{acyclic}(\rrf\cup\rpox)$ with $\text{acyclic}(\rrf\cup\lt)$. 
This is a more precise categorisation of thin-air behavior, and it allows aggressive compiler optimisations which would be erroneously forbidden by RC11's original No-Thin-Air axiom.
We show \pwter in action in Fig.~\ref{fig:tool}.
Finally, \pwter also allows us to toggle the completeness check of~\ref{def:complete}, providing an interface for understanding how fragments of code might compose by exposing preconditions and termination conditions which are not yet tautologies.

%% TODO: introduce some nice dogma about the value of a tool and what it's taught us.
%% The real friends are the tools we built along the way.

%% Notes for Simon and Mark:
% Intro
%  - Add contribution relating to tool

% Related work
%  - Mention tool for MRD & Event Structures, mention absence of tool for Promising

% This section
%  - tool for {PwT, PwT-PO, PwT-RC11}
%  - Validation of some tests? Which tests to include?

% C11 Section
%  - Discussion: C11 standard really needs a dep relation. MRD provides this, PwT provides this, Promising does not.
%  - There's an open question about how coherence should affect dep.
%    if(q) { x := 1; x := 2 } else { x := 2; x := 1 }
%  - What should the deps be?

% Conclusion
%  - Having a tool should be some kind of feel-good factor.

%% [Simon] I am not a fan of how this looks on the page.
\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.75\textwidth]{tool.png}
  \Description{Image shows tool running on a Mac. Java Causality Test Case 1 is printed, followed by an invocation of the tool as {\tt ./pomsets.exe --check --complete data/tests/jctc/jctc1.lit}, which shows that the tool passes the assertion listed in the test.}
  \caption{\label{fig:tool} Example output of tool, validating TC1~\cite{PughWebsite}.}
\end{center}
\end{figure}

### 2021-07-07 paper/tool.tex
Tool section: Fix broken macros, references, and citations

@@
-\newcommand{\pwter}{{\sc PwTer}}
-
-\section{\pwter: automatic litmus test evaluator}
+\section{\PwTerTITLE: automatic litmus test evaluator}
 \label{sec:tool}
 
-\pwter automatically and exhaustively calculates the allowed outcomes of litmus tests for the PwT, PwT-po and PwT-C11 models.
-It is built in OCaml, and uses Z3~\cite{???} to judge the truth of predicates constructed by the models.
-\pwter obviates the need for error-prone hand evaluation.
+\PwTer{} automatically and exhaustively calculates the allowed outcomes of litmus tests for the PwT, PwT-po and PwT-C11 models.
+It is built in OCaml, and uses Z3~\cite{Z3Solver} to judge the truth of predicates constructed by the models.
+\PwTer{} obviates the need for error-prone hand evaluation.
 
-\pwter allows several modes of evaluation: it can evaluate the rules of Fig.~\ref{fig:seq}, implementing PwT; it can generate program order according to Section~\ref{sec:c11}, implementing PwwT-po; and similar to \MRD~\cite{DBLP:conf/esop/PaviottiCPWOB20}, it can construct C11-style pre-executions and filter them according to the rules of C11.
+\PwTer{} allows several modes of evaluation: it can evaluate the rules of Fig.~\ref{fig:seq}, implementing PwT; it can generate program order according to Section~\ref{sec:c11}, implementing PwwT-po; and similar to \MRD~\cite{DBLP:conf/esop/PaviottiCPWOB20}, it can construct C11-style pre-executions and filter them according to the rules of C11.
 Our choice of C11 rules is careful: we take the rules of RC11~\cite{DBLP:conf/pldi/LahavVKHD17}, but we replace the overly strong No-Thin-Air restriction $\text{acyclic}(\rrf\cup\rpox)$ with $\text{acyclic}(\rrf\cup\lt)$.
 This is a more precise categorisation of thin-air behavior, and it allows aggressive compiler optimisations which would be erroneously forbidden by RC11's original No-Thin-Air axiom.
-We show \pwter in action in Fig.~\ref{fig:tool}.
-Finally, \pwter also allows us to toggle the completeness check of~\ref{def:complete}, providing an interface for understanding how fragments of code might compose by exposing preconditions and termination conditions which are not yet tautologies.
+We show \PwTer{} in action in Fig.~\ref{fig:tool}.
+Finally, \PwTer{} also allows us to toggle the complete check of~\ref{def:pomset}, providing an interface for understanding how fragments of code might compose by exposing preconditions and termination conditions which are not yet tautologies.
 
@@
 Java Causality Test Case 1 is printed, followed by an invocation of the tool as {\tt ./pomsets.exe --check --complete data/tests/jctc/jctc1.lit}, which shows that the tool passes the assertion listed in the test.}
-  \caption{\label{fig:tool} Example output of tool, validating TC1~\cite{PughWebsite}.}
+  \caption{\label{fig:tool} Example output of \PwTer, validating TC1~\cite{PughWebsite}.}
 \end{center}

### 2021-07-07 paper/abstract.tex
Abstract: include no-thin-air idea in other semantics' defficiencies

@@
 as a ``relaxed memory model.''  The relaxed memory models that have been
-proposed to date either fail to address sequential composition directly, or overly
-restrict processors and compilers.
+proposed to date either fail to address sequential composition directly, overly
+restrict processors and compilers, or permit nonsense thin-air behaviors which are
+unobservable in practice.
 

### 2021-07-07 paper/arm.tex
Minor spelling and editorial changes

@@
 \citet{armed} show that \EGC{} and \EC{} are both equivalent to the standard
-definition of Arm8.  They explain \EGC{} and \EC{} using the following
-example, which is allowed by Arm8.\footnote{We have changed an address
+definition of \armeight.  They explain \EGC{} and \EC{} using the following
+example, which is allowed by \armeight.\footnote{We have changed an address
   dependency in the first thread to a data dependency.}

### 2021-07-07 paper/intro.tex
Minor spelling and editorial changes

@@
 Pomsets with Preconditions (\PwP), which use preconditions and logic to
-calculate dependencies for a Java-like language on \mca{} hardware, such as
+calculate dependencies for a Java-like language on multicopy-atomic (\mca{}) hardware, such as
 \armeight{} \cite{DBLP:journals/pacmpl/PulteFDFSS18}.  However, neither paper
@@
 expect: programs write to, and read from, memory references.
-This can be thought of as a total order of reads and writes (black arrows),
-where each read has a matching \emph{fulfilling} write (green arrows),
+This can be thought of as a total order of reads and writes (pink arrows \tikz[baseline]{\draw [po] (0,0.1) -- (0.5,0.1);}),
+where each read has a matching \emph{fulfilling} write (green arrows \tikz[baseline]{\draw [rf] (0,0.1) -- (0.5,0.1);}),
 for example:

### 2021-07-07 paper/intro.tex
Adding text to introduction to plumb in tool

@@
 \item did not bother tracking dependencies, allowing ``thin air'' executions
-  \cite{Batty:2011:MCC:1926385.1926394},
+  --- as in C and C++~\cite{DBLP:conf/esop/BattyMNPS15},
 \item tracked dependencies conservatively, using syntax, requiring inefficient
   implementation of relaxed access
-  \cite{DBLP:conf/pldi/LahavVKHD17,Boehm:2014:OGA:2618128.2618134,DBLP:conf/oopsla/VafeiadisN13,DBLP:journals/corr/abs-1804-04214}---while
-  this may be fine for C, it's a non-started for safe languages like Java,
-  which lack C's ``plain'' access,
+  \cite{DBLP:conf/pldi/LahavVKHD17,Boehm:2014:OGA:2618128.2618134,DBLP:conf/oopsla/VafeiadisN13,DBLP:journals/corr/abs-1804-04214}--- a non-starter for safe languages like Java,
+  and an unacceptable cost for low-level languages like C,
 \item computed dependencies using non-compositional operational models over
@@
 we define \PwTc{}, which models \cXI, adapting the approach of
-\citet{DBLP:conf/esop/PaviottiCPWOB20}.  In \textsection\ref{sec:additional},
+\citet{DBLP:conf/esop/PaviottiCPWOB20}.  In \textsection\ref{sec:tool} we 
+provide a tool for automatic evaluation of litmus tests.  In \textsection\ref{sec:additional},
 we extend the semantics to include additional features, such as address
@@
 We provide a detailed comparison with these approaches in \textsection\ref{sec:promising}.
+
+Other work in relaxed memory has shown that tooling is especially useful to researchers, architects, and language specifiers, enabling them to build intuitions experimentally~\cite{DBLP:conf/esop/PaviottiCPWOB20,PrideMM,alglave,Batty:2011:MCC:1926385.1926394}.
+Unfortunately, it is not obvious that tools can be built for all thin-air free models, the calculation of \citet{Pichon-Pharabod:2016:CSR:2837614.2837616} does not have a termination proof for an arbitrary input, and the enormous state space for the operational models of \citet{DBLP:conf/popl/KangHLVD17} and \citet{DBLP:journals/pacmpl/ChakrabortyV19} is a daunting prospect for a tool builder -- and as yet no tool exists for automatically evaluating these models.
+We describe a tool, \PwTer, for automatically evaluating \PwT{} in \textsection\ref{sec:tool}.

### 2021-07-07 paper/arm.tex
edits

--- 2021-06-30 paper/arm.tex
@@
 \subsection{Arm executions}
+Our description of \armeight{} follows \citet{armed}, adapting the
+notation to our setting.
 \begin{definition}
@@
 
-We emphasize that \ref{pom-rf-le} does not hold for $\frf{\semrr{}}$:
- $\bEv\xrfx\aEv$ may not imply $\bEv\le\aEv$ when $\bEv$ and $\aEv$ come
-from different sides of a sequential composition.  This means that $\rrfx$
-must be verified during pomset construction, rather than post-hoc.  The
-following lemma gives a post-hoc verification technique for $\rrfx$, using
-program order ($\rpox$) and phantom events ($\fmrg{}$).  The construction is
-discussed in \textsection\ref{sec:c11}.
-\begin{lemma}
-  Any $\aPS$ in the image of $\frf{\semrr{}}$ is top-level iff
-  for every $\bEv\xrfx\aEv$ either
-  \begin{itemize}
-  \item external fulfillment:
-    $\bEv\le\aEv$ and if $\labeling(\cEv) \rblocks\labeling(\aEv)$ then either $\cEv\le\bEv$ or $\aEv\le\cEv$, or
-  \item internal fulfillment:
-    $(\exists\bEv'\mid\fmrg{\bEv'}=\bEv)$
-    $(\exists\aEv'\mid\fmrg{\aEv'}=\aEv)$
-\todo{The following is sketchy.}
-    $\bEv\xpox\aEv$ and $(\not\exists\cEv)$
-    $\labeling(\cEv) \rblocks \labeling(\aEv)$ and $\bEv\xpox\cEv\xpox\aEv$.
-  \end{itemize}
-\end{lemma}
 % \begin{enumerate}

### 2021-07-07 paper/mca1.tex
edits

--- 2021-07-06 paper/mca1.tex
@@
 \label{sec:mca}
-We define two models for \mca{} architectures: \PwTmca{1} and \PwTmca{2}.
-The first has a nicer semantics; the second has a more efficient lowering for
-acquiring reads on \armeight{}.  Both variants have optimal lowering for
-relaxed access.  We drop the subscript when referring to both models.
+
+We derive a model of concurrent computation by adding \emph{parallel
+  composition} and \emph{reads-from} to \reffig{fig:seq}.  To model coherence
+and synchronization, we add \emph{delay} to the rule for sequential
+composition.  For \mca{} architectures, it is sufficient to encode delay in
+the pomset order.  The resulting model, \PwTmca{1}, supports optimal lowering
+for relaxed access on \armeight{}, but requires extra synchronization for
+acquiring reads.
+
+A variant, \PwTmca{2}, supports optimal lowering for all access modes on
+\armeight{}.  To achieve this, \PwTmca{2} drops the global requirement that
+\emph{reads-from} implies pomset order \eqref{pom-rf-le}.
+
+The lowering proofs can be found in \textsection\ref{sec:arm}.  The proofs
+use recent alternative characterizations of \armeight{} \cite{armed}.
 
@@
 
-Note that reads-from implies order by \eqref{pom-rf-le}.
+We define \PwTmca{1} by extending \refdef{def:pomset} and \reffig{fig:seq}.
+The definition uses several relations over actions---$\rmatches$, $\rblocks$
+and $\rdelays$---as well a distinguished set of $\sread$ actions; see
+\textsection\ref{sec:actions}.
+\begin{definition}
+  \label{def:pwt:mca1}
+  A \PwTmca{1} is a \PwT{} (\refdef{def:pomset}) equipped
+  with a relation $\rrfx$ such that 
+  \begin{enumerate}[,label=(\textsc{m}\arabic*),ref=\textsc{m}\arabic*]
+    \setcounter{enumi}{\value{Brf}}
+  \item \label{pom-rf} \makecounter{rf} ${\rrfx} \subseteq \aEvs\times\aEvs$
+    is an injective relation capturing \emph{reads-from}, such that
+    % \end{enumerate}
+    % A pomset is a \emph{candidate} if there is an injective relation
+    % ${\rrfx} : \aEvs\times\aEvs$, capturing \emph{reads-from}, such that
+    \begin{enumerate}
+      % \begin{enumerate}[,label=(\textsc{i}\arabic*),ref=\textsc{i}\arabic*]
+      % \item \label{rf-injective}
+      %   if $\bEv\xrfx\aEv$ and $\cEv\xrfx\aEv$ then $\bEv=\cEv$, that is,
+      %   ${\rrfx}$ is injective,
+    \item \label{pom-rf-match} if $\bEv\xrfx\aEv$ then
+      $\labeling(\bEv) \rmatches \labeling(\aEv)$,
+    \item \label{pom-rf-block} if $\bEv\xrfx\aEv$ and
+      $\labeling(\cEv) \rblocks \labeling(\aEv)$ then either $\cEv\le\bEv$ or
+      $\aEv\le\cEv$,
+    \item \label{pom-rf-le} if $\bEv\xrfx\aEv$ then $\bEv\le\aEv$.
+    \end{enumerate}
+  \end{enumerate}
+
+  \noindent
+  \label{def:pwt:mca:complete}
+  A \PwTmca{} is \emph{complete} if %it is a complete \PwT{} (\refdef{def:pomset}) and
+  \begin{multicols}{2}
+    \begin{enumerate}[,label=(\textsc{c}\arabic*),ref=\textsc{c}\arabic*]
+    \item[\eqref{top-kappa}]
+      \eqref{top-term}\;
+      as in \refdef{def:pomset},
+      \setcounter{enumi}{\value{Brf}}
+    \item \label{top-rf}
+      if $\labeling(\aEv)$ is a $\sread$ then there is some $\bEv\xrfx\aEv$.
+    \end{enumerate}
+  \end{multicols}
+  \smallskip
+
+  %The semantic rules are as follows:
+  \input{fig-mca1.tex}
+\end{definition}
+Let
+\begin{math}
+  \sem{\aCmd_1 \LPAR[\bThrd] \aCmd_2} = \sLPAR{\sem{\aCmd_1}}{\sem{\aCmd_2}}.
+\end{math}
+We write $\semmca{1}{}$ for the semantic function when it is unclear from context.
+
+In complete pomsets, $\rrfx$ must pair every read with a matching write
+\eqref{top-rf}.  The requirements \ref{pom-rf-match}, \ref{pom-rf-block}, and
+\ref{pom-rf-le} guarantee that reads are \emph{fulfilled}, as in
+\cite[\textsection 2.7]{DBLP:journals/pacmpl/JagadeesanJR20}.
+
+The semantic rules are mostly straightforward: Parallel composition is disjoint
+union, and all constructs respect reads-from.  Only \ref{seq-le-delays}
+requires explanation.
+
 % \begin{lemma}
@@
 % \end{lemma}
-In top-level pomsets, every read must have a matching write in $\rrfx$
-\eqref{top-rf}.  Together with \ref{pom-rf-match} and \ref{pom-rf-block}, the lemma
-guarantees that reads are \emph{fulfilled} at top-level, as in
-\cite[\textsection 2.7]{DBLP:journals/pacmpl/JagadeesanJR20}.\footnote{The
-  basic model would be the same if we move $\rrfx$ from the model itself to
-  be existentially quantified in the definition of top-level pomset, along
-  with \ref{pom-rf-match} and \ref{pom-rf-block}.  This was the approach of
-  \citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} We include $\rrfx$
-  explicitly for use in \textsection\ref{sec:arm2}, where we introduce a
-  variant semantics $\frf{\semrr{}}$ where \ref{pom-rf-le} is not required.}
+% \footnote{The
+%   basic model would be the same if we move $\rrfx$ from the model itself to
+%   be existentially quantified in the definition of top-level pomset, along
+%   with \ref{pom-rf-match} and \ref{pom-rf-block}.  This was the approach of
+%   \citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} We include $\rrfx$
+%   explicitly for use in \textsection\ref{sec:arm2}, where we introduce a
+%   variant semantics $\frf{\semrr{}}$ where \ref{pom-rf-le} is not required.}
 
@@
 example, consider the following:\footnote{We use different colors for arrows representing order:
+  \begin{multicols}{2}
     \begin{itemize}  
-  \item \makebox{$\bEv\xpo\aEv$} arises from control/data/address \emph{dependency} (\ref{seq-kappa}, definition of $\labelingForm'_2(\bEv)$),
+    \item \makebox{$\bEv\xwki\aEv$} arises from $\eqreorderco$ \eqref{seq-le-delays},
     \item \makebox{$\bEv\xsync\aEv$} arises from $\reorderra$ or $\eqreordersc$ \eqref{seq-le-delays},
-  \item \makebox{$\bEv\xwki\aEv$} arises from $\eqreorderco$ \eqref{seq-le-delays},
+    \item \makebox{$\bEv\xpo\aEv$} arises from \makebox[0pt][l]{control/data/address \emph{dependency} (\ref{seq-kappa}, definition of $\labelingForm'_2(\bEv)$),}
     \item \makebox{$\bEv\xrf\aEv$} arises from \emph{reads-from} \eqref{pom-rf-match},
-  %\item \makebox{$\bEv\xsyncsc\aEv$} arises from \emph{strong fencing} \eqref{cand-lesync-sc},
+      % \item \makebox{$\bEv\xsyncsc\aEv$} arises from \emph{strong fencing} \eqref{cand-lesync-sc},
     \item \makebox{$\bEv\xwk\aEv$} arises from \emph{blocking} \eqref{pom-rf-block}.
     \end{itemize}    
-  In \textsection\ref{sec:arm2}, it is possible for $\rrf$ to contradict
+  \end{multicols}
+  In \PwTmca{2}, it is possible for $\rrf$ to contradict
   $\le$.  In this case, we use a dotted arrow for $\rrf$: $\bEv\xrfint\aEv$
@@
 elimination, and stronger than Java, in order to support local reasoning
-about data races.  See
-\cite[\textsection3.1]{DBLP:journals/pacmpl/JagadeesanJR20} for a discussion.
+about data races.  For further examples, see \textsection\ref{sec:extras} and
+\cite[\textsection3.1]{DBLP:journals/pacmpl/JagadeesanJR20}.
 
-\begin{definition}
-  \label{def:pwt:mca1}
-  A \PwTmca{1} is a \PwT{} (\refdef{def:pomset}) equipped
-  with a relation $\rrfx$ such that 
-  \begin{enumerate}[,label=(\textsc{m}\arabic*),ref=\textsc{m}\arabic*]
-    \setcounter{enumi}{\value{Brf}}
-  \item \label{pom-rf} \makecounter{rf} ${\rrfx} \subseteq \aEvs\times\aEvs$
-    is an injective relation capturing \emph{reads-from}, such that
-    % \end{enumerate}
-    % A pomset is a \emph{candidate} if there is an injective relation
-    % ${\rrfx} : \aEvs\times\aEvs$, capturing \emph{reads-from}, such that
-    \begin{enumerate}
-      % \begin{enumerate}[,label=(\textsc{i}\arabic*),ref=\textsc{i}\arabic*]
-      % \item \label{rf-injective}
-      %   if $\bEv\xrfx\aEv$ and $\cEv\xrfx\aEv$ then $\bEv=\cEv$, that is,
-      %   ${\rrfx}$ is injective,
-    \item \label{pom-rf-match} if $\bEv\xrfx\aEv$ then
-      $\labeling(\bEv) \rmatches \labeling(\aEv)$,
-    \item \label{pom-rf-block} if $\bEv\xrfx\aEv$ and
-      $\labeling(\cEv) \rblocks \labeling(\aEv)$ then either $\cEv\le\bEv$ or
-      $\aEv\le\cEv$,
-    \item \label{pom-rf-le} if $\bEv\xrfx\aEv$ then $\bEv\le\aEv$.
-    \end{enumerate}
-  \end{enumerate}
+Lemmas \ref{lem:monoid} and \ref{lem:if} mostly hold for \PwTmca{1}.  The
+exceptions are items \eqref{lem:ifelse:if:if1} and \eqref{lem:ifelse:if:if2},
+which become inclusions.  For example, \eqref{lem:ifelse:if:if1} becomes:
+\begin{displaymath} 
+  \xIFTHEN{\aForm}{\aPSS_1}{\aPSS_2}
+  \supseteq
+  \xSEMI{
+    \xIFTHEN{\aForm}{\aPSS_1}{}
+  }{
+    \xIFTHEN{\lnot\aForm}{\aPSS_2}{}
+  }
+\end{displaymath}
+The culprit is $\rdelay$, which introduces order regardless whether
+preconditions are disjoint.  As an example, 
+\begin{math}
+  \sem{\IF{r}
+  \THEN \PW{x}{1}
+  \ELSE \PW{x}{2}
+  \FI}
+\end{math}
+has an execution with
+\begin{math}
+  (r{=}0\mid\DW{x}{2})
+  \xwki
+  (r{\neq}0\mid\DW{x}{1}),
+\end{math}
+(using augmentation), whereas
+\begin{math}
+  \sem{
+    \IF{r} \THEN \PW{x}{1}\FI
+    \SEMI
+    \IF{\BANG r} \THEN \PW{x}{2}\FI
+  \FI}
+\end{math}
+has no such execution.
 
-  \label{def:pwt:mca:complete}
-  A \PwTmca{} is \emph{complete} if %it is a complete \PwT{} (\refdef{def:pomset}) and
-  \begin{multicols}{2}
-    \begin{enumerate}[,label=(\textsc{c}\arabic*),ref=\textsc{c}\arabic*]
-    \item[\eqref{top-kappa}]
-      \eqref{top-term}\;
-      as in \refdef{def:pomset},
-      \setcounter{enumi}{\value{Brf}}
-    \item \label{top-rf}
-      if $\labeling(\aEv)$ is a $\sread$ then there is some $\bEv\xrfx\aEv$.
-    \end{enumerate}
-  \end{multicols}
-\end{definition}
+For further discussion, see \textsection\ref{sec:delay}.
 
@@
 
-\input{fig-mca1.tex}
-\input{fig-mca2.tex}
-
-The semantic rules are given in \reffig{fig:mca1}.
-We write $\semmca{1}{}$ for the semantic function.
-Let
-\begin{math}
-  \semmca{1}{\aCmd_1 \LPAR[\bThrd] \aCmd_2} = \sLPAR{\semmca{1}{\aCmd_1}}{\semmca{1}{\aCmd_2}}.
-\end{math}
+%The semantic rules are given in \reffig{fig:mca1}.
 

### 2021-07-07 paper/mca2.tex
edits

@@
 
-%
-The use of $\rextendsdef{}{}$ in \ref{if-rf-extends}
-ensures that no $\rrfx$ is introduced between events in
-$\aEvs_1\cap\aEvs_2$ when coalescing.
+Lowering \PwTmcaTITLE{1} to \armeight{} requires a full fence after every
+acquiring read.  To see why, consider the following attempted
+execution, where the final values of both $x$ and $y$ are $2$.
+\begin{gather*}
+  \PW{x}{2}\SEMI 
+  \PR[\mACQ]{x}{r}\SEMI
+  \PW{y}{r{-}1} \PAR
+  \PW{y}{2}\SEMI
+  \PW[\mREL]{x}{1}
+  \\
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a}{\DW{x}{2}}{}
+      \raevent{b}{\DR[\mACQ]{x}{2}}{right=of a}
+      \event{c}{\DW{y}{1}}{right=of b}
+      \event{d}{\DW{y}{2}}{right=2.5em of c}
+      \raevent{e}{\DW[\mREL]{x}{1}}{right=of d}
+      \rfint{a}{b}
+      \sync{b}{c}
+      \wk{c}{d}
+      \sync{d}{e}
+      \wk[out=-165,in=-15]{e}{a}
+    \end{tikzinline}}
+\end{gather*}
+The execution is  allowed by \armeight, but disallowed by \PwTmca{1}, due to
+the cycle.
 
-In the second semantics, we weaken the relationship between $\rrfx$
-and $\le$ in \ref{pom-rf-le}.  Rather than ensuring that there is no
+\armeight{} allows the execution because the read of $x$ is internal to the
+thread.  This aspect of \armeight{} semantics is difficult to model locally.
+To capture this, we found it necessary to drop \ref{pom-rf-le} and relax
+\ref{seq-le-delays}, %from
+% \refdef{def:pwt:mca1},
+adding local constraints on $\rrfx$ to $\sLPAR{}{}$, $\sSEMI{}{}$ and
+$\sIF{}{}{}$.
+Rather than ensuring that there is no
 \emph{global} blocker for a sequentially fulfilled read \eqref{pom-rf-le}, we
@@
 \eqref{seq-rf-le-rf}.
+For \PwTmca{2}, internal reads don't necessarily contribute to order, and
+thus the above execution is allowed.
+
+\begin{definition}
+  \label{def:pwt:mca2}
+  A \PwTmca{2} is a \PwT{} (\refdef{def:pomset}) equipped with an injective
+  relation $\rrfx$ that satisfies requirements \ref{pom-rf-match} and
+  \ref{pom-rf-block} of \refdef{def:pwt:mca1}.
+
+  A A \PwTmca{2} is \emph{complete} if it satisfies
+  \ref{top-kappa}, \ref{top-term}, and \ref{top-rf}.
+
+  \input{fig-mca2.tex}
+\end{definition}
+% We write $\semmca{2}{}$ for the semantic function.
+A \PwTmca{2} need not satisfy requirement \ref{pom-rf-le}, and thus we may
+have $\bEv\xrfx\aEv$ and $\aEv\le\bEv$.
+
+% 
+% The use of $\rextendsdef{}{}$ in \ref{if-rf-extends}
+% ensures that no $\rrfx$ is introduced between events in
+% $\aEvs_1\cap\aEvs_2$ when coalescing.
+
+% In the second semantics, we weaken the relationship between $\rrfx$
+% and $\le$ in \ref{pom-rf-le}.  
 % This change both allows and requires us to weaken the definition of
@@
 
-\begin{definition}
-  \label{def:pwt:mca2}
-  A \PwTmca{2} is a \PwT{} (\refdef{def:pomset}) equipped with an injective
-  relation $\rrfx$ that satisfies requirements \ref{pom-rf-match} and
-  \ref{pom-rf-block} of \refdef{def:pwt:mca1}.
+\PwTmca{2} does not enforce \ref{pom-rf-le}: $\bEv\xrfx\aEv$ may not imply
+$\bEv\le\aEv$ when $\bEv$ and $\aEv$ come from different sides of a
+sequential composition.  This means that $\rrfx$ must be verified during
+pomset construction, rather than post-hoc.  In \textsection\ref{sec:c11}, we
+show how to construct program order ($\rpox$) for complete pomsets using
+phantom events ($\fmrg{}$).  Using this construction, the following lemma
+gives a post-hoc verification technique for $\rrfx$.
+\begin{lemma}
+  If $\aPS\in\frf{\semmca{2}{\aCmd}}$ is complete, then
+  for every $\bEv\xrfx\aEv$ either
+  \begin{itemize}
+  \item external fulfillment:
+    $\bEv\le\aEv$ and if $\labeling(\cEv) \rblocks\labeling(\aEv)$ then either $\cEv\le\bEv$ or $\aEv\le\cEv$, or
+  \item internal fulfillment:
+    $(\exists\bEv'\in\fmrginv{\bEv})$
+    $(\exists\aEv'\in\fmrginv{\aEv})$
+    $\bEv'\xpox\aEv'$ and $(\not\exists\cEv')$
+    $\labelingForm(\cEv)$ is a tautology and
+    $\labeling(\cEv) \rblocks \labeling(\aEv)$ and $\bEv'\xpox\cEv\xpox\aEv'$.
+  \end{itemize}
+\end{lemma}
+These mimic the \emph{external consistency} requirements of \armeight{}
+\cite{armed}.
 
-  A A \PwTmca{2} is \emph{complete} if it satisfies
-  \ref{top-kappa}, \ref{top-term}, and \ref{top-rf}.
 % \begin{enumerate}[,label=(\textsc{m}\arabic*),ref=\textsc{m}\arabic*]
@@
   % \end{enumerate}
-\end{definition}
-A \PwTmca{2} need not satisfy requirement \ref{pom-rf-le}, and thus we may
-have $\bEv\xrfx\aEv$ and $\aEv\le\bEv$.
 
@@
 
-The semantic rules are given in \reffig{fig:mca2}.
-We write $\semmca{2}{}$ for the semantic function.
+%The semantic rules are given in \reffig{fig:mca2}.

### 2021-07-07 paper/tool.tex

@@
 Our choice of C11 rules is careful: we take the rules of RC11~\cite{DBLP:conf/pldi/LahavVKHD17}, but we replace the overly strong No-Thin-Air restriction $\text{acyclic}(\rrf\cup\rpox)$ with $\text{acyclic}(\rrf\cup\lt)$.
-This is a more precise categorisation of thin-air behavior, and it allows aggressive compiler optimisations which would be erroneously forbidden by RC11's original No-Thin-Air axiom.
+This is a more precise categorisation of thin-air behavior, and it allows aggressive compiler optimizations which would be erroneously forbidden by RC11's original No-Thin-Air axiom.
 We show \PwTer{} in action in Fig.~\ref{fig:tool}.

### 2021-07-07 paper/oopsla.tex
drf hacking
cocktail time. only one todo left, and it is in an appendix

--- 2021-06-28 paper/oopsla.tex
@@
-\subsection{Comparison to ``A Promising Semantics 2.1'' [POPL 2017]}
+\subsection{Comparison to ``Promising Semantics'' [POPL 2017]}
 \label{sec:promising}
 
-[Todo.]
+\citet{DBLP:journals/pacmpl/JagadeesanJR20} note that the promising semantics
+(\PS{}) \cite{DBLP:conf/popl/KangHLVD17} and related models
+\citep{DBLP:conf/esop/JagadeesanPR10,DBLP:journals/pacmpl/ChakrabortyV19,Manson:2005:JMM:1047659.1040336},
+fail to validate compositional reasoning of temporal properties.  Consider
+their example \ref{oota4}:
+\begin{align}
+  \taglabel{oota4}
+  \begin{gathered}
+    \PW{y}{x}
+    \PAR
+    \PR{y}{r} \SEMI \IF{b}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI
+    \PAR
+    \PW{b}{1}
+    \\[-1ex]
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{rx}{\DR{x}{1}}{}
+        \event{wy}{\DW{y}{1}}{right=of rx}
+        \po{rx}{wy}
+        \event{ry}{\DR{y}{1}}{right=3em of wy} 
+        \event{wx}{\DW{x}{1}}{right=of ry}
+        \event{wz}{\DW{z}{1}}{right=of wx}
+        \event{rb}{\DR{b}{1}}{right=of wz}
+        \event{wb1}{\DW{b}{1}}{right=3em of rb}
+        \po{ry}{wx}
+        \rf{wb1}{rb}
+        \rf{wy}{ry}
+        \rf[out=-170,in=-10]{wx}{rx}
+        \po{rb}{wz}
+        \po[out=15,in=165]{ry}{wz}
+      \end{tikzinline}}
+  \end{gathered}  
+\end{align}
+Under \PS{}, this outcome is allowed by \emph{baiting} with the \texttt{else}
+branch, then \emph{switching} to the \texttt{then} branch, based on a coin
+flip $(\IF{b})$.\footnote{Call the threads \texttt{s}, \texttt{t}, and
+  \texttt{u}.  To get the result in the promising semantics, first execute
+  \texttt{u} to get message \texttt{<b:1@1>}.  Then \texttt{t} promises
+  \texttt{<x:1@1>}, which it can fulfill by reading \texttt{b}$=$\texttt{0}.
+  Then execute \texttt{s} to get message \texttt{<y:1@1>}.  Then execute
+  \texttt{t}, reading \texttt{b}$=$\texttt{1} and \texttt{y}$=$\texttt{1} and
+  fulfill the promise by writing \texttt{<x:1@1>}. The execution is exactly
+  the same in our speculative semantics \cite{DBLP:conf/esop/JagadeesanPR10},
+  removing timestamps and replacing the word \emph{promise} by
+  \emph{speculation}.}
 
-Case analysis gives very weak results when combined with thread inlining.
-See \cite[\textsection B.1]{DBLP:journals/pacmpl/ChakrabortyV19appendix}.
-These happen by performing transformations that: 
-(1) introduce conditionals,
-(2) inline two threads on both sides of the introduced conditional,
-(3) choose different orders for the two threads for the two sides of the conditional.
+Under all variants of \PwT{}, this outcome is disallowed, due to a cycle in
+${\rrfx}\cup{\lt}$. (Note that all reads cross threads in the example, so
+there is no difference between \PwTmca{1} and \PwTmca{2}.)
 
-Case analysis gives very weak results when combined with read introduction.
-See \cite{promising-ldrf}.
-These happen by performing transformations that: 
-(1) introduce reads,
-(2) introduce conditionals,
-(3) choose different values for the reads on the two sides of the conditional.
+\citet[Fig.~8]{DBLP:journals/toplas/Lochbihler13} notes that such violations
+of temporal reasoning could result in violations of the Java security
+architecture.
 
-The fact that the semantics is not verifiable a posteriori is something it
-shares with \weakestmo{}, where the justification relation must be built
-inductively.
+The difference between \PS{} and \PwT{} can be understood in terms of the
+valid program transformations.  \PS{} allows reads to be introduced, with
+subsequent case analysis (aka, if-closure) on the value read---effectively,
+this can turn one read into two, with different conditional branches taken
+for the two copies of the read.  For detailed examples, see
+\cite{promising-ldrf}.
 
-\weakestmo{} admits FADD, but \PS{} does not.
-\PS{} admits CohCYC, but \weakestmo{} does not.
+\PwT{} invalidates read introduction.  In return, \PwT{} enjoys
+compositionality for temporal safety properties, as discussed in
+\textsection\ref{sec:results} and
+\cite[\textsection6]{DBLP:journals/pacmpl/JagadeesanJR20}.
+
+This difference highlights the subtle tensions in relaxed memory models.
+It is not possible to have everything one wants, thus, one is forced
+to choose which optimizations and reasoning principles are most important.
+
+% \todo{Write this.}
+
+% Case analysis gives very weak results when combined with thread inlining.
+% See \cite[\textsection B.1]{DBLP:journals/pacmpl/ChakrabortyV19appendix}.
+% These happen by performing transformations that: 
+% (1) introduce conditionals,
+% (2) inline two threads on both sides of the introduced conditional,
+% (3) choose different orders for the two threads for the two sides of the conditional.
+
+% Case analysis gives very weak results when combined with read introduction.
+% See \cite{promising-ldrf}.
+% These happen by performing transformations that: 
+% (1) introduce reads,
+% (2) introduce conditionals,
+% (3) choose different values for the reads on the two sides of the conditional.
+
+% The fact that the semantics is not verifiable a posteriori is something it
+% shares with \weakestmo{}, where the justification relation must be built
+% inductively.
+
+% \weakestmo{} admits FADD, but \PS{} does not.
+% \PS{} admits CohCYC, but \weakestmo{} does not.
 
@@
 
+\endinput
+
 \myparagraph{Parallel Composition}

### 2021-07-07 paper/outro.tex
a stab at DRF
cocktail time. only one todo left, and it is in an appendix

@@
-\section{Future Work}
+\section{Conclusions} %Future Work}
 
@@
 We have not treated loops in this model, though we expect that the usual
-approach of showing continuity for all the semantic operations with respect to set inclusion
-would go through. \citet{DBLP:conf/esop/PaviottiCPWOB20} use step-indexing to account for
-loops; a similar approach could be applied here.
+approach of showing continuity for all the semantic operations with respect
+to set inclusion would go through. \citet{DBLP:conf/esop/PaviottiCPWOB20} use
+step-indexing to account for loops; a similar approach could be applied here.
 
-In \S\ref{sec:arm1} we presented a compilation strategy to \armeight{}
-for a simplified model, but which introduces fences to acquiring
-reads. These fences are not required in \S\ref{sec:arm2}, but at the
-cost of model complexity. It would be illuminating to find out what
+We've also not handled access elimination: store-forwarding and
+dead-write-removal are unsound.  We expect that these can be validated by
+allowing events with different labels to merge in some instances.
+
+\PwTmca{1} is a simpler model than \PwTmca{2}, but requires fences on
+acquiring reads for \armeight{}.  It would be illuminating to find out what
 the performance penalty is for these fences.
 
-The promising semantics (\PS{}) validates read introduction whereas \PwT{}
-does not.  As a result \PS{} admits behaviors that break reasoning about
-temporal safety properties (see \textsection\ref{sec:promising}).
-Nonetheless, read introduction is ubiquitous is some compilers.  It would be
-interesting to know if there is a performance penalty for banning read
-introduction.
+In a similar vein, the promising semantics (\PS{}) validates read
+introduction whereas \PwTmca{} does not.  As a result \PS{} admits behaviors
+that break reasoning about temporal safety properties (see
+\textsection\ref{sec:promising}).  Nonetheless, read introduction is
+ubiquitous is some compilers.  It would be interesting to know if there is a
+performance penalty for banning read introduction.
 
-An earlier version of this paper has been mechanized in Agda; it would be
-reassuring to update the mechanization to bring it in line with the current state.
-
-We don't handle access elimination.
+% An earlier version of this paper has been mechanized in Agda; it would be
+% reassuring to update the mechanization to bring it in line with the current state.
 

### 2021-07-07 paper/results.tex
edits
cocktail time. only one todo left, and it is in an appendix

@@
-\section{Results}
+\section{\PwTmcaTITLE{} Results}
 \label{sec:results}
 
-% \subsection{Reordering Transformations}
-% \label{sec:ex:valid}
+\PwP{} \cite{DBLP:journals/pacmpl/JagadeesanJR20} is a novel memory model,
+intended to serve as a semantic basis for a Java-like language, where all
+access is safe.  \PwTmca{} generalizes \PwP{}, making several small but
+significant changes.  As a result, we have had to re-prove most of the
+theorems from \PwP{}.
 
-% When $\sem{\aCmd} \supseteq \sem{\aCmd'}$, we say that $\aCmd'$ is a
-% \emph{valid transformation} of $\aCmd$.
-% In this subsection, we show the
-% validity of specific optimizations.
+In \textsection\ref{sec:arm}, we show that \PwTmca{1} supports the optimal
+lowering of relaxed accesses to \armeight{} and that \PwTmca{2} supports the
+optimal lowering of \emph{all} accesses to \armeight{}.  The proofs are based
+on two recent characterizations of \armeight{} \cite{armed}.  For \PwTmca{1},
+we use \emph{External Global Consistency}.  For \PwTmca{2}, we use
+\emph{External Consistency}.
 
-The semantics validates many peephole optimizations.  Most apply only to
-relaxed access.
+In \textsection\ref{sec:sc}, we prove sequential consistency for
+local-data-race-free programs.  The proof uses \emph{program order}, which we
+construct for \cXI{} in \textsection\ref{sec:c11}.  The same construction
+works for \PwTmca{}.  (This proof assumes there are no \RMW{} operations.)
+
+The semantics validates many peephole optimizations, such as the standard reorderings on relaxed access:
 \begin{align*}
@@
 \begin{align*}
-    %\tag{\textsc{roach1}}\label{AcqW}
+  % \tag{\textsc{roach1}}\label{AcqW}
   \sembase{x^\amode \GETS \aExp \SEMI\PR{y}{\bReg}} &\supseteq
@@
   \\
-    %\tag{\textsc{roach2}}\label{RelW}
+  % \tag{\textsc{roach2}}\label{RelW}
   \sembase{x \GETS \aExp \SEMI\PR[\amode]{y}{\bReg}} &\supseteq
@@
 \end{align*}
+
+Notably, the semantics does \emph{not} validate read introduction.  When
+combined with case analysis (\textsection\ref{sec:semca}), read introduction
+can break temporal reasoning.  This combination is allowed by speculative
+operational models.  See \textsection\ref{sec:promising} for a discussion.
+
+Prop.~6.1 of \cite{DBLP:journals/pacmpl/JagadeesanJR20} establishes a
+compositional principle for proving that programs validate formula in
+past-time temporal logic.  The principal is based entirely on the pomset
+order relation.  It's proof, and all of the no-thin-air examples in
+\cite[\textsection6]{DBLP:journals/pacmpl/JagadeesanJR20} hold equally for
+the models described here.
 

### 2021-07-07 paper/tool.tex

@@
 
-\PwTer{} automatically and exhaustively calculates the allowed outcomes of litmus tests for the PwT, PwT-po and PwT-C11 models.
+\PwTer{} automatically and exhaustively calculates the allowed outcomes of litmus tests for the \PwT, \PwTpo, and \PwTc{} models.
 It is built in OCaml, and uses Z3~\cite{Z3Solver} to judge the truth of predicates constructed by the models.
@@
 
-\PwTer{} allows several modes of evaluation: it can evaluate the rules of Fig.~\ref{fig:seq}, implementing PwT; it can generate program order according to Section~\ref{sec:c11}, implementing PwwT-po; and similar to \MRD~\cite{DBLP:conf/esop/PaviottiCPWOB20}, it can construct C11-style pre-executions and filter them according to the rules of C11.
+\PwTer{} allows several modes of evaluation: it can evaluate the rules of Fig.~\ref{fig:seq}, implementing \PwT; it can generate program order according to Section~\ref{sec:c11}, implementing \PwTpo; and similar to \MRD~\cite{DBLP:conf/esop/PaviottiCPWOB20}, it can construct C11-style pre-executions and filter them according to the rules of C11.
 Our choice of C11 rules is careful: we take the rules of RC11~\cite{DBLP:conf/pldi/LahavVKHD17}, but we replace the overly strong No-Thin-Air restriction $\text{acyclic}(\rrf\cup\rpox)$ with $\text{acyclic}(\rrf\cup\lt)$.

### 2021-07-07 paper/intro.tex
cocktail time. only one todo left, and it is in an appendix
added some comments from mark and simon

@@
 which provides a Java-like model for \mca{} hardware, similar to that of
-\citet{DBLP:journals/pacmpl/JagadeesanJR20}.  In \textsection\ref{sec:c11},
+\citet{DBLP:journals/pacmpl/JagadeesanJR20}.  In
+\textsection\ref{sec:results}, we summarize the results for this model. In \textsection\ref{sec:c11},
 we define \PwTc{}, which models \cXI, adapting the approach of
@@
 we extend the semantics to include additional features, such as address
-calculation.  In \textsection\ref{sec:results}, we summarize our results.
+calculation.
 
@@
   \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
-  \\[-2ex]
+  \\
   \nonumber
-  \hbox{\begin{tikzinline}[node distance=1ex and 1.2em]
-      \xform{rx0d}{(0{=}r) \limplies \bForm}{}
-      \event{rx0}{\DR{x}{0}}{right=of rx0d}
-      \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{right=of rx0}
-      \event{ry0}{\DR{y}{0}}{right=of rx0ry0d}
-      \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
-      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0d}
-      \xform{wz0i}{\bForm}{left=of rx0d}
-      \po[out=20,in=170]{ry0}{wz0}
+  \hbox{\begin{tikzinline}[node distance=2ex and 1.2em]
+      \xform{wz0i}{\bForm}{}
+      \xform{rx0d}{(0{=}r) \limplies \bForm}{right=of wz0i}
+      \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{right=of rx0d}
+      \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of rx0ry0d}
+      \event{rx0}{\DR{x}{0}}{above left=of rx0ry0d}
+      \event{ry0}{\DR{y}{0}}{above right=of rx0ry0d}
+      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
+      \po{ry0}{wz0}
       \xo{rx0}{rx0d}
@@
 \end{gather*}
+In the diagram, the dotted lines indicate set inclusion into the index of the transformer-family.
 As a sanity check, we can see that sequential composition is
@@
   \nonumber
-  \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
+  \hbox{\begin{tikzinline}[node distance=2ex and 1.5em]
       \xform{rx0d}{(0{=}r) \limplies \bForm}{}
@@
 
-\subsection{Related Work}
+\section{Related Work}
 

### 2021-07-08 paper/tool.tex
Remove duplication between C11 and Tool sections

@@
 
-\PwTer{} allows several modes of evaluation: it can evaluate the rules of Fig.~\ref{fig:seq}, implementing \PwT; it can generate program order according to Section~\ref{sec:c11}, implementing \PwTpo; and similar to \MRD~\cite{DBLP:conf/esop/PaviottiCPWOB20}, it can construct C11-style pre-executions and filter them according to the rules of C11.
-Our choice of C11 rules is careful: we take the rules of RC11~\cite{DBLP:conf/pldi/LahavVKHD17}, but we replace the overly strong No-Thin-Air restriction $\text{acyclic}(\rrf\cup\rpox)$ with $\text{acyclic}(\rrf\cup\lt)$.
-This is a more precise categorisation of thin-air behavior, and it allows aggressive compiler optimizations which would be erroneously forbidden by RC11's original No-Thin-Air axiom.
+\PwTer{} allows several modes of evaluation: it can evaluate the rules of Fig.~\ref{fig:seq}, implementing \PwT; it can generate program order according to \textsection~\ref{sec:c11}, implementing \PwTpo; and similar to \MRD~\cite{DBLP:conf/esop/PaviottiCPWOB20}, it can construct C11-style pre-executions and filter them according to the rules of \rcXI{} as described in \textsection\ref{sec:c11}.
+Finally, \PwTer{} also allows us to toggle the complete check of~\ref{def:complete}, providing an interface for understanding how fragments of code might compose by exposing preconditions and termination conditions that are not yet tautologies.
 We show \PwTer{} in action in Fig.~\ref{fig:tool}. 
-Finally, \PwTer{} also allows us to toggle the complete check of~\ref{def:pomset}, providing an interface for understanding how fragments of code might compose by exposing preconditions and termination conditions which are not yet tautologies.
+\PwTer{} will be made open source upon publication.
+%\PwTer{} is available online at \url{https://github.com/graymalkin/pomsets-with-predicate-transformers}.
 

### 2021-07-08 paper/tool.tex
Some commentary on C11 coherence/dependency interaction

@@
 
-\PwTer{} allows several modes of evaluation: it can evaluate the rules of Fig.~\ref{fig:seq}, implementing \PwT; it can generate program order according to \textsection~\ref{sec:c11}, implementing \PwTpo; and similar to \MRD~\cite{DBLP:conf/esop/PaviottiCPWOB20}, it can construct C11-style pre-executions and filter them according to the rules of \rcXI{} as described in \textsection\ref{sec:c11}.
+\PwTer{} allows several modes of evaluation: it can evaluate the rules of Fig.~\ref{fig:seq}, implementing \PwT; it can generate program order according to \textsection\ref{sec:c11}, implementing \PwTpo; and similar to \MRD~\cite{DBLP:conf/esop/PaviottiCPWOB20}, it can construct C11-style pre-executions and filter them according to the rules of \rcXI{} as described in \textsection\ref{sec:c11}.
 Finally, \PwTer{} also allows us to toggle the complete check of~\ref{def:complete}, providing an interface for understanding how fragments of code might compose by exposing preconditions and termination conditions that are not yet tautologies.

### 2021-07-08 paper/tool.tex
Comment out tool screenshot & reference sentence Not enough space D:

@@
 Finally, \PwTer{} also allows us to toggle the complete check of~\ref{def:complete}, providing an interface for understanding how fragments of code might compose by exposing preconditions and termination conditions that are not yet tautologies.
-We show \PwTer{} in action in Fig.~\ref{fig:tool}. 
+% We show \PwTer{} in action in Fig.~\ref{fig:tool}. 
 \PwTer{} will be made open source upon publication.
@@
 
-%% TODO: introduce some nice dogma about the value of a tool and what it's taught us.
-%% The real friends are the tools we built along the way.
-
-%% Notes for Simon and Mark:
-% Intro
-%  - Add contribution relating to tool
-
-% Related work
-%  - Mention tool for MRD & Event Structures, mention absence of tool for Promising
-
-% This section
-%  - tool for {PwT, PwT-PO, PwT-RC11}
-%  - Validation of some tests? Which tests to include?
-
-% C11 Section
-%  - Discussion: C11 standard really needs a dep relation. MRD provides this, PwT provides this, Promising does not.
-%  - There's an open question about how coherence should affect dep.
-%    if(q) { x := 1; x := 2 } else { x := 2; x := 1 }
-%  - What should the deps be?
-
-% Conclusion
-%  - Having a tool should be some kind of feel-good factor.
-
 %% [Simon] I am not a fan of how this looks on the page.
-\begin{figure}[t]
-\begin{center}
-  \includegraphics[width=0.75\textwidth]{tool.png}
-  \Description{Image shows tool running on a Mac.
+% \begin{figure}[t]
+% \begin{center}
+%   \includegraphics[width=0.75\textwidth]{tool.png}
+%   \Description{Image shows tool running on a Mac.
 Java Causality Test Case 1 is printed, followed by an invocation of the tool as {\tt ./pomsets.exe --check --complete data/tests/jctc/jctc1.lit}, which shows that the tool passes the assertion listed in the test.}
-  \caption{\label{fig:tool} Example output of \PwTer, validating TC1~\cite{PughWebsite}.}
-\end{center}
-\end{figure}
+%   \caption{\label{fig:tool} Example output of \PwTer, validating TC1~\cite{PughWebsite}.}
+% \end{center}
+% \end{figure}

### 2021-07-08 paper/intro.tex
edits

@@
 sequential composition using continuation-passing, and \PwP{} used prefixing,
-adding one event at a time on the left.  In both cases, the approaches
-require perfect knowledge of the future.
+adding one event at a time on the left.  In both cases, adding an event
+requires perfect knowledge of the future.
 
 In this paper, we show that \PwP{} can be extended with \emph{families of
-  predicate transformers (\PwT{})} to calculate sequential dependencies in a way that
+  predicate transformers} (\PwT{}) to calculate sequential dependencies in a way that
 is \emph{compositional} and \emph{direct}: \emph{compositional} in that the
@@
 
-To manage complexity, we have layered the definitions.  After an overview, we
-define sequential dependencies in \textsection\ref{sec:model}.  The next two
-sections add concurrency.  In \textsection\ref{sec:mca}, we define \PwTmca{},
+To manage complexity, we have layered the definitions.  After an overview and
+discussion of related work, we
+define sequential dependencies in \textsection\ref{sec:model}.  We then add
+concurrency.  In \textsection\ref{sec:mca}, we define \PwTmca{},
 which provides a Java-like model for \mca{} hardware, similar to that of
-\citet{DBLP:journals/pacmpl/JagadeesanJR20}.  In
-\textsection\ref{sec:results}, we summarize the results for this model. In \textsection\ref{sec:c11},
+\citet{DBLP:journals/pacmpl/JagadeesanJR20};
+\textsection\ref{sec:results} summarizes the results for this model. In \textsection\ref{sec:c11},
 we define \PwTc{}, which models \cXI, adapting the approach of
-\citet{DBLP:conf/esop/PaviottiCPWOB20}.  In \textsection\ref{sec:tool} we 
-provide a tool for automatic evaluation of litmus tests.  In \textsection\ref{sec:additional},
+\citet{DBLP:conf/esop/PaviottiCPWOB20};  \textsection\ref{sec:tool} 
+describes a tool for automatic evaluation of litmus tests.  In \textsection\ref{sec:additional},
 we extend the semantics to include additional features, such as address
@@
 ($\aLoc$--$\cLoc$), the obvious generalization
-\begin{multicols}{2}
+%\begin{multicols}{2}
 \begin{itemize}
-\item[{\labeltext[\textsc{d}2b]{(\textsc{d}2b)}{wp-write}}] $\fwp{\LET{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
-\item[{\labeltext[\textsc{d}2c]{(\textsc{d}2c)}{wp-read}}] $\fwp{\LET{\aReg}{\aLoc}}{\bForm} = \bForm[\aLoc/\aReg]$
+\item[{\labeltext[\textsc{d}2a]{(\textsc{d}2a)}{wp-write}}] $\fwp{\LET{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
+\item[{\labeltext[\textsc{d}2b]{(\textsc{d}2b)}{wp-read}}] $\fwp{\LET{\aReg}{\aLoc}}{\bForm} = \bForm[\aLoc/\aReg]$
 \end{itemize}
-\end{multicols}
+%\end{multicols}
 \noindent

### 2021-07-08 paper/outro.tex
edits
Add the image back for the tool, add a reference to PwT-C11 in Conclusion

@@
 sequential composition in a relaxed memory model which can be
-efficiently compiled to modern CPUs. There is, as usual, more research
-to be done.
+efficiently compiled to modern CPUs. We extract from this model a semantic dependency relation and use it to build \PwTc, a solution to the Out-of-Thin-Air problem in \cXI. There is, as usual,
+more research to be done.
 
@@
 
-We've also not handled access elimination: store-forwarding and
+\PwTmca{} does not validate access elimination: store-forwarding and
 dead-write-removal are unsound.  We expect that these can be validated by

### 2021-07-08 paper/oopsla.tex
Submission

@@
+\section{Discussion}
+\label{sec:discussion}
+
 \subsection{Comparison to ``Promising Semantics'' [POPL 2017]}
@@
 
-\citet{DBLP:journals/pacmpl/JagadeesanJR20} note that the promising semantics
-(\PS{}) \cite{DBLP:conf/popl/KangHLVD17} and related models
+Recently, \citet{promising-ldrf} showed that certain combinations of compiler
+optimizations are inconsistent with local \drf{} guarantees.  All of the
+examples that prove inconsistency have the same shape: they combine read
+introduction and case analysis (aka, if-closure).  Effectively, this turns
+one read into two, where different conditional branches can be taken for the
+two copies of the read.  This is reminiscent of the type of \emph{bait and
+  switch} behavior noted by \citet{DBLP:journals/pacmpl/JagadeesanJR20}: the
+promising semantics (\PS{}) \cite{DBLP:conf/popl/KangHLVD17} and related
+models
 \citep{DBLP:conf/esop/JagadeesanPR10,DBLP:journals/pacmpl/ChakrabortyV19,Manson:2005:JMM:1047659.1040336},
@@
     \PAR
-    \PR{y}{r} \SEMI \IF{b}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI
+    \PR{y}{r} \SEMI \PR{b}{s} \SEMI\IF{s}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI
     \PAR
@@
 branch, then \emph{switching} to the \texttt{then} branch, based on a coin
-flip $(\IF{b})$.\footnote{Call the threads \texttt{s}, \texttt{t}, and
-  \texttt{u}.  To get the result in the promising semantics, first execute
-  \texttt{u} to get message \texttt{<b:1@1>}.  Then \texttt{t} promises
-  \texttt{<x:1@1>}, which it can fulfill by reading \texttt{b}$=$\texttt{0}.
-  Then execute \texttt{s} to get message \texttt{<y:1@1>}.  Then execute
-  \texttt{t}, reading \texttt{b}$=$\texttt{1} and \texttt{y}$=$\texttt{1} and
-  fulfill the promise by writing \texttt{<x:1@1>}. The execution is exactly
-  the same in our speculative semantics \cite{DBLP:conf/esop/JagadeesanPR10},
-  removing timestamps and replacing the word \emph{promise} by
-  \emph{speculation}.}
+flip $(\IF{b})$.
+% Using a combination of read introduction and case analysis,
+% the middle thread can be rewritten to:
+% \begin{displaymath}
+%   \begin{array}{l}
+%     \PR{y}{r} \SEMI \IF{b}\THEN\\
+%     \quad \IF{b}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI\\
+%     \ELSE\\
+%     \quad \IF{b}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI\\
+%     \FI
+%   \end{array}
+% \end{displaymath}
+% \footnote{Call the threads \texttt{s}, \texttt{t}, and
+%   \texttt{u}.  To get the result in the promising semantics, first execute
+%   \texttt{u} to get message \texttt{<b:1@1>}.  Then \texttt{t} promises
+%   \texttt{<x:1@1>}, which it can fulfill by reading \texttt{b}$=$\texttt{0}.
+%   Then execute \texttt{s} to get message \texttt{<y:1@1>}.  Then execute
+%   \texttt{t}, reading \texttt{b}$=$\texttt{1} and \texttt{y}$=$\texttt{1} and
+%   fulfill the promise by writing \texttt{<x:1@1>}. The execution is exactly
+%   the same in our speculative semantics \cite{DBLP:conf/esop/JagadeesanPR10},
+%   removing timestamps and replacing the word \emph{promise} by
+%   \emph{speculation}.}
+% \citet[Fig.~8]{DBLP:journals/toplas/Lochbihler13} notes that such violations
+% of temporal reasoning could result in violations of the Java security
+% architecture.
 
@@
 
-\citet[Fig.~8]{DBLP:journals/toplas/Lochbihler13} notes that such violations
-of temporal reasoning could result in violations of the Java security
-architecture.
+The examples in \cite{promising-ldrf} are more complex.  Here is one, dubbed
+\textsc{ldrf-fail-ps}:
+\begin{gather*}  
+  %\taglabel{LDRF-Fail-PS}
+  \begin{gathered}
+  \IF{x}\THEN
+    \PFADD{w}{}{1}\SEMI
+    \PW{y}{1}\SEMI
+    \PW{z}{1}
+  \FI
+  \PAR
+  \IF{\BANG z}\THEN
+    \PW{x}{1}
+  \ELSE
+    \IF{\BANG \PFADD{w}{}{1}}\THEN
+      \PW{x}{\PR{y}{}}
+    \FI
+  \FI
+    \\
+    \hbox{\begin{tikzinlinesmall}[node distance=2em]
+        \event{a1}{\DR{x}{1}}{}
+        \event{a2}{\DR{w}{1}}{right=of a1}
+        \event{a3}{\DW{w}{2}}{right=of a2}
+        \event{a4}{\DW{y}{1}}{right=of a3}
+        \event{a5}{\DW{z}{1}}{right=of a4}
+        \event{b1}{\DR{z}{1}}{right=5em of a5}
+        \event{b2}{\DR{w}{0}}{right=of b1}
+        \event{b3}{\DW{w}{1}}{right=of b2}
+        \event{b4}{\DR{y}{1}}{right=of b3}
+        \event{b5}{\DW{x}{1}}{right=of b4}
+        \rmw{a2}{a3}
+        \po[out=15,in=165]{a1}{a3}
+        \po[out=15,in=165]{a1}{a4}
+        \po[out=15,in=165]{a1}{a5}        
+        \rmw{b2}{b3}
+        \po{b4}{b5}
+        \po[out=15,in=165]{b2}{b5}        
+        \po[out=15,in=165]{b1}{b3}
+        \rf{a5}{b1}
+        \rf[out=15,in=165]{a4}{b4}
+        \rf[out=-165,in=-15]{b3}{a2}
+        \rf[out=-165,in=-15]{b5}{a1}
+      \end{tikzinlinesmall}}
+  \end{gathered}
+\end{gather*}
+This execution is allowed by \PS{} and disallowed by \PwTmca{}.
 
@@
 valid program transformations.  \PS{} allows reads to be introduced, with
-subsequent case analysis (aka, if-closure) on the value read---effectively,
-this can turn one read into two, with different conditional branches taken
-for the two copies of the read.  For detailed examples, see
-\cite{promising-ldrf}.
+subsequent case analysis on the value read.  \PwT{} validates case analysis,
+but invalidates read introduction.  In return, \PwT{} enjoys compositionality
+for temporal safety properties, as discussed in \textsection\ref{sec:results}
+and \cite[\textsection6]{DBLP:journals/pacmpl/JagadeesanJR20}.
 
-\PwT{} invalidates read introduction.  In return, \PwT{} enjoys
-compositionality for temporal safety properties, as discussed in
-\textsection\ref{sec:results} and
-\cite[\textsection6]{DBLP:journals/pacmpl/JagadeesanJR20}.
+\citet{promising-ldrf} show that by restricting \RMW{}-store reorderings, one
+can establish local \drf{} for \PS{}.  We believe that no such restriction is
+required for \PwT{}.
 
-This difference highlights the subtle tensions in relaxed memory models.
-It is not possible to have everything one wants, thus, one is forced
-to choose which optimizations and reasoning principles are most important.
+These differences highlight the subtle tensions between compiler optimizations
+and program logics that are revealed by relaxed memory models.  It is not
+possible to have everything one wants, thus, one is forced to choose which
+optimizations and reasoning principles are most important.
 
@@
 The substitutions collapse $\aLoc$ and $\aReg$, allowing local invariant
-reasoning (\xLIR{}), as required by \jmm{} causality test case 1, discussed at the end of
-\textsection\ref{sec:ex:control}.  Without Skolemizing it is necessary to
+reasoning (\xLIR{}), as required by \jmm{} causality test case 1, discussed in
+\textsection\ref{sec:q}.  Without Skolemizing it is necessary to
 substitute $[\aLoc/\aReg]$, since the reverse substitution $[\aReg/\aLoc]$ is
@@
 \end{align*}
-
-\myparagraph{Augmentation of Preconditions}
-\jjr{} allows augmentation of preconditions.
-As discussed in \textsection\ref{sec:delay}, this causes associativity to fail
-for $\rdelay$, at least when attempting to validate \reflem{lem:if}\eqref{lem:ifelse:if:if1}--\eqref{lem:ifelse:if:if2}
-Thus, we use \emph{weakest} preconditions, rather than general preconditions.
-As a result, we fail to validate the following
-refinement:
-\begin{math}
-  \aPSS_1
-  \not\supseteq
-  \xIFTHEN{\aForm}{\aPSS_1}{}.
-\end{math}
 
@@
 \end{gather*}
-We discussed two approaches to this problem in \textsection\ref{sec:arm}.
+We discuss two approaches to this problem in \textsection\ref{sec:arm}.
 % The solution we have adopted is to allow an acquiring read to be downgraded
@@
 Contrary to the claim, redundant read elimination fails for \jjr{}.
-We discussed redundant read elimination in \textsection\ref{sec:semreg}.
-Consider JMM Causality Test Case 2, which we discussed there.
+We discuss redundant read elimination in \textsection\ref{sec:semreg}.
+Consider JMM Causality Test Case 2, which we describe there.
 \begin{gather*}
@@
 in \jjr{};
+
+\myparagraph{Augmentation of Preconditions}
+\jjr{} allows augmentation of preconditions.
+As discussed in \textsection\ref{sec:delay}, this causes associativity to fail
+for $\rdelay$, at least when attempting to validate \reflem{lem:if}\eqref{lem:ifelse:if:if1}--\eqref{lem:ifelse:if:if2}
+Thus, we use \emph{weakest} preconditions, rather than general preconditions.
+As a result, we fail to validate the following
+refinement:
+\begin{math}
+  \aPSS_1
+  \not\supseteq
+  \xIFTHEN{\aForm}{\aPSS_1}{}.
+\end{math}
 

### 2021-07-08 paper/outro.tex
Submission

--- 2021-07-07 paper/outro.tex
@@
 sequential composition in a relaxed memory model which can be
-efficiently compiled to modern CPUs. There is, as usual, more research
-to be done.
+efficiently compiled to modern CPUs. We extract from this model a semantic
+dependency relation and use it to build \PwTc, a solution to the
+Out-of-Thin-Air problem in \cXI, and \PwTmca{}, a model for Java-like
+languages.
+% There is, as usual, more research to be done.
 
@@
 
-We've also not handled access elimination: store-forwarding and
-dead-write-removal are unsound.  We expect that these can be validated by
-allowing events with different labels to merge in some instances.
+% \PwTmca{} does not validate access elimination: store-forwarding and
+% dead-write-removal are unsound.  We expect that these can be validated by
+% allowing events with different labels to merge in some instances.
 
-\PwTmca{1} is a simpler model than \PwTmca{2}, but requires fences on
-acquiring reads for \armeight{}.  It would be illuminating to find out what
-the performance penalty is for these fences.
+% \PwTmca{1} is a simpler model than \PwTmca{2}, but requires fences on
+% acquiring reads for \armeight{}.  It would be illuminating to find out what
+% the performance penalty is for these fences.
 
-In a similar vein, the promising semantics (\PS{}) validates read
-introduction whereas \PwTmca{} does not.  As a result \PS{} admits behaviors
-that break reasoning about temporal safety properties (see
+% In a similar vein,
+\PwT{} does not validate read introduction, whereas speculative operational
+semantics do.  Recent work shows a tension between read introduction and
+compositional reasoning for temporal safety properties (see
 \textsection\ref{sec:promising}).  Nonetheless, read introduction is

### 2021-07-09 paper/mca1.tex
cocktail time. only one todo left, and it is in an appendix
added some comments from mark and simon
edits
typos

@@
 \armeight{}.  To achieve this, \PwTmca{2} drops the global requirement that
-\emph{reads-from} implies pomset order \eqref{pom-rf-le}.
+\emph{reads-from} implies pomset order \eqref{pom-rf-le}.  The models are the
+same, except for \emph{internal reads}, where a thread reads its own write.
 
@@
   \end{multicols}
-  \smallskip
+  \bigskip
 
@@
 The semantic rules are mostly straightforward: Parallel composition is disjoint
-union, and all constructs respect reads-from.  Only \ref{seq-le-delays}
-requires explanation.
+union, and all constructs respect reads-from.  The monoid laws
+(\reflem{lem:monoid}) extend to parallel composition, with $\SKIP$ as right
+unit only due to the asymmetry of \ref{par-tau}.
 
+Only \ref{seq-le-delays} requires explanation.
 % \begin{lemma}
@@
 %   $\bEv\le\aEv$.
-
 %   \vspace{-.5\baselineskip}
@@
 %   variant semantics $\frf{\semrr{}}$ where \ref{pom-rf-le} is not required.}
-
 From \refdef{def:actions}, recall that $\aAct \rdelaysdef \bAct$ if
@@
 \end{displaymath}
-The culprit is $\rdelay$, which introduces order regardless whether
+The culprit is $\rdelay$, which introduces order regardless of whether
 preconditions are disjoint.  As an example, 

### 2021-07-09 paper/oopsla.tex
typos

@@
 fail to validate compositional reasoning of temporal properties.  Consider
-their example \ref{oota4}:
+example \ref{oota4} from \cite{DBLP:journals/pacmpl/JagadeesanJR20}:
 \begin{align}
@@
 but invalidates read introduction.  In return, \PwT{} enjoys compositionality
-for temporal safety properties, as discussed in \textsection\ref{sec:results}
-and \cite[\textsection6]{DBLP:journals/pacmpl/JagadeesanJR20}.
+for temporal safety properties, as discussed in \textsection\ref{sec:results}.
 
@@
 and program logics that are revealed by relaxed memory models.  It is not
-possible to have everything one wants, thus, one is forced to choose which
+possible to have everything one wants. Thus, one is forced to choose which
 optimizations and reasoning principles are most important.

### 2021-07-09 paper/outro.tex
typos

@@
 This paper is the first to present a direct denotational semantics for
-sequential composition in a relaxed memory model which can be
+sequential composition in a relaxed memory model that can be
 efficiently compiled to modern CPUs. We extract from this model a semantic

### 2021-07-15 paper/mca2.tex
typos and cleaning up notes

@@
 require only that there is no \emph{thread-local} blocker
-\eqref{seq-rf-le-rf}.
+\eqref{seq-le-rf-rf}.
 For \PwTmca{2}, internal reads don't necessarily contribute to order, and
@@
 
-  A A \PwTmca{2} is \emph{complete} if it satisfies
-  \ref{top-kappa}, \ref{top-term}, and \ref{top-rf}.
+  A \PwTmca{2} is \emph{complete} if it satisfies
+  \ref{top-kappa}, \ref{top-term}, and \ref{top-rf}---this is the same as for
+  \PwTmca{1}.
 
@@
 
-With the weakening of \ref{seq-rf-le-rf}, we must be careful not to allow
+\todo{Example using \ref{seq-le-delays-rf} and \ref{seq-le-rf-rf}. To make
+  space, \reflem{lem:mca2} could move to appendix.}
+
+With the weakening of \ref{seq-le-delays}, we must be careful not to allow
 spurious pairs to be added to the $\rrfx$ relation.  Thus we add
@@
 
-\PwTmca{2} does not enforce \ref{pom-rf-le}: $\bEv\xrfx\aEv$ may not imply
-$\bEv\le\aEv$ when $\bEv$ and $\aEv$ come from different sides of a
-sequential composition.  This means that $\rrfx$ must be verified during
+% \PwTmca{2} does not enforce \ref{pom-rf-le}: $\bEv\xrfx\aEv$ may not imply
+% $\bEv\le\aEv$ when $\bEv$ and $\aEv$ come from different sides of a
+% sequential composition.  This means that
+As a consequence of dropping \ref{pom-rf-le}, sequential $\rrfx$ must be validated during
 pomset construction, rather than post-hoc.  In \textsection\ref{sec:c11}, we
@@
 \begin{lemma}
+  \label{lem:mca2}
   If $\aPS\in\frf{\semmca{2}{\aCmd}}$ is complete, then

### 2021-07-15 paper/outro.tex
typos and cleaning up notes

@@
 
-% \PwTmca{} does not validate access elimination: store-forwarding and
-% dead-write-removal are unsound.  We expect that these can be validated by
-% allowing events with different labels to merge in some instances.
+\PwTmca{} does not validate access elimination: store-forwarding and
+dead-write-removal are unsound.  We expect that these can be validated by
+allowing events with different labels to merge. % in some instances.
 
-% \PwTmca{1} is a simpler model than \PwTmca{2}, but requires fences on
-% acquiring reads for \armeight{}.  It would be illuminating to find out what
-% the performance penalty is for these fences.
+\PwTmca{1} is a simpler model than \PwTmca{2}, but requires fences on
+acquiring reads for \armeight{}.  It would be illuminating to find out what
+the performance penalty is for these fences.
 

### 2021-07-16 paper/oopsla.tex
cleanup discussion of PS

@@
 example \ref{oota4} from \cite{DBLP:journals/pacmpl/JagadeesanJR20}:
-\begin{align}
+\begin{gather*}
   \taglabel{oota4}
@@
     \PAR
-    \PR{y}{r} \SEMI \PR{b}{s} \SEMI\IF{s}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI
+    \PR{y}{r} \SEMI\IF{b}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI
     \PAR
     \PW{b}{1}
+    % \\[-1ex]
+    % \hbox{\begin{tikzinline}[node distance=1.5em]
+    %     \event{rx}{\DR{x}{1}}{}
+    %     \event{wy}{\DW{y}{1}}{right=of rx}
+    %     \po{rx}{wy}
+    %     \event{ry}{\DR{y}{1}}{right=3em of wy} 
+    %     \event{wx}{\DW{x}{1}}{right=of ry}
+    %     \event{wz}{\DW{z}{1}}{right=of wx}
+    %     \event{rb}{\DR{b}{1}}{right=of wz}
+    %     \event{wb1}{\DW{b}{1}}{right=3em of rb}
+    %     \po{ry}{wx}
+    %     \rf{wb1}{rb}
+    %     \rf{wy}{ry}
+    %     \rf[out=-170,in=-10]{wx}{rx}
+    %     \po{rb}{wz}
+    %     \po[out=15,in=165]{ry}{wz}
+    %   \end{tikzinline}}
     \\[-1ex]
@@
         \po{rx}{wy}
-        \event{ry}{\DR{y}{1}}{right=3em of wy} 
+        \event{rb}{\DR{b}{1}}{right=3em of wy}
+        \event{ry}{\DR{y}{1}}{right=of rb} 
         \event{wx}{\DW{x}{1}}{right=of ry}
         \event{wz}{\DW{z}{1}}{right=of wx}
-        \event{rb}{\DR{b}{1}}{right=of wz}
-        \event{wb1}{\DW{b}{1}}{right=3em of rb}
+        \event{wb1}{\DW{b}{1}}{right=3em of wz}
         \po{ry}{wx}
-        \rf{wb1}{rb}
-        \rf{wy}{ry}
+        \rf[out=-170,in=-10]{wb1}{rb}
+        \rf[out=15,in=165]{wy}{ry}
         \rf[out=-170,in=-10]{wx}{rx}
-        \po{rb}{wz}
+        \po[out=15,in=165]{rb}{wz}
         \po[out=15,in=165]{ry}{wz}
@@
   \end{gathered}
-\end{align}
-Under \PS{}, this outcome is allowed by \emph{baiting} with the \texttt{else}
-branch, then \emph{switching} to the \texttt{then} branch, based on a coin
-flip $(\IF{b})$.
+\end{gather*}
+Under all variants of \PwT{}, this outcome is disallowed, due to the cycle
+involving $x$ and $y$.\footnote{All of the reads in \ref{oota4} are
+  cross-thread, so there is no difference between \PwTmca{1} and \PwTmca{2}.
+  For \PwTc, there is a cycle in ${\rrfx}\cup{\lt}$.}  Under \PS{}, this
+outcome is allowed by {baiting} with the \texttt{else} branch, then
+{switching} to the \texttt{then} branch, based on a coin flip (${b}$).
 % Using a combination of read introduction and case analysis,
@@
 
-Under all variants of \PwT{}, this outcome is disallowed, due to a cycle in
-${\rrfx}\cup{\lt}$. (Note that all reads cross threads in the example, so
-there is no difference between \PwTmca{1} and \PwTmca{2}.)
-
-The examples in \cite{promising-ldrf} are more complex.  Here is one, dubbed
-\textsc{ldrf-fail-ps}:
+\citet{promising-ldrf} introduce more complex examples to show that the
+promising semantics fails \ldrfsc{}.\footnote{\citet{promising-ldrf} show
+  that by restricting \RMW{}-store reorderings, one can establish \ldrfsc{}
+  for \PS{}.  We speculate that no such restriction is required for \PwT{}.
+  (We did not treat \RMW{}s in our proof of \ldrfsc{}.)}  Here is one, dubbed
+\labeltext{\textsc{ldrf-fail-ps}}{LDRF-Fail-PS}.
 \begin{gather*}  
@@
   \PAR
-  \IF{\BANG z}\THEN
-    \PW{x}{1}
-  \ELSE
+  \IF{z}\THEN
     \IF{\BANG \PFADD{w}{}{1}}\THEN
@@
     \FI
+  \ELSE
+    \PW{x}{1}
   \FI
     \\
-    \hbox{\begin{tikzinlinesmall}[node distance=2em]
+    \hbox{\begin{tikzinline}[node distance=1.5em]
         \event{a1}{\DR{x}{1}}{}
@@
         \event{a5}{\DW{z}{1}}{right=of a4}
-        \event{b1}{\DR{z}{1}}{right=5em of a5}
+        \event{b1}{\DR{z}{1}}{right=4em of a5}
         \event{b2}{\DR{w}{0}}{right=of b1}
@@
         \event{b5}{\DW{x}{1}}{right=of b4}
-        \rmw{a2}{a3}
         \po[out=15,in=165]{a1}{a3}
@@
         \po[out=15,in=165]{a1}{a5}        
-        \rmw{b2}{b3}
         \po{b4}{b5}
@@
         \rf{a5}{b1}
-        \rf[out=15,in=165]{a4}{b4}
-        \rf[out=-165,in=-15]{b3}{a2}
-        \rf[out=-165,in=-15]{b5}{a1}
-      \end{tikzinlinesmall}}
+        \rf[out=10,in=170]{a4}{b4}
+        \rf[out=-170,in=-10]{b3}{a2}
+        \rf[out=-170,in=-10]{b5}{a1}
+        \rmw{a2}{a3}
+        \rmw{b2}{b3}
+      \end{tikzinline}}
   \end{gathered}
 \end{gather*}
-This execution is allowed by \PS{} and disallowed by \PwTmca{}.
-
-The difference between \PS{} and \PwT{} can be understood in terms of the
-valid program transformations.  \PS{} allows reads to be introduced, with
+Again, all variants of \PwT{} disallow the outcome due to the cycle involving
+$x$ and $y$.  It is allowed by \PS{} by baiting the second thread with
+$\PW{x}{1}$ in the \texttt{else} branch, then switching to the \texttt{then}
+branch.  This shows some some structural resemblance to \ref{oota4}, with $z$
+replacing $b$.
+
+\citeauthor{promising-ldrf} argue that the outcome of \ref{LDRF-Fail-PS} is
+inevitable due to compiler optimizations.  The examples crucially involve the
+following sequence of operations:
+\begin{itemize}
+\item read introduction,
+\item if introduction, branching on the read just introduced.
+\end{itemize}
+We believe this combination of optimizations is unsound.  This is obviously
+the case in \cXI: read introduction may cause undefined behavior (\ub), due to
+the possible introduction of a data race.
+
+The situation is more delicate in \llvm.  The short version of the story is
+that load-hoisting followed by case analysis is unsound in \llvm, without freeze. 
+This happens because:
+\begin{itemize}
+\item read introduction may result in the undefined value $\UNDEFINED$, due
+  to the possible introduction of a data race \cite{DBLP:conf/cgo/ChakrabortyV17}, and
+\item branching on an undefined value in \llvm{} results in \ub. 
+\end{itemize}
+\llvm{} delays \ub{} using the undefined value.  This allows \llvm{} to
+perform optimizations such as load hoisting, where
+\begin{math}
+  \IF{C}\THEN \PR{x}{r}\FI
+\end{math}
+is rewritten to 
+\begin{math}
+  \PR{x}{s}\SEMI
+  \LET{r}{\TERNARY{C}{s}{r}}.
+\end{math}
+Despite this, other optimizations regularly performed by \llvm{} are
+unsound \cite{DBLP:conf/pldi/LeeKSHDMRL17}.  An example is loop switching,
+where
+\begin{math}
+  \WHILE{C_1}\DO \IF{C_2}\THEN \aCmd_1 \ELSE \aCmd_2 \FI \OD
+\end{math}
+is rewritten to 
+\begin{math}
+  \IF{C_2}\THEN \WHILE{C_1}\DO \aCmd_1 \OD \ELSE \WHILE{C_1}\DO \aCmd_2 \FI \OD.
+\end{math}
+Freeze was introduced in \llvm{} in order to make such optimizations sound by
+allowing branch on frozen $\UNDEFINED$ to give nondeterministic choice rather
+than \ub{}.  In the RFC for freeze, \citet{nuno} says: \emph{``Note that
+  having branch on poison not trigger \ub{} has its own problems.  We believe
+  this is a good tradeoff.''}  \ref{LDRF-Fail-PS} demonstrates a concrete
+problem with this tradeoff.  Other compilers, such as Compcert, are more
+conservative \cite[\textsection9]{DBLP:conf/pldi/LeeKSHDMRL17}.
+
+Thus, the difference between \PS{} and \PwT{} can be understood in terms of
+the valid program transformations.  \PS{} allows reads to be introduced, with
 subsequent case analysis on the value read.  \PwT{} validates case analysis,
-but invalidates read introduction.  In return, \PwT{} enjoys compositionality
-for temporal safety properties, as discussed in \textsection\ref{sec:results}.
-
-\citet{promising-ldrf} show that by restricting \RMW{}-store reorderings, one
-can establish local \drf{} for \PS{}.  We believe that no such restriction is
-required for \PwT{}.
-
-These differences highlight the subtle tensions between compiler optimizations
-and program logics that are revealed by relaxed memory models.  It is not
-possible to have everything one wants. Thus, one is forced to choose which
-optimizations and reasoning principles are most important.
+but invalidates read introduction.
+
+Allowing executions such as \ref{oota4} and \ref{LDRF-Fail-PS} also
+invalidates compositional reasoning for temporal safety properties (see
+\textsection\ref{sec:results}).
+
+These differences highlight the subtle tensions between compiler
+optimizations and program logics that are revealed by relaxed memory models.
+It is not possible to have everything one wants. Thus, one is forced to
+choose which optimizations and reasoning principles are most
+important.\footnote{Another example is the tension between load
+  hoisting---forbidden in \cXI{} but allowed by \llvm{}---and common
+  subexpression elimination over an acquiring lock---allowed by \cXI{} but
+  forbidden by \llvm{} \cite{DBLP:conf/cgo/ChakrabortyV17}.}
+
+Finally, we note that it is possible that \PS{} is properly weaker than \PwT{}.
 
@@
 % (2) inline two threads on both sides of the introduced conditional,
-% (3) choose different orders for the two threads for the two sides of the conditional.
+% (3) choose different orders for the two threads for the two sides of the ( conditional.
 
@@
 % \PS{} admits CohCYC, but \weakestmo{} does not.
+
+% \subsubsection{Load hoisting in LLVM}
+% Load-hoisting followed by case analysis is unsound in LLVM, without freeze.
+% Introducing a read may cause a race, resulting in read value $\UNDEFINED$.
+% Branch on $\UNDEFINED$ is UB.  Freeze was added to get around this...
+
+% Examples from \cite{promising-ldrf} show that freeze is bullshit.  Compcert
+% does not validate loop switching
+% \cite[\textsection9]{DBLP:conf/pldi/LeeKSHDMRL17}.
+
+% \cite[\textsection3.3]{DBLP:conf/pldi/LeeKSHDMRL17} Global Value Numbering
+% (GVN) and Loop Unswitching require different semantics for branch on
+% $\UNDEFINED$. \url{https://llvm.org/docs/LangRef.html#freeze-instruction}
+
+% \href{https://lists.llvm.org/pipermail/llvm-dev/2016-October/106182.html}{Purpose of Freeze}
+% \begin{quotation}
+%   Poison is propagated aggressively throughout. However, there are cases
+%   where this breaks certain optimizations, and therefore freeze is used to
+%   selectively stop poison from being propagated.
+
+%   A use of freeze is to enable speculative execution.  For example, loop
+%   switching does the following transformation:
+% \begin{verbatim}
+% \end{verbatim}
+% \begin{verbatim}
+% while (C) {        if (C2) {     
+%   if (C2) {           while (C)  
+%    A                     A       
+%   } else {   ==>   } else {      
+%    B                   while (C) 
+%   }                       B      
+% }                  }             
+% \end{verbatim}
+%   Here we are evaluating C2 before C.  If the original loop never executed
+%   then we had never evaluated C2, while now we do.  So we need to make sure
+%   there's no UB for branching on C2.  Freeze ensures that so we would
+%   actually have 'if (freeze(C2))' instead.  Note that
+%   \emph{having branch on poison not trigger UB has its own problems.}
+%   We believe this is a good tradeoff.
+% \end{quotation}
+
+% \begin{quotation}
+%   LLVM frequently performs such load introductions in the “simplify CFG”
+%   pass; e.g., when hoisting loads outside of conditionals.
+% \end{quotation}
+
+% case analysis happens in ``function specialization'' also known as
+% ``procedure cloning''
+
+% \subsubsection{Hoisting and CSE}
+% Example from Viktor's talk: ``Weak Memory Concurrency in C/C++11 and LLVM''
+
+% Load Hoisting:
+% \begin{displaymath}
+%   \IF{c}\THEN \PR{x}{a}\FI
+%   \rightsquigarrow
+%   \PR{x}{t}\SEMI
+%   \LET{a}{\TERNARY{c}{t}{a}}
+% \end{displaymath}
+
+% CSE over acquiring lock:
+% \begin{displaymath}
+%   \PR{x}{a}\SEMI
+%   \LOCK\SEMI
+%   \PR{x}{b}
+%   \rightsquigarrow
+%   \PR{x}{a}\SEMI
+%   \LOCK\SEMI
+%   \LET{b}{a}
+% \end{displaymath}
+
+% Having both is clearly wrong:
+% \begin{displaymath}
+%   \begin{array}{l}
+%     \IF{c}\THEN\\
+%     \quad\PR{x}{a}\FI\SEMI\\
+%     \LOCK\SEMI\\
+%     \PR{x}{b}    
+%   \end{array}
+%   \rightsquigarrow
+%   \begin{array}{l}
+%     \PR{x}{t}\SEMI\\
+%     \LET{a}{\TERNARY{c}{t}{a}}\SEMI\\
+%     \LOCK\SEMI\\
+%     \PR{x}{b}    
+%   \end{array}
+%   \rightsquigarrow
+%   \begin{array}{l}
+%     \PR{x}{t}\SEMI\\
+%     \LET{a}{\TERNARY{c}{t}{a}}\SEMI\\
+%     \LOCK\SEMI\\
+%     \LET{b}{a}    
+%   \end{array}
+% \end{displaymath}
+% When c is false, x is moved out of the critical region!
+
+% So we have to forbid one transformation.
+% \begin{itemize}
+% \item C11 forbids load hoisting, allows CSE over lock().
+% \item LLVM allows load hoisting, forbids CSE over lock().
+% \end{itemize}
 

### 2021-08-04 paper/drf.tex
change to strict partial orders.

@@
   \item \PwP{} allows preconditions that are stronger than the weakest precondition.
-  \item \PwP{} imposes \ref{pom-rf-le} ($\rrfx$ implies $\le$) and thus is
+  \item \PwP{} imposes \ref{pom-rf-le} ($\rrfx$ implies $\lt$) and thus is
     similar to \PwTmca{1}.  \PwTmca{2} is a weaker model that is new to this
@@
 Data races are defined using \emph{program} order $(\rpox)$, not
-\emph{pomset} order $(\le)$. %, and thus is stable with respect to augmentation.
+\emph{pomset} order $(\lt)$. %, and thus is stable with respect to augmentation.
 
@@
   pomset order in generation-minimal top-level pomsets.  First, we note that
-  $\le_*$ (the transitive reduction $\le$) can be decomposed into three
-  disjoint relations.  Let ${\rppo}=({\le_*}\cap{\rpox})$ denote
+  $\lt_*$ (the transitive reduction $\lt$) can be decomposed into three
+  disjoint relations.  Let ${\rppo}=({\lt_*}\cap{\rpox})$ denote
   \emph{preserved} program order, as required by sequential composition and
   conditional.  The other two relations are cross-thread subsets of
-  $({\le_*}\setminus{\rpox})$: $\rrfe$ (reads-from-external) orders writes
+  $({\lt_*}\setminus{\rpox})$: $\rrfe$ (reads-from-external) orders writes
   before reads, satisfying \ref{par-rf-le}; $\rcae$

### 2021-08-04 paper/mca1.tex
cleanup discussion of PS
change to strict partial orders.

@@
       $\aEv\le\cEv$,
-    \item \label{pom-rf-le} if $\bEv\xrfx\aEv$ then $\bEv\le\aEv$.
+    \item \label{pom-rf-le} if $\bEv\xrfx\aEv$ then $\bEv\lt\aEv$.
     \end{enumerate}
@@
 \begin{math}
-  \sem{\aCmd_1 \LPAR[\bThrd] \aCmd_2} = \sLPAR{\sem{\aCmd_1}}{\sem{\aCmd_2}}.
+  \sem{\aCmd_1 \RPAR \aCmd_2} = \sPAR{\sem{\aCmd_1}}{\sem{\aCmd_2}}.
 \end{math}
@@
   In \PwTmca{2}, it is possible for $\rrf$ to contradict
-  $\le$.  In this case, we use a dotted arrow for $\rrf$: $\bEv\xrfint\aEv$
-  indicates that $\aEv\le\bEv$.}
+  $\lt$.  In this case, we use a dotted arrow for $\rrf$: $\bEv\xrfint\aEv$
+  indicates that $\aEv\lt\bEv$.}
 \begin{gather*}
@@
 \ref{pom-rf-le} subsequently ensures that
-$\DWP[\mREL]{y}{1}\le \DRP[\mACQ]{y}{1}$.  The \emph{antidependency}
+$\DWP[\mREL]{y}{1}\lt \DRP[\mACQ]{y}{1}$.  The \emph{antidependency}
 $\DRP{x}{0}\xwk \DWP{x}{1}$ is required by \ref{pom-rf-block}.  (Alternatively,

### 2021-08-04 paper/mca2.tex
cleanup discussion of PS
change to strict partial orders.

@@
 % \refdef{def:pwt:mca1},
-adding local constraints on $\rrfx$ to $\sLPAR{}{}$, $\sSEMI{}{}$ and
+adding local constraints on $\rrfx$ to $\sPAR{}{}$, $\sSEMI{}{}$ and
 $\sIF{}{}{}$.
@@
 A \PwTmca{2} need not satisfy requirement \ref{pom-rf-le}, and thus we may
-have $\bEv\xrfx\aEv$ and $\aEv\le\bEv$.
+have $\bEv\xrfx\aEv$ and $\aEv\lt\bEv$.
 
@@
       \end{tikzinlinesmall}}}
-\end{math}, taking $\rrfx$ from the left and $\le$ from the right.
+\end{math}, taking $\rrfx$ from the left and $\lt$ from the right.
 
@@
   \item external fulfillment:
-    $\bEv\le\aEv$ and if $\labeling(\cEv) \rblocks\labeling(\aEv)$ then either $\cEv\le\bEv$ or $\aEv\le\cEv$, or
+    $\bEv\lt\aEv$ and if $\labeling(\cEv) \rblocks\labeling(\aEv)$ then either $\cEv\le\bEv$ or $\aEv\le\cEv$, or
   \item internal fulfillment:

### 2021-08-04 paper/oopsla.tex
change to strict partial orders.

@@
 \jjr{} imposes \emph{causal strengthening}, which requires for every pomset
-$\aPS$, if $\bEv\le\aEv$ then $\labelingForm(\aEv) \rimplies \labelingForm(\bEv)$. 
+$\aPS$, if $\bEv\lt\aEv$ then $\labelingForm(\aEv) \rimplies \labelingForm(\bEv)$. 
 \begin{scope}

### 2021-08-05 paper/intro.tex
Submission
add l-consistency to term

@@
     DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}---these
-  models validate many compiler optimizations, but also
+  models validate many compiler optimizations, but
   %allow some dubious executions 
@@
 Modular Relaxed Dependencies (\MRD), which use event structures to calculate
-dependencies for \cXI.  \citet{DBLP:journals/pacmpl/JagadeesanJR20} defined
+dependencies for \cXI, targeting the Intermediate Memory Model (\IMM)
+\cite{DBLP:journals/pacmpl/PodkopaevLV19}. \citet{DBLP:journals/pacmpl/JagadeesanJR20} defined
 Pomsets with Preconditions (\PwP), which use preconditions and logic to
-calculate dependencies for a Java-like language on multicopy-atomic (\mca{}) hardware, such as
+calculate dependencies for a Java-like language targeting multicopy-atomic (\mca{}) hardware, such as
 \armeight{} \cite{DBLP:journals/pacmpl/PulteFDFSS18}.  However, neither paper
@@
 we extend the semantics to include additional features, such as address
-calculation.
+calculation and \RMW{}s.
 

### 2021-08-09 paper/mca1.tex
adjusting address calculation to handle aliasing and TC12

@@
 
-Lemmas \ref{lem:monoid} and \ref{lem:if} mostly hold for \PwTmca{1}.  The
-exceptions are items \eqref{lem:ifelse:if:if1} and \eqref{lem:ifelse:if:if2},
-which become inclusions.  For example, \eqref{lem:ifelse:if:if1} becomes:
-\begin{displaymath} 
-  \xIFTHEN{\aForm}{\aPSS_1}{\aPSS_2}
-  \supseteq
-  \xSEMI{
-    \xIFTHEN{\aForm}{\aPSS_1}{}
-  }{
-    \xIFTHEN{\lnot\aForm}{\aPSS_2}{}
-  }
-\end{displaymath}
-The culprit is $\rdelay$, which introduces order regardless of whether
-preconditions are disjoint.  As an example, 
-\begin{math}
-  \sem{\IF{r}
-  \THEN \PW{x}{1}
-  \ELSE \PW{x}{2}
-  \FI}
-\end{math}
-has an execution with
-\begin{math}
-  (r{=}0\mid\DW{x}{2})
-  \xwki
-  (r{\neq}0\mid\DW{x}{1}),
-\end{math}
-(using augmentation), whereas
-\begin{math}
-  \sem{
-    \IF{r} \THEN \PW{x}{1}\FI
-    \SEMI
-    \IF{\BANG r} \THEN \PW{x}{2}\FI
-  \FI}
-\end{math}
-has no such execution.
-
-For further discussion, see \textsection\ref{sec:delay}.
+Lemmas \ref{lem:monoid} and \ref{lem:if} hold for \PwTmca{1}.  For further
+discussion of items \eqref{lem:ifelse:if:if1} and \eqref{lem:ifelse:if:if2},
+see \textsection\ref{sec:delay}.
 

### 2021-09-06 paper/POPL22_rebuttal_response.md
Started working on POPL22_rebuttal

We are thankful to reviewers for their work.

# Reviewer A. Why is it challenging to give a denotational semantics to sequential composition?
TODO

# Reviewers A and C. Correctness of the model, mechanization, relation to other models
We agree that the model is complicated and requires formal results to be sure in its correctness,
that is why we have been working on its mechanization.
Currently, we have mechanized in Coq Lemma 4.5 from the paper
(associativity of sequential composition and SKIP being left and right ids for sequential composition).

We also agree that a new memory model needs to be positioned against existing models.
The usual result here is a compilation correctness to hardware memory models.
We address this by showing compilation result for Armv8 model (§5)
(it could be easily extended to the x86-TSO model since x86-TSO has a stronger model than Armv8).
Developing a version of the semantics which could cover compilation to non-MCA models like IBM Power
is our immediate future work, which we decided not to include in this submission since it would require to put even
more to this (quite dense in our opinion) paper.

Direct comparing of behaviors allowed by other memory models for programming languages solving the OOTA problem
(Promising [Kang-al:POPL17], Weakestmo [Chakraborty-Vafeiadis:POPL19], MRD [Paviotti-al:ESOP20], PwP [Jagadeesan-al:OOPSLA20])
is not as interesting since all of them are incomparable, i.e., there are examples which are allowed by one/disallowed by another
and vice versa.
TODO: say something good about properties of our model comparing to these models.

### 2021-09-06 paper/POPL22_rebuttal_response.md
an answer to Reviewer B

@@
 TODO: say something good about properties of our model comparing to these models.
+
+# Reviewer B. Promising-ARM/RISC-V [Pulte et al. 2019] is evaluated on a concurrent queue implemented in 215 lines of C++ code. Does PwT-C11 work at this scale?
+Promising-ARM/RISC-V is a hardware memory model, that is, it is supposed to allow less behaviors comparing to programming language-level models like PwT-C11.
+TODO: more on testing in our model.
+
+# Reviewer B. Syntactic vs. semantic dependencies and their usage in compilers
+Yes, tracking semantic dependencies is intrinsically harder than syntactic ones, and hardware memory models stick with syntax.
+However, we cannot settle with syntactic dependencies for models of programming languages since common compiler optimizations
+used in almost all real compilers (including GCC and Clang) may remove syntactic dependencies (unlike semantic ones).
+That is, a programming language memory model which supports compiler optimizations and disallows OOTA (unlike the C/C++ memory model)
+has to track semantic dependencies in one way or another. All other proposed solutions to this problem do that
+(Promising [Kang-al:POPL17], Weakestmo [Chakraborty-Vafeiadis:POPL19], MRD [Paviotti-al:ESOP20], PwP [Jagadeesan-al:OOPSLA20]).
+The point of the submission is to make the idea of tracking dependencies with logic, proposed by Jagadeesan et. al [2020],
+to be compositional on the SEQ operator.
+TODO: more here

### 2021-09-06 paper/POPL22_rebuttal_response.md
minor in answer to Reviewer B

@@
 to be compositional on the SEQ operator.
+
+Also, a compiler does not have to use the proposed semantics directly, i.e., for calculating dependencies.
+Instead, the semantics has to be a wrapper over a reasonable set of compiler optimizations. That is, the optimizations
+could make more conservative assumptions about dependencies than the semantics. This is what we are trying to achieve.
+
 TODO: more here

### 2021-09-06 paper/POPL22_rebuttal_response.md
Mark's comments on reviewer A question

--- 2021-09-06 paper/POPL22_rebuttal_response.md
@@
 TODO
+[MB -- This question is mapping to the questions of motivation from Reviewers A and C. We should reiterate the broader motivation here:
+1. thin air is a problem with the semantics of all optimised concurrent languages
+2. there are only a handful of solutions out there and many of those are flawed
+3. a solution framed as an SDEP relation is the only approach compatible with the existing C++ concurrency definition
+4. sequential composition is necessary (but not sufficient) for scalable reasoning about big chunks of code, and no one else has that.
+]
 

### 2021-09-06 paper/POPL22_rebuttal_response.md
More comments from Mark
more minor changes to rebuttal

@@
 
-# Reviewer A. Why is it challenging to give a denotational semantics to sequential composition?
+# Reviewer A and C. Why is a denotational semantics for sequential composition challenging and worthwhile?
 TODO
@@
 (associativity of sequential composition and SKIP being left and right ids for sequential composition).
+Associativity supports the automatic evaluation tool -- PwTer chooses one order for evaluation and yet is complete by associativity.
 
@@
 to be compositional on the SEQ operator.
+[MB -- the last sentence casts the motivation of the paper in a minimising way.]
 

### 2021-09-06 paper/POPL22_rebuttal_response.md
Simon's comments for rebuttal

@@
 We are thankful to reviewers for their work.
+
+Since submission we have formalised this work in Coq.
+We have mechanically proved in Coq that Lemma 4.5 from the paper holds, which validates the compositionality of this model as well as being a necessary property the completeness of the tool.
+We have further proved in Coq that `skip` is the left and right unit for sequential composition, showing that the (SEQ, Skip) structure is a monoid over PwTs -- as expected.
+[SC: It's probably worth someone more qualified checking this last statement!]
 
@@
 ]
+
+To determine whether a dependency is present in some fragment of code, MRD requires that all following code is evaluated.
+With PwT, the presence or absence of a dependency can be understood in isolation.
+In practice, this means future applications where PwT can be used to modularly validate assumptions about program dependencies in larger blocks of code incrementally -- rather than the approach of MRD where evaluation must be done totally.
+
+Further, PwT-C11 is only the second semantics which interoperates with C++ through a semantic dependency relation, and the first one to be fully compositional.
+Semantic dependency is a worthwhile goal: a restriction of acyclic(sdep \cup rf) is a statement which is compatible with the existing C++ standard, subject to a good definition of sdep.
+With the exception of MRD, other thin-air free programming language memory models do not distil dependencies down to a relation compatible with the existing C++ standard.
 
@@
 Promising-ARM/RISC-V is a hardware memory model, that is, it is supposed to allow less behaviors comparing to programming language-level models like PwT-C11.
+It is worth commenting that the Promising arc of papers is now quite deep (by our count 1 POPL paper and 3 PLDI papers), expecting the same depth of results in a single paper is unreasonable -- even for POPL.
+[SC: this is likely too combative and should be dropped... ]
 TODO: more on testing in our model.

### 2021-09-06 paper/POPL22_rebuttal_response.md
grammar

@@
 With PwT, the presence or absence of a dependency can be understood in isolation.
-In practice, this means future applications where PwT can be used to modularly validate assumptions about program dependencies in larger blocks of code incrementally -- rather than the approach of MRD where evaluation must be done totally.
+In practice, this enables future applications where PwT can be used to modularly validate assumptions about program dependencies in larger blocks of code incrementally -- rather than the approach of MRD where evaluation must be done totally.
 
 Further, PwT-C11 is only the second semantics which interoperates with C++ through a semantic dependency relation, and the first one to be fully compositional.
-Semantic dependency is a worthwhile goal: a restriction of acyclic(sdep \cup rf) is a statement which is compatible with the existing C++ standard, subject to a good definition of sdep.
+Semantic dependency is a worthwhile goal: a restriction of $acyclic(sdep \cup rf)$ is a statement which is compatible with the existing C++ standard, subject to a good definition of sdep.
 With the exception of MRD, other thin-air free programming language memory models do not distil dependencies down to a relation compatible with the existing C++ standard.

### 2021-09-06 paper/POPL22_rebuttal_response.md
Added some text to the rebuttal

--- 2021-09-06 paper/POPL22_rebuttal_response.md
@@
 ]
+
+Both referees asked about the motivation for a compositional model of
+sequential composition, why not consider S1 and S2 equal if, for every
+continuation S0, we have [[ S1 ; S0 ]] = [[ S2 ; S0 ]]?  The problem
+is that this requires a quantification over all continuations S0. This
+quantification is problematic, both from a theoretical point of view
+(the syntax of programs is now mentioned in the definition of the
+semantics) and in practice (tools cannot quantify over infinite
+sets. This is a related problem to contextual equivalence, full
+abstraction and the CIU theorem.
+
+In addition, referee C asks "My worry is: is this a big enough "win"
+to be worth all this effort?" We would argue yes, that having a model
+behind peephole compiler optimizations is worth it, and that this
+requires a compositional treatment of sequential composition.
 

### 2021-09-06 paper/POPL22_rebuttal_response.md
Updates to response

--- 2021-09-06 paper/POPL22_rebuttal_response.md
@@
 We are thankful to reviewers for their work.
+In this response, we address the referees main concerns.
 
-# Reviewer A and C. Why is a denotational semantics for sequential composition challenging and worthwhile?
-TODO
-[MB -- This question is mapping to the questions of motivation from Reviewers A and C. We should reiterate the broader motivation here:
-1. thin air is a problem with the semantics of all optimised concurrent languages
-2. there are only a handful of solutions out there and many of those are flawed
-3. a solution framed as an SDEP relation is the only approach compatible with the existing C++ concurrency definition
-4. sequential composition is necessary (but not sufficient) for scalable reasoning about big chunks of code, and no one else has that.
-]
+First, we reiterate the broader context of our work: Thin air is a problem
+with the semantics of all optimized concurrent languages.  There are only a
+handful of solutions and all have deficiencies.  We focus on denotations with
+semantic dependency (SDEP) and sequential composition for the following
+reasons:
+- Using an SDEP relation is the only approach compatible with the existing
+  C++ concurrency definition.
+- Sequential composition is necessary for scalable reasoning about big chunks
+  of code.  No existing model captures it.
+- Existing models do not get at the essence of the problem.  By analogy: one
+  can use Landin's SECD machine to understand iteration, but Hoare logic and
+  Scott–Strachey semantics provide much more insight.
+
+## Reviewers A and C. Correctness of the model, mechanization, relation to other models.
+
+Since submission we have formalized this work in Coq.  This development is
+available for referees at https://fpl.cs.depaul.edu/jriely/PwT.zip
+(approximately 11,000 lines of code).
+
+- We have formally verified lemma 4.5, which states that SEQ is associative
+  and that SKIP is left and right unit.
+
+  This is a key result, showing that SEQ has the most important property
+  expected of sequential composition.  This result also contributes to
+  confidence in PwTer (§8), which runs litmus tests: PwTer uses a
+  normal form for programs, where sequential composition always associates to
+  the left.
+
+- We have established a compilation result for programs in a normal form.
+
+  This shows that our SDEP relation is not too strong to allow efficient
+  compilation for C11.  Thus, it alleviates the compilation problem of RC11,
+  as expected.
+
+  It remains to show that all programs are equivalent to some program in this
+  normal form.  Lemma 4.5 gets us halfway there.  We are currently working on
+  the reverse direction of Lemma 4.6(e), using the if-closure semantics
+  (§9.5).  Along with a bit of bookkeeping, this will complete the proof.
+
+We agree that a new memory model needs to be positioned against existing
+models.  The usual result here is a compilation correctness to hardware
+memory models.  For PwR-MCA, we address this by showing compilation result
+for Armv8 model (§5).  (This could be easily extended to the x86-TSO model
+since x86-TSO has a stronger model than Armv8.)  Developing a non-MCA version
+of this semantics which could cover compilation to PTX and IBM Power is our
+immediate future work, which we decided not to include in this submission
+since it would require to put even more to this (quite dense!) paper.
+
+Comparing software models, however, is unsatisfying: they are all
+incomparable, i.e., there are examples which are allowed by one/disallowed by
+the other and vice versa.
+
+Morally, our model sits between the strong models (exemplified by RC11
+[Lahav-al:PLDI17]) and the speculative models (exemplified by the promising
+semantics [Kang-al:POPL17]).  As we argue in §1:
+
+- The strong models require too much synchronization.
+
+- The speculative models allow thin air behaviors.
+
+Looking at the details, however, PwT-MCA is incomparable to both RC11 and
+promising semantics.  RC11 allows non-MCA behaviors that PwT-MCA disallows.
+PwT-MCA has a weaker notion of coherence than the promising semantics.
+
+Some differences reflect an attempt to fix a bug.  For example, Weakestmo
+[Chakraborty-Vafeiadis:POPL19] purposefully disallows _thin-air-like_
+behaviors of the promising semantics.
+
+Other differences reflect a different balance between allowed optimizations
+and reasoning principles.  There are fundamental conflicts, for example:
+
+- between Common Subexpression Elimination (CSE) and read-read coherence
+  (§D.1)
+
+- between if-introduction (aka, case analysis, if-closure) and java-style
+  final-field semantics.  (If-introduction requires that address dependencies
+  and control dependencies are the same.  Final-fields require that they be
+  different.)
+
+## Reviewers A and C. Why is a denotational semantics for sequential composition challenging and worthwhile?
 
@@
 
-# Reviewers A and C. Correctness of the model, mechanization, relation to other models
-We agree that the model is complicated and requires formal results to be sure in its correctness,
-that is why we have been working on its mechanization.
-Currently, we have mechanized in Coq Lemma 4.5 from the paper
-(associativity of sequential composition and SKIP being left and right ids for sequential composition).
-Associativity supports the automatic evaluation tool -- PwTer chooses one order for evaluation and yet is complete by associativity.
+To determine whether a dependency is present in some fragment of code, MRD
+and PwP require that all of following code is evaluated.  Suppose that you
+are writing system call code and you wish to know if you can reorder a couple
+of statements.  Using MRD or PwP, you cannot tell whether this is possible
+without having the calling code.  This is what we mean when we say that MRD
+and PwP require perfect knowledge of the future.
 
-We also agree that a new memory model needs to be positioned against existing models.
-The usual result here is a compilation correctness to hardware memory models.
-We address this by showing compilation result for Armv8 model (§5)
-(it could be easily extended to the x86-TSO model since x86-TSO has a stronger model than Armv8).
-Developing a version of the semantics which could cover compilation to non-MCA models like IBM Power
-is our immediate future work, which we decided not to include in this submission since it would require to put even
-more to this (quite dense in our opinion) paper.
+With PwT, the presence or absence of a dependency can be understood in isolation.
+In practice, this enables future applications where PwT can be used to modularly validate assumptions about program dependencies in larger blocks of code incrementally -- rather than the approach of MRD/PwP where evaluation must be done totally.
 
-Direct comparing of behaviors allowed by other memory models for programming languages solving the OOTA problem
-(Promising [Kang-al:POPL17], Weakestmo [Chakraborty-Vafeiadis:POPL19], MRD [Paviotti-al:ESOP20], PwP [Jagadeesan-al:OOPSLA20])
-is not as interesting since all of them are incomparable, i.e., there are examples which are allowed by one/disallowed by another
-and vice versa.
-TODO: say something good about properties of our model comparing to these models.
+Further, PwT-C11 is only the second semantics which interoperates with C++ through a semantic dependency relation, and the first one to be fully compositional.
+Semantic dependency is a worthwhile goal: a restriction of $acyclic(sdep \cup rf)$ is a statement which is compatible with the existing C++ standard, subject to a good definition of sdep.
+With the exception of MRD, other thin-air free programming language memory models do not distil dependencies down to a relation compatible with the existing C++ standard.
 
-# Reviewer B. Promising-ARM/RISC-V [Pulte et al. 2019] is evaluated on a concurrent queue implemented in 215 lines of C++ code. Does PwT-C11 work at this scale?
-Promising-ARM/RISC-V is a hardware memory model, that is, it is supposed to allow less behaviors comparing to programming language-level models like PwT-C11.
-TODO: more on testing in our model.
+## Reviewer B. Syntactic vs. semantic dependencies and their usage in compilers
 
-# Reviewer B. Syntactic vs. semantic dependencies and their usage in compilers
 Yes, tracking semantic dependencies is intrinsically harder than syntactic ones, and hardware memory models stick with syntax.
@@
 (Promising [Kang-al:POPL17], Weakestmo [Chakraborty-Vafeiadis:POPL19], MRD [Paviotti-al:ESOP20], PwP [Jagadeesan-al:OOPSLA20]).
-The point of the submission is to make the idea of tracking dependencies with logic, proposed by Jagadeesan et. al [2020],
-to be compositional on the SEQ operator.
-[MB -- the last sentence casts the motivation of the paper in a minimising way.]
 
-Also, a compiler does not have to use the proposed semantics directly, i.e., for calculating dependencies.
-Instead, the semantics has to be a wrapper over a reasonable set of compiler optimizations. That is, the optimizations
-could make more conservative assumptions about dependencies than the semantics. This is what we are trying to achieve.
+We have attempted to capture semantic dependencies in a meaningful way that
+admits sequential composition.  PwP [Jagadeesan-al:OOPSLA20] only got halfway
+there: No one would have ever been happy with Hoare logic if it failed to
+capture the meaning of sequential composition.
 
-TODO: more here
+Note also that a compiler does not have to use the proposed semantics
+directly, i.e., for calculating dependencies.  Instead, the semantics is
+meant to be a wrapper that validate some reasonable set of compiler
+optimizations. A compiler may make more conservative assumptions about
+dependencies than the semantics. This is explicitly allowed by Lemma 4.8
+(augment closure).
+
+## Reviewer B. Does the model scale?
+
+> Promising-ARM/RISC-V [Pulte et al. 2019] is evaluated on a concurrent queue
+> implemented in 215 lines of C++ code. Does PwT-C11 work at this scale?
+
+We have not attempted to run PwTer on programs as large as the Michael-Scott
+queue.  In the theory itself, there is no issue, but as [Pulte et al. 2019]
+point out, exploring such a large combinatorial state space is challenging.
+The task is simplified in a hardware model, such as Promising-ARM/RISC-V.  In
+a software model such as PwT-C11 or PwT-MCA, the state space is larger, and
+therefore the task is that much more difficult.
+
+## Reviewer A and B: Presentation
+
+We have tried to make §§1-4 readable, as these communicate the basic idea.
+It is true that §§5-9 are mostly definitions and results, but this seems
+inevitable given the space constraints.

### 2021-09-07 paper/POPL22_rebuttal_response.md
typo

@@
 models.  The usual result here is a compilation correctness to hardware
-memory models.  For PwR-MCA, we address this by showing compilation result
+memory models.  For PwT-MCA, we address this by showing compilation result
 for Armv8 model (§5).  (This could be easily extended to the x86-TSO model
@@
 
-Further, PwT-C11 is only the second semantics which interoperates with C++ through a semantic dependency relation, and the first one to be fully compositional.
-Semantic dependency is a worthwhile goal: a restriction of $acyclic(sdep \cup rf)$ is a statement which is compatible with the existing C++ standard, subject to a good definition of sdep.
+Further, PwT-C11 is only the second semantics to interoperate with C++ through a semantic dependency relation, and the first one to be fully compositional.
+Semantic dependency is a worthwhile goal: a restriction of $acyclic(SDEP \cup RF)$ is a statement which is compatible with the existing C++ standard, subject to a good definition of SDEP.
 With the exception of MRD, other thin-air free programming language memory models do not distil dependencies down to a relation compatible with the existing C++ standard.
@@
 directly, i.e., for calculating dependencies.  Instead, the semantics is
-meant to be a wrapper that validate some reasonable set of compiler
+meant to be a wrapper that validates some reasonable set of compiler
 optimizations. A compiler may make more conservative assumptions about

### 2021-09-07 paper/POPL22_rebuttal_response.md
Rebuttal, minor typo
Highlighted Coq

@@
 
-First, we reiterate the broader context of our work: Thin air is a problem
+Before that, we want to address one of the main concerns of referees A and C:
+
+* There is now a formalization of this work in Coq
+
+Details below.
+
+## Overall comments
+
+We reiterate the broader context of our work: Thin air is a problem
 with the semantics of all optimized concurrent languages.  There are only a
@@
 
-## Reviewer A and B: Presentation
+## Reviewers A and B: Presentation
 

### 2021-09-07 paper/POPL22_rebuttal_response.md
reordering response to optimize 500 word limit

@@
 We are thankful to reviewers for their work.
-In this response, we address the referees main concerns.
 
-Before that, we want to address one of the main concerns of referees A and C:
+First, let us address one of the main concerns of referees A and C:
 
-* There is now a formalization of this work in Coq
+## Coq formalization
 
-Details below.
+  Since submission we have formalized this work in Coq.  This development is
+  available for referees at https://fpl.cs.depaul.edu/jriely/PwT.zip
+  (approximately 11,000 lines of code).  In particular, we have formally
+  verified lemma 4.5, which states that SEQ is associative (SeqAssoc.v) and
+  that SKIP is left and right unit (SeqSkipId.v).
 
@@
 
-## Reviewers A and C. Correctness of the model, mechanization, relation to other models.
+## Reviewers A and C. Why is a denotational semantics for sequential composition challenging and worthwhile?
 
-Since submission we have formalized this work in Coq.  This development is
-available for referees at https://fpl.cs.depaul.edu/jriely/PwT.zip
-(approximately 11,000 lines of code).
+Both referees asked about the motivation for a compositional model of
+sequential composition, why not consider S1 and S2 equal if, for every
+continuation S0, we have [[ S1 ; S0 ]] = [[ S2 ; S0 ]]?  The problem
+is that this requires a quantification over all continuations S0. This
+quantification is problematic, both from a theoretical point of view
+(the syntax of programs is now mentioned in the definition of the
+semantics) and in practice (tools cannot quantify over infinite
+sets. This is a related problem to contextual equivalence, full
+abstraction and the CIU theorem.
 
-- We have formally verified lemma 4.5, which states that SEQ is associative
-  and that SKIP is left and right unit.
+In addition, referee C asks: _My worry is: is this a big enough "win"
+to be worth all this effort?_ We would argue yes, that having a model
+behind peephole compiler optimizations is worth it, and that this
+requires a compositional treatment of sequential composition.
 
-  This is a key result, showing that SEQ has the most important property
-  expected of sequential composition.  This result also contributes to
-  confidence in PwTer (§8), which runs litmus tests: PwTer uses a
-  normal form for programs, where sequential composition always associates to
-  the left.
+To determine whether a dependency is present in some fragment of code, MRD
+and PwP require that all of following code is evaluated.  Suppose that you
+are writing system call code and you wish to know if you can reorder a couple
+of statements.  Using MRD or PwP, you cannot tell whether this is possible
+without having the calling code.  This is what we mean when we say that MRD
+and PwP require perfect knowledge of the future.
 
-- We have established a compilation result for programs in a normal form.
+We have attempted to capture semantic dependencies in a meaningful way that
+admits sequential composition.  PwP/MRD only got halfway there: No one would
+have ever been happy with Hoare logic if it failed to capture the meaning of
+sequential composition.
 
-  This shows that our SDEP relation is not too strong to allow efficient
-  compilation for C11.  Thus, it alleviates the compilation problem of RC11,
-  as expected.
+With PwT, the presence or absence of a dependency can be understood in isolation.
+In practice, this enables future applications where PwT can be used to modularly validate assumptions about program dependencies in larger blocks of code incrementally -- rather than the approach of MRD/PwP where evaluation must be done totally.
 
-  It remains to show that all programs are equivalent to some program in this
-  normal form.  Lemma 4.5 gets us halfway there.  We are currently working on
-  the reverse direction of Lemma 4.6(e), using the if-closure semantics
-  (§9.5).  Along with a bit of bookkeeping, this will complete the proof.
+Further, PwT-C11 is only the second semantics to interoperate with C++ through a semantic dependency relation, and the first one to be fully compositional.
+Semantic dependency is a worthwhile goal: a restriction of $acyclic(SDEP \cup RF)$ is a statement which is compatible with the existing C++ standard, subject to a good definition of SDEP.
+With the exception of MRD, other thin-air free programming language memory models do not distil dependencies down to a relation compatible with the existing C++ standard.
+
+## Reviewers A and C. Correctness of the model, relating models.
+
+As noted above, the work has been formalized in Coq.
 
@@
 memory models.  For PwT-MCA, we address this by showing compilation result
-for Armv8 model (§5).  (This could be easily extended to the x86-TSO model
-since x86-TSO has a stronger model than Armv8.)  Developing a non-MCA version
-of this semantics which could cover compilation to PTX and IBM Power is our
-immediate future work, which we decided not to include in this submission
-since it would require to put even more to this (quite dense!) paper.
+for Armv8 model (§5).
 
@@
 
-## Reviewers A and C. Why is a denotational semantics for sequential composition challenging and worthwhile?
-
-Both referees asked about the motivation for a compositional model of
-sequential composition, why not consider S1 and S2 equal if, for every
-continuation S0, we have [[ S1 ; S0 ]] = [[ S2 ; S0 ]]?  The problem
-is that this requires a quantification over all continuations S0. This
-quantification is problematic, both from a theoretical point of view
-(the syntax of programs is now mentioned in the definition of the
-semantics) and in practice (tools cannot quantify over infinite
-sets. This is a related problem to contextual equivalence, full
-abstraction and the CIU theorem.
-
-In addition, referee C asks "My worry is: is this a big enough "win"
-to be worth all this effort?" We would argue yes, that having a model
-behind peephole compiler optimizations is worth it, and that this
-requires a compositional treatment of sequential composition.
-
-To determine whether a dependency is present in some fragment of code, MRD
-and PwP require that all of following code is evaluated.  Suppose that you
-are writing system call code and you wish to know if you can reorder a couple
-of statements.  Using MRD or PwP, you cannot tell whether this is possible
-without having the calling code.  This is what we mean when we say that MRD
-and PwP require perfect knowledge of the future.
-
-With PwT, the presence or absence of a dependency can be understood in isolation.
-In practice, this enables future applications where PwT can be used to modularly validate assumptions about program dependencies in larger blocks of code incrementally -- rather than the approach of MRD/PwP where evaluation must be done totally.
-
-Further, PwT-C11 is only the second semantics to interoperate with C++ through a semantic dependency relation, and the first one to be fully compositional.
-Semantic dependency is a worthwhile goal: a restriction of $acyclic(SDEP \cup RF)$ is a statement which is compatible with the existing C++ standard, subject to a good definition of SDEP.
-With the exception of MRD, other thin-air free programming language memory models do not distil dependencies down to a relation compatible with the existing C++ standard.
-
 ## Reviewer B. Syntactic vs. semantic dependencies and their usage in compilers
@@
 
-We have attempted to capture semantic dependencies in a meaningful way that
-admits sequential composition.  PwP [Jagadeesan-al:OOPSLA20] only got halfway
-there: No one would have ever been happy with Hoare logic if it failed to
-capture the meaning of sequential composition.
-
-Note also that a compiler does not have to use the proposed semantics
+A compiler does not have to use the proposed semantics
 directly, i.e., for calculating dependencies.  Instead, the semantics is

### 2021-10-05 paper/intro.tex
Removing some todos.

@@
 languages should guarantee sequential abstraction, despite the performance
-penalties.  In this paper, we take the opposite approach.  We have
+penalties (see also \cite{10.1145/3462206}).  In this paper, we take the opposite approach.  We have
 attempted to find the most intellectually tractable model that encompasses

### 2021-10-06 paper/tool.tex
Add the image back for the tool, add a reference to PwT-C11 in Conclusion
Removing some todos.
Add table for tool, and some explainatory text

@@
 
-\PwTer{} allows several modes of evaluation: it can evaluate the rules of Fig.~\ref{fig:seq}, implementing \PwT; it can generate program order according to \textsection\ref{sec:c11}, implementing \PwTpo; and similar to \MRD~\cite{DBLP:conf/esop/PaviottiCPWOB20}, it can construct C11-style pre-executions and filter them according to the rules of \rcXI{} as described in \textsection\ref{sec:c11}.
+\PwTer{} allows several modes of evaluation: it can evaluate the rules of Fig.~\ref{fig:seq}, implementing \PwT; it can generate program order according to \textsection\ref{sec:c11}, implementing \PwTpo; and similar to \MRD~\cite{DBLP:conf/esop/PaviottiCPWOB20}, it can construct C11-style pre-executions and filter them according to the rules of \rcXI{} as described in \textsection\ref{sec:c11}, implementing \PwTc{}.
 Finally, \PwTer{} also allows us to toggle the complete check of~\ref{def:complete}, providing an interface for understanding how fragments of code might compose by exposing preconditions and termination conditions that are not yet tautologies.
-% We show \PwTer{} in action in Fig.~\ref{fig:tool}. 
-\PwTer{} will be made open source upon publication.
-%\PwTer{} is available online at \url{https://github.com/graymalkin/pomsets-with-predicate-transformers}.
+We have run \PwTer{} over the Java Causality Tests~\cite{PughWebsite}, and tabulated the results in Figure~\ref{fig:tool}.
+
+It is noteworthy that the \PwTc{} version of the semantics does correctly identity semantic dependencies in tests which would have thin-air executions, but without a definition of reads from, and an axiom like $\textit{acyclic}(\rrfx \cup \rsdepx)$ it does not forbid the thin-air executions (as in test cases 4, 5 and 10).
+In the \PwTc{} model there is a restriction which forbids cycles in $\rrfx \cup \rsdepx$, so this correctly forbids thin-air executions.
+In jctc16, we see that the restrictions of \PwTc{} are too strong, this is because the coherence restrictions in C++ are stronger than those of Java.
+The execution times give a good indication the poor scaling of the tool with program size: for longer test cases the tool takes exponentially longer to compute, and in some cases simply fails.
+The compositional nature of the semantics makes tool building practical, but it is not enough to make it scalable for large tests.
+Unsurprisingly the execution time is dominated by the calculation of the denotation, with the additional axiomatic filtering step of \PwTc{} being within the margin of error. 
+% \PwTer{} will be made open source upon publication.
+\PwTer{} is available online at \url{https://github.com/graymalkin/pomsets-with-predicate-transformers}.
 
 %% [Simon] I am not a fan of how this looks on the page.
-% \begin{figure}[t]
-% \begin{center}
-%   \includegraphics[width=0.75\textwidth]{tool.png}
-%   \Description{Image shows tool running on a Mac.
-Java Causality Test Case 1 is printed, followed by an invocation of the tool as {\tt ./pomsets.exe --check --complete data/tests/jctc/jctc1.lit}, which shows that the tool passes the assertion listed in the test.}
-%   \caption{\label{fig:tool} Example output of \PwTer, validating TC1~\cite{PughWebsite}.}
-% \end{center}
-% \end{figure}
+\begin{figure}[t]
+\begin{center}
+  \begin{tabularx}{0.9\textwidth}{c||c|c|c|c}
+    \multirow{2}{*}{\bf Test name} & \multicolumn{2}{c|}{\PwT{}} & \multicolumn{2}{c}{\PwTc{}} \\
+    \cline{2-5}
+                                   & Result & Execution Time (s)  & Result & Execution Time (s)  \\
+    \cline{1-5}
+    jctc1                          & pass   & 2.608               & pass   & 2.397 \\
+    jctc2                          & pass   & 25.754              & pass   & 25.780 \\
+    jctc3                          & pass   & 205.120             & pass   & 196.935 \\
+    jctc4                          & fail   & 2.110               & pass   & 2.269 \\
+    jctc5                          & fail   & 69.441              & pass   & 63.714 \\
+    jctc6                          & pass   & 12.489              & pass   & 11.245 \\
+    jctc7                          & pass   & 96.099              & pass   & 88.250 \\
+    jctc8                          & pass   & 2.473               & pass   & 2.482 \\
+    jctc9                          & pass   & 15.384              & pass   & 13.592 \\
+    jctc10                         & fail   & 513.234             & pass   & 494.133 \\
+    jctc11                         & $\bot$ & --                  & $\bot$ & \\
+    jctc12                         & $\bot$ & --                  & $\bot$ & \\
+    jctc13                         & fail   & 2.247               & pass   & 2.101 \\
+    jctc14                         & unsupported   & --           & unsupported   & -- \\
+    jctc15                         & unsupported   & --           & unsupported   & -- \\
+    jctc16                         & pass   & 9.232               & fail   & 8.492 \\
+    jctc17                         & pass   & 186.228             & pass   & 178.304 \\
+    jctc18                         & pass   & 2.247               & pass   & 177.292 \\
+    jctc19                         & unsupported   & --           & unsupported   & -- \\
+    jctc20                         & unsupported   & --           & unsupported   & -- \\
+  \end{tabularx}
+  \caption{\label{fig:tool} Tool results for Java Causality Test Cases~\cite{PughWebsite}.
+$\bot$ indicates the tool failed to run for this test due to a memory overflow.
+Tests run on an Intel i9-9980HK with 64 GB of memory, execution times are the mean of 3 runs.}
+\end{center}
+\end{figure}

### 2021-10-06 paper/tool.tex
minor tool.tex wordsmith

@@
 Finally, \PwTer{} also allows us to toggle the complete check of~\ref{def:complete}, providing an interface for understanding how fragments of code might compose by exposing preconditions and termination conditions that are not yet tautologies.
-We have run \PwTer{} over the Java Causality Tests~\cite{PughWebsite}, and tabulated the results in Figure~\ref{fig:tool}.
+We have run \PwTer{} over the Java Causality Tests~\cite{PughWebsite} supported in the input syntax, and tabulated the results in Figure~\ref{fig:tool}.
 
-It is noteworthy that the \PwTc{} version of the semantics does correctly identity semantic dependencies in tests which would have thin-air executions, but without a definition of reads from, and an axiom like $\textit{acyclic}(\rrfx \cup \rsdepx)$ it does not forbid the thin-air executions (as in test cases 4, 5 and 10).
-In the \PwTc{} model there is a restriction which forbids cycles in $\rrfx \cup \rsdepx$, so this correctly forbids thin-air executions.
+It is noteworthy that the \PwT{} version of the semantics correctly identifies semantic dependencies in tests which would have thin-air executions, but without a definition of reads from and an axiom like $\textit{acyclic}(\rrfx \cup \rsdepx)$, \PwT{} does not forbid the thin-air executions (as in test cases 4, 5 and 10).
+In the \PwTc{} model there is a restriction which forbids cycles in $\rrfx \cup \rsdepx$, so this correctly rules out thin-air executions.
 In jctc16, we see that the restrictions of \PwTc{} are too strong, this is because the coherence restrictions in C++ are stronger than those of Java.
@@
     jctc13                         & fail   & 2.247               & pass   & 2.101 \\
-    jctc14                         & unsupported   & --           & unsupported   & -- \\
-    jctc15                         & unsupported   & --           & unsupported   & -- \\
+    % jctc14                         & unsupported   & --           & unsupported   & -- \\
+    % jctc15                         & unsupported   & --           & unsupported   & -- \\
     jctc16                         & pass   & 9.232               & fail   & 8.492 \\
@@
     jctc18                         & pass   & 2.247               & pass   & 177.292 \\
-    jctc19                         & unsupported   & --           & unsupported   & -- \\
-    jctc20                         & unsupported   & --           & unsupported   & -- \\
+    % jctc19                         & unsupported   & --           & unsupported   & -- \\
+    % jctc20                         & unsupported   & --           & unsupported   & -- \\
   \end{tabularx}

### 2021-10-06 paper/intro.tex
AEC updates

@@
 
-Introductory programmers are taught \emph{sequential abstraction}: that the
+Novice programmers are taught \emph{sequential abstraction}: that the
 program $\aCmd_1\SEMI \aCmd_2$ executes $\aCmd_1$ before $\aCmd_2$.  Since
@@
 expect: programs write to, and read from, memory references.
-This can be thought of as a total order of reads and writes (pink arrows \tikz[baseline]{\draw [po] (0,0.1) -- (0.5,0.1);}),
-where each read has a matching \emph{fulfilling} write (green arrows \tikz[baseline]{\draw [rf] (0,0.1) -- (0.5,0.1);}),
+This can be thought of as a total order over memory actions ($\xwki$),
+where each read has a matching \emph{fulfilling} write ($\xrf$),
 for example:
@@
       \event{wy2}{\DW{y}{2}}{right=of wx1}
-      \event{ry2}{\DR{y}{2}}{right=3em of wy2}
+      \event{ry2}{\DR{y}{2}}{right=of wy2}
       \event{rx1}{\DR{x}{1}}{right=of ry2}
@@
       \rf[out=15,in=165]{wx1}{rx1}
-      \po{wx0}{wx1}
-      \po{wx1}{wy2}
-      \po{wy2}{ry2}
-      \po{ry2}{rx1}
+      \wki{wx0}{wx1}
+      \wki{wx1}{wy2}
+      \wki{wy2}{ry2}
+      \wki{ry2}{rx1}
     \end{tikzinline}}
@@
 
-This model naturally extends to the case of shared-memory concurrency, leading to a \emph{sequentially consistent}
+This model extends naturally to the case of shared-memory concurrency, leading to a \emph{sequentially consistent}
 semantics \cite{Lamport:1979:MMC:1311099.1311750}, in which \emph{program order} inside a thread implies
@@
       \rf[out=15,in=165]{wx1}{rx1}
-      \po{wx0}{wx1}
-      \po{wx1}{wy2}
-      \po{ry2}{rx1}
+      \wki{wx0}{wx1}
+      \wki{wx1}{wy2}
+      \wki{ry2}{rx1}
     \end{tikzinline}}
 \end{gather*}
+In general, there will be many such executions, reflecting different
+interleavings of the threads.
 
@@
 This has led to \emph{relaxed memory models}, in which the requirement of
-sequential consistency is weakened to only apply \emph{per-location} and not globally
-over the whole program. This allows executions that
+sequential consistency is weakened to only apply \emph{per-location}. This allows executions that
 are inconsistent with program order, such as:
@@
       \rf[out=15,in=165]{wx0}{rx0}
-      \po{wx0}{wx1}
+      \wki{wx0}{wx1}
       \wk[out=-165,in=-15]{rx0}{wx1}
@@
 \end{gather*}
+Reads must be ordered between the available writes, creating \emph{antidependencies} $(\DR{x}{0}\xwk\DW{x}{1})$.
 
 In such models, the causal order between events is important,
-and includes control and data dependencies, to avoid
+and includes control and data dependencies ($\xpo$), to avoid
 paradoxical ``out of thin air'' examples such as:
@@
 Other work in relaxed memory has shown that tooling is especially useful to researchers, architects, and language specifiers, enabling them to build intuitions experimentally~\cite{DBLP:conf/esop/PaviottiCPWOB20,PrideMM,alglave,Batty:2011:MCC:1926385.1926394}.
-Unfortunately, it is not obvious that tools can be built for all thin-air free models, the calculation of \citet{Pichon-Pharabod:2016:CSR:2837614.2837616} does not have a termination proof for an arbitrary input, and the enormous state space for the operational models of \citet{DBLP:conf/popl/KangHLVD17} and \citet{DBLP:journals/pacmpl/ChakrabortyV19} is a daunting prospect for a tool builder -- and as yet no tool exists for automatically evaluating these models.
+Unfortunately, it is not obvious that tools can be built for all thin-air-free models, the calculation of \citet{Pichon-Pharabod:2016:CSR:2837614.2837616} does not have a termination proof for an arbitrary input, and the enormous state space for the operational models of \citet{DBLP:conf/popl/KangHLVD17} and \citet{DBLP:journals/pacmpl/ChakrabortyV19} is a daunting prospect for a tool builder -- and as yet no tool exists for automatically evaluating these models.
 We describe a tool, \PwTer, for automatically evaluating \PwT{} in \textsection\ref{sec:tool}.

### 2021-10-07 paper/drf.tex
Removing some todos.
AEC draft

@@
 \label{sec:sc}
+
+\todo{Remove this section?}
 
@@
   $({\lt_*}\setminus{\rpox})$: $\rrfe$ (reads-from-external) orders writes
-  before reads, satisfying \ref{par-rf-le}; $\rcae$
+  before reads, satisfying \ref{par-le-rf1} and \ref{par-le-rf2}; $\rcae$
   (coherence-after-external) orders read and write accesses before writes,
-  satisfying \ref{pom-rf-block}. (Within a thread, \ref{seq-le-mca2} induces
-  order that is included in ${\rppo}$.)
+  satisfying \ref{pom-rf-block}. (Within a thread, \ref{seq-le-delays-rf} and
+  \ref{seq-le-rf-rf} induce order that is included in ${\rppo}$.)
 

### 2021-10-07 paper/intro.tex
AEC draft

@@
 \cite{DBLP:journals/pacmpl/PodkopaevLV19}. \citet{DBLP:journals/pacmpl/JagadeesanJR20} defined
-Pomsets with Preconditions (\PwP), which use preconditions and logic to
-calculate dependencies for a Java-like language targeting multicopy-atomic (\mca{}) hardware, such as
+Pomsets\footnote{A pomset is a labeled partial order.} with Preconditions (\PwP), which use preconditions and logic to
+calculate dependencies for a Java-like language targeting {multicopy-atomic} (\mca{}) hardware, such as
 \armeight{} \cite{DBLP:journals/pacmpl/PulteFDFSS18}.  However, neither paper
@@
 $\aCmd_1$ and the denotation of $\aCmd_2$, and \emph{direct} in that these
-can be calculated independently.  The definition is associative: the
-denotation of $((\aCmd_1\SEMI \aCmd_2)\SEMI \aCmd_3)$ is the same as the
-denotation of $(\aCmd_1\SEMI (\aCmd_2\SEMI \aCmd_3))$.  It also validates
-expected laws concerning the interaction of sequencing and conditional
-execution.
+can be calculated independently.
+
+The model has been formalized in Coq.  We have formally verified that the
+sequential composition satisfies the expected monoid laws
+(\reflem{lem:monoid}).  In addition we have formally verified that
+\begin{math}
+  \sem{\xIFTHEN{\aForm}{
+      \xSEMI{\aCmd_1}{\aCmd_3}
+    }{
+      \xSEMI{\aCmd_2}{\aCmd_3}
+    }}
+  \supseteq
+  \sem{\xSEMI{
+      \xIFTHEN{\aForm}{\aCmd_1}{\aCmd_2}
+    }{
+      \aCmd_3
+    }}
+\end{math}
+(\reflem{lem:if}\ref{lem:if:seq}).
+
+% The
+% definition is associative: the denotation of
+% $((\aCmd_1\SEMI \aCmd_2)\SEMI \aCmd_3)$ is the same as the denotation of
+% $(\aCmd_1\SEMI (\aCmd_2\SEMI \aCmd_3))$.  It also validates expected laws
+% concerning the interaction of sequencing and conditional execution.
 
@@
 concurrency.  In \textsection\ref{sec:mca}, we define \PwTmca{},
-which provides a Java-like model for \mca{} hardware, similar to that of
+which provides a Java-like model for {multicopy-atomic} (\mca{}) hardware, similar to that of
 \citet{DBLP:journals/pacmpl/JagadeesanJR20};
@@
 sequential consistency is weakened to only apply \emph{per-location}. This allows executions that
-are inconsistent with program order, such as:
+are inconsistent with program order, such as the following, which contains an
+\emph{antidependency} $(\xwk)$:
 \begin{gather*}
@@
 \end{gather*}
-Reads must be ordered between the available writes, creating \emph{antidependencies} $(\DR{x}{0}\xwk\DW{x}{1})$.
+% Reads must be ordered between blocking writes, creating
+% \emph{antidependencies} such as $\DRP{x}{0}\xwk\DWP{x}{1}$.
 
@@
 To address this, \citet{DBLP:journals/pacmpl/JagadeesanJR20} introduced
-\emph{Pomsets with Preconditions (\PwP{})}, where events are labeled with logical
+\emph{Pomsets with Preconditions} (\PwP{}), where events are labeled with logical
 formulae.  Nontrivial preconditions are introduced by store actions (modeling
@@
   \hbox{\begin{tikzinline}[node distance=1.5em]
-      \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
+      \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \bigmid \DW{z}{0}}{}
     \end{tikzinline}}
 \end{gather*}
-Preconditions are discharged by being ordered after a read (we assume the
-usual precedence for logical operators---$\lnot$, $\land$, $\lor$, $\limplies$): 
+In this diagram, $(s {<} 1)$ is a control dependency and $(r{*}s){=}0$ is a data dependency. Preconditions are discharged by being ordered after a read (we assume the
+usual precedence for logical operators): 
 \begin{gather*}
@@
       \event{ry0}{\DR{y}{0}}{right=of rx0}
-      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
+      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \bigmid \DW{z}{0}}{right=of ry0}
       \po{ry0}{wz0}
@@
         \xform{rx0d}{(0{=}r) \limplies \bForm}{right=of rx0}
-        \xo{rx0}{rx0d}
+        %\xo{rx0}{rx0d}
       \end{tikzinline}}
@@
         \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
-        \xo{ry0}{ry0d}
+        %\xo{ry0}{ry0d}
       \end{tikzinline}}
@@
     \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
-        \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
+        \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \bigmid \DW{z}{0}}{}
         \xform{wz0d}{\bForm[r{*}s/z]}{right=of wz0}
-        \xo{wz0}{wz0d}
+        %\xo{wz0}{wz0d}
       \end{tikzinline}}
@@
 For a sequentially consistent semantics, sequential composition is
-straightforward: we apply each predicate transformer to the preconditions of
-subsequent events, composing the predicate transformers.  (In subsequent
-diagrams, we only show predicate transformers for reads.)
+straightforward: we apply each predicate transformer to subsequent
+preconditions, composing the predicate transformers.
 \begin{gather*}
   \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
-  \\[-1.5ex]
-  \nonumber
+  \\
   \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
@@
       \event{ry0}{\DR{y}{0}}{right=of rx0}
-      \event{wz0}{(0{=}r) \limplies (0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
-      \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{left=of rx0}
-      \po{rx0}{ry0}
-      \po{ry0}{wz0}
-      \xo{rx0}{rx0ry0d}
-      \xo[out=155,in=15]{ry0}{rx0ry0d}
+      \event{wz0}{(0{=}r) \limplies (0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \bigmid \DW{z}{0}}{right=of ry0}
+      \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm[r{*}s/z]}{right=of wz0}
+      \wki{rx0}{ry0}
+      \wki{ry0}{wz0}
+      %\xo{rx0}{rx0ry0d}
+      %\xo[out=155,in=15]{ry0}{rx0ry0d}
       % \xo{wz0}{rx0ry0d}
@@
 \end{gather*}
-This model works for the sequentially consistent case, but needs to be
-weakened for the relaxed case. The key observation of this paper is
+This works for the sequentially consistent case, but needs to be
+weakened for the relaxed case.
+
+The key observation of this paper is
 that rather than working with one predicate transformer, we should
@@
 of events.
-
 For example, for single-event pomsets, there are two predicate
@@
 We visualize this by including more than one transformed predicate,
-with an edge leading to the dependent one. For example:
+with a dotted edge leading to the dependent one ($\xxo$). For example:
 \begin{align*}
@@
   %   \hbox{\begin{tikzinline}[node distance=1ex and1.5em]
-  %     \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
+  %     \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \bigmid \DW{z}{0}}{}
   %     \xform{wz0i}{\bForm}{left=.4em of wz0}
@@
 which maps $\bForm$ to $(0{=}s) \limplies \bForm$.
+(In subsequent diagrams, we only show predicate transformers for reads.)
 \begin{gather*}
@@
   \\
-  \nonumber
   \hbox{\begin{tikzinline}[node distance=2ex and 1.2em]
@@
       \event{ry0}{\DR{y}{0}}{above right=of rx0ry0d}
-      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
+      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \bigmid \DW{z}{0}}{right=of ry0}
       \po{ry0}{wz0}
@@
 In the diagram, the dotted lines indicate set inclusion into the index of the transformer-family.
-As a sanity check, we can see that sequential composition is
+As a quick correctness test, we can see that sequential composition is
 associative in this case, since it does not matter whether we
-associate to the left, with intermediate step:
-\begin{gather*}
-  \PR{x}{r}\SEMI \PR{y}{s}
-  \\
-  \nonumber
-  \hbox{\begin{tikzinline}[node distance=2ex and 1.5em]
-      \xform{rx0d}{(0{=}r) \limplies \bForm}{}
-      \event{rx0}{\DR{x}{0}}{right=of rx0d}
-      \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{right=of rx0}
-      \event{ry0}{\DR{y}{0}}{right=of rx0ry0d}
-      \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
-      \xform{wz0i}{\bForm}{left=of rx0d}
-      \xo{rx0}{rx0d}
-      \xo{ry0}{ry0d}
-      \xo{rx0}{rx0ry0d}
-      \xo{ry0}{rx0ry0d}
-    \end{tikzinline}}
-\end{gather*}
-or to the right, with intermediate step:
+associate to the left, with the intermediate step eliding $\DWP{z}{0}$ in the
+diagram above,
+% \begin{gather*}
+%   \PR{x}{r}\SEMI \PR{y}{s}
+%   \\
+%   \nonumber
+%   \hbox{\begin{tikzinline}[node distance=2ex and 1.5em]
+%       \xform{rx0d}{(0{=}r) \limplies \bForm}{}
+%       \event{rx0}{\DR{x}{0}}{right=of rx0d}
+%       \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{right=of rx0}
+%       \event{ry0}{\DR{y}{0}}{right=of rx0ry0d}
+%       \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
+%       \xform{wz0i}{\bForm}{left=of rx0d}
+%       \xo{rx0}{rx0d}
+%       \xo{ry0}{ry0d}
+%       \xo{rx0}{rx0ry0d}
+%       \xo{ry0}{rx0ry0d}
+%     \end{tikzinline}}
+% \end{gather*}
+or to the right, with the intermediate step:
 \begin{gather*}
@@
       \event{ry0}{\DR{y}{0}}{}
-      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
+      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \bigmid \DW{z}{0}}{right=of ry0}
       \xform{ry0d}{(0{=}s) \limplies \bForm}{left=of ry0}

### 2021-10-07 paper/mca1.tex
AEC draft

@@
 
-We derive a model of concurrent computation by adding \emph{parallel
-  composition} and \emph{reads-from} to \reffig{fig:seq}.  To model coherence
+We derive a model of concurrent computation by adding
+% \emph{parallel composition} and 
+\emph{reads-from} to \reffig{fig:seq}.  To model coherence
 and synchronization, we add \emph{delay} to the rule for sequential
-composition.  For \mca{} architectures, it is sufficient to encode delay in
+composition.  For \emph{multicopy-atomic} (\mca{}) architectures, it is sufficient to encode delay in
 the pomset order.  The resulting model, \PwTmca{1}, supports optimal lowering
 for relaxed access on \armeight{}, but requires extra synchronization for
-acquiring reads.
+acquiring reads.  (\emph{Lowering} is the translation of language-level
+operators to machine instructions.  A lowering is \emph{optimal} if it
+provides the most efficient execution possible.)
 
@@
   \label{def:pwt:mca1}
-  A \PwTmca{1} is a \PwT{} (\refdef{def:pomset}) equipped
+  The definition of \PwTmca{1} extends that of \PwT{} %(\refdef{def:pomset})
   with a relation $\rrfx$ such that 
@@
   \label{def:pwt:mca:complete}
-  A \PwTmca{} is \emph{complete} if %it is a complete \PwT{} (\refdef{def:pomset}) and
+  The definition of completeness extends \refdef{def:pomset} as follows:
   \begin{multicols}{2}
     \begin{enumerate}[,label=(\textsc{c}\arabic*),ref=\textsc{c}\arabic*]
-    \item[\eqref{top-kappa}]
-      \eqref{top-term}\;
-      as in \refdef{def:pomset},
       \setcounter{enumi}{\value{Brf}}
@@
   \end{multicols}
-  \bigskip
 
-  %The semantic rules are as follows:
-  \input{fig-mca1.tex}
+  \noindent
+  The semantic function extends \reffig{fig:seq} as follows:
+  \begin{multicols}{2}
+    \begin{enumerate}[topsep=0pt,label=(\textsc{i}\arabic*),ref=\textsc{i}\arabic*]
+    \item[{\labeltext[\textsc{p}7]{(\textsc{p}7)}{par-rf}}]
+      \labeltext[\textsc{s}7]{(\textsc{s}7)}{seq-rf}\;
+      \labeltext[\textsc{i}7]{(\textsc{i}7)}{if-rf}\;
+      ${\rrfx}\rextends{\rrfx_1}{\rrfx_2}$,
+      \setcounter{enumi}{\value{le}}
+    \item[]
+      \begin{enumerate}[leftmargin=0pt]
+      \item \label{seq-le-delays}
+        if $\labeling_1(\bEv) \rdelays \labeling_2(\aEv)$ then $\bEv\le\aEv$.
+      \end{enumerate}
+    \end{enumerate}  
+  \end{multicols}
+  % The semantic rules are as follows:
+  % \input{fig-mca1.tex}
 \end{definition}
-Let
-\begin{math}
-  \sem{\aCmd_1 \RPAR \aCmd_2} = \sPAR{\sem{\aCmd_1}}{\sem{\aCmd_2}}.
-\end{math}
+% Let
+% \begin{math}
+%   \sem{\aCmd_1 \RPAR \aCmd_2} = \sPAR{\sem{\aCmd_1}}{\sem{\aCmd_2}}.
+% \end{math}
 We write $\semmca{1}{}$ for the semantic function when it is unclear from context.
@@
 Lemmas \ref{lem:monoid} and \ref{lem:if} hold for \PwTmca{1}.  For further
-discussion of items \eqref{lem:ifelse:if:if1} and \eqref{lem:ifelse:if:if2},
+discussion of item \eqref{lem:ifelse:if:if} %and \eqref{lem:ifelse:if:if2},
 see \textsection\ref{sec:delay}.

### 2021-10-07 paper/mca2.tex
in SEQ, dep respects dep1 and dep2 as needed for associativity with bimplication
AEC draft

@@
   \label{def:pwt:mca2}
-  A \PwTmca{2} is a \PwT{} (\refdef{def:pomset}) equipped with an injective
-  relation $\rrfx$ that satisfies requirements \ref{pom-rf-match} and
-  \ref{pom-rf-block} of \refdef{def:pwt:mca1}.
+  The definition of \PwTmca{2} is derived from that of \PwTmca{1} by
+  removing \ref{pom-rf-le} and adding the following:
 
-  A \PwTmca{2} is \emph{complete} if it satisfies
-  \ref{top-kappa}, \ref{top-term}, and \ref{top-rf}---this is the same as for
-  \PwTmca{1}.
+  \begin{enumerate}[topsep=0pt]
+  \item[{\labeltext[\textsc{p}6a]{(\textsc{p}6a)}{par-le-rf1}}]
+    if $\bEv\in\aEvs_1$, $\aEv\in\aEvs_2$ and $\bEv\xrfx\aEv$ then $\bEv\lt\aEv$,
+  \item[{\labeltext[\textsc{p}6b]{(\textsc{p}6b)}{par-le-rf2}}]
+    if $\bEv\in\aEvs_1$, $\aEv\in\aEvs_2$ and $\aEv\xrfx\bEv$ then $\aEv\lt\bEv$,
+  \item[{\labeltext[\textsc{s}6a]{(\textsc{s}6a)}{seq-le-delays-rf}}]
+    if $\bEv\in\aEvs_1$, $\aEv\in\aEvs_2$ and $\labeling_1(\bEv) \rdelays \labeling_2(\aEv)$ 
+    then either $\bEv\xrfx\aEv$ or $\bEv\le\aEv$,
+  \item[{\labeltext[\textsc{s}6b]{(\textsc{s}6b)}{seq-le-rf-rf}}]
+    if $\bEv\in\aEvs_1$, $\aEv\in\aEvs_2$ and $\labeling_1(\bEv) \rblocks \labeling_2(\aEv)$
+    then $\cEv\xrfx\aEv$ implies $\bEv\le\cEv$.
+    % \\
+    % if $\labeling_1(\cEv) \rblocks \labeling_2(\aEv)$ and $\bEv\xrfx\aEv$
+    % then $\cEv\le\bEv$.
+  % \item[{\labeltext[\textsc{i}6a]{(\textsc{i}6a)}{if-le-rf}}]
+  %   ${\rrfx}\subseteq {\rrfx_1}\cup {\rrfx_2}$
+  \end{enumerate}
+  % The definition of \PwTmca{2} is a \PwT{} (\refdef{def:pomset}) equipped with an injective
+  % relation $\rrfx$ that satisfies requirements \ref{pom-rf-match} and
+  % \ref{pom-rf-block} of \refdef{def:pwt:mca1}.
 
-  \input{fig-mca2.tex}
+  % A \PwTmca{2} is \emph{complete} if it satisfies
+  % \ref{top-kappa}, \ref{top-term}, and \ref{top-rf}---this is the same as for
+  % \PwTmca{1}.
+
+  %\input{fig-mca2.tex}
 \end{definition}
@@
 % 
-% The use of $\rextendsdef{}{}$ in \ref{if-rf-extends}
+% The use of $\rextendsdef{}{}$ in \ref{if-rf-project}
 % ensures that no $\rrfx$ is introduced between events in
@@
 
-\todo{Example using \ref{seq-le-delays-rf} and \ref{seq-le-rf-rf}. To make
-  space, \reflem{lem:mca2} could move to appendix.}
+\todo{Example using \ref{seq-le-delays-rf} and \ref{seq-le-rf-rf}. Perhaps
+  move \reflem{lem:mca2} to appendix.}
 
 With the weakening of \ref{seq-le-delays}, we must be careful not to allow
-spurious pairs to be added to the $\rrfx$ relation.  Thus we add
-\ref{par-rf-extends},
-\ref{seq-rf-extends}, and
-\ref{if-rf-extends}.
-For example, \ref{if-rf-extends} ensure that
+spurious pairs to be added to the $\rrfx$ relation.  For example, %\ref{if-rf-project} ensures that
 \begin{math}
@@
 \end{math}
-does not include 
+should not include 
 \begin{math}
@@
 \end{math}, taking $\rrfx$ from the left and $\lt$ from the right.
+The use of $\rextends{}{}$ ensures this.
 
@@
 phantom events ($\fmrg{}$).  Using this construction, the following lemma
-gives a post-hoc verification technique for $\rrfx$.
+gives a post-hoc verification technique for $\rrfx$.  Let $\fmrginv{}$ be the
+inverse of $\fmrg{}$.
 \begin{lemma}
@@
     $(\exists\aEv'\in\fmrginv{\aEv})$
-    $\bEv'\xpox\aEv'$ and $(\not\exists\cEv')$
+    $\bEv'\xpox\aEv'$ and $(\not\exists\cEv)$
     $\labelingForm(\cEv)$ is a tautology and

### 2021-10-07 paper/oopsla.tex
added example showing the need for resptec
AEC draft

@@
 optimizations are inconsistent with local \drf{} guarantees.  All of the
-examples that prove inconsistency have the same shape: they combine read
-introduction and case analysis (aka, if-closure).  Effectively, this turns
-one read into two, where different conditional branches can be taken for the
-two copies of the read.  This is reminiscent of the type of \emph{bait and
-  switch} behavior noted by \citet{DBLP:journals/pacmpl/JagadeesanJR20}: the
-promising semantics (\PS{}) \cite{DBLP:conf/popl/KangHLVD17} and related
-models
+examples that prove inconsistency have the same shape: they combine
+read-introduction and case analysis (aka, if-introduction).  Effectively,
+this turns one read into two, where different conditional branches can be
+taken for the two copies of the read.  This is reminiscent of the type of
+\emph{bait and switch} behavior noted by
+\citet{DBLP:journals/pacmpl/JagadeesanJR20}: the promising semantics (\PS{})
+\cite{DBLP:conf/popl/KangHLVD17} and related models
 \citep{DBLP:conf/esop/JagadeesanPR10,DBLP:journals/pacmpl/ChakrabortyV19,Manson:2005:JMM:1047659.1040336},
@@
 \begin{itemize}
-\item read introduction,
-\item if introduction, branching on the read just introduced.
+\item read-introduction,
+\item if-introduction, branching on the read just introduced.
 \end{itemize}
 We believe this combination of optimizations is unsound.  This is obviously
-the case in \cXI: read introduction may cause undefined behavior (\ub), due to
+the case in \cXI: read-introduction may cause undefined behavior (\ub), due to
 the possible introduction of a data race.
@@
 \begin{itemize}
-\item read introduction may result in the undefined value $\UNDEFINED$, due
+\item read-introduction may result in the undefined value $\UNDEFINED$, due
   to the possible introduction of a data race \cite{DBLP:conf/cgo/ChakrabortyV17}, and
@@
 subsequent case analysis on the value read.  \PwT{} validates case analysis,
-but invalidates read introduction.
+but invalidates read-introduction.
 
@@
         \event{a2}{\DR{x}{2}}{right=of a1}
-        \event{a3}{1{=}x\limplies 2{=}x\limplies x{<} x\mid\DW{y}{1}}{right=of a2}
+        \event{a3}{1{=}x\limplies 2{=}x\limplies x{<} x\bigmid\DW{y}{1}}{right=of a2}
         \po[out=20,in=160]{a1}{a3}
@@
         % \event{a1}{\DR{x}{1}}{right=of a0}
-        \event{a2}{r{\geq}0\mid\DW{y}{1}}{right=3em of a1}      
+        \event{a2}{r{\geq}0\bigmid\DW{y}{1}}{right=3em of a1}      
         % \wk{a0}{a1}
@@
         % \event{a1}{\DR{x}{1}}{right=of a0}
-        \event{a2}{0{\geq}0\mid\DW{y}{1}}{right=3em of a1}      
+        \event{a2}{0{\geq}0\bigmid\DW{y}{1}}{right=3em of a1}      
         % \wk{a0}{a1}
@@
         \hbox{\begin{tikzinline}[node distance=1em]
-            \event{a}{\aExp\mid\DW{x}{1}}{}
+            \event{a}{\aExp\bigmid\DW{x}{1}}{}
           \end{tikzinline}}
@@
         \hbox{\begin{tikzinline}[node distance=1em]
-            \event{a}{\lnot\aExp\mid\DW{x}{1}}{}
+            \event{a}{\lnot\aExp\bigmid\DW{x}{1}}{}
           \end{tikzinline}}
@@
         \hbox{\begin{tikzinline}[node distance=1em]
-            \event{a}{\aExp\mid\DW{y}{1}}{}
+            \event{a}{\aExp\bigmid\DW{y}{1}}{}
           \end{tikzinline}}
@@
         \hbox{\begin{tikzinline}[node distance=1em]
-            \event{a}{\lnot\aExp\mid\DW{y}{1}}{}
+            \event{a}{\lnot\aExp\bigmid\DW{y}{1}}{}
           \end{tikzinline}}
@@
         \hbox{\begin{tikzinline}[node distance=1em]
-            \event{a}{\lnot\aExp\mid\DW{x}{1}}{}
-            \event{b}{\aExp\mid\DW{y}{1}}{right=of a}
+            \event{a}{\lnot\aExp\bigmid\DW{x}{1}}{}
+            \event{b}{\aExp\bigmid\DW{y}{1}}{right=of a}
           \end{tikzinline}}
@@
       \hbox{\begin{tikzinline}[node distance=1em]
-          \event{a}{\aExp\mid\DR{x}{1}}{}
+          \event{a}{\aExp\bigmid\DR{x}{1}}{}
         \end{tikzinline}}
@@
       \hbox{\begin{tikzinline}[node distance=1em]
-          \event{a}{r{=}1\mid\DW{y}{1}}{}
+          \event{a}{r{=}1\bigmid\DW{y}{1}}{}
         \end{tikzinline}}
@@
       \hbox{\begin{tikzinline}[node distance=1em]
-          \event{a}{\lnot\aExp\mid\DR{x}{1}}{}
+          \event{a}{\lnot\aExp\bigmid\DR{x}{1}}{}
         \end{tikzinline}}
@@
       \hbox{\begin{tikzinline}[node distance=1em]
-          \event{a}{\aExp\mid\DR{x}{1}}{}
-          \event{b}{\aExp\mid\DW{y}{1}}{right=of a}
+          \event{a}{\aExp\bigmid\DR{x}{1}}{}
+          \event{b}{\aExp\bigmid\DW{y}{1}}{right=of a}
           \po{a}{b}
@@
           \event{a}{\DR{x}{1}}{}
-          \event{b}{\aExp\mid\DW{y}{1}}{right=of a}
+          \event{b}{\aExp\bigmid\DW{y}{1}}{right=of a}
           \po{a}{b}
@@
       \hbox{\begin{tikzinline}[node distance=1em]
-          \event{a}{\lnot\aExp\mid\DR{x}{1}}{}
-          \event{b}{r{=}1\mid\DW{y}{1}}{left=of a}
+          \event{a}{\lnot\aExp\bigmid\DR{x}{1}}{}
+          \event{b}{r{=}1\bigmid\DW{y}{1}}{left=of a}
         \end{tikzinline}}
@@
   not introduce thin air reads.  Associating to the right, the intermediate
-  state of the example in \textsection\ref{sec:addr} is:
+  state of \ref{addr2} (\textsection\ref{sec:addr}) is:
   \begin{align*}
@@
       \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-          \event{a2}{r\EQ2\mid\DR{\REF{2}}{1}}{}
-          \event{a3}{(r\EQ2\limplies 1\EQ s) \limplies s\EQ1\mid\DW{x}{1}}{right=of a2}
+          \event{a2}{r\EQ2\bigmid\DR{\REF{2}}{1}}{}
+          \event{a3}{(r\EQ2\limplies 1\EQ s) \limplies s\EQ1\bigmid\DW{x}{1}}{right=of a2}
           \po{a2}{a3}
@@
     %   \hbox{\begin{tikzinline}[node distance=.5em and 1.5em]
-    %     \event{b}{s\EQ1\mid\DW{x}{1}}{}
+    %     \event{b}{s\EQ1\bigmid\DW{x}{1}}{}
     %   \end{tikzinline}}
@@
           % \event{a1}{\DR{y}{2}}{}
-          \event{a2}{r\EQ2\mid\DR{\REF{2}}{1}}{}%right=of a1}
-          \event{a3}{r\EQ2\land\REF{2}\EQ1\mid\DW{x}{1}}{right=of a2}
+          \event{a2}{r\EQ2\bigmid\DR{\REF{2}}{1}}{}%right=of a1}
+          \event{a3}{r\EQ2\land\REF{2}\EQ1\bigmid\DW{x}{1}}{right=of a2}
           \po{a2}{a3}
@@
       \event{a2}{\DR{x}{1}}{right=of a1}
-      \event{a3}{1\EQ1\land1\EQ x \land x\EQ1 \land x=x\mid\DW{y}{1}}{right=of a2}
+      \event{a3}{1\EQ1\land1\EQ x \land x\EQ1 \land x=x\bigmid\DW{y}{1}}{right=of a2}
     \end{tikzinline}}

### 2021-10-07 paper/outro.tex
AEC updates
AEC draft

@@
 dead-write-removal are unsound.  We expect that these can be validated by
-allowing events with different labels to merge. % in some instances.
+allowing events with different actions to merge. % in some instances.
 
@@
 \textsection\ref{sec:promising}).  Nonetheless, read introduction is
-ubiquitous is some compilers.  It would be interesting to know if there is a
+ubiquitous in some compilers.  It would be interesting to know if there is a
 performance penalty for banning read introduction.

### 2021-10-15 paper/appendix.tex
hacking at address dependencies
proabably fixed addr dependencies
Working on appendices
Draft of Fork-Join
Sorted out bib
Updated defs
Updates

@@
-\documentclass[conference,compsoc]{IEEEtran}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Missing includes
-\usepackage{graphicx}
-\usepackage[prologue]{xcolor}
-\definecolor[named]{ACMBlue}{cmyk}{1,0.1,0,0.1}
-\definecolor[named]{ACMYellow}{cmyk}{0,0.16,1,0}
-\definecolor[named]{ACMOrange}{cmyk}{0,0.42,1,0.01}
-\definecolor[named]{ACMRed}{cmyk}{0,0.90,0.86,0}
-\definecolor[named]{ACMLightBlue}{cmyk}{0.49,0.01,0,0}
-\definecolor[named]{ACMGreen}{cmyk}{0.20,0,1,0.19}
-\definecolor[named]{ACMPurple}{cmyk}{0.55,1,0,0.15}
-\definecolor[named]{ACMDarkBlue}{cmyk}{1,0.58,0,0.21}
-\usepackage[bookmarksnumbered,unicode]{hyperref}
-\hypersetup{colorlinks,
-  linkcolor=ACMPurple,
-  citecolor=ACMPurple,
-  urlcolor=ACMDarkBlue,
-  filecolor=ACMDarkBlue}
-\usepackage{amsmath}
-\usepackage{amsthm}
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}
-%\newtheorem{conjecture}{Conjecture}
-\newtheorem{proposition}{Prop}
-\newtheorem{lemma}{Lemma}
-\theoremstyle{definition}
-\newtheorem{definition}{Def}
-\newtheorem{example}[definition]{Ex}
-\theoremstyle{remark}
-\newtheorem{remark}{Remark}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\RequirePackage{amssymb}  %% bizarre error previous def of Bbbk if after documentclass
+\documentclass[acmsmall,screen,anonymous,review]{acmart}\settopmatter{printfolios=true}
+\hypersetup{bookmarksnumbered,bookmarksopen=true,bookmarksdepth=3}
+\settopmatter{printfolios=true}
+\AtEndPreamble{%
+  \theoremstyle{acmdefinition}
+  \newtheorem{remark}[theorem]{Remark}
+  \newtheorem{candidate}[theorem]{Candidate}
+  \renewcommand{\theequation}{\fnsymbol{equation}}
+}
+\bibliographystyle{ACM-Reference-Format}
+\citestyle{acmauthoryear}   %% For author/year citations
 
-% \author{\IEEEauthorblockN{%
-%     Alan Jeffrey\IEEEauthorrefmark{1} and
-%     James Riely\IEEEauthorrefmark{2}
-% }
-% \IEEEauthorblockA{\IEEEauthorrefmark{1}Roblox}
-% \IEEEauthorblockA{\IEEEauthorrefmark{2}DePaul University}}
-\author{Anonymous\\Anonymous Institution}
-
-\title{Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers}
+\setcopyright{rightsretained}
+\acmPrice{}
+\acmDOI{}
+\acmYear{2022}
+\copyrightyear{2022}
+\acmSubmissionID{}
+\acmJournal{PACMPL}
+\acmVolume{0}
+\acmNumber{POPL}
+\acmArticle{0}
+\acmMonth{1}
+\startPage{1}
 
 \usepackage{macros}
-\usepackage[sectionbib,numbers,square]{natbib}
+% \showRAfalse
+\showSCOPEfalse
+\title{The Leaky Semicolon}
 
 \usepackage{xr}
-\newcounter{pomsetPreParCount}\setcounter{pomsetPreParCount}{8}
-\newcounter{pomsetXSemiCount}\setcounter{pomsetXSemiCount}{8}
 \externaldocument{paper}
+\makecounter{Bkappa} \setcounter{Bkappa}{2}
+\makecounter{Btau}   \setcounter{Btau}{3}
+\makecounter{Bterm}  \setcounter{Bterm}{4}
 \begin{document}
-\pagestyle{plain}
-\appendices
-\input{discussion}
-\bibliographystyle{abbrvnat}
+\appendix
+\input{oopsla.tex}
+\input{discussion.tex}
+\input{arm.tex}
+\input{drf.tex}
+\input{extra.tex}
 \bibliography{bib}
+\end{document}
 
-\end{document}
+Alex Aiken PLDI keynote:
+\begin{itemize}
+\item PAD Machines: Parallel, Accelerated (GPUs..), Distributed (memory)
+\item ``Registered'' memory: shared with network
+\item ``Zero-copy'' memory: shared with GPUs
+  \begin{itemize}
+  \item
+    \url{http://docs.nvidia.com/cuda/cuda-c-programming-guide/index.html#um-unified-memory-programming-hd}
+    Unified Memory offers a “single-pointer-to-data” model that is
+    conceptually similar to CUDA’s zero-copy memory. One key difference
+    between the two is that with zero-copy allocations the physical location
+    of memory is pinned in CPU system memory such that a program may have
+    fast or slow access to it depending on where it is being accessed
+    from. Unified Memory, on the other hand, decouples memory and execution
+    spaces so that all data accesses are fast.
+  \end{itemize}
+\item Software manages ``NUMA domains''
+\item In PAD machines, most computation happening on the accelerators.
+\item Bad Programming models:
+  \begin{itemize}
+  \item Data Analytics, Machine Learning: Single program in Spark Hadoop,
+    Etc.  Accessible to many people, Low performance
+  \item HPC: MPI, OpenMP, Vector Intrinsics, Cuda, etc...  Experts only, High performance
+  \end{itemize}
+\item Task based models are good.
+  \begin{itemize}
+  \item Asynchronous functions that get mapped to real resources.  May have
+    sub-tasks or not.  
+  \item Tasks capture locality:
+    \begin{itemize}
+    \item Task is co-located with its arguments.  
+    \item Tasks are a ``local address space'' model.
+    \end{itemize}
+  \item Tasks capture asynchrony
+  \end{itemize}
+\item Semantics
+  \begin{itemize}
+  \item \xmark Explicit Parallelism: X10, Chapel
+  \item \cmark Implicit Parallelism (dataflow) --- much easier to reason about
+    (sequential semantics) --- automated data movement
+    \begin{itemize}
+    \item Layout Explicitly programmed: StarPU 
+    \item Layout computed statically: PaRSEC/PTG, DPJ, Sequoia
+    \item Layout computed dynamically: Legion, PaRSEC/DTD, TensorFlow
+    \end{itemize}
+  \end{itemize}
+\item Partitioning:
+  \begin{itemize}
+  \item Flat disjoint partitions: TensorFlow, PyTorch.
+    \begin{itemize}
+    \item Not compositional: Problem is that different parts of the
+      application may want different views of the data.
+    \end{itemize}
+  \end{itemize}
+\item Hierarchical: StarPU
+  \begin{itemize}
+  \item Partitions created/destroyed
+  \item Partitions are disjoint
+  \item Only one partition of collection at a time
+  \item Only leaves can be used by application
+  \item Eager Repartitioning move data unnecessarily
+  \end{itemize}
+\item Legion allows multiple hierarchical partitions
+  \begin{itemize}
+  \item Compositional: data movement is lazy
+  \item Task dependence analysis requires alias analysis
+  \end{itemize}
+\end{itemize}
 

### 2021-10-15 paper/intro.tex
standardizing backarrows to be below
Updates

@@
 \cite{DBLP:journals/pacmpl/PodkopaevLV19}. \citet{DBLP:journals/pacmpl/JagadeesanJR20} defined
-Pomsets\footnote{A pomset is a labeled partial order.} with Preconditions (\PwP), which use preconditions and logic to
+Pomsets with Preconditions (\PwP), which use preconditions and logic to
 calculate dependencies for a Java-like language targeting {multicopy-atomic} (\mca{}) hardware, such as
@@
 \end{gather*}
-In general, there will be many such executions, reflecting different
-interleavings of the threads.
+We can represent such an execution as a labelled partial order, or
+\emph{pomset} \cite{GISCHER1988199}.  A program may give rise to many
+executions, each reflecting a different interleaving of the threads.
 
@@
       \rf[out=15,in=165]{wy1}{ry1}
-      \rf[out=165,in=15]{wx1}{rx1}
+      \rf[out=-165,in=-15]{wx1}{rx1}
       \po{rx1}{wy1}
@@
       \rf[out=15,in=165]{wy1}{ry1}
-      \rf[out=165,in=15]{wx1}{rx1}
+      \rf[out=-165,in=-15]{wx1}{rx1}
       \po{ry1}{wx1}

### 2021-10-18 paper/tool.tex
Add some text about complexity to tool.tex

@@
 
-It is noteworthy that the \PwT{} version of the semantics correctly identifies semantic dependencies in tests which would have thin-air executions, but without a definition of reads from and an axiom like $\textit{acyclic}(\rrfx \cup \rsdepx)$, \PwT{} does not forbid the thin-air executions (as in test cases 4, 5 and 10).
-In the \PwTc{} model there is a restriction which forbids cycles in $\rrfx \cup \rsdepx$, so this correctly rules out thin-air executions.
-In jctc16, we see that the restrictions of \PwTc{} are too strong, this is because the coherence restrictions in C++ are stronger than those of Java.
-The execution times give a good indication the poor scaling of the tool with program size: for longer test cases the tool takes exponentially longer to compute, and in some cases simply fails.
 The compositional nature of the semantics makes tool building practical, but it is not enough to make it scalable for large tests.
-Unsurprisingly the execution time is dominated by the calculation of the denotation, with the additional axiomatic filtering step of \PwTc{} being within the margin of error. 
+The execution times give a good indication the poor scaling of the tool with program size: for larger test cases, the tool takes exponentially longer to compute, and for the largest tests the memory footprint is too large for even a well-equipped computer.
+The definitions of $\sSEMI{\aPSS_1}{\aPSS_2}$ and $\sIF{\aForm}\sTHEN\aPSS_1\sELSE\aPSS_2\sFI$ -- in combination with the rules for reads and writes -- yields a complexity class of $\mathcal{O}((|\Val| + 1)^{n})$ where $n$ is the length of the longest chain of sequential compositions in the program.
+Some further optimisations may be possible within the tool to improve the situation, such as killing 'dead-end' pomsets at each sequence operator, or by doing a directed search for particular execution outcomes.
+Unsurprisingly the execution time is dominated by the calculation of the denotation, with the additional axiomatic filtering step of \PwTc{} being within margin of error difference of just calculating the \PwT{} semantics.
 % \PwTer{} will be made open source upon publication.
@@
 \begin{center}
-  \begin{tabularx}{0.9\textwidth}{c||c|c|c|c}
-    \multirow{2}{*}{\bf Test name} & \multicolumn{2}{c|}{\PwT{}} & \multicolumn{2}{c}{\PwTc{}} \\
-    \cline{2-5}
-                                   & Result & Execution Time (s)  & Result & Execution Time (s)  \\
-    \cline{1-5}
-    jctc1                          & pass   & 2.608               & pass   & 2.397 \\
-    jctc2                          & pass   & 25.754              & pass   & 25.780 \\
-    jctc3                          & pass   & 205.120             & pass   & 196.935 \\
-    jctc4                          & fail   & 2.110               & pass   & 2.269 \\
-    jctc5                          & fail   & 69.441              & pass   & 63.714 \\
-    jctc6                          & pass   & 12.489              & pass   & 11.245 \\
-    jctc7                          & pass   & 96.099              & pass   & 88.250 \\
-    jctc8                          & pass   & 2.473               & pass   & 2.482 \\
-    jctc9                          & pass   & 15.384              & pass   & 13.592 \\
-    jctc10                         & fail   & 513.234             & pass   & 494.133 \\
-    jctc11                         & $\bot$ & --                  & $\bot$ & \\
-    jctc12                         & $\bot$ & --                  & $\bot$ & \\
-    jctc13                         & fail   & 2.247               & pass   & 2.101 \\
-    % jctc14                         & unsupported   & --           & unsupported   & -- \\
-    % jctc15                         & unsupported   & --           & unsupported   & -- \\
-    jctc16                         & pass   & 9.232               & fail   & 8.492 \\
-    jctc17                         & pass   & 186.228             & pass   & 178.304 \\
-    jctc18                         & pass   & 2.247               & pass   & 177.292 \\
-    % jctc19                         & unsupported   & --           & unsupported   & -- \\
-    % jctc20                         & unsupported   & --           & unsupported   & -- \\
+  \begin{tabularx}{0.7\textwidth}{c||c|c|c}
+    \multirow{2}{*}{\bf Causality Test}   & \multicolumn{2}{c|}{\PwTc{}}    & \PwT{}\\
+    \cline{2-4}
+                                     & Result & Execution Time (s)  & Execution Time (s) \\
+    \cline{1-4}
+    jctc1                            & pass   & 2.397               & 2.608              \\
+    jctc2                            & pass   & 25.780              & 25.754             \\
+    jctc3                            & pass   & 196.935             & 205.120            \\
+    jctc4                            & pass   & 2.269               & 2.110              \\
+    jctc5                            & pass   & 63.714              & 69.441             \\
+    jctc6                            & pass   & 11.245              & 12.489             \\
+    jctc7                            & pass   & 88.250              & 96.099             \\
+    jctc8                            & pass   & 2.482               & 2.473              \\
+    jctc9                            & pass   & 13.592              & 15.384             \\
+    jctc10                           & pass   & 494.133             & 513.234            \\
+    jctc11                           & $\bot$ & --                  & --                 \\
+    jctc12                           & $\bot$ & --                  & --                 \\
+    jctc13                           & pass   & 2.101               & 2.247              \\
+    jctc17                           & pass   & 178.304             & 186.228            \\
+    jctc18                           & pass   & 177.292             & 2.247              \\
   \end{tabularx}
-  \caption{\label{fig:tool} Tool results for Java Causality Test Cases~\cite{PughWebsite}.
+  \caption{\label{fig:tool} Tool results for supported Java Causality Test Cases~\cite{PughWebsite}.
 $\bot$ indicates the tool failed to run for this test due to a memory overflow.

### 2021-10-18 paper/tool.tex
Revert "Add some text about complexity to tool.tex"
This reverts commit bd93a204e11ab22df60e343434c751dba2c096a1.

@@
 
+The execution times give a good indication the poor scaling of the tool with program size: for larger test cases, the tool takes exponentially longer to compute, and in some cases simply fails.
 The compositional nature of the semantics makes tool building practical, but it is not enough to make it scalable for large tests.
-The execution times give a good indication the poor scaling of the tool with program size: for larger test cases, the tool takes exponentially longer to compute, and for the largest tests the memory footprint is too large for even a well-equipped computer.
-The definitions of $\sSEMI{\aPSS_1}{\aPSS_2}$ and $\sIF{\aForm}\sTHEN\aPSS_1\sELSE\aPSS_2\sFI$ -- in combination with the rules for reads and writes -- yields a complexity class of $\mathcal{O}((|\Val| + 1)^{n})$ where $n$ is the length of the longest chain of sequential compositions in the program.
-Some further optimisations may be possible within the tool to improve the situation, such as killing 'dead-end' pomsets at each sequence operator, or by doing a directed search for particular execution outcomes.
 Unsurprisingly the execution time is dominated by the calculation of the denotation, with the additional axiomatic filtering step of \PwTc{} being within margin of error difference of just calculating the \PwT{} semantics.

### 2021-10-19 paper/tool.tex
Some better text about complexity without a bad guess at the precise complexity class.

@@
 
-The execution times give a good indication the poor scaling of the tool with program size: for larger test cases, the tool takes exponentially longer to compute, and in some cases simply fails.
+The execution times give a good indication the poor scaling of the tool with program size: for larger test cases, the tool takes exponentially longer to compute, and for the largest tests the memory footprint is too large for even a well-equipped computer.
 The compositional nature of the semantics makes tool building practical, but it is not enough to make it scalable for large tests.
+The definitions of $\sSEMI{\aPSS_1}{\aPSS_2}$ and $\sIF{\aForm}\sTHEN\aPSS_1\sELSE\aPSS_2\sFI$ -- in combination with the rules for reads and writes has exponential complexity.
+This is compounded by the hidden complexity of calculating the possible merges between pomsets through union in rules \ref{seq-E} and \ref{if-E}.
+Significant effort has been put into throwing away spurious merges early in \PwTer, so that executing the tool remains manageable for small examples.
+Some further optimisations may be possible within the tool to improve the situation further, such as killing 'dead-end' pomsets at each sequence operator, or by doing a directed search for particular execution outcomes.
 Unsurprisingly the execution time is dominated by the calculation of the denotation, with the additional axiomatic filtering step of \PwTc{} being within margin of error difference of just calculating the \PwT{} semantics.

### 2021-10-25 paper/abstract.tex
Integrating referee comments.

--- 2021-06-24 paper/abstract.tex
@@
 % \end{math}
-Program logics and semantics tell us that when executing $(S_1; S_2)$ starting in state $s_0$,
-we execute $S_1$ in $s_0$ to arrive at $s_1$, then execute $S_2$ in $s_1$ to
-arrive at the final state $s_2$.  This is, of course, an abstraction.  Processors
-execute instructions out of order, due to pipelines and caches, and compilers
-reorder programs even more dramatically.  All of this reordering is meant to
-be unobservable in single-threaded code, but is observable in multi-threaded code.
-A formal attempt to understand the resulting mess is known
-as a ``relaxed memory model.''  The relaxed memory models that have been
-proposed to date either fail to address sequential composition directly, or overly
-restrict processors and compilers.
+Program logics and semantics tell a pleasant story about sequential
+composition: when executing $(S_1; S_2)$, we first execute $S_1$ then $S_2$.
+To improve performance, however, processors execute instructions out of
+order, and compilers reorder programs even more dramatically.  By design,
+single-threaded systems cannot observe these reorderings; however,
+multiple-threaded systems can, making the story considerably less pleasant.
+A formal attempt to understand the resulting mess is known as a ``relaxed
+memory model.''  Prior models either fail to address sequential composition
+directly, or overly restrict processors and compilers, or permit nonsense
+thin-air behaviors which are unobservable in practice.
 
-To support sequential composition while targeting modern hardware, we propose
-using preconditions and families of predicate transformers.  When composing
-$(S_1;S_2)$, the predicate transformers used to validate the preconditions of
-events in $S_2$ are chosen based on the semantic dependencies from events in
-$S_1$ to events in $S_2$.  We apply this approach to two existing memory
-models: ``Modular Relaxed Dependencies'' for C11 and ``Pomsets with
-Preconditions.''
+To support sequential composition while targeting modern hardware, we enrich
+the standard event-based approach with \emph{preconditions} and
+\emph{families of predicate transformers}.
+%
+% When calculating the precondition of an event $e$ in $\sem{S_1;S_2}$, from
+% $\sem{S_2}$ is derived by applying a predicate transformer chosen based on
+% events of $\sem{S_1}$ upon which $e$ depends.
+When calculating $\sem{S_1;S_2}$, the predicate transformer applied to the
+precondition of an event $e$ from $\sem{S_2}$ is chosen based on the set of
+events in $\sem{S_1}$ upon which $e$ depends.  We apply this approach to two
+existing memory models.
+% : ``Modular Relaxed Dependencies'' for C11 and
+% ``Pomsets with Preconditions.''
 

### 2021-10-25 paper/intro.tex
Integrating referee comments.

@@
 \cite{DBLP:conf/snapl/MarinoMMNS15}.  When these approaches are not
-available, however, we are left with an enormous gap in our understanding of
-one of the most basic elements of computing: the humble semicolon.  Until
-recently, existing approaches either
+available, however, the humble semicolon becomes shrouded in mystery, covered
+in the cloak of something known as a \emph{memory model}.  Every language has
+such a model: For each read operation, it determines the set of available
+values.  Compilers and runtime systems are allowed to chose any value in the
+set.  To allow efficient implementation, the set must not be too small.  To
+allow invariant reasoning, the set must not be too large.
+
+For optimized concurrent languages, it is surprising difficult to define a
+model that allows common compiler optimizations and hardware reorderings yet
+disallows nonsense behaviors that don't arise in practice.  The latter are
+commonly known as ``thin air'' behaviors \cite{DBLP:conf/esop/BattyMNPS15}.
+There are only a handful of solutions, and all have deficiencies.  These can
+be classified by their approach to dependency tracking (from strongest to
+weakest):
 \begin{itemize}
-\item did not bother tracking dependencies, allowing ``thin air'' executions
-  --- as in C and C++~\cite{DBLP:conf/esop/BattyMNPS15},
-\item tracked dependencies conservatively, using syntax, requiring inefficient
-  implementation of relaxed access
-  \cite{DBLP:conf/pldi/LahavVKHD17,Boehm:2014:OGA:2618128.2618134,DBLP:conf/oopsla/VafeiadisN13,DBLP:journals/corr/abs-1804-04214}--- a non-starter for safe languages like Java,
-  and an unacceptable cost for low-level languages like C,
-\item computed dependencies using non-compositional operational models over
-  alternate worlds \cite{Manson:2005:JMM:1047659.1040336,
-    DBLP:conf/esop/JagadeesanPR10,
-    DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}---these
-  models validate many compiler optimizations, but
-  %allow some dubious executions 
-  fail to validate temporal safety properties
-  (see \textsection\ref{sec:promising}).
+\item Syntactic dependencies
+  \cite{DBLP:conf/pldi/LahavVKHD17,Boehm:2014:OGA:2618128.2618134,DBLP:conf/oopsla/VafeiadisN13,DBLP:journals/corr/abs-1804-04214}.
+  These models require inefficient implementation of relaxed access.  This is
+  a non-starter for safe languages like Java and Javascript, and may be an
+  unacceptable cost for low-level languages like \cXI{}.
+\item Semantic dependencies \cite{%
+    Manson:2005:JMM:1047659.1040336, DBLP:conf/esop/JagadeesanPR10,
+    DBLP:conf/popl/KangHLVD17, DBLP:journals/pacmpl/ChakrabortyV19,
+    DBLP:conf/pldi/LeeCPCHLV20, promising-ldrf}. These models compute
+  dependencies operationally using alternate worlds, making it impossible to
+  understood a single execution in isolation; they also allow executions that
+  violate temporal reasoning (see \textsection\ref{sec:related}).
+\item No dependencies, as in \cXI{} \cite{DBLP:conf/esop/BattyMNPS15} and
+  Javascript \cite{DBLP:journals/pacmpl/WattRP19}.  This allows thin air
+  executions.
 \end{itemize}
 
-Recently, two denotational models have been proposed that compute sequential
-dependencies semantically. \citet{DBLP:conf/esop/PaviottiCPWOB20} defined
-Modular Relaxed Dependencies (\MRD), which use event structures to calculate
-dependencies for \cXI, targeting the Intermediate Memory Model (\IMM)
-\cite{DBLP:journals/pacmpl/PodkopaevLV19}. \citet{DBLP:journals/pacmpl/JagadeesanJR20} defined
-Pomsets with Preconditions (\PwP), which use preconditions and logic to
-calculate dependencies for a Java-like language targeting {multicopy-atomic} (\mca{}) hardware, such as
-\armeight{} \cite{DBLP:journals/pacmpl/PulteFDFSS18}.  However, neither paper
-treated sequential composition as a first-class citizen.  \MRD{} encoded
-sequential composition using continuation-passing, and \PwP{} used prefixing,
-adding one event at a time on the left.  In both cases, adding an event
-requires perfect knowledge of the future.
+These models are all non-compositional in the sense that in order to
+calculate the meaning of any thread, all threads must be known.  Using the
+axiomatic approach of \cXI{}, for example, execution graphs are first
+constructed for each thread, using an operational semantics that allows a
+read to see any value.  The combined graphs are then filtered using a set of
+acyclicity axioms that determine which reads are valid.  These axioms use
+existentially defined global relations, such as memory order $\rmo$, which
+must be a per-location total order on write actions.
+
+Part of this non-compositionality is essential: In a concurrent system, the
+complete set of writes is known only at top-level.  However, much of it is
+incidental.  Two recent models have attempted to limit non-compositionality.
+\citet{DBLP:journals/pacmpl/JagadeesanJR20} defined Pomsets with
+Preconditions (\PwP), which use preconditions and logic to calculate
+dependencies for a Java-like language.
+\citet{DBLP:conf/esop/PaviottiCPWOB20} defined Modular Relaxed Dependencies
+(\MRD), which use event structures to calculate a semantic dependency
+relation ($\rsdep$).  \PwP{} is defined using (acyclic) labelled partial
+orders, or \emph{pomsets} \cite{GISCHER1988199}.  \MRD{} adds a causality
+axiom to \cXI{}, stating that $({\rsdep}\cup{\rrfx})$ must be acyclic.  In
+both approaches, acyclicity enables inductive reasoning.
+
+% While largely compositional, avoiding relations such as $\rmo$, \PwP{} still
+% requires some top-level filtering to ensure that every read has a matching
+% write---this seems unavoidable.  \MRD{} calculates $\rsdep$ compositionally,
+% but otherwise suffers from the non-compositionality of \cXI{}.
+
+While \PwP{} and \MRD{} both treat \emph{concurrency} compositionally, neither gives
+a compositional account of \emph{sequentiality}.  \PwP{} uses prefixing,
+adding one event at a time on the left.  \MRD{} encodes sequential
+composition using continuation-passing.  In both, adding an event requires
+perfect knowledge of the future.  For example, suppose that you are writing
+system call code and you wish to know if you can reorder a couple of
+statements.  Using \PwP{} or \MRD{}, you cannot tell whether this is possible
+without having the calling code!  More formally,
+\citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} state the equivalence
+allowing reordering independent writes as follows:
+\begin{align*}
+  \sem{\aLoc \GETS \aExp \SEMI \bLoc  \GETS \bExp\SEMI\aCmd} &=
+  \sem{\bLoc  \GETS \bExp\SEMI \aLoc \GETS \aExp\SEMI\aCmd} \;\;\text{if}\; \aLoc\neq\bLoc
+\end{align*}
+This requires a quantification over all continuations $\aCmd$. This is
+problematic, both from a theoretical point of view---the syntax of programs
+is now mentioned in the definition of the semantics---and in practice---tools
+cannot quantify over infinite sets. This problem is related to contextual
+equivalence, full abstraction
+\cite{DBLP:journals/tcs/Milner77,DBLP:journals/tcs/Plotkin77} and the CIU
+theorem of \citet{DBLP:conf/lics/MasonT92}.
 
@@
 $\aCmd_1$ and the denotation of $\aCmd_2$, and \emph{direct} in that these
-can be calculated independently.
+can be calculated independently.  With this formulation, we can show:
+\begin{align*}
+  \sem{\aLoc \GETS \aExp \SEMI \bLoc  \GETS \bExp} &=
+  \sem{\bLoc  \GETS \bExp\SEMI \aLoc \GETS \aExp} \;\;\text{if}\; \aLoc\neq\bLoc
+\end{align*}
+Then the equivalence holds in any context---this form of the equivalence
+enables reasoning about peephole optimizations.  Unlike prior work, \PwT{}
+allows the presence or absence of a dependency can be understood in
+isolation.  This enables incremental and modular validation of assumptions
+about program dependencies in larger blocks of code.
 
-The model has been formalized in Coq.  We have formally verified that the
+Our main insight is that for language models, \emph{sequentiality} is the
+hard part.  \emph{Concurrency} is easy!  Or at least, it is no more difficult
+than it is for hardware.  Compilers make the difference, since they typically
+do little optimization between threads.  We motivate our approach to
+sequential dependencies in \textsection\ref{sec:overview} and provide formal
+definitions in \textsection\ref{sec:model}.  We extend the model to include
+additional features, such as address calculation and \RMW{}s, in
+\textsection\ref{sec:additional}.  We further discuss related work in
+\textsection\ref{sec:related}.
+
+We extend \PwT{} to a full memory model in \textsection\ref{sec:mca}, based
+on \PwP{} \cite{DBLP:journals/pacmpl/JagadeesanJR20}.
+\textsection\ref{sec:results} summarizes the results for this model.  The
+dependency relation calculated by \PwT{} can also be used with off-the-shelf
+models.  For example, in \textsection\ref{sec:c11} we show that it can be
+used as an $\rsdep$ relation for \cXI, adapting the approach of \MRD{}
+\cite{DBLP:conf/esop/PaviottiCPWOB20}.  % This is only the second semantics
+% to distill dependencies down to a relation compatible with the existing
+% \cpp{} standard, and the first one to be fully compositional.
+\textsection\ref{sec:tool} describes a tool for automatic evaluation of
+litmus tests in this model.  \cXI{} allows thin-air in order to avoid
+overhead in the implementation of relaxed reads.  Safe language like OCaml
+\cite{Dolan:2018:BDR:3192366.3192421} have typically made the opposite choice.  Just
+\PwT{} can be used to strengthen \cXI{}, it could also be used to weaken these
+models, allowing optimal lowering for relaxed reads while banning thin-air.
+
+% The
+% definition is associative: the denotation of
+% $((\aCmd_1\SEMI \aCmd_2)\SEMI \aCmd_3)$ is the same as the denotation of
+% $(\aCmd_1\SEMI (\aCmd_2\SEMI \aCmd_3))$.  It also validates expected laws
+% concerning the interaction of sequencing and conditional execution.
+
+% To manage complexity, we have layered the definitions.  After an overview and
+% discussion of related work, we
+% define sequential dependencies in \textsection\ref{sec:model}.  We then add
+% concurrency.  In \textsection\ref{sec:mca}, we define \PwTmca{},
+% which provides a Java-like model for {multicopy-atomic} (\mca{}) hardware, similar to that of
+% \citet{DBLP:journals/pacmpl/JagadeesanJR20};
+% \textsection\ref{sec:results} summarizes the results for this model. In \textsection\ref{sec:c11},
+% we define \PwTc{}, which models \cXI, adapting the approach of
+% \citet{DBLP:conf/esop/PaviottiCPWOB20};  \textsection\ref{sec:tool} 
+% describes a tool for automatic evaluation of litmus tests.  In \textsection\ref{sec:additional},
+% we extend the semantics to include additional features, such as address
+% calculation and \RMW{}s.
+
+\PwT{} has been formalized in Coq.  We have formally verified that the
 sequential composition satisfies the expected monoid laws
@@
 
-% The
-% definition is associative: the denotation of
-% $((\aCmd_1\SEMI \aCmd_2)\SEMI \aCmd_3)$ is the same as the denotation of
-% $(\aCmd_1\SEMI (\aCmd_2\SEMI \aCmd_3))$.  It also validates expected laws
-% concerning the interaction of sequencing and conditional execution.
-
-To manage complexity, we have layered the definitions.  After an overview and
-discussion of related work, we
-define sequential dependencies in \textsection\ref{sec:model}.  We then add
-concurrency.  In \textsection\ref{sec:mca}, we define \PwTmca{},
-which provides a Java-like model for {multicopy-atomic} (\mca{}) hardware, similar to that of
-\citet{DBLP:journals/pacmpl/JagadeesanJR20};
-\textsection\ref{sec:results} summarizes the results for this model. In \textsection\ref{sec:c11},
-we define \PwTc{}, which models \cXI, adapting the approach of
-\citet{DBLP:conf/esop/PaviottiCPWOB20};  \textsection\ref{sec:tool} 
-describes a tool for automatic evaluation of litmus tests.  In \textsection\ref{sec:additional},
-we extend the semantics to include additional features, such as address
-calculation and \RMW{}s.
-
-% \todo{flesh this out}
 % \begin{itemize}
@@
 
-% % \todo{}
 % % We show how predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75} can
@@
 $\fwp{\aCmd}{\bForm}$ for loop-free code below (where $\aReg$--$\bReg$ range over thread-local \emph{registers}
-and $\aExp$--$\bExp$ range over side-effect-free \emph{expressions}). 
-%\begin{multicols}{2}
-  \begin{enumerate}[,label=(\textsc{d}\arabic*),ref=\textsc{d}\arabic*]
-  \item
-    \begin{math}
-      \fwp{\SKIP}{\bForm} = \bForm
-    \end{math}
-  \item \label{wp-let}
-    $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
+and $\aExp$--$\bExp$ range over side-effect-free
+\emph{expressions}). 
+% \begin{multicols}{3}
+%   \begin{enumerate}[,label=(\textsc{d}\arabic*),ref=\textsc{d}\arabic*]
+%   \item \label{wp-let}
+%     $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
+%   \item
+%     \makebox[0cm][l]{
+%     \begin{math}
+%       \fwp{\IF{\aExp}\THEN \aCmd_1\ELSE \aCmd_2\FI}{\bForm}= {}
+%     \end{math}
+%     \begin{math}
+%       ((\aExp{\ne}0) \limplies \fwp{\aCmd_1}{\bForm}) \land
+%       ((\aExp{=}0) \limplies \fwp{\aCmd_2}{\bForm})
+%     \end{math}}
+%     % \item 
+%     %   \begin{math}
+%     %     \fwp{\ABORT}{\bForm} = \FALSE
+%     %   \end{math}
+%     %   \stepcounter{enumi}
+%     % \item[] \labeltext[\textsc{d}2]{}{wp-assign}
+%     %   \begin{enumerate}[leftmargin=0pt]
+%     %   \item \label{wp-write}
+%     %     $\fwp{\PW{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
+%     %   \item \label{wp-let}
+%     %     $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
+%     %   \item \label{wp-read}
+%     %     $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \aLoc{=}\aReg\limplies\bForm$ %\bForm[\aLoc/\aReg]$ %
+%     %   \end{enumerate}
+%     % \item 
+%     %   \begin{math}
+%     %     \fwp{\LET{x}{\aExp}}{\bForm} = \bForm[\aExp/x]
+%     %   \end{math}
+%   \columnbreak
+%   \item
+%     \makebox[0cm][l]{
+%     \begin{math}
+%       \fwp{\aCmd_1;\aCmd_2}{\bForm} = \fwp{\aCmd_1}{\fwp{\aCmd_2}{\bForm}}
+%     \end{math}}
+%   \columnbreak
 %   \item
 %     \begin{math}
-    %     \fwp{\ABORT}{\bForm} = \FALSE
+%       \fwp{\SKIP}{\bForm} = \bForm
 %     \end{math}
-    %   \stepcounter{enumi}
-    % \item[] \labeltext[\textsc{d}2]{}{wp-assign}
-    %   \begin{enumerate}[leftmargin=0pt]
-    %   \item \label{wp-write}
-    %     $\fwp{\PW{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
-    %   \item \label{wp-let}
-    %     $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
-    %   \item \label{wp-read}
-    %     $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \aLoc{=}\aReg\limplies\bForm$ %\bForm[\aLoc/\aReg]$ %
 %   \end{enumerate}
-    % \item 
-    %   \begin{math}
-    %     \fwp{\LET{x}{\aExp}}{\bForm} = \bForm[\aExp/x]
-    %   \end{math}
-  \item
-    \begin{math}
+% \end{multicols}
+\begin{gather*}
+  \fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]
+  \qquad
   \fwp{\aCmd_1;\aCmd_2}{\bForm} = \fwp{\aCmd_1}{\fwp{\aCmd_2}{\bForm}}
-    \end{math}
-  \item
-    \begin{math}
-      \fwp{\IF{\aExp}\THEN \aCmd_1\ELSE \aCmd_2\FI}{\bForm}= {}
-    \end{math}
-    \begin{math}
+  \qquad
+  \fwp{\SKIP}{\bForm} = \bForm
+  \\
+  \fwp{\IF{\aExp}\THEN \aCmd_1\ELSE \aCmd_2\FI}{\bForm}= 
   ((\aExp{\ne}0) \limplies \fwp{\aCmd_1}{\bForm}) \land
   ((\aExp{=}0) \limplies \fwp{\aCmd_2}{\bForm})
-    \end{math}
-  \end{enumerate}
-%\end{multicols}
-For this language, the Hoare triple
+\end{gather*}    
+Without loops, the Hoare triple
 $\hoare{\aForm}{\aCmd}{\bForm}$ holds exactly when $\aForm \limplies
@@
 This is an elegant explanation of sequential computation in a sequential
-context. Note that \ref{wp-let} is sound because a read from a thread-local
+context. Note that the assignment rule %\ref{wp-let}
+is sound because a read from a thread-local
 register must be fulfilled by a preceding write in the same thread.
 In a concurrent context, with shared variables
-($\aLoc$--$\cLoc$), the obvious generalization
-%\begin{multicols}{2}
-\begin{itemize}
-\item[{\labeltext[\textsc{d}2a]{(\textsc{d}2a)}{wp-write}}] $\fwp{\LET{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
-\item[{\labeltext[\textsc{d}2b]{(\textsc{d}2b)}{wp-read}}] $\fwp{\LET{\aReg}{\aLoc}}{\bForm} = \bForm[\aLoc/\aReg]$
-\end{itemize}
-%\end{multicols}
-\noindent
-is unsound!  In particular, a read from a shared memory location may be fulfilled by a write
-in another thread, invalidating \ref{wp-read}.  (We assume that expressions
-% $\aExp$--$\bExp$
-do \emph{not} include shared variables.) % $\aLoc$--$\cLoc$.)
+($\aLoc$--$\cLoc$), the obvious generalization of the assignment rule
+% \eqref{wp-let}
+for reads,
+$\fwp{\LET{\aReg}{\aLoc}}{\bForm} = \bForm[\aLoc/\aReg]$,
+% \begin{multicols}{2}
+% \begin{itemize}
+% \item[{\labeltext[\textsc{d}1a]{(\textsc{d}1a)}{wp-write}}] $\fwp{\LET{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
+% \item[{\labeltext[\textsc{d}1b]{(\textsc{d}1b)}{wp-read}}] $\fwp{\LET{\aReg}{\aLoc}}{\bForm} = \bForm[\aLoc/\aReg]$
+% \end{itemize}
+% \end{multicols}
+% \allowbreak
+% \noindent
+is unsound!
+% (Expressions
+% % $\aExp$--$\bExp$
+% do \emph{not} include shared variables.) % $\aLoc$--$\cLoc$.)
+In particular, a read from a shared memory location may be fulfilled by a write
+in another thread.
 
@@
 We can represent such an execution as a labelled partial order, or
-\emph{pomset} \cite{GISCHER1988199}.  A program may give rise to many
+\emph{pomset} \cite{DBLP:conf/lop/Pratt85,GISCHER1988199}.  A program may give rise to many
 executions, each reflecting a different interleaving of the threads.
@@
 
-In such models, the causal order between events is important,
-and includes control and data dependencies ($\xpo$), to avoid
-paradoxical ``out of thin air'' examples such as:
+In such models, the causal order between events is important, and includes
+control and data dependencies ($\xpo$) to avoid paradoxical ``out of thin
+air'' examples such as the following.  (We routinely elide initializing
+writes when they are uninteresting.)
 \begin{gather*}
@@
 \end{gather*}
-In this diagram, $(s {<} 1)$ is a control dependency and $(r{*}s){=}0$ is a data dependency. Preconditions are discharged by being ordered after a read (we assume the
+In this diagram, $(s {<} 1)$ is a control dependency and $(r{*}s){=}0$ is a
+data dependency. Preconditions are updated as events are prepended (we assume the
 usual precedence for logical operators): 
@@
 \end{gather*}
-Note that there is dependency order from $(\DR{y}{0})$ to $(\DW{z}{0})$
-so the precondition for $(\DW{z}{0})$ only has to be satisfied assuming the hypothesis
-$(0{=}s)$. There is no matching order from $(\DR{x}{0})$ to $(\DW{z}{0})$
-which is why we do not assume the hypothesis $(0{=}r)$. Nonetheless, the precondition on
-$(\DW{z}{0})$ is a tautology, and so can be elided in the diagram:
+In this diagram there are two reads.  As evidenced by the arrow, the read of
+$y$ is ordered before the write, reflecting possible dependency; the read of
+$x$ is not, reflecting independency.  The dependent read of $y$ allows the
+precondition of the write to weaken: now the old precondition need only be
+satisfied assuming the hypothesis $(0{=}s)$.  The independent read of $x$
+allows no such weakening.  Nonetheless, the precondition of the write is now
+a tautology, and so can be elided in the diagram:
 \begin{gather*}
@@
 \end{gather*}
+In order for a pomset to be \emph{complete}, all preconditions must be
+tautologies.  Thus, an execution of \eqref{eq1} that reads $\DR{y}{2}$ must
+elide the write action to be considered complete.  For this reason, we must
+allow read and write to generate empty pomsets, in addition to singletons.
 
@@
 
-Pomsets with Preconditions show how the logical approach to sequential
-dependency calculation can be mixed into a relaxed memory model.  However,
-\citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} do not provide a model of
-sequential composition.  Instead, their model uses \emph{prefixing}, which
-requires that the model is built from right to left: events are prepended one
-at a time, with perfect knowledge of the future.  This makes reasoning about
-sequential program fragments difficult.  For example,
-\citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} state the equivalence
-allowing reordering independent writes as follows,
-\begin{align*}
-  \sem{\aLoc \GETS \aExp \SEMI \bLoc  \GETS \bExp\SEMI\aCmd} &=
-  \sem{\bLoc  \GETS \bExp\SEMI \aLoc \GETS \aExp\SEMI\aCmd} \;\;\text{if}\; \aLoc\neq\bLoc
-\end{align*}
-where $\aCmd$ is the entire future computation!  By formalizing sequential
-composition, we can show:
-\begin{align*}
-  \sem{\aLoc \GETS \aExp \SEMI \bLoc  \GETS \bExp} &=
-  \sem{\bLoc  \GETS \bExp\SEMI \aLoc \GETS \aExp} \;\;\text{if}\; \aLoc\neq\bLoc
-\end{align*}
-Then the equivalence holds in any
-%% (sequential)
-context.
+\PwP{} shows how the logical approach to sequential dependency calculation
+can be mixed into a relaxed memory model.  Our contribution is to extend
+\PwP{} with predicate transformers to arrive at a model of sequential
+composition.  Predicate transformers are a good fit for logical models of
+dependency calculation, since both are concerned with
+preconditions.% and how they are
+% transformed by sequential composition.
+
+% Instead, their model uses \emph{prefixing}, which
+% requires that the model is built from right to left: events are prepended one
+% at a time, with perfect knowledge of the future.  This makes reasoning about
+% sequential program fragments difficult.  For example,
+% \citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} state the equivalence
+% allowing reordering independent writes as follows,
+% \begin{align*}
+%   \sem{\aLoc \GETS \aExp \SEMI \bLoc  \GETS \bExp\SEMI\aCmd} &=
+%   \sem{\bLoc  \GETS \bExp\SEMI \aLoc \GETS \aExp\SEMI\aCmd} \;\;\text{if}\; \aLoc\neq\bLoc
+% \end{align*}
+% where $\aCmd$ is the entire future computation!  By formalizing sequential
+% composition, we can show:
+% \begin{align*}
+%   \sem{\aLoc \GETS \aExp \SEMI \bLoc  \GETS \bExp} &=
+%   \sem{\bLoc  \GETS \bExp\SEMI \aLoc \GETS \aExp} \;\;\text{if}\; \aLoc\neq\bLoc
+% \end{align*}
+% Then the equivalence holds in any
+% %% (sequential)
+% context.
 
@@
 
-Predicate transformers are a good fit for logical models of dependency
-calculation, since both are concerned with preconditions and how they are
-transformed by sequential composition. Our first attempt is to associate a
+Our first attempt is to associate a
 predicate transformer with each pomset. We visualize this in diagrams by
@@
 \end{align*}
-The predicate transformer from the write matches
-\citeauthor{DBLP:journals/cacm/Dijkstra75}'s \ref{wp-write}.  For the reads,
-however, \ref{wp-read} defines the transformer of $\PR{x}{r}$ to be
+The predicate transformer for a write $\PW{z}{\aExp}$ matches
+\citeauthor{DBLP:journals/cacm/Dijkstra75}: taking $\bForm$ to
+$\bForm[\aExp/z]$.  For a read $\PR{\aLoc}{\aReg}$, however,
+\citeauthor{DBLP:journals/cacm/Dijkstra75} would transform $\bForm$ to
 $\bForm[x/r]$, which is equivalent to $(x{=}r) \limplies \bForm$ under the
-assumption that registers are assigned at most once.
-Instead, we use $(0{=}r) \limplies \bForm$, reflecting the fact that $0$ may
-come from a concurrent write.  The obligation to find a matching write is
-moved from the sequential semantics of \emph{substitution} and
-\emph{implication} to the concurrent semantics of \emph{fulfillment}.
+assumption that registers are assigned at most once.  Instead, we use
+$(0{=}r) \limplies \bForm$, reflecting the fact that $0$ may come from a
+concurrent write.  The obligation to find a matching write is moved from the
+sequential semantics of \emph{substitution} and \emph{implication} to the
+concurrent semantics of \emph{fulfillment}.
 
@@
 This is an instance of the general result that sequential composition forms a monoid.
-
-\section{Related Work}
-
-\citet{DBLP:conf/snapl/MarinoMMNS15} argue that the ``silently shifting
-semicolon'' is sufficiently problematic for programmers that concurrent
-languages should guarantee sequential abstraction, despite the performance
-penalties (see also \cite{10.1145/3462206}).  In this paper, we take the opposite approach.  We have
-attempted to find the most intellectually tractable model that encompasses
-all of the messiness of relaxed memory.
-
-There are few prior studies of relaxed memory that include sequential
-composition and/or precise calculation of semantic dependencies.
-\citet{DBLP:journals/pacmpl/JagadeesanJR20} give a denotational semantics,
-using prefixing rather than sequential compositions.
-\citet{DBLP:conf/esop/PaviottiCPWOB20} give a denotational semantics,
-calculating dependencies using event structures rather than logic.  They give
-the semantics of sequential composition in continuation passing style,
-whereas we give it in direct style.  This paper provides a general technique
-for computing sequential dependencies and applies it to these two approaches.
-We provide a detailed comparison with
-\cite{DBLP:journals/pacmpl/JagadeesanJR20} in \textsection\ref{sec:diff}.
-
-\citet{DBLP:journals/corr/abs-1804-04214} define a semantics
-using pomsets without preconditions. Instead, their model uses syntactic
-dependencies, thus invalidating many compiler optimizations.  They also
-require a fence after every relaxed read on \armeight{}.
-%
-\citet{Pichon-Pharabod:2016:CSR:2837614.2837616} use event structures to
-calculate dependencies, combined with an operational semantics that
-incorporates program transformations.  This approach seems to require
-whole-program analysis.
-
-Other studies of relaxed memory can be categorized by their approach to
-dependency calculation.  Hardware models use syntactic dependencies
-\cite{alglave}.  Many software models do not bother with dependencies at all
-\cite{Batty:2011:MCC:1926385.1926394, DBLP:journals/pacmpl/WattRP19,
-  DBLP:conf/pldi/WattPPBDFPG20, goMM}.  Others have strong dependencies that
-disallow compiler optimizations and efficient implementation, typically
-requiring fences for every relaxed read on Arm
-\cite{Lamport:1979:MMC:1311099.1311750, DBLP:conf/pldi/LahavVKHD17,
-  Dolan:2018:BDR:3192366.3192421, DBLP:conf/lics/JeffreyR16,
-  Boehm:2014:OGA:2618128.2618134}. %, DBLP:journals/corr/abs-1804-04214}.
-%
-Many of the most prominent models are operational, whole-program models based
-on speculative execution \cite{Manson:2005:JMM:1047659.1040336,
-  DBLP:conf/esop/JagadeesanPR10,
-  DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}.
-We provide a detailed comparison with these approaches in \textsection\ref{sec:promising}.
-
-Other work in relaxed memory has shown that tooling is especially useful to researchers, architects, and language specifiers, enabling them to build intuitions experimentally~\cite{DBLP:conf/esop/PaviottiCPWOB20,PrideMM,alglave,Batty:2011:MCC:1926385.1926394}.
-Unfortunately, it is not obvious that tools can be built for all thin-air-free models, the calculation of \citet{Pichon-Pharabod:2016:CSR:2837614.2837616} does not have a termination proof for an arbitrary input, and the enormous state space for the operational models of \citet{DBLP:conf/popl/KangHLVD17} and \citet{DBLP:journals/pacmpl/ChakrabortyV19} is a daunting prospect for a tool builder -- and as yet no tool exists for automatically evaluating these models.
-We describe a tool, \PwTer, for automatically evaluating \PwT{} in \textsection\ref{sec:tool}.

### 2021-10-25 paper/mca1.tex
Updates
Integrating referee comments.

@@
 
-We derive a model of concurrent computation by adding
+In concurrent hardware, each processor has two views of each memory location:
+local and global.  In a \emph{multicopy-atomic} (\mca{}) architecture, there
+is only one global view, shared by all processors.  These two views are
+neatly reflected in \PwT{}: the local view corresponds to the preconditions.
+The global view corresponds to the partial order.
+
+In this section, we develop a model of concurrent computation by adding
 % \emph{parallel composition} and 
@@
 and synchronization, we add \emph{delay} to the rule for sequential
-composition.  For \emph{multicopy-atomic} (\mca{}) architectures, it is sufficient to encode delay in
+composition.  For  \mca{} architectures, it is sufficient to encode delay in
 the pomset order.  The resulting model, \PwTmca{1}, supports optimal lowering
@@
 same, except for \emph{internal reads}, where a thread reads its own write.
+We show an example at the beginning of \textsection\ref{sec:mca2}.
 
-The lowering proofs can be found in \textsection\ref{sec:arm}.  The proofs
+The lowering proofs can be found in \cite{appendix}.  The proofs
 use recent alternative characterizations of \armeight{} \cite{armed}.
@@
         \rf{wy1}{ry1}
-        \wk[out=170,in=10]{rx0}{wx1}
+        \wk[out=-170,in=-10]{rx0}{wx1}
         \wki{wx0}{wx1}
@@
 elimination, and stronger than Java, in order to support local reasoning
-about data races.  For further examples, see \textsection\ref{sec:extras} and
+about data races.  For further examples, see %\textsection\ref{sec:extras} and
 \cite[\textsection3.1]{DBLP:journals/pacmpl/JagadeesanJR20}.
@@
 discussion of item \eqref{lem:ifelse:if:if} %and \eqref{lem:ifelse:if:if2},
-see \textsection\ref{sec:delay}.
+see \cite{appendix}.
+%see \textsection\ref{sec:delay}.
 

### 2021-10-25 paper/outro.tex
Integrating referee comments.

@@
 \section{Conclusions} %Future Work}
+\label{sec:outro}
 
@@
 allowing events with different actions to merge. % in some instances.
+Nor does \PwTmca{} validate the reverse inclusions for
+\reflem{lem:if}\eqref{lem:ifelse:if:if}. The culprit is $\rdelay$, which
+introduces order regardless of whether preconditions are disjoint.  As an
+example, using augmentation, 
+\begin{math}
+  \sem{\IF{r}
+  \THEN \PW{x}{1}
+  \ELSE \PW{x}{2}
+  \FI}
+\end{math}
+has an execution with
+\begin{math}
+  (r{=}0\mid\DW{x}{2})
+  \xwki
+  (r{\neq}0\mid\DW{x}{1}),
+\end{math}
+whereas
+\begin{math}
+  \sem{
+    \IF{r} \THEN \PW{x}{1}\FI
+    \SEMI
+    \IF{\BANG r} \THEN \PW{x}{2}\FI
+  \FI}
+\end{math}
+has no such execution.  We expect that this can be remedied by encoding
+$\rdelay$ in the logic.
+%\todo{Comment on \PwTpo{} and \PwTc{}?}
 
@@
 compositional reasoning for temporal safety properties (see
-\textsection\ref{sec:promising}).  Nonetheless, read introduction is
+\textsection\ref{sec:related}).  Nonetheless, read introduction is
 ubiquitous in some compilers.  It would be interesting to know if there is a
@@
 
-% \subsection*{Acknowledgements}
-% % This paper has been greatly improved by the comments of the anonymous reviewers.
-% Riely was supported by the National Science Foundation under
-% grant No.~CCR-1617175.
+\subsection*{Acknowledgements}
+This paper has been greatly improved by the comments of the anonymous reviewers.
+
+Riely was supported by the National Science Foundation under
+grant No.~CCR-1617175.
+
+\todo{Add support for Mark\&Simon.}
+
+\todo{Add support for Anton\&Ilya.}
 

### 2021-10-25 paper/related.tex
Integrating referee comments.

\section{Related Work}
\label{sec:related}

\citet{DBLP:conf/snapl/MarinoMMNS15} argue that the ``silently shifting
semicolon'' is sufficiently problematic for programmers that concurrent
languages should guarantee sequential abstraction, despite the performance
penalties (see also \citet{10.1145/3462206}).  In this paper, we take the
opposite approach.  We have attempted to find the most intellectually
tractable model that encompasses all of the messiness of relaxed memory.

There are few prior studies of relaxed memory that include sequential
composition and/or precise calculation of semantic dependencies.
\citet{DBLP:journals/pacmpl/JagadeesanJR20} give a denotational semantics,
using prefixing rather than sequential compositions.
\citet{DBLP:conf/esop/PaviottiCPWOB20} give a denotational semantics,
calculating dependencies using event structures rather than logic.  They give
the semantics of sequential composition in continuation passing style,
whereas we give it in direct style.  This paper provides a general technique
for computing sequential dependencies and applies it to these two approaches.

\todo{Text about PwP.}

\todo{By the way, I note that PwP doesn't use these termination conditions,
  and there's no explanation in the paper of why this part of the model has
  changed, even in the appendix -- could you add it?}

% We provide a detailed comparison with
% \cite{DBLP:journals/pacmpl/JagadeesanJR20} in \textsection\ref{sec:diff}.

\citet{DBLP:journals/corr/abs-1804-04214} define a semantics
using pomsets without preconditions. Instead, their model uses syntactic
dependencies, thus invalidating many compiler optimizations.  They also
require a fence after every relaxed read on \armeight{}.
%
\citet{Pichon-Pharabod:2016:CSR:2837614.2837616} use event structures to
calculate dependencies, combined with an operational semantics that
incorporates program transformations.  This approach seems to require
whole-program analysis.

Other studies of relaxed memory can be categorized by their approach to
dependency calculation.  Hardware models use syntactic dependencies
\cite{alglave}.  Many software models do not bother with dependencies at all
\cite{Batty:2011:MCC:1926385.1926394, DBLP:journals/pacmpl/WattRP19,
  DBLP:conf/pldi/WattPPBDFPG20, goMM}.  Others have strong dependencies that
disallow compiler optimizations and efficient implementation, typically
requiring fences for every relaxed read on Arm
\cite{Lamport:1979:MMC:1311099.1311750, DBLP:conf/pldi/LahavVKHD17,
  Dolan:2018:BDR:3192366.3192421, DBLP:conf/lics/JeffreyR16,
  Boehm:2014:OGA:2618128.2618134}. %, DBLP:journals/corr/abs-1804-04214}.
%
Many of the most prominent models are operational, whole-program models based
on speculative execution \cite{Manson:2005:JMM:1047659.1040336,
  DBLP:conf/esop/JagadeesanPR10,
  DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}.
% We provide a detailed comparison with these approaches in \textsection\ref{sec:promising}.

\todo{We agree that a new memory model needs to be positioned against existing
models.  The usual result here is a compilation correctness to hardware
memory models.  For PwT-MCA, we address this by showing compilation result
for Armv8 model (§5).
%
Comparing software models, however, is unsatisfying: they are all
incomparable, i.e., there are examples which are allowed by one/disallowed by
the other and vice versa.
%
Morally, our model sits between the strong models (exemplified by RC11
[Lahav-al:PLDI17]) and the speculative models (exemplified by the promising
semantics [Kang-al:POPL17]).  As we argue in §1:
%
- The strong models require too much synchronization.
%
- The speculative models allow thin air behaviors.
%
Looking at the details, however, PwT-MCA is incomparable to both RC11 and
promising semantics.  RC11 allows non-MCA behaviors that PwT-MCA disallows.
PwT-MCA has a weaker notion of coherence than the promising semantics.
%
Some differences reflect an attempt to fix a bug.  For example, Weakestmo
[Chakraborty-Vafeiadis:POPL19] purposefully disallows thin-air-like
behaviors of the promising semantics.
%
Other differences reflect a different balance between allowed optimizations
and reasoning principles.  There are fundamental conflicts, for example:
%
- between Common Subexpression Elimination (CSE) and read-read coherence
  (§D.1)
%
- between if-introduction (aka, case analysis, if-closure) and java-style
  final-field semantics.  (If-introduction requires that address dependencies
  and control dependencies are the same.  Final-fields require that they be
  different.)
}
\todo{Text about promising.}

Other work in relaxed memory has shown that tooling is especially useful to
researchers, architects, and language specifiers, enabling them to build
intuitions
experimentally~\cite{DBLP:conf/esop/PaviottiCPWOB20,PrideMM,alglave,Batty:2011:MCC:1926385.1926394}.
Unfortunately, it is not obvious that tools can be built for all
thin-air-free models, the calculation of
\citet{Pichon-Pharabod:2016:CSR:2837614.2837616} does not have a termination
proof for an arbitrary input, and the enormous state space for the
operational models of \citet{DBLP:conf/popl/KangHLVD17} and
\citet{DBLP:journals/pacmpl/ChakrabortyV19} is a daunting prospect for a tool
builder -- and as yet no tool exists for automatically evaluating these
models.  We described a tool, \PwTer, for automatically evaluating \PwT{} in
\textsection\ref{sec:tool}.

### 2021-10-25 paper/results.tex
AEC draft
Integrating referee comments.

@@
 
-\PwP{} \cite{DBLP:journals/pacmpl/JagadeesanJR20} is a novel memory model,
+\PwP{} %\cite{DBLP:journals/pacmpl/JagadeesanJR20}
+is a novel memory model,
 intended to serve as a semantic basis for a Java-like language, where all
@@
 significant changes.  As a result, we have had to re-prove most of the
-theorems from \PwP{}.
+theorems from \PwP{}.  The proofs can be found in \cite{appendix}.
 
-In \textsection\ref{sec:arm}, we show that \PwTmca{1} supports the optimal
+% In \textsection\ref{sec:arm},
+We show that \PwTmca{1} supports the optimal
 lowering of relaxed accesses to \armeight{} and that \PwTmca{2} supports the
@@
 
-In \textsection\ref{sec:sc}, we prove sequential consistency for
+%In \textsection\ref{sec:sc},
+We prove sequential consistency for
 local-data-race-free programs.  The proof uses \emph{program order}, which we
@@
 can break temporal reasoning.  This combination is allowed by speculative
-operational models.  See \textsection\ref{sec:promising} for a discussion.
+operational models.  See \textsection\ref{sec:related} for a discussion.
 
@@
 past-time temporal logic.  The principal is based entirely on the pomset
-order relation.  It's proof, and all of the no-thin-air examples in
+order relation.  Its proof, and all of the no-thin-air examples in
 \cite[\textsection6]{DBLP:journals/pacmpl/JagadeesanJR20} hold equally for

### 2021-10-26 paper/mca1.tex
removing todos

@@
 % \emph{parallel composition} and 
-\emph{reads-from} to \reffig{fig:seq}.  To model coherence
-and synchronization, we add \emph{delay} to the rule for sequential
-composition.  For  \mca{} architectures, it is sufficient to encode delay in
-the pomset order.  The resulting model, \PwTmca{1}, supports optimal lowering
-for relaxed access on \armeight{}, but requires extra synchronization for
-acquiring reads.  (\emph{Lowering} is the translation of language-level
-operators to machine instructions.  A lowering is \emph{optimal} if it
-provides the most efficient execution possible.)
+\emph{reads-from} to \reffig{fig:seq}.  To model coherence and
+synchronization, we add \emph{delay} to the rule for sequential composition.
+For \mca{} architectures, it is sufficient to encode delay in the pomset
+order.  The resulting model, \PwTmca{1}, supports optimal lowering for
+relaxed access on \armeight{}, but requires extra synchronization for
+acquiring reads.\footnote{\citet{DBLP:journals/pacmpl/JagadeesanJR20}
+  erroneously elide the required synchronization on acquiring reads.}
+(\emph{Lowering} is the translation of language-level operators to machine
+instructions.  A lowering is \emph{optimal} if it provides the most efficient
+execution possible.)
 
@@
 
-Lemmas \ref{lem:monoid} and \ref{lem:if} hold for \PwTmca{1}.  For further
-discussion of item \eqref{lem:ifelse:if:if} %and \eqref{lem:ifelse:if:if2},
-see \cite{appendix}.
-%see \textsection\ref{sec:delay}.
+Lemmas \ref{lem:monoid} and \ref{lem:if} hold for \PwTmca{1}.  We discuss
+\ref{lem:if}\ref{lem:ifelse:if:if} %and \eqref{lem:ifelse:if:if2},
+further in \textsection\ref{sec:outro}.
 

### 2021-10-26 paper/oopsla.tex
Updates
removing todos

@@
       \po{b1}{b2}
-      \rf[out=169,in=11]{b2}{a2}
-      \rf[out=169,in=11]{b2}{a1}
+      \rf[out=-169,in=-11]{b2}{a2}
+      \rf[out=-169,in=-11]{b2}{a1}
     \end{tikzinline}}
@@
 
-\endinput
-
-\myparagraph{Parallel Composition}
+\myparagraph{Termination Conditions and Parallel Composition}
 
@@
 termination actions.  The predicates should be joined using $\land$, not
-$\lor$.
+$\lor$.  Here we have used termination conditions rather than termination
+actions so that termination is handled separately.
 
@@
 \myparagraph{Augmentation of Preconditions}
-\jjr{} allows augmentation of preconditions.
-As discussed in \textsection\ref{sec:delay}, this causes associativity to fail
-for $\rdelay$, at least when attempting to validate \reflem{lem:if}\eqref{lem:ifelse:if:if1}--\eqref{lem:ifelse:if:if2}
-Thus, we use \emph{weakest} preconditions, rather than general preconditions.
-As a result, we fail to validate the following
-refinement:
-\begin{math}
-  \aPSS_1
-  \not\supseteq
-  \xIFTHEN{\aForm}{\aPSS_1}{}.
-\end{math}
+\jjr{} allows arbitrary augmentation of preconditions.  Here we are more
+conservative, only allowing augmentation of preconditions in the semantics of
+primitive actions, as in \textsection\ref{sec:semca}.
+As discussed in \textsection\ref{sec:delay}, allowing arbitrary augmentation
+causes associativity to fail when encoding $\rdelay$ logically. 
+% Thus, we use \emph{weakest} preconditions, rather than general preconditions.
+% As a result, we fail to validate the following
+% refinement:
+% \begin{math}
+%   \aPSS_1
+%   \not\supseteq
+%   \xIFTHEN{\aForm}{\aPSS_1}{}.
+% \end{math}
 

### 2021-10-26 paper/outro.tex
removing todos

@@
 % In a similar vein,
-\PwT{} does not validate read introduction, whereas speculative operational
-semantics do.  Recent work shows a tension between read introduction and
-compositional reasoning for temporal safety properties (see
-\textsection\ref{sec:related}).  Nonetheless, read introduction is
-ubiquitous in some compilers.  It would be interesting to know if there is a
-performance penalty for banning read introduction.
+% \PwT{} does not validate read introduction, whereas speculative operational
+% semantics do.  Recent work shows a tension between read introduction and
+% compositional reasoning for temporal safety properties (see
+% \textsection\ref{sec:related}).  Nonetheless, read introduction is
+% ubiquitous in some compilers.  It would be interesting to know if there is a
+% performance penalty for banning read introduction.
 

### 2021-10-26 paper/related.tex
removing todos

@@
 
-\todo{Text about PwP.}
-
-\todo{By the way, I note that PwP doesn't use these termination conditions,
-  and there's no explanation in the paper of why this part of the model has
-  changed, even in the appendix -- could you add it?}
-
-% We provide a detailed comparison with
-% \cite{DBLP:journals/pacmpl/JagadeesanJR20} in \textsection\ref{sec:diff}.
+There are many differences between \PwP{}
+\cite{DBLP:journals/pacmpl/JagadeesanJR20} and \PwT{}.  Two of these are
+motivated by requirements unique to \PwT{}: First, \PwP{} uses substitution
+to define read actions; \PwT{} Skolemizes in order to satisfy disjunction
+closure for predicate transformers (\textsection\ref{sec:ex:assoc}).  Second,
+\PwP{} requires \emph{consistency} and \emph{causal strengthening}; \PwT{}
+drops these in order to satisfy associativity (\reflem{lem:monoid}).  Other
+differences are stylistic: For example, we replaced the termination
+\emph{actions} of \PwP{} by termination \emph{conditions}---our formulation
+fixes an error in \citeauthor{DBLP:journals/pacmpl/JagadeesanJR20}'s
+definition of parallel composition.  We also fix an error in their treatment
+of redundant read elimination.
 
@@
 %
-Many of the most prominent models are operational, whole-program models based
+Many of the most prominent models are operational models based
 on speculative execution \cite{Manson:2005:JMM:1047659.1040336,
@@
   DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}.
+Morally, \PwT{} fits between the strong models and the speculative ones.
 % We provide a detailed comparison with these approaches in \textsection\ref{sec:promising}.
 
-\todo{We agree that a new memory model needs to be positioned against existing
-models.  The usual result here is a compilation correctness to hardware
-memory models.  For PwT-MCA, we address this by showing compilation result
-for Armv8 model (§5).
-%
-Comparing software models, however, is unsatisfying: they are all
-incomparable, i.e., there are examples which are allowed by one/disallowed by
-the other and vice versa.
-%
-Morally, our model sits between the strong models (exemplified by RC11
-[Lahav-al:PLDI17]) and the speculative models (exemplified by the promising
-semantics [Kang-al:POPL17]).  As we argue in §1:
-%
-- The strong models require too much synchronization.
-%
-- The speculative models allow thin air behaviors.
-%
-Looking at the details, however, PwT-MCA is incomparable to both RC11 and
-promising semantics.  RC11 allows non-MCA behaviors that PwT-MCA disallows.
-PwT-MCA has a weaker notion of coherence than the promising semantics.
-%
-Some differences reflect an attempt to fix a bug.  For example, Weakestmo
-[Chakraborty-Vafeiadis:POPL19] purposefully disallows thin-air-like
-behaviors of the promising semantics.
-%
-Other differences reflect a different balance between allowed optimizations
-and reasoning principles.  There are fundamental conflicts, for example:
-%
-- between Common Subexpression Elimination (CSE) and read-read coherence
-  (§D.1)
-%
-- between if-introduction (aka, case analysis, if-closure) and java-style
-  final-field semantics.  (If-introduction requires that address dependencies
-  and control dependencies are the same.  Final-fields require that they be
-  different.)
-}
-\todo{Text about promising.}
+% \todo{We agree that a new memory model needs to be positioned against existing
+% models.  The usual result here is a compilation correctness to hardware
+% memory models.  For PwT-MCA, we address this by showing compilation result
+% for Armv8 model (§5).
+% %
+% Comparing software models, however, is unsatisfying: they are all
+% incomparable, i.e., there are examples which are allowed by one/disallowed by
+% the other and vice versa.
+% %
+% Morally, our model sits between the strong models (exemplified by RC11
+% [Lahav-al:PLDI17]) and the speculative models (exemplified by the promising
+% semantics [Kang-al:POPL17]).  As we argue in §1:
+% %
+% - The strong models require too much synchronization.
+% %
+% - The speculative models allow thin air behaviors.
+% %
+% Looking at the details, however, PwT-MCA is incomparable to both RC11 and
+% promising semantics.  RC11 allows non-MCA behaviors that PwT-MCA disallows.
+% PwT-MCA has a weaker notion of coherence than the promising semantics.
+% %
+% Some differences reflect an attempt to fix a bug.  For example, Weakestmo
+% [Chakraborty-Vafeiadis:POPL19] purposefully disallows thin-air-like
+% behaviors of the promising semantics.
+% %
+% Other differences reflect a different balance between allowed optimizations
+% and reasoning principles.  There are fundamental conflicts, for example:
+% %
+% - between Common Subexpression Elimination (CSE) and read-read coherence
+%   (§D.1)
+% %
+% - between if-introduction (aka, case analysis, if-closure) and java-style
+%   final-field semantics.  (If-introduction requires that address dependencies
+%   and control dependencies are the same.  Final-fields require that they be
+%   different.)
+% }
+\citet{DBLP:journals/pacmpl/JagadeesanJR20} argue that the speculative models
+allow too many executions, resulting in a failure of temporal reasoning and
+potentially jeopardizing type safety and other security properties.  In a
+similar vein, \citet{promising-ldrf} argue that local \drf{} guarantees are
+violated when read-introduction is followed by if-introduction, branching on
+the read just introduced.  These optimizations are validated by the
+speculative models---\citeauthor{promising-ldrf} manage to avoid the problem
+by adopting a sub-optimal lowering for \RMW{}s.  \PwT{} does not suffer from
+this problem, since \PwT{} does not validate read-introduction.  Nonetheless,
+read introduction is ubiquitous in some compilers
+\cite{DBLP:conf/pldi/LeeKSHDMRL17}.  There appears to be a genuine tension
+between temporal reasoning, as supported by \PwT{}, and read-introduction, as
+supported by the speculative models.
 

### 2021-10-27 paper/tool.tex
Integrating referee comments.
update tool runtimes, add MRD

@@
 Unsurprisingly the execution time is dominated by the calculation of the denotation, with the additional axiomatic filtering step of \PwTc{} being within margin of error difference of just calculating the \PwT{} semantics.
+For context, we have included the results of \MRD{} for the Java Causality tests~\cite{DBLP:conf/esop/PaviottiCPWOB20}.
 % \PwTer{} will be made open source upon publication.
 \PwTer{} is available online at \url{https://github.com/graymalkin/pomsets-with-predicate-transformers}.
+
+\todo{Remove link? --- perhaps we put in the appendices.}
+
+\todo{Verify that referee comments sufficiently addressed regarding ``scalability challenges''.  Perhaps mention that it is easier for hardware models?}
+
+\todo{Shrink by using two columns and dump \PwT{}?}
 
@@
 \begin{center}
-  \begin{tabularx}{0.7\textwidth}{c||c|c|c}
-    \multirow{2}{*}{\bf Causality Test}   & \multicolumn{2}{c|}{\PwTc{}}    & \PwT{}\\
-    \cline{2-4}
-                                     & Result & Execution Time (s)  & Execution Time (s) \\
-    \cline{1-4}
-    jctc1                            & pass   & 2.397               & 2.608              \\
-    jctc2                            & pass   & 25.780              & 25.754             \\
-    jctc3                            & pass   & 196.935             & 205.120            \\
-    jctc4                            & pass   & 2.269               & 2.110              \\
-    jctc5                            & pass   & 63.714              & 69.441             \\
-    jctc6                            & pass   & 11.245              & 12.489             \\
-    jctc7                            & pass   & 88.250              & 96.099             \\
-    jctc8                            & pass   & 2.482               & 2.473              \\
-    jctc9                            & pass   & 13.592              & 15.384             \\
-    jctc10                           & pass   & 494.133             & 513.234            \\
-    jctc11                           & $\bot$ & --                  & --                 \\
-    jctc12                           & $\bot$ & --                  & --                 \\
-    jctc13                           & pass   & 2.101               & 2.247              \\
-    jctc17                           & pass   & 178.304             & 186.228            \\
-    jctc18                           & pass   & 177.292             & 2.247              \\
+  \begin{tabularx}{0.8\textwidth}{c||c|c|c|c|c|c}
+    \multirow{2}{*}{\bf Causality Test}   & \multicolumn{2}{c|}{\PwTc{}}    & \PwT{}     & \multicolumn{3}{c}{\MRD}  \\
+    \cline{2-7}
+                                     & Result & Time (s)  &  Time (s) & \MRD   & \MRDIMM & \MRDRC  \\
+    \cline{1-7}
+    jctc1                            & pass   & 2.181     & 2.112     & pass    & pass    & pass   \\
+    jctc2                            & pass   & 25.605    & 21.404    & pass    & pass    & pass   \\
+    jctc3                            & pass   & 181.052   & 185.913   & pass    & pass    & pass   \\
+    jctc4                            & pass   & 1.943     & 1.906     & pass    & pass    & pass   \\
+    jctc5                            & pass   & 57.599    & 57.352    & pass    & pass    & pass   \\
+    jctc6                            & pass   & 10.297    & 10.271    & pass    & pass    & pass   \\
+    jctc7                            & pass   & 79.916    & 85.777    & pass    & pass    & pass   \\
+    jctc8                            & pass   & 2.211     & 2.171     & pass    & pass    & pass   \\
+    jctc9                            & pass   & 14.638    & 13.167    & pass    & pass    & pass   \\
+    jctc10                           & pass   & 538.614   & 526.774   & pass    & pass    & pass   \\
+    jctc11                           & $\bot$ & --        & --        & pass    & pass    & pass   \\
+    jctc12                           & $\bot$ & --        & --        & --      & --      & --     \\
+    jctc13                           & pass   & 2.225     & 2.154     & pass    & pass    & pass   \\
+    jctc17                           & pass   & 188.318   & 200.829   & fail    & pass    & fail   \\
+    jctc18                           & pass   & 182.788   & 170.625   & fail    & pass    & fail   \\
   \end{tabularx}
@@
 $\bot$ indicates the tool failed to run for this test due to a memory overflow.
-Tests run on an Intel i9-9980HK with 64 GB of memory, execution times are the mean of 3 runs.}
+Tests run on an Intel i9-9980HK with 64 GB of memory, execution times are the mean of 3 runs.
+For context, results for the various flavours of \MRD{} are included~\cite{DBLP:conf/esop/PaviottiCPWOB20}.}
 \end{center}

### 2021-10-27 paper/intro.tex
done with section 2

@@
   \\[-.4ex]
-  \nonumber
   \hbox{\begin{tikzinline}[node distance=1.5em]
@@
 \begin{gather*}
-  \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
+  \IF{s{>}0} \THEN \PW{z}{r{*}(s{-}1)} \FI
   \\
-  \nonumber
   \hbox{\begin{tikzinline}[node distance=1.5em]
-      \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \bigmid \DW{z}{0}}{}
+      \event{wz0}{(s {>} 0) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{}
     \end{tikzinline}}
 \end{gather*}
-In this diagram, $(s {<} 1)$ is a control dependency and $(r{*}s){=}0$ is a
+In this diagram, $(s {>} 0)$ is a control dependency and $(r{*}(s{-}1)){=}0$ is a
 data dependency. Preconditions are updated as events are prepended (we assume the
@@
 \begin{gather*}
-  \tag{$\dagger$}\label{eq1}
-  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
+  %\tag{$\dagger$}\label{eq1}
+  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{>}0} \THEN \PW{z}{r{*}(s{-}1)} \FI
   \\
-  \nonumber
   \hbox{\begin{tikzinline}[node distance=1.5em]
-      \event{rx0}{\DR{x}{0}}{}
-      \event{ry0}{\DR{y}{0}}{right=of rx0}
-      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \bigmid \DW{z}{0}}{right=of ry0}
+      \event{rx0}{\DR{x}{1}}{}
+      \event{ry0}{\DR{y}{1}}{right=of rx0}
+      \event{wz0}{(1{=}s) \limplies (s{>}0) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
       \po{ry0}{wz0}
@@
 precondition of the write to weaken: now the old precondition need only be
-satisfied assuming the hypothesis $(0{=}s)$.  The independent read of $x$
+satisfied assuming the hypothesis $(1{=}s)$.  The independent read of $x$
 allows no such weakening.  Nonetheless, the precondition of the write is now
-a tautology, and so can be elided in the diagram:
+a tautology, and so can be elided in the diagram.
+
+We can complete the execution by adding the required writes:
 \begin{gather*}
-  % \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
-  % \\
-  \nonumber
+  \PW{x}{1}
+  \SEMI
+  \PW{y}{1}
+  \PAR
+  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{>}0} \THEN \PW{z}{r{*}(s{-}1)} \FI
+  \\
   \hbox{\begin{tikzinline}[node distance=1.5em]
-      \event{rx0}{\DR{x}{0}}{}
-      \event{ry0}{\DR{y}{0}}{right=of rx0}
-      \event{wz0}{\DW{z}{0}}{right=of ry0}
-      \po{ry0}{wz0}
+      \event{wx}{\DW{x}{1}}{}
+      \event{wy}{\DW{y}{1}}{right=of wx}
+      \event{rx}{\DR{x}{1}}{right=3em of wy}
+      \event{ry}{\DR{y}{1}}{right=of rx}
+      \event{wz}{\DW{z}{0}}{right=of ry}
+      \po{ry}{wz}
+      \rf[out=10,in=170]{wx}{rx}
+      \rf[out=10,in=170]{wy}{ry}
     \end{tikzinline}}
 \end{gather*}
-In order for a pomset to be \emph{complete}, all preconditions must be
-tautologies.  Thus, an execution of \eqref{eq1} that reads $\DR{y}{2}$ must
-elide the write action to be considered complete.  For this reason, we must
-allow read and write to generate empty pomsets, in addition to singletons.
+In order for a \PwP{} to be \emph{complete}, all preconditions must be
+tautologies and all reads must be fulfilled by matching writes.  The first
+requirement captures the sequential semantics.  The second requirement
+captures the concurrent semantics.  These correspond to two views of memory
+for each thread: thread-local and global.  In a
+\emph{multicopy-atomic} (\mca{}) architecture, there is only one global view,
+shared by all processors, which is neatly captured by the order of the
+pomset.
+
+An untaken conditional produces no events.  \PwP{} models this by including
+the empty pomset in the semantics of every program fragment.  To then ensure
+that $\SKIP$ is not a refinement of $\PW{x}{1}$, \PwP{} include the
+\emph{termination} action, $\aTerm{}$, which we have elided in the examples
+above.
 
@@
 %   \\
-%   \nonumber
 %   \hbox{\begin{tikzinline}[node distance=2ex and 1.5em]
@@
   \\
-  \nonumber
   \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]

### 2021-10-27 paper/mca1.tex
done with section 2

@@
 \label{sec:mca}
-
-In concurrent hardware, each processor has two views of each memory location:
-local and global.  In a \emph{multicopy-atomic} (\mca{}) architecture, there
-is only one global view, shared by all processors.  These two views are
-neatly reflected in \PwT{}: the local view corresponds to the preconditions.
-The global view corresponds to the partial order.
 

### 2021-10-27 paper/related.tex
removing todos
updated related wokr

@@
 There are few prior studies of relaxed memory that include sequential
-composition and/or precise calculation of semantic dependencies.
-\citet{DBLP:journals/pacmpl/JagadeesanJR20} give a denotational semantics,
-using prefixing rather than sequential compositions.
-\citet{DBLP:conf/esop/PaviottiCPWOB20} give a denotational semantics,
-calculating dependencies using event structures rather than logic.  They give
-the semantics of sequential composition in continuation passing style,
-whereas we give it in direct style.  This paper provides a general technique
-for computing sequential dependencies and applies it to these two approaches.
-
-There are many differences between \PwP{}
-\cite{DBLP:journals/pacmpl/JagadeesanJR20} and \PwT{}.  Two of these are
-motivated by requirements unique to \PwT{}: First, \PwP{} uses substitution
-to define read actions; \PwT{} Skolemizes in order to satisfy disjunction
-closure for predicate transformers (\textsection\ref{sec:ex:assoc}).  Second,
-\PwP{} requires \emph{consistency} and \emph{causal strengthening}; \PwT{}
-drops these in order to satisfy associativity (\reflem{lem:monoid}).  Other
-differences are stylistic: For example, we replaced the termination
-\emph{actions} of \PwP{} by termination \emph{conditions}---our formulation
-fixes an error in \citeauthor{DBLP:journals/pacmpl/JagadeesanJR20}'s
-definition of parallel composition.  We also fix an error in their treatment
-of redundant read elimination.
+composition and/or precise calculation of semantic dependencies.  No prior
+work gives the semantics of sequential composition in direct style.
+\citet{DBLP:conf/esop/PaviottiCPWOB20} defined \MRD{}, which calculates
+dependencies using event structures rather than logic.  This strategy is
+brittler than ours, leading to false positives (\textsection\ref{sec:lir}).
+\citet{DBLP:journals/pacmpl/JagadeesanJR20} defined \PwP{}, using logical
+entailment to define dependency.  Although \PwT{} is based on \PwP{}, there
+are many differences.  Two of these are motivated by requirements unique to
+\PwT{}: First, \PwP{} uses substitution to define read actions; \PwT{}
+Skolemizes in order to satisfy disjunction closure for predicate transformers
+(\textsection\ref{sec:ex:assoc}).  Second, \PwP{} requires \emph{consistency}
+and \emph{causal strengthening}; \PwT{} drops these in order to satisfy
+associativity (\reflem{lem:monoid}).  Other differences are stylistic: For
+example, we use termination \emph{conditions} rather than termination
+\emph{actions}---our formulation fixes an error in
+\citeauthor{DBLP:journals/pacmpl/JagadeesanJR20}'s definition of parallel
+composition.  We also fix an error in their treatment of redundant read
+elimination (\textsection\ref{sec:semreg}).
 
@@
 %
-Many of the most prominent models are operational models based
-on speculative execution \cite{Manson:2005:JMM:1047659.1040336,
+Many of the most prominent models are operational models based on speculative
+execution \cite{Manson:2005:JMM:1047659.1040336,
   DBLP:conf/esop/JagadeesanPR10,
   DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}.
+
 Morally, \PwT{} fits between the strong models and the speculative ones.
+Looking at the details, however, \PwTmca{} is incomparable to both \rcXI{}
+\cite{DBLP:conf/pldi/LahavVKHD17} and the promising semantics
+\cite{DBLP:conf/popl/KangHLVD17}, to take two examples.  \rcXI{} allows
+non-\mca{} behaviors that \PwTmca{} disallows.  \PwTmca{} has a weaker notion
+of coherence than the promising semantics.
+
 % We provide a detailed comparison with these approaches in \textsection\ref{sec:promising}.

### 2021-10-27 paper/mca2.tex
Updates
removing todos
messing in 4.2

@@
 
-Lowering \PwTmcaTITLE{1} to \armeight{} requires a full fence after every
+Lowering \PwTmca{1} to \armeight{} requires a full fence after every
 acquiring read.  To see why, consider the following attempted
@@
 \begin{gather*}
+  \taglabel{internal-acq}
+  \begin{gathered}
   \PW{x}{2}\SEMI 
@@
       \sync{d}{e}
-      \wk[out=-165,in=-15]{e}{a}
+      \wk[out=-170,in=-10]{e}{a}
     \end{tikzinline}}
+\end{gathered}
 \end{gather*}
@@
     if $\bEv\in\aEvs_1$, $\aEv\in\aEvs_2$ and $\aEv\xrfx\bEv$ then $\aEv\lt\bEv$,
-  \item[{\labeltext[\textsc{s}6a]{(\textsc{s}6a)}{seq-le-delays-rf}}]
+  \item[{\labeltext[\textsc{s}6a$'$]{(\textsc{s}6a$'$)}{seq-le-delays-rf}}]
     if $\bEv\in\aEvs_1$, $\aEv\in\aEvs_2$ and $\labeling_1(\bEv) \rdelays \labeling_2(\aEv)$ 
@@
 \end{definition}
-% We write $\semmca{2}{}$ for the semantic function.
-A \PwTmca{2} need not satisfy requirement \ref{pom-rf-le}, and thus we may
-have $\bEv\xrfx\aEv$ and $\aEv\lt\bEv$.
+\ref{par-le-rf1} and \ref{par-le-rf2} ensure that $\bEv\xrfx\aEv$ implies
+$\bEv\lt\aEv$ when the actions come from different threads.  However, we may
+have $\bEv\xrfx\aEv$ and $\aEv\lt\bEv$ within a thread, as between
+$\DWP{x}{2}$ to $\DRP[\mACQ]{x}{2}$ in \ref{internal-acq}.  To allow this,
+\ref{seq-le-delays-rf} relaxes \ref{seq-le-delays} from \PwTmca{1}.
+\ref{seq-le-rf-rf} is then necessary to stop stale reads within a thread.
+For example, it prevents a read of $1$ in
+\begin{math}
+  \PW{x}{1}
+  \SEMI
+  \PW{x}{2}
+  \SEMI
+  \PR{x}{r}.
+\end{math}
+\todo{Is s6b necessary?}
 
@@
 % \emph{delays} to drop write-to-read order from $\eqreorderco$.
-
-\todo{Example using \ref{seq-le-delays-rf} and \ref{seq-le-rf-rf}. Perhaps
-  move \reflem{lem:mca2} to appendix.}
 

### 2021-10-28 paper/mca2.tex
removing todos. Integrated consistency and causal strengtheining.

@@
 $\sIF{}{}{}$.
-Rather than ensuring that there is no
-\emph{global} blocker for a sequentially fulfilled read \eqref{pom-rf-le}, we
-require only that there is no \emph{thread-local} blocker
-\eqref{seq-le-rf-rf}.
-For \PwTmca{2}, internal reads don't necessarily contribute to order, and
-thus the above execution is allowed.
+% Rather than ensuring that there is no
+% \emph{global} blocker for a sequentially fulfilled read \eqref{pom-rf-le}, we
+% require only that there is no \emph{thread-local} blocker
+% \eqref{seq-le-delays-rf}.
+% For \PwTmca{2}, internal reads don't necessarily contribute to order, and
+% thus the above execution is allowed.
+(For parallelism, we explicitly specify the domain of $\bEv$ and $\aEv$ in
+\ref{seq-le-delays-rf}.)
 
@@
   The definition of \PwTmca{2} is derived from that of \PwTmca{1} by
-  removing \ref{pom-rf-le} and adding the following:
+  removing \ref{pom-rf-le} and \ref{seq-le-delays} and adding the following:
 
@@
     then either $\bEv\xrfx\aEv$ or $\bEv\le\aEv$,
-  \item[{\labeltext[\textsc{s}6b]{(\textsc{s}6b)}{seq-le-rf-rf}}]
-    if $\bEv\in\aEvs_1$, $\aEv\in\aEvs_2$ and $\labeling_1(\bEv) \rblocks \labeling_2(\aEv)$
-    then $\cEv\xrfx\aEv$ implies $\bEv\le\cEv$.
+  % \item[{\labeltext[\textsc{s}6b]{(\textsc{s}6b)}{seq-le-rf-rf}}]
+  %   if $\bEv\in\aEvs_1$, $\aEv\in\aEvs_2$ and $\labeling_1(\bEv) \rblocks \labeling_2(\aEv)$
+  %   then $\cEv\xrfx\aEv$ implies $\bEv\le\cEv$.
     % \\
@@
 have $\bEv\xrfx\aEv$ and $\aEv\lt\bEv$ within a thread, as between
-$\DWP{x}{2}$ to $\DRP[\mACQ]{x}{2}$ in \ref{internal-acq}.  To allow this,
-\ref{seq-le-delays-rf} relaxes \ref{seq-le-delays} from \PwTmca{1}.
-\ref{seq-le-rf-rf} is then necessary to stop stale reads within a thread.
-For example, it prevents a read of $1$ in
+$\DWP{x}{2}$ to $\DRP[\mACQ]{x}{2}$ in \ref{internal-acq}, thus allowing this
+execution.
+% To accomplish this, \ref{seq-le-delays-rf} relaxes
+% \ref{seq-le-delays} from \PwTmca{1}.
+\ref {pom-rf-block} and \ref{seq-le-delays-rf} are sufficient to stop stale
+reads within a thread.  For example, it prevents a read of $1$ in
 \begin{math}
@@
 \end{math}
-\todo{Is s6b necessary?}
 
@@
         \event{b}{\DW{x}{1}}{right=of a}
-        \rfint[out=165,in=15]{b}{a}
-        \wki{a}{b}
+        \rfint[out=-165,in=-15]{b}{a}
+        \wki[out=15,in=165]{a}{b}
       \end{tikzinlinesmall}}}
 \end{math}, taking $\rrfx$ from the left and $\lt$ from the right.
-The use of $\rextends{}{}$ ensures this.
+The use of ``$\rextends{}{}$'' in \ref{if-le} and \ref{if-rf} ensures this.
 
@@
   \begin{itemize}
-  \item external fulfillment:
+  \item external fulfillment:\;\,
     $\bEv\lt\aEv$ and if $\labeling(\cEv) \rblocks\labeling(\aEv)$ then either $\cEv\le\bEv$ or $\aEv\le\cEv$, or
   \item internal fulfillment:
-    $(\exists\bEv'\in\fmrginv{\bEv})$
-    $(\exists\aEv'\in\fmrginv{\aEv})$
-    $\bEv'\xpox\aEv'$ and $(\not\exists\cEv)$
-    $\labelingForm(\cEv)$ is a tautology and
-    $\labeling(\cEv) \rblocks \labeling(\aEv)$ and $\bEv'\xpox\cEv\xpox\aEv'$.
+    \begin{math}
+      \begin{array}[t]{l}
+        (\exists\bEv'\in\fmrginv{\bEv})
+        (\exists\aEv'\in\fmrginv{\aEv})
+        \\ 
+        \bEv'\xpox\aEv' \textand (\not\exists\cEv)\;
+        % \labelingForm(\cEv) is a tautology \textand
+        \labeling(\cEv) \rblocks \labeling(\aEv) \textand
+        \bEv'\xpox\cEv\xpox\aEv'.        
+      \end{array}
+    \end{math}
   \end{itemize}

### 2021-10-28 paper/outro.tex
Acknowledgements: Anton
done with section 2
removing todos. Integrated consistency and causal strengtheining.

@@
-\section{Conclusions} %Future Work}
+\section{Limitations and Future Work}
 \label{sec:outro}
@@
 This paper is the first to present a direct denotational semantics for
-sequential composition in a relaxed memory model that can be
-efficiently compiled to modern CPUs. We extract from this model a semantic
-dependency relation and use it to build \PwTc, a solution to the
-Out-of-Thin-Air problem in \cXI, and \PwTmca{}, a model for Java-like
-languages.
-% There is, as usual, more research to be done.
+sequential composition in a relaxed memory model that can be efficiently
+compiled to modern CPUs. We extract from this model a semantic dependency
+relation and use it to build \PwTc, a solution to the Out-of-Thin-Air problem
+in \cXI, and \PwTmca{}, a model intended for safe languages such as Java and
+Javascript.  Our work has several limitations, providing opportunities for
+future work.
 
-We have not treated loops in this model, though we expect that the usual
-approach of showing continuity for all the semantic operations with respect
-to set inclusion would go through. \citet{DBLP:conf/esop/PaviottiCPWOB20} use
-step-indexing to account for loops; a similar approach could be applied here.
+We have mechanized some proofs, but not all.  In particular, we have only a
+pen-and-paper proof showing that \PwTmca{} supports optimal lowering to
+\armeight{}.  The same is true for local data race-freedom (\ldrfsc{}).
+Additionally, our proof sketch for \ldrfsc{} elides \RMW{}s, which have
+caused complications in other models \cite{promising-ldrf}.
+
+We have not treated loops, although we expect that the usual approach of
+showing continuity for all the semantic operations with respect to set
+inclusion would go through. \citet{DBLP:conf/esop/PaviottiCPWOB20} use
+step-indexing to account for loops; perhaps this approach could be adapted.
 
@@
 acquiring reads for \armeight{}.  It would be illuminating to find out what
-the performance penalty is for these fences.
+the performance penalty is for these fences.  In a similar vein, it would be
+interesting to know if there is a performance penalty for banning
+read-introduction, which is not valid for \PwT{}.
 
-% In a similar vein,
-% \PwT{} does not validate read introduction, whereas speculative operational
-% semantics do.  Recent work shows a tension between read introduction and
-% compositional reasoning for temporal safety properties (see
-% \textsection\ref{sec:related}).  Nonetheless, read introduction is
+\PwTc{} can be lowered efficiently to any architecture supported by \cXI{},
+but inherits the top-level axioms of \rcXI{}, compromising compositionality.
+\PwTmca{} is as a compositional as a model of concurrent imperative
+programming can be, but is limited to \mca{} architectures for optimal
+lowering.  It would be interesting to explore the middle ground to find a
+fully compositional model that supports optimal lowering to all modern
+architectures.
+
+% In a similar vein, \PwT{} does not validate read-introduction, whereas
+% speculative operational semantics do.  Recent work shows a tension between
+% read-introduction and compositional reasoning for temporal safety properties
+% (see \textsection\ref{sec:related}).  Nonetheless, read introduction is
 % ubiquitous in some compilers.  It would be interesting to know if there is a
-% performance penalty for banning read introduction.
+% performance penalty for banning read-introduction.
 
@@
 
+Supplementary material for this paper available at
+\url{https://weakmemory.github.io/pwt}.
+
 \subsection*{Acknowledgements}
@@
 
-Riely was supported by the National Science Foundation under
+James Riely was supported by the National Science Foundation under
 grant No.~CCR-1617175.
@@
 
-\todo{Add support for Anton\&Ilya.}
+Anton Podkopaev was supported by JetBrains Research.
 

### 2021-10-28 paper/related.tex
removing todos. Integrated consistency and causal strengtheining.

@@
 entailment to define dependency.  Although \PwT{} is based on \PwP{}, there
-are many differences.  Two of these are motivated by requirements unique to
-\PwT{}: First, \PwP{} uses substitution to define read actions; \PwT{}
-Skolemizes in order to satisfy disjunction closure for predicate transformers
-(\textsection\ref{sec:ex:assoc}).  Second, \PwP{} requires \emph{consistency}
-and \emph{causal strengthening}; \PwT{} drops these in order to satisfy
-associativity (\reflem{lem:monoid}).  Other differences are stylistic: For
+are many differences.
+Some of these are motivated by requirements unique to
+\PwT{} (see \textsection\ref{sec:ex:assoc}).
+% Two of these are motivated by requirements unique to
+% \PwT{}: First, \PwP{} uses substitution to define read actions; \PwT{}
+% Skolemizes in order to satisfy disjunction closure for predicate transformers
+% (\textsection\ref{sec:ex:assoc}).  Second, \PwP{} requires \emph{consistency}
+% and \emph{causal strengthening}; \PwT{} drops these in order to satisfy
+% associativity (\reflem{lem:monoid}).
+Other differences are stylistic: For
 example, we use termination \emph{conditions} rather than termination
@@
 this problem, since \PwT{} does not validate read-introduction.  Nonetheless,
-read introduction is ubiquitous in some compilers
+read-introduction is ubiquitous in some compilers
 \cite{DBLP:conf/pldi/LeeKSHDMRL17}.  There appears to be a genuine tension

### 2021-10-28 paper/results.tex
removing todos. Integrated consistency and causal strengtheining.

@@
 
-\PwP{} %\cite{DBLP:journals/pacmpl/JagadeesanJR20}
-is a novel memory model,
-intended to serve as a semantic basis for a Java-like language, where all
-access is safe.  \PwTmca{} generalizes \PwP{}, making several small but
-significant changes.  As a result, we have had to re-prove most of the
-theorems from \PwP{}.  The proofs can be found in \cite{appendix}.
+Prop.~6.1 of \cite{DBLP:journals/pacmpl/JagadeesanJR20} establishes a
+compositional principle for proving that programs validate formula in
+past-time temporal logic.  The principal is based entirely on the pomset
+order relation.  Its proof, and all of the no-thin-air examples in
+\cite[\textsection6]{DBLP:journals/pacmpl/JagadeesanJR20} hold equally for
+the models described here.
 
-% In \textsection\ref{sec:arm},
-We show that \PwTmca{1} supports the optimal
+% We \PwP{} %\cite{DBLP:journals/pacmpl/JagadeesanJR20}
+% is a novel memory model, intended to serve as a semantic basis for a safe
+% languages such as Java and Javascript.  \PwTmca{} generalizes \PwP{}, making
+% several small but significant changes.  As a result, we have had to re-prove
+% most of the theorems from \PwP{}.  These can be found in proofs can be found
+% in \refAppendices{}.
+
+In \refAppendix{sec:arm}, we show that \PwTmca{1} supports the optimal
 lowering of relaxed accesses to \armeight{} and that \PwTmca{2} supports the
@@
 
-%In \textsection\ref{sec:sc},
-We prove sequential consistency for
+In \refAppendix{sec:sc}, we also sketch a proof of sequential consistency for
 local-data-race-free programs.  The proof uses \emph{program order}, which we
 construct for \cXI{} in \textsection\ref{sec:c11}.  The same construction
-works for \PwTmca{}.  (This proof assumes there are no \RMW{} operations.)
+works for \PwTmca{}.  (This proof sketch assumes there are no \RMW{} operations.)
 
-The semantics validates many peephole optimizations, such as the standard reorderings on relaxed access:
+The semantics validates many peephole optimizations, such as reorderings on relaxed access:
 \begin{align*}
@@
 %\ref{5} imposes no order between events in \ref{RR}--\ref{RW}.  %Note that \ref{RR} allows aliasing.
-Here $\free(\aCmd)$ is the set of locations and registers that occur in $\aCmd$.
+Here $\free(\aExp)$ is the set of locations and registers that occur in $\aExp$.
 Using augmentation closure, the semantics also validates roach-motel reorderings \cite{SevcikThesis}.  For
@@
 \end{align*}
-
-Notably, the semantics does \emph{not} validate read introduction.  When
-combined with case analysis (\textsection\ref{sec:semca}), read introduction
+Notably, the semantics does \emph{not} validate read-introduction.  When
+combined with if-introduction (\textsection\ref{sec:semca}), read-introduction
 can break temporal reasoning.  This combination is allowed by speculative
 operational models.  See \textsection\ref{sec:related} for a discussion.
-
-Prop.~6.1 of \cite{DBLP:journals/pacmpl/JagadeesanJR20} establishes a
-compositional principle for proving that programs validate formula in
-past-time temporal logic.  The principal is based entirely on the pomset
-order relation.  Its proof, and all of the no-thin-air examples in
-\cite[\textsection6]{DBLP:journals/pacmpl/JagadeesanJR20} hold equally for
-the models described here.
 

### 2021-10-28 paper/related.tex
wordsmithing conclusion

@@
 
-There are few prior studies of relaxed memory that include sequential
-composition and/or precise calculation of semantic dependencies.  No prior
-work gives the semantics of sequential composition in direct style.
-\citet{DBLP:conf/esop/PaviottiCPWOB20} defined \MRD{}, which calculates
+There are two prior studies of relaxed memory that include precise
+calculation of semantic dependencies---neither gives the semantics of
+sequential composition in direct style.
+First, \citet{DBLP:conf/esop/PaviottiCPWOB20} defined \MRD{}, which calculates
 dependencies using event structures rather than logic.  This strategy is
 brittler than ours, leading to false positives (\textsection\ref{sec:lir}).
-\citet{DBLP:journals/pacmpl/JagadeesanJR20} defined \PwP{}, using logical
+Second, \citet{DBLP:journals/pacmpl/JagadeesanJR20} defined \PwP{}, using logical
 entailment to define dependency.  Although \PwT{} is based on \PwP{}, there

### 2021-10-28 paper/intro.tex
removing todos
Did a pass through Secs 1-3 (but not 3.8)
removing todos. Integrated consistency and causal strengtheining.
finished with 3.8

@@
 \cite{OHearn:2007:RCL:1235896.1236121}, or to accept inefficient
-implementation in order to obtain sequential consistency
+implementation in order to obtain sequential consistency (SC)
 \cite{DBLP:conf/snapl/MarinoMMNS15}.  When these approaches are not
@@
 
+Supplementary material for this paper available at
+\url{https://weakmemory.github.io/pwt}.
+
 \section{Overview}
@@
 \subsection{Memory Models}
+\label{sec:intro:mm}
 
@@
 preconditions, composing the predicate transformers.
-\begin{gather*}
+\begin{equation}
+\begin{gathered}
   \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
@@
   \end{tikzinline}}
-\end{gather*}
+\end{gathered}
+\label{eq1}
+\end{equation}
 This works for the sequentially consistent case, but needs to be

### 2021-10-28 paper/oopsla.tex
removing todos. Integrated consistency and causal strengtheining.
finished with 3.8

@@
 
-\subsection{Comparison to ``Promising Semantics'' [POPL 2017]}
+\subsection{Further Comparison to ``Promising Semantics'' [POPL 2017]}
 \label{sec:promising}
@@
 examples that prove inconsistency have the same shape: they combine
-read-introduction and case analysis (aka, if-introduction).  Effectively,
+read-introduction and if-introduction (aka, case analysis).  Effectively,
 this turns one read into two, where different conditional branches can be
@@
 {switching} to the \texttt{then} branch, based on a coin flip (${b}$).
-% Using a combination of read introduction and case analysis,
+% Using a combination of read-introduction and if-introduction,
 % the middle thread can be rewritten to:
@@
 
-% Case analysis gives very weak results when combined with read introduction.
+% Case analysis gives very weak results when combined with read-introduction.
 % See \cite{promising-ldrf}.
@@
 % \begin{quotation}
-%   LLVM frequently performs such load introductions in the “simplify CFG”
+%   LLVM frequently performs such load-introductions in the “simplify CFG”
 %   pass; e.g., when hoisting loads outside of conditionals.
@@
 
-\subsection{Comparison to ``Pomsets with Preconditions'' [OOPSLA 2020]}
+\subsection{Further Comparison to ``Pomsets with Preconditions'' [OOPSLA 2020]}
 \label{sec:diff}

### 2021-10-29 paper/tool.tex
tool.tex: refine table, remove timings, fix text

--- 2021-10-19 paper/tool.tex
@@
 Some further optimisations may be possible within the tool to improve the situation further, such as killing 'dead-end' pomsets at each sequence operator, or by doing a directed search for particular execution outcomes.
-Unsurprisingly the execution time is dominated by the calculation of the denotation, with the additional axiomatic filtering step of \PwTc{} being within margin of error difference of just calculating the \PwT{} semantics.
-% \PwTer{} will be made open source upon publication.
-\PwTer{} is available online at \url{https://github.com/graymalkin/pomsets-with-predicate-transformers}.
+For context, we have included the results of \MRD{} for the Java Causality tests~\cite{DBLP:conf/esop/PaviottiCPWOB20}.
+\PwTer{} is available with this paper's supplementary materials.
+
+\todo{Verify that referee comments sufficiently addressed regarding ``scalability challenges''.  Perhaps mention that it is easier for hardware models?}
 
@@
 \begin{figure}[t]
-\begin{center}
-  \begin{tabularx}{0.7\textwidth}{c||c|c|c}
-    \multirow{2}{*}{\bf Causality Test}   & \multicolumn{2}{c|}{\PwTc{}}    & \PwT{}\\
-    \cline{2-4}
-                                     & Result & Execution Time (s)  & Execution Time (s) \\
-    \cline{1-4}
-    jctc1                            & pass   & 2.397               & 2.608              \\
-    jctc2                            & pass   & 25.780              & 25.754             \\
-    jctc3                            & pass   & 196.935             & 205.120            \\
-    jctc4                            & pass   & 2.269               & 2.110              \\
-    jctc5                            & pass   & 63.714              & 69.441             \\
-    jctc6                            & pass   & 11.245              & 12.489             \\
-    jctc7                            & pass   & 88.250              & 96.099             \\
-    jctc8                            & pass   & 2.482               & 2.473              \\
-    jctc9                            & pass   & 13.592              & 15.384             \\
-    jctc10                           & pass   & 494.133             & 513.234            \\
-    jctc11                           & $\bot$ & --                  & --                 \\
-    jctc12                           & $\bot$ & --                  & --                 \\
-    jctc13                           & pass   & 2.101               & 2.247              \\
-    jctc17                           & pass   & 178.304             & 186.228            \\
-    jctc18                           & pass   & 177.292             & 2.247              \\
+\noindent\begin{center}
+  %\hspace{-9ex}%
+  \begin{tabularx}{\textwidth}{@{}c||c|c|c|c @{\extracolsep{5mm}}c||@{\extracolsep{1mm}}c|c|c|c}
+    % \multirow{2}{*}{\bf Test} &  & \multicolumn{3}{c}{\MRD} & & \multirow{2}{*}{\bf Test} &  & \multicolumn{3}{c}{\MRD}\\
+    % \cline{2-5} \cline{8-11}
+    {\bf Test} & \PwTc{}  & \MRD   & \MRDIMM & \MRDRC & {\bf Test} &\PwTc{} & \MRD & \MRDIMM & \MRDRC \\
+    \cline{1-5} \cline{6-10}
+    jctc1      & pass    & pass    & pass    & pass   & jctc9      & pass   & pass & pass     & pass \\
+    jctc2      & pass    & pass    & pass    & pass   & jctc10     & pass   & pass & pass     & pass \\
+    jctc3      & pass    & pass    & pass    & pass   & jctc11     & $\bot$ & pass & pass     & pass \\
+    jctc4      & pass    & pass    & pass    & pass   & jctc12     & $\bot$ & --   & --       & --   \\
+    jctc5      & pass    & pass    & pass    & pass   & jctc13     & pass   & pass & pass     & pass \\
+    jctc6      & pass    & pass    & pass    & pass   & jctc17     & pass   & fail & pass     & fail \\
+    jctc7      & pass    & pass    & pass    & pass   & jctc18     & pass   & fail & pass     & fail \\
+    jctc8      & pass    & pass    & pass    & pass
   \end{tabularx}
@@
 $\bot$ indicates the tool failed to run for this test due to a memory overflow.
-Tests run on an Intel i9-9980HK with 64 GB of memory, execution times are the mean of 3 runs.}
+Tests run on an Intel i9-9980HK with 64 GB of memory.
+For context, results for the \MRD, \MRDIMM, and \MRDRC{} are also included~\cite{DBLP:conf/esop/PaviottiCPWOB20}.}
 \end{center}

### 2021-10-29 paper/intro.tex
pass over intro

@@
 Then the equivalence holds in any context---this form of the equivalence
-enables reasoning about peephole optimizations.  Unlike prior work, \PwT{}
-allows the presence or absence of a dependency can be understood in
-isolation.  This enables incremental and modular validation of assumptions
+enables reasoning about peephole optimizations.  Said differently, unlike prior work, \PwT{}
+allows the presence or absence of a dependency to be understood in
+isolation---this enables incremental and modular validation of assumptions
 about program dependencies in larger blocks of code.
@@
 additional features, such as address calculation and \RMW{}s, in
-\textsection\ref{sec:additional}.  We further discuss related work in
-\textsection\ref{sec:related}.
+\textsection\ref{sec:additional}.  We discuss related and future work in
+\textsection\ref{sec:related}--\ref{sec:outro}.
 
@@
 on \PwP{} \cite{DBLP:journals/pacmpl/JagadeesanJR20}.
-\textsection\ref{sec:results} summarizes the results for this model.  The
-dependency relation calculated by \PwT{} can also be used with off-the-shelf
-models.  For example, in \textsection\ref{sec:c11} we show that it can be
-used as an $\rsdep$ relation for \cXI, adapting the approach of \MRD{}
+\textsection\ref{sec:results} summarizes the results for this model.  In
+addition to powering such a bespoke model, the dependency relation calculated
+by \PwT{} can also be used with off-the-shelf models.  For example, in
+\textsection\ref{sec:c11} we show that it can be used as an $\rsdep$ relation
+for \cXI, adapting the approach of \MRD{}
 \cite{DBLP:conf/esop/PaviottiCPWOB20}.  % This is only the second semantics
@@
 overhead in the implementation of relaxed reads.  Safe language like OCaml
-\cite{Dolan:2018:BDR:3192366.3192421} have typically made the opposite choice.  Just
-\PwT{} can be used to strengthen \cXI{}, it could also be used to weaken these
-models, allowing optimal lowering for relaxed reads while banning thin-air.
+\cite{Dolan:2018:BDR:3192366.3192421} have typically made the opposite
+choice, accepting a performance penalty in order to avoid thin-air.  Just as
+\PwT{} can be used to strengthen \cXI{}, it could also be used to weaken
+these models, allowing optimal lowering for relaxed reads while banning
+thin-air.
 
@@
 
-Supplementary material for this paper available at
+Supplementary material for this paper is available at
 \url{https://weakmemory.github.io/pwt}.

### 2021-10-29 paper/tool.tex
improve table typesetting, fix last todos.

@@
 Finally, \PwTer{} also allows us to toggle the complete check of~\ref{def:complete}, providing an interface for understanding how fragments of code might compose by exposing preconditions and termination conditions that are not yet tautologies.
-We have run \PwTer{} over the Java Causality Tests~\cite{PughWebsite} supported in the input syntax, and tabulated the results in Figure~\ref{fig:tool}.
+We have run \PwTer{} over the Java Causality Tests~\cite{PughWebsite} supported in the input syntax, and tabulated the results in Table~\ref{fig:tool}.
 
@@
 For context, we have included the results of \MRD{} for the Java Causality tests~\cite{DBLP:conf/esop/PaviottiCPWOB20}.
+Of note is that test cases 17 and 18 for \MRD{} and \MRDRC{} do not give the correct outcome, this is for similar reasons to the example given in \textsection\ref{sec:lir}, where local invariant reasoning in \MRD{} is too constrained.
 \PwTer{} is available with this paper's supplementary materials.
 
-\todo{Verify that referee comments sufficiently addressed regarding ``scalability challenges''.  Perhaps mention that it is easier for hardware models?}
+% \todo{Verify that referee comments sufficiently addressed regarding ``scalability challenges''.  Perhaps mention that it is easier for hardware models?}
 
 %% [Simon] I am not a fan of how this looks on the page.
-\begin{figure}[t]
-\noindent\begin{center}
-  %\hspace{-9ex}%
-  \begin{tabularx}{\textwidth}{@{}c||c|c|c|c @{\extracolsep{5mm}}c||@{\extracolsep{1mm}}c|c|c|c}
-    % \multirow{2}{*}{\bf Test} &  & \multicolumn{3}{c}{\MRD} & & \multirow{2}{*}{\bf Test} &  & \multicolumn{3}{c}{\MRD}\\
-    % \cline{2-5} \cline{8-11}
-    {\bf Test} & \PwTc{}  & \MRD   & \MRDIMM & \MRDRC & {\bf Test} &\PwTc{} & \MRD & \MRDIMM & \MRDRC \\
-    \cline{1-5} \cline{6-10}
-    jctc1      & pass    & pass    & pass    & pass   & jctc9      & pass   & pass & pass     & pass \\
-    jctc2      & pass    & pass    & pass    & pass   & jctc10     & pass   & pass & pass     & pass \\
-    jctc3      & pass    & pass    & pass    & pass   & jctc11     & $\bot$ & pass & pass     & pass \\
-    jctc4      & pass    & pass    & pass    & pass   & jctc12     & $\bot$ & --   & --       & --   \\
-    jctc5      & pass    & pass    & pass    & pass   & jctc13     & pass   & pass & pass     & pass \\
-    jctc6      & pass    & pass    & pass    & pass   & jctc17     & pass   & fail & pass     & fail \\
-    jctc7      & pass    & pass    & pass    & pass   & jctc18     & pass   & fail & pass     & fail \\
-    jctc8      & pass    & pass    & pass    & pass
-  \end{tabularx}
+\newcommand{\good}{{\color{green!70!black}\ding{52}}}
+\newcommand{\nogood}{{\color{red!90!black}\ding{56}}}
+\begin{table}[t]
   \caption{\label{fig:tool} Tool results for supported Java Causality Test Cases~\cite{PughWebsite}.
@@
 For context, results for the \MRD, \MRDIMM, and \MRDRC{} are also included~\cite{DBLP:conf/esop/PaviottiCPWOB20}.}
+
+\noindent\begin{center}
+  \begin{minipage}[t]{0.49\textwidth}
+  \begin{tabular}{l*{4}c}
+    \toprule
+    {\bf Test} & \PwTc & \MRD  & \MRDIMM & \MRDRC \\
+    \midrule
+    jctc1      & \good & \good & \good   & \good \\
+    jctc2      & \good & \good & \good   & \good \\
+    jctc3      & \good & \good & \good   & \good \\
+    jctc4      & \good & \good & \good   & \good \\
+    jctc5      & \good & \good & \good   & \good \\
+    jctc6      & \good & \good & \good   & \good \\
+    jctc7      & \good & \good & \good   & \good \\
+    jctc8      & \good & \good & \good   & \good \\
+    \bottomrule
+  \end{tabular}
+  \end{minipage}\hfill
+  \begin{minipage}[t]{0.49\textwidth}
+  \begin{tabular}{l*{4}c}
+    \toprule
+    {\bf Test} & \PwTc  & \MRD    & \MRDIMM & \MRDRC \\
+    \midrule
+    jctc9      & \good  & \good   & \good   & \good \\
+    jctc10     & \good  & \good   & \good   & \good \\
+    jctc11     & $\bot$ & \good   & \good   & \good \\
+    jctc12     & $\bot$ & --      & --      & --   \\
+    jctc13     & \good  & \good   & \good   & \good \\
+    jctc17     & \good  & \nogood & \good   & \nogood \\
+    jctc18     & \good  & \nogood & \good   & \nogood \\
+    \bottomrule
+  \end{tabular}
+  \end{minipage}
 \end{center}
-\end{figure}
+\end{table}

### 2021-10-29 paper/tool.tex
Word smithing tool.tex

@@
 We have run \PwTer{} over the Java Causality Tests~\cite{PughWebsite} supported in the input syntax, and tabulated the results in Table~\ref{fig:tool}.
+For context, we have included the results of \MRD{} for the Java Causality tests~\cite{DBLP:conf/esop/PaviottiCPWOB20}.
+Of note is that test cases 17 and 18 for \MRD{} and \MRDRC{} do not give the correct outcome, this is for similar reasons to the example given in \textsection\ref{sec:lir}, where local invariant reasoning in \MRD{} is too constrained.
 
@@
 Significant effort has been put into throwing away spurious merges early in \PwTer, so that executing the tool remains manageable for small examples.
-Some further optimisations may be possible within the tool to improve the situation further, such as killing 'dead-end' pomsets at each sequence operator, or by doing a directed search for particular execution outcomes.
-For context, we have included the results of \MRD{} for the Java Causality tests~\cite{DBLP:conf/esop/PaviottiCPWOB20}.
-Of note is that test cases 17 and 18 for \MRD{} and \MRDRC{} do not give the correct outcome, this is for similar reasons to the example given in \textsection\ref{sec:lir}, where local invariant reasoning in \MRD{} is too constrained.
+Some further optimisations may be possible within the tool to improve the situation further, such as killing `dead-end' pomsets at each sequence operator, or by doing a directed search for particular execution outcomes.
 \PwTer{} is available with this paper's supplementary materials.

### 2021-10-29 paper/intro.tex
updates
cleanup

@@
 is acyclic and hence allowed. For example dropping the control dependency
-results in:
+results in the following execution, which should be allowed:
 \begin{gather*}
@@
 While syntactic dependency calculation suffices for hardware models, it is
-not preserved by common compiler optimizations. For example, if we calculate
-control dependencies syntactically, then there is a dependency from
-$(\DR{x}{1})$ to $(\DW{y}{1})$, and therefore a cycle in, the candidate
-execution:
+not preserved by common compiler optimizations. For example, consider the
+following program:
 \begin{gather*}
@@
   \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
-  \\[-.4ex]
+\end{gather*}
+Because $\PW{y}{1}$ occurs on both branches of the conditional, a compiler
+may lift it out.  With the dependency removed, the compiler could reorder the
+read of $x$ and write to $y$, allowing both reads to see $1$.  Attempting to
+generate this execution with syntactic dependencies, however, results in the
+following candidate execution, which has a cycle and therefore is disallowed:
+\begin{gather*}
   \hbox{\begin{tikzinline}[node distance=1.5em]
@@
 \end{gather*}
-A compiler may lift the assignment $\PW{y}{1}$ out of the conditional,
-thus removing the dependency.
 
@@
 shared by all processors, which is neatly captured by the order of the
-pomset.
+pomset (see \textsection\ref{sec:mca}).
 
@@
 the empty pomset in the semantics of every program fragment.  To then ensure
-that $\SKIP$ is not a refinement of $\PW{x}{1}$, \PwP{} include the
+that $\SKIP$ is not a refinement of $\PW{x}{1}$, \PwP{} include a
 \emph{termination} action, $\aTerm{}$, which we have elided in the examples
@@
     \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
-        \event{rx0}{\DR{x}{0}}{}
-        \xform{rx0d}{(0{=}r) \limplies \bForm}{right=of rx0}
+        \event{rx0}{\DR{x}{1}}{}
+        \xform{rx0d}{(1{=}r) \limplies \bForm}{right=of rx0}
         %\xo{rx0}{rx0d}
@@
     \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
-        \event{ry0}{\DR{y}{0}}{}
-        \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
+        \event{ry0}{\DR{y}{1}}{}
+        \xform{ry0d}{(1{=}s) \limplies \bForm}{right=of ry0}
         %\xo{ry0}{ry0d}
@@
     \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
-        \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \bigmid \DW{z}{0}}{}
+        \event{wz0}{(s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{}
         \xform{wz0d}{\bForm[r{*}s/z]}{right=of wz0}
@@
 assumption that registers are assigned at most once.  Instead, we use
-$(0{=}r) \limplies \bForm$, reflecting the fact that $0$ may come from a
+$(1{=}r) \limplies \bForm$, reflecting the fact that $1$ may come from a
 concurrent write.  The obligation to find a matching write is moved from the
@@
 \begin{gathered}
-  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
+  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}(s{-}1)} \FI
   \\
   \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
-      \event{rx0}{\DR{x}{0}}{}
-      \event{ry0}{\DR{y}{0}}{right=of rx0}
-      \event{wz0}{(0{=}r) \limplies (0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \bigmid \DW{z}{0}}{right=of ry0}
-      \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm[r{*}s/z]}{right=of wz0}
+      \event{rx0}{\DR{x}{1}}{}
+      \event{ry0}{\DR{y}{1}}{right=of rx0}
+      \event{wz0}{(1{=}r) \limplies (1{=}s) \limplies (s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
+      \xform{rx0ry0d}{(1{=}r) \limplies(1{=}s) \limplies \bForm[r{*}(s{-}1)/z]}{right=of wz0}
       \wki{rx0}{ry0}
@@
     \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
-        \event{rx0}{\DR{x}{0}}{}
+        \event{rx0}{\DR{x}{1}}{}
         \xform{rx0i}{\bForm}{left=.4em of rx0}
-        \xform{rx0d}{(0{=}r) \limplies \bForm}{right=of rx0}
+        \xform{rx0d}{(1{=}r) \limplies \bForm}{right=of rx0}
         \xo{rx0}{rx0d}
@@
     \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
-        \event{ry0}{\DR{y}{0}}{}
+        \event{ry0}{\DR{y}{1}}{}
         \xform{ry0i}{\bForm}{left=.4em of ry0}
-        \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
+        \xform{ry0d}{(1{=}s) \limplies \bForm}{right=of ry0}
         \xo{ry0}{ry0d}
@@
   %   \hbox{\begin{tikzinline}[node distance=1ex and1.5em]
-  %     \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \bigmid \DW{z}{0}}{}
+  %     \event{wz0}{(s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{}
   %     \xform{wz0i}{\bForm}{left=.4em of wz0}
@@
 for \eqref{eq1} by choosing
-the predicate transformer which is independent of $(\DR x0)$
-but dependent on $(\DR y0)$, which is the transformer
-which maps $\bForm$ to $(0{=}s) \limplies \bForm$.
+the predicate transformer which is independent of $(\DR x1)$
+but dependent on $(\DR y1)$, which is the transformer
+which maps $\bForm$ to $(1{=}s) \limplies \bForm$.
 (In subsequent diagrams, we only show predicate transformers for reads.)
 \begin{gather*}
-  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
+  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}(s{-}1)} \FI
   \\
@@
       \xform{wz0i}{\bForm}{}
-      \xform{rx0d}{(0{=}r) \limplies \bForm}{right=of wz0i}
-      \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{right=of rx0d}
-      \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of rx0ry0d}
-      \event{rx0}{\DR{x}{0}}{above left=of rx0ry0d}
-      \event{ry0}{\DR{y}{0}}{above right=of rx0ry0d}
-      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \bigmid \DW{z}{0}}{right=of ry0}
+      \xform{rx0d}{(1{=}r) \limplies \bForm}{right=of wz0i}
+      \xform{rx0ry0d}{(1{=}r) \limplies(1{=}s) \limplies \bForm}{right=of rx0d}
+      \xform{ry0d}{(1{=}s) \limplies \bForm}{right=of rx0ry0d}
+      \event{rx0}{\DR{x}{1}}{above left=of rx0ry0d}
+      \event{ry0}{\DR{y}{1}}{above right=of rx0ry0d}
+      \event{wz0}{(1{=}s) \limplies (s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
       \po{ry0}{wz0}
@@
 In the diagram, the dotted lines indicate set inclusion into the index of the transformer-family.
-As a quick correctness test, we can see that sequential composition is
+As a quick correctness check, we can see that sequential composition is
 associative in this case, since it does not matter whether we
-associate to the left, with the intermediate step eliding $\DWP{z}{0}$ in the
-diagram above,
+associate to the left---with the intermediate step as in the diagram above,
+eliding the write action---
 % \begin{gather*}
@@
 % \end{gather*}
-or to the right, with the intermediate step:
+or to the right---with the intermediate step:
 \begin{gather*}
-  \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
+  \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}(s{-}1)} \FI
   \\
   \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
-      \event{ry0}{\DR{y}{0}}{}
-      \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \bigmid \DW{z}{0}}{right=of ry0}
-      \xform{ry0d}{(0{=}s) \limplies \bForm}{left=of ry0}
+      \event{ry0}{\DR{y}{1}}{}
+      \event{wz0}{(1{=}s) \limplies (s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
+      \xform{ry0d}{(1{=}s) \limplies \bForm}{left=of ry0}
       \xform{wz0i}{\bForm}{left=of ry0d}
@@
 \end{gather*}
-This is an instance of the general result that sequential composition forms a monoid.
+This is an instance of the general result that sequential composition forms a
+monoid (\reflem{lem:monoid}).

### 2021-10-29 paper/tool.tex
cleanup

@@
+\input{fig-pwter.tex}
 \section{\PwTerTITLE: automatic litmus test evaluator}
@@
 % \todo{Verify that referee comments sufficiently addressed regarding ``scalability challenges''.  Perhaps mention that it is easier for hardware models?}
-
-%% [Simon] I am not a fan of how this looks on the page.
-\newcommand{\good}{{\color{green!70!black}\ding{52}}}
-\newcommand{\nogood}{{\color{red!90!black}\ding{56}}}
-\begin{table}[t]
-  \caption{\label{fig:tool} Tool results for supported Java Causality Test Cases~\cite{PughWebsite}.
-$\bot$ indicates the tool failed to run for this test due to a memory overflow.
-Tests run on an Intel i9-9980HK with 64 GB of memory.
-For context, results for the \MRD, \MRDIMM, and \MRDRC{} are also included~\cite{DBLP:conf/esop/PaviottiCPWOB20}.}
-
-\noindent\begin{center}
-  \begin{minipage}[t]{0.49\textwidth}
-  \begin{tabular}{l*{4}c}
-    \toprule
-    {\bf Test} & \PwTc & \MRD  & \MRDIMM & \MRDRC \\
-    \midrule
-    jctc1      & \good & \good & \good   & \good \\
-    jctc2      & \good & \good & \good   & \good \\
-    jctc3      & \good & \good & \good   & \good \\
-    jctc4      & \good & \good & \good   & \good \\
-    jctc5      & \good & \good & \good   & \good \\
-    jctc6      & \good & \good & \good   & \good \\
-    jctc7      & \good & \good & \good   & \good \\
-    jctc8      & \good & \good & \good   & \good \\
-    \bottomrule
-  \end{tabular}
-  \end{minipage}\hfill
-  \begin{minipage}[t]{0.49\textwidth}
-  \begin{tabular}{l*{4}c}
-    \toprule
-    {\bf Test} & \PwTc  & \MRD    & \MRDIMM & \MRDRC \\
-    \midrule
-    jctc9      & \good  & \good   & \good   & \good \\
-    jctc10     & \good  & \good   & \good   & \good \\
-    jctc11     & $\bot$ & \good   & \good   & \good \\
-    jctc12     & $\bot$ & --      & --      & --   \\
-    jctc13     & \good  & \good   & \good   & \good \\
-    jctc17     & \good  & \nogood & \good   & \nogood \\
-    jctc18     & \good  & \nogood & \good   & \nogood \\
-    \bottomrule
-  \end{tabular}
-  \end{minipage}
-\end{center}
-\end{table}

### 2021-11-01 paper/abstract.tex
accepted!

@@
 % events of $\sem{S_1}$ upon which $e$ depends.
-When calculating $\sem{S_1;S_2}$, the predicate transformer applied to the
-precondition of an event $e$ from $\sem{S_2}$ is chosen based on the set of
-events in $\sem{S_1}$ upon which $e$ depends.  We apply this approach to two
+When calculating the meaning of $(S_1;S_2)$, the predicate transformer applied to the
+precondition of an event $e$ from ${S_2}$ is chosen based on the set of
+events in ${S_1}$ upon which $e$ depends.  We apply this approach to two
 existing memory models.

### 2021-11-01 paper/drf.tex
removing todos. Integrated consistency and causal strengtheining.
accepted!

@@
 % \section{Local Data Race Freedom and Sequential Consistency}
-\section{LDRF-SC for \PwTmcaTITLE{}}
+\section{Proof Sketch: LDRF-SC for \PwTmcaTITLE{}}
 \label{sec:sc}
 
-\todo{Remove this section?}
+%\todo{Remove this section?}
 
 \begin{changed}
-  In this appendix, we establish a \drfsc{} for \PwTmca{2}.  We prove an
+  In this appendix, we sketch a proof of \drfsc{} for \PwTmca{2}.  We prove an
   \emph{external} result, where the notion of \emph{data-race} is independent
@@
 
-\bookmark{Definitions}
+\subsection{Definitions}
 \begin{changed}
@@
 
-\bookmark{Theorem and Proof Sketch}
+\subsection{Theorem and Proof Sketch}
 \begin{theorem}
@@
   (coherence-after-external) orders read and write accesses before writes,
-  satisfying \ref{pom-rf-block}. (Within a thread, \ref{seq-le-delays-rf} and
-  \ref{seq-le-rf-rf} induce order that is included in ${\rppo}$.)
+  satisfying \ref{pom-rf-block}. (Within a thread, \ref{seq-le-delays-rf} 
+  %and \ref{seq-le-rf-rf}
+  induces order that is included in ${\rppo}$.)
 
@@
 \end{proof}
+
+\subsection{A Note on Prior Work}
+
+In preparing this paper, we came across the following example, which appears
+to invalidate Theorem 4.1 of \cite{DBLP:conf/ppopp/DongolJR19}.
+\begin{gather}
+  \nonumber
+  \PW{x}{1}\SEMI
+  \PW[\mREL]{y}{1}\SEMI
+  \PR[\mACQ]{x}{r}
+  \PAR
+  \IF{\PR[\mACQ]{y}{}}\THEN \PW[\mREL]{x}{2}\FI
+  \\
+  \tag{\P}
+  \label{mix1}
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a1}{\DW{x}{1}}{}
+      \raevent{a2}{\DW[\mREL]{y}{1}}{right=of a1}
+      \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
+      \raevent{b1}{\DR[\mACQ]{y}{1}}{right=3em of a3}
+      % \raevent{b1}{\DR[\mACQ]{y}{1}}{below=of a1}
+      \raevent{b2}{\DW[\mREL]{x}{2}}{right=of b1}
+      \sync{a1}{a2}
+      \rf[out=20,in=160]{a1}{a3}
+      \rf[out=20,in=160]{a2}{b1}
+      \wk[out=20,in=160]{a3}{b2}
+      \sync{b1}{b2}
+      % \node(ai)[left=3em of a1]{};
+      % \bgoval[yellow!50]{(ai)}{P}
+      % \bgoval[pink!50]{(a1)(a2)(b1)(b2)}{P'\setminus P}
+      % \bgoval[green!10]{(a3)}{P'''\setminus P'}
+    \end{tikzinline}}
+  \\
+  \nonumber
+  % \label{mix2}
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a1}{\DW{x}{1}}{}
+      \raevent{a2}{\DW[\mREL]{y}{1}}{right=of a1}
+      \raevent{a3}{\DR[\mACQ]{x}{2}}{right=of a2}
+      \raevent{b1}{\DR[\mACQ]{y}{1}}{right=3em of a3}
+      \raevent{b2}{\DW[\mREL]{x}{2}}{right=of b1}
+      \sync{a1}{a2}
+      \rf[out=20,in=160]{a2}{b1}
+      \rf[out=-160,in=-20]{b2}{a3}
+      \sync{b1}{b2}
+    \end{tikzinline}}
+\end{gather}
+The program is data-race free.  The two executions shown are the only
+top-level executions that include $\DWP[\mREL]{x}{2}$.
+
+Theorem 4.1 of \cite{DBLP:conf/ppopp/DongolJR19} is stated by extending
+execution sequences.  In the terminology of
+\cite{DBLP:conf/ppopp/DongolJR19}, a read is \emph{$L$-weak} if it is
+sequentially stale.  Let
+\begin{math}
+  \rho=\DWP{x}{1}\allowbreak
+  \DWP[\mREL]{y}{1}\allowbreak
+  \DRP[\mACQ]{y}{1}\allowbreak
+  \DWP[\mREL]{x}{2}
+\end{math}
+be a sequence and
+\begin{math}
+  \alpha=\DRP[\mACQ]{x}{1}.
+\end{math}
+$\rho$ is $L$-sequential and $\alpha$ is $L$-weak in $\rho\alpha$.  But there
+is no execution of this program that includes a data race, contradicting the
+theorem.  The error seems to be in Lemma A.4 of
+\cite{DBLP:conf/ppopp/DongolJR19}, which states that if $\alpha$ is $L$-weak
+after an $L$-sequential $\rho$, then $\alpha$ must be in a data race.  That
+is clearly false here, since $\DRP[\mACQ]{x}{1}$ is stale, but the program is
+data race free.
+
+In proving the SC-LDRF result in \cite[\textsection8]{DBLP:journals/pacmpl/JagadeesanJR20}, we noted that our proof
+technique is more robust than that of \cite{DBLP:conf/ppopp/DongolJR19},
+because it limits the prefixes that must be considered.  In \eqref{mix1}, the
+induction hypothesis requires that we add $\DRP[\mACQ]{x}{1}$ before
+$\DWP[\mREL]{x}{2}$ since $\DRP[\mACQ]{x}{1}\xwk\DWP[\mREL]{x}{2}$.  In
+particular,
+\begin{gather*}
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a1}{\DW{x}{1}}{}
+      \raevent{a2}{\DW[\mREL]{y}{1}}{right=of a1}
+      % \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
+      \raevent{b1}{\DR[\mACQ]{y}{1}}{right=3em of a3}
+      % \raevent{b1}{\DR[\mACQ]{y}{1}}{below=of a1}
+      \raevent{b2}{\DW[\mREL]{x}{2}}{right=of b1}
+      \sync{a1}{a2}
+      % \rf[out=20,in=160]{a1}{a3}
+      \rf{a2}{b1}
+      % \wk[out=20,in=160]{a3}{b2}
+      \sync{b1}{b2}
+      % \node(ai)[left=3em of a1]{};
+      % \bgoval[yellow!50]{(ai)}{P}
+      % \bgoval[pink!50]{(a1)(a2)(b1)(b2)}{P'\setminus P}
+      % \bgoval[green!10]{(a3)}{P'''\setminus P'}
+    \end{tikzinline}}
+\end{gather*}
+is not a downset of \eqref{mix1}, because
+$\DRP[\mACQ]{x}{1}\xwk\DWP[\mREL]{x}{2}$.  As noted in \cite[\textsection8]{DBLP:journals/pacmpl/JagadeesanJR20},
+this affects the inductive order in which we move across pomsets, but does
+not affect the set of pomsets that are considered.  In particular,
+\begin{gather*}
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a1}{\DW{x}{1}}{}
+      \raevent{a2}{\DW[\mREL]{y}{1}}{right=of a1}
+      % \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
+      \raevent{b1}{\DR[\mACQ]{y}{1}}{right=3em of a3}
+      % \raevent{b1}{\DR[\mACQ]{y}{1}}{below=of a1}
+      % \raevent{b2}{\DW[\mREL]{x}{2}}{right=of b1}
+      \sync{a1}{a2}
+      % \rf[out=20,in=160]{a1}{a3}
+      \rf{a2}{b1}
+      % \wk[out=-20,in=-160]{a3}{b2}
+      % \sync{b1}{b2}
+      % \node(ai)[left=3em of a1]{};
+      % \bgoval[yellow!50]{(ai)}{P}
+      % \bgoval[pink!50]{(a1)(a2)(b1)(b2)}{P'\setminus P}
+      % \bgoval[green!10]{(a3)}{P'''\setminus P'}
+    \end{tikzinline}}
+\end{gather*}
+is a downset of \eqref{mix1}.

### 2021-11-01 paper/intro.tex
accepted!

@@
 such a model: For each read operation, it determines the set of available
-values.  Compilers and runtime systems are allowed to chose any value in the
+values.  Compilers and runtime systems are allowed to choose any value in the
 set.  To allow efficient implementation, the set must not be too small.  To
@@
 disallows nonsense behaviors that don't arise in practice.  The latter are
-commonly known as ``thin air'' behaviors \cite{DBLP:conf/esop/BattyMNPS15}.
+commonly known as ``thin-air'' behaviors \cite{DBLP:conf/esop/BattyMNPS15}.
 There are only a handful of solutions, and all have deficiencies.  These can
@@
 \item No dependencies, as in \cXI{} \cite{DBLP:conf/esop/BattyMNPS15} and
-  Javascript \cite{DBLP:journals/pacmpl/WattRP19}.  This allows thin air
+  Javascript \cite{DBLP:journals/pacmpl/WattRP19}.  This allows thin-air
   executions.
@@
 litmus tests in this model.  \cXI{} allows thin-air in order to avoid
-overhead in the implementation of relaxed reads.  Safe language like OCaml
+overhead in the implementation of relaxed reads.  Safe languages like OCaml
 \cite{Dolan:2018:BDR:3192366.3192421} have typically made the opposite

### 2021-11-01 paper/oopsla.tex
accepted!

@@
 \label{sec:discussion}
+
+\subsection{Read-Read Dependencies, If-Introduction, and Java Final Field Semantics}
+
+One might worry that the lack of read-read dependencies could cause \drfsc{}
+to fail.  For example, the following execution has a control dependency
+between the reads of the last thread, but this order is not enforced, neither
+by our model, nor \armeight.
+\begin{gather*}
+  \PW{z}{1}\SEMI\PW[\mREL]{y}{1}
+  \PAR
+  \PR[\mACQ]{y}{r}\SEMI\PW[\mREL]{x}{1}
+  \PAR
+  \IF{\PR{x}{}}\THEN\PR{z}{s}\FI  
+  \\
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a}{\DW{z}{1}}{}
+      \event{b}{\DW[\mREL]{y}{1}}{right=of a}
+      \event{c}{\DR[\mACQ]{y}{1}}{right=3em of b}
+      \event{d}{\DW[\mREL]{x}{1}}{right=of c}
+      \event{e}{\DR{x}{1}}{right=3em of d}
+      \event{f}{\DR{z}{0}}{right=of e}
+      \sync{a}{b}
+      \rf{b}{c}
+      \sync{c}{d}
+      \rf{d}{e}
+      % \ctrl{e}{f}
+      \wk[out=-170,in=-10]{f}{a}
+    \end{tikzinline}}
+\end{gather*}
+If the first read of the last thread is acquiring, then the execution is
+disallowed, since acquiring reads are ordered with respect to the reads that
+follow.
+\begin{gather*}
+  \PW{z}{1}\SEMI\PW[\mREL]{y}{1}
+  \PAR
+  \PR[\mACQ]{y}{r}\SEMI\PW[\mREL]{x}{1}
+  \PAR
+  \IF{\PR[\mACQ]{x}{}}\THEN\PR{z}{s}\FI  
+  \\
+  \hbox{\begin{tikzinline}[node distance=1.5em]
+      \event{a}{\DW{z}{1}}{}
+      \event{b}{\DW[\mREL]{y}{1}}{right=of a}
+      \event{c}{\DR[\mACQ]{y}{1}}{right=3em of b}
+      \event{d}{\DW[\mREL]{x}{1}}{right=of c}
+      \event{e}{\DR[\mACQ]{x}{1}}{right=3em of d}
+      \event{f}{\DR{z}{0}}{right=of e}
+      \sync{a}{b}
+      \rf{b}{c}
+      \sync{c}{d}
+      \rf{d}{e}
+      \po{e}{f}
+      \wk[out=-170,in=-10]{f}{a}
+    \end{tikzinline}}
+\end{gather*}
+
+\armeight{} enforces address dependencies between reads, but not control
+dependencies.  To support if-introduction (\AKA{} case-analysis), we drop all
+dependencies between reads.  This, in turn, invalidates Java's final field
+semantics.
+\begin{gather*}
+  \taglabel{addr2}
+  \begin{gathered}
+    \PBR{
+      \LET{r}{1} \SEMI
+      \PW{\REF{r}}{0} \SEMI
+      \PW{\REF{r}}{1} \SEMI
+      \PW[\mREL]{x}{r}
+    }
+    \PAR
+    \PBR{
+      \PR[\mACQ]{x}{r} \SEMI
+      \PR{\REF{r}}{s}
+    }
+    \\
+    \hbox{\begin{tikzinline}[node distance=1.5em]
+        \event{a1}{\DW{\REF{1}}{0}}{}
+        \event{a2}{\DW{\REF{1}}{1}}{right=of a1}
+        \wk{a1}{a2}
+        \event{a4}{\DWRel{x}{1}}{right=of a2}
+        \sync{a2}{a4}
+        \event{b1}{\DR[\mRA]{x}{1}}{right=3em of a4}
+        \event{b2}{\DR{\REF{1}}{0}}{right=of b1}
+        \sync{b1}{b2}
+        \rf{a4}{b1}
+        \wk[out=-170,in=-10]{b2}{a2}
+      \end{tikzinline}}
+  \end{gathered}
+\end{gather*}
+The acquire annotation is required to ensure publication.  If address
+dependencies were enforced between reads then the acquire annotation could be
+dropped.  However, the compiler would need to track address dependencies in
+order to ensure that if-introduction did not convert them to control
+dependencies.
 
@@
 
-\myparagraph{Substitution}
+\subsubsection*{Substitution}
 
@@
 
-\myparagraph[p:downset]{Downset closure}
-
+% \myparagraph[p:downset]{Downset closure}
+\subsubsection*{Downset closure}
+\label{p:downset}
 \jjr{} enforces downset closure in the prefixing rule.  Even without this,
@@
 
-\myparagraph{Consistency}
+\subsubsection*{Consistency}
 \jjr{} imposes \emph{consistency}, which requires that for every pomset
@@
 
-\myparagraph{Causal Strengthening}
+\subsubsection*{Causal Strengthening}
 % \labeltext[]{Causal Strengthening}{xCausal}
@@
 
-% \myparagraph{Causal Strengthening and Address Dependencies}
+% \subsubsection*{Causal Strengthening and Address Dependencies}
 % \labeltext[]{Causal Strengthening and Address Dependencies}{xADDRxRRD}
@@
   One use of causal strengthening is to ensure that address dependencies do
-  not introduce thin air reads.  Associating to the right, the intermediate
+  not introduce thin-air reads.  Associating to the right, the intermediate
   state of \ref{addr2} (\textsection\ref{sec:addr}) is:
@@
 
-\myparagraph{Internal Acquiring Reads}
+\subsubsection*{Internal Acquiring Reads}
 
@@
 
-\myparagraph{Redundant Read Elimination}
+\subsubsection*{Redundant Read Elimination}
 
@@
 
-\myparagraph{Termination Conditions and Parallel Composition}
+\subsubsection*{Termination Conditions and Parallel Composition}
 
@@
 
-\myparagraph{Read-Modify-Write Actions}
+\subsubsection*{Read-Modify-Write Actions}
 
@@
 
-\myparagraph{Data Race Freedom}
+\subsubsection*{Data Race Freedom}
 
@@
 
-\myparagraph{Augmentation of Preconditions}
+\subsubsection*{Augmentation of Preconditions}
 \jjr{} allows arbitrary augmentation of preconditions.  Here we are more

### 2021-11-01 paper/results.tex
accepted!

@@
 
-Prop.~6.1 of \cite{DBLP:journals/pacmpl/JagadeesanJR20} establishes a
+Prop.~6.1 of \citet{DBLP:journals/pacmpl/JagadeesanJR20} establishes a
 compositional principle for proving that programs validate formula in

### 2021-11-08 paper/appendix.tex
accepted!
supplementary material
make appendices a combo

@@
 \RequirePackage{amssymb}  %% bizarre error previous def of Bbbk if after documentclass
-\documentclass[acmsmall,screen,anonymous,review]{acmart}\settopmatter{printfolios=true}
+\documentclass[acmsmall,screen]{acmart}\settopmatter{printfolios=true}
 \hypersetup{bookmarksnumbered,bookmarksopen=true,bookmarksdepth=3}
@@
 \showSCOPEfalse
-\title{The Leaky Semicolon}
+\title{The Leaky Semicolon (Supplementary Material)}
 
@@
 \begin{document}
+\setcounter{page}{31}
 \appendix
+\noindent
+{\LARGE \textsf{\textbf{The Leaky Semicolon (Supplementary Material)}}}
+\bigskip
+
+\input{arm.tex}
 \input{oopsla.tex}
 \input{discussion.tex}
-\input{arm.tex}
+\input{extra.tex}
 \input{drf.tex}
-\input{extra.tex}
 \bibliography{bib}
 \end{document}
-
-Alex Aiken PLDI keynote:
-\begin{itemize}
-\item PAD Machines: Parallel, Accelerated (GPUs..), Distributed (memory)
-\item ``Registered'' memory: shared with network
-\item ``Zero-copy'' memory: shared with GPUs
-  \begin{itemize}
-  \item
-    \url{http://docs.nvidia.com/cuda/cuda-c-programming-guide/index.html#um-unified-memory-programming-hd}
-    Unified Memory offers a “single-pointer-to-data” model that is
-    conceptually similar to CUDA’s zero-copy memory. One key difference
-    between the two is that with zero-copy allocations the physical location
-    of memory is pinned in CPU system memory such that a program may have
-    fast or slow access to it depending on where it is being accessed
-    from. Unified Memory, on the other hand, decouples memory and execution
-    spaces so that all data accesses are fast.
-  \end{itemize}
-\item Software manages ``NUMA domains''
-\item In PAD machines, most computation happening on the accelerators.
-\item Bad Programming models:
-  \begin{itemize}
-  \item Data Analytics, Machine Learning: Single program in Spark Hadoop,
-    Etc.  Accessible to many people, Low performance
-  \item HPC: MPI, OpenMP, Vector Intrinsics, Cuda, etc...  Experts only, High performance
-  \end{itemize}
-\item Task based models are good.
-  \begin{itemize}
-  \item Asynchronous functions that get mapped to real resources.  May have
-    sub-tasks or not.  
-  \item Tasks capture locality:
-    \begin{itemize}
-    \item Task is co-located with its arguments.  
-    \item Tasks are a ``local address space'' model.
-    \end{itemize}
-  \item Tasks capture asynchrony
-  \end{itemize}
-\item Semantics
-  \begin{itemize}
-  \item \xmark Explicit Parallelism: X10, Chapel
-  \item \cmark Implicit Parallelism (dataflow) --- much easier to reason about
-    (sequential semantics) --- automated data movement
-    \begin{itemize}
-    \item Layout Explicitly programmed: StarPU 
-    \item Layout computed statically: PaRSEC/PTG, DPJ, Sequoia
-    \item Layout computed dynamically: Legion, PaRSEC/DTD, TensorFlow
-    \end{itemize}
-  \end{itemize}
-\item Partitioning:
-  \begin{itemize}
-  \item Flat disjoint partitions: TensorFlow, PyTorch.
-    \begin{itemize}
-    \item Not compositional: Problem is that different parts of the
-      application may want different views of the data.
-    \end{itemize}
-  \end{itemize}
-\item Hierarchical: StarPU
-  \begin{itemize}
-  \item Partitions created/destroyed
-  \item Partitions are disjoint
-  \item Only one partition of collection at a time
-  \item Only leaves can be used by application
-  \item Eager Repartitioning move data unnecessarily
-  \end{itemize}
-\item Legion allows multiple hierarchical partitions
-  \begin{itemize}
-  \item Compositional: data movement is lazy
-  \item Task dependence analysis requires alias analysis
-  \end{itemize}
-\end{itemize}
 

### 2021-11-10 paper/abstract.tex
edit sections 1-3

@@
 \endinput
+
+Program logics and semantics tell a pleasant story about sequential composition: when executing (S1;S2), we first execute S1 then S2.
+To improve performance, however, processors execute instructions out of order, and compilers reorder programs even more dramatically.
+By design, single-threaded systems cannot observe these reorderings; however, multiple-threaded systems can, making the story considerably less pleasant.
+A formal attempt to understand the resulting mess is known as a “relaxed memory model.” Prior models either fail to address sequential composition directly, or overly restrict processors and compilers, or permit nonsense thin-air behaviors which are unobservable in practice.
+
+To support sequential composition while targeting modern hardware, we enrich the standard event-based approach with preconditions and families of predicate transformers.
+When calculating the meaning of (S1; S2), the predicate transformer applied to the precondition of an event e from S2 is chosen based on the set of events in S1 upon which e depends.
+We apply this approach to two existing memory models.

### 2021-11-10 paper/intro.tex
edit sections 1-3

@@
 acyclicity axioms that determine which reads are valid.  These axioms use
-existentially defined global relations, such as memory order $\rmo$, which
+existentially defined global relations, such as memory order ($\rmo$), which
 must be a per-location total order on write actions.
@@
 sequential dependencies in \textsection\ref{sec:overview} and provide formal
-definitions in \textsection\ref{sec:model}.  We extend the model to include
-additional features, such as address calculation and \RMW{}s, in
-\textsection\ref{sec:additional}.  We discuss related and future work in
+definitions in \textsection\ref{sec:model}.  In
+\textsection\ref{sec:additional}, we extend the model to include
+additional features, such as address calculation and \RMW{}s.  We discuss related and future work in
 \textsection\ref{sec:related}--\ref{sec:outro}.
@@
 preconditions, composing the predicate transformers.
-\begin{equation}
-\begin{gathered}
+\begin{gather*}
   \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}(s{-}1)} \FI
+  \tag{*}\label{eq1}
   \\
@@
   \end{tikzinline}}
-\end{gathered}
-\label{eq1}
-\end{equation}
+\end{gather*}
 This works for the sequentially consistent case, but needs to be
@@
 associate to the left---with the intermediate step as in the diagram above,
-eliding the write action---
+eliding the write action---or to the right---with the intermediate step:
 % \begin{gather*}
@@
 % \end{gather*}
-or to the right---with the intermediate step:
 \begin{gather*}

### 2021-11-10 paper/outro.tex
add Simon and Mark's funding acknowledgement
pass over intro
pass through 9-10

@@
 This paper is the first to present a direct denotational semantics for
-sequential composition in a relaxed memory model that can be efficiently
-compiled to modern CPUs. We extract from this model a semantic dependency
-relation and use it to build \PwTc, a solution to the Out-of-Thin-Air problem
-in \cXI, and \PwTmca{}, a model intended for safe languages such as Java and
-Javascript.  Our work has several limitations, providing opportunities for
-future work.
+sequential composition that can be efficiently compiled to modern \CPU{}s.  We
+defined two models: \PwTc{} solves the out-of-thin-air problem for \cXI, and
+\PwTmca{} solves it for safe languages such as Java and Javascript.
 
-We have mechanized some proofs, but not all.  In particular, we have only a
-pen-and-paper proof showing that \PwTmca{} supports optimal lowering to
-\armeight{}.  The same is true for local data race-freedom (\ldrfsc{}).
-Additionally, our proof sketch for \ldrfsc{} elides \RMW{}s, which have
-caused complications in other models \cite{promising-ldrf}.
+Our work has several limitations, providing opportunities for future work:
 
-We have not treated loops, although we expect that the usual approach of
-showing continuity for all the semantic operations with respect to set
-inclusion would go through. \citet{DBLP:conf/esop/PaviottiCPWOB20} use
-step-indexing to account for loops; perhaps this approach could be adapted.
+\PwTc{} can be lowered efficiently to any architecture supported by \cXI{},
+but inherits the top-level axioms of \rcXI{}, compromising compositionality.
+\PwTmca{} is as a compositional as a model of concurrent imperative
+programming can be, but is limited to \mca{} architectures for optimal
+lowering.  It would be interesting to explore the middle ground to find a
+fully compositional model that supports optimal lowering to all modern
+architectures.
 
-\PwTmca{} does not validate access elimination: store-forwarding and
-dead-write-removal are unsound.  We expect that these can be validated by
-allowing events with different actions to merge. % in some instances.
-Nor does \PwTmca{} validate the reverse inclusions for
-\reflem{lem:if}\eqref{lem:ifelse:if:if}. The culprit is $\rdelay$, which
-introduces order regardless of whether preconditions are disjoint.  As an
-example, using augmentation, 
+As mentioned in \textsection\ref{sec:related}, some safety guarantees may be
+violated when read-introduction is followed by if-introduction, branching on
+the read just introduced.  Nonetheless, read-introduction is ubiquitous in
+some compilers \cite{DBLP:conf/pldi/LeeKSHDMRL17}.  It would be interesting
+to know the cost of restricting this optimization.
+%  
+In a similar vein, \PwTmca{1} is a simpler model than \PwTmca{2}, but
+requires fences on acquiring reads for \armeight{}.  It would be illuminating
+to find out what the performance penalty is for these fences.
+
+We have defined the soundness of compiler optimizations in the model, rather
+than contextually: $\aCmd'$ is a sound refinement of $\aCmd$ if
+$\sem{\aCmd'}\subseteq\sem{\aCmd}$.  This approach has several
+advantages---for example, it is immediate that a sound optimization is sound
+in any context.  It also has a disadvantage: some optimizations complicate
+the semantics.  For example, \PwTmca{} does not validate access elimination,
+such as store-forwarding and dead-write-removal---consider that complete
+executions of
+\begin{math}
+  \sem{\PW{x}{1}\SEMI\PR{x}{r}}
+\end{math}
+must include a read action and that complete executions of
+\begin{math}
+  \sem{\PW{x}{1}\SEMI\PW{x}{2}}
+\end{math}
+must include two write actions.  As another example, \PwTmca{} does not
+validate the reverse inclusions for
+\reflem{lem:if}\eqref{lem:ifelse:if:if}---consider that
 \begin{math}
@@
 \end{math}
-has an execution with
+has an augmented (\reflem{lem:augment}) execution with
 \begin{math}
@@
 \end{math}
-has no such execution.  We expect that this can be remedied by encoding
-$\rdelay$ in the logic.
+has no such execution.  We expect that these optimizations can be validated,
+at the cost of complicating the semantics.  For access elimination, it is
+likely sufficient to allow events with different actions to merge.  For
+\reflem{lem:if}\eqref{lem:ifelse:if:if}, it is likely sufficient to encoding
+$\rdelay$ in the logic---the problem in the execution above is that $\rdelay$
+introduces order regardless of whether preconditions are disjoint.
 %\todo{Comment on \PwTpo{} and \PwTc{}?}
 
-\PwTmca{1} is a simpler model than \PwTmca{2}, but requires fences on
-acquiring reads for \armeight{}.  It would be illuminating to find out what
-the performance penalty is for these fences.  In a similar vein, it would be
-interesting to know if there is a performance penalty for banning
-read-introduction, which is not valid for \PwT{}.
+We have not treated loops, although we expect that the usual approach of
+showing continuity for all the semantic operations with respect to set
+inclusion would go through. \citet{DBLP:conf/esop/PaviottiCPWOB20} use
+step-indexing to account for loops; perhaps this approach could be adapted.
 
-\PwTc{} can be lowered efficiently to any architecture supported by \cXI{},
-but inherits the top-level axioms of \rcXI{}, compromising compositionality.
-\PwTmca{} is as a compositional as a model of concurrent imperative
-programming can be, but is limited to \mca{} architectures for optimal
-lowering.  It would be interesting to explore the middle ground to find a
-fully compositional model that supports optimal lowering to all modern
-architectures.
+While we have mechanized some proofs, most of our proofs are informal.
+In particular, we have only a pen-and-paper proof showing that \PwTmca{}
+supports optimal lowering to \armeight{}.  The same is true for local data
+race-freedom (\ldrfsc{})---additionally, our proof sketch for \ldrfsc{}
+elides \RMW{}s, which have caused complications in other models
+\cite{promising-ldrf}.
 
@@
 This paper has been greatly improved by the comments of the anonymous reviewers.
-
+%
 James Riely was supported by the National Science Foundation under
 grant No.~CCR-1617175.
-
-\todo{Add support for Mark\&Simon.}
-
+%
+Mark Batty and Simon Cooksey were supported by the EPSRC under grant
+Nos.~EP/V000470/1 and~EP/R032971/1, and by VeTSS.
+%
 Anton Podkopaev was supported by JetBrains Research.

### 2021-11-10 paper/related.tex
accepted!
pass through 9-10

@@
 disallow compiler optimizations and efficient implementation, typically
-requiring fences for every relaxed read on Arm
+requiring fences for every relaxed read on \armeight{}
 \cite{Lamport:1979:MMC:1311099.1311750, DBLP:conf/pldi/LahavVKHD17,
@@
 % %
-% - The speculative models allow thin air behaviors.
+% - The speculative models allow thin-air behaviors.
 % %
@@
 by adopting a sub-optimal lowering for \RMW{}s.  \PwT{} does not suffer from
-this problem, since \PwT{} does not validate read-introduction.  Nonetheless,
-read-introduction is ubiquitous in some compilers
-\cite{DBLP:conf/pldi/LeeKSHDMRL17}.  There appears to be a genuine tension
+this problem, since \PwT{} does not validate read-introduction.  There appears to be a genuine tension
 between temporal reasoning, as supported by \PwT{}, and read-introduction, as
@@
 Unfortunately, it is not obvious that tools can be built for all
-thin-air-free models, the calculation of
+thin-air-free models: the calculation of
 \citet{Pichon-Pharabod:2016:CSR:2837614.2837616} does not have a termination
-proof for an arbitrary input, and the enormous state space for the
+proof for an arbitrary input; the enormous state space for the
 operational models of \citet{DBLP:conf/popl/KangHLVD17} and
-\citet{DBLP:journals/pacmpl/ChakrabortyV19} is a daunting prospect for a tool
-builder -- and as yet no tool exists for automatically evaluating these
-models.  We described a tool, \PwTer, for automatically evaluating \PwT{} in
+\citet{DBLP:journals/pacmpl/ChakrabortyV19} is daunting for a tool
+builder---and as yet no tool exists for automatically evaluating these
+models.  We described a tool for automatically evaluating \PwT{} in
 \textsection\ref{sec:tool}.

### 2021-11-10 paper/mca1.tex
removing todos. Integrated consistency and causal strengtheining.
accepted!
pass through 4-5

@@
 relaxed access on \armeight{}, but requires extra synchronization for
-acquiring reads.\footnote{\citet{DBLP:journals/pacmpl/JagadeesanJR20}
-  erroneously elide the required synchronization on acquiring reads.}
-(\emph{Lowering} is the translation of language-level operators to machine
-instructions.  A lowering is \emph{optimal} if it provides the most efficient
-execution possible.)
+acquiring reads.  (\emph{Lowering} is the translation of language-level
+operators to machine instructions.  A lowering is \emph{optimal} if it
+provides the most efficient execution possible.)
 
@@
 We show an example at the beginning of \textsection\ref{sec:mca2}.
-
-The lowering proofs can be found in \cite{appendix}.  The proofs
+%
+The lowering proofs can be found in \refAppendix{sec:arm}.  The proofs
 use recent alternative characterizations of \armeight{} \cite{armed}.
@@
 
+  \label{def:pwt:mca:complete}
   \noindent
-  \label{def:pwt:mca:complete}
+  \begin{minipage}{1.0\linewidth}
     The definition of completeness extends \refdef{def:pomset} as follows:
@@
     \end{multicols}
+  \end{minipage}
+  \smallskip
 
   \noindent
+  \begin{minipage}{1.0\linewidth}
     The semantic function extends \reffig{fig:seq} as follows:
     \begin{multicols}{2}
-    \begin{enumerate}[topsep=0pt,label=(\textsc{i}\arabic*),ref=\textsc{i}\arabic*]
-    \item[{\labeltext[\textsc{p}7]{(\textsc{p}7)}{par-rf}}]
-      \labeltext[\textsc{s}7]{(\textsc{s}7)}{seq-rf}\;
-      \labeltext[\textsc{i}7]{(\textsc{i}7)}{if-rf}\;
-      ${\rrfx}\rextends{\rrfx_1}{\rrfx_2}$,
+      \begin{enumerate}[topsep=0pt,label=(\textsc{i}\arabic*),ref=\textsc{s}\arabic*]
         \setcounter{enumi}{\value{le}}
@@
         \item \label{seq-le-delays}
-        if $\labeling_1(\bEv) \rdelays \labeling_2(\aEv)$ then $\bEv\le\aEv$.
+          if $\labeling_1(\bEv) \rdelays \labeling_2(\aEv)$ then $\bEv\le\aEv$,
         \end{enumerate}
+      \item[{\labeltext[\textsc{p}7]{(\textsc{p}7)}{par-rf}}]
+        \labeltext[\textsc{s}7]{(\textsc{s}7)}{seq-rf}\;
+        \labeltext[\textsc{i}7]{(\textsc{i}7)}{if-rf}\;
+        ${\rrfx}\rextends{\rrfx_1}{\rrfx_2}$.
       \end{enumerate}  
     \end{multicols}
+  \end{minipage}
+  \smallskip
   % The semantic rules are as follows:
@@
 % \end{math}
-We write $\semmca{1}{}$ for the semantic function when it is unclear from context.
+%We write $\semmca{1}{}$ for the semantic function when it is unclear from context.
 
-In complete pomsets, $\rrfx$ must pair every read with a matching write
-\eqref{top-rf}.  The requirements \ref{pom-rf-match}, \ref{pom-rf-block}, and
-\ref{pom-rf-le} guarantee that reads are \emph{fulfilled}, as in
-\cite[\textsection 2.7]{DBLP:journals/pacmpl/JagadeesanJR20}.
+In complete pomsets, reads-from ($\rrfx$) must pair every read with a
+matching write \eqref{top-rf}.  The requirements \ref{pom-rf-match},
+\ref{pom-rf-block}, and \ref{pom-rf-le} guarantee that reads are
+\emph{fulfilled}, as in \cite[\textsection
+2.7]{DBLP:journals/pacmpl/JagadeesanJR20}.  Parallel composition, sequential
+composition, and the conditional respect reads-from (\ref{par-rf},
+\ref{seq-rf}, \ref{if-rf}).
 
-The semantic rules are mostly straightforward: Parallel composition is disjoint
-union, and all constructs respect reads-from.  The monoid laws
-(\reflem{lem:monoid}) extend to parallel composition, with $\SKIP$ as right
-unit only due to the asymmetry of \ref{par-tau}.
+% The semantic rules are mostly straightforward: Parallel composition is disjoint
+% union, and all constructs respect reads-from.  The monoid laws
+% (\reflem{lem:monoid}) extend to parallel composition, with $\SKIP$ as right
+% unit only due to the asymmetry of \ref{par-tau}.
 
-Only \ref{seq-le-delays} requires explanation.
+%Only \ref{seq-le-delays} requires explanation.
 % \begin{lemma}
@@
 between SC accesses ($\eqreordersc$).  Combined with the fulfillment
-requirements (\ref{pom-rf-match}, \ref{pom-rf-block} and \ref{pom-rf-le}),
+requirements (\ref{pom-rf-match}, \ref{pom-rf-block}, \ref{pom-rf-le}),
 these ensure coherence, publication, subscription and other idioms.  For

### 2021-11-10 paper/mca2.tex
pass through 4-5

@@
 
-Lowering \PwTmca{1} to \armeight{} requires a full fence after every
-acquiring read.  To see why, consider the following attempted
-execution, where the final values of both $x$ and $y$ are $2$.
+Lowering \PwTmca{1} to \armeight{} requires a full fence before every
+acquiring read.\footnote{\citet{DBLP:journals/pacmpl/JagadeesanJR20}
+  erroneously elide the required synchronization on acquiring reads.}  To see
+why, consider the following attempted execution, where the final values of
+both $x$ and $y$ are $2$.
 \begin{gather*}

### 2021-11-10 paper/outro.tex
pass through 4-5

@@
 validate the reverse inclusions for
-\reflem{lem:if}\eqref{lem:ifelse:if:if}---consider that
+\reflem{lem:if}\ref{lem:ifelse:if:if}---consider that
 \begin{math}
@@
 likely sufficient to allow events with different actions to merge.  For
-\reflem{lem:if}\eqref{lem:ifelse:if:if}, it is likely sufficient to encoding
+\reflem{lem:if}\ref{lem:ifelse:if:if}, it is likely sufficient to encoding
 $\rdelay$ in the logic---the problem in the execution above is that $\rdelay$

### 2021-11-10 paper/tool.tex
pass through 6-7

@@
 
-\PwTer{} automatically and exhaustively calculates the allowed outcomes of litmus tests for the \PwT, \PwTpo, and \PwTc{} models.
-It is built in OCaml, and uses Z3~\cite{Z3Solver} to judge the truth of predicates constructed by the models.
-\PwTer{} obviates the need for error-prone hand evaluation.
+\PwTer{} automatically and exhaustively calculates the allowed outcomes of litmus tests for the \PwT, \PwTpo, and \PwTc{} models, obviating the need for error-prone hand evaluation.
+It is built in OCaml, using Z3~\cite{Z3Solver} to judge the truth of predicates.
+% constructed by the models.
 
 \PwTer{} allows several modes of evaluation: it can evaluate the rules of Fig.~\ref{fig:seq}, implementing \PwT; it can generate program order according to \textsection\ref{sec:c11}, implementing \PwTpo; and similar to \MRD~\cite{DBLP:conf/esop/PaviottiCPWOB20}, it can construct C11-style pre-executions and filter them according to the rules of \rcXI{} as described in \textsection\ref{sec:c11}, implementing \PwTc{}.
-Finally, \PwTer{} also allows us to toggle the complete check of~\ref{def:complete}, providing an interface for understanding how fragments of code might compose by exposing preconditions and termination conditions that are not yet tautologies.
+Finally, \PwTer{} also allows us to toggle the complete check of~\refdef{def:complete}, providing an interface for understanding how fragments of code might compose by exposing preconditions and termination conditions that are not yet tautologies.
+
 We have run \PwTer{} over the Java Causality Tests~\cite{PughWebsite} supported in the input syntax, and tabulated the results in Table~\ref{fig:tool}.
 For context, we have included the results of \MRD{} for the Java Causality tests~\cite{DBLP:conf/esop/PaviottiCPWOB20}.
-Of note is that test cases 17 and 18 for \MRD{} and \MRDRC{} do not give the correct outcome, this is for similar reasons to the example given in \textsection\ref{sec:lir}, where local invariant reasoning in \MRD{} is too constrained.
+Note that \MRD{} and \MRDRC{} do not give the correct outcome on jctc17--18---the reason is that local invariant reasoning in \MRD{} is too constrained (see \textsection\ref{sec:lir}).
 
-The execution times give a good indication the poor scaling of the tool with program size: for larger test cases, the tool takes exponentially longer to compute, and for the largest tests the memory footprint is too large for even a well-equipped computer.
+%The execution times give a good indication of the poor scaling of the tool with program size: 
+For larger test cases, the tool takes exponentially longer to compute, and for the largest tests the memory footprint is too large for even a well-equipped computer.
 The compositional nature of the semantics makes tool building practical, but it is not enough to make it scalable for large tests.
-The definitions of $\sSEMI{\aPSS_1}{\aPSS_2}$ and $\sIF{\aForm}\sTHEN\aPSS_1\sELSE\aPSS_2\sFI$ -- in combination with the rules for reads and writes has exponential complexity.
+In combination with the rules for reads and writes, the definitions of $\sSEMI{\aPSS_1}{\aPSS_2}$ and $\sIF{\aForm}\sTHEN\aPSS_1\sELSE\aPSS_2\sFI$ have exponential complexity.
 This is compounded by the hidden complexity of calculating the possible merges between pomsets through union in rules \ref{seq-E} and \ref{if-E}.
 Significant effort has been put into throwing away spurious merges early in \PwTer, so that executing the tool remains manageable for small examples.
-Some further optimisations may be possible within the tool to improve the situation further, such as killing `dead-end' pomsets at each sequence operator, or by doing a directed search for particular execution outcomes.
-\PwTer{} is available with this paper's supplementary materials.
+Some further optimizations may be possible within the tool to improve the situation further, such as killing ``dead-end'' pomsets at each sequence operator, or by doing a directed search for particular execution outcomes.
+\PwTer{} is available in the supplementary material.
 

### 2021-11-10 paper/tool.tex
pass through 8

@@
 For context, we have included the results of \MRD{} for the Java Causality tests~\cite{DBLP:conf/esop/PaviottiCPWOB20}.
-Note that \MRD{} and \MRDRC{} do not give the correct outcome on jctc17--18---the reason is that local invariant reasoning in \MRD{} is too constrained (see \textsection\ref{sec:lir}).
+Note that \MRD{} and \MRDRC{} do not give the correct outcome on TC17--18---the reason is that local invariant reasoning in \MRD{} is too constrained (see \textsection\ref{sec:lir}).
 

### 2021-11-12 paper/outro.tex
typo in 10

@@
 likely sufficient to allow events with different actions to merge.  For
-\reflem{lem:if}\ref{lem:ifelse:if:if}, it is likely sufficient to encoding
+\reflem{lem:if}\ref{lem:ifelse:if:if}, it is likely sufficient to encode
 $\rdelay$ in the logic---the problem in the execution above is that $\rdelay$
-introduces order regardless of whether preconditions are disjoint.
+introduces order even when the preconditions are disjoint.
 %\todo{Comment on \PwTpo{} and \PwTc{}?}

### 2021-12-30 paper/intro.tex
Fixing typos

@@
   These models require inefficient implementation of relaxed access.  This is
-  a non-starter for safe languages like Java and Javascript, and may be an
+  a non-starter for safe languages like Java and JavaScript, and may be an
   unacceptable cost for low-level languages like \cXI{}.
-\item Semantic dependencies \cite{%
-    Manson:2005:JMM:1047659.1040336, DBLP:conf/esop/JagadeesanPR10,
-    DBLP:conf/popl/KangHLVD17, DBLP:journals/pacmpl/ChakrabortyV19,
-    DBLP:conf/pldi/LeeCPCHLV20, promising-ldrf}. These models compute
+\item Semantic dependencies \cite{Manson:2005:JMM:1047659.1040336,DBLP:conf/esop/JagadeesanPR10,DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}.
+These models compute
   dependencies operationally using alternate worlds, making it impossible to
@@
 \item No dependencies, as in \cXI{} \cite{DBLP:conf/esop/BattyMNPS15} and
-  Javascript \cite{DBLP:journals/pacmpl/WattRP19}.  This allows thin-air
+  JavaScript \cite{DBLP:journals/pacmpl/WattRP19}.  This allows thin-air
   executions.
@@
 (\MRD), which use event structures to calculate a semantic dependency
-relation ($\rsdep$).  \PwP{} is defined using (acyclic) labelled partial
+relation ($\rsdep$).  \PwP{} is defined using (acyclic) labeled partial
 orders, or \emph{pomsets} \cite{GISCHER1988199}.  \MRD{} adds a causality
@@
 % \end{itemize}
-% % Acknowledgements go here, once we're not double-blinded.
+% % Acknowledgments go here, once we're not double-blinded.
 % % The definitions in this paper have been formalized in Agda.
@@
 % should guarantee sequential abstraction.  But processor and language
-% implementors balk at the prospect, due to significant costs.
+% implementers balk at the prospect, due to significant costs.
 
@@
 \end{gather*}
-We can represent such an execution as a labelled partial order, or
+We can represent such an execution as a labeled partial order, or
 \emph{pomset} \cite{DBLP:conf/lop/Pratt85,GISCHER1988199}.  A program may give rise to many
@@
   \begin{gathered}
-    \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
+    \IF{s} \THEN \PW{z}{r{*}(s{-}1)} \FI
     \\
     \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
-        \event{wz0}{(s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{}
+        \event{wz0}{(s{\neq}0) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{}
         \xform{wz0d}{\bForm[r{*}s/z]}{right=of wz0}
@@
 \begin{gather*}
-  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}(s{-}1)} \FI
+  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s} \THEN \PW{z}{r{*}(s{-}1)} \FI
   \tag{*}\label{eq1}
@@
       \event{ry0}{\DR{y}{1}}{right=of rx0}
-      \event{wz0}{(1{=}r) \limplies (1{=}s) \limplies (s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
+      \event{wz0}{(1{=}r) \limplies (1{=}s) \limplies (s{\neq}0) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
       \xform{rx0ry0d}{(1{=}r) \limplies(1{=}s) \limplies \bForm[r{*}(s{-}1)/z]}{right=of wz0}
@@
   % \begin{gathered}
-  %   \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
+  %   \IF{s} \THEN \PW{z}{r{*}s} \FI
   %   \\
   %   \hbox{\begin{tikzinline}[node distance=1ex and1.5em]
-  %     \event{wz0}{(s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{}
+  %     \event{wz0}{(s{\neq}0) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{}
   %     \xform{wz0i}{\bForm}{left=.4em of wz0}
@@
 \begin{gather*}
-  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}(s{-}1)} \FI
+  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s} \THEN \PW{z}{r{*}(s{-}1)} \FI
   \\
@@
       \event{ry0}{\DR{y}{1}}{above right=of rx0ry0d}
-      \event{wz0}{(1{=}s) \limplies (s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
+      \event{wz0}{(1{=}s) \limplies (s{\neq}0) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
       \po{ry0}{wz0}
@@
 \begin{gather*}
-  \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}(s{-}1)} \FI
+  \PR{y}{s}\SEMI \IF{s} \THEN \PW{z}{r{*}(s{-}1)} \FI
   \\
@@
       \event{ry0}{\DR{y}{1}}{}
-      \event{wz0}{(1{=}s) \limplies (s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
+      \event{wz0}{(1{=}s) \limplies (s{\neq}0) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
       \xform{ry0d}{(1{=}s) \limplies \bForm}{left=of ry0}

### 2021-12-30 paper/intro.tex
another typo

@@
   dependencies operationally using alternate worlds, making it impossible to
-  understood a single execution in isolation; they also allow executions that
+  understand a single execution in isolation; they also allow executions that
   violate temporal reasoning (see \textsection\ref{sec:related}).
